### 2018-11-13 paper/paper_PE.tex

\documentclass[aps,showpacs,twocolumn,prd,superscriptaddress,nofootinbib]{revtex4}

\usepackage{amsmath}
\usepackage{amsfonts}
\usepackage{amssymb}
\usepackage{latexsym}
\usepackage{graphicx}
\usepackage{bm}
%\usepackage{color}
\usepackage{enumerate}
\usepackage{ulem}

\usepackage{color}
\usepackage[usenames,dvipsnames,svgnames,table]{xcolor}
%\usepackage[colorlinks=true,
%            linkcolor=green,
%            urlcolor=blue,
%            citecolor=red]{hyperref}
\usepackage[colorlinks=true,
            linkcolor=YellowOrange,
            urlcolor=RoyalBlue,
            citecolor=RedViolet]{hyperref}

\newcommand{\be}{\begin{equation}}
\newcommand{\ee}{\end{equation}}
\newcommand{\bsub}{\begin{subequations}}
\newcommand{\esub}{\end{subequations}}
\newcommand\ud{{\mathrm{d}}}
\newcommand\uD{{\mathrm{D}}}
\newcommand\calO{{\mathcal{O}}}
\newcommand\bfx{\mathbf{x}}
\newcommand{\ov}[1]{\overline{#1}}
\newcommand{\ph}[1]{\phantom{#1}}
\newcommand{\cte}{\mathrm{cte}}
\newcommand{\nn}{\nonumber}
\newcommand{\hatk}{\hat{k}}
\newcommand{\Hz}{\,\mathrm{Hz}}
\newcommand{\sinc}{\,\mathrm{sinc}}
\newcommand{\Msol}{M_{\odot}}
\newcommand\calL{{\mathcal{L}}}
\newcommand\calA{{\mathcal{A}}}
\newcommand\betaL{{\beta_{L}}}
\newcommand\lambdaL{{\lambda_{L}}}
\newcommand\varphiL{{\varphi_{L}}}
\newcommand\psiL{{\psi_{L}}}

\newcommand{\SM}[1]{{\color{Blue} #1}}
\newcommand{\jgb}[1]{{\color{DarkGreen} #1}}

\begin{document}

\title{Bayesian methods for black hole merger parameter estimation with LISA.}

\author{John G. Baker}
\affiliation{Gravitational Astrophysics Laboratory, NASA Goddard Space Flight Center, 8800 Greenbelt Rd., Greenbelt, MD 20771, USA}
\author{Sylvain Marsat}
\affiliation{Department of Physics, University of Maryland, College Park, MD 20742, USA}
\affiliation{Gravitational Astrophysics Laboratory, NASA Goddard Space Flight Center, 8800 Greenbelt Rd., Greenbelt, MD 20771, USA}
\affiliation{Max Planck Institute for Gravitational Physics (Albert Einstein Institute), Am M\"uhlenberg 1, Potsdam-Golm, 14476, Germany}

\date{\today}

\begin{abstract}

[Abstract]

\end{abstract}

\pacs{
04.70.Bw, % classical black holes
04.80.Nn, % Gravitational wave detectors and experiments
95.30.Sf, % relativity and gravitation
95.55.Ym, % Gravitational radiation detectors
97.60.Lf  % black holes (astrophysics)
}

\maketitle

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

\section{Introduction}
\label{sec:intro}

[Introduction]

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

\section{Methodology}
\label{sec:intro}

\subsection{Fast frequency domain LISA response}
\label{sec:response}

-- Geometric setting, definitions, summarize the FD response

\subsection{Reduced order model for EOBNRv2 waveforms.}
\label{sec:waveforms}

-- Waveforms: EOBNRv2HMROM + PNext

\subsection{Fast likelihood computation}
\label{sec:likelihood}

-- Likelihood computation: fast overlaps

\subsection{Fisher matrix parameter estimation}
\label{sec:Fisher}

-- Fisher matrices computation: step self-tuning,...

\subsection{Bayesian sampling}
\label{sec:samplers}

-- Bayesian samplers: Multinest, PTMCMC

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

\section{Morphology of the signals}
\label{sec:morph}

- Morphology of the response

\subsection{Temporal development of SNR}
\label{sec:timeSNR}

-- SNR accumulation with time or frequency: inspiral/MRD balance across mass range

\subsection{LISAframe parameters}
\label{sec:LISAframe}

-- LISA frame angles

\subsection{The low-frequency limit}
\label{sec:low-freq}

-- make the connection to low-f response (2 interferometers)

\subsection{A simplified likelihood model}
\label{sec:simple-like}

In this section, we investigate a simplified form of the extrinsic likelihood and the structure of its degeneracies.

It will be useful to introduce the following notation for the patter functions associated to the independent channels A and E:
\bsub\label{eq:defDaDe}
\begin{align}
	D_{a}^{+} &= \frac{1}{2} \left( 1 + \sin^{2}\betaL \right) \cos\left( 2\lambdaL - \frac{\pi}{3} \right) \,, \\
	D_{a}^{\times} &= \sin \betaL \sin\left( 2\lambdaL - \frac{\pi}{3} \right) \,, \\
	D_{e}^{+} &= - \frac{1}{2} \left( 1 + \sin^{2}\betaL \right) \sin\left( 2\lambdaL - \frac{\pi}{3} \right) \,, \\
	D_{e}^{\times} &= \sin \betaL \cos\left( 2\lambdaL - \frac{\pi}{3} \right) \,.
\end{align}
\esub
Note that the expressions for the channel E can be obtained from the expressions for A with the replacement $\lambdaL \rightarrow \lambdaL + \pi/4$.

In each channel, we separate the contribution of the modes $h_{22}$ and $h_{2,-2}$ as
\bsub\label{eq:defsase}
\begin{align}
	s_{a} &= a_{22} + a_{2,-2} \,, \\
	s_{e} &= e_{22} + e_{2,-2} \,,
\end{align}
\esub
with
\bsub\label{eq:defa22a2m2}
\begin{align}
	a_{22} &= \frac{3i}{4d} \sqrt{\frac{5}{\pi}} \cos^{4}\frac{\iota}{2} e^{2i(-\varphiL-\psiL)} \left( D_{a}^{+} + i D_{a}^{\times} \right) \,, \\
	a_{2,-2} &= \frac{3i}{4d} \sqrt{\frac{5}{\pi}} \sin^{4}\frac{\iota}{2} e^{2i(-\varphiL+\psiL)} \left( D_{a}^{+} - i D_{a}^{\times} \right) \,.
\end{align}
\esub
and similarly for the E channel with~\eqref{eq:defDe} instead of~\eqref{eq:defDa}
\bsub\label{eq:defe22e2m2}
\begin{align}
	e_{22} &= \frac{3i}{4d} \sqrt{\frac{5}{\pi}} \cos^{4}\frac{\iota}{2} e^{2i(-\varphiL-\psiL)} \left( D_{e}^{+} + i D_{e}^{\times} \right) \,, \\
	e_{2,-2} &= \frac{3i}{4d} \sqrt{\frac{5}{\pi}} \sin^{4}\frac{\iota}{2} e^{2i(-\varphiL+\psiL)} \left( D_{e}^{+} - i D_{e}^{\times} \right) \,.
\end{align}
\esub
Here, we introduced the dimensionless luminosity distance ratio $d \equiv D / D^{\rm inj}$.

In this simplified response, an exact symmetry relation is manifest: changing simultaneously
\be\label{eq:symmetryresponse}
\betaL \rightarrow -\betaL\,, \quad \iota \rightarrow \pi - \iota \,, \quad \psiL \rightarrow \pi - \psiL
\ee
leaves the signal unchanged. We can also rewrite the likelihood in a way that makes clearer the symmetry between the two pairs of parameters $(\iota, \varphiL)$ and $(\betaL, \lambdaL)$ \SM{[TODO]}.

The simplified extrinsic likelihood for a frozen LISA then takes the form
\be\label{eq:simplelikelihood}
	\ln \calL = - \frac{1}{2} \Lambda \left( \left| s_{a} - s_{a}^{\rm inj} \right|^{2} + \left| s_{e} - s_{e}^{\rm inj} \right|^{2} \right) \,,
\ee
where $\Lambda$ appears as an overall normalization factor, playing the role of the square SNR of the signal.

We can now explore possible degeneracies in the likelihood given by~\eqref{eq:simplelikelihood}, \eqref{eq:defsase} and \eqref{eq:defa22a2m2}-\eqref{eq:defe22e2m2}. We note that the relative contributions of the modes $h_{2,2}$ and $h_{2,-2}$ are strongly separated in the face-on or face-off limit, leading to a simplification of the response and an enhanced degeneracy. Due to the symmetry~\eqref{eq:symmetryresponse}, we will focus on the face-on limit $\iota \rightarrow 0$. In this limit, the $h_{2,-2}$ contributions die off rapidly as $a_{2,-2}\,, e_{2,-2} \sim \sin^{4} (\iota/2)$. If they are assumed to be negligible, then we can rewrite
\bsub\label{eq:saseapprox}
\begin{align}
	s_{a} \simeq i\calA e^{2 i \xi} \left( D_{a}^{+} + i D_{a}^{\times} \right) \,, \\
	s_{e} \simeq i\calA e^{2 i \xi} \left( D_{e}^{+} + i D_{e}^{\times} \right) \,,
\end{align}
\esub
where we introduced the notations $\xi (\varphiL, \psiL) \equiv -\varphiL - \psiL$ and $\calA(d, \iota) \equiv 3/4\sqrt{5/\pi}\cos^{4}(\iota/2)/d $ for the common phase and amplitude factors.

For a degeneracy to occur, both complex quantities $s_{a}$ and $s_{e}$ must be close to the injection values $s_{a}^{\rm inj}$ and $s_{e}^{\rm inj}$, which can be achieved as follows if~\eqref{eq:saseapprox} is valid. Defining the complex ratio
\be\label{eq:defr}
	r(\lambdaL, \betaL) = \frac{D_{a}^{+} + i D_{a}^{\times}}{D_{e}^{+} + i D_{e}^{\times} } \,,
\ee
if the position in the sky $(\lambda_{L}^{*}, \beta_{L}^{*})$ can be chosen so that
\be\label{eq:rcondition}
	r(\lambda_{L}^{*}, \beta_{L}^{*}) = \frac{s_{a}^{\rm inj}}{s_{e}^{\rm inj}} \,,
\ee
then adjusting $\calA$ and $\xi$ so that
\be
	i\calA e^{2 i \xi} = \frac{s_{a}^{\rm inj}}{\left( D_{a}^{+} + i D_{a}^{\times} \right)(\lambda_{L}^{*}, \beta_{L}^{*})}
\ee
generates degenerate points in parameter space with $\ln \calL \simeq 0$.

The conditions~\eqref{eq:defr}-\eqref{eq:rcondition} can be inverted for the sky position as follows. For a given complex value of $r$, $r(\lambda_{L}^{*}, \beta_{L}^{*}) = r$ can be rewritten as
\be\label{eq:eqtosolveforlambdabetaL}
	e^{4i\lambda_{L}^{*}}e^{i\frac{\pi}{3}} \frac{\left( 1 +\sin\beta_{L}^{*} \right)^{2}}{\left( 1 - \sin\beta_{L}^{*} \right)^{2}} = \frac{1+ i r}{1 - i r} \,.
\ee
Solving for the modulus of~\eqref{eq:eqtosolveforlambdabetaL} then yields
\be\label{eq:solutionbetaL}
	\sin\beta_{L}^{*} = \frac{\rho - 1}{\rho + 1} \,,
\ee
for
\be
	\rho = \sqrt{\left| \frac{1+ i r}{1 - i r} \right|} \,.
\ee
Solving for the argument of~\eqref{eqtosolveforlambdabetaL} gives four solutions for $\lambda_{L}^{*}$ as
\be\label{eq:solutionlambdaL}
	\lambda_{L}^{*} = - \frac{\pi}{12} + \frac{1}{4}\mathrm{Arg} \frac{1+ i r}{1 - i r} + \frac{k \pi}{2} \quad \mathrm{for} \quad k = 0,\dots,3 \,.
\ee

Repeating the argument for the other branch of solutions $\iota \rightarrow \pi$ (or directly considering the symmery~\eqref{eq:symmetryresponse}) gives four other solutions in the sky with identical sky longitudes $\lambda_{L}^{*}$, with symmetrized sky latitudes $\beta_{L}^{*} \rightarrow -\beta_{L}^{*}$, and with $\xi (\varphiL, \psiL) \equiv -\varphiL + \psiL$, $\calA(d, \iota) \equiv 3/4\sqrt{5/\pi}\sin^{4}(\iota/2)/d $.

Thus, with~\eqref{eq:solutionbetaL}, \eqref{eq:solutionlambdaL} and $\beta_{L}^{*} \rightarrow -\beta_{L}^{*}$, we arrive at eight degenerate positions in the sky $(\lambda_{L}^{*}, \beta_{L}^{*})$. For each of these sky positions, an exact line degeneracy exists between $\varphiL$ and $\psiL$ as long as $\xi = \mathrm{const}$, as well as an approximate line degeneracy between $d$ and $\iota$, as long as $\calA = \mathrm{const}$, limited to inclination values where $\sin^{4}(\iota/2)$ (or $\cos^{4}(\iota/2)$, for the other branch of solutions) remains negligible. The combination of these degeneracies generates an extended structure in the multi-dimensional parameter space. As a result, when considering the marginal posterior for the sky position, these degenerate spots in the sky can have a lot of probability support.

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

\section{Illustrating examples}
\label{sec:examples}

- Demonstration/examples

\subsection{A non-degenerate case}
\label{sec:high-M-non-deg}

-- High-mass system, non-degenerate, 22 and HM + Fisher - stacked SNR contours
eg:

Case 12?
m1,    m2,    t0,D,     phi0,i,   lam,  bet, psi
8.89e5,1.11e5,0, 3.67e4,pi/3,pi/2,3pi/4,pi/3,pi/3

Case 0

-- Performance: waveform and likelihood costs, number of sampler evaluations

\subsection{Stellar origin black holes}
\label{sec:SOBH}

-- SOBH system, SNR sufficient for clean analysis + Fisher - stacked SNR contours
-- Performance: waveform and likelihood costs, number of sampler evaluations

\subsection{Highlighting degeneracies}
\label{sec:degen}

- Highlighting degeneracies

-- High-mass degenerate example, 22 vs HM - stacked SNR contours
-- show degenerate waveforms
-- simplified response and closed-form likelihood, full degeneracies
-- (explore simplified response further: parameter maps, other off-the-shelf samplers ?)

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

\section{Discussion}
\label{sec:discussion}

- Summary/conclusions

----------------
Other questions:
- special orientations/sky locations ?
- Fisher matrix qualification - explore parameter choices
- continuously increase SNR - include on these examples
- continuously accumulate information over time
- full exploration of parameter space for all of the above...

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

\end{document}

### 2018-11-15 data/t-series-mcmc-results/skyplot.py
Basic analysis of t-series.

@@ -1,0 +1,18 @@
+import numpy as np
+import pylab as pl
+
+fig = pl.figure()
+ax = fig.add_subplot(1, 1, 1, projection='hammer')
+
+labels = ['40 h', '2.5 h', '7 m', 'Post-merger']
+
+for i, label in zip([64, 16, 4, 1], labels):
+    data = np.loadtxt('STP_09_lm55_t{:02d}_joint_resampled.dat'.format(i))
+    ra, dec = data[:,8], data[:,9]
+    ra -= np.pi
+
+    ax.plot(ra, dec, 'o', markeredgewidth=0, markersize=2, label=label)
+
+ax.grid()
+ax.legend()
+pl.savefig('skymap.png', dpi=200)

### 2018-11-16 data/t-series-mcmc-results/skyplot.py
Skymap plot with contours

@@ -1,17 +1,39 @@
 import numpy as np
 import pylab as pl
+from sklearn.neighbors.kde import KernelDensity
+
 
 fig = pl.figure()
 ax = fig.add_subplot(1, 1, 1, projection='hammer')
 
 labels = ['40 h', '2.5 h', '7 m', 'Post-merger']
+colors = ['g', 'b', 'r', 'k']
+bandwidths = [0.2, 0.05, 0.01, 0.005]
 
-for i, label in zip([64, 16, 4, 1], labels):
+for i, label, color, bw in zip([64, 16, 4, 1], labels, colors, bandwidths):
+    print(i)
     data = np.loadtxt('STP_09_lm55_t{:02d}_joint_resampled.dat'.format(i))
     ra, dec = data[:,8], data[:,9]
     ra -= np.pi
 
-    ax.plot(ra, dec, 'o', markeredgewidth=0, markersize=2, label=label)
+    # create the KDE estimator
+    # we could use grid-search cross validation to estimate the bandwidth here
+    radec = np.vstack((dec, ra)).T
+    kde = KernelDensity(kernel='tophat', bandwidth=bw, metric='haversine').fit(radec)
+
+    # find the density levels corresponding to 1, 2, 3 sigmas
+    n = 10000
+    sample = kde.sample(n)
+    sample_densities = np.sort(np.exp(kde.score_samples(sample)))
+    levels = [sample_densities[int(n * (1 - p))] for p in [0.9973, 0.9545, 0.6827]]
+
+    gra = np.linspace(-np.pi, np.pi, 300)
+    gdec = np.linspace(-np.pi/2, np.pi/2, 300)
+    ggra, ggdec = np.meshgrid(gra, gdec)
+    d = np.exp(kde.score_samples(np.vstack((ggdec.ravel(), ggra.ravel())).T))
+    d = np.reshape(d, ggra.shape)
+    cs = ax.contour(gra, gdec, d, colors=color, linewidths=[0.5, 1, 1.5], levels=levels)
+    cs.collections[0].set_label(label)
 
 ax.grid()
 ax.legend()

### 2018-11-20 paper/paper_PE.tex
Paper : reorganized sections
Paper : pushed TODOs for plots in README and changed section structure.
Paper : small edit.

@@
 %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
 
@@
 %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
 
-\section{Methodology}
-\label{sec:intro}
-
-\subsection{Fast frequency domain LISA response}
+\section{Instrument response}
 \label{sec:response}
-
--- Geometric setting, definitions, summarize the FD response
-
-\subsection{Reduced order model for EOBNRv2 waveforms.}
-\label{sec:waveforms}
-
--- Waveforms: EOBNRv2HMROM + PNext
-
-\subsection{Fast likelihood computation}
-\label{sec:likelihood}
-
--- Likelihood computation: fast overlaps
-
-\subsection{Fisher matrix parameter estimation}
-\label{sec:Fisher}
-
--- Fisher matrices computation: step self-tuning,...
-
-\subsection{Bayesian sampling}
-\label{sec:samplers}
-
--- Bayesian samplers: Multinest, PTMCMC
 
@@
 
-\section{Morphology of the signals}
-\label{sec:morph}
+\subsection{The LISA response}
+\label{sec:lisaresponse}
 
-- Morphology of the response
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
 
-\subsection{Temporal development of SNR}
-\label{sec:timeSNR}
-
--- SNR accumulation with time or frequency: inspiral/MRD balance across mass range
-
-\subsection{LISAframe parameters}
+\subsection{LISA-frame parameters}
 \label{sec:LISAframe}
@@
 -- LISA frame angles
+
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
 
@@
 -- make the connection to low-f response (2 interferometers)
+
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
 
@@
 \esub
-and similarly for the E channel with~\eqref{eq:defDe} instead of~\eqref{eq:defDa}
+and similarly for the E channel with $D_{e}^{+,\times}$ instead of $D_{a}^{+, \times}$ as
 \bsub\label{eq:defe22e2m2}
@@
 where $\Lambda$ appears as an overall normalization factor, playing the role of the square SNR of the signal.
+
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+
+\section{Methodology}
+\label{sec:method}
+
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+
+\subsection{Frequency domain LISA response}
+\label{sec:response}
+
+-- Geometric setting, definitions, summarize the FD response
+
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+
+\subsection{Reduced order model for EOBNRv2 waveforms.}
+\label{sec:waveforms}
+
+-- Waveforms: EOBNRv2HMROM + PNext
+
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+
+\subsection{Likelihood computation}
+\label{sec:likelihood}
+
+-- Likelihood computation: fast overlaps
+
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+
+\subsection{Fisher matrix parameter estimation}
+\label{sec:Fisher}
+
+-- Fisher matrices computation: step self-tuning,...
+
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+
+\subsection{Bayesian sampling}
+\label{sec:samplers}
+
+-- Bayesian samplers: Multinest, PTMCMC
+
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+
+\section{Morphology of the signals}
+\label{sec:morph}
+
+- Morphology of the response
+
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+
+\subsection{Signals and transfer functions}
+\label{sec:signaltransfer}
+
+-- SNR accumulation with time or frequency: inspiral/MRD balance across mass range
+
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+
+\subsection{Time and frequency accumulation of SNR}
+\label{sec:timefreqSNR}
+
+-- SNR accumulation with time or frequency: inspiral/MRD balance across mass range
+
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+
+\section{Supermassive black holes}
+\label{sec:SMBH}
+
+- Demonstration/examples
+
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+
+\subsection{A non-degenerate case}
+\label{sec:SMBHPEnondeg}
+
+-- High-mass system, non-degenerate, 22 and HM + Fisher - stacked SNR contours
+eg:
+
+Case 12?
+m1,    m2,    t0,D,     phi0,i,   lam,  bet, psi
+8.89e5,1.11e5,0, 3.67e4,pi/3,pi/2,3pi/4,pi/3,pi/3
+
+Case 0
+
+-- Performance: waveform and likelihood costs, number of sampler evaluations
+
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+
+\subsection{Accumulating signal with time}
+\label{sec:SMBHPEacctime}
+
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+
+\subsection{SNR dependence of parameter estimation}
+\label{sec:SMBHPEdistance}
+
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+
+\subsection{Highlighting degeneracies}
+\label{sec:SMBHPEdegen}
+
+- Highlighting degeneracies
+
+-- High-mass degenerate example, 22 vs HM - stacked SNR contours
+-- show degenerate waveforms
+-- simplified response and closed-form likelihood, full degeneracies
+-- (explore simplified response further: parameter maps, other off-the-shelf samplers ?)
 
@@
 
-The conditions~\eqref{eq:defr}-\eqref{eq:rcondition} can be inverted for the sky position as follows. For a given complex value of $r$, $r(\lambda_{L}^{*}, \beta_{L}^{*}) = r$ can be rewritten as
+The conditions~\eqref{eq:defr}-\eqref{eq:rcondition} can be inverted for the sky position as follows.
+For a given complex value of $r = s_{a}^{\rm inj} / s_{e}^{\rm inj}$, $r(\lambda_{L}^{*}, \beta_{L}^{*}) = r$ can be rewritten as
 \be\label{eq:eqtosolveforlambdabetaL}
@@
 \ee
-Solving for the argument of~\eqref{eqtosolveforlambdabetaL} gives four solutions for $\lambda_{L}^{*}$ as
+Solving for the argument of~\eqref{eq:eqtosolveforlambdabetaL} gives four solutions for $\lambda_{L}^{*}$ as
 \be\label{eq:solutionlambdaL}
@@
 %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
 
-\section{Illustrating examples}
-\label{sec:examples}
-
-- Demonstration/examples
-
-\subsection{A non-degenerate case}
-\label{sec:high-M-non-deg}
-
--- High-mass system, non-degenerate, 22 and HM + Fisher - stacked SNR contours
-eg:
-
-Case 12?
-m1,    m2,    t0,D,     phi0,i,   lam,  bet, psi
-8.89e5,1.11e5,0, 3.67e4,pi/3,pi/2,3pi/4,pi/3,pi/3
-
-Case 0
-
--- Performance: waveform and likelihood costs, number of sampler evaluations
-
-\subsection{Stellar origin black holes}
+\section{Stellar origin black holes}
 \label{sec:SOBH}
@@
 -- Performance: waveform and likelihood costs, number of sampler evaluations
-
-\subsection{Highlighting degeneracies}
-\label{sec:degen}
-
-- Highlighting degeneracies
-
--- High-mass degenerate example, 22 vs HM - stacked SNR contours
--- show degenerate waveforms
--- simplified response and closed-form likelihood, full degeneracies
--- (explore simplified response further: parameter maps, other off-the-shelf samplers ?)
 

### 2018-11-28 paper/paper_PE.tex
Work on the paper.

@@
-\documentclass[aps,showpacs,twocolumn,prd,superscriptaddress,nofootinbib]{revtex4}
+\documentclass[aps,showpacs,twocolumn,prd,superscriptaddress,nofootinbib]{revtex4-1}
 
@@
 \usepackage{ulem}
+\usepackage{tabularx}
+\usepackage{braket}
 
@@
 \newcommand\calO{{\mathcal{O}}}
+\newcommand\calM{{\mathcal{M}}}
+\newcommand\calF{{\mathcal{F}}}
+\newcommand\calT{{\mathcal{T}}}
+\newcommand\calD{{\mathcal{D}}}
+\newcommand\calA{{\mathcal{A}}}
+\newcommand\calL{{\mathcal{L}}}
 \newcommand\bfx{\mathbf{x}}
@@
 \newcommand{\Msol}{M_{\odot}}
-\newcommand\calL{{\mathcal{L}}}
-\newcommand\calA{{\mathcal{A}}}
+\newcommand{\Mchirp}{M_{c}}
 \newcommand\betaL{{\beta_{L}}}
@@
 \newcommand\psiL{{\psi_{L}}}
+\newcommand{\tf}{t_{f}}
+\newcommand{\Tf}{T_{f}}
+\newcommand{\tfd}{t_{f}^{d}}
+\newcommand{\tfSPA}{t_{f}^{\rm SPA}}
+\newcommand{\TfSPA}{T_{f}^{\rm SPA}}
+
+\newcolumntype{C}[1]{>{\centering\arraybackslash}p{#1}}
+\newcolumntype{L}[1]{>{\raggedright\arraybackslash}p{#1}}
 
@@
 
-\section{Instrument response}
+\section{The LISA instrument response}
 \label{sec:response}
@@
 
-\subsection{The LISA response}
+\subsection{The LISA response - OLD}
 \label{sec:lisaresponse}
+
+\SM{------------------------------------------------------}
+
+\SM{[copy-paste from previous paper -- just here to copy needed formulas / refs]}
+
+We begin by detailing the model that we use for the response of a LISA-like detector, together with the assumptions used and their limitations.
+Since our aim is to assess the accuracy of our direct Fourier-domain treatment of the response, we can focus on the gravitational-wave contribution to the basic single-link observables.
+We therefore use a somewhat simplified model for the time-domain response~\cite{Krolak+04}, ignoring corrections that would be crucial from the point of view of noise cancellations, keeping in mind that the response model can be enriched later without affecting the conclusions of the present analysis.
+
+The frequency-shift response for a single link~\eqref{eq:yslr} was derived in Ref.~\cite{EW75} (see also~\cite{CR02, RCP04, Finn08, Cornish09}).
+Several assumptions enter the result as written in~\eqref{eq:yslr}: (i) effects of the order $v/c$ are neglected, including for instance the special relativistic Doppler effect created by the relative speeds of the spacecraft on their orbits (ii) the propagation is assumed to take place in a flat spacetime, perturbed only by the gravitational wave; thus the gravitational redshift as well as the deflection of light created by the gravitational potential of the Sun is ignored (iii) all geometric factors are evaluated at a single time, whereas one should consider the beam as propagating from the position of the first spacecraft at the time of emission to the position of the second scapecraft at the time of reception, leading to a point-ahead effect.
+
+Additionally, we limit ourselves to a rigid model for the orbits of the constellation, namely we assume that the constellation remains in an equilateral configuration with fixed armlengths.
+These simplified orbits neglect (iv) effects of order $e^{2}$ from the eccentricity of the individual Keplerian orbits, (v) the effect of gravitational perturbations coming from other celestial bodies, such as the Earth, the quadrupole of the Sun, and the other planets.
+Note that although we can neglect all the effects (i)-(v) for our present study, keeping track of these corrections is crucial for the purpose of laser noise cancellations, and led to the development of new generations of TDI observables~\cite{Tintoliving}.
+
+It is natural to split the response~\eqref{eq:yslr} into two steps: first the orbital delay related to the orbit around the Sun of the whole constellation, and then the constellation response.
+The baselines for the delays are indeed very different in the two cases.
+Geometrical projection factors aside, we have for the orbit around the Sun $R=1\text{au}=1.5\times 10^{8} \text{km}$, while the detector armlength is $L=2.5\times 10^{6}\text{km}$ (in the configuration proposed in~\cite{LISA17}).
+It is natural to define two transfer frequencies for the two relevant length scales for the delays, defined such that a wavelength fits within this length scale, i.e.
+$2\pi f d = 1$.
+This gives
+\begin{subequations}\label{eq:transferfrequencies}
+\begin{align}
+	f_{R} &= 3.2\times10^{-4}\Hz \,,\\
+	f_{L} &= 1.9\times 10^{-2}\Hz \,.
+\end{align}
+\end{subequations}
+The LISA response will behave qualitatively differently on the three frequency bands $f \leq f_{R}$, $f_{R} \leq f \leq f_{L}$ and $f_{L} \leq f$.
+
+The first stage of the response, the orbital delay, consists simply in applying the varying time delay to bring the wavefront sampling point from the SSB reference to the center of the LISA triangular constellation, common to all $y_{slr}$ observables.
+For $h^{\rm TT}$ the transverse-traceless gravitational waveform in matrix form, we write this orbital time delay as
+\be\label{eq:defresponse0}
+	h_{0}^{\rm TT} (t) = h^{\rm TT}(t-\hatk\cdot p_{0}) \,,
+\ee
+with $p_{0}$ the position of the constellation center, which follows the Earth orbit around the Sun.
+The second stage of the response calculation comprises the remaining, constellation-centered response.
+For the single-link contribution to the response, for the laser link from spacecraft $s$ to spacecraft $r$ along a path in direction $n_l$, we write
+\begin{align}\label{eq:defresponseL}
+	y_{slr} &= \frac{1}{2} \frac{1}{1 - \hatk\cdot n_{l}} \nn\\
+	& \cdot n_{l}\cdot \left[ h_{0}^{\rm TT}(t - L - \hatk\cdot p^{L}_{s}) - h_{0}^{\rm TT}(t - \hatk\cdot p^{L}_{r}) \right] \cdot n_{l}\,,
+\end{align}
+where we reference the positions of the spacecraft relative to the center of the constellation, $p^{L}_{A} \equiv p_{A} - p_{0}$.
+
+As described in App.~\ref{app:notation} We will decompose the full signal in the contributions of the individual spin-weighted spherical modes $h_{\ell m}$, whose Fourier transforms are assumed to have a smooth amplitude and phase.
+First, we define the matrices $P_{+},P_{\times}$ such that, in the sense of matrices,
+\be
+	h^{\rm TT} = h_{+}P_{+} + h_{\times}P_{\times} \,.
+\ee
+We focus only on positive frequencies.
+Assuming that the approximation~\eqref{eq:zeronegativef} applies, we consider a single mode contribution, $h=h_{\ell m}$ with $m>0$.
+For each given mode we define a complex matrix $P_{\ell m}$ incorporating the spin-weighted spherical harmonic constant factor as
+\be
+	P_{\ell m} =
+	\begin{cases}
+	\frac{1}{2} {}_{-2}Y_{\ell m} \left( P_{+} + i P_{\times} \right) \text{ for } m>0\,,\\
+	\frac{1}{2} {}_{-2}Y_{\ell m}^{*} \left( P_{+} - i P_{\times} \right) \text{ for } m<0\,.
+	\end{cases}
+\ee
+
+We now turn to the transformation of~\eqref{eq:defresponse0} and~\eqref{eq:defresponseL} to the Fourier domain. Applying a pure delay as in~\eqref{eq:defresponse0} translates into
+\be\label{eq:G0}
+	G_{0}(f, t) = e^{-2i\pi f d_{0}(t)} \,,
+\ee
+with $d_{0} = -\hatk \cdot p_{0}$ the delay associated to the orbit around the Sun.
+For the leading-order response~\eqref{eq:transferlocal}, this gives a Fourier-domain transfer function common to all modes, that is a pure phase factor, proportional to the frequency but also $t_{f}$-dependent:
+\be\label{eq:transfer0local}
+	\calT_{0}^{\rm local}(f) = G_{0}(f, \tf)\,.
+\ee
+If $(\lambda, \beta)$ are the ecliptic longitude and latitude of the source in the sky, and if the orbital phase is set by convention to $0$ at $t=0$, the orbital delay has the simple expression
+\be\label{eq:delay0}
+	d_{0}(t) = -R \cos\beta \cos\left(\Omega_{0}t - \lambda\right)\,.
+\ee
+
+For the constellation part of response~\eqref{eq:defresponseL}, treated separately from the delay~\eqref{eq:delay0}, we write
+\begin{align}\label{eq:decomposeGslr}
+	F_{slr}^{L}(t) &= \frac{1}{2} \frac{1}{1 - \hatk\cdot n_{l}(t)} n_{l}(t) \cdot P_{\ell m} \cdot n_{l} (t) \,,\nn\\
+	d_{s}(t) &= - k\cdot p_{s}^{L}(t) \,, \quad d_{r}(t) = - k\cdot p_{r}^{L}(t) \,,\nn\\
+	G_{slr}^{L}(f,t) &=  F_{slr}^{L}(t) \left( e^{-2i\pi f (d_{s,L}(t) + L)} - e^{-2i\pi f d_{r}(t)} \right) \,.
+\end{align}
+The superscript $L$ indicates that the orbital delay~\eqref{eq:delay0} is not included.
+Since we also assume the rigid approximation for the constellation, where the armlengths are fixed, a particular simplification occurs when combining these individual delays, thanks to the relation $p^{L}_{r} - p^{L}_{s} =  L n_{l}$:
+\begin{align}\label{eq:GslrL}
+	G_{slr}^{L}(f,t) &= \frac{i \pi f L}{2} \sinc \left[ \pi f L\left(1-\hatk\cdot n_{l} \right) \right] \nn\\
+	& \quad \cdot \exp\left[ i \pi f \left( L + \hatk\cdot \left( p_{1}^{L} + p_{2}^{L} \right) \right) \right]  n_{l} \cdot P_{\ell m} \cdot n_{l} \,,
+\end{align}
+with all time-dependent vectors evaluated at $t$.
+This expression is well known as describing the frequency-dependency in the LISA response~\cite{Larson+99, Cornish01, CR02, RCP04}.
+In the local approximation~\eqref{eq:transferlocal}, the Fourier-domain transfer function then reads
+\begin{align}\label{eq:transferLlocal}
+	\calT_{slr}^{L, \mathrm{local}}(f) &= G_{slr}^{L}(f, \tf) \,.
+\end{align}
+For plotting purposes, we will also define
+\be\label{eq:transferLenvelope}
+	\overline{\calT}_{slr}^{L} (f) = \frac{i \pi f L}{2} n_{l} \cdot P_{\ell m} \cdot n_{l} (\tf)
+\ee
+which will serve as an estimate for the enveloppe function of the response, devoid of the zero-crossings at high frequencies of the $\sinc$ term in~\eqref{eq:GslrL}.
+
+Note that, if the corrections of Sec.~\ref{subsec:delays}, for non-negligible $\dot d$, are included for the constellation delays, the transfer function will not have this simple form anymore, as $\dot{d}$ will have a different velocity-dependent expression for the sending and receiving spacecraft.
+One must then separately handle $d_{s}$ and $d_{r}$ in~\eqref{eq:decomposeGslr} to compute the corrections.
+
+The orbital response~\eqref{eq:transfer0local}-\eqref{eq:delay0} takes a simple analytic form, but the phase contribution of this delay is significant across most of the frequency band and can be large for $f \gg f_{R}$.
+
+The constellation response~\eqref{eq:GslrL}-\eqref{eq:transferLlocal} can be interpreted as the Fourier-domain translation of a discrete derivative taken on the waveform.
+The leading factor in~\eqref{eq:GslrL} shows that the amplitude of the response is proportional to $f$ in the low-frequency limit $f\ll f_{L}$, where the other factors are essentially unity.
+For $f\gtrsim f_{L}$, the $\sinc$ and the phase of the exponential generate additional structure in the response, including zero-crossings when the projected armlength is an integer number of wavelengths.
+From~\eqref{eq:transferLlocal}, an expansion for small $f\ll f_{L}$ yields back a Fourier-domain analog of the low-frequency approximation of the response~\cite{Cutler97, RCP04}, which is equivalent to having two LIGO-type interferometers turned by $\pi/4$ and set in motion.
+
+For analysis of the response we need a concrete set of gravitational waveforms.
+We will use the PhenomD model~\cite{Khan+15,Husa+15}, which provides Fourier-domain inspiral-merger-ringdown waveforms for aligned spins.
+We refer to App.~\ref{app:precLISA} for a brief discussion of the prospects for applying our formalism for the LISA response to precessing Fourier-domain waveforms.
+
+\SM{------------------------------------------------------}
 
@@
 
-\subsection{LISA-frame parameters}
+\subsection{The GW signal}
+\label{sec:gwsignal}
+
+We start by introducing a source-frame $(\hat{x}_{S}, \hat{y}_{S}, \hat{z}_{S})$ attached to the binary system emitting gravitational waves.
+For comparable-mass systems without spin, the natural choice is to take the normal to the orbital plane as $\hat{z}_{S}$, and we will assume that the remaining rotation around $\hat{z}_{S}$ is fixed by the phase convention of the waveform model (see Sec.~\ref{}).
+
+Introducing $k$ the wave propagation vector going from the source towards the observer, we define the inclination $\iota$ and the observer phase $\varphi$ simply as being its spherical angular coordinates $(\theta_{S}, \phi_{S})$, so that in the source-frame
+\be
+	k_{S} = (\sin\iota \cos\varphi, \sin\iota \sin\varphi, \cos\iota) \,.
+\ee
+
+By convention, for our polarization vectors $p$ and $q$ we will use the spherical coordinate vectors $p = e_{\theta}^{S}$, $q = e_{\phi}^{S}$. In terms of only $\hat{z}_{S}$,
+\bsub
+\begin{align}
+	p &= q \times k \,,\\
+	q &= \hat{z}_{S} \times k / |\hat{z}_{S} \times k|\,.
+\end{align}
+\esub
+If $H_{ij} = h_{ij}^{\rm TT}$ represents the gravitational wave signal in the transverse-traceless gauge,
+\be
+	H = h_{+} P_{+} + h_{\times} P_{\times}
+\ee
+with the polarization tensors
+\bsub
+\begin{align}
+	P_{+} &= p \otimes p - q \otimes q \,,\\
+	P_{\times} &= p \otimes q + q \otimes p \,.
+\end{align}
+\esub
+Conversely, the polarizations are
+\bsub
+\begin{align}
+	h_{+} &= \frac{1}{2} \left(p \otimes p - q \otimes q \right) : H \,,\\
+	h_{\times} &= \frac{1}{2} \left( p \otimes q + q \otimes p \right) : H \,.
+\end{align}
+\esub
+
+We will further decompose the gravitational wave signal, seen as a function of the direction of emission $(\iota, \varphi)$, in spin-weighted spherical harmonics as
+\be
+	h_{+} - i h_{\times} = \sum_{\ell \geq 2} \sum_{m = -\ell}^{\ell} {}_{-2}Y_{\ell m} (\iota, \varphi) h_{\ell m} \,,
+\ee
+where the explicit expression of the ${}_{-2}Y_{\ell m}$ is given in~\cite{}.
+
+The vectors $(p,q,k)$ form the Wave-frame, such that the gravitational wave takes in this frame the familiar form
+\be
+	H_{W} = \begin{pmatrix}
+		h_{+} & h_{\times} & 0 \\
+		h_{\times} & -h_{+} & 0 \\
+		0 & 0 & 0
+		\end{pmatrix} \,.
+\ee
+
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+
+\subsection{LISA trajectories}
+\label{sec:lisatraj}
+
+For completeness, we gather in this appendix explicit expressions for the trajectory of the LISA constellation and the rotation matrices entering the response.
+
+The orbits of the three spacecrafts around the Sun can be chosen so that, at leading order in the eccentricity of the orbits, the constellation retains the shape of an equilateral triangle in a cartwheeling motion following the Earth orbit.
+We denote by $a$ the semi-major axis, $e$ the eccentricity and pose $\alpha = \Omega_{0} (t-t_{0}))$, with $t_{0}$ a reference time for the initial position ($t_{0} = 0$ in our case).
+The reference SSB frame is $(\hat{x}, \hat{y}, \hat{z})$.
+
+Using the notation $c,s = \cos, \sin \alpha$, the trajectory of the center of the constellation is simply
+\be
+	p_{0} = ac \hat{x} + as \hat{y} \,,
+\ee
+and we take $a\equiv R = 1 \, \mathrm{au}$.
+We denote the positions of the spacecrafts as $p_{A}$ for $A=1,2,3$, and the positions relative to the constellation center as $p_{A}^{L} = p_{A} - p_{0}$.
+Setting $\beta_{A} = 2(A-1)\pi/3 + \beta_{0}$ (with $\beta_{0}$ and initial condition set to 0 in our case), the Cartesian coordinates in the SSB-frame of the position of the spacecrafts read~\cite{}
+\bsub
+\begin{align}
+	p_{A}^{L} &= a e \left[ \sin \beta_{A} c s - \cos\beta_{A} \left( 1 + s^{2} \right) \right] \hat{x} \nn\\
+	& + a e \left[ \cos \beta_{A} c s - \sin\beta_{A} \left( 1 + c^{2} \right) \right] \hat{y} \nn\\
+	& - a e \sqrt{3} \cos(\alpha - \beta_{A}) \hat{z} \,,
+\end{align}
+\esub
+The armlength, constant in this approximation, is related to the eccentricity and semi-major axis by
+\be
+	L = 2\sqrt{3} a e \,.
+\ee
+The rigid approximation for the constellation, at first order in $e$, can be seen as being at first order in $L/R \simeq 0.017$ for a $2.5 \, \mathrm{Gm}$ armlength~\cite{}.
+
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+
+\subsection{The LISA frame}
 \label{sec:LISAframe}
 
--- LISA frame angles
+It will be very useful to introduce a time-dependent frame $(\hat{x}_{L}, \hat{y}_{L}, \hat{z}_{L})(t)$ following the detector in its motion.
+Specifically, we choose this LISA frame such that at any time,
+\bsub
+\begin{align}
+	p_{1}^{L} &= - \frac{L}{\sqrt{3}} \hat{x}_{L} \,,\\
+	p_{2}^{L} &= \frac{L}{2\sqrt{3}} \hat{x}_{L} - \frac{L}{2} \hat{y}_{L} \,,\\
+	p_{3}^{L} &= \frac{L}{2\sqrt{3}} \hat{x}_{L} + \frac{L}{2} \hat{y}_{L} \,.\\
+\end{align}
+\esub
+
+This frame also provides us with an equivalent representation of the trajectories making use of rotation matrices.
+Since in our rigid approximation the constellation remains an equilateral triangle in its cartwheeling motion around the Sun, the configuration of the constellation at a later time is given by a rotation around the Sun composed with a rotation of the constellation around its symmetry axis.
+
+If $R(v,\varpi)$ denotes the matrix of an active rotation around the vector $v$ by an angle $\varpi$, and if we denote by $R_{L}$ the active rotation from the SSB-frame to the L-frame such that for each of the three basis vectors $(\hat{x}_{L}, \hat{y}_{L}, \hat{z}_{L}) = R_{L} (\hat{x}, \hat{y}, \hat{z})$, we have
+\be\label{eq:RL}
+	R_{L} = R(z, \alpha) \cdot R(y, -\pi/3) \cdot R(z, -\alpha) \,,
+\ee
+where we recall that $\alpha = \Omega_{0} (t-t_{0}) $.
+For all vectors $X$ among $p_{A}^{L}$, $n_{A}$ we have $X(t) = R_{L} \cdot X(t=t_{0})$.
+If a vector $X$ is given by its components in the SSB-frame, in the L-frame the components are $X_{L} = R_{L}^{-1} \cdot X$.
+
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+
+\subsection{From the Wave frame to the SSB frame}
+\label{sec:wavessbframe}
+
+We can now relate the Wave-frame to the SSB-frame.
+The source position in the sky is given by $(\lambda, \beta)$, the ecliptic longitude and latitude in the SSB-frame.
+Like with the Source-frame, we can introduce spherical vectors $(e_{r}^{SSB}, e_{\theta}^{SSB}, e_{\phi}^{SSB})$.
+Since the propagation vector is $k = - e_{r}^{SSB}$,
+\be
+	k = (- \cos\beta\cos\lambda, - \cos\beta\sin\lambda, -\sin\beta) \,.
+\ee
+
+The last degree of freedom between the frames represents a rotation along the line-of-sight, parametrized by the polarization angle $\psi$.
+We introduce reference polarization vectors in the SSB-frame as
+\bsub
+\begin{align}
+	u &= \hat{z} \times k / |\hat{z} \times k| \,,\\
+	v &= k \times u \,.
+\end{align}
+\esub
+In terms of spherical vectors, $(k, u, v) = (-e_{r}^{SSB}, -e_{\phi}^{SSB}, -e_{\theta}^{SSB})$.
+The polarization angle $\psi$ is then defined such that $(p,q)$ are obtained by rotating $(u,v)$ by the angle $\psi$ around $k$, i.e.
+\be
+	(p,q) = R(k, \psi) \cdot (u,v) \,.
+\ee
+
+These relations can be summarized by the active rotation matrix from the SSB-frame to the Wave-frame $R_{W}$:
+\be\label{eq:RW}
+	R_{W} = R(z, \lambda - \pi/2) \cdot R(x, \beta + \pi/2 ) \cdot R(z, \psi)\,.
+\ee
+
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+
+\subsection{Parametrizing the signal in the LISA-frame}
+\label{sec:LISAframeparams}
+
+While long-lasting signals like SOBHs will see a strong imprint of the LISA orbital motion over the course of observation, SMBH signals are dominated in SNR by a short-lived burst of emission at merger.
+For such signals, it will be useful to use a parametrization based on the L-frame at the time of merger instead of the SSB frame.
+
+The new parameters are defined as playing the same role as the SSB parameters, but relative to the L-frame.
+Namely, we define $R_{LW}$ the active rotation matrix from the L-frame to the Wave-frame, expressed in the L-frame basis, so that for each basis vector expressed in the L-frame basis $(\hat{x}_{W}, \hat{y}_{W}, \hat{z}_{W})_{L} = R_{LW} \cdot (\hat{x}_{L}, \hat{y}_{L}, \hat{z}_{L})_{L}$.
+The defining condition on $(\lambda_{L}, \beta_{L}, \psi_{L})$ is
+\be\label{eq:defRLW}
+	R_{LW} = R(z, \lambda_{L} - \pi/2) \cdot R(x, \beta_{L} + \pi/2 ) \cdot R(z, \psi_{L}) \,.
+\ee
+
+Coming back to vectors expressed in the SSB-basis, $R_{L}^{-1} \cdot (\hat{x}_{W}, \hat{y}_{W}, \hat{z}_{W}) = R_{LW} \cdot R_{L}^{-1} \cdot (\hat{x}_{L}, \hat{y}_{L}, \hat{z}_{L})$.
+Since $ (\hat{x}_{W}, \hat{y}_{W}, \hat{z}_{W}) = R_{W} \cdot (\hat{x}, \hat{y}, \hat{z})$ and $ (\hat{x}_{L}, \hat{y}_{L}, \hat{z}_{L}) = R_{L} \cdot (\hat{x}, \hat{y}, \hat{z})$, we arrive at
+\be\label{eq:RLW}
+	R_{LW} = R_{L}^{-1} \cdot R_{W} \,.
+\ee
+
+Combining~\eqref{eq:RLW}, \eqref{eq:RL}, \eqref{eq:RW} and~\eqref{eq:defRLW} yields the following expressions for the L-frame angular parameters\footnote{The convention here is that the point of coordinates $(x,y)$ in the plane has for argument $\arctan\left[ x, y \right]$.}:
+\bsub
+\begin{align}
+	\beta_{L} &= \arcsin \left[ \cos \frac{\pi}{3} \sin \beta - \sin \frac{\pi}{3} \cos \beta \cos \left( \lambda - \alpha \right) \right] \,,\\
+	\lambda_{L} &= \arctan \left[ \cos\beta \cos\lambda \left( \cos \frac{\pi}{3} \cos^{2}\alpha + \sin^{2}\alpha \right) \right. \nn\\
+	& \qquad\quad\;\; \left. + \cos\beta \sin\lambda \cos\alpha \sin\alpha \left( \cos\frac{\pi}{3} - 1 \right) \right. \nn\\
+	& \qquad\quad\;\; \left. + \sin\frac{\pi}{3} \sin\beta \cos\alpha , \right. \nn\\
+	& \qquad\qquad\; \left. \cos\beta \sin\lambda \left( \cos \frac{\pi}{3} \sin^{2}\alpha + \cos^{2}\alpha \right) \right. \nn\\
+	& \qquad\quad\;\; \left. + \cos\beta \cos\lambda \cos\alpha \sin\alpha \left( \cos\frac{\pi}{3} - 1 \right) \right. \nn\\
+	& \qquad\quad\;\; \left. + \sin\frac{\pi}{3} \sin\beta \sin\alpha \right] \,,\\
+	\lambda_{L} &= \arctan \left[ \cos\psi \left( \cos\frac{\pi}{3}\cos\beta + \sin\frac{\pi}{3} \sin\beta \cos(\lambda - \alpha)\right) \right. \nn\\
+	& \qquad\quad\;\; \left. + \sin \psi \sin\frac{\pi}{3} \sin(\lambda - \alpha), \right. \nn\\
+	& \qquad\qquad\; \left. \sin\psi \left( \cos\frac{\pi}{3}\cos\beta + \sin\frac{\pi}{3} \sin\beta \cos(\lambda - \alpha)\right) \right. \nn\\
+	& \qquad\quad\;\; \left. - \cos \psi \sin\frac{\pi}{3} \sin(\lambda - \alpha) \right] \,.
+\end{align}
+\esub
+
+The definition of the reference phase $\phi_{\rm Ref}$, \dots
 
@@
 
-\subsection{A simplified likelihood model}
+\subsection{A simple extrinsic likelihood}
 \label{sec:simple-like}
@@
 \begin{align}
-	a_{22} &= \frac{3i}{4d} \sqrt{\frac{5}{\pi}} \cos^{4}\frac{\iota}{2} e^{2i(-\varphiL-\psiL)} \left( D_{a}^{+} + i D_{a}^{\times} \right) \,, \\
-	a_{2,-2} &= \frac{3i}{4d} \sqrt{\frac{5}{\pi}} \sin^{4}\frac{\iota}{2} e^{2i(-\varphiL+\psiL)} \left( D_{a}^{+} - i D_{a}^{\times} \right) \,.
+	a_{22} &= \frac{3i}{4d} \sqrt{\frac{5}{\pi}} \cos^{4}\frac{\iota}{2} e^{2i(-\varphi-\psiL)} \left( D_{a}^{+} + i D_{a}^{\times} \right) \,, \\
+	a_{2,-2} &= \frac{3i}{4d} \sqrt{\frac{5}{\pi}} \sin^{4}\frac{\iota}{2} e^{2i(-\varphi+\psiL)} \left( D_{a}^{+} - i D_{a}^{\times} \right) \,.
 \end{align}
@@
 \begin{align}
-	e_{22} &= \frac{3i}{4d} \sqrt{\frac{5}{\pi}} \cos^{4}\frac{\iota}{2} e^{2i(-\varphiL-\psiL)} \left( D_{e}^{+} + i D_{e}^{\times} \right) \,, \\
-	e_{2,-2} &= \frac{3i}{4d} \sqrt{\frac{5}{\pi}} \sin^{4}\frac{\iota}{2} e^{2i(-\varphiL+\psiL)} \left( D_{e}^{+} - i D_{e}^{\times} \right) \,.
+	e_{22} &= \frac{3i}{4d} \sqrt{\frac{5}{\pi}} \cos^{4}\frac{\iota}{2} e^{2i(-\varphi-\psiL)} \left( D_{e}^{+} + i D_{e}^{\times} \right) \,, \\
+	e_{2,-2} &= \frac{3i}{4d} \sqrt{\frac{5}{\pi}} \sin^{4}\frac{\iota}{2} e^{2i(-\varphi+\psiL)} \left( D_{e}^{+} - i D_{e}^{\times} \right) \,.
 \end{align}
@@
 %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+
+\appendix
+
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+
+\section{Numerical integration with spline-interpolated phases}
+\label{sec:numintegration}
+
+[describe implementation of 0-noise integrals]
+
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+
+\section{Reduced-order model for EOBNRv2HMROM}
+\label{sec:rom}
+
+[describe implementation of the ROM]
+
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+
+\bibliography{references.bib}
 

### 2018-12-04 paper/paper_PE.tex
Paper: work on the response.

@@
 \newcommand{\TfSPA}{T_{f}^{\rm SPA}}
+\newcommand{\sYlm}{{}_{-2}Y_{\ell m}}
+\newcommand{\sYlmstar}{{}_{-2}Y_{\ell, -m}^{*}}
+\newcommand{\sYlminusmstar}{{}_{-2}Y_{\ell, -m}^{*}}
 
@@
 We will further decompose the gravitational wave signal, seen as a function of the direction of emission $(\iota, \varphi)$, in spin-weighted spherical harmonics as
-\be
+\be\label{eq:hpcmodes}
 	h_{+} - i h_{\times} = \sum_{\ell \geq 2} \sum_{m = -\ell}^{\ell} {}_{-2}Y_{\ell m} (\iota, \varphi) h_{\ell m} \,,
@@
 \ee
+
+We will both generate the waveforms and apply the response directly in the Fourier domain. Our convention for the Fourier transform is
+\be\label{eq:defFourier}
+	\tilde{F}(f) = \int dt \; e^{2 i \pi f t} F(t) \,.
+\ee
+Note that this differs by a change $f\rightarrow -f$ from the most usual one (used e.g. in \texttt{LAL}).
+
+The definition of the reference time and phase \dots
 
@@
 
-We can now relate the Wave-frame to the SSB-frame.
+We can now relate the Wave-frame to the SSB-frame, which will fix our convention for the sky position and polarization angle.
 The source position in the sky is given by $(\lambda, \beta)$, the ecliptic longitude and latitude in the SSB-frame.
@@
 
-These relations can be summarized by the active rotation matrix from the SSB-frame to the Wave-frame $R_{W}$:
+These relations can be summarized by the active rotation matrix $R_{W}$ from the SSB-frame to the Wave-frame:
 \be\label{eq:RW}
@@
 	& \qquad\quad\;\; \left. + \sin\frac{\pi}{3} \sin\beta \sin\alpha \right] \,,\\
-	\lambda_{L} &= \arctan \left[ \cos\psi \left( \cos\frac{\pi}{3}\cos\beta + \sin\frac{\pi}{3} \sin\beta \cos(\lambda - \alpha)\right) \right. \nn\\
+	\psi_{L} &= \arctan \left[ \cos\psi \left( \cos\frac{\pi}{3}\cos\beta + \sin\frac{\pi}{3} \sin\beta \cos(\lambda - \alpha)\right) \right. \nn\\
 	& \qquad\quad\;\; \left. + \sin \psi \sin\frac{\pi}{3} \sin(\lambda - \alpha), \right. \nn\\
@@
 
+\subsection{Mode decomposition and polarization angle}
+\label{sec:modespol}
+
+We now translate the mode decomposition~\eqref{eq:hpcmodes} in the Fourier domain, and make more explicit the dependence of the signal in the polarization $\psi$, so as to make manifest its purely extrinsic nature.
+Indeed, $\psi$ enters the observables only through constant prefactors, which is a bit obscured by the rotation matrix formulation~\eqref{eq:RW}.
+
+The mode decomposition~\eqref{eq:hpcmodes} gives (dropping $(\iota, \varphi)$)
+\bsub
+\begin{align}
+	h_{+} = \frac{1}{2} \sum_{\ell, m} \left( \sYlm h_{\ell, m} + \sYlmstar h_{\ell, m}^{*} \right) \,,\\
+	h_{+} = \frac{i}{2} \sum_{\ell, m} \left( \sYlm h_{\ell, m} - \sYlmstar h_{\ell, m}^{*} \right) \,,
+\end{align}
+\esub
+which is valid in general. Now, for non-precessing binary systems, an exact symmetry relation between modes reads
+\be\label{eq:nonprecsymmetry}
+	h_{\ell, -m} = (-1)^{\ell} h_{\ell, m}^{*} \,.
+\ee
+Using this symmetry, we can write
+\be
+	h_{+,\times} = \sum_{\ell, m} K_{\ell m}^{+, \times} h_{\ell m} \,, 
+\ee
+with
+\bsub
+\begin{align}
+	K_{\ell m}^{+} =\frac{1}{2} \left( \sYlm + (-1)^{\ell} \sYlminusmstar \right) \,,\\
+	K_{\ell m}^{\times} = \frac{i}{2} \left( \sYlm - (-1)^{\ell} \sYlminusmstar \right) \,.
+\end{align}
+\esub
+
+Going to the Fourier domain, an approximation often used is to neglect support for negative/positive frequencies according to
+\be\label{eq:approxnegf}
+	\tilde{h}_{\ell m} (f) \simeq 0 \;\; \text{for} \;\; m<0, \; f>0 \;\; ( m>0, \; f<0 )\,,
+\ee
+and neglecting modes $h_{\ell 0}$.
+We will use this approximation throughout this paper.
+Note that we picked our Fourier convention~\eqref{eq:defFourier} to ensure that the mode $\tilde{h}_{22} (f)$ has support for $f>0$.
+Using~\eqref{eq:approxnegf},
+\be
+	\tilde{h}_{+,\times} = \sum_{\ell} \sum_{m>0} K_{\ell m}^{+, \times} \tilde{h}_{\ell m} \,.
+\ee
+
+Next, it is convenient to introduce mode-by-mode polarization matrices
+\be
+	P_{\ell m} = P_{+} K_{\ell m}^{+} + P_{\times} K_{\ell m}^{\times} \,,
+\ee
+so that
+\be
+	H = \sum_{\ell m} P_{\ell m} h_{\ell m} \,.
+\ee
+To make explicit the dependence in polarization, we can define polarization tensors for 0 polarization angle as
+\bsub
+\begin{align}
+	P_{+}^{0} &= P_{+}(\psi = 0) = u \otimes u - v \otimes v \,,\\
+	P_{\times}^{0} &= P_{\times}(\psi = 0) = u \otimes v + v \otimes u \,.
+\end{align}
+\esub
+This allows to write the dependence in polarization as
+\be
+	P_{+} + i P_{\times} = e^{-2 i \psi} \left( P_{+}^{0} + i P_{\times}^{0} \right) \,,
+\ee
+or explicitly in $P_{\ell m}$ as
+\begin{align}
+	P_{\ell m} &= \frac{1}{2} \sYlm e^{-2 i \psi} \left( P_{+}^{0} + i P_{\times}^{0} \right) \nn\\
+	& + \frac{1}{2} (-1)^{\ell} \sYlminusmstar e^{+2 i \psi} \left( P_{+}^{0} - i P_{\times}^{0} \right) \,.
+\end{align}
+
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+
+\subsection{The LISA observables}
+\label{sec:LISAobservables}
+
+The single-link frequency shift observable $y_{slr} = (\nu_{r} - \nu_{s})/\nu$ between sending spacecraft $s$ and receiving spacecraft $r$ along the link $l$ takes the form 
+\be\label{eq:defyslr}
+	y_{slr} = \frac{1}{2} \frac{n_{l} \otimes n_{l}}{1 - k\cdot n_{l}} : \left[ H(t - L - k\cdot p_{s}) - H(t - k\cdot p_{r}) \right] \,.
+\ee
+Several assumptions enter this form of the response: \dots.
+In the frequency domain, the corresponding form of the kernel $G^{\ell m}_{slr}(f, t)$ for the response to a given mode $h_{\ell m}$ in this approximation is
+\begin{align}\label{eq:Gslr}
+	G_{slr}^{\ell m}(f,t) &= \frac{i \pi f L}{2} \sinc \left[ \pi f L\left(1-k\cdot n_{l} \right) \right] \nn\\
+	& \quad \cdot \exp\left[ i \pi f \left( L + k\cdot \left( p_{1} + p_{2} \right) \right) \right]  n_{l} \cdot P_{\ell m} \cdot n_{l} \,.
+\end{align}
+
+In our rigid approximation, delays are simply all constant and constant to $L$. Using the shortcut notation $y_{sr,L} = y_{slr}(t - L)$, the first-generation TDI Michelson observable $X$ reads
+\begin{align}
+	X &= y_{31} + y_{13,L} + \left( y_{21} + y_{12,L} \right)_{2L} \nn\\
+	& - \left( y_{21} + y_{12,L} \right) - \left( y_{31} + y_{13,L} \right)_{2L} \,,
+\end{align}
+with the other Michelson observables $Y$, $Z$ being obtained by cyclic permutation. The independent combinations $A$, $E$ and $T$ are then defined as
+\bsub
+\begin{align}
+	A &= \frac{1}{\sqrt{2}} \left( Z - X \right) \,,\\
+	E &= \frac{1}{\sqrt{6}} \left( X - 2Y + Z \right) \,,\\
+	T &= \frac{1}{\sqrt{3}} \left( X + Y + Z \right) \,.
+\end{align}
+\esub
+These channels are independent under the approximation of an uncorrrelated noise between the arms. With constant delays in the rigid approximation, the TDI combinations take the form
+\bsub\label{eq:aet}
+\begin{align}
+	\tilde{a} &= \left[ (1+z) \left( \tilde{y}_{31} + \tilde{y}_{13} \right) - \tilde{y}_{23} - z \tilde{y}_{32} - \tilde{y}_{21} - z \tilde{y}_{12} \right] \,,\\
+	\tilde{e} &= \frac{1}{\sqrt{3}} \left[ (1-z)\left( \tilde{y}_{13} - \tilde{y}_{31} \right) + (2+z) \left( \tilde{y}_{12} - \tilde{y}_{32} \right) \right. \nn\\
+	&\qquad\quad \left. + (1+2z) \left( \tilde{y}_{21} - \tilde{y}_{23} \right) \right] \,,\\
+	\tilde{t} &= \frac{\sqrt{2}}{\sqrt{3}} \left[ \tilde{y}_{21} - \tilde{y}_{12} + \tilde{y}_{32} - \tilde{y}_{23} + \tilde{y}_{13} - \tilde{y}_{31} \right] \,,
+\end{align}
+\esub
+where we have introduced the rescalings
+\bsub\label{eq:scalingAET}
+\begin{align}
+	\tilde{a}, \tilde{e} &= \frac{1}{i \sqrt{2} z \sin 2\pi f L}\times \tilde{A}, \tilde{E} \,,\\
+	\tilde{t} &= \frac{1}{2\sqrt{2} z^{3} \sin \pi f L \sin 2\pi f L} \times \tilde{T} \,.
+\end{align}
+\esub
+Scaling out the same square factors from the noise power spectral density as
+\bsub
+\begin{align}
+	S_{n}^{A}, S_{n}^{E} &= 2 \sin^{2} 2\pi f L \times S_{n}^{a}, S_{n}^{e} \,,\\
+	S_{n}^{T} &= 8 \sin^{2} \pi f L \sin^{2} 2\pi f L S_{n}^{t} \,,
+\end{align}
+\esub
+the reduced PSD for the three channels take the form
+\bsub
+\begin{align}
+	S_{n}^{a} = S_{n}^{e} &= 2 \left( 3 + 2\cos 2\pi f L + \cos 4 \pi f L \right) S^{\rm pm}(f) \nn\\
+	& \quad + \left( 2 + \cos 2\pi f L \right) S^{\rm op}(f) \,,\\
+	S_{n}^{t} &= 4 \sin^{2} 2\pi f L S^{\rm pm}(f) + S^{\rm op}(f) \,,
+\end{align}
+\esub
+with $S^{\rm pm}$ the test-mass noise PSD and $S^{\rm op}$ the optical noise PSD.
+Here we ignore the confusion noise coming from the background of galactic binaries in the  LISA band.
+Note that the prefactors~\eqref{eq:scalingAET} are oscillatory and have zero-crossings at high frequencies, and scaling them out from both the signal and the noise is an approximation, as we can expect that the real delays would lead to imperfect cancellations in the vicinity of the zero-crossings.
+
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+
 \subsection{The low-frequency limit}
@@
 
--- make the connection to low-f response (2 interferometers)
+As is well known~\cite{Cutler97}, in the low-frequency limit (also called the long-wavelength approximation), the finite-length effects of the arms of LISA disappear, and the response of the detectors falls back to the one of two LIGO-type detectors set in motion, rotated from each other by $\pi/4$.
+
+For $f \ll f_{L}$, we have $2\pi f L \ll 1$ and the kernel~\eqref{eq:defGslr} reduces to
+\be\label{eq:Gslrlowf}
+	G^{\ell m}_{slr} \simeq \frac{i \pi f L}{2} \exp\left[ 2 i \pi f k\cdot p_{0} \right] n_{l} \otimes n_{l} : P_{\ell m}\,.
+\ee
+For even lower frequencies, when $f \ll f_{R}$ as well, $2\pi f R \ll 1$ and the delay term is negligible: $\exp\left[ 2 i \pi f k\cdot p_{0} \right] \simeq 1$.
+
+In the following, we will also apply frequency-domain response at leading order in the separation of timescales~\cite{MB18}, so that the transfer functions are $\calT(f) = G(f, \tf)$ with $\tf$ the time-of-frequency function~\eqref{eq:deftf}.
+For $2\pi f L \ll 1$, we have $z\simeq 1$, and the link reversal symmetry $\tilde{y}_{slr} \simeq \tilde{y}_{r-ls}$, so that~\eqref{eq:aet} becomes
+\bsub\label{eq:aetlowf}
+\begin{align}
+	\tilde{a} &\simeq 4\tilde{y}_{31} - 2 \tilde{y}_{23} - 2 \tilde{y}_{12} \,,\\
+	\tilde{e} &\simeq 2\sqrt{3} \left[ \tilde{y}_{12}  - \tilde{y}_{23} \right] \,,\\
+	\tilde{t} &\simeq 0 \,,
+\end{align}
+\esub
+with the $T$-channel becoming negligible in this limit. Using~\eqref{eq:Gslrlowf}, we can write
+\be\label{eq:aelowfmodes}
+	\tilde{a}, \tilde{e} = \sum_{\ell, m>0} (-2i\pi f) \exp\left[ 2 i \pi f k\cdot p_{0} \right] \tilde{h}_{\ell m} D_{a,e} : P_{\ell m} \,,
+\ee
+where we introduced the detector tensors
+\bsub\label{eq:DaDe}
+\begin{align}
+	D_{a} &= \frac{L}{2} \left( n_{1}\otimes n_{1} + n_{3} \otimes n_{3} - 2 n_{2} \otimes n_{2} \right) \,,\\
+	D_{e} &= \frac{L\sqrt{3}}{2} \left( n_{1} \otimes n_{1} - n_{3} \otimes n_{3} \right) \,.
+\end{align}
+\esub
+Here, we have made apparent factors $(-2i\pi f)$ corresponding to the Fourier-domain translation of a time derivative.
+Indeed, the observables $\Delta \nu / \nu$ used here are time derivatives of the laser phase more commonly used for ground-based detectors.
+
+We can now map these two channels to two fictitious LIGO-type detectors as follows.
+For an orthogonal detector of the same setup as the ground-based LIGO and Virgo with armlength $L_{D}$, rotated by an angle $\epsilon_{D}$ from the basis vectors $x,y$, the detector tensor is
+\begin{align}
+	D &= \frac{L_{D}}{2} \left[ \cos 2\epsilon_{D} \left( x\otimes x - y\otimes y \right) \right. \nn\\
+	& \qquad\quad + \left. \sin 2\epsilon_{D} \left( x\otimes y + y \otimes x\right) \right] \,.
+\end{align}
+By comparison with~\eqref{eq:DaDe}, we obtain
+\be
+	L_{a} = L_{e} = \frac{3L}{2} \,, \; \epsilon_{a} = \frac{2\pi}{3} \,, \; \epsilon_{e} = \frac{5\pi}{12} \,.
+\ee
+Note a degeneracy $\pm \pi$ in how we choose to orient these effective detectors. We can recover pattern functions (for $\psi_{L}=0$) by factoring out the effective length, defining
+\be
+	F_{a,e}^{+,\times} \equiv \frac{2}{3L} D_{a,e} : P_{+,\times}^{0} \,,
+\ee
+which gives in terms of the LISA-frame sky position angles
+\bsub\label{eq:FapcFepc}
+\begin{align}
+	F_{a}^{+} &= \frac{1}{2} \left( 1 + \sin^{2}\beta_{L} \right) \sin \left(2\lambda_{L} + \frac{\pi}{6} \right) \,,\\
+	F_{a}^{\times} &= -\sin\beta_{L} \cos \left(2\lambda_{L} + \frac{\pi}{6} \right) \,,\\
+	F_{e}^{+} &= \frac{1}{2} \left( 1 + \sin^{2}\beta_{L} \right) \cos \left(2\lambda_{L} + \frac{\pi}{6} \right) \,,\\
+	F_{e}^{\times} &= \sin\beta_{L} \sin \left(2\lambda_{L} + \frac{\pi}{6} \right)  \,.
+\end{align}
+\esub
 
@@
 
-\subsection{A simple extrinsic likelihood}
+\subsection{TO UPDATE -- A simple extrinsic likelihood}
 \label{sec:simple-like}

### 2019-02-12 paper/paper_PE.tex
Paper: correcting typos

@@
 \newcommand{\sYlm}{{}_{-2}Y_{\ell m}}
-\newcommand{\sYlmstar}{{}_{-2}Y_{\ell, -m}^{*}}
+\newcommand{\sYlmstar}{{}_{-2}Y_{\ell m}^{*}}
 \newcommand{\sYlminusmstar}{{}_{-2}Y_{\ell, -m}^{*}}
@@
 \author{Sylvain Marsat}
-\affiliation{Department of Physics, University of Maryland, College Park, MD 20742, USA}
+\affiliation{APC, AstroParticule et Cosmologie, Universit\'{e} Paris Diderot, CNRS/IN2P3, CEA/Irfu, Observatoire de Paris, Sorbonne Paris Cit\'{e}, 10, rue Alice Domon et L\'{e}onie Duquet 75205 PARIS Cedex 13, France}
+\author{Tito Dal Canton}
 \affiliation{Gravitational Astrophysics Laboratory, NASA Goddard Space Flight Center, 8800 Greenbelt Rd., Greenbelt, MD 20771, USA}
-\affiliation{Max Planck Institute for Gravitational Physics (Albert Einstein Institute), Am M\"uhlenberg 1, Potsdam-Golm, 14476, Germany}
 
@@
 \begin{align}
-	h_{+} = \frac{1}{2} \sum_{\ell, m} \left( \sYlm h_{\ell, m} + \sYlmstar h_{\ell, m}^{*} \right) \,,\\
-	h_{+} = \frac{i}{2} \sum_{\ell, m} \left( \sYlm h_{\ell, m} - \sYlmstar h_{\ell, m}^{*} \right) \,,
+	h_{+} = \frac{1}{2} \sum_{\ell, m} \left( \sYlm h_{\ell m} + \sYlmstar h_{\ell m}^{*} \right) \,,\\
+	h_{\times} = \frac{i}{2} \sum_{\ell, m} \left( \sYlm h_{\ell m} - \sYlmstar h_{\ell m}^{*} \right) \,,
 \end{align}
@@
 
+\begin{figure*}
+  \centering
+  \includegraphics[width=.98\linewidth]{../plots/corner_smbh_case0_ptmcmc_22_hm.png}
+  \caption{.}
+  \label{fig:PEsmbh22hmCase0}
+\end{figure*}
+
 -- High-mass system, non-degenerate, 22 and HM + Fisher - stacked SNR contours
@@
 
+-- Move to Appendix ?
+
 %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
@@
 \label{sec:SMBHPEdegen}
+
+\begin{figure*}
+  \centering
+  \includegraphics[width=.98\linewidth]{../plots/corner_smbh_case9_ptmcmc_22_hm.png}
+  \caption{.}
+  \label{fig:PEsmbh22hmCase0}
+\end{figure*}
 

### 2019-03-13 paper/paper_PE.tex
Some initial text for SOBH section

@@
 \newcommand{\jgb}[1]{{\color{DarkGreen} #1}}
+\newcommand{\tdc}[1]{{\color{red} #1}}
 
@@
 
+Stellar-origin black-hole (SOBH) mergers have been recently recognized as an
+important source of LISA detections \cite{Sesana16} although full parameter
+estimation studies are not available yet. In this section we present
+three case studies of parameter inference for such sources. The parameters of
+the simulated mergers are taken from the Radler LISA data challenge and are
+listed in Table \ref{table:sobh_params}.
+\begin{table}
+	\begin{tabular}{|c||c|c|c|}
+		\hline
+		Radler entry 		& 1 	& 12 	& 52 \\
+		\hline
+		Mass 1 ($\Msol$) 	& 21.4	& 63.5	& 57.0 \\
+		\hline
+		Mass 2 ($\Msol$) 	& 20.1	& 45.1	& 36.8 \\
+		\hline
+		Distance (Mpc) 		& 49.1	& 318	& 308 \\
+		\hline
+		LISA SNR 			& 26.6	& 11.9	& 16.0 \\
+		\hline
+	\end{tabular}
+	\caption{Parameters of the simulated SOBH mergers. \tdc{Maybe we can show
+			 these on the upper right of each corner plot instead.}}
+	\label{table:sobh_params}
+\end{table}
+
+The waveform model used to work with such systems is the phenomenological
+approximant IMRPhenomD, extended at low frequency with a post-Newtonian
+frequency-domain waveform. In practice, because LISA only observes the early
+inspiral of SOBH systems, only the post-Newtonian waveform plays a role here.
+We do not expect radiation modes beyond the quadrupole to give a significant
+contribution to the parameter estimates, so we only include the $(2,2)$ mode
+for simplicity. We do not include any effect arising from black-hole spins.
+
+As done in the previous section, we apply our parameter inference procedure to
+these systems using both BAMBI and parallel-tempered MCMC sampling, and we
+numerically evaluate the Fisher matrix at the parameters of the injection.
+
+Due to the duration of an SOBH inspiral in the LISA band, and the complete lack
+of merger and ringdown, the chirp mass $\Mchirp$ is measured much better than
+other mass parameters and the component masses have a much stronger correlation
+than usually found with ground-based gravitational-wave detectors. In order for
+the sampling to converge robustly in a reasonable time, we find it is necessary
+to express the mass parameters in terms of $\Mchirp$ and symmetric mass ratio
+$\eta$, and a relatively tight prior around the injected parameters is required
+as well.
+
+In order to generate these results, our PTMCMC sampler requires between
+$5\times 10^5$ and $3\times 10^6$ steps with 100 temperatures, taking on the
+order of days on a single CPU core without careful optimization. The BAMBI
+sampling, on the other hand, completes in tens of minutes only.
+
+The resulting posterior distributions are shown in Fig.~\ref{fig:sobh_corner_1}
+and \ref{fig:sobh_corner_12}. We find close agreement between the BAMBI and
+PTMCMC samplers. The full joint distributions are fairly complex, but the
+complexity is mostly restricted to multimodalities and degeneracies in the
+angular parameters only. The typical degeneracy between inclination and distance
+is evident. Masses, merger time, distance, inclination and sky location, which
+are arguably the most astrophysically useful parameters, have simple marginal
+distributions with a well-defined mode and close to Gaussian, even at these
+moderate SNR values. The Fisher matrix estimates represent the marginal
+distributions of these parameters at least to the order of magnitude. In some
+cases the Fisher estimates approximate very precisely even the joint
+distributions of pairs of these parameters.
+
+\begin{figure*}
+	\includegraphics[width=\textwidth]{../plots/corner_sobh_tdc1_ptmcmc_bambi}
+	\caption{Inferred parameter posterior distribution for SOBH system 1.}
+	\label{fig:sobh_corner_1}
+\end{figure*}
+
+\begin{figure*}
+	\includegraphics[width=\textwidth]{../plots/corner_sobh_tdc2_ptmcmc_bambi}
+	\caption{Inferred parameter posterior distribution for SOBH system 12.}
+	\label{fig:sobh_corner_12}
+\end{figure*}
+
 %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

### 2019-03-15 paper/paper_PE.tex
Added plots to the paper

--- 2019-02-12 paper/paper_PE.tex
@@
 
-\section{Morphology of the signals}
-\label{sec:morph}
+\section{Supermassive black holes}
+\label{sec:SMBH}
 
-- Morphology of the response
+- Demonstration/examples
 
@@
 
--- SNR accumulation with time or frequency: inspiral/MRD balance across mass range
+\begin{figure*}
+  \centering
+  \includegraphics[width=.98\linewidth]{../plots/plot_smbh9_alltdi.pdf}
+  \caption{\SM{TODO: rethink the visualization, maybe plot only transfer functions}.}
+  \label{fig:tranferSMBHCase9}
+\end{figure*}
+
+\begin{figure*}
+  \centering
+  \includegraphics[width=.7\linewidth]{../plots/plot_tdia_hm.pdf}
+  \caption{\SM{TODO: make figure smaller, single-column.}}
+  \label{fig:FDHMSMBHCase9}
+\end{figure*}
 
@@
 -- SNR accumulation with time or frequency: inspiral/MRD balance across mass range
-
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-
-\section{Supermassive black holes}
-\label{sec:SMBH}
-
-- Demonstration/examples
 
@@
 
+\begin{figure*}
+  \centering
+  \begin{minipage}{.32\linewidth}
+      \includegraphics[width=.99\linewidth]{../plots/corner_smbh_case9_hm_tseries_m1m2.png}
+   \end{minipage}
+   \begin{minipage}{.32\linewidth}
+      \includegraphics[width=.99\linewidth]{../plots/corner_smbh_case9_hm_tseries_Dinc.png}
+   \end{minipage}
+   \begin{minipage}{.32\linewidth}
+      \includegraphics[width=.99\linewidth]{../plots/corner_smbh_case9_hm_tseries_tphi.png}
+   \end{minipage}
+  \caption{.}
+  \label{fig:smbhCornerZoomHMCase9}
+\end{figure*}
+
+\begin{figure*}
+  \centering
+  \includegraphics[width=.8\linewidth]{../plots/plot_mollweide_tseries.png}
+  \caption{.}
+  \label{fig:MollweidesmbhTserieshmCase9}
+\end{figure*}
+
 %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
@@
 \end{figure*}
+
+\begin{figure*}
+  \centering
+  \includegraphics[width=.8\linewidth]{../plots/corner_smbh_case9_simplelike_ptmcmc_bambi.png}
+  \caption{.}
+  \label{fig:PEsmbh22hmSimpleLikeCase0}
+\end{figure*}
+
+\begin{figure}
+  \centering
+  \begin{minipage}{.49\linewidth}
+  \includegraphics[width=.99\linewidth]{../plots/corner_skydown_smbh_case9_22_ptmcmc_bambi_withdegen.png}
+  \end{minipage}
+  \begin{minipage}{.49\linewidth}
+  \includegraphics[width=.99\linewidth]{../plots/corner_skydown_smbh_case9_simplelike_ptmcmc_bambi_withdegen.png}
+  \end{minipage}
+  \caption{.}
+  \label{fig:Skydownsmbh22hmSimpleLikeCase0}
+\end{figure}
 
@@
 %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+
+\subsection{Signals and transfer functions}
+\label{subsec:signaltransferSOBH}
+
+\begin{figure*}
+  \centering
+  \includegraphics[width=.98\linewidth]{../plots/plot_sobh1_alltdi.pdf}
+  \caption{\SM{TODO: rethink the visualization, maybe plot only transfer functions}.}
+  \label{fig:PEsmbh22hmCase0}
+\end{figure*}
+
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
 %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

### 2019-04-02 paper/paper_PE.tex
Work on the paper

@@
 
-The definition of the reference phase $\phi_{\rm Ref}$, \dots
+The definition of the reference phase $\phi_{\rm Ref}$, \dots \SM{[COMPLETE]}
 
@@
 so that
-\be
+\be\label{eq:Hsummodes}
 	H = \sum_{\ell m} P_{\ell m} h_{\ell m} \,.
@@
 \end{align}
+Writing the $P_{\ell m}$ matrices in this way allows to factor out all dependencies in the extrinsic parameters $(\iota, \phi, \psi)$.
+Together with the luminosity distance $D_{L}$, these parameters appear in combinations of constant factors multiplying the whole response.
 
@@
 \ee
-Several assumptions enter this form of the response: \dots.
-In the frequency domain, the corresponding form of the kernel $G^{\ell m}_{slr}(f, t)$ for the response to a given mode $h_{\ell m}$ in this approximation is
+Several assumptions enter this form of the response: \dots \SM{[COMPLETE]}.
+
+In the frequency domain, we will use the analysis of~\cite{} and write the response in individual observables $y_{slr}$ with a transfer function for each spherical harmonic mode as
+\be
+	\tilde{y}_{slr} = \sum_{\ell, m}\calT_{slr}^{\ell m}(f) \tilde{h}_{\ell m} \,.
+\ee
+With the decomposition~\eqref{eq:Hsummodes}, the corresponding form of the kernel $G^{\ell m}_{slr}(f, t)$ in the formalism of~\cite{} is
 \begin{align}\label{eq:Gslr}
@@
 \end{align}
+At leading order in the separation of timescales, we have simply
+\be
+	\calT_{slr}^{\ell m}(f) = G_{slr}^{\ell m}(f, t_{f}^{\ell m}) \,,
+\ee
+with $t_{f}^{\ell m}$ the effective time-to-frequency correspondence, defined across the whole frequency band, including in the merger-ringdown, as $t_{f}^{\ell m} = -1/(2\pi) d\Psi_{\ell m} / df$ for a mode $\tilde{h}_{\ell m} = A_{\ell m}(f) \exp\left[-i \Psi_{\ell m}\right]$.
+The analysis of~\cite{} has shown that higher-order corrections are small for SMBH systems, and for SOBH systems that are close enough to merger as will be the case for the examples chosen here.
+The fact that we use the same Fourier-domain treatment of the transfer functions for both the signal and the templates should also mitigate errors.
 
@@
 Note a degeneracy $\pm \pi$ in how we choose to orient these effective detectors. We can recover pattern functions (for $\psi_{L}=0$) by factoring out the effective length, defining
-\be
+\be\label{eq:defFapcFepc}
 	F_{a,e}^{+,\times} \equiv \frac{2}{3L} D_{a,e} : P_{+,\times}^{0} \,,
@@
 \begin{align}
-	F_{a}^{+} &= \frac{1}{2} \left( 1 + \sin^{2}\beta_{L} \right) \sin \left(2\lambda_{L} + \frac{\pi}{6} \right) \,,\\
-	F_{a}^{\times} &= -\sin\beta_{L} \cos \left(2\lambda_{L} + \frac{\pi}{6} \right) \,,\\
+	F_{a}^{+} &= \frac{1}{2} \left( 1 + \sin^{2}\beta_{L} \right) \cos \left(2\lambda_{L} - \frac{\pi}{3} \right) \,,\\
+	F_{a}^{\times} &= \sin\beta_{L} \sin \left(2\lambda_{L} - \frac{\pi}{3} \right) \,,\\
 	F_{e}^{+} &= \frac{1}{2} \left( 1 + \sin^{2}\beta_{L} \right) \cos \left(2\lambda_{L} + \frac{\pi}{6} \right) \,,\\
@@
 \esub
-
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-
-\subsection{TO UPDATE -- A simple extrinsic likelihood}
-\label{sec:simple-like}
-
-In this section, we investigate a simplified form of the extrinsic likelihood and the structure of its degeneracies.
-
-It will be useful to introduce the following notation for the patter functions associated to the independent channels A and E:
-\bsub\label{eq:defDaDe}
-\begin{align}
-	D_{a}^{+} &= \frac{1}{2} \left( 1 + \sin^{2}\betaL \right) \cos\left( 2\lambdaL - \frac{\pi}{3} \right) \,, \\
-	D_{a}^{\times} &= \sin \betaL \sin\left( 2\lambdaL - \frac{\pi}{3} \right) \,, \\
-	D_{e}^{+} &= - \frac{1}{2} \left( 1 + \sin^{2}\betaL \right) \sin\left( 2\lambdaL - \frac{\pi}{3} \right) \,, \\
-	D_{e}^{\times} &= \sin \betaL \cos\left( 2\lambdaL - \frac{\pi}{3} \right) \,.
-\end{align}
-\esub
-Note that the expressions for the channel E can be obtained from the expressions for A with the replacement $\lambdaL \rightarrow \lambdaL + \pi/4$.
-
-In each channel, we separate the contribution of the modes $h_{22}$ and $h_{2,-2}$ as
-\bsub\label{eq:defsase}
-\begin{align}
-	s_{a} &= a_{22} + a_{2,-2} \,, \\
-	s_{e} &= e_{22} + e_{2,-2} \,,
-\end{align}
-\esub
-with
-\bsub\label{eq:defa22a2m2}
-\begin{align}
-	a_{22} &= \frac{3i}{4d} \sqrt{\frac{5}{\pi}} \cos^{4}\frac{\iota}{2} e^{2i(-\varphi-\psiL)} \left( D_{a}^{+} + i D_{a}^{\times} \right) \,, \\
-	a_{2,-2} &= \frac{3i}{4d} \sqrt{\frac{5}{\pi}} \sin^{4}\frac{\iota}{2} e^{2i(-\varphi+\psiL)} \left( D_{a}^{+} - i D_{a}^{\times} \right) \,.
-\end{align}
-\esub
-and similarly for the E channel with $D_{e}^{+,\times}$ instead of $D_{a}^{+, \times}$ as
-\bsub\label{eq:defe22e2m2}
-\begin{align}
-	e_{22} &= \frac{3i}{4d} \sqrt{\frac{5}{\pi}} \cos^{4}\frac{\iota}{2} e^{2i(-\varphi-\psiL)} \left( D_{e}^{+} + i D_{e}^{\times} \right) \,, \\
-	e_{2,-2} &= \frac{3i}{4d} \sqrt{\frac{5}{\pi}} \sin^{4}\frac{\iota}{2} e^{2i(-\varphi+\psiL)} \left( D_{e}^{+} - i D_{e}^{\times} \right) \,.
-\end{align}
-\esub
-Here, we introduced the dimensionless luminosity distance ratio $d \equiv D / D^{\rm inj}$.
-
-In this simplified response, an exact symmetry relation is manifest: changing simultaneously
-\be\label{eq:symmetryresponse}
-\betaL \rightarrow -\betaL\,, \quad \iota \rightarrow \pi - \iota \,, \quad \psiL \rightarrow \pi - \psiL
-\ee
-leaves the signal unchanged. We can also rewrite the likelihood in a way that makes clearer the symmetry between the two pairs of parameters $(\iota, \varphiL)$ and $(\betaL, \lambdaL)$ \SM{[TODO]}.
-
-The simplified extrinsic likelihood for a frozen LISA then takes the form
-\be\label{eq:simplelikelihood}
-	\ln \calL = - \frac{1}{2} \Lambda \left( \left| s_{a} - s_{a}^{\rm inj} \right|^{2} + \left| s_{e} - s_{e}^{\rm inj} \right|^{2} \right) \,,
-\ee
-where $\Lambda$ appears as an overall normalization factor, playing the role of the square SNR of the signal.
+Note that the expressions for the channel $e$ can be obtained from the expressions for $a$ with the replacement $\lambdaL \rightarrow \lambdaL + \pi/4$.
 
@@
 
+\subsection{Bayesian setting}
+\label{subsec:defbayes}
+
+-- definitions relevant to the Bayesian setting (likelihood, ...)
+
+Introducing the standard matched-filter inner product as
+\be
+	\left( \tilde{a} | \tilde{b} \right) = 4 \mathrm{Re} \int_{0}^{+\infty} df \; \frac{\tilde{a}(f) \tilde{b}^{*}(f)}{S_{n}(f)} \,,
+\ee
+for stationary Gaussian noise with noise PSD $S_{n}$, the likelihood $\calL = p(d | \bm{\theta})$ for a given gravitational wave channel takes the form
+\be\label{eq:deflnL}
+	\ln \calL = -\frac{1}{2} \left( h(\bm{\theta}) - d | h(\bm{\theta}) - d \right) \,,
+\ee
+with the data stream being a superposition of the gravitational wave signal for the true parameters $\bm{\theta}_{0}$ and the noise realization in the experiment, $d = s(\bm{\theta}_{0}) + n$.
+The posterior distribution for the physical parameters $\bm{\theta}$ given the observed data $d$ is then
+\be\label{eq:defBayes}
+	p\left( \bm{\theta} | d \right) = \frac{p(d | \bm{\theta})p(\bm{\theta})}{p(d)}\,,
+\ee
+with $p(\bm{\theta})$ the priors on the parameters, and $p(d)$ the evidence that we will ignore here as it enters as a multiplicative constant in the posterior.
+
+In our analysis, we will use the so-called zero-noise approximation.
+This corresponds simply to performing a Bayesian analysis with~\eqref{eq:deflnL} and ~\eqref{eq:defBayes}, but setting the noise realization to zero, $n=0$.
+It allows us to accelerate the likelihood computation, as will be explained in~\ref{}.
+For exploratory parameter estimation runs, this approximation allows us to explore the structure of the likelihood with its possible parameter degeneracies (\SM{[cite past use of this approx.]}), and is a significant improvement over Fisher matrix estimates in that sense.
+We leave the extension of our analysis to include noise realizations for future work.
+We will also ignore waveform systematic errors and assume our waveform templates match exactly the real signals.
+
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+
 \subsection{Frequency domain LISA response}
@@
 
-\subsection{Reduced order model for EOBNRv2 waveforms.}
+\subsection{Reduced order model for EOBNRv2HM waveforms.}
 \label{sec:waveforms}
@@
 
-\subsection{Highlighting degeneracies}
+\subsection{Investigating degeneracies}
 \label{sec:SMBHPEdegen}
@@
   \caption{.}
-  \label{fig:PEsmbh22hmSimpleLikeCase0}
+  \label{fig:PEsmbh22hmSimpleLikeCase9}
 \end{figure*}
@@
   \caption{.}
-  \label{fig:Skydownsmbh22hmSimpleLikeCase0}
+  \label{fig:Skydownsmbh22hmSimpleLikeCase9}
 \end{figure}
 
-- Highlighting degeneracies
-
--- High-mass degenerate example, 22 vs HM - stacked SNR contours
--- show degenerate waveforms
--- simplified response and closed-form likelihood, full degeneracies
--- (explore simplified response further: parameter maps, other off-the-shelf samplers ?)
-
-We can now explore possible degeneracies in the likelihood given by~\eqref{eq:simplelikelihood}, \eqref{eq:defsase} and \eqref{eq:defa22a2m2}-\eqref{eq:defe22e2m2}.
+In this section, we investigate the cause of the degeneracies appearing in the posterior of Case~II shown in~\ref{subsec:SMBHPE}.
+Understanding the structure of these features could prove useful to inform Bayesian samplers, for instance by building adapted jump proposal allowing a much faster mixing of MCMC chains than naive methods.
+a simplified form of the extrinsic likelihood and the structure of its degeneracies.
+
+An important clue about the appearance of these degeneracies is the length of the observable signal.
+As shown in Fig.~\ref{}, for our example SMBH systems the SNR accumulates in a matter of days before merger, which means that there is only little impact of the modulations caused by the motion of the detector on the TDI transfer functions~\eqref{}.
+If we can neglect entirely the motion of the detector (which is not quite true as we will see later on), then we are in a regime where the information about the sky position of the source comes essentially from the pattern functions affecting this high-SNR signal.
+Using the the low-frequency approximation (see~\ref{}) and the time-at-LISA $t_{L}$ as a time variable, there is no frequency dependence in the response.
+A further simplification occurs when pinning the masses and the time of coalescence to their true values.
+We found indeed that these parameters correlate only weakly to the other ones, so that we can hope to understand the degeneracy features by restricting the analysis to the six extrinsic parameters $(D_{L}, \iota, \varphi, \lambdaL, \betaL, \psiL)$.
+
+Under all these simplifying assumptions, the likelihood function becomes purely extrinsic, and takes the form of trivial geometric factors multiplying a constant overlap that can be precomputed.
+Using the subscript $\mathrm{inj}$ for the injected signal, the likelihood~\eqref{eq:defliklihood} is
+\be
+	\ln \calL = -\frac{1}{2} \left( \tilde{a} - \tilde{a}_{\rm inj} | \tilde{a} - \tilde{a}_{\rm inj} \right) -\frac{1}{2} \left( \tilde{e} - \tilde{e}_{\rm inj} | \tilde{e} - \tilde{e}_{\rm inj} \right) \,.
+\ee
+From the low-frequency response \eqref{eq:aelowfmodes}, pinning the time-at-LISA so that we ca ignore the factor $\exp [2 i \pi f k\cdot p_{0}]$ as a pure constant corresponding to a redefinition of time, we obtain
+\begin{align}
+	\ln \calL &= -\frac{1}{2} \sum_{\ell m} \sum_{\ell' m'}   \left[ \left( s_{a}^{\ell m} - s_{a, \mathrm{inj}}^{\ell m}\right) \left( s_{a}^{\ell' m'} - s_{a, \mathrm{inj}}^{\ell' m'}\right)^{*} \right.
+\nn\\
+	& \qquad \left.
++ \left( s_{e}^{\ell m} - s_{e, \mathrm{inj}}^{\ell m}\right) \left( s_{e}^{\ell' m'} - s_{e, \mathrm{inj}}^{\ell' m'}\right)^{*} \right] \left\langle \ell m | \ell' m' \right\rangle \,,
+\end{align}
+where we introduced the following extrinsic signal functions
+\be\label{eq:defsaelm}
+	s_{a,e}^{\ell m} = \frac{1}{d} \frac{2}{3L} D_{a,e} : P_{\ell m} \,,
+\ee
+with $d = D / D^{\rm inj}$ the dimensionless ratio of luminosity distances.
+The prefactors $\left\langle \ell m | \ell' m' \right\rangle$ are pure normalization constants, and represent mode overlaps as
+\be
+	\left\langle \ell m | \ell' m' \right\rangle = 4\mathrm{Re} \int \frac{df}{S_{n}^{a,e}} \left( 3 \pi f L\right)^{2} \tilde{h}_{\ell m} \tilde{h}_{\ell' m'}^{*} \,.
+\ee
+Here, the noise PSD $S_{n}^{a,e}$ is given by~\eqref{} and is identical between the channels $a$ and $e$.
+Having fixed the masses and the time, these overlaps are to be computed with the modes $\tilde{h}_{\ell m}^{\rm inj}$ of the injected signal.
+
+In the case where we only include the dominant harmonic $h_{22}$, the functions~\eqref{eq:defsaelm} take the explicit form
+\begin{align}
+	s_{a,e}^{22} &= \frac{1}{4d} \sqrt{\frac{5}{\pi}} \cos^{4}\frac{\iota}{2} e^{2i(\varphi-\psiL)} \left( F_{a,e}^{+} + i F_{a,e}^{\times} \right) \\
+	&+ \frac{1}{4d} \sqrt{\frac{5}{\pi}} \sin^{4}\frac{\iota}{2} e^{2i(\varphi+\psiL)} \left( F_{a,e}^{+} - i F_{a,e}^{\times} \right) \,,
+\end{align}
+with the pattern functions for the channels $a$ and $e$ defined in \eqref{eq:defFapcFepc}-\eqref{eq:FapcFepc}.
+
+In order to make more manifest the degeneracies in the problem, it will be useful to introduce the following notations.
+First, since $\iota \in [0, \pi]$ while $\beta_{L} \in [-\pi/2, \pi/2]$, it will be more convenient to work with $\theta_{L} = \pi/2 - \beta_{L}$.
+We will further abbreviate notations by using the variables
+\be
+	t_{\iota} \equiv \tan \iota/2 \,, \quad t_{\theta} \equiv \tan \theta_{L} / 2 \,.
+\ee
+We also introduce the azimuthal angles $\lambda_{a} = \lambda_{L} - \pi/6$, $\lambda_{e} = \lambda_{L} + \pi/12 = \lambda_{a} + \pi/4$, and form the combinations
+\be
+	\sigma_{\pm} \equiv s_{a}^{22} \pm i s_{e}^{22} \,.
+\ee
+In these notations, we obtain
+\bsub\label{eq:sigmapm}
+\begin{align}
+	\sigma_{+} &= \rho e^{2i\varphi} \left[ t_{\theta}^{4} e^{-2 i \psiL} + t_{\iota}^{4} e^{2 i \psiL} \right] e^{-2i \lambda_{a}} \,, \\
+	\sigma_{-} &= \rho e^{2i\varphi} \left[ e^{-2 i \psiL} + t_{\theta}^{4} t_{\iota}^{4} e^{2 i \psiL} \right] e^{2i \lambda_{a}} \,,
+\end{align}
+\esub
+with a common prefactor
+\be\label{eq:sigmafactorrho}
+	\rho(d, \iota, \theta_{L}) = \frac{1}{4d} \sqrt{\frac{5}{\pi}} \frac{1}{\left( 1 + t_{\iota}^{2} \right)^{2} \left(1 + t_{\theta}^{2} \right)^{2}} \,.
+\ee
+Finding points in the parameter space that are degenerate with the injection, with $\ln \calL \simeq 0$, amounts to try to obtain both the same value of $\sigma_{+}$ and $\sigma_{-}$ as the injection.
+
+We can now use the simplified response written in the form~\eqref{eq:sigmapm}-\eqref{eq:sigmafactorrho} to look for symmetries and degeneracies.
+First, we note the following manifest symmetry: changing simultaneously
+\be\label{eq:symmetryresponse}
+	\betaL \rightarrow -\betaL\,, \quad \iota \rightarrow \pi - \iota \,, \quad \psiL \rightarrow \pi - \psiL
+\ee
+leaves the likelihood exactly invariant.
+There is also a form of symmetry obatained by exchanging $\iota \leftrightarrow \theta_{L}$.
+This conjugate the term inside bracket for $\sigma_{+}$, which can be corrected withe the phase terms $\varphi \pm \lambda_{a}$.
+If $\Phi = \mathrm{Arg} \left[ t_{\theta}^{4} e^{-2 i \psiL} + t_{\iota}^{4} e^{2 i \psiL} \right]$, we obtain the symmetry
+\begin{align}
+	\iota' = \theta_{L} \,, \quad \theta_{L}' = \iota \,, \nn\\
+	\varphi' = \varphi + \frac{1}{2} \Phi \; \mathrm{mod} \; \pi\,, \nn\\
+	\lambda_{a}' = \lambda_{a} - \frac{1}{2} \Phi \; \mathrm{mod} \; \pi\,.
+\end{align}
+
+We can now explain the degenerate features we found as certain limits in this simplified likelihood.
+In the face-on or face-off limit $\iota \rightarrow 0$, $|t_{\iota}| \ll 1$.
+This amounts to the $(\ell,m)=(2,2)$ contribution dominating the $(\ell, m)=(2,-2)$ contribution.
+Neglecting the terms in~\eqref{eq:sigmapm} that are quartic in $t_{\iota}$, we have
+\bsub\label{eq:sigmapmapprox}
+\begin{align}
+	\sigma_{+} &\simeq \rho e^{2i(\varphi - \psiL)} t_{\theta}^{4} e^{-2i \lambda_{a}} \,, \\
+	\sigma_{-} &\simeq \rho e^{2i(\varphi - \psiL)} e^{2i \lambda_{a}} \,,
+\end{align}
+\esub
+Defining the ratio
+\be
+	r = \frac{\sigma_{+}}{\sigma_{-}} \,,
+\ee
+we see that finding $(\lambda_{a}^{*}, \theta_{L}^{*})$ verifying
+\be\label{eq:conditionsky}
+	\frac{\sigma_{+}}{\sigma_{-}} (\lambda_{a}^{*}, \theta_{L}^{*}) = r_{\rm inj} \,,
+\ee
+with the approximation~\eqref{eq:sigmapmapprox} for the left-hand side, guarantees that we can then match both $\sigma_{+}$ and $\sigma_{-}$ of the injection with the condition
+\be\label{eq:conditionprefactor}
+	\rho e^{2i(\varphi - \psiL)} e^{2i \lambda_{a}^{*}} = \sigma_{-}^{\rm inj}\,.
+\ee
+In the limit $|t_{\iota}| \ll 1$,
+\be
+	\frac{\sigma_{+}}{\sigma_{-}} \simeq t_{\theta}^{4} e^{-4i\lambda_{a}} \,,
+\ee
+so that~\eqref{eq:conditionsky} is verified for special positions in the sky at (coming back to the original variables $(\lambdaL, \betaL)$)
+\bsub\label{eq:soldegensky}
+\begin{align}
+	\beta_{L}^{*} &= \frac{\pi}{2} - 2 \mathrm{Arctan} \; \left| r_{\rm inj}\right|^{1/4} \,, \\
+	\lambda_{L}^{*} &= \frac{\pi}{6} -\frac{1}{4} \mathrm{Arg} \; r_{\rm inj} \;\mathrm{mod}\; \frac{\pi}{2} \,.
+\end{align}
+\esub
+The other condition gives then a single phase constraint for $\varphi - \psiL$, and a single amplitude constraint for $\rho(d, \iota, \theta_{L}^{*})$.
+The limit $\iota \rightarrow \pi$ (where the $(2,-2)$ contribution dominates the $(2,2)$) is very similar, as can be seen most easily by the change of variable $\iota \rightarrow \pi - \iota$, with $-\psiL$ and $\pi - \theta_{L}$ playing the roles of $\psiL$ and $\theta_{L}$.
+
+With the $\lambdaL \rightarrow \lambdaL + k \pi/2$ and $\betaL \rightarrow - \betaL$ symmetries, we arrive at eight sky positions where the degeneracy occurs, with an exact degeneracy for the pair of parameters $(\varphi, \psiL)$, as well as an approximate degeneracy for the pair $(d, \iota)$ as long as we remain in the regime $t_{\iota}^{4} \ll 1$ or $t_{\iota}^{4} \gg 1$ (which is quite extended, thanks to the quartic power).
+For each of these sky positions built from~\eqref{eq:soldegensky}, many different values of $(\varphi, \psiL)$ and $(d, \iota)$ produce a waveform very close to the injection, resulting in apparent peaks in the marginalized posterior distribution for the sky position, located at the special sky positions $(\lambda_{a}^{*}, \theta_{L}^{*})$, which are offset from the injected value.
+
+We can further find the same mechanism, with a few differences, by exchanging the roles of $(\iota, \varphi)$ and $(\theta_{L}, \lambda_{a})$, as follows.
+In the limit $\theta_{L} \rightarrow 0$ with $|t_{\theta}| \ll 1$, we have this time
+\be
+	\frac{\sigma_{+}}{\sigma_{-}} \simeq t_{\iota}^{4} e^{-4i(\lambda_{a} - \psi_{L})} \,,
+\ee
+which gives in the same way a degeneracy if
+\bsub\label{eq:soldegeniotaphi}
+\begin{align}
+	\iota^{*} &= 2 \mathrm{Arctan} \; \left| r_{\rm inj}\right|^{1/4} \,, \\
+	(\lambda_{a} - \psiL)^{*} &= -\frac{1}{4} \mathrm{Arg} \; r_{\rm inj} \;\mathrm{mod}\; \frac{\pi}{2} \,,
+\end{align}
+\esub
+with a condition on prefactors that is now concerning $\rho e^{2 i \varphi}$.
+Thus, we find again a degenerate pattern, with the difference that \eqref{eq:soldegeniotaphi} imposes a condition on the pair $\lambda_{a} - \psiL$ instead of $\lambda_{a}$, and the extra condition imposes a condition on $\varphi$ instead of the pair $\varphi - \psiL$.
+
+We illustrate our findings by running a Bayesian parameter estimation for the simplified likelihood. The results are shown in Fig.~\ref{fig:PEsmbh22hmSimpleLikeCase9}, with the two samplers.
+
+Focusing on the sky position, we investigate the degenerate peak in the sky that is the closest to the injection in Fig.~\ref{fig:Skydownsmbh22hmSimpleLikeCase9}.
+We compare here the sky position posterior distribution generated by the two samplers for the true likelihood
+
+%\be\label{eq:symmetryresponse}
+%\betaL \rightarrow -\betaL\,, \quad \iota \rightarrow \pi - \iota \,, \quad \psiL \rightarrow \pi - \psiL
+%\ee
+%leaves the signal unchanged. We can also rewrite the likelihood in a way that makes clearer the symmetry between the two pairs of parameters $(\iota, \varphiL)$ and $(\betaL, \lambdaL)$ \SM{[TODO]}.
+%
+%The simplified extrinsic likelihood for a frozen LISA then takes the form
+%\be\label{eq:simplelikelihood}
+%	\ln \calL = - \frac{1}{2} \Lambda \left( \left| s_{a} - s_{a}^{\rm inj} \right|^{2} + \left| s_{e} - s_{e}^{\rm inj} \right|^{2} \right) \,,
+%\ee
+%where $\Lambda$ appears as an overall normalization factor, playing the role of the square SNR of the signal.
+
+%We can now explore possible degeneracies in the likelihood given by~\eqref{eq:simplelikelihood}, \eqref{eq:defsase} and \eqref{eq:defa22a2m2}-\eqref{eq:defe22e2m2}.
 We note that the relative contributions of the modes $h_{2,2}$ and $h_{2,-2}$ are strongly separated in the face-on or face-off limit, leading to a simplification of the response and an enhanced degeneracy.
@@
 If they are assumed to be negligible, then we can rewrite
-\bsub\label{eq:saseapprox}
-\begin{align}
-	s_{a} \simeq i\calA e^{2 i \xi} \left( D_{a}^{+} + i D_{a}^{\times} \right) \,, \\
-	s_{e} \simeq i\calA e^{2 i \xi} \left( D_{e}^{+} + i D_{e}^{\times} \right) \,,
-\end{align}
-\esub
-where we introduced the notations $\xi (\varphiL, \psiL) \equiv -\varphiL - \psiL$ and $\calA(d, \iota) \equiv 3/4\sqrt{5/\pi}\cos^{4}(\iota/2)/d $ for the common phase and amplitude factors.
-
-For a degeneracy to occur, both complex quantities $s_{a}$ and $s_{e}$ must be close to the injection values $s_{a}^{\rm inj}$ and $s_{e}^{\rm inj}$, which can be achieved as follows if~\eqref{eq:saseapprox} is valid.
+%\bsub\label{eq:saseapprox}
+%\begin{align}
+%	s_{a} \simeq i\calA e^{2 i \xi} \left( D_{a}^{+} + i D_{a}^{\times} \right) \,, \\
+%	s_{e} \simeq i\calA e^{2 i \xi} \left( D_{e}^{+} + i D_{e}^{\times} \right) \,,
+%\end{align}
+%\esub
+%where we introduced the notations $\xi (\varphiL, \psiL) \equiv -\varphiL - \psiL$ and $\calA(d, \iota) \equiv 3/4\sqrt{5/\pi}\cos^{4}(\iota/2)/d $ for the common phase and amplitude factors.
+%
+%For a degeneracy to occur, both complex quantities $s_{a}$ and $s_{e}$ must be close to the injection values $s_{a}^{\rm inj}$ and $s_{e}^{\rm inj}$, which can be achieved as follows if~\eqref{eq:saseapprox} is valid.
 Defining the complex ratio
-\be\label{eq:defr}
-	r(\lambdaL, \betaL) = \frac{D_{a}^{+} + i D_{a}^{\times}}{D_{e}^{+} + i D_{e}^{\times} } \,,
-\ee
-if the position in the sky $(\lambda_{L}^{*}, \beta_{L}^{*})$ can be chosen so that
-\be\label{eq:rcondition}
-	r(\lambda_{L}^{*}, \beta_{L}^{*}) = \frac{s_{a}^{\rm inj}}{s_{e}^{\rm inj}} \,,
-\ee
-then adjusting $\calA$ and $\xi$ so that
-\be
-	i\calA e^{2 i \xi} = \frac{s_{a}^{\rm inj}}{\left( D_{a}^{+} + i D_{a}^{\times} \right)(\lambda_{L}^{*}, \beta_{L}^{*})}
-\ee
-generates degenerate points in parameter space with $\ln \calL \simeq 0$.
-
-The conditions~\eqref{eq:defr}-\eqref{eq:rcondition} can be inverted for the sky position as follows.
+%\be\label{eq:defr}
+%	r(\lambdaL, \betaL) = \frac{D_{a}^{+} + i D_{a}^{\times}}{D_{e}^{+} + i D_{e}^{\times} } \,,
+%\ee
+%if the position in the sky $(\lambda_{L}^{*}, \beta_{L}^{*})$ can be chosen so that
+%\be\label{eq:rcondition}
+%	r(\lambda_{L}^{*}, \beta_{L}^{*}) = \frac{s_{a}^{\rm inj}}{s_{e}^{\rm inj}} \,,
+%\ee
+%then adjusting $\calA$ and $\xi$ so that
+%\be
+%	i\calA e^{2 i \xi} = \frac{s_{a}^{\rm inj}}{\left( D_{a}^{+} + i D_{a}^{\times} \right)(\lambda_{L}^{*}, \beta_{L}^{*})}
+%\ee
+%generates degenerate points in parameter space with $\ln \calL \simeq 0$.
+%
+%The conditions~\eqref{eq:defr}-\eqref{eq:rcondition} can be inverted for the sky position as follows.
 For a given complex value of $r = s_{a}^{\rm inj} / s_{e}^{\rm inj}$, $r(\lambda_{L}^{*}, \beta_{L}^{*}) = r$ can be rewritten as
-\be\label{eq:eqtosolveforlambdabetaL}
-	e^{4i\lambda_{L}^{*}}e^{i\frac{\pi}{3}} \frac{\left( 1 +\sin\beta_{L}^{*} \right)^{2}}{\left( 1 - \sin\beta_{L}^{*} \right)^{2}} = \frac{1+ i r}{1 - i r} \,.
-\ee
-Solving for the modulus of~\eqref{eq:eqtosolveforlambdabetaL} then yields
-\be\label{eq:solutionbetaL}
-	\sin\beta_{L}^{*} = \frac{\rho - 1}{\rho + 1} \,,
-\ee
-for
-\be
-	\rho = \sqrt{\left| \frac{1+ i r}{1 - i r} \right|} \,.
-\ee
-Solving for the argument of~\eqref{eq:eqtosolveforlambdabetaL} gives four solutions for $\lambda_{L}^{*}$ as
-\be\label{eq:solutionlambdaL}
-	\lambda_{L}^{*} = - \frac{\pi}{12} + \frac{1}{4}\mathrm{Arg} \frac{1+ i r}{1 - i r} + \frac{k \pi}{2} \quad \mathrm{for} \quad k = 0,\dots,3 \,.
-\ee
-
-Repeating the argument for the other branch of solutions $\iota \rightarrow \pi$ (or directly considering the symmery~\eqref{eq:symmetryresponse}) gives four other solutions in the sky with identical sky longitudes $\lambda_{L}^{*}$, with symmetrized sky latitudes $\beta_{L}^{*} \rightarrow -\beta_{L}^{*}$, and with $\xi (\varphiL, \psiL) \equiv -\varphiL + \psiL$, $\calA(d, \iota) \equiv 3/4\sqrt{5/\pi}\sin^{4}(\iota/2)/d $.
-
-Thus, with~\eqref{eq:solutionbetaL}, \eqref{eq:solutionlambdaL} and $\beta_{L}^{*} \rightarrow -\beta_{L}^{*}$, we arrive at eight degenerate positions in the sky $(\lambda_{L}^{*}, \beta_{L}^{*})$.
+%\be\label{eq:eqtosolveforlambdabetaL}
+%	e^{4i\lambda_{L}^{*}}e^{i\frac{\pi}{3}} \frac{\left( 1 +\sin\beta_{L}^{*} \right)^{2}}{\left( 1 - \sin\beta_{L}^{*} \right)^{2}} = \frac{1+ i r}{1 - i r} \,.
+%\ee
+%Solving for the modulus of~\eqref{eq:eqtosolveforlambdabetaL} then yields
+%\be\label{eq:solutionbetaL}
+%	\sin\beta_{L}^{*} = \frac{\rho - 1}{\rho + 1} \,,
+%\ee
+%for
+%\be
+%	\rho = \sqrt{\left| \frac{1+ i r}{1 - i r} \right|} \,.
+%\ee
+%Solving for the argument of~\eqref{eq:eqtosolveforlambdabetaL} gives four solutions for $\lambda_{L}^{*}$ as
+%\be\label{eq:solutionlambdaL}
+%	\lambda_{L}^{*} = - \frac{\pi}{12} + \frac{1}{4}\mathrm{Arg} \frac{1+ i r}{1 - i r} + \frac{k \pi}{2} \quad \mathrm{for} \quad k = 0,\dots,3 \,.
+%\ee
+%
+%Repeating the argument for the other branch of solutions $\iota \rightarrow \pi$ (or directly considering the symmery~\eqref{eq:symmetryresponse}) gives four other solutions in the sky with identical sky longitudes $\lambda_{L}^{*}$, with symmetrized sky latitudes $\beta_{L}^{*} \rightarrow -\beta_{L}^{*}$, and with $\xi (\varphiL, \psiL) \equiv -\varphiL + \psiL$, $\calA(d, \iota) \equiv 3/4\sqrt{5/\pi}\sin^{4}(\iota/2)/d $.
+%
+%Thus, with~\eqref{eq:solutionbetaL}, \eqref{eq:solutionlambdaL} and $\beta_{L}^{*} \rightarrow -\beta_{L}^{*}$, we arrive at eight degenerate positions in the sky $(\lambda_{L}^{*}, \beta_{L}^{*})$.
 For each of these sky positions, an exact line degeneracy exists between $\varphiL$ and $\psiL$ as long as $\xi = \mathrm{const}$, as well as an approximate line degeneracy between $d$ and $\iota$, as long as $\calA = \mathrm{const}$, limited to inclination values where $\sin^{4}(\iota/2)$ (or $\cos^{4}(\iota/2)$, for the other branch of solutions) remains negligible.

### 2019-04-04 paper/paper_PE.tex
Work on the paper

@@
 
-\section{The LISA instrument response}
-\label{sec:response}
-
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-
-\subsection{The LISA response - OLD}
-\label{sec:lisaresponse}
-
-\SM{------------------------------------------------------}
-
-\SM{[copy-paste from previous paper -- just here to copy needed formulas / refs]}
-
-We begin by detailing the model that we use for the response of a LISA-like detector, together with the assumptions used and their limitations.
-Since our aim is to assess the accuracy of our direct Fourier-domain treatment of the response, we can focus on the gravitational-wave contribution to the basic single-link observables.
-We therefore use a somewhat simplified model for the time-domain response~\cite{Krolak+04}, ignoring corrections that would be crucial from the point of view of noise cancellations, keeping in mind that the response model can be enriched later without affecting the conclusions of the present analysis.
-
-The frequency-shift response for a single link~\eqref{eq:yslr} was derived in Ref.~\cite{EW75} (see also~\cite{CR02, RCP04, Finn08, Cornish09}).
-Several assumptions enter the result as written in~\eqref{eq:yslr}: (i) effects of the order $v/c$ are neglected, including for instance the special relativistic Doppler effect created by the relative speeds of the spacecraft on their orbits (ii) the propagation is assumed to take place in a flat spacetime, perturbed only by the gravitational wave; thus the gravitational redshift as well as the deflection of light created by the gravitational potential of the Sun is ignored (iii) all geometric factors are evaluated at a single time, whereas one should consider the beam as propagating from the position of the first spacecraft at the time of emission to the position of the second scapecraft at the time of reception, leading to a point-ahead effect.
-
-Additionally, we limit ourselves to a rigid model for the orbits of the constellation, namely we assume that the constellation remains in an equilateral configuration with fixed armlengths.
-These simplified orbits neglect (iv) effects of order $e^{2}$ from the eccentricity of the individual Keplerian orbits, (v) the effect of gravitational perturbations coming from other celestial bodies, such as the Earth, the quadrupole of the Sun, and the other planets.
-Note that although we can neglect all the effects (i)-(v) for our present study, keeping track of these corrections is crucial for the purpose of laser noise cancellations, and led to the development of new generations of TDI observables~\cite{Tintoliving}.
-
-It is natural to split the response~\eqref{eq:yslr} into two steps: first the orbital delay related to the orbit around the Sun of the whole constellation, and then the constellation response.
-The baselines for the delays are indeed very different in the two cases.
-Geometrical projection factors aside, we have for the orbit around the Sun $R=1\text{au}=1.5\times 10^{8} \text{km}$, while the detector armlength is $L=2.5\times 10^{6}\text{km}$ (in the configuration proposed in~\cite{LISA17}).
-It is natural to define two transfer frequencies for the two relevant length scales for the delays, defined such that a wavelength fits within this length scale, i.e.
-$2\pi f d = 1$.
-This gives
-\begin{subequations}\label{eq:transferfrequencies}
-\begin{align}
-	f_{R} &= 3.2\times10^{-4}\Hz \,,\\
-	f_{L} &= 1.9\times 10^{-2}\Hz \,.
-\end{align}
-\end{subequations}
-The LISA response will behave qualitatively differently on the three frequency bands $f \leq f_{R}$, $f_{R} \leq f \leq f_{L}$ and $f_{L} \leq f$.
-
-The first stage of the response, the orbital delay, consists simply in applying the varying time delay to bring the wavefront sampling point from the SSB reference to the center of the LISA triangular constellation, common to all $y_{slr}$ observables.
-For $h^{\rm TT}$ the transverse-traceless gravitational waveform in matrix form, we write this orbital time delay as
-\be\label{eq:defresponse0}
-	h_{0}^{\rm TT} (t) = h^{\rm TT}(t-\hatk\cdot p_{0}) \,,
-\ee
-with $p_{0}$ the position of the constellation center, which follows the Earth orbit around the Sun.
-The second stage of the response calculation comprises the remaining, constellation-centered response.
-For the single-link contribution to the response, for the laser link from spacecraft $s$ to spacecraft $r$ along a path in direction $n_l$, we write
-\begin{align}\label{eq:defresponseL}
-	y_{slr} &= \frac{1}{2} \frac{1}{1 - \hatk\cdot n_{l}} \nn\\
-	& \cdot n_{l}\cdot \left[ h_{0}^{\rm TT}(t - L - \hatk\cdot p^{L}_{s}) - h_{0}^{\rm TT}(t - \hatk\cdot p^{L}_{r}) \right] \cdot n_{l}\,,
-\end{align}
-where we reference the positions of the spacecraft relative to the center of the constellation, $p^{L}_{A} \equiv p_{A} - p_{0}$.
-
-As described in App.~\ref{app:notation} We will decompose the full signal in the contributions of the individual spin-weighted spherical modes $h_{\ell m}$, whose Fourier transforms are assumed to have a smooth amplitude and phase.
-First, we define the matrices $P_{+},P_{\times}$ such that, in the sense of matrices,
-\be
-	h^{\rm TT} = h_{+}P_{+} + h_{\times}P_{\times} \,.
-\ee
-We focus only on positive frequencies.
-Assuming that the approximation~\eqref{eq:zeronegativef} applies, we consider a single mode contribution, $h=h_{\ell m}$ with $m>0$.
-For each given mode we define a complex matrix $P_{\ell m}$ incorporating the spin-weighted spherical harmonic constant factor as
-\be
-	P_{\ell m} =
-	\begin{cases}
-	\frac{1}{2} {}_{-2}Y_{\ell m} \left( P_{+} + i P_{\times} \right) \text{ for } m>0\,,\\
-	\frac{1}{2} {}_{-2}Y_{\ell m}^{*} \left( P_{+} - i P_{\times} \right) \text{ for } m<0\,.
-	\end{cases}
-\ee
-
-We now turn to the transformation of~\eqref{eq:defresponse0} and~\eqref{eq:defresponseL} to the Fourier domain. Applying a pure delay as in~\eqref{eq:defresponse0} translates into
-\be\label{eq:G0}
-	G_{0}(f, t) = e^{-2i\pi f d_{0}(t)} \,,
-\ee
-with $d_{0} = -\hatk \cdot p_{0}$ the delay associated to the orbit around the Sun.
-For the leading-order response~\eqref{eq:transferlocal}, this gives a Fourier-domain transfer function common to all modes, that is a pure phase factor, proportional to the frequency but also $t_{f}$-dependent:
-\be\label{eq:transfer0local}
-	\calT_{0}^{\rm local}(f) = G_{0}(f, \tf)\,.
-\ee
-If $(\lambda, \beta)$ are the ecliptic longitude and latitude of the source in the sky, and if the orbital phase is set by convention to $0$ at $t=0$, the orbital delay has the simple expression
-\be\label{eq:delay0}
-	d_{0}(t) = -R \cos\beta \cos\left(\Omega_{0}t - \lambda\right)\,.
-\ee
-
-For the constellation part of response~\eqref{eq:defresponseL}, treated separately from the delay~\eqref{eq:delay0}, we write
-\begin{align}\label{eq:decomposeGslr}
-	F_{slr}^{L}(t) &= \frac{1}{2} \frac{1}{1 - \hatk\cdot n_{l}(t)} n_{l}(t) \cdot P_{\ell m} \cdot n_{l} (t) \,,\nn\\
-	d_{s}(t) &= - k\cdot p_{s}^{L}(t) \,, \quad d_{r}(t) = - k\cdot p_{r}^{L}(t) \,,\nn\\
-	G_{slr}^{L}(f,t) &=  F_{slr}^{L}(t) \left( e^{-2i\pi f (d_{s,L}(t) + L)} - e^{-2i\pi f d_{r}(t)} \right) \,.
-\end{align}
-The superscript $L$ indicates that the orbital delay~\eqref{eq:delay0} is not included.
-Since we also assume the rigid approximation for the constellation, where the armlengths are fixed, a particular simplification occurs when combining these individual delays, thanks to the relation $p^{L}_{r} - p^{L}_{s} =  L n_{l}$:
-\begin{align}\label{eq:GslrL}
-	G_{slr}^{L}(f,t) &= \frac{i \pi f L}{2} \sinc \left[ \pi f L\left(1-\hatk\cdot n_{l} \right) \right] \nn\\
-	& \quad \cdot \exp\left[ i \pi f \left( L + \hatk\cdot \left( p_{1}^{L} + p_{2}^{L} \right) \right) \right]  n_{l} \cdot P_{\ell m} \cdot n_{l} \,,
-\end{align}
-with all time-dependent vectors evaluated at $t$.
-This expression is well known as describing the frequency-dependency in the LISA response~\cite{Larson+99, Cornish01, CR02, RCP04}.
-In the local approximation~\eqref{eq:transferlocal}, the Fourier-domain transfer function then reads
-\begin{align}\label{eq:transferLlocal}
-	\calT_{slr}^{L, \mathrm{local}}(f) &= G_{slr}^{L}(f, \tf) \,.
-\end{align}
-For plotting purposes, we will also define
-\be\label{eq:transferLenvelope}
-	\overline{\calT}_{slr}^{L} (f) = \frac{i \pi f L}{2} n_{l} \cdot P_{\ell m} \cdot n_{l} (\tf)
-\ee
-which will serve as an estimate for the enveloppe function of the response, devoid of the zero-crossings at high frequencies of the $\sinc$ term in~\eqref{eq:GslrL}.
-
-Note that, if the corrections of Sec.~\ref{subsec:delays}, for non-negligible $\dot d$, are included for the constellation delays, the transfer function will not have this simple form anymore, as $\dot{d}$ will have a different velocity-dependent expression for the sending and receiving spacecraft.
-One must then separately handle $d_{s}$ and $d_{r}$ in~\eqref{eq:decomposeGslr} to compute the corrections.
-
-The orbital response~\eqref{eq:transfer0local}-\eqref{eq:delay0} takes a simple analytic form, but the phase contribution of this delay is significant across most of the frequency band and can be large for $f \gg f_{R}$.
-
-The constellation response~\eqref{eq:GslrL}-\eqref{eq:transferLlocal} can be interpreted as the Fourier-domain translation of a discrete derivative taken on the waveform.
-The leading factor in~\eqref{eq:GslrL} shows that the amplitude of the response is proportional to $f$ in the low-frequency limit $f\ll f_{L}$, where the other factors are essentially unity.
-For $f\gtrsim f_{L}$, the $\sinc$ and the phase of the exponential generate additional structure in the response, including zero-crossings when the projected armlength is an integer number of wavelengths.
-From~\eqref{eq:transferLlocal}, an expansion for small $f\ll f_{L}$ yields back a Fourier-domain analog of the low-frequency approximation of the response~\cite{Cutler97, RCP04}, which is equivalent to having two LIGO-type interferometers turned by $\pi/4$ and set in motion.
-
-For analysis of the response we need a concrete set of gravitational waveforms.
-We will use the PhenomD model~\cite{Khan+15,Husa+15}, which provides Fourier-domain inspiral-merger-ringdown waveforms for aligned spins.
-We refer to App.~\ref{app:precLISA} for a brief discussion of the prospects for applying our formalism for the LISA response to precessing Fourier-domain waveforms.
-
-\SM{------------------------------------------------------}
-
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-
-\subsection{The GW signal}
-\label{sec:gwsignal}
-
-We start by introducing a source-frame $(\hat{x}_{S}, \hat{y}_{S}, \hat{z}_{S})$ attached to the binary system emitting gravitational waves.
-For comparable-mass systems without spin, the natural choice is to take the normal to the orbital plane as $\hat{z}_{S}$, and we will assume that the remaining rotation around $\hat{z}_{S}$ is fixed by the phase convention of the waveform model (see Sec.~\ref{}).
-
-Introducing $k$ the wave propagation vector going from the source towards the observer, we define the inclination $\iota$ and the observer phase $\varphi$ simply as being its spherical angular coordinates $(\theta_{S}, \phi_{S})$, so that in the source-frame
-\be
-	k_{S} = (\sin\iota \cos\varphi, \sin\iota \sin\varphi, \cos\iota) \,.
-\ee
-
-By convention, for our polarization vectors $p$ and $q$ we will use the spherical coordinate vectors $p = e_{\theta}^{S}$, $q = e_{\phi}^{S}$. In terms of only $\hat{z}_{S}$,
-\bsub
-\begin{align}
-	p &= q \times k \,,\\
-	q &= \hat{z}_{S} \times k / |\hat{z}_{S} \times k|\,.
-\end{align}
-\esub
-If $H_{ij} = h_{ij}^{\rm TT}$ represents the gravitational wave signal in the transverse-traceless gauge,
-\be
-	H = h_{+} P_{+} + h_{\times} P_{\times}
-\ee
-with the polarization tensors
-\bsub
-\begin{align}
-	P_{+} &= p \otimes p - q \otimes q \,,\\
-	P_{\times} &= p \otimes q + q \otimes p \,.
-\end{align}
-\esub
-Conversely, the polarizations are
-\bsub
-\begin{align}
-	h_{+} &= \frac{1}{2} \left(p \otimes p - q \otimes q \right) : H \,,\\
-	h_{\times} &= \frac{1}{2} \left( p \otimes q + q \otimes p \right) : H \,.
-\end{align}
-\esub
-
-We will further decompose the gravitational wave signal, seen as a function of the direction of emission $(\iota, \varphi)$, in spin-weighted spherical harmonics as
-\be\label{eq:hpcmodes}
-	h_{+} - i h_{\times} = \sum_{\ell \geq 2} \sum_{m = -\ell}^{\ell} {}_{-2}Y_{\ell m} (\iota, \varphi) h_{\ell m} \,,
-\ee
-where the explicit expression of the ${}_{-2}Y_{\ell m}$ is given in~\cite{}.
-
-The vectors $(p,q,k)$ form the Wave-frame, such that the gravitational wave takes in this frame the familiar form
-\be
-	H_{W} = \begin{pmatrix}
-		h_{+} & h_{\times} & 0 \\
-		h_{\times} & -h_{+} & 0 \\
-		0 & 0 & 0
-		\end{pmatrix} \,.
-\ee
-
-We will both generate the waveforms and apply the response directly in the Fourier domain. Our convention for the Fourier transform is
-\be\label{eq:defFourier}
-	\tilde{F}(f) = \int dt \; e^{2 i \pi f t} F(t) \,.
-\ee
-Note that this differs by a change $f\rightarrow -f$ from the most usual one (used e.g. in \texttt{LAL}).
-
-The definition of the reference time and phase \dots
-
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-
-\subsection{LISA trajectories}
-\label{sec:lisatraj}
-
-For completeness, we gather in this appendix explicit expressions for the trajectory of the LISA constellation and the rotation matrices entering the response.
-
-The orbits of the three spacecrafts around the Sun can be chosen so that, at leading order in the eccentricity of the orbits, the constellation retains the shape of an equilateral triangle in a cartwheeling motion following the Earth orbit.
-We denote by $a$ the semi-major axis, $e$ the eccentricity and pose $\alpha = \Omega_{0} (t-t_{0}))$, with $t_{0}$ a reference time for the initial position ($t_{0} = 0$ in our case).
-The reference SSB frame is $(\hat{x}, \hat{y}, \hat{z})$.
-
-Using the notation $c,s = \cos, \sin \alpha$, the trajectory of the center of the constellation is simply
-\be
-	p_{0} = ac \hat{x} + as \hat{y} \,,
-\ee
-and we take $a\equiv R = 1 \, \mathrm{au}$.
-We denote the positions of the spacecrafts as $p_{A}$ for $A=1,2,3$, and the positions relative to the constellation center as $p_{A}^{L} = p_{A} - p_{0}$.
-Setting $\beta_{A} = 2(A-1)\pi/3 + \beta_{0}$ (with $\beta_{0}$ and initial condition set to 0 in our case), the Cartesian coordinates in the SSB-frame of the position of the spacecrafts read~\cite{}
-\bsub
-\begin{align}
-	p_{A}^{L} &= a e \left[ \sin \beta_{A} c s - \cos\beta_{A} \left( 1 + s^{2} \right) \right] \hat{x} \nn\\
-	& + a e \left[ \cos \beta_{A} c s - \sin\beta_{A} \left( 1 + c^{2} \right) \right] \hat{y} \nn\\
-	& - a e \sqrt{3} \cos(\alpha - \beta_{A}) \hat{z} \,,
-\end{align}
-\esub
-The armlength, constant in this approximation, is related to the eccentricity and semi-major axis by
-\be
-	L = 2\sqrt{3} a e \,.
-\ee
-The rigid approximation for the constellation, at first order in $e$, can be seen as being at first order in $L/R \simeq 0.017$ for a $2.5 \, \mathrm{Gm}$ armlength~\cite{}.
-
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-
-\subsection{The LISA frame}
-\label{sec:LISAframe}
-
-It will be very useful to introduce a time-dependent frame $(\hat{x}_{L}, \hat{y}_{L}, \hat{z}_{L})(t)$ following the detector in its motion.
-Specifically, we choose this LISA frame such that at any time,
-\bsub
-\begin{align}
-	p_{1}^{L} &= - \frac{L}{\sqrt{3}} \hat{x}_{L} \,,\\
-	p_{2}^{L} &= \frac{L}{2\sqrt{3}} \hat{x}_{L} - \frac{L}{2} \hat{y}_{L} \,,\\
-	p_{3}^{L} &= \frac{L}{2\sqrt{3}} \hat{x}_{L} + \frac{L}{2} \hat{y}_{L} \,.\\
-\end{align}
-\esub
-
-This frame also provides us with an equivalent representation of the trajectories making use of rotation matrices.
-Since in our rigid approximation the constellation remains an equilateral triangle in its cartwheeling motion around the Sun, the configuration of the constellation at a later time is given by a rotation around the Sun composed with a rotation of the constellation around its symmetry axis.
-
-If $R(v,\varpi)$ denotes the matrix of an active rotation around the vector $v$ by an angle $\varpi$, and if we denote by $R_{L}$ the active rotation from the SSB-frame to the L-frame such that for each of the three basis vectors $(\hat{x}_{L}, \hat{y}_{L}, \hat{z}_{L}) = R_{L} (\hat{x}, \hat{y}, \hat{z})$, we have
-\be\label{eq:RL}
-	R_{L} = R(z, \alpha) \cdot R(y, -\pi/3) \cdot R(z, -\alpha) \,,
-\ee
-where we recall that $\alpha = \Omega_{0} (t-t_{0}) $.
-For all vectors $X$ among $p_{A}^{L}$, $n_{A}$ we have $X(t) = R_{L} \cdot X(t=t_{0})$.
-If a vector $X$ is given by its components in the SSB-frame, in the L-frame the components are $X_{L} = R_{L}^{-1} \cdot X$.
-
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-
-\subsection{From the Wave frame to the SSB frame}
-\label{sec:wavessbframe}
-
-We can now relate the Wave-frame to the SSB-frame, which will fix our convention for the sky position and polarization angle.
-The source position in the sky is given by $(\lambda, \beta)$, the ecliptic longitude and latitude in the SSB-frame.
-Like with the Source-frame, we can introduce spherical vectors $(e_{r}^{SSB}, e_{\theta}^{SSB}, e_{\phi}^{SSB})$.
-Since the propagation vector is $k = - e_{r}^{SSB}$,
-\be
-	k = (- \cos\beta\cos\lambda, - \cos\beta\sin\lambda, -\sin\beta) \,.
-\ee
-
-The last degree of freedom between the frames represents a rotation along the line-of-sight, parametrized by the polarization angle $\psi$.
-We introduce reference polarization vectors in the SSB-frame as
-\bsub
-\begin{align}
-	u &= \hat{z} \times k / |\hat{z} \times k| \,,\\
-	v &= k \times u \,.
-\end{align}
-\esub
-In terms of spherical vectors, $(k, u, v) = (-e_{r}^{SSB}, -e_{\phi}^{SSB}, -e_{\theta}^{SSB})$.
-The polarization angle $\psi$ is then defined such that $(p,q)$ are obtained by rotating $(u,v)$ by the angle $\psi$ around $k$, i.e.
-\be
-	(p,q) = R(k, \psi) \cdot (u,v) \,.
-\ee
-
-These relations can be summarized by the active rotation matrix $R_{W}$ from the SSB-frame to the Wave-frame:
-\be\label{eq:RW}
-	R_{W} = R(z, \lambda - \pi/2) \cdot R(x, \beta + \pi/2 ) \cdot R(z, \psi)\,.
-\ee
-
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-
-\subsection{Parametrizing the signal in the LISA-frame}
-\label{sec:LISAframeparams}
-
-While long-lasting signals like SOBHs will see a strong imprint of the LISA orbital motion over the course of observation, SMBH signals are dominated in SNR by a short-lived burst of emission at merger.
-For such signals, it will be useful to use a parametrization based on the L-frame at the time of merger instead of the SSB frame.
-
-The new parameters are defined as playing the same role as the SSB parameters, but relative to the L-frame.
-Namely, we define $R_{LW}$ the active rotation matrix from the L-frame to the Wave-frame, expressed in the L-frame basis, so that for each basis vector expressed in the L-frame basis $(\hat{x}_{W}, \hat{y}_{W}, \hat{z}_{W})_{L} = R_{LW} \cdot (\hat{x}_{L}, \hat{y}_{L}, \hat{z}_{L})_{L}$.
-The defining condition on $(\lambda_{L}, \beta_{L}, \psi_{L})$ is
-\be\label{eq:defRLW}
-	R_{LW} = R(z, \lambda_{L} - \pi/2) \cdot R(x, \beta_{L} + \pi/2 ) \cdot R(z, \psi_{L}) \,.
-\ee
-
-Coming back to vectors expressed in the SSB-basis, $R_{L}^{-1} \cdot (\hat{x}_{W}, \hat{y}_{W}, \hat{z}_{W}) = R_{LW} \cdot R_{L}^{-1} \cdot (\hat{x}_{L}, \hat{y}_{L}, \hat{z}_{L})$.
-Since $ (\hat{x}_{W}, \hat{y}_{W}, \hat{z}_{W}) = R_{W} \cdot (\hat{x}, \hat{y}, \hat{z})$ and $ (\hat{x}_{L}, \hat{y}_{L}, \hat{z}_{L}) = R_{L} \cdot (\hat{x}, \hat{y}, \hat{z})$, we arrive at
-\be\label{eq:RLW}
-	R_{LW} = R_{L}^{-1} \cdot R_{W} \,.
-\ee
-
-Combining~\eqref{eq:RLW}, \eqref{eq:RL}, \eqref{eq:RW} and~\eqref{eq:defRLW} yields the following expressions for the L-frame angular parameters\footnote{The convention here is that the point of coordinates $(x,y)$ in the plane has for argument $\arctan\left[ x, y \right]$.}:
-\bsub
-\begin{align}
-	\beta_{L} &= \arcsin \left[ \cos \frac{\pi}{3} \sin \beta - \sin \frac{\pi}{3} \cos \beta \cos \left( \lambda - \alpha \right) \right] \,,\\
-	\lambda_{L} &= \arctan \left[ \cos\beta \cos\lambda \left( \cos \frac{\pi}{3} \cos^{2}\alpha + \sin^{2}\alpha \right) \right. \nn\\
-	& \qquad\quad\;\; \left. + \cos\beta \sin\lambda \cos\alpha \sin\alpha \left( \cos\frac{\pi}{3} - 1 \right) \right. \nn\\
-	& \qquad\quad\;\; \left. + \sin\frac{\pi}{3} \sin\beta \cos\alpha , \right. \nn\\
-	& \qquad\qquad\; \left. \cos\beta \sin\lambda \left( \cos \frac{\pi}{3} \sin^{2}\alpha + \cos^{2}\alpha \right) \right. \nn\\
-	& \qquad\quad\;\; \left. + \cos\beta \cos\lambda \cos\alpha \sin\alpha \left( \cos\frac{\pi}{3} - 1 \right) \right. \nn\\
-	& \qquad\quad\;\; \left. + \sin\frac{\pi}{3} \sin\beta \sin\alpha \right] \,,\\
-	\psi_{L} &= \arctan \left[ \cos\psi \left( \cos\frac{\pi}{3}\cos\beta + \sin\frac{\pi}{3} \sin\beta \cos(\lambda - \alpha)\right) \right. \nn\\
-	& \qquad\quad\;\; \left. + \sin \psi \sin\frac{\pi}{3} \sin(\lambda - \alpha), \right. \nn\\
-	& \qquad\qquad\; \left. \sin\psi \left( \cos\frac{\pi}{3}\cos\beta + \sin\frac{\pi}{3} \sin\beta \cos(\lambda - \alpha)\right) \right. \nn\\
-	& \qquad\quad\;\; \left. - \cos \psi \sin\frac{\pi}{3} \sin(\lambda - \alpha) \right] \,.
-\end{align}
-\esub
-
-The definition of the reference phase $\phi_{\rm Ref}$, \dots \SM{[COMPLETE]}
-
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-
-\subsection{Mode decomposition and polarization angle}
-\label{sec:modespol}
-
-We now translate the mode decomposition~\eqref{eq:hpcmodes} in the Fourier domain, and make more explicit the dependence of the signal in the polarization $\psi$, so as to make manifest its purely extrinsic nature.
-Indeed, $\psi$ enters the observables only through constant prefactors, which is a bit obscured by the rotation matrix formulation~\eqref{eq:RW}.
-
-The mode decomposition~\eqref{eq:hpcmodes} gives (dropping $(\iota, \varphi)$)
-\bsub
-\begin{align}
-	h_{+} = \frac{1}{2} \sum_{\ell, m} \left( \sYlm h_{\ell m} + \sYlmstar h_{\ell m}^{*} \right) \,,\\
-	h_{\times} = \frac{i}{2} \sum_{\ell, m} \left( \sYlm h_{\ell m} - \sYlmstar h_{\ell m}^{*} \right) \,,
-\end{align}
-\esub
-which is valid in general. Now, for non-precessing binary systems, an exact symmetry relation between modes reads
-\be\label{eq:nonprecsymmetry}
-	h_{\ell, -m} = (-1)^{\ell} h_{\ell, m}^{*} \,.
-\ee
-Using this symmetry, we can write
-\be
-	h_{+,\times} = \sum_{\ell, m} K_{\ell m}^{+, \times} h_{\ell m} \,, 
-\ee
-with
-\bsub
-\begin{align}
-	K_{\ell m}^{+} =\frac{1}{2} \left( \sYlm + (-1)^{\ell} \sYlminusmstar \right) \,,\\
-	K_{\ell m}^{\times} = \frac{i}{2} \left( \sYlm - (-1)^{\ell} \sYlminusmstar \right) \,.
-\end{align}
-\esub
-
-Going to the Fourier domain, an approximation often used is to neglect support for negative/positive frequencies according to
-\be\label{eq:approxnegf}
-	\tilde{h}_{\ell m} (f) \simeq 0 \;\; \text{for} \;\; m<0, \; f>0 \;\; ( m>0, \; f<0 )\,,
-\ee
-and neglecting modes $h_{\ell 0}$.
-We will use this approximation throughout this paper.
-Note that we picked our Fourier convention~\eqref{eq:defFourier} to ensure that the mode $\tilde{h}_{22} (f)$ has support for $f>0$.
-Using~\eqref{eq:approxnegf},
-\be
-	\tilde{h}_{+,\times} = \sum_{\ell} \sum_{m>0} K_{\ell m}^{+, \times} \tilde{h}_{\ell m} \,.
-\ee
-
-Next, it is convenient to introduce mode-by-mode polarization matrices
-\be
-	P_{\ell m} = P_{+} K_{\ell m}^{+} + P_{\times} K_{\ell m}^{\times} \,,
-\ee
-so that
-\be\label{eq:Hsummodes}
-	H = \sum_{\ell m} P_{\ell m} h_{\ell m} \,.
-\ee
-To make explicit the dependence in polarization, we can define polarization tensors for 0 polarization angle as
-\bsub
-\begin{align}
-	P_{+}^{0} &= P_{+}(\psi = 0) = u \otimes u - v \otimes v \,,\\
-	P_{\times}^{0} &= P_{\times}(\psi = 0) = u \otimes v + v \otimes u \,.
-\end{align}
-\esub
-This allows to write the dependence in polarization as
-\be
-	P_{+} + i P_{\times} = e^{-2 i \psi} \left( P_{+}^{0} + i P_{\times}^{0} \right) \,,
-\ee
-or explicitly in $P_{\ell m}$ as
-\begin{align}
-	P_{\ell m} &= \frac{1}{2} \sYlm e^{-2 i \psi} \left( P_{+}^{0} + i P_{\times}^{0} \right) \nn\\
-	& + \frac{1}{2} (-1)^{\ell} \sYlminusmstar e^{+2 i \psi} \left( P_{+}^{0} - i P_{\times}^{0} \right) \,.
-\end{align}
-Writing the $P_{\ell m}$ matrices in this way allows to factor out all dependencies in the extrinsic parameters $(\iota, \phi, \psi)$.
-Together with the luminosity distance $D_{L}$, these parameters appear in combinations of constant factors multiplying the whole response.
-
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-
-\subsection{The LISA observables}
-\label{sec:LISAobservables}
-
-The single-link frequency shift observable $y_{slr} = (\nu_{r} - \nu_{s})/\nu$ between sending spacecraft $s$ and receiving spacecraft $r$ along the link $l$ takes the form 
-\be\label{eq:defyslr}
-	y_{slr} = \frac{1}{2} \frac{n_{l} \otimes n_{l}}{1 - k\cdot n_{l}} : \left[ H(t - L - k\cdot p_{s}) - H(t - k\cdot p_{r}) \right] \,.
-\ee
-Several assumptions enter this form of the response: \dots \SM{[COMPLETE]}.
-
-In the frequency domain, we will use the analysis of~\cite{} and write the response in individual observables $y_{slr}$ with a transfer function for each spherical harmonic mode as
-\be
-	\tilde{y}_{slr} = \sum_{\ell, m}\calT_{slr}^{\ell m}(f) \tilde{h}_{\ell m} \,.
-\ee
-With the decomposition~\eqref{eq:Hsummodes}, the corresponding form of the kernel $G^{\ell m}_{slr}(f, t)$ in the formalism of~\cite{} is
-\begin{align}\label{eq:Gslr}
-	G_{slr}^{\ell m}(f,t) &= \frac{i \pi f L}{2} \sinc \left[ \pi f L\left(1-k\cdot n_{l} \right) \right] \nn\\
-	& \quad \cdot \exp\left[ i \pi f \left( L + k\cdot \left( p_{1} + p_{2} \right) \right) \right]  n_{l} \cdot P_{\ell m} \cdot n_{l} \,.
-\end{align}
-At leading order in the separation of timescales, we have simply
-\be
-	\calT_{slr}^{\ell m}(f) = G_{slr}^{\ell m}(f, t_{f}^{\ell m}) \,,
-\ee
-with $t_{f}^{\ell m}$ the effective time-to-frequency correspondence, defined across the whole frequency band, including in the merger-ringdown, as $t_{f}^{\ell m} = -1/(2\pi) d\Psi_{\ell m} / df$ for a mode $\tilde{h}_{\ell m} = A_{\ell m}(f) \exp\left[-i \Psi_{\ell m}\right]$.
-The analysis of~\cite{} has shown that higher-order corrections are small for SMBH systems, and for SOBH systems that are close enough to merger as will be the case for the examples chosen here.
-The fact that we use the same Fourier-domain treatment of the transfer functions for both the signal and the templates should also mitigate errors.
-
-In our rigid approximation, delays are simply all constant and constant to $L$. Using the shortcut notation $y_{sr,L} = y_{slr}(t - L)$, the first-generation TDI Michelson observable $X$ reads
-\begin{align}
-	X &= y_{31} + y_{13,L} + \left( y_{21} + y_{12,L} \right)_{2L} \nn\\
-	& - \left( y_{21} + y_{12,L} \right) - \left( y_{31} + y_{13,L} \right)_{2L} \,,
-\end{align}
-with the other Michelson observables $Y$, $Z$ being obtained by cyclic permutation. The independent combinations $A$, $E$ and $T$ are then defined as
-\bsub
-\begin{align}
-	A &= \frac{1}{\sqrt{2}} \left( Z - X \right) \,,\\
-	E &= \frac{1}{\sqrt{6}} \left( X - 2Y + Z \right) \,,\\
-	T &= \frac{1}{\sqrt{3}} \left( X + Y + Z \right) \,.
-\end{align}
-\esub
-These channels are independent under the approximation of an uncorrrelated noise between the arms. With constant delays in the rigid approximation, the TDI combinations take the form
-\bsub\label{eq:aet}
-\begin{align}
-	\tilde{a} &= \left[ (1+z) \left( \tilde{y}_{31} + \tilde{y}_{13} \right) - \tilde{y}_{23} - z \tilde{y}_{32} - \tilde{y}_{21} - z \tilde{y}_{12} \right] \,,\\
-	\tilde{e} &= \frac{1}{\sqrt{3}} \left[ (1-z)\left( \tilde{y}_{13} - \tilde{y}_{31} \right) + (2+z) \left( \tilde{y}_{12} - \tilde{y}_{32} \right) \right. \nn\\
-	&\qquad\quad \left. + (1+2z) \left( \tilde{y}_{21} - \tilde{y}_{23} \right) \right] \,,\\
-	\tilde{t} &= \frac{\sqrt{2}}{\sqrt{3}} \left[ \tilde{y}_{21} - \tilde{y}_{12} + \tilde{y}_{32} - \tilde{y}_{23} + \tilde{y}_{13} - \tilde{y}_{31} \right] \,,
-\end{align}
-\esub
-where we have introduced the rescalings
-\bsub\label{eq:scalingAET}
-\begin{align}
-	\tilde{a}, \tilde{e} &= \frac{1}{i \sqrt{2} z \sin 2\pi f L}\times \tilde{A}, \tilde{E} \,,\\
-	\tilde{t} &= \frac{1}{2\sqrt{2} z^{3} \sin \pi f L \sin 2\pi f L} \times \tilde{T} \,.
-\end{align}
-\esub
-Scaling out the same square factors from the noise power spectral density as
-\bsub
-\begin{align}
-	S_{n}^{A}, S_{n}^{E} &= 2 \sin^{2} 2\pi f L \times S_{n}^{a}, S_{n}^{e} \,,\\
-	S_{n}^{T} &= 8 \sin^{2} \pi f L \sin^{2} 2\pi f L S_{n}^{t} \,,
-\end{align}
-\esub
-the reduced PSD for the three channels take the form
-\bsub
-\begin{align}
-	S_{n}^{a} = S_{n}^{e} &= 2 \left( 3 + 2\cos 2\pi f L + \cos 4 \pi f L \right) S^{\rm pm}(f) \nn\\
-	& \quad + \left( 2 + \cos 2\pi f L \right) S^{\rm op}(f) \,,\\
-	S_{n}^{t} &= 4 \sin^{2} 2\pi f L S^{\rm pm}(f) + S^{\rm op}(f) \,,
-\end{align}
-\esub
-with $S^{\rm pm}$ the test-mass noise PSD and $S^{\rm op}$ the optical noise PSD.
-Here we ignore the confusion noise coming from the background of galactic binaries in the  LISA band.
-Note that the prefactors~\eqref{eq:scalingAET} are oscillatory and have zero-crossings at high frequencies, and scaling them out from both the signal and the noise is an approximation, as we can expect that the real delays would lead to imperfect cancellations in the vicinity of the zero-crossings.
-
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-
-\subsection{The low-frequency limit}
-\label{sec:low-freq}
-
-As is well known~\cite{Cutler97}, in the low-frequency limit (also called the long-wavelength approximation), the finite-length effects of the arms of LISA disappear, and the response of the detectors falls back to the one of two LIGO-type detectors set in motion, rotated from each other by $\pi/4$.
-
-For $f \ll f_{L}$, we have $2\pi f L \ll 1$ and the kernel~\eqref{eq:defGslr} reduces to
-\be\label{eq:Gslrlowf}
-	G^{\ell m}_{slr} \simeq \frac{i \pi f L}{2} \exp\left[ 2 i \pi f k\cdot p_{0} \right] n_{l} \otimes n_{l} : P_{\ell m}\,.
-\ee
-For even lower frequencies, when $f \ll f_{R}$ as well, $2\pi f R \ll 1$ and the delay term is negligible: $\exp\left[ 2 i \pi f k\cdot p_{0} \right] \simeq 1$.
-
-In the following, we will also apply frequency-domain response at leading order in the separation of timescales~\cite{MB18}, so that the transfer functions are $\calT(f) = G(f, \tf)$ with $\tf$ the time-of-frequency function~\eqref{eq:deftf}.
-For $2\pi f L \ll 1$, we have $z\simeq 1$, and the link reversal symmetry $\tilde{y}_{slr} \simeq \tilde{y}_{r-ls}$, so that~\eqref{eq:aet} becomes
-\bsub\label{eq:aetlowf}
-\begin{align}
-	\tilde{a} &\simeq 4\tilde{y}_{31} - 2 \tilde{y}_{23} - 2 \tilde{y}_{12} \,,\\
-	\tilde{e} &\simeq 2\sqrt{3} \left[ \tilde{y}_{12}  - \tilde{y}_{23} \right] \,,\\
-	\tilde{t} &\simeq 0 \,,
-\end{align}
-\esub
-with the $T$-channel becoming negligible in this limit. Using~\eqref{eq:Gslrlowf}, we can write
-\be\label{eq:aelowfmodes}
-	\tilde{a}, \tilde{e} = \sum_{\ell, m>0} (-2i\pi f) \exp\left[ 2 i \pi f k\cdot p_{0} \right] \tilde{h}_{\ell m} D_{a,e} : P_{\ell m} \,,
-\ee
-where we introduced the detector tensors
-\bsub\label{eq:DaDe}
-\begin{align}
-	D_{a} &= \frac{L}{2} \left( n_{1}\otimes n_{1} + n_{3} \otimes n_{3} - 2 n_{2} \otimes n_{2} \right) \,,\\
-	D_{e} &= \frac{L\sqrt{3}}{2} \left( n_{1} \otimes n_{1} - n_{3} \otimes n_{3} \right) \,.
-\end{align}
-\esub
-Here, we have made apparent factors $(-2i\pi f)$ corresponding to the Fourier-domain translation of a time derivative.
-Indeed, the observables $\Delta \nu / \nu$ used here are time derivatives of the laser phase more commonly used for ground-based detectors.
-
-We can now map these two channels to two fictitious LIGO-type detectors as follows.
-For an orthogonal detector of the same setup as the ground-based LIGO and Virgo with armlength $L_{D}$, rotated by an angle $\epsilon_{D}$ from the basis vectors $x,y$, the detector tensor is
-\begin{align}
-	D &= \frac{L_{D}}{2} \left[ \cos 2\epsilon_{D} \left( x\otimes x - y\otimes y \right) \right. \nn\\
-	& \qquad\quad + \left. \sin 2\epsilon_{D} \left( x\otimes y + y \otimes x\right) \right] \,.
-\end{align}
-By comparison with~\eqref{eq:DaDe}, we obtain
-\be
-	L_{a} = L_{e} = \frac{3L}{2} \,, \; \epsilon_{a} = \frac{2\pi}{3} \,, \; \epsilon_{e} = \frac{5\pi}{12} \,.
-\ee
-Note a degeneracy $\pm \pi$ in how we choose to orient these effective detectors. We can recover pattern functions (for $\psi_{L}=0$) by factoring out the effective length, defining
-\be\label{eq:defFapcFepc}
-	F_{a,e}^{+,\times} \equiv \frac{2}{3L} D_{a,e} : P_{+,\times}^{0} \,,
-\ee
-which gives in terms of the LISA-frame sky position angles
-\bsub\label{eq:FapcFepc}
-\begin{align}
-	F_{a}^{+} &= \frac{1}{2} \left( 1 + \sin^{2}\beta_{L} \right) \cos \left(2\lambda_{L} - \frac{\pi}{3} \right) \,,\\
-	F_{a}^{\times} &= \sin\beta_{L} \sin \left(2\lambda_{L} - \frac{\pi}{3} \right) \,,\\
-	F_{e}^{+} &= \frac{1}{2} \left( 1 + \sin^{2}\beta_{L} \right) \cos \left(2\lambda_{L} + \frac{\pi}{6} \right) \,,\\
-	F_{e}^{\times} &= \sin\beta_{L} \sin \left(2\lambda_{L} + \frac{\pi}{6} \right)  \,.
-\end{align}
-\esub
-Note that the expressions for the channel $e$ can be obtained from the expressions for $a$ with the replacement $\lambdaL \rightarrow \lambdaL + \pi/4$.
-
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-
 \section{Methodology}
@@
 Introducing the standard matched-filter inner product as
-\be
-	\left( \tilde{a} | \tilde{b} \right) = 4 \mathrm{Re} \int_{0}^{+\infty} df \; \frac{\tilde{a}(f) \tilde{b}^{*}(f)}{S_{n}(f)} \,,
+\be\label{eq:definnerproduct}
+	( \tilde{a} | \tilde{b} ) = 4 \mathrm{Re} \int_{0}^{+\infty} df \; \frac{\tilde{a}(f) \tilde{b}^{*}(f)}{S_{n}(f)} \,,
 \ee
@@
 
-\begin{figure*}
+\begin{figure}
   \centering
-  \includegraphics[width=.7\linewidth]{../plots/plot_tdia_hm.pdf}
-  \caption{\SM{TODO: make figure smaller, single-column.}}
+  \includegraphics[width=.99\linewidth]{../plots/plot_hc_tdia_hm.pdf}
+  \caption{.}
   \label{fig:FDHMSMBHCase9}
-\end{figure*}
-
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-
-\subsection{Time and frequency accumulation of SNR}
-\label{sec:timefreqSNR}
-
--- SNR accumulation with time or frequency: inspiral/MRD balance across mass range
-
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-
-\subsection{A non-degenerate case}
-\label{sec:SMBHPEnondeg}
+\end{figure}
+
+-- add plot for SNR accumulation with time or frequency: inspiral/MRD balance across mass range
+
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+
+\subsection{Parameter estimation}
+\label{sec:SMBHPE}
 
@@
 
+\begin{figure*}
+  \centering
+  \includegraphics[width=.98\linewidth]{../plots/corner_smbh_case9_ptmcmc_22_hm.png}
+  \caption{.}
+  \label{fig:PEsmbh22hmCase0}
+\end{figure*}
+
 -- High-mass system, non-degenerate, 22 and HM + Fisher - stacked SNR contours
@@
 
-\subsection{Accumulating signal with time}
+\subsection{Accumulation of signal with time}
 \label{sec:SMBHPEacctime}
@@
 
-\subsection{SNR dependence of parameter estimation}
-\label{sec:SMBHPEdistance}
-
--- Move to Appendix ?
-
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-
 \subsection{Investigating degeneracies}
 \label{sec:SMBHPEdegen}
-
-\begin{figure*}
-  \centering
-  \includegraphics[width=.98\linewidth]{../plots/corner_smbh_case9_ptmcmc_22_hm.png}
-  \caption{.}
-  \label{fig:PEsmbh22hmCase0}
-\end{figure*}
 
@@
   \includegraphics[width=.8\linewidth]{../plots/corner_smbh_case9_simplelike_ptmcmc_bambi.png}
-  \caption{.}
+  \caption{Posterior distributions obtained with the two samplers \texttt{PTMCMC} (blue) and $\texttt{MultiNest}$ (green) for the simplified likelihood~\eqref{eq:simplelikelihoodhm}, using only the dominant mode $(\ell, m) = (2,2)$.
+Masses and time are pinned to the injected value, and the motion of the detector is neglected.
+We also superimpose (red) the contours obtained with \texttt{MultiNest} by including all harmonics $(2,2)$, $(2,1)$, $(3,3)$, $(4,4)$, $(5,5)$.}
   \label{fig:PEsmbh22hmSimpleLikeCase9}
@@
   \end{minipage}
-  \caption{.}
+  \caption{Sky position posterior in presence of degeneracies, zooming close to the injected value.
+The left panel shows the complete likelihood, while the right panel shows the simplified likelihood~\eqref{eq:simplelikelihoodhm} with a motionless LISA and fixed masses and time.
+The results of \texttt{PTMCMC} (blue) and \texttt{MultiNest} are superimposed.
+The black cross is the injected value, and the grey cross is the analytic prediction~\eqref{eq:soldegensky} for the offset peak created by degeneracies.}
   \label{fig:Skydownsmbh22hmSimpleLikeCase9}
@@
 From the low-frequency response \eqref{eq:aelowfmodes}, pinning the time-at-LISA so that we ca ignore the factor $\exp [2 i \pi f k\cdot p_{0}]$ as a pure constant corresponding to a redefinition of time, we obtain
-\begin{align}
+\begin{align}\label{eq:simplelikelihoodhm}
 	\ln \calL &= -\frac{1}{2} \sum_{\ell m} \sum_{\ell' m'}   \left[ \left( s_{a}^{\ell m} - s_{a, \mathrm{inj}}^{\ell m}\right) \left( s_{a}^{\ell' m'} - s_{a, \mathrm{inj}}^{\ell' m'}\right)^{*} \right.
@@
 The prefactors $\left\langle \ell m | \ell' m' \right\rangle$ are pure normalization constants, and represent mode overlaps as
-\be
+\be\label{eq:innerproductlmlpmp}
 	\left\langle \ell m | \ell' m' \right\rangle = 4\mathrm{Re} \int \frac{df}{S_{n}^{a,e}} \left( 3 \pi f L\right)^{2} \tilde{h}_{\ell m} \tilde{h}_{\ell' m'}^{*} \,.
@@
 
-In the case where we only include the dominant harmonic $h_{22}$, the functions~\eqref{eq:defsaelm} take the explicit form
+In the case where we only include the dominant harmonic $h_{22}$, the likelihood~\eqref{eq:simplelikelihoodhm} simplifies to
+\be\label{eq:simplelikelihood22}
+	\ln \calL = -\frac{1}{2} \left\langle 22 | 22 \right\rangle  \left[ \left| s_{a}^{22} - s_{a, \mathrm{inj}}^{22} \right|^{2} + \left| s_{e}^{22} - s_{e, \mathrm{inj}}^{22}\right|^{2} \right] \,.
+\ee
+The functions~\eqref{eq:defsaelm} take the explicit form
 \begin{align}
@@
 we see that finding $(\lambda_{a}^{*}, \theta_{L}^{*})$ verifying
-\be\label{eq:conditionsky}
+\be\label{eq:degenconditionsky}
 	\frac{\sigma_{+}}{\sigma_{-}} (\lambda_{a}^{*}, \theta_{L}^{*}) = r_{\rm inj} \,,
@@
 with the approximation~\eqref{eq:sigmapmapprox} for the left-hand side, guarantees that we can then match both $\sigma_{+}$ and $\sigma_{-}$ of the injection with the condition
-\be\label{eq:conditionprefactor}
+\be\label{eq:degenconditionprefactor}
 	\rho e^{2i(\varphi - \psiL)} e^{2i \lambda_{a}^{*}} = \sigma_{-}^{\rm inj}\,.
@@
 \ee
-so that~\eqref{eq:conditionsky} is verified for special positions in the sky at (coming back to the original variables $(\lambdaL, \betaL)$)
+so that~\eqref{eq:degenconditionsky} is verified for special positions in the sky at (coming back to the original variables $(\lambdaL, \betaL)$)
 \bsub\label{eq:soldegensky}
@@
 \esub
-The other condition gives then a single phase constraint for $\varphi - \psiL$, and a single amplitude constraint for $\rho(d, \iota, \theta_{L}^{*})$.
+The other condition~\eqref{eq:conditionprefactor} gives then a single phase constraint for $\varphi - \psiL$, and a single amplitude constraint for $\rho(d, \iota, \theta_{L}^{*})$.
 The limit $\iota \rightarrow \pi$ (where the $(2,-2)$ contribution dominates the $(2,2)$) is very similar, as can be seen most easily by the change of variable $\iota \rightarrow \pi - \iota$, with $-\psiL$ and $\pi - \theta_{L}$ playing the roles of $\psiL$ and $\theta_{L}$.
@@
 \esub
-with a condition on prefactors that is now concerning $\rho e^{2 i \varphi}$.
+with a condition on $\rho$ and $\varphi$ that is now
+\be
+	\rho e^{2i\varphi} e^{2i (\lambda_{a}^{*} - \psi_{L}^{*})} = \sigma_{-}^{\rm inj} \,.
+\ee
 Thus, we find again a degenerate pattern, with the difference that \eqref{eq:soldegeniotaphi} imposes a condition on the pair $\lambda_{a} - \psiL$ instead of $\lambda_{a}$, and the extra condition imposes a condition on $\varphi$ instead of the pair $\varphi - \psiL$.
 
-We illustrate our findings by running a Bayesian parameter estimation for the simplified likelihood. The results are shown in Fig.~\ref{fig:PEsmbh22hmSimpleLikeCase9}, with the two samplers.
-
-Focusing on the sky position, we investigate the degenerate peak in the sky that is the closest to the injection in Fig.~\ref{fig:Skydownsmbh22hmSimpleLikeCase9}.
-We compare here the sky position posterior distribution generated by the two samplers for the true likelihood
+Finally, when considering more harmonics beyond the dominant mode $h_{22}$, the degeneracies will be broken easily.
+As in the case of the full likelihood, the simplified likelihood~\eqref{eq:simplelikelihoodhm} will have several terms with different inclination and phase dependencies.
+If the signal is loud enough for at least two modes to be detected, then the inclination $\iota$ and the phase $\varphi$ are fixed by the relative amplitude and phase of these modes.
+This will break the degeneracies $(d, \iota)$ and the degeneracy $(\varphi, \psiL)$.
+
+We illustrate our findings by running a Bayesian parameter estimation for the simplified likelihood~\eqref{eq:simplelikelihood22} with the two samplers \texttt{MultiNest} and \texttt{PTMCMC} described in Sec.~\ref{}.
+Since the inner products~\eqref{eq:innerproductlmlpmp} are constants computed only once, the external factors become trivially inexpensive, and the likelihood is very fast to compute.
+We take advantage of this to run \texttt{PTMCMC} for a very large number of steps, ensuring that we do not miss degenerate regions of the parameter space.
+The resulting posterior distribution is extremely degenerate, as shown in Fig.~\ref{fig:PEsmbh22hmSimpleLikeCase9}.
+The sampler \texttt{MultiNest} fails to resolve this structure, stopping after having only explored the degenerate region $\iota \rightarrow 0$ and $\iota \rightarrow \pi$ and the eight sky positions~\eqref{eq:soldegensky}.
+In accordance with~\eqref{eq:degenconditionsky}-\eqref{eq:degenconditionprefactor}, this corresponds to lines in $\varphi \pm \psiL = \mathrm{const}$.
+The sampler \texttt{PTMCMC} finds more degenerate regions, in particular it finds similar structures with the roles of $\iota$ and $\betaL$ exchanged (as shown notably by the $(d,\iota)$ and $(d,\betaL)$ panels), as well as intermediate regions connecting all these features.
+
+Fig.~\ref{fig:PEsmbh22hmSimpleLikeCase9} also shows the result obtained when including all available harmonics~\eqref{} in~\eqref{eq:simplelikelihoodhm}.
+The degenerate structures collapse to leave only eight modes corresponding to $\betaL \rightarrow - \betaL$ and $\lambdaL \rightarrow \lambdaL + k\pi/2$.
+We found that this degeneracy-breaking occurs already when including a single subdominant harmonic $h_{21}$.
+
+We can now come back to explaining the results of Sec.~\ref{}, and in particular Fig.~\ref{}, obtained with the true likelihood function for the mode $h_{22}$.
+First, the difference between the two Bayesian samplers has been made more extreme by going to the more degenerate likelihood~\eqref{eq:simplelikelihood22}, allowing us to attribute their disagreement to an apparent shortcoming of \texttt{MultiNest}\footnote{We recall that we only used standard values for the metaparameters \SM{[complete]} for this sampler, and we do not exclude that it would be possible to improve results with minor changes.} for resolving extended degenerate regions.
+Focusing on the sky position, we zoom close to the injection in Fig.~\ref{fig:Skydownsmbh22hmSimpleLikeCase9}.
+We compare here the posterior distribution generated with the true likelihood with the one obtained with the simplified likelihood~\eqref{eq:simplelikelihood22}, for both samplers.
+We also superimpose the prediction~\eqref{eq:soldegensky} for the degeneracy-induced peak.
+We see that, although ,
 
@@
 
+\section{The LISA instrument response}
+\label{sec:response}
+
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+%
+%\subsection{The LISA response - OLD}
+%\label{sec:lisaresponse}
+%
+%\SM{------------------------------------------------------}
+%
+%\SM{[copy-paste from previous paper -- just here to copy needed formulas / refs]}
+%
+%We begin by detailing the model that we use for the response of a LISA-like detector, together with the assumptions used and their limitations.
+Since our aim is to assess the accuracy of our direct Fourier-domain treatment of the response, we can focus on the gravitational-wave contribution to the basic single-link observables.
+We therefore use a somewhat simplified model for the time-domain response~\cite{Krolak+04}, ignoring corrections that would be crucial from the point of view of noise cancellations, keeping in mind that the response model can be enriched later without affecting the conclusions of the present analysis.
+%
+%The frequency-shift response for a single link~\eqref{eq:yslr} was derived in Ref.~\cite{EW75} (see also~\cite{CR02, RCP04, Finn08, Cornish09}).
+Several assumptions enter the result as written in~\eqref{eq:yslr}: (i) effects of the order $v/c$ are neglected, including for instance the special relativistic Doppler effect created by the relative speeds of the spacecraft on their orbits (ii) the propagation is assumed to take place in a flat spacetime, perturbed only by the gravitational wave; thus the gravitational redshift as well as the deflection of light created by the gravitational potential of the Sun is ignored (iii) all geometric factors are evaluated at a single time, whereas one should consider the beam as propagating from the position of the first spacecraft at the time of emission to the position of the second scapecraft at the time of reception, leading to a point-ahead effect.
+%
+%Additionally, we limit ourselves to a rigid model for the orbits of the constellation, namely we assume that the constellation remains in an equilateral configuration with fixed armlengths.
+These simplified orbits neglect (iv) effects of order $e^{2}$ from the eccentricity of the individual Keplerian orbits, (v) the effect of gravitational perturbations coming from other celestial bodies, such as the Earth, the quadrupole of the Sun, and the other planets.
+Note that although we can neglect all the effects (i)-(v) for our present study, keeping track of these corrections is crucial for the purpose of laser noise cancellations, and led to the development of new generations of TDI observables~\cite{Tintoliving}.
+%
+%It is natural to split the response~\eqref{eq:yslr} into two steps: first the orbital delay related to the orbit around the Sun of the whole constellation, and then the constellation response.
+The baselines for the delays are indeed very different in the two cases.
+Geometrical projection factors aside, we have for the orbit around the Sun $R=1\text{au}=1.5\times 10^{8} \text{km}$, while the detector armlength is $L=2.5\times 10^{6}\text{km}$ (in the configuration proposed in~\cite{LISA17}).
+It is natural to define two transfer frequencies for the two relevant length scales for the delays, defined such that a wavelength fits within this length scale, i.e.
+$2\pi f d = 1$.
+This gives
+%\begin{subequations}\label{eq:transferfrequencies}
+%\begin{align}
+%	f_{R} &= 3.2\times10^{-4}\Hz \,,\\
+%	f_{L} &= 1.9\times 10^{-2}\Hz \,.
+%\end{align}
+%\end{subequations}
+%The LISA response will behave qualitatively differently on the three frequency bands $f \leq f_{R}$, $f_{R} \leq f \leq f_{L}$ and $f_{L} \leq f$.
+%
+%The first stage of the response, the orbital delay, consists simply in applying the varying time delay to bring the wavefront sampling point from the SSB reference to the center of the LISA triangular constellation, common to all $y_{slr}$ observables.
+For $h^{\rm TT}$ the transverse-traceless gravitational waveform in matrix form, we write this orbital time delay as
+%\be\label{eq:defresponse0}
+%	h_{0}^{\rm TT} (t) = h^{\rm TT}(t-\hatk\cdot p_{0}) \,,
+%\ee
+%with $p_{0}$ the position of the constellation center, which follows the Earth orbit around the Sun.
+The second stage of the response calculation comprises the remaining, constellation-centered response.
+For the single-link contribution to the response, for the laser link from spacecraft $s$ to spacecraft $r$ along a path in direction $n_l$, we write
+%\begin{align}\label{eq:defresponseL}
+%	y_{slr} &= \frac{1}{2} \frac{1}{1 - \hatk\cdot n_{l}} \nn\\
+%	& \cdot n_{l}\cdot \left[ h_{0}^{\rm TT}(t - L - \hatk\cdot p^{L}_{s}) - h_{0}^{\rm TT}(t - \hatk\cdot p^{L}_{r}) \right] \cdot n_{l}\,,
+%\end{align}
+%where we reference the positions of the spacecraft relative to the center of the constellation, $p^{L}_{A} \equiv p_{A} - p_{0}$.
+%
+%As described in App.~\ref{app:notation} We will decompose the full signal in the contributions of the individual spin-weighted spherical modes $h_{\ell m}$, whose Fourier transforms are assumed to have a smooth amplitude and phase.
+First, we define the matrices $P_{+},P_{\times}$ such that, in the sense of matrices,
+%\be
+%	h^{\rm TT} = h_{+}P_{+} + h_{\times}P_{\times} \,.
+%\ee
+%We focus only on positive frequencies.
+Assuming that the approximation~\eqref{eq:zeronegativef} applies, we consider a single mode contribution, $h=h_{\ell m}$ with $m>0$.
+For each given mode we define a complex matrix $P_{\ell m}$ incorporating the spin-weighted spherical harmonic constant factor as
+%\be
+%	P_{\ell m} =
+%	\begin{cases}
+%	\frac{1}{2} {}_{-2}Y_{\ell m} \left( P_{+} + i P_{\times} \right) \text{ for } m>0\,,\\
+%	\frac{1}{2} {}_{-2}Y_{\ell m}^{*} \left( P_{+} - i P_{\times} \right) \text{ for } m<0\,.
+%	\end{cases}
+%\ee
+%
+%We now turn to the transformation of~\eqref{eq:defresponse0} and~\eqref{eq:defresponseL} to the Fourier domain. Applying a pure delay as in~\eqref{eq:defresponse0} translates into
+%\be\label{eq:G0}
+%	G_{0}(f, t) = e^{-2i\pi f d_{0}(t)} \,,
+%\ee
+%with $d_{0} = -\hatk \cdot p_{0}$ the delay associated to the orbit around the Sun.
+For the leading-order response~\eqref{eq:transferlocal}, this gives a Fourier-domain transfer function common to all modes, that is a pure phase factor, proportional to the frequency but also $t_{f}$-dependent:
+%\be\label{eq:transfer0local}
+%	\calT_{0}^{\rm local}(f) = G_{0}(f, \tf)\,.
+%\ee
+%If $(\lambda, \beta)$ are the ecliptic longitude and latitude of the source in the sky, and if the orbital phase is set by convention to $0$ at $t=0$, the orbital delay has the simple expression
+%\be\label{eq:delay0}
+%	d_{0}(t) = -R \cos\beta \cos\left(\Omega_{0}t - \lambda\right)\,.
+%\ee
+%
+%For the constellation part of response~\eqref{eq:defresponseL}, treated separately from the delay~\eqref{eq:delay0}, we write
+%\begin{align}\label{eq:decomposeGslr}
+%	F_{slr}^{L}(t) &= \frac{1}{2} \frac{1}{1 - \hatk\cdot n_{l}(t)} n_{l}(t) \cdot P_{\ell m} \cdot n_{l} (t) \,,\nn\\
+%	d_{s}(t) &= - k\cdot p_{s}^{L}(t) \,, \quad d_{r}(t) = - k\cdot p_{r}^{L}(t) \,,\nn\\
+%	G_{slr}^{L}(f,t) &=  F_{slr}^{L}(t) \left( e^{-2i\pi f (d_{s,L}(t) + L)} - e^{-2i\pi f d_{r}(t)} \right) \,.
+%\end{align}
+%The superscript $L$ indicates that the orbital delay~\eqref{eq:delay0} is not included.
+Since we also assume the rigid approximation for the constellation, where the armlengths are fixed, a particular simplification occurs when combining these individual delays, thanks to the relation $p^{L}_{r} - p^{L}_{s} =  L n_{l}$:
+%\begin{align}\label{eq:GslrL}
+%	G_{slr}^{L}(f,t) &= \frac{i \pi f L}{2} \sinc \left[ \pi f L\left(1-\hatk\cdot n_{l} \right) \right] \nn\\
+%	& \quad \cdot \exp\left[ i \pi f \left( L + \hatk\cdot \left( p_{1}^{L} + p_{2}^{L} \right) \right) \right]  n_{l} \cdot P_{\ell m} \cdot n_{l} \,,
+%\end{align}
+%with all time-dependent vectors evaluated at $t$.
+This expression is well known as describing the frequency-dependency in the LISA response~\cite{Larson+99, Cornish01, CR02, RCP04}.
+In the local approximation~\eqref{eq:transferlocal}, the Fourier-domain transfer function then reads
+%\begin{align}\label{eq:transferLlocal}
+%	\calT_{slr}^{L, \mathrm{local}}(f) &= G_{slr}^{L}(f, \tf) \,.
+%\end{align}
+%For plotting purposes, we will also define
+%\be\label{eq:transferLenvelope}
+%	\overline{\calT}_{slr}^{L} (f) = \frac{i \pi f L}{2} n_{l} \cdot P_{\ell m} \cdot n_{l} (\tf)
+%\ee
+%which will serve as an estimate for the enveloppe function of the response, devoid of the zero-crossings at high frequencies of the $\sinc$ term in~\eqref{eq:GslrL}.
+%
+%Note that, if the corrections of Sec.~\ref{subsec:delays}, for non-negligible $\dot d$, are included for the constellation delays, the transfer function will not have this simple form anymore, as $\dot{d}$ will have a different velocity-dependent expression for the sending and receiving spacecraft.
+One must then separately handle $d_{s}$ and $d_{r}$ in~\eqref{eq:decomposeGslr} to compute the corrections.
+%
+%The orbital response~\eqref{eq:transfer0local}-\eqref{eq:delay0} takes a simple analytic form, but the phase contribution of this delay is significant across most of the frequency band and can be large for $f \gg f_{R}$.
+%
+%The constellation response~\eqref{eq:GslrL}-\eqref{eq:transferLlocal} can be interpreted as the Fourier-domain translation of a discrete derivative taken on the waveform.
+The leading factor in~\eqref{eq:GslrL} shows that the amplitude of the response is proportional to $f$ in the low-frequency limit $f\ll f_{L}$, where the other factors are essentially unity.
+For $f\gtrsim f_{L}$, the $\sinc$ and the phase of the exponential generate additional structure in the response, including zero-crossings when the projected armlength is an integer number of wavelengths.
+From~\eqref{eq:transferLlocal}, an expansion for small $f\ll f_{L}$ yields back a Fourier-domain analog of the low-frequency approximation of the response~\cite{Cutler97, RCP04}, which is equivalent to having two LIGO-type interferometers turned by $\pi/4$ and set in motion.
+%
+%For analysis of the response we need a concrete set of gravitational waveforms.
+We will use the PhenomD model~\cite{Khan+15,Husa+15}, which provides Fourier-domain inspiral-merger-ringdown waveforms for aligned spins.
+We refer to App.~\ref{app:precLISA} for a brief discussion of the prospects for applying our formalism for the LISA response to precessing Fourier-domain waveforms.
+%
+%\SM{------------------------------------------------------}
+
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+
+\subsection{The GW signal}
+\label{sec:gwsignal}
+
+We start by introducing a source-frame $(\hat{x}_{S}, \hat{y}_{S}, \hat{z}_{S})$ attached to the binary system emitting gravitational waves.
+For comparable-mass systems without spin, the natural choice is to take the normal to the orbital plane as $\hat{z}_{S}$, and we will assume that the remaining rotation around $\hat{z}_{S}$ is fixed by the phase convention of the waveform model (see Sec.~\ref{}).
+
+Introducing $k$ the wave propagation vector going from the source towards the observer, we define the inclination $\iota$ and the observer phase $\varphi$ simply as being its spherical angular coordinates $(\theta_{S}, \phi_{S})$, so that in the source-frame
+\be
+	k_{S} = (\sin\iota \cos\varphi, \sin\iota \sin\varphi, \cos\iota) \,.
+\ee
+
+By convention, for our polarization vectors $p$ and $q$ we will use the spherical coordinate vectors $p = e_{\theta}^{S}$, $q = e_{\phi}^{S}$. In terms of only $\hat{z}_{S}$,
+\bsub
+\begin{align}
+	p &= q \times k \,,\\
+	q &= \hat{z}_{S} \times k / |\hat{z}_{S} \times k|\,.
+\end{align}
+\esub
+If $H_{ij} = h_{ij}^{\rm TT}$ represents the gravitational wave signal in the transverse-traceless gauge,
+\be
+	H = h_{+} P_{+} + h_{\times} P_{\times}
+\ee
+with the polarization tensors
+\bsub
+\begin{align}
+	P_{+} &= p \otimes p - q \otimes q \,,\\
+	P_{\times} &= p \otimes q + q \otimes p \,.
+\end{align}
+\esub
+Conversely, the polarizations are
+\bsub
+\begin{align}
+	h_{+} &= \frac{1}{2} \left(p \otimes p - q \otimes q \right) : H \,,\\
+	h_{\times} &= \frac{1}{2} \left( p \otimes q + q \otimes p \right) : H \,.
+\end{align}
+\esub
+
+We will further decompose the gravitational wave signal, seen as a function of the direction of emission $(\iota, \varphi)$, in spin-weighted spherical harmonics as
+\be\label{eq:hpcmodes}
+	h_{+} - i h_{\times} = \sum_{\ell \geq 2} \sum_{m = -\ell}^{\ell} {}_{-2}Y_{\ell m} (\iota, \varphi) h_{\ell m} \,,
+\ee
+where the explicit expression of the ${}_{-2}Y_{\ell m}$ is given in~\cite{}.
+
+The vectors $(p,q,k)$ form the Wave-frame, such that the gravitational wave takes in this frame the familiar form
+\be
+	H_{W} = \begin{pmatrix}
+		h_{+} & h_{\times} & 0 \\
+		h_{\times} & -h_{+} & 0 \\
+		0 & 0 & 0
+		\end{pmatrix} \,.
+\ee
+
+We will both generate the waveforms and apply the response directly in the Fourier domain. Our convention for the Fourier transform is
+\be\label{eq:defFourier}
+	\tilde{F}(f) = \int dt \; e^{2 i \pi f t} F(t) \,.
+\ee
+Note that this differs by a change $f\rightarrow -f$ from the most usual one (used e.g. in \texttt{LAL}).
+
+The definition of the reference time and phase \dots
+
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+
+\subsection{LISA trajectories}
+\label{sec:lisatraj}
+
+For completeness, we gather in this appendix explicit expressions for the trajectory of the LISA constellation and the rotation matrices entering the response.
+
+The orbits of the three spacecrafts around the Sun can be chosen so that, at leading order in the eccentricity of the orbits, the constellation retains the shape of an equilateral triangle in a cartwheeling motion following the Earth orbit.
+We denote by $a$ the semi-major axis, $e$ the eccentricity and pose $\alpha = \Omega_{0} (t-t_{0}))$, with $t_{0}$ a reference time for the initial position ($t_{0} = 0$ in our case).
+The reference SSB frame is $(\hat{x}, \hat{y}, \hat{z})$.
+
+Using the notation $c,s = \cos, \sin \alpha$, the trajectory of the center of the constellation is simply
+\be
+	p_{0} = ac \hat{x} + as \hat{y} \,,
+\ee
+and we take $a\equiv R = 1 \, \mathrm{au}$.
+We denote the positions of the spacecrafts as $p_{A}$ for $A=1,2,3$, and the positions relative to the constellation center as $p_{A}^{L} = p_{A} - p_{0}$.
+Setting $\beta_{A} = 2(A-1)\pi/3 + \beta_{0}$ (with $\beta_{0}$ and initial condition set to 0 in our case), the Cartesian coordinates in the SSB-frame of the position of the spacecrafts read~\cite{}
+\bsub
+\begin{align}
+	p_{A}^{L} &= a e \left[ \sin \beta_{A} c s - \cos\beta_{A} \left( 1 + s^{2} \right) \right] \hat{x} \nn\\
+	& + a e \left[ \cos \beta_{A} c s - \sin\beta_{A} \left( 1 + c^{2} \right) \right] \hat{y} \nn\\
+	& - a e \sqrt{3} \cos(\alpha - \beta_{A}) \hat{z} \,,
+\end{align}
+\esub
+The armlength, constant in this approximation, is related to the eccentricity and semi-major axis by
+\be
+	L = 2\sqrt{3} a e \,.
+\ee
+The rigid approximation for the constellation, at first order in $e$, can be seen as being at first order in $L/R \simeq 0.017$ for a $2.5 \, \mathrm{Gm}$ armlength~\cite{}.
+
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+
+\subsection{The LISA frame}
+\label{sec:LISAframe}
+
+It will be very useful to introduce a time-dependent frame $(\hat{x}_{L}, \hat{y}_{L}, \hat{z}_{L})(t)$ following the detector in its motion.
+Specifically, we choose this LISA frame such that at any time,
+\bsub
+\begin{align}
+	p_{1}^{L} &= - \frac{L}{\sqrt{3}} \hat{x}_{L} \,,\\
+	p_{2}^{L} &= \frac{L}{2\sqrt{3}} \hat{x}_{L} - \frac{L}{2} \hat{y}_{L} \,,\\
+	p_{3}^{L} &= \frac{L}{2\sqrt{3}} \hat{x}_{L} + \frac{L}{2} \hat{y}_{L} \,.\\
+\end{align}
+\esub
+
+This frame also provides us with an equivalent representation of the trajectories making use of rotation matrices.
+Since in our rigid approximation the constellation remains an equilateral triangle in its cartwheeling motion around the Sun, the configuration of the constellation at a later time is given by a rotation around the Sun composed with a rotation of the constellation around its symmetry axis.
+
+If $R(v,\varpi)$ denotes the matrix of an active rotation around the vector $v$ by an angle $\varpi$, and if we denote by $R_{L}$ the active rotation from the SSB-frame to the L-frame such that for each of the three basis vectors $(\hat{x}_{L}, \hat{y}_{L}, \hat{z}_{L}) = R_{L} (\hat{x}, \hat{y}, \hat{z})$, we have
+\be\label{eq:RL}
+	R_{L} = R(z, \alpha) \cdot R(y, -\pi/3) \cdot R(z, -\alpha) \,,
+\ee
+where we recall that $\alpha = \Omega_{0} (t-t_{0}) $.
+For all vectors $X$ among $p_{A}^{L}$, $n_{A}$ we have $X(t) = R_{L} \cdot X(t=t_{0})$.
+If a vector $X$ is given by its components in the SSB-frame, in the L-frame the components are $X_{L} = R_{L}^{-1} \cdot X$.
+
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+
+\subsection{From the Wave frame to the SSB frame}
+\label{sec:wavessbframe}
+
+We can now relate the Wave-frame to the SSB-frame, which will fix our convention for the sky position and polarization angle.
+The source position in the sky is given by $(\lambda, \beta)$, the ecliptic longitude and latitude in the SSB-frame.
+Like with the Source-frame, we can introduce spherical vectors $(e_{r}^{SSB}, e_{\theta}^{SSB}, e_{\phi}^{SSB})$.
+Since the propagation vector is $k = - e_{r}^{SSB}$,
+\be
+	k = (- \cos\beta\cos\lambda, - \cos\beta\sin\lambda, -\sin\beta) \,.
+\ee
+
+The last degree of freedom between the frames represents a rotation along the line-of-sight, parametrized by the polarization angle $\psi$.
+We introduce reference polarization vectors in the SSB-frame as
+\bsub
+\begin{align}
+	u &= \hat{z} \times k / |\hat{z} \times k| \,,\\
+	v &= k \times u \,.
+\end{align}
+\esub
+In terms of spherical vectors, $(k, u, v) = (-e_{r}^{SSB}, -e_{\phi}^{SSB}, -e_{\theta}^{SSB})$.
+The polarization angle $\psi$ is then defined such that $(p,q)$ are obtained by rotating $(u,v)$ by the angle $\psi$ around $k$, i.e.
+\be
+	(p,q) = R(k, \psi) \cdot (u,v) \,.
+\ee
+
+These relations can be summarized by the active rotation matrix $R_{W}$ from the SSB-frame to the Wave-frame:
+\be\label{eq:RW}
+	R_{W} = R(z, \lambda - \pi/2) \cdot R(x, \beta + \pi/2 ) \cdot R(z, \psi)\,.
+\ee
+
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+
+\subsection{Parametrizing the signal in the LISA-frame}
+\label{sec:LISAframeparams}
+
+While long-lasting signals like SOBHs will see a strong imprint of the LISA orbital motion over the course of observation, SMBH signals are dominated in SNR by a short-lived burst of emission at merger.
+For such signals, it will be useful to use a parametrization based on the L-frame at the time of merger instead of the SSB frame.
+
+The new parameters are defined as playing the same role as the SSB parameters, but relative to the L-frame.
+Namely, we define $R_{LW}$ the active rotation matrix from the L-frame to the Wave-frame, expressed in the L-frame basis, so that for each basis vector expressed in the L-frame basis $(\hat{x}_{W}, \hat{y}_{W}, \hat{z}_{W})_{L} = R_{LW} \cdot (\hat{x}_{L}, \hat{y}_{L}, \hat{z}_{L})_{L}$.
+The defining condition on $(\lambda_{L}, \beta_{L}, \psi_{L})$ is
+\be\label{eq:defRLW}
+	R_{LW} = R(z, \lambda_{L} - \pi/2) \cdot R(x, \beta_{L} + \pi/2 ) \cdot R(z, \psi_{L}) \,.
+\ee
+
+Coming back to vectors expressed in the SSB-basis, $R_{L}^{-1} \cdot (\hat{x}_{W}, \hat{y}_{W}, \hat{z}_{W}) = R_{LW} \cdot R_{L}^{-1} \cdot (\hat{x}_{L}, \hat{y}_{L}, \hat{z}_{L})$.
+Since $ (\hat{x}_{W}, \hat{y}_{W}, \hat{z}_{W}) = R_{W} \cdot (\hat{x}, \hat{y}, \hat{z})$ and $ (\hat{x}_{L}, \hat{y}_{L}, \hat{z}_{L}) = R_{L} \cdot (\hat{x}, \hat{y}, \hat{z})$, we arrive at
+\be\label{eq:RLW}
+	R_{LW} = R_{L}^{-1} \cdot R_{W} \,.
+\ee
+
+Combining~\eqref{eq:RLW}, \eqref{eq:RL}, \eqref{eq:RW} and~\eqref{eq:defRLW} yields the following expressions for the L-frame angular parameters\footnote{The convention here is that the point of coordinates $(x,y)$ in the plane has for argument $\arctan\left[ x, y \right]$.}:
+\bsub
+\begin{align}
+	\beta_{L} &= \arcsin \left[ \cos \frac{\pi}{3} \sin \beta - \sin \frac{\pi}{3} \cos \beta \cos \left( \lambda - \alpha \right) \right] \,,\\
+	\lambda_{L} &= \arctan \left[ \cos\beta \cos\lambda \left( \cos \frac{\pi}{3} \cos^{2}\alpha + \sin^{2}\alpha \right) \right. \nn\\
+	& \qquad\quad\;\; \left. + \cos\beta \sin\lambda \cos\alpha \sin\alpha \left( \cos\frac{\pi}{3} - 1 \right) \right. \nn\\
+	& \qquad\quad\;\; \left. + \sin\frac{\pi}{3} \sin\beta \cos\alpha , \right. \nn\\
+	& \qquad\qquad\; \left. \cos\beta \sin\lambda \left( \cos \frac{\pi}{3} \sin^{2}\alpha + \cos^{2}\alpha \right) \right. \nn\\
+	& \qquad\quad\;\; \left. + \cos\beta \cos\lambda \cos\alpha \sin\alpha \left( \cos\frac{\pi}{3} - 1 \right) \right. \nn\\
+	& \qquad\quad\;\; \left. + \sin\frac{\pi}{3} \sin\beta \sin\alpha \right] \,,\\
+	\psi_{L} &= \arctan \left[ \cos\psi \left( \cos\frac{\pi}{3}\cos\beta + \sin\frac{\pi}{3} \sin\beta \cos(\lambda - \alpha)\right) \right. \nn\\
+	& \qquad\quad\;\; \left. + \sin \psi \sin\frac{\pi}{3} \sin(\lambda - \alpha), \right. \nn\\
+	& \qquad\qquad\; \left. \sin\psi \left( \cos\frac{\pi}{3}\cos\beta + \sin\frac{\pi}{3} \sin\beta \cos(\lambda - \alpha)\right) \right. \nn\\
+	& \qquad\quad\;\; \left. - \cos \psi \sin\frac{\pi}{3} \sin(\lambda - \alpha) \right] \,.
+\end{align}
+\esub
+
+The definition of the reference phase $\phi_{\rm Ref}$, \dots \SM{[COMPLETE]}
+
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+
+\subsection{Mode decomposition and polarization angle}
+\label{sec:modespol}
+
+We now translate the mode decomposition~\eqref{eq:hpcmodes} in the Fourier domain, and make more explicit the dependence of the signal in the polarization $\psi$, so as to make manifest its purely extrinsic nature.
+Indeed, $\psi$ enters the observables only through constant prefactors, which is a bit obscured by the rotation matrix formulation~\eqref{eq:RW}.
+
+The mode decomposition~\eqref{eq:hpcmodes} gives (dropping $(\iota, \varphi)$)
+\bsub
+\begin{align}
+	h_{+} = \frac{1}{2} \sum_{\ell, m} \left( \sYlm h_{\ell m} + \sYlmstar h_{\ell m}^{*} \right) \,,\\
+	h_{\times} = \frac{i}{2} \sum_{\ell, m} \left( \sYlm h_{\ell m} - \sYlmstar h_{\ell m}^{*} \right) \,,
+\end{align}
+\esub
+which is valid in general. Now, for non-precessing binary systems, an exact symmetry relation between modes reads
+\be\label{eq:nonprecsymmetry}
+	h_{\ell, -m} = (-1)^{\ell} h_{\ell, m}^{*} \,.
+\ee
+Using this symmetry, we can write
+\be
+	h_{+,\times} = \sum_{\ell, m} K_{\ell m}^{+, \times} h_{\ell m} \,, 
+\ee
+with
+\bsub
+\begin{align}
+	K_{\ell m}^{+} =\frac{1}{2} \left( \sYlm + (-1)^{\ell} \sYlminusmstar \right) \,,\\
+	K_{\ell m}^{\times} = \frac{i}{2} \left( \sYlm - (-1)^{\ell} \sYlminusmstar \right) \,.
+\end{align}
+\esub
+
+Going to the Fourier domain, an approximation often used is to neglect support for negative/positive frequencies according to
+\be\label{eq:approxnegf}
+	\tilde{h}_{\ell m} (f) \simeq 0 \;\; \text{for} \;\; m<0, \; f>0 \;\; ( m>0, \; f<0 )\,,
+\ee
+and neglecting modes $h_{\ell 0}$.
+We will use this approximation throughout this paper.
+Note that we picked our Fourier convention~\eqref{eq:defFourier} to ensure that the mode $\tilde{h}_{22} (f)$ has support for $f>0$.
+Using~\eqref{eq:approxnegf},
+\be
+	\tilde{h}_{+,\times} = \sum_{\ell} \sum_{m>0} K_{\ell m}^{+, \times} \tilde{h}_{\ell m} \,.
+\ee
+
+Next, it is convenient to introduce mode-by-mode polarization matrices
+\be
+	P_{\ell m} = P_{+} K_{\ell m}^{+} + P_{\times} K_{\ell m}^{\times} \,,
+\ee
+so that
+\be\label{eq:Hsummodes}
+	H = \sum_{\ell m} P_{\ell m} h_{\ell m} \,.
+\ee
+To make explicit the dependence in polarization, we can define polarization tensors for 0 polarization angle as
+\bsub
+\begin{align}
+	P_{+}^{0} &= P_{+}(\psi = 0) = u \otimes u - v \otimes v \,,\\
+	P_{\times}^{0} &= P_{\times}(\psi = 0) = u \otimes v + v \otimes u \,.
+\end{align}
+\esub
+This allows to write the dependence in polarization as
+\be
+	P_{+} + i P_{\times} = e^{-2 i \psi} \left( P_{+}^{0} + i P_{\times}^{0} \right) \,,
+\ee
+or explicitly in $P_{\ell m}$ as
+\begin{align}
+	P_{\ell m} &= \frac{1}{2} \sYlm e^{-2 i \psi} \left( P_{+}^{0} + i P_{\times}^{0} \right) \nn\\
+	& + \frac{1}{2} (-1)^{\ell} \sYlminusmstar e^{+2 i \psi} \left( P_{+}^{0} - i P_{\times}^{0} \right) \,.
+\end{align}
+Writing the $P_{\ell m}$ matrices in this way allows to factor out all dependencies in the extrinsic parameters $(\iota, \phi, \psi)$.
+Together with the luminosity distance $D_{L}$, these parameters appear in combinations of constant factors multiplying the whole response.
+
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+
+\subsection{The LISA observables}
+\label{sec:LISAobservables}
+
+The single-link frequency shift observable $y_{slr} = (\nu_{r} - \nu_{s})/\nu$ between sending spacecraft $s$ and receiving spacecraft $r$ along the link $l$ takes the form 
+\be\label{eq:defyslr}
+	y_{slr} = \frac{1}{2} \frac{n_{l} \otimes n_{l}}{1 - k\cdot n_{l}} : \left[ H(t - L - k\cdot p_{s}) - H(t - k\cdot p_{r}) \right] \,.
+\ee
+Several assumptions enter this form of the response: \dots \SM{[COMPLETE]}.
+
+In the frequency domain, we will use the analysis of~\cite{} and write the response in individual observables $y_{slr}$ with a transfer function for each spherical harmonic mode as
+\be
+	\tilde{y}_{slr} = \sum_{\ell, m}\calT_{slr}^{\ell m}(f) \tilde{h}_{\ell m} \,.
+\ee
+With the decomposition~\eqref{eq:Hsummodes}, the corresponding form of the kernel $G^{\ell m}_{slr}(f, t)$ in the formalism of~\cite{} is
+\begin{align}\label{eq:Gslr}
+	G_{slr}^{\ell m}(f,t) &= \frac{i \pi f L}{2} \sinc \left[ \pi f L\left(1-k\cdot n_{l} \right) \right] \nn\\
+	& \quad \cdot \exp\left[ i \pi f \left( L + k\cdot \left( p_{1} + p_{2} \right) \right) \right]  n_{l} \cdot P_{\ell m} \cdot n_{l} \,.
+\end{align}
+At leading order in the separation of timescales, we have simply
+\be
+	\calT_{slr}^{\ell m}(f) = G_{slr}^{\ell m}(f, t_{f}^{\ell m}) \,,
+\ee
+with $t_{f}^{\ell m}$ the effective time-to-frequency correspondence, defined across the whole frequency band, including in the merger-ringdown, as $t_{f}^{\ell m} = -1/(2\pi) d\Psi_{\ell m} / df$ for a mode $\tilde{h}_{\ell m} = A_{\ell m}(f) \exp\left[-i \Psi_{\ell m}\right]$.
+The analysis of~\cite{} has shown that higher-order corrections are small for SMBH systems, and for SOBH systems that are close enough to merger as will be the case for the examples chosen here.
+The fact that we use the same Fourier-domain treatment of the transfer functions for both the signal and the templates should also mitigate errors.
+
+In our rigid approximation, delays are simply all constant and constant to $L$. Using the shortcut notation $y_{sr,L} = y_{slr}(t - L)$, the first-generation TDI Michelson observable $X$ reads
+\begin{align}
+	X &= y_{31} + y_{13,L} + \left( y_{21} + y_{12,L} \right)_{2L} \nn\\
+	& - \left( y_{21} + y_{12,L} \right) - \left( y_{31} + y_{13,L} \right)_{2L} \,,
+\end{align}
+with the other Michelson observables $Y$, $Z$ being obtained by cyclic permutation. The independent combinations $A$, $E$ and $T$ are then defined as
+\bsub
+\begin{align}
+	A &= \frac{1}{\sqrt{2}} \left( Z - X \right) \,,\\
+	E &= \frac{1}{\sqrt{6}} \left( X - 2Y + Z \right) \,,\\
+	T &= \frac{1}{\sqrt{3}} \left( X + Y + Z \right) \,.
+\end{align}
+\esub
+These channels are independent under the approximation of an uncorrrelated noise between the arms. With constant delays in the rigid approximation, the TDI combinations take the form
+\bsub\label{eq:aet}
+\begin{align}
+	\tilde{a} &= \left[ (1+z) \left( \tilde{y}_{31} + \tilde{y}_{13} \right) - \tilde{y}_{23} - z \tilde{y}_{32} - \tilde{y}_{21} - z \tilde{y}_{12} \right] \,,\\
+	\tilde{e} &= \frac{1}{\sqrt{3}} \left[ (1-z)\left( \tilde{y}_{13} - \tilde{y}_{31} \right) + (2+z) \left( \tilde{y}_{12} - \tilde{y}_{32} \right) \right. \nn\\
+	&\qquad\quad \left. + (1+2z) \left( \tilde{y}_{21} - \tilde{y}_{23} \right) \right] \,,\\
+	\tilde{t} &= \frac{\sqrt{2}}{\sqrt{3}} \left[ \tilde{y}_{21} - \tilde{y}_{12} + \tilde{y}_{32} - \tilde{y}_{23} + \tilde{y}_{13} - \tilde{y}_{31} \right] \,,
+\end{align}
+\esub
+where we have introduced the rescalings
+\bsub\label{eq:scalingAET}
+\begin{align}
+	\tilde{a}, \tilde{e} &= \frac{1}{i \sqrt{2} z \sin 2\pi f L}\times \tilde{A}, \tilde{E} \,,\\
+	\tilde{t} &= \frac{1}{2\sqrt{2} z^{3} \sin \pi f L \sin 2\pi f L} \times \tilde{T} \,.
+\end{align}
+\esub
+Scaling out the same square factors from the noise power spectral density as
+\bsub
+\begin{align}
+	S_{n}^{A}, S_{n}^{E} &= 2 \sin^{2} 2\pi f L \times S_{n}^{a}, S_{n}^{e} \,,\\
+	S_{n}^{T} &= 8 \sin^{2} \pi f L \sin^{2} 2\pi f L S_{n}^{t} \,,
+\end{align}
+\esub
+the reduced PSD for the three channels take the form
+\bsub\label{eq:Snaet}
+\begin{align}
+	S_{n}^{a} = S_{n}^{e} &= 2 \left( 3 + 2\cos 2\pi f L + \cos 4 \pi f L \right) S^{\rm pm}(f) \nn\\
+	& \quad + \left( 2 + \cos 2\pi f L \right) S^{\rm op}(f) \,,\\
+	S_{n}^{t} &= 4 \sin^{2} 2\pi f L S^{\rm pm}(f) + S^{\rm op}(f) \,,
+\end{align}
+\esub
+with $S^{\rm pm}$ the test-mass noise PSD and $S^{\rm op}$ the optical noise PSD.
+Here we ignore the confusion noise coming from the background of galactic binaries in the  LISA band.
+Note that the prefactors~\eqref{eq:scalingAET} are oscillatory and have zero-crossings at high frequencies, and scaling them out from both the signal and the noise is an approximation, as we can expect that the real delays would lead to imperfect cancellations in the vicinity of the zero-crossings.
+
+Finally, the likelihood function that we use, with a zero noise realization, is
+\begin{align}
+	\ln\calL &= -\frac{1}{2} \left[ \left( \tilde{a} - \tilde{a}_{\rm inj} | \tilde{a} - \tilde{a}_{\rm inj}  \right) + \left( \tilde{e} - \tilde{e}_{\rm inj} | \tilde{e} - \tilde{e}_{\rm inj}  \right) \right.
+\nn\\
+	& \qquad\qquad + \left. \left( \tilde{t} - \tilde{t}_{\rm inj} | \tilde{t} - \tilde{t}_{\rm inj} \right) \right] \,,
+\end{align}
+with the $\rm inj$ subscript indicating the injected signals, and where the inner products $(\cdot | \cdot)$ are given by~\eqref{eq:definnerproduct} with the noise PSDs~\eqref{eq:Snaet}.
+
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+
+\subsection{The low-frequency limit}
+\label{sec:low-freq}
+
+As is well known~\cite{Cutler97}, in the low-frequency limit (also called the long-wavelength approximation), the finite-length effects of the arms of LISA disappear, and the response of the detectors falls back to the one of two LIGO-type detectors set in motion, rotated from each other by $\pi/4$.
+
+For $f \ll f_{L}$, we have $2\pi f L \ll 1$ and the kernel~\eqref{eq:defGslr} reduces to
+\be\label{eq:Gslrlowf}
+	G^{\ell m}_{slr} \simeq \frac{i \pi f L}{2} \exp\left[ 2 i \pi f k\cdot p_{0} \right] n_{l} \otimes n_{l} : P_{\ell m}\,.
+\ee
+For even lower frequencies, when $f \ll f_{R}$ as well, $2\pi f R \ll 1$ and the delay term is negligible: $\exp\left[ 2 i \pi f k\cdot p_{0} \right] \simeq 1$.
+
+In the following, we will also apply frequency-domain response at leading order in the separation of timescales~\cite{MB18}, so that the transfer functions are $\calT(f) = G(f, \tf)$ with $\tf$ the time-of-frequency function~\eqref{eq:deftf}.
+For $2\pi f L \ll 1$, we have $z\simeq 1$, and the link reversal symmetry $\tilde{y}_{slr} \simeq \tilde{y}_{r-ls}$, so that~\eqref{eq:aet} becomes
+\bsub\label{eq:aetlowf}
+\begin{align}
+	\tilde{a} &\simeq 4\tilde{y}_{31} - 2 \tilde{y}_{23} - 2 \tilde{y}_{12} \,,\\
+	\tilde{e} &\simeq 2\sqrt{3} \left[ \tilde{y}_{12}  - \tilde{y}_{23} \right] \,,\\
+	\tilde{t} &\simeq 0 \,,
+\end{align}
+\esub
+with the $T$-channel becoming negligible in this limit. Using~\eqref{eq:Gslrlowf}, we can write
+\be\label{eq:aelowfmodes}
+	\tilde{a}, \tilde{e} = \sum_{\ell, m>0} (-2i\pi f) \exp\left[ 2 i \pi f k\cdot p_{0} \right] \tilde{h}_{\ell m} D_{a,e} : P_{\ell m} \,,
+\ee
+where we introduced the detector tensors
+\bsub\label{eq:DaDe}
+\begin{align}
+	D_{a} &= \frac{L}{2} \left( n_{1}\otimes n_{1} + n_{3} \otimes n_{3} - 2 n_{2} \otimes n_{2} \right) \,,\\
+	D_{e} &= \frac{L\sqrt{3}}{2} \left( n_{1} \otimes n_{1} - n_{3} \otimes n_{3} \right) \,.
+\end{align}
+\esub
+Here, we have made apparent factors $(-2i\pi f)$ corresponding to the Fourier-domain translation of a time derivative.
+Indeed, the observables $\Delta \nu / \nu$ used here are time derivatives of the laser phase more commonly used for ground-based detectors.
+
+We can now map these two channels to two fictitious LIGO-type detectors as follows.
+For an orthogonal detector of the same setup as the ground-based LIGO and Virgo with armlength $L_{D}$, rotated by an angle $\epsilon_{D}$ from the basis vectors $x,y$, the detector tensor is
+\begin{align}
+	D &= \frac{L_{D}}{2} \left[ \cos 2\epsilon_{D} \left( x\otimes x - y\otimes y \right) \right. \nn\\
+	& \qquad\quad + \left. \sin 2\epsilon_{D} \left( x\otimes y + y \otimes x\right) \right] \,.
+\end{align}
+By comparison with~\eqref{eq:DaDe}, we obtain
+\be
+	L_{a} = L_{e} = \frac{3L}{2} \,, \; \epsilon_{a} = \frac{2\pi}{3} \,, \; \epsilon_{e} = \frac{5\pi}{12} \,.
+\ee
+Note a degeneracy $\pm \pi$ in how we choose to orient these effective detectors. We can recover pattern functions (for $\psi_{L}=0$) by factoring out the effective length, defining
+\be\label{eq:defFapcFepc}
+	F_{a,e}^{+,\times} \equiv \frac{2}{3L} D_{a,e} : P_{+,\times}^{0} \,,
+\ee
+which gives in terms of the LISA-frame sky position angles
+\bsub\label{eq:FapcFepc}
+\begin{align}
+	F_{a}^{+} &= \frac{1}{2} \left( 1 + \sin^{2}\beta_{L} \right) \cos \left(2\lambda_{L} - \frac{\pi}{3} \right) \,,\\
+	F_{a}^{\times} &= \sin\beta_{L} \sin \left(2\lambda_{L} - \frac{\pi}{3} \right) \,,\\
+	F_{e}^{+} &= \frac{1}{2} \left( 1 + \sin^{2}\beta_{L} \right) \cos \left(2\lambda_{L} + \frac{\pi}{6} \right) \,,\\
+	F_{e}^{\times} &= \sin\beta_{L} \sin \left(2\lambda_{L} + \frac{\pi}{6} \right)  \,.
+\end{align}
+\esub
+Note that the expressions for the channel $e$ can be obtained from the expressions for $a$ with the replacement $\lambdaL \rightarrow \lambdaL + \pi/4$.
+
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+
 \section{Numerical integration with spline-interpolated phases}
@@
 
+\section{SNR dependence of parameter estimation}
+\label{sec:SMBHPEdistance}
+
+-- show series of analyses varying only the distance/SNR
+
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+
 \bibliography{references.bib}

### 2019-04-04 paper/paper_PE.tex
Fisher and start of sampling.

--- 2019-04-02 paper/paper_PE.tex
@@
 
+\subsection{Bayesian sampling}
+\label{sec:samplers}
+
+For this exploratory work we take a ``naive'' approach to  estimating posterior-probability distributed parameter samples, not applying any specific knowledge or expectations about distributions which might plausibly lead to artifically robust-seeming but incomplete results.
+We apply two independent approaches, an parallel tempering MCMC sampler with generically wtih amotivated proposal distribution, and nested sampling.
+\jgb{More to come}
+
+-- Bayesian samplers: Multinest, PTMC
+
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+
 \subsection{Fisher matrix parameter estimation}
@@
 
--- Fisher matrices computation: step self-tuning,...
-
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-
-\subsection{Bayesian sampling}
-\label{sec:samplers}
-
--- Bayesian samplers: Multinest, PTMCMC
+For comparison with our Bayesian inference results we also compute estimates of parameter estimation precision using the Fisher information matrix the common workhorse of LISA science studies despite its limitations.~\cite{Vallisneri08}  For measurements with additive Gaussian noise, as assumed here, the Fisher information matrix can be computed by
+\be
+F_{ij}=\left( \partial_{\theta_i} h| \partial_{\theta_i} h \right).
+\ee
+The inverse of this matrix estimates the covariance matrix describing parameter uncertainty. 
+The derivatives are computed by second-order finite differences, with the signal inner-products expanded and computed before differencing.
+For these calculations, where numerical smoothness is more of a concern than speed we compute the inner products explicitly on a fine grid.
+
+Generally, finite difference Fisher matrices can be problematic, with small numerical defects potentially having outsized impact, so the finite-difference step size $\epsilon_i$ used in each parameter derivative $\partial_{\theta_i}h$ must be chosen carefully.
+We target the step-size to be scaled off the diagonal Fisher matrix elements by ${\epsilon_i}\approx\delta/\sqrt{F_{ii}}$ using $\delta=0.001$ \jgb{This is \texttt{fisher\_err\_target}, was it different for some runs?}  for the results shown here.
+To achieve this we set begin with an initial choice of the $\epsilon_i$ scaled of parameter values or prior widths, then we estimate the Fisher diagonals and iterate until convergence.
+In this process we also constrain that the step is never more than $10^{-10}$ times the initial scaling.
+\jgb{There are some more special-case details, but probably not worth elaborating.
+We could also discuss how we do matrix inversion, but I don't think there is anything interesting there.}
 

### 2019-04-04 paper/paper_PE.tex

--- 2019-04-04 paper/paper_PE.tex
@@
 
+\subsection{Bayesian sampling}
+\label{sec:samplers}
+
+For this exploratory work we take a ``naive'' approach to  estimating posterior-probability distributed parameter samples, not applying any specific knowledge or expectations about distributions which might plausibly lead to artifically robust-seeming but incomplete results.
+We apply two independent approaches, an parallel tempering MCMC sampler with generically wtih amotivated proposal distribution, and nested sampling.
+\jgb{More to come}
+
+-- Bayesian samplers: Multinest, PTMC
+
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+
+\subsection{Fisher matrix parameter estimation}
+\label{sec:Fisher}
+
+For comparison with our Bayesian inference results we also compute estimates of parameter estimation precision using the Fisher information matrix the common workhorse of LISA science studies despite its limitations.~\cite{Vallisneri08}  For measurements with additive Gaussian noise, as assumed here, the Fisher information matrix can be computed by
+\be
+F_{ij}=\left( \partial_{\theta_i} h| \partial_{\theta_i} h \right).
+\ee
+The inverse of this matrix estimates the covariance matrix describing parameter uncertainty. 
+The derivatives are computed by second-order finite differences, with the signal inner-products expanded and computed before differencing.
+For these calculations, where numerical smoothness is more of a concern than speed we compute the inner products explicitly on a fine grid.
+
+Generally, finite difference Fisher matrices can be problematic, with small numerical defects potentially having outsized impact, so the finite-difference step size $\epsilon_i$ used in each parameter derivative $\partial_{\theta_i}h$ must be chosen carefully.
+We target the step-size to be scaled off the diagonal Fisher matrix elements by ${\epsilon_i}\approx\delta/\sqrt{F_{ii}}$ using $\delta=0.001$ \jgb{This is \texttt{fisher\_err\_target}, was it different for some runs?}  for the results shown here.
+To achieve this we set begin with an initial choice of the $\epsilon_i$ scaled of parameter values or prior widths, then we estimate the Fisher diagonals and iterate until convergence.
+In this process we also constrain that the step is never more than $10^{-10}$ times the initial scaling.
+\jgb{There are some more special-case details, but probably not worth elaborating.
+We could also discuss how we do matrix inversion, but I don't think there is anything interesting there.}
+
 \section{Stellar origin black holes}

### 2019-04-04 paper/paper_PE.tex
Updates to SOBH section

@@
 %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-
-\subsection{Bayesian sampling}
-\label{sec:samplers}
-
-For this exploratory work we take a ``naive'' approach to  estimating posterior-probability distributed parameter samples, not applying any specific knowledge or expectations about distributions which might plausibly lead to artifically robust-seeming but incomplete results.
-We apply two independent approaches, an parallel tempering MCMC sampler with generically wtih amotivated proposal distribution, and nested sampling.
-\jgb{More to come}
-
--- Bayesian samplers: Multinest, PTMC
-
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-
-\subsection{Fisher matrix parameter estimation}
-\label{sec:Fisher}
-
-For comparison with our Bayesian inference results we also compute estimates of parameter estimation precision using the Fisher information matrix the common workhorse of LISA science studies despite its limitations.~\cite{Vallisneri08}  For measurements with additive Gaussian noise, as assumed here, the Fisher information matrix can be computed by
-\be
-F_{ij}=\left( \partial_{\theta_i} h| \partial_{\theta_i} h \right).
-\ee
-The inverse of this matrix estimates the covariance matrix describing parameter uncertainty. 
-The derivatives are computed by second-order finite differences, with the signal inner-products expanded and computed before differencing.
-For these calculations, where numerical smoothness is more of a concern than speed we compute the inner products explicitly on a fine grid.
-
-Generally, finite difference Fisher matrices can be problematic, with small numerical defects potentially having outsized impact, so the finite-difference step size $\epsilon_i$ used in each parameter derivative $\partial_{\theta_i}h$ must be chosen carefully.
-We target the step-size to be scaled off the diagonal Fisher matrix elements by ${\epsilon_i}\approx\delta/\sqrt{F_{ii}}$ using $\delta=0.001$ \jgb{This is \texttt{fisher\_err\_target}, was it different for some runs?}  for the results shown here.
-To achieve this we set begin with an initial choice of the $\epsilon_i$ scaled of parameter values or prior widths, then we estimate the Fisher diagonals and iterate until convergence.
-In this process we also constrain that the step is never more than $10^{-10}$ times the initial scaling.
-\jgb{There are some more special-case details, but probably not worth elaborating.
-We could also discuss how we do matrix inversion, but I don't think there is anything interesting there.}
-
 \section{Stellar origin black holes}
@@
 important source of LISA detections \cite{Sesana16} although full parameter
-estimation studies are not available yet. In this section we present
-three case studies of parameter inference for such sources. The parameters of
-the simulated mergers are taken from the Radler LISA data challenge and are
-listed in Table \ref{table:sobh_params}.
+estimation studies are not available yet. In this section we present two case
+studies of parameter inference for such sources using the formalism and methods
+described earlier. The parameters of the simulated mergers are taken from the
+Radler LISA data challenge and are listed in Table \ref{table:sobh_params}.
 \begin{table}
-	\begin{tabular}{|c||c|c|c|}
+	\begin{tabular}{|c||c|c|}
 		\hline
-		Radler entry 		& 1 	& 12 	& 52 \\
+		Radler entry 		& 1 	& 12   \\
 		\hline
-		Mass 1 ($\Msol$) 	& 21.4	& 63.5	& 57.0 \\
+		Mass 1 ($\Msol$) 	& 21.4	& 63.5 \\
 		\hline
-		Mass 2 ($\Msol$) 	& 20.1	& 45.1	& 36.8 \\
+		Mass 2 ($\Msol$) 	& 20.1	& 45.1 \\
 		\hline
-		Distance (Mpc) 		& 49.1	& 318	& 308 \\
+		Distance (Mpc) 		& 49.1	& 318  \\
 		\hline
-		LISA SNR 			& 26.6	& 11.9	& 16.0 \\
+		LISA SNR 			& 26.6	& 11.9 \\
 		\hline
@@
 
-The waveform model used to work with such systems is the phenomenological
-approximant IMRPhenomD, extended at low frequency with a post-Newtonian
+We caution the reader that our purpose here is to show the feasibility of
+Bayesian inference for SOBH sources in LISA using the methodology presented in
+this paper. It is by no means a final answer on how to do so in an optimal or
+convenient way.  For this reason, the problem is posed in a simplified
+way here: in particular, the SOBH signal is assumed to merge roughly at the end of
+the observation epoch, and the finite observation time is enforced by
+converting it to a minimum analyzed Fourier frequency via the time-frequency
+relation of a simple Newtonian inspiral. In a more realistic situation, the
+merger time would likely be free to range from the beginning of the observation
+to years after its end, and the frequency of the signal at the beginning of the
+observation would be another sampling parameter.
+
+The waveform model used to work with SOBH systems is the same phenomenological
+approximant PhenomD used earlier, extended at low frequency with a post-Newtonian
 frequency-domain waveform. In practice, because LISA only observes the early
@@
 \end{figure*}
+
+Such posterior distributions should not be taken literally as expectations of
+LISA's capability for SOBH systems. One reason is the simplified treatment of
+the finite observation time described earlier. A second reason is the assumption
+of zero black-hole spins, which artificially reduces the width of the posterior
+distribution due to the strong degeneracy between spins and symmetric mass
+ratio \cite{Baird+2013}.
 

### 2019-04-08 paper/paper_PE.tex
Work on the paper, pushed d-series plots

--- 2019-04-04 paper/paper_PE.tex
@@
 
-Introducing the standard matched-filter inner product as
+Introducing the standard matched-filter inner product (also called overlap) as
 \be\label{eq:definnerproduct}
@@
 \ee
-with the data stream being a superposition of the gravitational wave signal for the true parameters $\bm{\theta}_{0}$ and the noise realization in the experiment, $d = s(\bm{\theta}_{0}) + n$.
+with the data stream being a superposition of the gravitational wave signal for the true parameters $\bm{\theta}_{0}$ and the noise realization in the experiment, $d = h(\bm{\theta}_{0}) + n$.
+In general, the templates $h(\theta)$ used for the analysis will only be an approximation of the physical signals entering the data stream $d$; we will ignore entirely this distinction and neglect the systematic errors in the waveform model.
 The posterior distribution for the physical parameters $\bm{\theta}$ given the observed data $d$ is then
@@
 \subsection{Frequency domain LISA response}
-\label{sec:response}
+\label{subsec:response}
 
@@
 \subsection{Reduced order model for EOBNRv2HM waveforms.}
-\label{sec:waveforms}
+\label{subsec:ROM}
 
@@
 
--- Likelihood computation: fast overlaps
+Setting the noise realization to zero in~\eqref{eq:deflnL}, having at our disposal a compact amplitude/phase respresentation of the signals allows us to compute likelihoods in an accelerated fashion.
+In general, both the template and the injection in~\eqref{eq:deflnL} are given as frequency series sampled with the Nyquist criterion $\Delta f = 1/(2T)$ with $T$ the duration of the signal, and the overlaps~\eqref{eq:definnerproduct} are simply computed as discrete sums over those frequency samples.
+Although transforming~\eqref{eq:definnerproduct} in a discrete is a simple operation, the size of the frequency series alone can represent a significant cost.
+In the context of binary neutron star observations.
+In the LISA.
+
+Applying the response as in~\ref{subsec:response} mode-by-mode on the sparsely sampled amplitude and phase generated as in~\ref{subsec:ROM}, we obtain a complete representation of both the waveform and of the instrument transfer functions~\eqref{} on a few hundred points in the Fourier domain, with the full signal being implicitly reconstructed with a standard cubic spline interpolation over frequencies.
+Decomposing~\eqref{eq:deflnL} as
+\be\label{eq:lnLoverlapstructure}
+	\ln \calL = \left( h(\bm{\theta}) | h(\bm{\theta}_{0}) \right)- \frac{1}{2} \left( h(\bm{\theta}) | h(\bm{\theta}) \right) - \frac{1}{2} \left( h(\bm{\theta}_{0}) | h(\bm{\theta}_{0}) \right) \,,
+\ee
+we have to compute inner products of the form~\eqref{eq:definnerproduct}.
+Note that individual terms in~\eqref{eq:lnLoverlapstructure} can be individually large, and computing $\ln \calL$ relies on accurate cancellations.
+Decomposing further in TDI channels as in~\eqref{} and in harmonics as in~\eqref{}, we have a sum of terms with the structure
+\be
+	\int \frac{df}{S_{n}} \; \calT_{1}^{\ell m} \calT_{2}^{\ell' m'} A_{1,\ell, m} A_{2, \ell' m'} e^{i(\Phi_{1,\ell m} - \Phi_{2,\ell' m'})} \,.
+\ee
+The difficulty of computing such an overlap depends on the phase difference.
+A large phase difference causes the integrand to be very oscillatory, requiring more frequency resolution than non-oscillatory integrands; on the other hand, oscillatory integrands give raise to a small integral due to cancellation effects, while non-oscillatory integrands contribute the most.
+For a single mode, SNR terms of the type $(h|h)$ have a zero phase difference by construction, but including different modes generates cross-terms $(\ell, m) \neq (\ell', m')$ with a large phase difference.
+In a Bayesian analysis, a large fraction of the time (after burn-in) will be spent exploring signals that are rather close to the injection.
+
+Thus, one can envision several approaches to compute integrals of the type~\eqref{eq:lnLoverlapstructure}, such as factoring out a guess of the injected signal~\cite{} and ignoring cross-terms between modes.
+In this work, we use a generic numerical treatment for the integrals~\eqref{eq:lnLoverlapstructure} and keep all terms with no further approximation.
+
+First, we built a joint sparse frequency grid suitable to represent both signals and the transfer functions.
+We then resample the integrand on this grid, separating prefactors (noise PSD, transfer function without the Doppler phase~\eqref{eq:defPhiR}, signal amplitude) from phases (Doppler phases, signal phase difference).
+We build a cubic spline for the prefactor, and a quadratic spline for the phase.
+We obtain the structure
+\be\label{eq:fastlikestructure}
+	\sum_{j \in \mathrm{grid}} \int_{f_{j}}^{f_{j+1}} df \; P_{j} (f) e^{i \phi_{j} (f)} \,,
+\ee
+where in each grid interval $P_{j}$ is a cubic polynomial and $\phi_{j}$ a quadratic polynomial.
+We can then compute the elementary integrals~\eqref{eq:fastlikestructure} using a combination of integrations by parts and asymptotic expansions, as explained in~\ref{app:numintegration}.
+The number of grid intervals is typically several hundreds \SM{[COMPLETE]}.
+
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+
+\subsection{Bayesian sampling}
+\label{sec:samplers}
+
+For this exploratory work we take a ``naive'' approach to  estimating posterior-probability distributed parameter samples, not applying any specific knowledge or expectations about distributions which might plausibly lead to artifically robust-seeming but incomplete results.
+We apply two independent approaches, an parallel tempering MCMC sampler with generically wtih amotivated proposal distribution, and nested sampling.
+\jgb{More to come}
+
+-- Bayesian samplers: Multinest, PTMC
 
@@
   \centering
-  \includegraphics[width=.98\linewidth]{../plots/plot_smbh9_alltdi.pdf}
+  \includegraphics[width=.98\linewidth]{../plots/plot_smbh9_tdi.pdf}
   \caption{\SM{TODO: rethink the visualization, maybe plot only transfer functions}.}
@@
 \ee
-From the low-frequency response \eqref{eq:aelowfmodes}, pinning the time-at-LISA so that we ca ignore the factor $\exp [2 i \pi f k\cdot p_{0}]$ as a pure constant corresponding to a redefinition of time, we obtain
+From the low-frequency response \eqref{eq:aelowfmodes}, pinning the time-at-LISA~\eqref{} \SM{[COMPLETE $t_{L}$]} so that we can ignore the factor $\exp [2 i \pi f k\cdot p_{0}]$ as a pure constant corresponding to a redefinition of time, we obtain
 \begin{align}\label{eq:simplelikelihoodhm}
@@
   \centering
-  \includegraphics[width=.98\linewidth]{../plots/plot_sobh1_alltdi.pdf}
+  \includegraphics[width=.98\linewidth]{../plots/plot_sobh1_tdi.pdf}
   \caption{\SM{TODO: rethink the visualization, maybe plot only transfer functions}.}
@@
 For even lower frequencies, when $f \ll f_{R}$ as well, $2\pi f R \ll 1$ and the delay term is negligible: $\exp\left[ 2 i \pi f k\cdot p_{0} \right] \simeq 1$.
+It can be convenient to introduce the following notations for this delay phase:
+\be\label{eq:defPhiR}
+	\Phi_{R} \equiv 2 \pi f k\cdot p_{0} \,, \quad \Delta \Phi_{R} = \Phi_{R} - \Phi_{R}(t=t_{f}^{\rm peak}) \,.
+\ee
+Here, $\Phi_{R}$ with $t=t_{f}$ is what is called often the Doppler phase in the literature, and $\Delta \Phi_{R}$ subtracts the contribution that corresponds to shifting the origin of times to the time of the peak seen at the LISA location \SM{[COMPLETE $t_{L}$]}, which is equivalent to using the time variable $t_{L}$ as defined in~\eqref{}.
 
@@
 \section{Numerical integration with spline-interpolated phases}
-\label{sec:numintegration}
+\label{app:numintegration}
 
@@
 
+\begin{figure}
+  \centering
+  \includegraphics[width=.99\linewidth]{../plots/plot_MM_ROM.pdf}
+  \caption{Unfaithfulness of ROM waveforms compared to the original EOBNRv2HM waveforms, computed as in~\eqref{}.
+We show here the results for the orientations \SM{[complete]}, for 200 random mass ratio values between 1 and 12, for masses $10 \Msol$, $50 \Msol$ and $250 \Msol$.
+The inclination angle is $\iota = \pi/2$, maximizing the effect of higher harmonics.}
+  \label{fig:MMROM}
+\end{figure}
+
 %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
@@
 
+\begin{figure*}
+  \centering
+  \begin{minipage}{.32\linewidth}
+      \includegraphics[width=.99\linewidth]{../plots/corner_smbh_case9_hm_dseries_m1m2.png}
+   \end{minipage}
+   \begin{minipage}{.32\linewidth}
+      \includegraphics[width=.99\linewidth]{../plots/corner_smbh_case9_hm_dseries_Dinc.png}
+   \end{minipage}
+   \begin{minipage}{.32\linewidth}
+      \includegraphics[width=.99\linewidth]{../plots/corner_smbh_case9_hm_dseries_tphi.png}
+   \end{minipage}
+  \caption{.}
+  \label{fig:smbhCornerZoomHMCase9}
+\end{figure*}
+
+\begin{figure*}
+  \centering
+  \includegraphics[width=.8\linewidth]{../plots/plot_mollweide_dseries.png}
+  \caption{.}
+  \label{fig:MollweidesmbhTserieshmCase9}
+\end{figure*}
+
 %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

### 2019-04-09 paper/paper_PE.tex
Work on the paper

@@
 
--- Geometric setting, definitions, summarize the FD response
+We will give in this section a brief summary of the instrument response of LISA, and of our Fourier-domain treatment of the modulations and delays.
+We refer the reader to App.~\ref{app:response} for comprehensive notations and definitions.
+
+The higher harmonics $h_{\ell, m}$ parametrize the decomposition of the gravitational wave signal $h_{+}$, $h_{\times}$ in spin-weighted spherical harmonics as
+\be\label{eq:hpcmodes}
+	h_{+} - i h_{\times} = \sum_{\ell \geq 2} \sum_{m = -\ell}^{\ell} {}_{-2}Y_{\ell m} (\iota, \varphi) h_{\ell m} \,,
+\ee
+where $(\iota, \varphi)$ is the direction of emission in a frame attached to the source by convention, and the ${}_{-2}Y_{\ell m}$ are spin-weighted spherical harmonics~\cite{}.
+
+The LISA response can be built from single-link observables $y_{slr} = (\nu_{r} - \nu_{s})/\nu$ respresenting a frequency shift between the spacecrafts $s$ and $r$ along the link $l$.
+We use the expression~\cite{}
+\be\label{eq:defyslr}
+	y_{slr} = \frac{1}{2} \frac{n_{l} \otimes n_{l}}{1 - k\cdot n_{l}} : \left[ H(t - L - k\cdot p_{s}) - H(t - k\cdot p_{r}) \right] \,,
+\ee
+with $H$ the transverse-traceless matrix representing the gravitational wave, $L$ the delay along one arm\footnote{We use units with $G=c=1$.}, taken to be fixed, $k$ the propagation vector, $n_{l}$ the link unit vectors and $p_{r}$, $p_{s}$ the positions of the spacecrafts.
+
+In the frequency domain, modes will be respresented with their amplitude and phase as\footnote{Note our Fourier convention~\eqref{eq:defFourier}.}
+\be\label{eq:hlmampphase}
+	\tilde{h}_{\ell m} (f) = A_{\ell m} (f) e^{-i\Psi_{\ell m} (f)}\,.
+\ee
+We will use the analysis of~\cite{MB18} and write the response in individual observables $y_{slr}$ with a transfer function for each spherical harmonic mode as
+\be
+	\tilde{y}_{slr} = \sum_{\ell, m}\calT_{slr}^{\ell m}(f) \tilde{h}_{\ell m} \,.
+\ee
+With the decomposition~\eqref{eq:Hsummodes}, the corresponding form of the kernel $G^{\ell m}_{slr}(f, t)$ in the formalism of~\cite{MB18} is
+\begin{align}\label{eq:Gslr}
+	G_{slr}^{\ell m}(f,t) &= \frac{i \pi f L}{2} \sinc \left[ \pi f L\left(1-k\cdot n_{l} \right) \right] \nn\\
+	& \quad \cdot \exp\left[ i \pi f \left( L + k\cdot \left( p_{1} + p_{2} \right) \right) \right]  n_{l} \cdot P_{\ell m} \cdot n_{l} \,.
+\end{align}
+At leading order in the perturbative formalism of~\cite{MB18}, we have simply
+\be
+	\calT_{slr}^{\ell m}(f) = G_{slr}^{\ell m}(f, t_{f}^{\ell m}) \,,
+\ee
+with $t_{f}^{\ell m}$ the effective time-to-frequency correspondence, defined across the whole frequency band, including in the merger-ringdown, as $t_{f}^{\ell m} = -1/(2\pi) d\Psi_{\ell m} / df$.
+The analysis of~\cite{} has shown that higher-order corrections are small for SMBH systems, and for SOBH systems provided they are close enough to merger as will be the case for the examples chosen here.
+The fact that we use the same Fourier-domain treatment of the transfer functions for both the signal and the templates should also mitigate the importance of modelling errors.
 
@@
 
--- Waveforms: EOBNRv2HMROM + PNext
+As opposed to previous studies of Bayesian parameter recovery for LISA, we wish to use full inspiral-merger-ringdown signals.
+Since higher harmonics will play a crucial role in our analysis, in this study we use the Effective-One-Body (EOB) waveform model \texttt{EOBNRv2HM}~\cite{Pan+11}.
+These waveforms are base on the EOB formalism~\cite{BD99, BD00}, and are calibrated on a set of numerical relativity waveforms.
+The model is limited to non-spinning systems on quasicircular orbits, but includes the most important higher harmonics in the signal:
+\be
+	(\ell, m) = (2,2) ,\, (2,1) ,\, (3,3) ,\, (4,4) ,\, (5,5) \,.
+\ee
+The waveform are generated in time domain by integrating a system of ODEs, which can become quite slow for long signals (up to seconds of integration).
+We developed a reduced order model (ROM) for these waveforms, \texttt{EOBNRv2HM}, following the methods of~\cite{Puerrer14}, allowing a much accelerated generation of the Fourier-domain amplitudes and phases of the modes~\eqref{eq:hlmampphase}.
+
+We note that the limitation to non-spinning waveforms is a stringent one.
+In particular, it means that our results ignore the well-known degeneracy between mass ratio and spin, thus underestimating the uncertainty in the recovered mass ratio.
+By ignoring misaligned spins, we also leave precession effects aside.
+However, consistently including the merger and ringdown in the signal, as well as the higher harmonics, represents a significant improvement over previous studies.
+We also expect intrinsic and extrinsic parameters to correlate weakly, at least in the absence of precession, so that the conclusions drawn from Sec.~\ref{sec:SMBHPE} below, notably the analysis of possible degeneracies in the extrinsic part of the likelihood, should hold for aligned spin systems.
+
+Fast waveforms including both the aligned components of the spins and higher harmonics are becoming available, in the phenomenological waveform family~\cite{London+17}, as EOB models~\cite{Cotesta+18}\footnote{A ROM is under construction to accelerate the waveform model of~\cite{Cotesta+18}.}, and as Numerical Relativity surrogates~\cite{Varma+18}.
+We plan to include the effects of aligned spins, and of misaligned spins causing precession effects, in future studies.
 
@@
 where in each grid interval $P_{j}$ is a cubic polynomial and $\phi_{j}$ a quadratic polynomial.
-We can then compute the elementary integrals~\eqref{eq:fastlikestructure} using a combination of integrations by parts and asymptotic expansions, as explained in~\ref{app:numintegration}.
+We can then compute the elementary integrals~\eqref{eq:fastlikestructure} using a combination of integrations by parts and asymptotic expansions, as explained in App.~\ref{app:numintegration}.
 The number of grid intervals is typically several hundreds \SM{[COMPLETE]}.
@@
 \subsection{Investigating degeneracies}
-\label{sec:SMBHPEdegen}
+\label{subsec:SMBHPEdegen}
 
@@
 The waveform model used to work with SOBH systems is the same phenomenological
-approximant PhenomD used earlier, extended at low frequency with a post-Newtonian
+approximant \texttt{EOBNRv2HMROM} used earlier, extended at low frequency with a post-Newtonian
 frequency-domain waveform. In practice, because LISA only observes the early
@@
 \section{The LISA instrument response}
-\label{sec:response}
+\label{app:response}
 
@@
 
+For completeness, we provide in this Appendix comprehensive definitions and notations for the gravitational wave signal as well as the LISA response.
+We hope to spare the reader the effort of navigating between the various conventions used in the literature, and to provide a point of reference for future works.
+Having these definitions at hand will also help us derive the approximations linking the full response to the simplified response that we used in Sec.~\ref{subsec:SMBHPEdegen}.
+
 %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
@@
 We will further decompose the gravitational wave signal, seen as a function of the direction of emission $(\iota, \varphi)$, in spin-weighted spherical harmonics as
-\be\label{eq:hpcmodes}
+\be\label{eq:apphpcmodes}
 	h_{+} - i h_{\times} = \sum_{\ell \geq 2} \sum_{m = -\ell}^{\ell} {}_{-2}Y_{\ell m} (\iota, \varphi) h_{\ell m} \,,
@@
 
-The definition of the reference time and phase \dots
+The definition of the reference time and phase \SM{[\dots COMPLETE]}
 
@@
 
+\SM{[avoid repetition with methodology section]}
+
 The single-link frequency shift observable $y_{slr} = (\nu_{r} - \nu_{s})/\nu$ between sending spacecraft $s$ and receiving spacecraft $r$ along the link $l$ takes the form 
-\be\label{eq:defyslr}
+\be\label{eq:appdefyslr}
 	y_{slr} = \frac{1}{2} \frac{n_{l} \otimes n_{l}}{1 - k\cdot n_{l}} : \left[ H(t - L - k\cdot p_{s}) - H(t - k\cdot p_{r}) \right] \,.
@@
 
-In our rigid approximation, delays are simply all constant and constant to $L$. Using the shortcut notation $y_{sr,L} = y_{slr}(t - L)$, the first-generation TDI Michelson observable $X$ reads
+In our rigid approximation, delays are simply all constant and constant to $L$. Using the shortcut notation $y_{sr,L} = y_{slr}(t - L)$, the first-generation TDI Michelson observable $X$ reads~\cite{}
 \begin{align}

### 2019-04-09 paper/paper_PE.tex
Updated sampling section.

@@
 
-Fast waveforms including both the aligned components of the spins and higher harmonics are becoming available, in the phenomenological waveform family~\cite{London+17}, as EOB models~\cite{Cotesta+18}\footnote{A ROM is under construction to accelerate the waveform model of~\cite{Cotesta+18}.}, and as Numerical Relativity surrogates~\cite{Varma+18}.
+Fast waveforms including both the aligned components of the spins and higher harmonics are becoming available, in the phenomenological waveform family~\cite{London+17}, as EOB models~\cite{Cotesta+18}~\footnote{A ROM is under construction to accelerate the waveform model of~\cite{Cotesta+18}.}, and as Numerical Relativity surrogates~\cite{Varma+18}.
 We plan to include the effects of aligned spins, and of misaligned spins causing precession effects, in future studies.
@@
 
-For this exploratory work we take a ``naive'' approach to  estimating posterior-probability distributed parameter samples, not applying any specific knowledge or expectations about distributions which might plausibly lead to artifically robust-seeming but incomplete results.
-We apply two independent approaches, an parallel tempering MCMC sampler with generically wtih amotivated proposal distribution, and nested sampling.
-\jgb{More to come}
-
--- Bayesian samplers: Multinest, PTMC
+For this exploratory work we take a ``naive'' approach to  estimating posterior-probability distributed parameter samples, without applying specific knowledge or expectations about distributions which might plausibly lead to artifically robust-seeming but incomplete results.
+We apply two independent approaches, a parallel tempering Markov-chain Monte Carlo (MCMC) sampler and nested sampling.
+
+Our MCMC code \texttt{PTMCMC}\footnote{Open source at github.com/JohnGBaker/ptmcmc.} has been developed through several astrophysics projects, tested on several sample problems, and modularly designed with the aim of minizing opportunities for errors with new projects.
+The code performs parallel tempered MCMC\cite{Swendsen+1986,Littenberg2009} with temperatures slowly adjusted to achieve nearly equal exchange rates among all pairs of adjacent temperatures.
+Most runs here performed comparably with either 80 or 240 parallel temperature chains.
+We applied a general-purpose proposal distribution composed of a weighted set of sub-proposals, including differential-evolution steps~\cite{Vrugt+2008} (although the proposal in this case is based on each single chain, not an ensemble) and a collection of several Gaussian step draws of varying sizes scaled off the prior domain at about 0.01--1\% of its scale and weighted favor the smaller scale proposal draws.
+Effective sample sizes (ESS) were estimated from the number of (post-burn-in) samples, reduced by the autocorrelation length, with the burn-in size chosen to optimize the ESS.
+In most cases, the runs were continued until achieving an estimated ESS~$>2000$.
+
+For comparison and verification we also computed posterior samples using the nested sampling code \texttt{BAMBI}\cite{Graff+2012} a variant of \texttt{MultiNest}\cite{Feroz+2009b} with additional options (unused here).
+The nested sampling algorithm evolves a set of 'live points' toward regions of high probability by iteratively resampling the lowest probability point from within an ever-narrowing region covering the set of live-point samples.
+The computation proceeds until the estimated Bayesian evidence, or marginal probability, within the region covered by the live points is smaller than some threshold.
+Posterior-distributed samples are then drawn based on the set of sub-regions and sample posterior values thereby generated.
+In our runs we used 4000 live points and with an stopping evidence threshhod of 1.0.
+\jgb{Sylvain needs to check these values.}
 
@@
 8.89e5,1.11e5,0, 3.67e4,pi/3,pi/2,3pi/4,pi/3,pi/3
+
+--enumerate priors
 

### 2019-04-23 paper/paper_PE.tex
Add my acknowledgments
Just some comments so far.

@@
 \author{Tito Dal Canton}
+\thanks{NASA Postdoctoral Program fellow}
 \affiliation{Gravitational Astrophysics Laboratory, NASA Goddard Space Flight Center, 8800 Greenbelt Rd., Greenbelt, MD 20771, USA}
@@
 \ee
-From the low-frequency response \eqref{eq:aelowfmodes}, pinning the time-at-LISA~\eqref{} \SM{[COMPLETE $t_{L}$]} so that we can ignore the factor $\exp [2 i \pi f k\cdot p_{0}]$ as a pure constant corresponding to a redefinition of time, we obtain
+From the low-frequency response \eqref{eq:aelowfmodes}, pinning the time-at-LISA~\eqref{} \SM{[COMPLETE $t_{L}$]} so that we can ignore the factor $\exp [2 i \pi f k\cdot p_{0}]$ as a pure constant corresponding to a redefinition of time, we obtain \jgb{Should we use notation $\Phi_R$ here and in \eqref{eq:aelowfmodes}?}
 \begin{align}\label{eq:simplelikelihoodhm}
@@
 \end{align}
-where we introduced the following extrinsic signal functions
+where we introduced the following extrinsic signal functions \jgb{Since neither $D_{a,e}$ nor $P_{\ell m}$ are defined in the main text, may be better to write in terms of a version of $F$ as in \eqref{eq:defFapcFepc}}
 \be\label{eq:defsaelm}
@@
 with $d = D / D^{\rm inj}$ the dimensionless ratio of luminosity distances.
-The prefactors $\left\langle \ell m | \ell' m' \right\rangle$ are pure normalization constants, and represent mode overlaps as
+The prefactors $\left\langle \ell m | \ell' m' \right\rangle$ \sout{are pure normalization constants, and} represent mode overlaps as
 \be\label{eq:innerproductlmlpmp}
@@
 Together with the luminosity distance $D_{L}$, these parameters appear in combinations of constant factors multiplying the whole response.
+\jgb{Would it be better to include the angle dependence explicitly here in the $\sYlm$ here for clarity, since the point is to account for parameter dependencies.}
 
@@
 \ee
-Several assumptions enter this form of the response: \dots \SM{[COMPLETE]}.
+Several assumptions enter this form of the response: \dots \SM{[COMPLETE]} \jgb{Define $n_l$. Also double dot looks like colon, needing some clarification.}.
 
@@
 \ee
-Note a degeneracy $\pm \pi$ in how we choose to orient these effective detectors. We can recover pattern functions (for $\psi_{L}=0$) by factoring out the effective length, defining
+Note a degeneracy $\pm \pi$ in how we choose to orient these effective detectors.
+We can recover pattern functions (for $\psi_{L}=0$) by factoring out the effective length, we define
+\jgb{Would it be better to just include a new version of expression \eqref{eq:aelowfmodes} here using $F_{a,e}$, such as:
+  $\tilde{a}, \tilde{e} = (-3i\pi f L)\left[ \tilde{h}_{\ell m}^\times F_{a,e}^+ + \tilde{h}_{\ell m}^\times F_{a,e}^\times\right]$.
+Also an $(\ell,m)$ version of $F_{a,e}$ would be useful in the main text.}
 \be\label{eq:defFapcFepc}
@@
 
+\begin{acknowledgments}
+
+TDC was supported by an appointment to the NASA Postdoctoral Program at the
+Goddard Space Flight Center, administered by Universities Space Research
+Association under contract with NASA.
+
+\end{acknowledgments}
+
 \bibliography{references.bib}

### 2019-04-26 paper/paper_PE.tex
Proposed changes and comments after Fridayss read

@@
 \newcommand{\jgb}[1]{{\color{DarkGreen} #1}}
-\newcommand{\tdc}[1]{{\color{red} #1}}
+\newcommand{\tdc}[1]{{\color{red} (TD: #1)}}
 
@@
 \ee
-With the decomposition~\eqref{eq:Hsummodes}, the corresponding form of the kernel $G^{\ell m}_{slr}(f, t)$ in the formalism of~\cite{MB18} is
+Applying the perturbative formalism of~\cite{MB18} to leading order, we have simply
+\be
+	\calT_{slr}^{\ell m}(f) = G_{slr}^{\ell m}(f, t_{f}^{\ell m}) \,,
+\ee
+where
 \begin{align}\label{eq:Gslr}
 	G_{slr}^{\ell m}(f,t) &= \frac{i \pi f L}{2} \sinc \left[ \pi f L\left(1-k\cdot n_{l} \right) \right] \nn\\
-	& \quad \cdot \exp\left[ i \pi f \left( L + k\cdot \left( p_{1} + p_{2} \right) \right) \right]  n_{l} \cdot P_{\ell m} \cdot n_{l} \,.
-\end{align}
-At leading order in the perturbative formalism of~\cite{MB18}, we have simply
-\be
-	\calT_{slr}^{\ell m}(f) = G_{slr}^{\ell m}(f, t_{f}^{\ell m}) \,,
-\ee
-with $t_{f}^{\ell m}$ the effective time-to-frequency correspondence, defined across the whole frequency band, including in the merger-ringdown, as $t_{f}^{\ell m} = -1/(2\pi) d\Psi_{\ell m} / df$.
+	& \quad \cdot \exp\left[ i \pi f \left( L + k\cdot \left( p_{1} + p_{2} \right) \right) \right] \nn\\
+	& \quad \cdot n_{l} \cdot P_{\ell m} \cdot n_{l} \,
+\end{align}
+\tdc{what are $p_1$ and $p_2$ here? Are they the same as $p_r$ and $p_s$?}
+is the kernel from~\cite{MB18}, $P_{\ell m}$ is defined by the decomposition~\eqref{eq:Hsummodes}, and
+\be
+	t_{f}^{\ell m} = -\frac{1}{2\pi} \frac{d\Psi_{\ell m}}{df}
+\ee
+is the effective frequency$\to$time correspondence, defined across the whole frequency band and including the merger and ringdown.
 The analysis of~\cite{}  has shown that higher-order corrections are small for SMBH systems, and for SOBH systems provided they are close enough to merger as will be the case for the examples chosen here.
@@
 Since higher harmonics will play a crucial role in our analysis, in this study we use the Effective-One-Body (EOB) waveform model \texttt{EOBNRv2HM}~\cite{Pan+11}.
-These waveforms are base on the EOB formalism~\cite{BD99, BD00}, and are calibrated on a set of numerical relativity waveforms.
+These waveforms are based on the EOB formalism~\cite{BD99, BD00}, and are calibrated on a set of numerical relativity waveforms.
 The model is limited to non-spinning systems on quasicircular orbits, but includes the most important higher harmonics in the signal:
@@
 \ee
-The waveform are generated in time domain by integrating a system of ODEs, which can become quite slow for long signals (up to seconds of integration).
-We developed a reduced order model (ROM) for these waveforms, \texttt{EOBNRv2HM}, following the methods of~\cite{Puerrer14}, allowing a much accelerated generation of the Fourier-domain amplitudes and phases of the modes~\eqref{eq:hlmampphase}.
-
-We note that the limitation to non-spinning waveforms is a stringent one.
-In particular, it means that our results ignore the well-known degeneracy between mass ratio and spin, thus underestimating the uncertainty in the recovered mass ratio.
-By ignoring misaligned spins, we also leave precession effects aside.
-However, consistently including the merger and ringdown in the signal, as well as the higher harmonics, represents a significant improvement over previous studies.
+The waveforms are generated in the time domain by integrating a system of ODEs, an operation requiring up to seconds of computing time for long signals.
+For this reason, we developed a reduced order model (ROM) for these waveforms, \texttt{EOBNRv2HM}, following the methods of~\cite{Puerrer14}, enabling a much faster generation of the Fourier-domain amplitudes and phases of the modes~\eqref{eq:hlmampphase}.
+
+We note that the assumption of non-spinning black holes is an important limitation to a realistic assessment of parameter uncertainties.
+In particular, it means that our results ignore the well-known degeneracy between mass ratio and spin \cite{Baird+2013}, thus underestimating the uncertainty in the recovered mass ratio.
+By ignoring misaligned spins, we also neglect the possible effects of orbital precession on parameter degeneracies.
+Nevertheless, consistently including merger, ringdown and higher harmonics already represents a significant improvement over previous studies.
 We also expect intrinsic and extrinsic parameters to correlate weakly, at least in the absence of precession, so that the conclusions drawn from Sec.~\ref{sec:SMBHPE} below, notably the analysis of possible degeneracies in the extrinsic part of the likelihood, should hold for aligned spin systems.
-
-Fast waveforms including both the aligned components of the spins and higher harmonics are becoming available, in the phenomenological waveform family~\cite{London+17}, as EOB models~\cite{Cotesta+18}~\footnote{A ROM is under construction to accelerate the waveform model of~\cite{Cotesta+18}.}, and as Numerical Relativity surrogates~\cite{Varma+18}.
-We plan to include the effects of aligned spins, and of misaligned spins causing precession effects, in future studies.
+\tdc{the last sentence would fit well in the conclusion.}
+
+Fast waveforms including both the aligned components of the spins and higher harmonics are becoming available in the phenomenological waveform family~\cite{London+17}, as EOB models~\cite{Cotesta+18}, and as Numerical Relativity surrogates~\cite{Varma+18}.
+ROMs to accelerate the waveform model of~\cite{Cotesta+18} are under construction, and we plan to include the effects of aligned and misaligned spins in future studies.
+\tdc{this paragraph also sounds like conclusion material.}
 
@@
 In the LISA.
+\tdc{this paragraph seems to need more work.
+I would start with the existing general case, then say that by using a zero-noise realization we can write the amplitude/phase representation and use a sparse grid and get a large speedup.}
 
@@
 Note that individual terms in~\eqref{eq:lnLoverlapstructure} can be individually large, and computing $\ln \calL$ relies on accurate cancellations.
-Decomposing further in TDI channels as in~\eqref{} and in harmonics as in~\eqref{}, we have a sum of terms with the structure
+Decomposing further in TDI channels as in~\eqref{} and in harmonics as in~\eqref{}, we have a sum of terms with the structure \tdc{we need one or two more equations describing explicitly how to go from $\tilde{y}_{slr}$ to $\ln \calL$}
 \be
@@
 \ee
+\tdc{should this notation be aligned to Eq 6 and 7?}
 The difficulty of computing such an overlap depends on the phase difference.
@@
 
-For this exploratory work we take a ``naive'' approach to  estimating posterior-probability distributed parameter samples, without applying specific knowledge or expectations about distributions which might plausibly lead to artifically robust-seeming but incomplete results.
-We apply two independent approaches, a parallel tempering Markov-chain Monte Carlo (MCMC) sampler and nested sampling.
-
-Our MCMC code \texttt{PTMCMC}\footnote{Open source at github.com/JohnGBaker/ptmcmc.} has been developed through several astrophysics projects, tested on several sample problems, and modularly designed with the aim of minizing opportunities for errors with new projects.
-The code performs parallel tempered MCMC\cite{Swendsen+1986,Littenberg2009} with temperatures slowly adjusted to achieve nearly equal exchange rates among all pairs of adjacent temperatures.
+For this exploratory work we take a ``naive'' \tdc{``brute force''?} approach to inferring the posterior probability distribution of the parameters, without applying specific knowledge or expectations about distributions and degeneracies.
+Such assumptions might in fact lead to an incomplete exploration of the posterior probability in ways that would be difficult to recognize and diagnose.
+In order to demonstrate that the problem is tractable without critically depending on the details of the methodology, we also use two independent approaches to sample the posterior distributions: parallel tempering Markov-chain Monte Carlo (MCMC) and nested sampling.
+
+Our MCMC code \texttt{PTMCMC}\footnote{https://github.com/JohnGBaker/ptmcmc} has been developed through several astrophysics projects, tested on several sampling problems, and modularly designed with the aim of minizing opportunities for errors with new applications.
+The code performs parallel tempered MCMC \cite{Swendsen+1986,Littenberg2009} with temperatures slowly adjusted to achieve nearly equal exchange rates among all pairs of adjacent temperatures.
 Most runs here performed comparably with either 80 or 240 parallel temperature chains.
 We applied a general-purpose proposal distribution composed of a weighted set of sub-proposals, including differential-evolution steps~\cite{Vrugt+2008} (although the proposal in this case is based on each single chain, not an ensemble) and a collection of several Gaussian step draws of varying sizes scaled off the prior domain at about 0.01--1\% of its scale and weighted favor the smaller scale proposal draws.
-Effective sample sizes (ESS) were estimated from the number of (post-burn-in) samples, reduced by the autocorrelation length, with the burn-in size chosen to optimize the ESS.
+Effective sample sizes (ESS) were estimated from the number of post-burn-in samples, reduced by the autocorrelation length, with the burn-in size chosen to optimize \tdc{maximize?} the ESS.
 In most cases, the runs were continued until achieving an estimated ESS~$>2000$.
 
-For comparison and verification we also computed posterior samples using the nested sampling code \texttt{BAMBI}\cite{Graff+2012} a variant of \texttt{MultiNest}\cite{Feroz+2009b} with additional options (unused here).
+For comparison and verification we also computed posterior samples using the nested sampling code \texttt{BAMBI} \cite{Graff+2012}, a variant of \texttt{MultiNest} \cite{Feroz+2009b} with additional options (unused here).
 The nested sampling algorithm evolves a set of 'live points' toward regions of high probability by iteratively resampling the lowest probability point from within an ever-narrowing region covering the set of live-point samples.
@@
 Posterior-distributed samples are then drawn based on the set of sub-regions and sample posterior values thereby generated.
-In our runs we used 4000 live points and with an stopping evidence threshhod of 1.0.
+In our runs we used 4000 live points and a stopping evidence threshold of 1.0.
 \jgb{Sylvain needs to check these values.}
@@
 
-For comparison with our Bayesian inference results we also compute estimates of parameter estimation precision using the Fisher information matrix the common workhorse of LISA science studies despite its limitations.~\cite{Vallisneri08}  For measurements with additive Gaussian noise, as assumed here, the Fisher information matrix can be computed by
+For comparison with our Bayesian inference results we also compute estimates of parameter uncertainties using the Fisher information matrix, which despite its limitations is the common workhorse of LISA science studies to date ~\cite{Vallisneri08}.
+For measurements with additive Gaussian noise, as assumed here, the Fisher information matrix can be computed by
 \be
@@
 If we can neglect entirely the motion of the detector (which is not quite true as we will see later on), then we are in a regime where the information about the sky position of the source comes essentially from the pattern functions affecting this high-SNR signal.
-Using the the low-frequency approximation (see~\ref{}) and the time-at-LISA $t_{L}$ as a time variable, there is no frequency dependence in the response.
+Using the low-frequency approximation (see~\ref{}) and the time-at-LISA $t_{L}$ as a time variable, there is no frequency dependence in the response.
 A further simplification occurs when pinning the masses and the time of coalescence to their true values.
 We found indeed that these parameters correlate only weakly to the other ones, so that we can hope to understand the degeneracy features by restricting the analysis to the six extrinsic parameters $(D_{L}, \iota, \varphi, \lambdaL, \betaL, \psiL)$.
+\tdc{we could make the connection with a similar decoupling in LIGO/Virgo, which is actually what makes low-latency sky localization possible.}
 
@@
 We will further abbreviate notations by using the variables
+\tdc{it is unclear if these mean $\tan (\iota/2)$ or $(\tan \iota)/2$}
 \be
@@
 leaves the likelihood exactly invariant.
-There is also a form of symmetry obatained by exchanging $\iota \leftrightarrow \theta_{L}$.
+There is also a form of \tdc{also exact?} symmetry obtained by exchanging $\iota \leftrightarrow \theta_{L}$.
 This conjugate the term inside bracket for $\sigma_{+}$, which can be corrected withe the phase terms $\varphi \pm \lambda_{a}$.
@@
 
-Finally, when considering more harmonics beyond the dominant mode $h_{22}$, the degeneracies will be broken easily.
+When considering more harmonics beyond the dominant mode $h_{22}$, such degeneracies will be broken easily.
 As in the case of the full likelihood, the simplified likelihood~\eqref{eq:simplelikelihoodhm} will have several terms with different inclination and phase dependencies.

### 2019-05-10 paper/paper_PE.tex
degen

--- 2019-04-23 paper/paper_PE.tex
@@
 \esub
-with a common prefactor
+with a common prefactor \jgb{This seems to need to  be multiplied by -6 to get what I have in my notes. Trace to change in def of $s_{a,e}$?}
 \be\label{eq:sigmafactorrho}
@@
 \ee
-Finding points in the parameter space that are degenerate with the injection, with $\ln \calL \simeq 0$, amounts to try to obtain both the same value of $\sigma_{+}$ and $\sigma_{-}$ as the injection.
+Finding points in the parameter space that are degenerate with the injection, i.e.
+with $\ln \calL \simeq 0$, amounts finding choices for the parameters which obtain the same values for both  $\sigma_{+}$ and $\sigma_{-}$ as the injection.
 
@@
 There is also a form of symmetry obatained by exchanging $\iota \leftrightarrow \theta_{L}$.
-This conjugate the term inside bracket for $\sigma_{+}$, which can be corrected withe the phase terms $\varphi \pm \lambda_{a}$.
+While leaving $\sigma_-$ unchanged, this conjugates the factor inside brackets for $\sigma_{+}$, which can be compensated using the phase terms $\varphi \pm \lambda_{a}$.
 If $\Phi = \mathrm{Arg} \left[ t_{\theta}^{4} e^{-2 i \psiL} + t_{\iota}^{4} e^{2 i \psiL} \right]$, we obtain the symmetry
 \begin{align}
-	\iota' = \theta_{L} \,, \quad \theta_{L}' = \iota \,, \nn\\
-	\varphi' = \varphi + \frac{1}{2} \Phi \; \mathrm{mod} \; \pi\,, \nn\\
-	\lambda_{a}' = \lambda_{a} - \frac{1}{2} \Phi \; \mathrm{mod} \; \pi\,.
-\end{align}
-
-We can now explain the degenerate features we found as certain limits in this simplified likelihood.
+	\iota' = \theta_{L} \,, \qquad \theta_{L}' = \iota \,, \nn\\
+	\varphi' = \varphi + \frac{1}{2} \Phi \quad \mathrm{mod} \; \pi\,, \nn\\
+	\lambda_{a}' = \lambda_{a} - \frac{1}{2} \Phi \quad \mathrm{mod} \; \pi\,.
+\end{align}
+
+\jgb{Don't have to resort to limits.}
+\sout{We can now explain the degenerate features we found as certain limits in this simplified likelihood.
 In the face-on or face-off limit $\iota \rightarrow 0$, $|t_{\iota}| \ll 1$.
 This amounts to the $(\ell,m)=(2,2)$ contribution dominating the $(\ell, m)=(2,-2)$ contribution.
-Neglecting the terms in~\eqref{eq:sigmapm} that are quartic in $t_{\iota}$, we have
-\bsub\label{eq:sigmapmapprox}
-\begin{align}
-	\sigma_{+} &\simeq \rho e^{2i(\varphi - \psiL)} t_{\theta}^{4} e^{-2i \lambda_{a}} \,, \\
-	\sigma_{-} &\simeq \rho e^{2i(\varphi - \psiL)} e^{2i \lambda_{a}} \,,
-\end{align}
-\esub
+Neglecting the terms in~\eqref{eq:sigmapm} that are quartic in $t_{\iota}$, we have}
+%\bsub\label{eq:sigmapmapprox}
+%\begin{align}
+%	\sigma_{+} &\simeq \rho e^{2i(\varphi - \psiL)} t_{\theta}^{4} e^{-2i \lambda_{a}} \,, \\
+%	\sigma_{-} &\simeq \rho e^{2i(\varphi - \psiL)} e^{2i \lambda_{a}} \,,
+%\end{align}
+%\esub
+
+Beyond these discrete symetries, since likelihood function dependence on the six extrinsic parameters is funneled through just two complex functions, we should expect a two-dimensional degenerate subspace.
+Indeed we can explicitly find a general solution for parameter values that solve $\sigma_{\pm}((D_{L}, t_\iota, t_\theta \varphi, \lambda_a, \psiL)=\sigma_{\pm,{\rm inj}}$, which we just sketch here, as needed to explain features of the degeneracies.
+
 Defining the ratio
 \be
-	r = \frac{\sigma_{+}}{\sigma_{-}} \,,
-\ee
-we see that finding $(\lambda_{a}^{*}, \theta_{L}^{*})$ verifying
-\be\label{eq:degenconditionsky}
-	\frac{\sigma_{+}}{\sigma_{-}} (\lambda_{a}^{*}, \theta_{L}^{*}) = r_{\rm inj} \,,
-\ee
-with the approximation~\eqref{eq:sigmapmapprox} for the left-hand side, guarantees that we can then match both $\sigma_{+}$ and $\sigma_{-}$ of the injection with the condition
-\be\label{eq:degenconditionprefactor}
-	\rho e^{2i(\varphi - \psiL)} e^{2i \lambda_{a}^{*}} = \sigma_{-}^{\rm inj}\,.
-\ee
-In the limit $|t_{\iota}| \ll 1$,
-\be
-	\frac{\sigma_{+}}{\sigma_{-}} \simeq t_{\theta}^{4} e^{-4i\lambda_{a}} \,,
-\ee
-so that~\eqref{eq:degenconditionsky} is verified for special positions in the sky at (coming back to the original variables $(\lambdaL, \betaL)$)
-\bsub\label{eq:soldegensky}
-\begin{align}
-	\beta_{L}^{*} &= \frac{\pi}{2} - 2 \mathrm{Arctan} \; \left| r_{\rm inj}\right|^{1/4} \,, \\
-	\lambda_{L}^{*} &= \frac{\pi}{6} -\frac{1}{4} \mathrm{Arg} \; r_{\rm inj} \;\mathrm{mod}\; \frac{\pi}{2} \,.
-\end{align}
-\esub
+	r(\lambda_a, t_\iota, t_\theta, \psiL) = \frac{\sigma_{+}}{\sigma_{-}} \,,
+\ee
+makes clear that $r=r_{\rm inj}$ provides one complex condition on 4 real unknown parameters, eliminating parameters $D_{L}$ and $\varphi$ while retaining the features of the full degenerate subspace.
+Going further, $|r|^2=|r_{\rm inj}|^2$ yields one condition on 3 parameters, eliminating $\lambda_a$.
+In fact, with a little rearrangement, this can be written as a quadratic expression for either $t_\iota^4$ or $t_\theta^4$ for an explicit solution.
+Depending on the value of $|r_{\rm inj}|^2$, values of $\psiL$ and the independent variable ($t_\theta$ or $t_i$) can always be chosen to guarantee a real, positive solution.
+
+Given a solution for $(t_\iota,t_\theta,\psi_L)$, the rest of the solution then proceeds backwards.
+An explicit expression for commensurate $\lambda$ comes from solving $r=r_{\rm inj}$, and then $D_L$ and $\varphi$ are obtained from solving, e.g., $s_+=s_{+,{\rm inj}}$.
+
+In the limit that the independent variable is given a value far from unity (near polar $\iota$ ot $\beta_L$), then the solution for the dependent variable approaches a fixed constant special values $t_{\iota,\theta}^4={|r_{\rm inj}|^{\pm1}}$, with leading order corrections proportional to  $t_{\theta,\iota}^{4}$.
+A large part of the degenerate subspace volume then tends to found with parameters $\iota$ ot $\beta_L$ near these special values with corresponding constraints on $\lambda$ coming from the solution for the complex argument of $r$.
+
+In terms of the original parameters then, if we consider inferences with the parameter $\iota$ that is not near the source equator, then we find a significant part of the degenerate maximum-likelihood parameter space volume near the special sky positions
+\bsub
+\begin{align}
+  \beta_L&\approx{\beta_L}^{*}=\pm\left(\frac{\pi}{2} - 2 \mathrm{Arctan} \; \left| r_{\rm inj}\right|^{1/4} \right)\,, \\
+  \lambda_a&\approx {\lambda_a}^{*}=\frac{\pi}{6} -\frac{1}{4} \mathrm{Arg} \; r_{\rm inj} \;\mathrm{mod}\; \frac{\pi}{2} \,,\\
+  \rho e^{2i(\varphi-\psiL)} &\approx e^{-2i\lambda_{a}^{*}} \sigma_{-}^{\rm inj} \,.
+\end{align}
+\esub
+With the $\lambdaL \rightarrow \lambdaL + k \pi/2$ and $\betaL \rightarrow - \betaL$ symmetries, we arrive at eight sky positions where the degeneracy occurs, with an exact degeneracy for the pair of parameters $(\varphi, \psiL)$, as well as an approximate degeneracy for the pair $(d, \iota)$ as long as we remain in the regime $t_{\iota}^{4} \ll 1$ or $t_{\iota}^{4} \gg 1$ (which is quite extended, thanks to the quartic power).
+For each of these sky positions built from~\eqref{eq:soldegensky}, many different values of $(\varphi, \psiL)$ and $(d, \iota)$ produce a waveform very close to the injection, resulting in apparent peaks in the marginalized posterior distribution for the sky position, located at the special sky positions $(\lambda_{a}^{*}, \theta_{L}^{*})$, which are offset from the injected value.
+
+Without special tuning, the injected sky position which led to the given values of $s_{a,\mathrm{inj}}$ and $s_{e,\mathrm{inj}}$ is correspondingly likely to have been near one of these sky locations, but it may not have been.
+In those cases the posterior\ldots
+
+Similarly, considering inferred locations away from the LISA-constellation plane, we find a significant part of the degenerate parameter space near special values for inclination and correspondingly for $\lambda_L+\varphi$
+\bsub
+\begin{align}
+	\iota&\approx\iota^{*} = 2 \mathrm{Arctan} \; \left| r_{\rm inj}\right|^{1/4} \,, \\
+	\lambda_a-\psiL&\approx (\lambda_{a} - \psiL)^{*} = -\frac{1}{4} \mathrm{Arg} \; r_{\rm inj} \;\mathrm{mod}\; \frac{\pi}{2} \,\\
+        \rho e^{2i\varphi} &\approx e^{-2i (\lambda_{a} - \psi_{L})^{*}} \sigma_{-}^{\rm inj} \,.
+\end{align}
+\esub
+
+%finding $(\lambda_{a}^{*}, \theta_{L}^{*})$ verifying
+%\be\label{eq:degenconditionsky}
+%	\frac{\sigma_{+}}{\sigma_{-}} (\lambda_{a}^{*}, \theta_{L}^{*}) = r_{\rm inj} \,,
+%\ee
+%with the approximation~\eqref{eq:sigmapmapprox} for the left-hand side, guarantees that we can then match both $\sigma_{+}$ and $\sigma_{-}$ of the injection with the condition
+%\be\label{eq:degenconditionprefactor}
+%	\rho e^{2i(\varphi - \psiL)} e^{2i \lambda_{a}^{*}} = \sigma_{-}^{\rm inj}\,.
+%\ee
+%In the limit $|t_{\iota}| \ll 1$,
+%\be
+%	\frac{\sigma_{+}}{\sigma_{-}} \simeq t_{\theta}^{4} e^{-4i\lambda_{a}} \,,
+%\ee
+%so that~\eqref{eq:degenconditionsky} is verified for special positions in the sky at (coming back to the original variables $(\lambdaL, \betaL)$)
+%\bsub\label{eq:soldegensky}
+%\begin{align}
+%\beta_{L}^{*} &= \frac{\pi}{2} - 2 \mathrm{Arctan} \; \left| r_{\rm inj}\right|^{1/4} \,, \\
+%	\lambda_{L}^{*} &= \frac{\pi}{6} -\frac{1}{4} \mathrm{Arg} \; r_{\rm inj} \;\mathrm{mod}\; \frac{\pi}{2} \,.
+%\end{align}
+%\esub
+\jgb{Keep something from this?}
 The other condition~\eqref{eq:conditionprefactor} gives then a single phase constraint for $\varphi - \psiL$, and a single amplitude constraint for $\rho(d, \iota, \theta_{L}^{*})$.
@@
 
-With the $\lambdaL \rightarrow \lambdaL + k \pi/2$ and $\betaL \rightarrow - \betaL$ symmetries, we arrive at eight sky positions where the degeneracy occurs, with an exact degeneracy for the pair of parameters $(\varphi, \psiL)$, as well as an approximate degeneracy for the pair $(d, \iota)$ as long as we remain in the regime $t_{\iota}^{4} \ll 1$ or $t_{\iota}^{4} \gg 1$ (which is quite extended, thanks to the quartic power).
-For each of these sky positions built from~\eqref{eq:soldegensky}, many different values of $(\varphi, \psiL)$ and $(d, \iota)$ produce a waveform very close to the injection, resulting in apparent peaks in the marginalized posterior distribution for the sky position, located at the special sky positions $(\lambda_{a}^{*}, \theta_{L}^{*})$, which are offset from the injected value.
-
-We can further find the same mechanism, with a few differences, by exchanging the roles of $(\iota, \varphi)$ and $(\theta_{L}, \lambda_{a})$, as follows.
+%We can further find the same mechanism, with a few differences, by exchanging the roles of $(\iota, \varphi)$ and $(\theta_{L}, \lambda_{a})$, as follows.
 In the limit $\theta_{L} \rightarrow 0$ with $|t_{\theta}| \ll 1$, we have this time
-\be
-	\frac{\sigma_{+}}{\sigma_{-}} \simeq t_{\iota}^{4} e^{-4i(\lambda_{a} - \psi_{L})} \,,
-\ee
-which gives in the same way a degeneracy if
-\bsub\label{eq:soldegeniotaphi}
-\begin{align}
-	\iota^{*} &= 2 \mathrm{Arctan} \; \left| r_{\rm inj}\right|^{1/4} \,, \\
-	(\lambda_{a} - \psiL)^{*} &= -\frac{1}{4} \mathrm{Arg} \; r_{\rm inj} \;\mathrm{mod}\; \frac{\pi}{2} \,,
-\end{align}
-\esub
-with a condition on $\rho$ and $\varphi$ that is now
-\be
-	\rho e^{2i\varphi} e^{2i (\lambda_{a}^{*} - \psi_{L}^{*})} = \sigma_{-}^{\rm inj} \,.
-\ee
+%\be
+%	\frac{\sigma_{+}}{\sigma_{-}} \simeq t_{\iota}^{4} e^{-4i(\lambda_{a} - \psi_{L})} \,,
+%\ee
+%which gives in the same way a degeneracy if
+%\bsub\label{eq:soldegeniotaphi}
+%\begin{align}
+%	\iota^{*} &= 2 \mathrm{Arctan} \; \left| r_{\rm inj}\right|^{1/4} \,, \\
+%	(\lambda_{a} - \psiL)^{*} &= -\frac{1}{4} \mathrm{Arg} \; r_{\rm inj} \;\mathrm{mod}\; \frac{\pi}{2} \,,
+%\end{align}
+%\esub
+%with a condition on $\rho$ and $\varphi$ that is now
+%\be
+%\rho e^{2i\varphi} e^{2i (\lambda_{a}^{*} - \psi_{L}^{*})} = \sigma_{-}^{\rm inj} \,.
+%\ee
+\jgb{Keep something from this?}
 Thus, we find again a degenerate pattern, with the difference that \eqref{eq:soldegeniotaphi} imposes a condition on the pair $\lambda_{a} - \psiL$ instead of $\lambda_{a}$, and the extra condition imposes a condition on $\varphi$ instead of the pair $\varphi - \psiL$.

### 2019-05-10 paper/paper_PE.tex

@@
 leaves the likelihood exactly invariant.
-There is also a form of symmetry obatained by exchanging $\iota \leftrightarrow \theta_{L}$.
+There is also a symmetry based on exchanging $\iota \leftrightarrow \theta_{L}$.
 While leaving $\sigma_-$ unchanged, this conjugates the factor inside brackets for $\sigma_{+}$, which can be compensated using the phase terms $\varphi \pm \lambda_{a}$.

### 2019-05-10 paper/paper_PE.tex
Merge

@@
 
-\newcommand{\SM}[1]{{\color{Blue} #1}}
-\newcommand{\jgb}[1]{{\color{DarkGreen} #1}}
-\newcommand{\tdc}[1]{{\color{red} (TD: #1)}}
+\newcommand{\SM}[1]{{\color{Blue} [SM: #1]}}
+\newcommand{\JGB}[1]{{\color{DarkGreen} [JGB: #1]}}
+\newcommand{\TDC}[1]{{\color{red} [TDC: #1]}}
 
@@
 
-\title{Bayesian methods for black hole merger parameter estimation with LISA.}
-
+%\title{Bayesian methods for black hole merger parameter estimation with LISA.}
+\title{Exploring Bayesian parameter estimation of binary black holes with LISA}
+
+\author{Sylvain Marsat}
+\affiliation{APC, AstroParticule et Cosmologie, Universit\'{e} Paris Diderot, CNRS/IN2P3, CEA/Irfu, Observatoire de Paris, Sorbonne Paris Cit\'{e}, 10, rue Alice Domon et L\'{e}onie Duquet 75205 PARIS Cedex 13, France}
 \author{John G. Baker}
 \affiliation{Gravitational Astrophysics Laboratory, NASA Goddard Space Flight Center, 8800 Greenbelt Rd., Greenbelt, MD 20771, USA}
-\author{Sylvain Marsat}
-\affiliation{APC, AstroParticule et Cosmologie, Universit\'{e} Paris Diderot, CNRS/IN2P3, CEA/Irfu, Observatoire de Paris, Sorbonne Paris Cit\'{e}, 10, rue Alice Domon et L\'{e}onie Duquet 75205 PARIS Cedex 13, France}
 \author{Tito Dal Canton}
@@
 
-[Abstract]
+The space-based gravitational wave detector LISA will observe mergers of supermassive black hole binary systems (SMBHs) to cosmological distances, as well as inspiralling stellar-origin (or stellar-mass) binaries (SOBHs) years before they enter the LIGO/Virgo band.
+Much remains to be explored for the parameter recovery of both class of systems.
+Previous SMBH analyses relied on inspiral-only signals and/or a simplified Fisher matrix analysis, while SOBHs have not yet been analyzed with Bayesian methods.
+We accelerate likelihood computations by (i) using a Fourier-domain response of the LISA instrument, (ii) using a reduced-order model for non-spinning waveforms that include a merger-ringdown and higher harmonics, (iii) setting the noise realization to zero and computing overlaps in the amplitude/phase representation.
+We present the first simulations of Bayesian inference for the parameters of supermassive black hole systems including consistently the merger and ringdown of the signal, as well as higher harmonics.
+We find that restricting the merger-dominated signal to its dominant harmonic can make the extrinsic likelihood very degenerate.
+Including higher harmonics proves to be crucial to break degeneracies and considerably improves the localization of the source, with a surviving bimodality in the sky position.
+We also present the first simulations of Bayesian inference for SOBHs, and show that although unimodal, their posterior distributions can have non-Gaussian features.
 
@@
 
-[Introduction]
+%The future gravitational wave detector LISA~\cite{LISA2017}.
+%
+%, the network of ground-based detectors will have expanded from the current LIGO~\cite{} and Virgo~\cite{} detectors to include KAGRA~\cite{} and Indigo, which will have continued improving their sensitivity.
+By the time LISA will fly, a third generation of detectors, like Einstein Telescope~\cite{} and Cosmic Explorer~\cite{}, might be operating.
+%
+%LISA will provide complementary informations.
+By targeting a much lower frequency band, will be sensitive to different sources.
+Among the sources, mergers of SMBHs, seen at high SNR and anywhere in the observable universe, and inspirals of LIGO-type sources (SOBHs), opening the possibility of multi-band gravitational wave astronomy.
+%
+%Much remained to be explored for the data analysis of LISA.
+Different challenges : high SNR for merger-dominated SMBH signals, SNR-limited regime for very long SOBH signals.
+The motion of the detector leaves an imprint on the signal, while the response also has frequency-dependency at high frequencies where the long-wavelength approx is not valid anymore.
+Finally, many other challenges : calibration and understanding noise on the fly, non-stationarity of the noise, dealing with glitches, and most importantly global fit: superposition of long-lived sources in a single datastream.
 
@@
 It allows us to accelerate the likelihood computation, as will be explained in~\ref{}.
-For exploratory parameter estimation runs, this approximation allows us to explore the structure of the likelihood with its possible parameter degeneracies (\SM{[cite past use of this approx.]}), and is a significant improvement over Fisher matrix estimates in that sense.
+For exploratory parameter estimation runs, this approximation allows us to explore the structure of the likelihood with its possible parameter degeneracies (\SM{cite past use of this approx.}), and is a significant improvement over Fisher matrix estimates in that sense.
 We leave the extension of our analysis to include noise realizations for future work.
@@
 	G_{slr}^{\ell m}(f,t) &= \frac{i \pi f L}{2} \sinc \left[ \pi f L\left(1-k\cdot n_{l} \right) \right] \nn\\
-	& \quad \cdot \exp\left[ i \pi f \left( L + k\cdot \left( p_{1} + p_{2} \right) \right) \right] \nn\\
+	& \quad \cdot \exp\left[ i \pi f \left( L + k\cdot \left( p_{r} + p_{s} \right) \right) \right] \nn\\
 	& \quad \cdot n_{l} \cdot P_{\ell m} \cdot n_{l} \,
 \end{align}
-\tdc{what are $p_1$ and $p_2$ here? Are they the same as $p_r$ and $p_s$?}
 is the kernel from~\cite{MB18}, $P_{\ell m}$ is defined by the decomposition~\eqref{eq:Hsummodes}, and
@@
 We also expect intrinsic and extrinsic parameters to correlate weakly, at least in the absence of precession, so that the conclusions drawn from Sec.~\ref{sec:SMBHPE} below, notably the analysis of possible degeneracies in the extrinsic part of the likelihood, should hold for aligned spin systems.
-\tdc{the last sentence would fit well in the conclusion.}
+\TDC{the last sentence would fit well in the conclusion.}
 
@@
 ROMs to accelerate the waveform model of~\cite{Cotesta+18} are under construction, and we plan to include the effects of aligned and misaligned spins in future studies.
-\tdc{this paragraph also sounds like conclusion material.}
+\TDC{this paragraph also sounds like conclusion material.}
 
@@
 In the LISA.
-\tdc{this paragraph seems to need more work.
+\TDC{this paragraph seems to need more work.
 I would start with the existing general case, then say that by using a zero-noise realization we can write the amplitude/phase representation and use a sparse grid and get a large speedup.}
@@
 Note that individual terms in~\eqref{eq:lnLoverlapstructure} can be individually large, and computing $\ln \calL$ relies on accurate cancellations.
-Decomposing further in TDI channels as in~\eqref{} and in harmonics as in~\eqref{}, we have a sum of terms with the structure \tdc{we need one or two more equations describing explicitly how to go from $\tilde{y}_{slr}$ to $\ln \calL$}
+Decomposing further in TDI channels as in~\eqref{} and in harmonics as in~\eqref{}, we have a sum of terms with the structure \TDC{we need one or two more equations describing explicitly how to go from $\tilde{y}_{slr}$ to $\ln \calL$}
 \be
@@
 \ee
-\tdc{should this notation be aligned to Eq 6 and 7?}
+\TDC{should this notation be aligned to Eq 6 and 7?}
 The difficulty of computing such an overlap depends on the phase difference.
@@
 We can then compute the elementary integrals~\eqref{eq:fastlikestructure} using a combination of integrations by parts and asymptotic expansions, as explained in App.~\ref{app:numintegration}.
-The number of grid intervals is typically several hundreds \SM{[COMPLETE]}.
+The number of grid intervals is typically several hundreds \SM{to be completed}.
 
@@
 
-For this exploratory work we take a ``naive'' \tdc{``brute force''?} approach to inferring the posterior probability distribution of the parameters, without applying specific knowledge or expectations about distributions and degeneracies.
+For this exploratory work we take a ``naive'' \TDC{``brute force''?} approach to inferring the posterior probability distribution of the parameters, without applying specific knowledge or expectations about distributions and degeneracies.
 Such assumptions might in fact lead to an incomplete exploration of the posterior probability in ways that would be difficult to recognize and diagnose.
@@
 We applied a general-purpose proposal distribution composed of a weighted set of sub-proposals, including differential-evolution steps~\cite{Vrugt+2008} (although the proposal in this case is based on each single chain, not an ensemble) and a collection of several Gaussian step draws of varying sizes scaled off the prior domain at about 0.01--1\% of its scale and weighted favor the smaller scale proposal draws.
-Effective sample sizes (ESS) were estimated from the number of post-burn-in samples, reduced by the autocorrelation length, with the burn-in size chosen to optimize \tdc{maximize?} the ESS.
+Effective sample sizes (ESS) were estimated from the number of post-burn-in samples, reduced by the autocorrelation length, with the burn-in size chosen to optimize \TDC{maximize?} the ESS.
 In most cases, the runs were continued until achieving an estimated ESS~$>2000$.
@@
 In our runs we used 4000 live points and a stopping evidence threshold of 1.0.
-\jgb{Sylvain needs to check these values.}
+\JGB{Sylvain needs to check these values.}
 
@@
 Generally, finite difference Fisher matrices can be problematic, with small numerical defects potentially having outsized impact, so the finite-difference step size $\epsilon_i$ used in each parameter derivative $\partial_{\theta_i}h$ must be chosen carefully.
-We target the step-size to be scaled off the diagonal Fisher matrix elements by ${\epsilon_i}\approx\delta/\sqrt{F_{ii}}$ using $\delta=0.001$ \jgb{This is \texttt{fisher\_err\_target}, was it different for some runs?}  for the results shown here.
+We target the step-size to be scaled off the diagonal Fisher matrix elements by ${\epsilon_i}\approx\delta/\sqrt{F_{ii}}$ using $\delta=0.001$ \JGB{This is \texttt{fisher\_err\_target}, was it different for some runs?}  for the results shown here.
 To achieve this we set begin with an initial choice of the $\epsilon_i$ scaled of parameter values or prior widths, then we estimate the Fisher diagonals and iterate until convergence.
 In this process we also constrain that the step is never more than $10^{-10}$ times the initial scaling.
-\jgb{There are some more special-case details, but probably not worth elaborating.
+\JGB{There are some more special-case details, but probably not worth elaborating.
 We could also discuss how we do matrix inversion, but I don't think there is anything interesting there.}
@@
 We found indeed that these parameters correlate only weakly to the other ones, so that we can hope to understand the degeneracy features by restricting the analysis to the six extrinsic parameters $(D_{L}, \iota, \varphi, \lambdaL, \betaL, \psiL)$.
-\tdc{we could make the connection with a similar decoupling in LIGO/Virgo, which is actually what makes low-latency sky localization possible.}
+\TDC{we could make the connection with a similar decoupling in LIGO/Virgo, which is actually what makes low-latency sky localization possible.}
 
@@
 \ee
-From the low-frequency response \eqref{eq:aelowfmodes}, pinning the time-at-LISA~\eqref{} \SM{[COMPLETE $t_{L}$]} so that we can ignore the factor $\exp [2 i \pi f k\cdot p_{0}]$ as a pure constant corresponding to a redefinition of time, we obtain \jgb{Should we use notation $\Phi_R$ here and in \eqref{eq:aelowfmodes}?}
+From the low-frequency response \eqref{eq:aelowfmodes}, pinning the time-at-LISA~\eqref{} \SM{to be completed for $t_{L}$} so that we can ignore the factor $\exp [2 i \pi f k\cdot p_{0}]$ as a pure constant corresponding to a redefinition of time, we obtain \JGB{Should we use notation $\Phi_R$ here and in \eqref{eq:aelowfmodes}?}
 \begin{align}\label{eq:simplelikelihoodhm}
@@
 \end{align}
-where we introduced the following extrinsic signal functions \jgb{Since neither $D_{a,e}$ nor $P_{\ell m}$ are defined in the main text, may be better to write in terms of a version of $F$ as in \eqref{eq:defFapcFepc}}
+where we introduced the following extrinsic signal functions \JGB{Since neither $D_{a,e}$ nor $P_{\ell m}$ are defined in the main text, may be better to write in terms of a version of $F$ as in \eqref{eq:defFapcFepc}}
 \be\label{eq:defsaelm}
@@
 We will further abbreviate notations by using the variables
-\tdc{it is unclear if these mean $\tan (\iota/2)$ or $(\tan \iota)/2$}
-\be
-	t_{\iota} \equiv \tan \iota/2 \,, \quad t_{\theta} \equiv \tan \theta_{L} / 2 \,.
+\TDC{it is unclear if these mean $\tan (\iota/2)$ or $(\tan \iota)/2$}
+\be
+	t_{\iota} \equiv \tan \frac{\iota}{2} \,, \quad t_{\theta} \equiv \tan \frac{\theta_{L}}{2} \,.
 \ee
@@
 	\end{tabular}
-	\caption{Parameters of the simulated SOBH mergers. \tdc{Maybe we can show
+	\caption{Parameters of the simulated SOBH mergers. \TDC{Maybe we can show
 			 these on the upper right of each corner plot instead.}}
@@
 Together with the luminosity distance $D_{L}$, these parameters appear in combinations of constant factors multiplying the whole response.
-\jgb{Would it be better to include the angle dependence explicitly here in the $\sYlm$ here for clarity, since the point is to account for parameter dependencies.}
+\JGB{Would it be better to include the angle dependence explicitly here in the $\sYlm$ here for clarity, since the point is to account for parameter dependencies.}
 
@@
 \ee
-Several assumptions enter this form of the response: \dots \SM{[COMPLETE]} \jgb{Define $n_l$. Also double dot looks like colon, needing some clarification.}.
+Several assumptions enter this form of the response: \dots \SM{[COMPLETE]} \JGB{Define $n_l$. Also double dot looks like colon, needing some clarification.}.
 
@@
 We can recover pattern functions (for $\psi_{L}=0$) by factoring out the effective length, we define
-\jgb{Would it be better to just include a new version of expression \eqref{eq:aelowfmodes} here using $F_{a,e}$, such as:
+\JGB{Would it be better to just include a new version of expression \eqref{eq:aelowfmodes} here using $F_{a,e}$, such as:
   $\tilde{a}, \tilde{e} = (-3i\pi f L)\left[ \tilde{h}_{\ell m}^\times F_{a,e}^+ + \tilde{h}_{\ell m}^\times F_{a,e}^\times\right]$.

### 2019-05-10 paper/paper_PE.tex
Initial notes about intro

@@
 
+Gravitational waves are awesome.
+
+Literature review (can base on GOAT study).
+
+Main points:
+\begin{itemize}
+    \item Fast likelihood calculation of mergers, including ROM model with FD LISA response and techniques for fast overlaps.
+    \item Example cases of MBH near sweet spot and SOBH.
+    \item Compare Bayesian and Fisher.
+    \item Study effect of higher modes for MBH.
+    \item Evolution of information over time to merger.
+    \item Analytic intuitive explanation of complicated degeneracies.
+\end{itemize}
+
 %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
@@
 \esub
-with a common prefactor \jgb{This seems to need to  be multiplied by -6 to get what I have in my notes. Trace to change in def of $s_{a,e}$?}
+with a common prefactor \JGB{This seems to need to  be multiplied by -6 to get what I have in my notes. Trace to change in def of $s_{a,e}$?}
 \be\label{eq:sigmafactorrho}
@@
 
-\jgb{Don't have to resort to limits.}
+\JGB{Don't have to resort to limits.}
 \sout{We can now explain the degenerate features we found as certain limits in this simplified likelihood.
@@
 %\esub
-\jgb{Keep something from this?}
+\JGB{Keep something from this?}
 The other condition~\eqref{eq:conditionprefactor} gives then a single phase constraint for $\varphi - \psiL$, and a single amplitude constraint for $\rho(d, \iota, \theta_{L}^{*})$.
@@
 %\ee
-\jgb{Keep something from this?}
+\JGB{Keep something from this?}
 Thus, we find again a degenerate pattern, with the difference that \eqref{eq:soldegeniotaphi} imposes a condition on the pair $\lambda_{a} - \psiL$ instead of $\lambda_{a}$, and the extra condition imposes a condition on $\varphi$ instead of the pair $\varphi - \psiL$.

### 2019-05-10 paper/paper_PE.tex
fixes

@@
 
-Gravitational waves are awesome.
-
-Literature review (can base on GOAT study).
-
-Main points:
-\begin{itemize}
-    \item Fast likelihood calculation of mergers, including ROM model with FD LISA response and techniques for fast overlaps.
-    \item Example cases of MBH near sweet spot and SOBH.
-    \item Compare Bayesian and Fisher.
-    \item Study effect of higher modes for MBH.
-    \item Evolution of information over time to merger.
-    \item Analytic intuitive explanation of complicated degeneracies.
-\end{itemize}
-
 %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

### 2019-05-10 paper/paper_PE.tex
More intro notes, remove outdated note
Work on intro

@@
 
+Gravitational waves from the coalescence of black hole binaries are routinely observed \cite{GWTC-1} by the ground-based interferometers Advanced LIGO \cite{LIGO}, Advanced Virgo \cite{Virgo} and, soon, KAGRA \cite{KAGRA}.
+
+Ground-based interferometers are fundamentally limited at low frequency by terrestrial noise, and cannot be used to study mergers of compact objects much heavier than $\approx 10^3 \Msol$.
+The LISA mission \cite{LISA2017} is designed to overcome this limitation, enabling the observation and precise characterization of mergers of black holes ranging in mass from tens of $\Msol$ (stellar-origin black holes, SOBH) to millions of $\Msol$ (supermassive black holes, SMBH).
+Black hole mergers, in fact, will be a major class of sources in the LISA frequency band.
+
+In order to realize the full scientific objective of LISA with respect to black hole mergers, adequate data analysis tools must be prepared in advance.
+Similarly to what happens in LIGO and Virgo, one has to first identify the presence of a merger waveform in the LISA data, and then infer the distribution of the physical parameters of its source from the data, enabling the construction of a catalog of black hole mergers.
+Inferring the parameters of binaries, in particular their distance and sky location, will also be necessary to organize multimessenger observations of their late inspirals and mergers \cite{TAP-LISA} and for multiband gravitational-wave astronomy \cite{Sesana16}.
+In this paper we focus on the latter part of the analysis problem, i.e.~the inference of the black hole parameters from the LISA data, once we have reasons to believe the presence of the signal in the data.
+
+The inference problem amounts to computing a likelihood function, i.e.~the probability of observing the measured data given a model of the source and detector, and a particular parameter vector for the model.
+In LISA, this problem is more challenging than in kilometer-scale ground-based interferometers.
+Depending on the masses and spins of the black holes, the complexity arises from the much longer duration of the waveform, the much larger expected signal-to-noise ratios (SNRs), or the much more complex response of the LISA constellation to a given waveform.
+The large number of cycles in the LISA band and the large SNR require very accurate modeling of the waveform, including the merger and ringdown regimes, the effects of spin and corrections of higher order than the quadrupole.
+Contrary to LIGO and Virgo, which are approximately fixed in an inertial frame during the observation of any merger, the LISA constellation will move around the Sun and change its orientation appreciably during the observation of a large fraction of the inspirals, leading to additional modulations of the signals.
+Furthermore, due to the LISA arm length, for many sources the long-wavelength approximation will not hold.
+As we will see in this study, a proper treatment of these effects is necessary in order to break severe degeneracies between the source parameters.
+
+\TDC{Add literature review on BBH PE here, can base on GOAT study.}
+
+Here we demonstrate that standard Bayesian inference can be performed for SMBHs and SOBHs using a self-consistent waveform model that includes inspiral, merger, ringdown and higher-order modes (albeit neglecting spins) and a full model of the LISA response.
+We start by introducing a fast method to calculate the likelihood for a black hole merger.
+The method includes: a fast computation of a Fourier-domain inspiral-merger-ringdown waveform with higher-order modes based on reduced-order modeling; a fast Fourier-domain computation of the LISA response based on \cite{MB18}; and a fast method for computing the noise-weighted product between waveforms.
+We couple our likelihood with two codes for Bayesian inference based on different sampling techniques, constructing an inference engine which allows us to perform a variety of investigations with simulated black hole mergers.
+Both samplers are first shown to converge to comparable posterior distributions of the source parameters for a few case studies, and we highlight the challenges we encounter in the sampling.
+The full posterior distributions are then compared with their Fisher approximations, widely used in gravitational-wave analysis.
+Focusing on massive black holes, we show next the effect of higher-order modes on the degeneracies between some of the parameters.
+We derive an analytic and intuitive explanation of the origin of such degeneracies via various approximations.
+We show examples of how the posterior distributions evolve over time as more and more data becomes available for the inference.
+Finally, we repeat some of our investigations for stellar-origin black hole mergers as well.
+
 %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
@@
 We will further abbreviate notations by using the variables
-\TDC{it is unclear if these mean $\tan (\iota/2)$ or $(\tan \iota)/2$}
 \be

### 2019-05-21 paper/paper_PE.tex
Add note about LISA trajectories
Draft of time accumulation

@@
   \centering
+  \begin{minipage}{.48\linewidth}
+      \includegraphics[width=.99\linewidth]{../plots/t-of-F.pdf}
+   \end{minipage}
+   \begin{minipage}{.48\linewidth}
+      \includegraphics[width=.99\linewidth]{../plots/t-of-SNR.pdf}
+   \end{minipage}
+  \caption{Relations of time, Fourier frequency, and accumulated SNR for $M=2\times 10^6$,$q=3$ merger at $z=4$.
+The left panel shows times computed from $\tilde h(f)$ using Eq.~eref{eq:t-of-f}, and are approximately common among the waveform component modes up to minutes before merger.
+We use this relation to simulate time-cuts of the signals a various points before merger.
+The right panel shows how SNR accumulates with time, though the full signal has a high SNR of about 666.0, most SNR accumulates very late.
+The curves shown extend from approximately SNR=1 until $t(f)$ is 40s after nominal merger.
+The stars show points where the SNR has reached $\{1/4,1/16,1/64\}$ of the full-signal total, where we have studied parameter estimate for temporally incomplete signals.}
+  \label{fig:t-f-snr-Case9}
+\end{figure*}
+
+The rate at which parameter information accumulates LISA merger observations is crucial for planning strategies for multi-messenger observations and in guiding the latency demands of the LISA down-link and data processing.
+We explore how parameter information accumulates on approach to merger by performing parameter estimation studies temporally truncated signals, cut at points where the accumulated SNR is about $\{10,42,166\}$ corresponding to $\{1/64,1/16,1/4\}$ of the total SNR=$666.0$, and times $\{41 \mathrm{hr}, 2.5 \mathrm{hr}, 7 \mathrm{min}\}$ before merger.
+
+We implement these temporal cuts in the Fourier domain, using Eq.XXXX, the $t(f)$ relation inferred from the smoothly chirping harmonic mode components of the waveforms as part of the LISA response.
+This is a good approximation for temporal cuts, and consistent among the various modes for our cuts, which are sufficiently early before merger, as shown in Fig.~\ref{fig:t-f-snr-Case9}.
+
+The right panel of the figure also shows the times before merger at which a given value of SNR is reached.
+In this case the signal begins being relevant, reaching SNR~1 about 34 days before merger, but reaches a plausible first-detectability threshold about 41 hours before merger.
+Most of the SNR accumulates over the last few minutes of signal.
+
+In Fig.~\ref{fig:smbhCornerZoomHMCase9} and Fig.~\ref{fig:MollweidesmbhTserieshmCase9} we show how parameter information accumulates at these time points.
+Individual masses are within a factor of 2 at first detection and better than $10\%$ by 41 hours before merger.
+Distance within a factor of 2 at 41 hours and within roughly $10\%$ at 2.5 hours prior to merger.
+Merger time is estimated to about an hour at first detection, and within a few minutes by 2.5 hours before merger.
+Sky position...
+
+\begin{figure*}
+  \centering
   \begin{minipage}{.32\linewidth}

### 2019-11-15 paper/paper_PE.tex
Work on the paper

@@
 \newcommand{\hatk}{\hat{k}}
+\newcommand{\SNR}{\,\mathrm{SNR}}
 \newcommand{\Hz}{\,\mathrm{Hz}}
+\newcommand{\mHz}{\,\mathrm{mHz}}
+\newcommand{\au}{\,\mathrm{au}}
 \newcommand{\sinc}{\,\mathrm{sinc}}
@@
 
-The inference problem amounts to computing a likelihood function, i.e.~the probability of observing the measured data given a model of the source and detector, and a particular parameter vector for the model.
-In LISA, this problem is more challenging than in kilometer-scale ground-based interferometers.
-Depending on the masses and spins of the black holes, the complexity arises from the much longer duration of the waveform, the much larger expected signal-to-noise ratios (SNRs), or the much more complex response of the LISA constellation to a given waveform.
-The large number of cycles in the LISA band and the large SNR require very accurate modeling of the waveform, including the merger and ringdown regimes, the effects of spin and corrections of higher order than the quadrupole.
+The inference problem amounts to produce samples from the posterior distribution for sourcee parameters, related by Baye's theroem to the likelihood function, i.e.~the probability of observing the measured data given the source parameters and a model of the source and detector.
+In LISA, this problem is more complex than in kilometer-scale ground-based interferometers.
+Depending on the masses and spins of the black holes, additional challenges arises from the much larger expected signal-to-noise ratios (SNRs) for high-mass systems, or the much longer duration of the waveform for low-mass systems.
+For SMBHBs, the large SNR will require a very accurate modeling of the waveform, including the merger and ringdown regimes, the effects of spin and corrections of higher harmonics in the signal beyond the quadrupole.
+
+The response of the LISA instrument to a given waveform shows a time- and frequency-dependence.
 Contrary to LIGO and Virgo, which are approximately fixed in an inertial frame during the observation of any merger, the LISA constellation will move around the Sun and change its orientation appreciably during the observation of a large fraction of the inspirals, leading to additional modulations of the signals.
-Furthermore, due to the LISA arm length, for many sources the long-wavelength approximation will not hold.
-As we will see in this study, a proper treatment of these effects is necessary in order to break severe degeneracies between the source parameters.
+Furthermore, due to the LISA arm length, for many sources the long-wavelength approximation will not hold, which introduces a frequency-dependence in the response.
+As we will see in this study, a proper treatment of these effects is necessary in order to understand degeneracies between the source parameters.
+
+Previous parameter recovery studies with LISA
 
@@
 Both samplers are first shown to converge to comparable posterior distributions of the source parameters for a few case studies, and we highlight the challenges we encounter in the sampling.
-The full posterior distributions are then compared with their Fisher approximations, widely used in gravitational-wave analysis.
+The full posterior distributions are then compared with their Fisher approximations, widely used in gravitational-wave analyses.
 Focusing on massive black holes, we show next the effect of higher-order modes on the degeneracies between some of the parameters.
@@
 
+\section{LISA instrument response}
+\label{sec:response}
+
+\SM{add mini-intro}
+
+We will use units with $G=c=1$.
+
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+
+\subsection{The GW signal}
+\label{sec:gwsignal}
+
+To describe the gravitational wave signal, we first need to define a conventional source frame associated to the binary system, as detailed in App.~\ref{app:conventions}.
+The gravitational wave propagation vector $k$, pointing from the source to the observer, has polar angles $(\iota, \varphi)$ in this source frame.
+Next, one introduces polarization vectors $p$, $q$ so that $(p, q, k)$ is a direct triad.
+The precise choice of $(p,q)$ is a matter of convention, for which we refer to App.~\ref{app:conventions}.
+
+The gravitational waveform in the transverse-traceless gauge is described by the two polarizations $h_{+}$, $h_{\times}$.
+If $H_{ij} = h_{ij}^{\rm TT}$ represents the gravitational wave signal in matrix form,
+\be
+	H = h_{+} P_{+} + h_{\times} P_{\times}
+\ee
+with the polarization tensors
+\bsub
+\begin{align}
+	P_{+} &= p \otimes p - q \otimes q \,,\\
+	P_{\times} &= p \otimes q + q \otimes p \,.
+\end{align}
+\esub
+Conversely, the polarizations are
+\bsub
+\begin{align}
+	h_{+} &= \frac{1}{2} \left(p \otimes p - q \otimes q \right) : H \,,\\
+	h_{\times} &= \frac{1}{2} \left( p \otimes q + q \otimes p \right) : H \,,
+\end{align}
+\esub
+with the notation $A : B = A_{ij}B_{ij}$.
+
+One can further decompose the gravitational wave signal, seen as a function of the direction of emission $(\iota, \varphi)$ in the source frame, in spin-weighted spherical harmonics as
+\be\label{eq:apphpcmodes}
+	h_{+} - i h_{\times} = \sum_{\ell \geq 2} \sum_{m = -\ell}^{\ell} {}_{-2}Y_{\ell m} (\iota, \varphi) h_{\ell m} \,,
+\ee
+where the explicit expression of the ${}_{-2}Y_{\ell m}$ is given in~\cite{}.
+In the following, we will use exclusively this parametrization of the waveform as a set of modes $h_{\ell m}$.
+The dominant harmonic is $h_{22}$, while the others are called higher modes (HM) or higher harmonics.
+
+We will both generate the waveforms and apply the response directly in the Fourier domain. Our convention for the Fourier transform of a function $F$ is
+\be\label{eq:defFourier}
+	\tilde{F}(f) = \int dt \; e^{2 i \pi f t} F(t) \,.
+\ee
+Note that this differs by a change $f\rightarrow -f$ from the more usual convention used e.g. in \texttt{LAL}~\cite{}.
+
+The definition of the reference time and phase \SM{[\dots COMPLETE]}
+
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+
+\subsection{Mode decomposition and polarization angle}
+\label{sec:modespol}
+
+We now translate the mode decomposition~\eqref{eq:hpcmodes} in the Fourier domain. We have
+\bsub
+\begin{align}
+	h_{+} = \frac{1}{2} \sum_{\ell, m} \left( \sYlm h_{\ell m} + \sYlmstar h_{\ell m}^{*} \right) \,,\\
+	h_{\times} = \frac{i}{2} \sum_{\ell, m} \left( \sYlm h_{\ell m} - \sYlmstar h_{\ell m}^{*} \right) \,,
+\end{align}
+\esub
+which is valid in general (we dropped the $(\iota, \varphi)$ arguments of the $\sYlm$). Now, for non-precessing binary systems, an exact symmetry relation between modes reads
+\be\label{eq:nonprecsymmetry}
+	h_{\ell, -m} = (-1)^{\ell} h_{\ell, m}^{*} \,.
+\ee
+Using this symmetry, we can write
+\be
+	h_{+,\times} = \sum_{\ell, m} K_{\ell m}^{+, \times} h_{\ell m} \,,
+\ee
+with
+\bsub
+\begin{align}
+	K_{\ell m}^{+} =\frac{1}{2} \left( \sYlm + (-1)^{\ell} \sYlminusmstar \right) \,,\\
+	K_{\ell m}^{\times} = \frac{i}{2} \left( \sYlm - (-1)^{\ell} \sYlminusmstar \right) \,.
+\end{align}
+\esub
+
+Going to the Fourier domain, an approximation often used is to neglect support for negative/positive frequencies according to
+\be\label{eq:approxnegf}
+	\tilde{h}_{\ell m} (f) \simeq 0 \;\; \text{for} \;\; m<0, \; f>0 \;\; ( m>0, \; f<0 )\,,
+\ee
+and neglecting modes $h_{\ell 0}$.
+We will use this approximation throughout this paper.
+Note that we picked our Fourier convention~\eqref{eq:defFourier} to ensure that modes $\tilde{h}_{\ell m} (f)$ with $m>0$ have support for $f>0$.
+Using~\eqref{eq:approxnegf}, for $f>0$ we have
+\be
+	\tilde{h}_{+,\times} = \sum_{\ell} \sum_{m>0} K_{\ell m}^{+, \times} \tilde{h}_{\ell m} \,.
+\ee
+
+Next, it is convenient to introduce mode-by-mode polarization matrices
+\be
+	P_{\ell m} = P_{+} K_{\ell m}^{+} + P_{\times} K_{\ell m}^{\times} \,,
+\ee
+so that
+\be\label{eq:Hsummodes}
+	H = \sum_{\ell, m} P_{\ell m} h_{\ell m} \,.
+\ee
+
+The polarization angle $\psi$ can be seen as detailed in App~\ref{app:conventions} as a degree of freedom in the relation between the source frame and the detector frame, parametrizing a rotation around the wave vector $k$.
+To define this angle, one introduces reference vectors $(u,v)$ orthogonal to $k$, that define the zero of the polarization angle as $(p,q)(\psi=0) = (u,v)$.
+Our convention for $(u,v)$ is detailed in App~\ref{app:conventions}.
+
+To make explicit the dependence in polarization, we can define polarization tensors for 0 polarization angle as
+\bsub
+\begin{align}
+	P_{+}^{0} &= P_{+}(\psi = 0) = u \otimes u - v \otimes v \,,\\
+	P_{\times}^{0} &= P_{\times}(\psi = 0) = u \otimes v + v \otimes u \,.
+\end{align}
+\esub
+This allows to write the dependence in polarization as
+\be
+	P_{+} + i P_{\times} = e^{-2 i \psi} \left( P_{+}^{0} + i P_{\times}^{0} \right) \,,
+\ee
+or explicitly in $P_{\ell m}$ as
+\begin{align}
+	P_{\ell m} (\iota, \varphi, \psi) &= \frac{1}{2} \sYlm(\iota, \varphi) e^{-2 i \psi} \left( P_{+}^{0} + i P_{\times}^{0} \right) \nn\\
+	& + \frac{1}{2} (-1)^{\ell} \sYlminusmstar (\iota, \varphi) e^{+2 i \psi} \left( P_{+}^{0} - i P_{\times}^{0} \right) \,.
+\end{align}
+Writing the $P_{\ell m}$ matrices in this way allows to factor out explicitly all dependencies in the extrinsic parameters $(\iota, \phi, \psi)$.
+Together with the luminosity distance $D_{L}$ scaling the overall amplitude of the signal, these parameters enter as constant (time and frequency independent) prefactors in the response for each mode $h_{\ell m}$.
+
+Since only $2\psi$ intervenes, the polarization $\psi$ has an exact $\pi$-degeneracy; by convention we set $\psi \in [0,\pi]$.
+
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+
+\subsection{Frequency domain LISA response}
+\label{subsec:FDresponse}
+
+The LISA response can be built from single-link observables $y_{slr} = (\nu_{r} - \nu_{s})/\nu$ respresenting a laser frequency shift between the sending spacecraft $s$ and the receiving spacecraft $r$ along the link $l$.
+We use the expression~\cite{}
+\be\label{eq:defyslr}
+	y_{slr} = \frac{1}{2} \frac{n_{l} \otimes n_{l}}{1 - k\cdot n_{l}} : \left[ H(t - L - k\cdot p_{s}) - H(t - k\cdot p_{r}) \right] \,,
+\ee
+with $H$ the transverse-traceless matrix representing the gravitational wave, $L$ the delay along one arm\footnote{We use units with $G=c=1$.}, taken to be fixed, $k$ the propagation vector, $n_{l}$ the link unit vectors (from $s$ to $r$) and $p_{s}$, $p_{r}$ the positions of the spacecrafts.
+Here $n_{l}$, $p_{s}$ and $p_{r}$ are evaluated at the same time $t$.
+
+In the frequency domain, modes will be represented with their amplitude and phase as\footnote{Note our Fourier convention~\eqref{eq:defFourier}.}
+\be\label{eq:hlmampphase}
+	\tilde{h}_{\ell m} (f) = A_{\ell m} (f) e^{-i\Psi_{\ell m} (f)}\,.
+\ee
+We will use the analysis of~\cite{MB18} and write the response in individual observables $y_{slr}$ with a transfer function for each spherical harmonic mode as
+\be
+	\tilde{y}_{slr} = \sum_{\ell, m}\calT_{slr}^{\ell m}(f) \tilde{h}_{\ell m} \,.
+\ee
+Applying the perturbative formalism of~\cite{MB18} to leading order, we have simply
+\be\label{eq:Tlmslr}
+	\calT_{slr}^{\ell m}(f) = G_{slr}^{\ell m}(f, t_{f}^{\ell m}) \,,
+\ee
+where
+\begin{align}\label{eq:Gslr}
+	G_{slr}^{\ell m}(f,t) &= \frac{i \pi f L}{2} \sinc \left[ \pi f L\left(1-k\cdot n_{l} \right) \right] \nn\\
+	& \quad \cdot \exp\left[ i \pi f \left( L + k\cdot \left( p_{r} + p_{s} \right) \right) \right] \; n_{l} \cdot P_{\ell m} \cdot n_{l} \,,
+\end{align}
+is the kernel from~\cite{MB18} with $n_{l}$, $p_{s}$, $p_{r}$ evaluated at $t$, with $P_{\ell m}$ defined by the decomposition~\eqref{eq:Hsummodes}. In~\eqref{eq:Tlmslr},
+\be\label{eq:deftflm}
+	t_{f}^{\ell m} = -\frac{1}{2\pi} \frac{d\Psi_{\ell m}}{df}
+\ee
+is the effective frequency$\to$time correspondence, defined across the whole frequency band and including the merger and ringdown.
+
+The analysis of~\cite{MB18} has shown that higher-order corrections are small for SMBH systems, as well as for SOBH systems provided they are close enough to merger as will be the case for the examples chosen here.
+The fact that we use the same Fourier-domain treatment of the transfer functions for both the signal and the templates should also mitigate the importance of modelling errors.
+We will therefore limit ourselves to the leading order in the treatment of~\cite{MB18} in the rest of this paper.
+
+We note an important feature of~\eqref{eq:Gslr}: frequency-dependent terms only feature the projections
+\be
+	k_{\parallel} \cdot n_{l}, \; k_{\parallel} \cdot p_{s,r} \,, 
+\ee
+with $k_{\parallel}$ the projection of the wave vector $k$ in the (instantaneous) LISA plane, thus the frequency-dependent factors are invariant when we reflect $k$ across this plane.
+This will remain true when constructing TDI variables below\footnote{Strictly speaking it will not be true in the most general response formalism of~\cite{MB18}, where corrections are either slightly non-local in time or involve velocities, that are out of the LISA plane.}.
+We will investigate sky degeneracies in detail in Sections~\ref{subsec:MBHBPEdegen} and~\ref{subsec:MBHBacctimebreakdegen}.
+
+In~\eqref{eq:defyslr} and~\eqref{eq:Gslr}, one can distinguish two types of delays: on one hand the delay associated to the position $p_{0}$ of the center of the constellation on its oarbit around the Sun, with baseline $R=1 \au$, and on the other hand the delays associated to the individual spacecraft positions in the constellation, with baseline $L=2.5 \cdot 10^{6 }\mathrm{km}$.
+This defines the transfer frequencies
+\bsub\label{eq:transferfrequencies}
+\begin{align}
+	f_{R} &= 1/R = 2.0\times10^{-3}\Hz \,,\\
+	f_{L} &= 1/L = 0.12\Hz \,.
+\end{align}
+\esub
+The transfer frequency $f_{L}$ correspond to fitting a full wavelength in the armlength; as we will see below in Sec.~\ref{}, the long-wavelength approximation starts to break down at significantly lower frequencies.
+
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+
+\subsection{Time Delay Interferometry}
+\label{sec:TDIresponse}
+
+To mitigate laser noise, LISA requires Time Delay Interferometry (TDI)~\cite{}.
+These TDI observables are built from delayed combinations of the one-arm basic observables~\eqref{eq:defyslr}.
+Detailed cancellations for a flexing and rotating constellation prompted the development of higher-generation TDI schemes~\cite{}.
+However, these refinements affect only marginally the response to the gravitational waves.
+We therefore limit ourselves to first-generation TDI and adopt a rigid approximation for the cancellation, where delays are all constant and set to $L$.
+Using the notation $y_{slr,L} = y_{slr}(t - L)$, the first-generation TDI Michelson observable $X$ reads~\cite{}
+\begin{align}
+	X &= y_{31} + y_{13,L} + \left( y_{21} + y_{12,L} \right)_{,2L} \nn\\
+	& - \left( y_{21} + y_{12,L} \right) - \left( y_{31} + y_{13,L} \right)_{,2L} \,,
+\end{align}
+with the other Michelson observables $Y$, $Z$ being obtained by cyclic permutation. Uncorrelated combinations $A$, $E$ and $T$ are then defined as
+\bsub
+\begin{align}
+	A &= \frac{1}{\sqrt{2}} \left( Z - X \right) \,,\\
+	E &= \frac{1}{\sqrt{6}} \left( X - 2Y + Z \right) \,,\\
+	T &= \frac{1}{\sqrt{3}} \left( X + Y + Z \right) \,.
+\end{align}
+\esub
+These channels are independent under the assumption of an identical and uncorrrelated \SM{check} noise in the detector arms.
+With constant delays in the rigid approximation, with the notation $z\equiv \exp[2i\pi fL]$ the TDI combinations take the form
+\bsub\label{eq:aet}
+\begin{align}
+	\tilde{a} &= \left[ (1+z) \left( \tilde{y}_{31} + \tilde{y}_{13} \right) - \tilde{y}_{23} - z \tilde{y}_{32} - \tilde{y}_{21} - z \tilde{y}_{12} \right] \,,\\
+	\tilde{e} &= \frac{1}{\sqrt{3}} \left[ (1-z)\left( \tilde{y}_{13} - \tilde{y}_{31} \right) + (2+z) \left( \tilde{y}_{12} - \tilde{y}_{32} \right) \right. \nn\\
+	&\qquad\quad \left. + (1+2z) \left( \tilde{y}_{21} - \tilde{y}_{23} \right) \right] \,,\\
+	\tilde{t} &= \frac{\sqrt{2}}{\sqrt{3}} \left[ \tilde{y}_{21} - \tilde{y}_{12} + \tilde{y}_{32} - \tilde{y}_{23} + \tilde{y}_{13} - \tilde{y}_{31} \right] \,,
+\end{align}
+\esub
+where we have eliminated frequency-dependent prefactors that are common to the signal and to the noise by introducing the rescalings \SM{for $\tilde{a}$, $\tilde{e}$ is it $1/z^{2}$ ?}
+\bsub\label{eq:scalingAET}
+\begin{align}
+	\tilde{a}, \tilde{e} &= \frac{1}{i \sqrt{2} z \sin (2\pi f L)}\times \tilde{A}, \tilde{E} \,,\\
+	\tilde{t} &= \frac{1}{2\sqrt{2} z^{3} \sin (\pi f L) \sin (2\pi f L)} \times \tilde{T} \,.
+\end{align}
+\esub
+We will use mode-by-mode transfer functions for these reduced channels as 
+\be\label{eq:defTlmaet}
+	\tilde{a}, \tilde{e}, \tilde{t} = \sum_{\ell, m} \calT_{a,e,t}^{\ell m} \tilde{h}_{\ell m}\,.
+\ee
+To make the connection between these observables and the familiar strain observed by ground-based intruments (see Sec.~\ref{} detailing the low-frequency approximation), we also introduce the notation
+\be\label{eq:defhaet}
+	\tilde{h}_{a,e,t} \equiv \frac{1}{-6i \pi f L} \times \tilde{a}, \tilde{e}, \tilde{t} \,,
+\ee
+and similarly the transfer functions $\calT_{h_{a}, h_{e}, h_{t}}^{\ell m} = \calT_{a,e,t}^{\ell m} / (-6 i \pi f L)$.
+
+Scaling out the same square factors from the noise power spectral density as
+\bsub
+\begin{align}
+	S_{n}^{A}, S_{n}^{E} &= 2 \sin^{2} (2\pi f L) \times S_{n}^{a}, S_{n}^{e} \,,\\
+	S_{n}^{T} &= 8 \sin^{2} (\pi f L) \sin^{2} (2\pi f L) S_{n}^{t} \,,
+\end{align}
+\esub
+the reduced PSD for the three channels take the form
+\bsub\label{eq:Snaet}
+\begin{align}
+	S_{n}^{a} = S_{n}^{e} &= 2 \left( 3 + 2\cos (2\pi f L) + \cos (4 \pi f L) \right) S^{\rm pm}(f) \nn\\
+	& \quad + \left( 2 + \cos (2\pi f L) \right) S^{\rm op}(f) \,,\\
+	S_{n}^{t} &= 4 \sin^{2} (2\pi f L) S^{\rm pm}(f) + S^{\rm op}(f) \,,
+\end{align}
+\esub
+with $S^{\rm pm}$ the test-mass noise PSD and $S^{\rm op}$ the optical noise PSD, described in~\ref{} \SM{complete}.
+Note that the confusion noise coming from the background of galactic binaries in the  LISA band is to be added to the point-mass and optical noises.
+We can define strain-like noise PSD associated to the strain-like TDI observables~\eqref{eq:defhaet} as
+\be\label{eq:defShaet}
+	S_{h}^{a,e,t} (f) = \frac{S_{n}^{a,e,t}(f)}{(6 \pi f L)^{2}}\,.
+\ee
+
+The prefactors~\eqref{eq:scalingAET} are oscillatory and have zero-crossings at high frequencies, with the first one occuring at $f_{L}/2 = 0.06 \mathrm{Hz}$, which is why it is convenient to eliminate them to avoid $0/0$ numerical instabilities.
+However, scaling these factors out from both the signal and the noise is an approximation, as we can expect that the real delays would lead to imperfect cancellations in the vicinity of the zero-crossings.
+
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+
+\subsection{The low-frequency limit}
+\label{subsec:lowfresponse}
+
+As is well known~\cite{Cutler97}, in the low-frequency limit (also called the long-wavelength approximation), the finite-length effects of the arms of LISA disappear, and the response of the detectors falls back to the one of two LIGO-type detectors set in motion, rotated from each other by $\pi/4$.
+
+For $f \ll f_{L}$, we have $2\pi f L \ll 1$ and the kernel~\eqref{eq:Gslr} reduces to
+\be\label{eq:Gslrlowf}
+	G^{\ell m}_{slr} \simeq \frac{i \pi f L}{2} \exp\left[ 2 i \pi f k\cdot p_{0} \right] n_{l} \otimes n_{l} : P_{\ell m}\,.
+\ee
+For even lower frequencies, when $f \ll f_{R}$ as well, we have $2\pi f R \ll 1$ and the delay term is negligible: $\exp\left[ 2 i \pi f k\cdot p_{0} \right] \simeq 1$.
+It can be convenient to introduce the following notations for this delay phase:
+\be\label{eq:defPhiR}
+	\Phi_{R} \equiv 2 \pi f k\cdot p_{0} \,, \quad \Delta \Phi_{R} = \Phi_{R} - \Phi_{R}(t=t_{f}^{\rm peak}) \,.
+\ee
+The quantity $\Phi_{R}(\tf)$ is often the Doppler phase in the literature; it may appear to be large, but in the limit of short-lived signals it corresponds to the fixed delay between the time of arrival at the SSB and the time of arrival at the LISA constellation.
+Only the variations of the Doppler phase carry useful information about the sky position.
+The phase difference $\Delta \Phi_{R}$ subtracts this constant time delay contribution \SM{[complete for $t_{L}$]}, which is equivalent to parametrizing the signal using the time variable $t_{L}$ as defined in~\eqref{}.
+
+For $2\pi f L \ll 1$, we have $z\simeq 1$, and the link reversal symmetry $\tilde{y}_{slr} \simeq \tilde{y}_{r-ls}$, so that~\eqref{eq:aet} becomes (dropping $l$ indices and symmetrizing $r,s$)
+\bsub\label{eq:aetlowf}
+\begin{align}
+	\tilde{a} &\simeq 4\tilde{y}_{31} - 2 \tilde{y}_{23} - 2 \tilde{y}_{12} \,,\\
+	\tilde{e} &\simeq 2\sqrt{3} \left[ \tilde{y}_{12}  - \tilde{y}_{23} \right] \,,\\
+	\tilde{t} &\simeq 0 \,,
+\end{align}
+\esub
+with the $T$-channel becoming negligible in this limit. Using~\eqref{eq:Gslrlowf}, we can write
+\be\label{eq:aelowfmodes}
+	\tilde{a}, \tilde{e} = \sum_{\ell, m>0} (-2i\pi f) \exp\left[ 2 i \pi f k\cdot p_{0} \right] \tilde{h}_{\ell m} D_{a,e} : P_{\ell m} \,,
+\ee
+where we introduced the detector tensors
+\bsub\label{eq:DaDe}
+\begin{align}
+	D_{a} &= \frac{L}{2} \left( n_{1}\otimes n_{1} + n_{3} \otimes n_{3} - 2 n_{2} \otimes n_{2} \right) \,,\\
+	D_{e} &= \frac{L\sqrt{3}}{2} \left( n_{1} \otimes n_{1} - n_{3} \otimes n_{3} \right) \,.
+\end{align}
+\esub
+Here, we have made apparent factors $(-2i\pi f)$ corresponding to the Fourier-domain translation of a time derivative.
+Indeed, the observables $\Delta \nu / \nu$ used here are time derivatives of the laser phase more commonly used for ground-based detectors.
+
+We can now map these two channels to two fictitious LIGO-type detectors as follows.
+For an orthogonal detector of the same setup as the ground-based LIGO and Virgo with armlength $L_{D}$, rotated by an angle $\epsilon_{D}$ from the basis vectors $x,y$, the detector tensor is
+\begin{align}
+	D &= \frac{L_{D}}{2} \left[ \cos 2\epsilon_{D} \left( x\otimes x - y\otimes y \right) \right. \nn\\
+	& \qquad\quad + \left. \sin 2\epsilon_{D} \left( x\otimes y + y \otimes x\right) \right] \,.
+\end{align}
+By comparison with~\eqref{eq:DaDe}, using the LISA frame as detector frame (see App.~\ref{app:conventions}) we obtain
+\be
+	L_{a} = L_{e} = \frac{3L}{2} \,, \; \epsilon_{a} = \frac{2\pi}{3} \,, \; \epsilon_{e} = \frac{5\pi}{12} \,.
+\ee
+Note a degeneracy $\pm \pi$ in $\epsilon_{a,e}$, reflecting a freedom in the choice of the orientation of these effective detectors.
+
+We can recover familiar pattern functions analogous to the ones of ground-based detectors (for $\psi_{L}=0$) by factoring out the effective length, defining
+\be\label{eq:defFapcFepc}
+	F_{a,e}^{+,\times} \equiv \frac{1}{3L} D_{a,e} : P_{+,\times}^{0} \,,
+\ee
+which gives in terms of the LISA-frame sky position angles $\lambdaL$, $\betaL$:
+\bsub\label{eq:FapcFepc}
+\begin{align}
+	F_{a}^{+} &= \frac{1}{2} \left( 1 + \sin^{2}\beta_{L} \right) \cos \left(2\lambda_{L} - \frac{\pi}{3} \right) \,,\\
+	F_{a}^{\times} &= \sin\beta_{L} \sin \left(2\lambda_{L} - \frac{\pi}{3} \right) \,,\\
+	F_{e}^{+} &= \frac{1}{2} \left( 1 + \sin^{2}\beta_{L} \right) \cos \left(2\lambda_{L} + \frac{\pi}{6} \right) \,,\\
+	F_{e}^{\times} &= \sin\beta_{L} \sin \left(2\lambda_{L} + \frac{\pi}{6} \right)  \,.
+\end{align}
+\esub
+One can check that the expressions for the channel $e$ is obtained from the expressions for $a$ with the replacement $\lambdaL \rightarrow \lambdaL + \pi/4$.
+
+If we introduce
+\begin{align}\label{eq:Flmae}
+	F_{a,e}^{\ell m} &= \frac{1}{2} \sYlm e^{-2 i \psi_{L}} \left( F_{a,e}^{+} + i F_{a,e}^{\times}\right) \nn\\
+	& + \frac{1}{2} (-1)^{\ell} \sYlminusmstar e^{+2 i \psi_{L}} \left( F_{a,e}^{+} - i F_{a,e}^{\times} \right) \,,
+\end{align}
+the mode transfer functions~\eqref{eq:defTlmaet} become
+\be\label{eq:transferlowfae}
+	\tilde{a}, \tilde{e} = \sum_{\ell,m} \calT^{\ell m}_{a,e} \tilde{h}_{\ell m} = \left(-6 i \pi f L\right) \exp[2i \pi f k\cdot p_{0}] F_{a,e}^{\ell m} \tilde{h}_{\ell m}\,.
+\ee
+We see that scaling out $(-6 i \pi fL)$ as in~\eqref{eq:defhaet} brings us back to strain observables.
+The only difference with the response of a ground-based observatory is then the time dependency entering $p_{0}(t)$ and the LISA frame angles $\lambda_{L}(t)$, $\beta_{L}(t)$, $\psi_{L}(t)$ (see~\eqref{eq:Lframeangles}), with the time evaluated at $t_{f}^{\ell m}$.
+
+Finally, we explain how to write a response fully analogous to ground-based detectors.
+Above we have chosen to write $\psi$ in a prefactor outside the pattern functions, but we could write $\psi$-dependent patter functions as
+\bsub
+\begin{align}
+	F^{+}_{a,e}(\psi) &= \cos 2\psi F^{+}_{a,e} + \sin 2\psi F^{\times}_{a,e} \,,\\
+	F^{\times}_{a,e}(\psi) &= -\sin 2\psi F^{+}_{a,e} + \cos 2\psi F^{\times}_{a,e} \,,
+\end{align}
+\esub
+so that in the low-frequency approximation the response for the strain-like observables~\eqref{eq:defhaet} is
+\be
+	\tilde{h}_{a,e} = F^{+}_{a,e}(\psi) \tilde{h}_{+} + F^{\times}_{a,e}(\psi) \tilde{h}_{\times} \,,
+\ee
+if we ignore the delay $k \cdot p_{0}$ and treat the angles as constants.
+
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+
 \section{Methodology}
@@
 \label{subsec:defbayes}
-
--- definitions relevant to the Bayesian setting (likelihood, ...)
 
@@
 
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-
-\subsection{Frequency domain LISA response}
-\label{subsec:response}
-
-We will give in this section a brief summary of the instrument response of LISA, and of our Fourier-domain treatment of the modulations and delays.
-We refer the reader to App.~\ref{app:response} for comprehensive notations and definitions.
-
-The higher harmonics $h_{\ell, m}$ parametrize the decomposition of the gravitational wave signal $h_{+}$, $h_{\times}$ in spin-weighted spherical harmonics as
-\be\label{eq:hpcmodes}
-	h_{+} - i h_{\times} = \sum_{\ell \geq 2} \sum_{m = -\ell}^{\ell} {}_{-2}Y_{\ell m} (\iota, \varphi) h_{\ell m} \,,
-\ee
-where $(\iota, \varphi)$ is the direction of emission in a frame attached to the source by convention, and the ${}_{-2}Y_{\ell m}$ are spin-weighted spherical harmonics~\cite{}.
-
-The LISA response can be built from single-link observables $y_{slr} = (\nu_{r} - \nu_{s})/\nu$ respresenting a frequency shift between the spacecrafts $s$ and $r$ along the link $l$.
-We use the expression~\cite{}
-\be\label{eq:defyslr}
-	y_{slr} = \frac{1}{2} \frac{n_{l} \otimes n_{l}}{1 - k\cdot n_{l}} : \left[ H(t - L - k\cdot p_{s}) - H(t - k\cdot p_{r}) \right] \,,
-\ee
-with $H$ the transverse-traceless matrix representing the gravitational wave, $L$ the delay along one arm\footnote{We use units with $G=c=1$.}, taken to be fixed, $k$ the propagation vector, $n_{l}$ the link unit vectors and $p_{r}$, $p_{s}$ the positions of the spacecrafts.
-
-In the frequency domain, modes will be respresented with their amplitude and phase as\footnote{Note our Fourier convention~\eqref{eq:defFourier}.}
-\be\label{eq:hlmampphase}
-	\tilde{h}_{\ell m} (f) = A_{\ell m} (f) e^{-i\Psi_{\ell m} (f)}\,.
-\ee
-We will use the analysis of~\cite{MB18} and write the response in individual observables $y_{slr}$ with a transfer function for each spherical harmonic mode as
-\be
-	\tilde{y}_{slr} = \sum_{\ell, m}\calT_{slr}^{\ell m}(f) \tilde{h}_{\ell m} \,.
-\ee
-Applying the perturbative formalism of~\cite{MB18} to leading order, we have simply
-\be
-	\calT_{slr}^{\ell m}(f) = G_{slr}^{\ell m}(f, t_{f}^{\ell m}) \,,
-\ee
-where
-\begin{align}\label{eq:Gslr}
-	G_{slr}^{\ell m}(f,t) &= \frac{i \pi f L}{2} \sinc \left[ \pi f L\left(1-k\cdot n_{l} \right) \right] \nn\\
-	& \quad \cdot \exp\left[ i \pi f \left( L + k\cdot \left( p_{r} + p_{s} \right) \right) \right] \nn\\
-	& \quad \cdot n_{l} \cdot P_{\ell m} \cdot n_{l} \,
-\end{align}
-is the kernel from~\cite{MB18}, $P_{\ell m}$ is defined by the decomposition~\eqref{eq:Hsummodes}, and
-\be
-	t_{f}^{\ell m} = -\frac{1}{2\pi} \frac{d\Psi_{\ell m}}{df}
-\ee
-is the effective frequency$\to$time correspondence, defined across the whole frequency band and including the merger and ringdown.
-The analysis of~\cite{}  has shown that higher-order corrections are small for SMBH systems, and for SOBH systems provided they are close enough to merger as will be the case for the examples chosen here.
-The fact that we use the same Fourier-domain treatment of the transfer functions for both the signal and the templates should also mitigate the importance of modelling errors.
+\SM{complete} The likelihood function that we use, with a zero noise realization, is
+\begin{align}\label{eq:deflnLaet}
+	\ln\calL &= -\frac{1}{2} \left[ \left( \tilde{a} - \tilde{a}_{\rm inj} | \tilde{a} - \tilde{a}_{\rm inj}  \right) + \left( \tilde{e} - \tilde{e}_{\rm inj} | \tilde{e} - \tilde{e}_{\rm inj}  \right) \right.
+\nn\\
+	& \qquad\qquad + \left. \left( \tilde{t} - \tilde{t}_{\rm inj} | \tilde{t} - \tilde{t}_{\rm inj} \right) \right] \,,
+\end{align}
+with the $\rm inj$ subscript indicating the injected signals, and where the inner products $(\cdot | \cdot)$ are given by~\eqref{eq:definnerproduct} with the noise PSDs~\eqref{eq:Snaet}.
 
@@
 
-\subsection{Likelihood computation}
+\subsection{Likelihood computation with zero noise}
 \label{sec:likelihood}
@@
 In the LISA.
-\TDC{this paragraph seems to need more work.
+\SM{complete} \TDC{this paragraph seems to need more work.
 I would start with the existing general case, then say that by using a zero-noise realization we can write the amplitude/phase representation and use a sparse grid and get a large speedup.}
@@
 Note that individual terms in~\eqref{eq:lnLoverlapstructure} can be individually large, and computing $\ln \calL$ relies on accurate cancellations.
-Decomposing further in TDI channels as in~\eqref{} and in harmonics as in~\eqref{}, we have a sum of terms with the structure \TDC{we need one or two more equations describing explicitly how to go from $\tilde{y}_{slr}$ to $\ln \calL$}
-\be
-	\int \frac{df}{S_{n}} \; \calT_{1}^{\ell m} \calT_{2}^{\ell' m'} A_{1,\ell, m} A_{2, \ell' m'} e^{i(\Phi_{1,\ell m} - \Phi_{2,\ell' m'})} \,.
-\ee
-\TDC{should this notation be aligned to Eq 6 and 7?}
-The difficulty of computing such an overlap depends on the phase difference.
+Decomposing further in TDI channels as in~\eqref{} and in harmonics as in~\eqref{}, we have a sum of terms with the structure (symbolically) \TDC{we need one or two more equations describing explicitly how to go from $\tilde{y}_{slr}$ to $\ln \calL$}
+\be
+	\sum_{\ell m, \ell'm'}\int \frac{df}{S_{n}} \; \calT_{1}^{\ell m} \calT_{2}^{\ell' m'} A_{1,\ell, m} A_{2, \ell' m'} e^{i(\Phi_{1,\ell m} - \Phi_{2,\ell' m'})} \,,
+\ee
+with transfer functions $\calT$, amplitudes $A$, and phases $\Phi$. \TDC{should this notation be aligned to Eq 6 and 7?}
+The difficulty of computing numerically such an overlap depends on the phase difference.
 A large phase difference causes the integrand to be very oscillatory, requiring more frequency resolution than non-oscillatory integrands; on the other hand, oscillatory integrands give raise to a small integral due to cancellation effects, while non-oscillatory integrands contribute the most.
@@
 We can then compute the elementary integrals~\eqref{eq:fastlikestructure} using a combination of integrations by parts and asymptotic expansions, as explained in App.~\ref{app:numintegration}.
-The number of grid intervals is typically several hundreds \SM{to be completed}.
+The number of grid intervals is typically several hundreds \SM{to be completed, report speed}.
 
@@
 
-\section{Supermassive black holes}
+\section{Massive black holes signals}
 \label{sec:SMBH}
-
-- Demonstration/examples
 
@@
 
-\begin{figure*}
-  \centering
-  \includegraphics[width=.98\linewidth]{../plots/plot_smbh9_tdi.pdf}
-  \caption{\SM{TODO: rethink the visualization, maybe plot only transfer functions}.}
-  \label{fig:tranferSMBHCase9}
-\end{figure*}
+We focus our analysis on a single set of mass parameters, representative of canonical MBHB sources expected for LISA~\cite{Proposal}: we pick a total redshifted mass $M=m_{1}+m_{2} = 2\times10^{6} M_{\odot}$ and a mass ratio $q=m_{1}/m_{2} = 3$, and place the source at a redshift $z=4$.
+The inclination plays an important role in deciding the importance of higher harmonics; we set it to the median value $\iota = \pi/3$.
+We will study in detail two such systems, named cases I and II, stemming from a series of twelve runs with randomized orientation parameters, and chosen to exemplify qualitative differences in the parameter estimation.
+The parameters are summarized in Table~\ref{tab:MBHBparams}, along with the SNR of these signals with and without higher harmonics.
+The different randomized orientations and the inclusion of higher harmonics both have a sizeable impact on the SNR.
+
+We note that, by contrast with current ground-based observations that are horizon-limited and that have a strong selection bias towards systems with favorable orientations (face-on or face-off), the LISA horizon for MBHB systems extends to essentially the entirety of the observable universe; it is therefore natural to consider randomized orientations.
+
+\begin{table}\label{tab:MBHBparams}
+	\begin{tabular}{|c||c|c|}
+		\hline
+		Identifier 		& I 	& II   \\
+		\hline
+		\hline
+		Mass 1 ($\Msol$) 	& \multicolumn{2}{|c|}{$1.5\times10^{6}$} \\
+		\hline
+		Mass 2 ($\Msol$) 	&  \multicolumn{2}{|c|}{$0.5\times10^{6}$}\\
+		\hline
+		Redshift 		&  \multicolumn{2}{|c|}{$4.$}  \\
+		\hline
+		Lum. Distance (Mpc) 		&  \multicolumn{2}{|c|}{$36594.3$} \\
+		\hline
+		Inclination (rad) 		& \multicolumn{2}{|c|}{$\pi/3$}  \\
+		\hline
+		Phase (rad) 			& $2.140$	& $-1.249$ \\
+		\hline
+		Ecliptic longitude (rad) 			& $3.335$	& $2.275$ \\
+		\hline
+		Ecliptic latitude (rad) 			& $1.468$	& $-1.376$ \\
+		\hline
+		Polarization (rad) 			& $2.237$	& $1.635$ \\
+		\hline
+		\hline
+		LISA SNR $h_{22}$			& 857.4 & 645.7 \\
+		\hline
+		LISA SNR $h_{\ell m}$			& 944.8 & 666.0 \\
+		\hline
+	\end{tabular}
+	\caption{Parameters of the simulated MBHB mergers. \TDC{Maybe we can show
+			 these on the upper right of each corner plot instead.}}
+\end{table}
 
@@
   \caption{.}
-  \label{fig:FDHMSMBHCase9}
+  \label{fig:hctdiahm}
 \end{figure}
 
--- add plot for SNR accumulation with time or frequency: inspiral/MRD balance across mass range
-
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-
-\subsection{Parameter estimation}
+\begin{figure*}
+  \centering
+  \includegraphics[width=.98\linewidth]{../plots/plot_smbh9_tdi.pdf}
+  \caption{\SM{TODO: rethink the visualization, maybe plot only transfer functions and Doppler phase}.}
+  \label{fig:tranferSMBHCase9}
+\end{figure*}
+
+We illustrate the Fourier-domain signal in Fig.~\ref{hctdiahm}, with its higher harmonics, by showing mode-by-mode contributions to the characteristic strain, as well as the characteristic noise PSD, defined to be
+\bsub
+\begin{align}
+	\tilde{h}_{a,e,t}^{c} (f) &= 2 f \tilde{h}_{a,e,t} (f) \,,\\
+	S^{a,e,t}_{c} (f) &= f S^{a,e,t}_{h} (f) \,,
+\end{align}
+\esub
+with the strain-like observables and noise PSD are defind in~\eqref{eq:defhaet} and~\eqref{eq:defShaet}.
+We represent the $A$ TDI channel (the $E$ channel being similar, and the $T$ channel negligible at low frequencies).
+This characteristic strain and associated PSD are often used for plotting~\cite{}, since one can visualize the $\SNR$ by integrating by eye under the curve with a logarithmic $x$-axis.
+The full signal shows oscillatory features in frequency that are due to beatings between the harmonics that have smooth amplitudes individually.
+
+We also highlight in Fig.~\ref{fig:hctdiahm} different time intervals before the merger (see the next Section~\ref{subsec:MBHBacctime}).
+These times correspond each to a different frequency across modes, by virtue of the approximate relation between instantaneous mode frequencies in the inspiral:
+\be\label{eq:fscalinglm}
+	f_{\ell m}(t) \simeq \frac{m}{2} \times f_{22}(t) \quad (\mathrm{Inspiral})\,.
+\ee
+We also highlight the frequency at merger, also denoted peak frequency.
+Since we represent mode contributions to the TDI observables, the response is included in this Figure; we see that all modes display structure at low frequencies that corresponds to the time modulation resulting of the LISA motion.
+
+Fig.~\ref{fig:tranferSMBHCase9} shows the detail of the transfer functions themselves.
+The Doppler phase variation $\Delta \Phi_{R}$ is a small effect at low frequencies, being suppressed by a factor $\sim 2\pi f R$; it is also small at higher frequencies, because the chirp happens over a short time interval during which $\Phi_{R}$ does not have time to vary.
+We also show transfer functions for the mode $h_{22}$ in the three TDI channels $A,E,T$, ($T$ is suppressed at low frequencies), after factoring out the Doppler phase.
+The low-frequency features show the time-dependency of the response created by the LISA motion, while frequency-dependency in the response appears at high frequencies.
+We will decompose further these different physical effects below in Sec.~\ref{subsec:MBHBresponseapprox}
+
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+
+\subsection{Accumulation of signal with time}
+\label{subsec:MBHBacctime}
+
+\begin{figure*}
+  \centering
+  \includegraphics[width=.98\linewidth]{../plots/overlapmodebymode_maxfscaledlm_t.pdf}
+  \caption{.}
+  \label{fig:modeoverlapsoftime}
+\end{figure*}
+
+We now illustrate the accumulation of the gravitational wave signal with time, which is crucial to understand the ability of the instrument to indentify and localize MBHB signals in advance of their coalescence, as well as the requirements put on the instrumental configuration (data gaps and downlink cadence).
+As will be shown in more details in a separate publication~\cite{}, within the uncertainties of astrophysical models we can expect the bulk of MBHB signals to be detectable only a few days prior to merger, while a tail of more favorable events could be observable for significantly longer times, up to months.
+
+We will focus on the system II of Table~\ref{tab:MBHBparams}, and we will highlight four different epochs, corresponding to the time before merger where a certain fraction of the total $\SNR = 666$ has been accumulated.
+We include higher harmonics here.
+These epochs are:
+\begin{itemize}
+	\item $\SNR/64$: $\sim 42\mathrm{h}$ before merger, which corresponds roughly, with $\SNR \simeq 10$, to the first time we could confidently claim a detection;
+	\item $\SNR/16$: $\sim 2.5\mathrm{h}$ before merger;
+	\item $\SNR/4$: $\sim 7\mathrm{min}$ before merger.
+\end{itemize}
+
+To represent the observations stopping with a certain time to merger left, we will approximate the cut in time using simply the time-to-frequency correspondence~\eqref{eq:deftflm} to translate it as a cut in frequency.
+In a more realistic setting, an abrupt interruption of the signal would translate into non-local features in the Fourier domain unless tapering is applied, but we ignore these complications here.
+
+The time-to-frequency correspondence~\eqref{eq:deftflm} gives for each mode time as a function of frequency.
+It is accurate for the inspiral, as such a correspondence is at the heart of the SPA \SM{check SPA defined}.
+As a function it also extends formally through merger, while losing in accuracy and in physical interpretation, becoming non-monotonous with frequency (see the discussion in~\cite{MB18}).
+Due to the scaling $h_{\ell m} \propto e^{-i m \phi_{\rm orb}}$ with $\phi_{\rm orb}$ the orbital phase, a given time corresponds to different frequencies $f_{\ell m}(t) \simeq m/2 \times f_{22}(t)$ \SM{update with previous section} for different modes during the inspiral.
+At merger and this relation will be lost and the non-monotonicity of $\tf$ makes it impossible to invert $\tf$ to build $f(t)$.
+Equivalently,
+\be\label{eq:relationtflm}
+	t_{f}^{\ell m} \left( \frac{m}{2} f\right) \simeq t_{f}^{22} (f) \quad (\mathrm{Inspiral}) \,.
+\ee
+This relation is illustrated in Fig.~\ref{}. We see that it holds almost up to merger \SM{complete}, while different modes start to depart after this point.
+
+Fig.~\ref{fig:modeoverlapsoftime} shows the accumulated contributions to the total SNR of individual mode combinations.
+For a generic signal $s = \sum_{\ell m} s_{\ell m}$ with mode contributions $s_{\ell m}$ ($s$ is any among the channels $\tilde{a}$, $\tilde{e}$, $\tilde{t}$)
+\be
+	\SNR^{2} = (s|s) = \sum_{\ell m} \sum_{\ell' m'} (s_{\ell m} | s_{\ell' m'}) \,.
+\ee
+We choose to show contributions to $\SNR^{2}$ because this is the scale on which the contributions of different modes and channels are additive, and because it is the relevant scale for the log-likelihood~\eqref{eq:deflnL}, since $\ln \calL \sim \SNR^{2}$.
+We also sum over the three channels ($T$ being essentially negligible).
+Together with the total contribution to $\SNR^{2}$, the three panels of Fig.~\ref{fig:modeoverlapsoftime} show diagonal terms $(s_{\ell m} | s_{\ell m})$, cross-terms involving the dominant quadrupolar mode $(s_{22} | s_{\ell m})$, and finally other cross-terms between subdominant modes.
+Results are displayed as a function of the time to merger, and we show separately the result obtained for the full post-merger signal.
+
+We can draw several conclusions from Fig.~\ref{fig:modeoverlapsoftime}.
+The signal reaches detectability at $\SNR=10$ about two days before merger.
+We see that the SNR accumulates rapidly in the last instants before merger, as shown in particular by reaching $\SNR/4$ only 7 minutes before coalescence.
+Stopping the signal at a given time in the inspiral changes somewhat the hierarchy between subdominant modes: contrasting Fig.~\ref{fig:modeoverlapsoftime} with Fig.~\ref{fig:hctdiahm}, we see e.g that $h_{21}$ is now subdominant before merger (and increases significantly in strength when including the post-merger signal).
+This is because modes $h_{\ell m}$ with a higher $m$ will reach a higher frequency at any given time, while the weighting by the noise favors higher frequencies.
+This is also visible in Fig.~\ref{fig:hctdiahm} where ticks translate a cut in time into a different cut in frequency for different modes.
+
+We also see that, as even the most subdominant modes contribute significantly to $\SNR^{2}$, our set $(22, 21, 33, 44, 55)$ is incomplete; we will need waveforms with a richer set of higher harmonics to analyze real LISA data.
+Diagonal terms $(s_{\ell m} | s_{\ell m})$ and cross terms with $(\ell m) \neq (\ell' m')$ are qualitatively different.
+Diagonal terms accumulate coherently, while cross terms are oscillatory, as they feature two modes with different phasings.
+This oscillatory character tends to suppress the contribution of the cross terms; however, we see they are not negligible, in particular the ones involving the dominant harmonic and a subdominant harmonic.
+
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+
+\subsection{Decomposing the instrument response}
+\label{subsec:MBHBresponseapprox}
+
+\begin{figure}
+  \centering
+  \includegraphics[width=.98\linewidth]{../plots/responseapprox_tdia_22.pdf}
+  \caption{Response approximations.}
+  \label{fig:responseapproxtdia22}
+\end{figure}
+
+The LISA instrument response, as recalled in Section~\ref{sec:response}, features both time- and frequency-dependency.
+As we will see, these features play an important role in breaking degeneracies in the parameter estimation of the source, notably allowing us to localize the source in the sky.
+It is therefore important to understand these features, also in the light of the pre-merger accumulation of signal with time.
+
+The time-dependency in the response mirrors the motion of the LISA constellation.
+In the low-frequency picture of~Sec.~\ref{subsec:lowfresponse}, the time dependency enters both in the Doppler phase term~\eqref{} (LISA moves in the wavefront) and in the time-dependent LISA-frame angles~\eqref{} (the LISA arms change orientation).
+However, we have seen in the previous section that the SMBH signals we consider here are quite short (less than two days).
+This limits the effect of the LISA motion: between the time where the signal becomes detectable and the merger, LISA has barely changed in orientation and position.
+Moreover, the region where the signal accumulates most of its SNR is even shorter, as shown in Fig.~{}.
+
+To this end, we distinguish four different versions of the response:
+\begin{enumerate}[(i)]
+	\item \textit{Full}: in this case, we keep the full response of Secs.~\ref{subsec:FDresponse}-\ref{subsec:TDIresponse}, with its complete time and frequency dependency;
+	\item \textit{Frozen}: we neglect the LISA motion by evaluating all time-dependent vectors at $\tf = t_{f}^{\rm peak}$, effectively freezing LISA in its orbit, while keeping the frequency dependency in the response;
+	\item \textit{Low-$f$}: we implement the low-frequency (long-wavelength) approximation as in Sec.~\ref{subsec:lowfresponse}, while still keeping the time-dependency due to the motion;
+	\item \textit{Frozen low-$f$}: we neglect both the time and frequency dependency in the response.
+\end{enumerate}
+
+Case (iii), \textit{Low-$f$}, has been extensively used in the past for the analysis of MBHB signals.
+In particular, this simplification has been used in previous MLDC studies.
+Being based on the $f \ll f_{L}$ approximation, it is less inaccurate for inspiral-only signals, to which most previous studies were limited, as it is for a full IMR signal.
+
+In the case (iv), \textit{Frozen low-$f$}, the response is equivalent to two LIGO-type detectors, lying motionless in the same location and rotated from each other by $\pi/4$, and there is no other information on the source's sky position other than the pattern functions of the two effective detectors.
+Contrarily to networks of ground-based observatories, we have no triangulation information from times of arrival at different detectors.
+This case is the most degenerate, and will be useful to get an anlytical understanding of approximate degeneracies occuring when using the more complicated full response (i).
+
+The effect on the response of adopting these approximations is shown in Fig.~\ref{fig:responseapproxtdia22}.
+We display the strain transfer function $\calT_{h_{a}}^{22}(f)$ as defined in~\eqref{eq:defhaet} for the mode $h_{22}$, with the four lines showing the four approximations (i)-(iv).
+Vertical lines also indicate the times to merger highlighted in Sec.~\ref{subsec:MBHBacctime}, as well as the peak frequency.
+Paralleling this figure to Figs.~\ref{fig:hctdiahm} and~\ref{fig:modeoverlapsoftime}, we stress that most of the SNR is accumulated in the very last instants before merger and at the merger itself, due to the noise normalization not visible at the level of the transfer function.
+
+The \textit{Frozen low-$f$} transfer function is just a constant factor, similarly to the LIGO response where $h_{+,\times}$ are simply multiplied by pattern functions.
+The \textit{Low-$f$} transfer function goes to the same constant at high frequencies, where the signal chirps so rapidly that the LISA motion is negligible.
+At low frequencies, however, modulations due to the LISA motion appear.
+The \textit{Frozen} transfer function asymptotes to the constant of the \textit{Frozen low-$f$} case at low frequencies, for $f \ll f_{L}$.
+At higher frequencies, we see a growing departure from this approximation.
+As noted in Sec.~\ref{subsec:lowfresponse}, when reaching the armlength transfer frequency $f_{L} = 0.12 \Hz$ the long-wavelength approximation has completely broken down; departures start to be significant at much lower frequencies.
+Finally, the \textit{Full} transfer function displays all the features we discussed.
+
+We note finally a coincidence in Fig.~\ref{fig:responseapproxtdia22}: the frequency below which we see the imprint of the LISA motion and the frequency above which we see the breakdown of the long-wavelength approximation appear to be the same.
+This will not be true in general: the former is essentially a measure of the time-to-frequency correspondence $\tf$, with lower-mass signals having support and showing these features at higher frequencies, while the latter only marks the magnitude of $2\pi f L$ factors and is source-independent\footnote{The effect in the transfer functions is source-independent; however, the SNR accumulation with frequency is source-dependent and will influence the impact of these features on the parameter estimation.}.
+
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+
+\section{Massive black holes parameter estimation}
 \label{sec:SMBHPE}
+
+\subsection{Analysis of IMR signals}
+\label{subsec:SMBHPEfull}
 
@@
   \caption{.}
-  \label{fig:PEsmbh22hmCase0}
+  \label{fig:PEsmbh22hmCase9}
 \end{figure*}
 
--- High-mass system, non-degenerate, 22 and HM + Fisher - stacked SNR contours
-eg:
-
-Case 12?
-m1,    m2,    t0,D,     phi0,i,   lam,  bet, psi
-8.89e5,1.11e5,0, 3.67e4,pi/3,pi/2,3pi/4,pi/3,pi/3
-
---enumerate priors
-
-Case 0
-
--- Performance: waveform and likelihood costs, number of sampler evaluations
-
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-
-\subsection{Accumulation of signal with time}
-\label{sec:SMBHPEacctime}
+We performed Bayesian parameter estimation for the two MBHB sources listed in Table.~\ref{tab:MBHBparams}, using the methodology summarized in Sec.~\ref{sec:method}.
+In particular, we set the noise realization to zero, used accelerated likelihood based on an amplitude and phase representation of the signals.
+The priors used are logarithmic in mass, flat in luminosity distance, and flat for the angles: flat on the sphere for the pairs of angles $(\iota, \varphi)$ and $(\lambda, \beta)$, and flat for the polarization $\psi$.
+We expect priors to be unimportant for the masses, since they are well determined in this very high SNR limit.
+The luminosity distance, as we will see, can be less well determined in the absence of higher harmonics, and one should keep in mind that the prior choice does affect the posterior there.
+
+To illustrate the role of higher harmonics in the analysis, we will present two classes of results: ``22'', where we inject and recover signals including only the dominant harmonic $h_{22}$, and ``HM'', where we both inject and recover with the higher harmonics $(h_{22}, h_{21}, h_{33}, h_{44}, h_{55})$.
+We do not attempt to recover an injection with higher modes using 22-mode waveforms; one expects parameter biases in this case, induced by the inadequacy of the simplified waveforms.
+A related question would be to assess the mode content necessary for waveform models to mitigate systematic biases when analyzing full GR signals.
+We leave such waveform systematics studies for future work; we only note that, as shown in Fig.~\ref{fig:modeoverlapsoftime}, the set of five modes we include in the present study is quite obviously incomplete.
+
+Figs.~\ref{fig:PEsmbh22hmCase0} and~\ref{fig:PEsmbh22hmCase9} show posterior distributions for all parameters for Cases I and II respectively.
+We overlay ``22'' and ``HM'' posteriors.
+All results for MBHB systems are presented using LISA-frame parameters, as defined in App.~\ref{app:conventions}.
+We should keep in mind the difference in SNR of Cases I and II ($\SNR \simeq 945$ versus $\SNR \simeq 666$ with higher harmonics, see Table~\ref{tab:MBHBparams}), resulting from their different orientation, however their posteriors have qualitative differences that go beyond a simple scaling of the errors as $\sim 1/\SNR$ \SM{redundant ?}.
+
+In Case I, the ``22'' posterior appears to be well represented by a multimodal Gaussian, with multimodality for the angular parameters.
+The sky position, most notably, admits a degenerate mode at $\betaL \rightarrow -\betaL$, although it has less weight than the main mode around the injection.
+By contrast, in Case II, the ``22'' posterior is much more degenerate.
+The distance and inclination are very degenerate with each other, with a support extending all the way to $\iota=0$ (or $\iota=\pi$).
+The phase $\varphi$ and polarization $\psiL$ have a distinct extended degeneracy along lines of constant $\varphi + \psiL$ and $\varphi - \psiL$.
+Most notably, the sky position, if retaining the same overall bimodality $\betaL \rightarrow -\betaL$ as in Case I, shows here a curious feature: the marginalized sky posterior peaks away from the injected value.
+Looking at other panels, we see that the shifted peak corresponds to the region of high distance (and extremal inclination).
+This is a genuine featuer of the multidimensional zero-noise posterior and will be explained in Sec.~\ref{subsec:MBHBPEdegen22}.
+
+In both Cases I and II, including the higher harmonics has a major effect on the parameter recovery, much beyond what we would expect from the sole gain in total SNR shown in Table.~\ref{tab:MBHBparams} (and a scaling of statistical errors $\sim 1/\SNR$).
+The marginalized posterior of the masses is narrower, although not qualitatively different; the major change is in the extrinsic parameters.
+
+We can understand the effect of the higher harmonics by noting that, in a signal of the form $\sum_{\ell m} \sYlm(\iota, \varphi)$, measuring independently two separate contributions $(\ell m) \neq (\ell' m')$ gives us an independent measurement of the inclination $\iota$ by the relative amplitude of the two mode contributions
+
+First, including the higher harmonics in the analysis breaks the degeneracy between distance and inclination, an effect that is well known in the context of ground-based observations.
+
+Second, the phase $\varphi$ also becomes independently measured.
+
+\SM{Add sky area numbers ? Add table of $\pm 1\sigma$ intervals ?}
+
+\SM{Add comparison between the two samplers + Fisher: Cases I \& II, 22 \& hm ($\times 4$), m1-m2, D-iota, lambda-beta}
+
+\SM{Performance: waveform and likelihood costs, number of sampler evaluations}
+
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+
+\subsection{Degeneracies in the sky}
+\label{subsec:MBHBPEdegen}
+
+The most remarkable feature in Figs.~\ref{fig:PEsmbh22hmCase0} and~\ref{fig:PEsmbh22hmCase9} is the existence of a degenerate mode (or secondary maximum) in the posterior distribution for the sky position $(\lambda_{L}, \beta_{L})$, located at $\beta_{L}^{*} = -\beta_{L}^{\rm inj}$, i.e.
+by reflecting the wave vector $k$ across the plane of the LISA constellation.
+This secondary mode has less weight than the main mode around the injection, but is present in both cases.
+In particular, it survives when including higher harmonics.
+
+We can gain insight about degeneracies in the parameter space by considering the simple expressions representing the instrument response in the \textit{Frozen low-$f$} approximation described in Sec.~\ref{subsec:responseapprox}, appropriate for the low-frequency limit and signals short enough that the LISA motion can be neglected; we will explore in details the respective influence of these neglected effects in Sec.~\ref{sec:MBHBPEacctime}.
+
+The \textit{Frozen low-$f$} response is given by~\eqref{eq:FapcFepc}-\eqref{eq:transferlowfae}, ignoring the $k\cdot p_{0}$ delay as a mere constant delay, and neglecting time-dependency in the LISA frame angles $(\lambda_{L}, \beta_{L}, \psi_{L})$.
+The response is then summarized by the pattern functions for harmonics $F^{\ell m}_{a,e} (\iota, \varphi, \lambda_{L}, \beta_{L}, \psi_{L})$ given in~\eqref{eq:Flmae}, which we reproduce here:
+\begin{align}\label{eq:Flmaerepeat}
+	F_{a,e}^{\ell m} &= \frac{1}{2} \sYlm (\iota, \varphi) e^{-2 i \psi_{L}} \left( F_{a,e}^{+} + i F_{a,e}^{\times}\right) (\lambda_{L}, \beta_{L}) \nn\\
+	& + \frac{1}{2} (-1)^{\ell} \sYlminusmstar (\iota, \varphi) e^{+2 i \psi_{L}} \left( F_{a,e}^{+} - i F_{a,e}^{\times} \right) (\lambda_{L}, \beta_{L}) \,.
+\end{align}
+
+Changing the sign of $\beta_{L}$ in the pattern functions $F_{a,e}^{+,\times}$ given in~\eqref{eq:FapcFepc} changes the sign of $F_{a,e}^{\times}$ for both channels, leaving $F_{a,e}^{+}$ invariant.
+In~\eqref{eq:Flmaerepeat} the $F_{a,e}^{+,\times}$ factors of the two lines are exchanged.
+Since spin-weighted spherical harmonics~\cite{} obey the relation
+\be
+	\sYlm (\pi - \iota, \varphi) = (-1)^{\ell} \sYlminusmstar (\iota, \varphi) \,,
+\ee
+we see that changing $\iota \rightarrow \pi-\iota$, $\psi_{L} \rightarrow \pi - \psi_{L}$ leaves $F_{a}^{\ell m}$, $F_{e}^{\ell m}$ invariant.
+
+This defines a transformation of extrinsic parameters yielding an exact degeneracy in the \textit{Frozen low-$f$} approximation, which we call the \textit{reflected} sky position (for a reflection with respect to the LISA plane):
+\begin{align} \label{eq:degenreflection}
+	\lambdaL^{*} &= \lambdaL \,, \nn \\
+	\betaL^{*} &= -\betaL \,, \nn \\
+	\psiL^{*} &= \pi - \psiL \,, \nn \\
+	\iota^{*} &= \pi - \iota \,, \nn \\
+	\varphi^{*} &= \varphi \,,
+\end{align}
+where we chose the transformation for $\psiL$ to keep this parameter in the range $[0, \pi]$.
+
+From the structure of~\eqref{eq:Flmaerepeat}, other points in parameter space are degenerate in the low-frequency limit.
+First, $\lambdaL \rightarrow \lambdaL + \pi$ leaves the pattern functions $F_{a,e}^{+,\times}$ invariant.
+For $\lambdaL \rightarrow \lambdaL \pm \pi/2$, these pattern functions acquire an overall minus sign.
+Such an overall minus sign is readily compensated by a shift $\psiL \rightarrow \psiL \pm \pi/2$, so that we have the other transformation
+\begin{align} \label{eq:degenlambdapi2}
+	\lambdaL^{(k)} &= \lambdaL + \frac{k \pi}{2} \mod 2\pi \,, \, k = 0, \dots, 3 \,, \nn \\
+	\psiL^{(k)} &= \psiL + \frac{k \pi}{2} \mod \pi \,, \, k = 0, \dots, 3 \,.
+\end{align}
+
+Combining~\eqref{eq:degenreflection} and~\eqref{eq:degenlambdapi2}, we arrive at eight different degenerate positions in the sky, equally spaced in $\lambdaL$ and symmetric above and below the LISA plane, with an inclination $\pi - \iota$ for the reflected positions and various values for the polarization $\psiL$.
+Among these eight secondary modes in the sky, two play special roles: first, the \textit{reflected} mode~\eqref{eq:degenreflection} already mentioned; and second, the \textit{antipodal} mode with
+\begin{align} \label{eq:degenantipodal}
+	\lambdaL^{(a)} &= \lambdaL + \pi \,, \nn \\
+	\betaL^{(a)} &= -\betaL \,, \nn \\
+	\psiL^{(a)} &= \pi - \psiL \,, \nn \\
+	\iota^{(a)} &= \pi - \iota \,, \nn \\
+	\varphi^{(a)} &= \varphi \,.
+\end{align}
+
+Importantly, when moving away from the \textit{Frozen low-$f$} approximation, taking into account the frequency-dependence in the \textit{Frozen} response, the \textit{reflected} is the only mode remaining degenerate.
+Indeed, as noted in Sec.~\ref{subsec:FDresponse}, in~\eqref{eq:Gslr} the frequency-dependent terms feature $k\cdot n_{l}$, $k \cdot p_{s,r}$ projections, that depend only on the projection of the wave vector $k$ in the plane of LISA, invariant for $\betaL \rightarrow -\betaL$.
+The LISA motion, however, will break the degeneracy.
+
+When taking into account the LISA motion in the \textit{Low-$f$} response, the \textit{antipodal} mode is the only one that keeps pattern functions that are exactly degenerate with the injection: all other modes will move together with the time-dependent LISA frame.
+The antipodal mode is not moving, because the antipode of the true direction of the arriving signal  is defined as $k \rightarrow -k$ independently of the orientation of LISA; the only degeneracy-breaking term is then the orbital delay $k \cdot p_{0}$.
+With the \textit{Full} response, by contrast, frequency dependency terms in~\eqref{eq:Gslr} featuring $k\cdot n_{l}$, $k \cdot p_{s,r}$ readily break this antipodal degeneracy.
+
+\begin{table}\label{tab:MBHBdegen}
+	\begin{tabularx}{.48\textwidth}{|X||c|c|c|c|}
+		\hline
+		Sky mode & \textit{Full} & \textit{Frozen} & \textit{Low-$f$} & \textit{Frozen low-$f$} \\
+		\hline
+		\textit{reflected}: \newline $-\betaL$, $\lambdaL$ & $t$-dep. & degen. & $t$-dep. & degen.  \\
+		\hline
+		\textit{antipodal}: \newline $-\betaL$, $\lambdaL + \pi$ & $f$-dep.$+\Delta \Phi_{R}$ & $f$-dep. & $\Delta \Phi_{R}$ & degen.  \\
+		\hline
+		$\betaL$, $\lambdaL + \pi/2$ & $t$-$f$-dep. & $f$-dep. & $t$-dep. & degen.  \\
+		\hline
+		$\betaL$, $\lambdaL + \pi$ & $t$-$f$-dep. & $f$-dep. & $t$-dep. & degen.  \\
+		\hline
+		$\betaL$, $\lambdaL - \pi/2$ & $t$-$f$-dep. & $f$-dep. & $t$-dep. & degen.  \\
+		\hline
+		$-\betaL$, $\lambdaL + \pi/2$ & $t$-$f$-dep. & $f$-dep. & $t$-dep. & degen.  \\
+		\hline
+		$-\betaL$, $\lambdaL - \pi/2$ & $t$-$f$-dep. & $f$-dep. & $t$-dep. & degen.  \\
+		\hline
+	\end{tabularx}
+	\caption{Sketch of the degeneracy structure of the eight modes in the sky, with the cases where the degeneracy is exact (degen.), and with the qualitative effects that break the degeneracy with the injected signal: $t$-dependency, $f$-dependency, both, or only the Doppler phase $\Delta \Phi_{R}$.
+This qualitative structure is the same with and without higher harmonics.}
+\end{table}
+
+Thus, we have found that in the \textit{Frozen low-$f$} approximation for the response we expect a pattern of degeneracy of eight degenerate positions in the sky (with certain rules for inclination and polarization).
+Table~\ref{tab:MBHBdegen} summarizes the qualitative picture of degeneracy breaking by the features of the response.
+We will see in Sec.~\ref{sec:SMBHPEacctime} that the eight-mode degeneracy pattern indeed appears when doing a pre-merger analysis; on the other hand, in results Figs.~\ref{fig:PEsmbh22hmCase0} and~\ref{fig:PEsmbh22hmCase9} for full IMR signals, out of the eight possible sky modes only the \textit{reflected} mode~\eqref{eq:degenreflection} survives.
+We will investigate in detail in Sec.~\ref{sec:SMBHPEacctime} how time-dependence and frequency-dependence in the response break part, but not all, of the degeneracies.
+
+However, this analysis does not explain why one obtains, with a zero noise realization, marginalized posteriors for the sky positions that appear biased from the injected signal, when ignoring higher harmonics.
+This is the question that we will address in the next Section.
+
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+
+\subsection{Apparent biases in the absence of higher harmonics}
+\label{subsec:MBHBPEdegen22}
+
+\begin{figure}
+  \centering
+  \begin{minipage}{.49\linewidth}
+  \includegraphics[width=.99\linewidth]{../plots/corner_skydown_smbh_case9_22_ptmcmc_bambi_withdegen.png}
+  \end{minipage}
+  \begin{minipage}{.49\linewidth}
+  \includegraphics[width=.99\linewidth]{../plots/corner_skydown_smbh_case9_simplelike_ptmcmc_bambi_withdegen.png}
+  \end{minipage}
+  \caption{Sky position posterior in presence of degeneracies, zooming close to the injected value.
+The left panel shows the complete likelihood, while the right panel shows the simplified likelihood~\eqref{eq:simplelikelihoodhm} with a motionless LISA and fixed masses and time.
+The results of \texttt{PTMCMC} (blue) and \texttt{MultiNest} are superimposed.
+The black cross is the injected value, and the grey cross is the analytic prediction~\eqref{eq:soldegensky} for the offset peak created by degeneracies.}
+  \label{fig:Skydownsmbh22hmSimpleLikeCase9}
+\end{figure}
+
+In this section, we investigate the cause of the degeneracies appearing in the posterior of Case~II shown in~\ref{subsec:SMBHPE}.
+Understanding the structure of these features could prove useful to inform Bayesian samplers, for instance by building adapted jump proposal allowing a much faster mixing of MCMC chains than naive methods.
+a simplified form of the extrinsic likelihood and the structure of its degeneracies.
+
+An important clue about the appearance of these degeneracies is the length of the observable signal.
+As shown in Fig.~\ref{}, for our example SMBH systems the SNR accumulates in a matter of days before merger, which means that there is only little impact of the modulations caused by the motion of the detector on the TDI transfer functions~\eqref{}.
+If we can neglect entirely the motion of the detector (which is not quite true as we will see later on), then we are in a regime where the information about the sky position of the source comes essentially from the pattern functions affecting this high-SNR signal.
+Using the low-frequency approximation (see~\ref{}) and the time-at-LISA $t_{L}$ as a time variable, there is no frequency dependence in the response.
+A further simplification occurs when pinning the masses and the time of coalescence to their true values.
+We found indeed that these parameters correlate only weakly to the other ones, so that we can hope to understand the degeneracy features by restricting the analysis to the six extrinsic parameters $(D_{L}, \iota, \varphi, \lambdaL, \betaL, \psiL)$.
+\TDC{we could make the connection with a similar decoupling in LIGO/Virgo, which is actually what makes low-latency sky localization possible.}
+
+Under all these simplifying assumptions, the likelihood function becomes purely extrinsic, and takes the form of trivial geometric factors multiplying a constant overlap that can be precomputed.
+Using the subscript $\mathrm{inj}$ for the injected signal, the likelihood~\eqref{eq:defliklihood} is
+\be
+	\ln \calL = -\frac{1}{2} \left( \tilde{a} - \tilde{a}_{\rm inj} | \tilde{a} - \tilde{a}_{\rm inj} \right) -\frac{1}{2} \left( \tilde{e} - \tilde{e}_{\rm inj} | \tilde{e} - \tilde{e}_{\rm inj} \right) \,.
+\ee
+From the low-frequency response \eqref{eq:aelowfmodes}, pinning the time-at-LISA~\eqref{} \SM{to be completed for $t_{L}$} so that we can ignore the factor $\exp [2 i \pi f k\cdot p_{0}]$ as a pure constant corresponding to a redefinition of time, we obtain \JGB{Should we use notation $\Phi_R$ here and in \eqref{eq:aelowfmodes}?}
+\begin{align}\label{eq:simplelikelihoodhm}
+	\ln \calL &= -\frac{1}{2} \sum_{\ell m} \sum_{\ell' m'}   \left[ \left( s_{a}^{\ell m} - s_{a, \mathrm{inj}}^{\ell m}\right) \left( s_{a}^{\ell' m'} - s_{a, \mathrm{inj}}^{\ell' m'}\right)^{*} \right.
+\nn\\
+	& \qquad \left.
++ \left( s_{e}^{\ell m} - s_{e, \mathrm{inj}}^{\ell m}\right) \left( s_{e}^{\ell' m'} - s_{e, \mathrm{inj}}^{\ell' m'}\right)^{*} \right] \left\langle \ell m | \ell' m' \right\rangle \,,
+\end{align}
+where we introduced the following extrinsic signal functions \JGB{Since neither $D_{a,e}$ nor $P_{\ell m}$ are defined in the main text, may be better to write in terms of a version of $F$ as in \eqref{eq:defFapcFepc}}
+\be\label{eq:defsaelm}
+	s_{a,e}^{\ell m} = \frac{1}{d} \frac{2}{3L} D_{a,e} : P_{\ell m} \,,
+\ee
+with $d = D_{L} / D_{L}^{\rm inj}$ the dimensionless ratio of luminosity distances.
+The prefactors $\left\langle \ell m | \ell' m' \right\rangle$ \sout{are pure normalization constants, and} represent mode overlaps as
+\be\label{eq:innerproductlmlpmp}
+	\left\langle \ell m | \ell' m' \right\rangle = 4\mathrm{Re} \int \frac{df}{S_{n}^{a,e}} \left( 3 \pi f L\right)^{2} \tilde{h}_{\ell m} \tilde{h}_{\ell' m'}^{*} \,.
+\ee
+Here, the noise PSD $S_{n}^{a,e}$ is given by~\eqref{} and is identical between the channels $a$ and $e$.
+Having fixed the masses and the time, these overlaps are to be computed with the modes $\tilde{h}_{\ell m}^{\rm inj}$ of the injected signal.
+
+In the case where we only include the dominant harmonic $h_{22}$, the likelihood~\eqref{eq:simplelikelihoodhm} simplifies to
+\be\label{eq:simplelikelihood22}
+	\ln \calL = -\frac{1}{2} \left\langle 22 | 22 \right\rangle  \left[ \left| s_{a}^{22} - s_{a, \mathrm{inj}}^{22} \right|^{2} + \left| s_{e}^{22} - s_{e, \mathrm{inj}}^{22}\right|^{2} \right] \,.
+\ee
+The functions~\eqref{eq:defsaelm} take the explicit form
+\begin{align}
+	s_{a,e}^{22} &= \frac{1}{4d} \sqrt{\frac{5}{\pi}} \cos^{4}\frac{\iota}{2} e^{2i(\varphi-\psiL)} \left( F_{a,e}^{+} + i F_{a,e}^{\times} \right) \\
+	&+ \frac{1}{4d} \sqrt{\frac{5}{\pi}} \sin^{4}\frac{\iota}{2} e^{2i(\varphi+\psiL)} \left( F_{a,e}^{+} - i F_{a,e}^{\times} \right) \,,
+\end{align}
+with the pattern functions for the channels $a$ and $e$ defined in \eqref{eq:defFapcFepc}-\eqref{eq:FapcFepc}.
+
+In order to make more manifest the degeneracies in the problem, it will be useful to introduce the following notations.
+First, since $\iota \in [0, \pi]$ while $\beta_{L} \in [-\pi/2, \pi/2]$, it will be more convenient to work with $\theta_{L} = \pi/2 - \beta_{L}$.
+We will further abbreviate notations by using the variables
+\be
+	t_{\iota} \equiv \tan \frac{\iota}{2} \,, \quad t_{\theta} \equiv \tan \frac{\theta_{L}}{2} \,.
+\ee
+We also introduce the azimuthal angles $\lambda_{a} = \lambda_{L} - \pi/6$, $\lambda_{e} = \lambda_{L} + \pi/12 = \lambda_{a} + \pi/4$, and form the combinations
+\be
+	\sigma_{\pm} \equiv \frac{1}{2} \left[ s_{a}^{22} \pm i s_{e}^{22} \right] \,.
+\ee
+In these notations, we obtain
+\bsub\label{eq:sigmapm}
+\begin{align}
+	\sigma_{+} &= \rho e^{2i\varphi} \left[ t_{\theta}^{4} e^{-2 i \psiL} + t_{\iota}^{4} e^{2 i \psiL} \right] e^{-2i \lambda_{a}} \,, \\
+	\sigma_{-} &= \rho e^{2i\varphi} \left[ e^{-2 i \psiL} + t_{\theta}^{4} t_{\iota}^{4} e^{2 i \psiL} \right] e^{2i \lambda_{a}} \,,
+\end{align}
+\esub
+with a common prefactor \JGB{This seems to need to  be multiplied by -6 to get what I have in my notes. Trace to change in def of $s_{a,e}$?}
+\be\label{eq:sigmafactorrho}
+	\rho(d, \iota, \theta_{L}) = \frac{1}{4d} \sqrt{\frac{5}{\pi}} \frac{1}{\left( 1 + t_{\iota}^{2} \right)^{2} \left(1 + t_{\theta}^{2} \right)^{2}} \,.
+\ee
+Finding points in the parameter space that are degenerate with the injection, i.e.
+with $\ln \calL \simeq 0$, amounts finding choices for the parameters which obtain the same values as the injection for both quantities $\sigma_{+}$ and $\sigma_{-}$.
+
+We can now use the simplified response written in the form~\eqref{eq:sigmapm}-\eqref{eq:sigmafactorrho} to look for symmetries and degeneracies.
+First, we note the following manifest symmetry: changing simultaneously
+\be\label{eq:symmetryresponse}
+	\betaL \rightarrow -\betaL\,, \quad \iota \rightarrow \pi - \iota \,, \quad \psiL \rightarrow \pi - \psiL
+\ee
+leaves the likelihood exactly invariant.
+There is also a symmetry based on exchanging $\iota \leftrightarrow \theta_{L}$.
+While leaving $\sigma_-$ unchanged, this conjugates the factor inside brackets for $\sigma_{+}$, which can be compensated using the phase terms $\varphi \pm \lambda_{a}$.
+If $\Phi = \mathrm{Arg} \left[ t_{\theta}^{4} e^{-2 i \psiL} + t_{\iota}^{4} e^{2 i \psiL} \right]$, we obtain the symmetry
+\begin{align}
+	\iota' = \theta_{L} \,, \qquad \theta_{L}' = \iota \,, \nn\\
+	\varphi' = \varphi + \frac{1}{2} \Phi \quad \mathrm{mod} \; \pi\,, \nn\\
+	\lambda_{a}' = \lambda_{a} - \frac{1}{2} \Phi \quad \mathrm{mod} \; \pi\,.
+\end{align}
+
+\JGB{Don't have to resort to limits.}
+\sout{We can now explain the degenerate features we found as certain limits in this simplified likelihood.
+In the face-on or face-off limit $\iota \rightarrow 0$, $|t_{\iota}| \ll 1$.
+This amounts to the $(\ell,m)=(2,2)$ contribution dominating the $(\ell, m)=(2,-2)$ contribution.
+Neglecting the terms in~\eqref{eq:sigmapm} that are quartic in $t_{\iota}$, we have}
+%\bsub\label{eq:sigmapmapprox}
+%\begin{align}
+%	\sigma_{+} &\simeq \rho e^{2i(\varphi - \psiL)} t_{\theta}^{4} e^{-2i \lambda_{a}} \,, \\
+%	\sigma_{-} &\simeq \rho e^{2i(\varphi - \psiL)} e^{2i \lambda_{a}} \,,
+%\end{align}
+%\esub
+
+Beyond these discrete symetries, since likelihood function dependence on the six extrinsic parameters is funneled through just two complex functions, we should expect a two-dimensional degenerate subspace.
+Indeed we can explicitly find a general solution for parameter values that solve $\sigma_{\pm}((D_{L}, t_\iota, t_\theta \varphi, \lambda_a, \psiL)=\sigma_{\pm,{\rm inj}}$, which we just sketch here, as needed to explain features of the degeneracies.
+
+Defining the ratio
+\be
+	r(\lambda_a, t_\iota, t_\theta, \psiL) = \frac{\sigma_{+}}{\sigma_{-}} \,,
+\ee
+makes clear that $r=r_{\rm inj}$ provides one complex condition on 4 real unknown parameters, eliminating parameters $D_{L}$ and $\varphi$ while retaining the features of the full degenerate subspace.
+Going further, $|r|^2=|r_{\rm inj}|^2$ yields one condition on 3 parameters, eliminating $\lambda_a$.
+In fact, with a little rearrangement, this can be written as a quadratic expression for either $t_\iota^4$ or $t_\theta^4$ for an explicit solution.
+Depending on the value of $|r_{\rm inj}|^2$, values of $\psiL$ and the independent variable ($t_\theta$ or $t_i$) can always be chosen to guarantee a real, positive solution.
+
+Given a solution for $(t_\iota,t_\theta,\psi_L)$, the rest of the solution then proceeds backwards.
+An explicit expression for commensurate $\lambda$ comes from solving $r=r_{\rm inj}$, and then $D_L$ and $\varphi$ are obtained from solving, e.g., $s_+=s_{+,{\rm inj}}$.
+
+In the limit that the independent variable is given a value far from unity (near polar $\iota$ ot $\beta_L$), then the solution for the dependent variable approaches a fixed constant special values $t_{\iota,\theta}^4={|r_{\rm inj}|^{\pm1}}$, with leading order corrections proportional to  $t_{\theta,\iota}^{4}$.
+A large part of the degenerate subspace volume then tends to found with parameters $\iota$ ot $\beta_L$ near these special values with corresponding constraints on $\lambda$ coming from the solution for the complex argument of $r$.
+
+In terms of the original parameters then, if we consider inferences with the parameter $\iota$ that is not near the source equator, then we find a significant part of the degenerate maximum-likelihood parameter space volume near the special sky positions
+\bsub
+\begin{align}
+  \beta_L&\approx{\beta_L}^{*}=\pm\left(\frac{\pi}{2} - 2 \mathrm{Arctan} \; \left| r_{\rm inj}\right|^{1/4} \right)\,, \\
+  \lambda_a&\approx {\lambda_a}^{*}=\frac{\pi}{6} -\frac{1}{4} \mathrm{Arg} \; r_{\rm inj} \;\mathrm{mod}\; \frac{\pi}{2} \,,\\
+  \rho e^{2i(\varphi-\psiL)} &\approx e^{-2i\lambda_{a}^{*}} \sigma_{-}^{\rm inj} \,.
+\end{align}
+\esub
+With the $\lambdaL \rightarrow \lambdaL + k \pi/2$ and $\betaL \rightarrow - \betaL$ symmetries, we arrive at eight sky positions where the degeneracy occurs, with an exact degeneracy for the pair of parameters $(\varphi, \psiL)$, as well as an approximate degeneracy for the pair $(d, \iota)$ as long as we remain in the regime $t_{\iota}^{4} \ll 1$ or $t_{\iota}^{4} \gg 1$ (which is quite extended, thanks to the quartic power).
+For each of these sky positions built from~\eqref{eq:soldegensky}, many different values of $(\varphi, \psiL)$ and $(d, \iota)$ produce a waveform very close to the injection, resulting in apparent peaks in the marginalized posterior distribution for the sky position, located at the special sky positions $(\lambda_{a}^{*}, \theta_{L}^{*})$, which are offset from the injected value.
+
+Without special tuning, the injected sky position which led to the given values of $s_{a,\mathrm{inj}}$ and $s_{e,\mathrm{inj}}$ is correspondingly likely to have been near one of these sky locations, but it may not have been.
+In those cases the posterior\ldots
+
+Similarly, considering inferred locations away from the LISA-constellation plane, we find a significant part of the degenerate parameter space near special values for inclination and correspondingly for $\lambda_L+\varphi$
+\bsub
+\begin{align}
+	\iota&\approx\iota^{*} = 2 \mathrm{Arctan} \; \left| r_{\rm inj}\right|^{1/4} \,, \\
+	\lambda_a-\psiL&\approx (\lambda_{a} - \psiL)^{*} = -\frac{1}{4} \mathrm{Arg} \; r_{\rm inj} \;\mathrm{mod}\; \frac{\pi}{2} \,\\
+        \rho e^{2i\varphi} &\approx e^{-2i (\lambda_{a} - \psi_{L})^{*}} \sigma_{-}^{\rm inj} \,.
+\end{align}
+\esub
+
+%finding $(\lambda_{a}^{*}, \theta_{L}^{*})$ verifying
+%\be\label{eq:degenconditionsky}
+%	\frac{\sigma_{+}}{\sigma_{-}} (\lambda_{a}^{*}, \theta_{L}^{*}) = r_{\rm inj} \,,
+%\ee
+%with the approximation~\eqref{eq:sigmapmapprox} for the left-hand side, guarantees that we can then match both $\sigma_{+}$ and $\sigma_{-}$ of the injection with the condition
+%\be\label{eq:degenconditionprefactor}
+%	\rho e^{2i(\varphi - \psiL)} e^{2i \lambda_{a}^{*}} = \sigma_{-}^{\rm inj}\,.
+%\ee
+%In the limit $|t_{\iota}| \ll 1$,
+%\be
+%	\frac{\sigma_{+}}{\sigma_{-}} \simeq t_{\theta}^{4} e^{-4i\lambda_{a}} \,,
+%\ee
+%so that~\eqref{eq:degenconditionsky} is verified for special positions in the sky at (coming back to the original variables $(\lambdaL, \betaL)$)
+%\bsub\label{eq:soldegensky}
+%\begin{align}
+%\beta_{L}^{*} &= \frac{\pi}{2} - 2 \mathrm{Arctan} \; \left| r_{\rm inj}\right|^{1/4} \,, \\
+%	\lambda_{L}^{*} &= \frac{\pi}{6} -\frac{1}{4} \mathrm{Arg} \; r_{\rm inj} \;\mathrm{mod}\; \frac{\pi}{2} \,.
+%\end{align}
+%\esub
+\JGB{Keep something from this?}
+The other condition~\eqref{eq:conditionprefactor} gives then a single phase constraint for $\varphi - \psiL$, and a single amplitude constraint for $\rho(d, \iota, \theta_{L}^{*})$.
+The limit $\iota \rightarrow \pi$ (where the $(2,-2)$ contribution dominates the $(2,2)$) is very similar, as can be seen most easily by the change of variable $\iota \rightarrow \pi - \iota$, with $-\psiL$ and $\pi - \theta_{L}$ playing the roles of $\psiL$ and $\theta_{L}$.
+
+%We can further find the same mechanism, with a few differences, by exchanging the roles of $(\iota, \varphi)$ and $(\theta_{L}, \lambda_{a})$, as follows.
+In the limit $\theta_{L} \rightarrow 0$ with $|t_{\theta}| \ll 1$, we have this time
+%\be
+%	\frac{\sigma_{+}}{\sigma_{-}} \simeq t_{\iota}^{4} e^{-4i(\lambda_{a} - \psi_{L})} \,,
+%\ee
+%which gives in the same way a degeneracy if
+%\bsub\label{eq:soldegeniotaphi}
+%\begin{align}
+%	\iota^{*} &= 2 \mathrm{Arctan} \; \left| r_{\rm inj}\right|^{1/4} \,, \\
+%	(\lambda_{a} - \psiL)^{*} &= -\frac{1}{4} \mathrm{Arg} \; r_{\rm inj} \;\mathrm{mod}\; \frac{\pi}{2} \,,
+%\end{align}
+%\esub
+%with a condition on $\rho$ and $\varphi$ that is now
+%\be
+%\rho e^{2i\varphi} e^{2i (\lambda_{a}^{*} - \psi_{L}^{*})} = \sigma_{-}^{\rm inj} \,.
+%\ee
+\JGB{Keep something from this?}
+Thus, we find again a degenerate pattern, with the difference that \eqref{eq:soldegeniotaphi} imposes a condition on the pair $\lambda_{a} - \psiL$ instead of $\lambda_{a}$, and the extra condition imposes a condition on $\varphi$ instead of the pair $\varphi - \psiL$.
+
+When considering more harmonics beyond the dominant mode $h_{22}$, such degeneracies will be broken easily.
+As in the case of the full likelihood, the simplified likelihood~\eqref{eq:simplelikelihoodhm} will have several terms with different inclination and phase dependencies.
+If the signal is loud enough for at least two modes to be detected, then the inclination $\iota$ and the phase $\varphi$ are fixed by the relative amplitude and phase of these modes.
+This will break the degeneracies $(d, \iota)$ and the degeneracy $(\varphi, \psiL)$.
+
+We illustrate our findings by running a Bayesian parameter estimation for the simplified likelihood~\eqref{eq:simplelikelihood22} with the two samplers \texttt{MultiNest} and \texttt{PTMCMC} described in Sec.~\ref{}.
+Since the inner products~\eqref{eq:innerproductlmlpmp} are constants computed only once, the external factors become trivially inexpensive, and the likelihood is very fast to compute.
+We take advantage of this to run \texttt{PTMCMC} for a very large number of steps, ensuring that we do not miss degenerate regions of the parameter space.
+The resulting posterior distribution is extremely degenerate, as shown in Fig.~\ref{fig:PEsmbh22hmSimpleLikeCase9}.
+The sampler \texttt{MultiNest} fails to resolve this structure, stopping after having only explored the degenerate region $\iota \rightarrow 0$ and $\iota \rightarrow \pi$ and the eight sky positions~\eqref{eq:soldegensky}.
+In accordance with~\eqref{eq:degenconditionsky}-\eqref{eq:degenconditionprefactor}, this corresponds to lines in $\varphi \pm \psiL = \mathrm{const}$.
+The sampler \texttt{PTMCMC} finds more degenerate regions, in particular it finds similar structures with the roles of $\iota$ and $\betaL$ exchanged (as shown notably by the $(d,\iota)$ and $(d,\betaL)$ panels), as well as intermediate regions connecting all these features.
+
+Fig.~\ref{fig:PEsmbh22hmSimpleLikeCase9} also shows the result obtained when including all available harmonics~\eqref{} in~\eqref{eq:simplelikelihoodhm}.
+The degenerate structures collapse to leave only eight modes corresponding to $\betaL \rightarrow - \betaL$ and $\lambdaL \rightarrow \lambdaL + k\pi/2$.
+We found that this degeneracy-breaking occurs already when including a single subdominant harmonic $h_{21}$.
+
+We can now come back to explaining the results of Sec.~\ref{}, and in particular Fig.~\ref{}, obtained with the true likelihood function for the mode $h_{22}$.
+First, the difference between the two Bayesian samplers has been made more extreme by going to the more degenerate likelihood~\eqref{eq:simplelikelihood22}, allowing us to attribute their disagreement to an apparent shortcoming of \texttt{MultiNest}\footnote{We recall that we only used standard values for the metaparameters \SM{[complete]} for this sampler, and we do not exclude that it would be possible to improve results with minor changes.} for resolving extended degenerate regions.
+Focusing on the sky position, we zoom close to the injection in Fig.~\ref{fig:Skydownsmbh22hmSimpleLikeCase9}.
+We compare here the posterior distribution generated with the true likelihood with the one obtained with the simplified likelihood~\eqref{eq:simplelikelihood22}, for both samplers.
+We also superimpose the prediction~\eqref{eq:soldegensky} for the degeneracy-induced peak.
+We see that, although ,
+
+%\be\label{eq:symmetryresponse}
+%\betaL \rightarrow -\betaL\,, \quad \iota \rightarrow \pi - \iota \,, \quad \psiL \rightarrow \pi - \psiL
+%\ee
+%leaves the signal unchanged. We can also rewrite the likelihood in a way that makes clearer the symmetry between the two pairs of parameters $(\iota, \varphiL)$ and $(\betaL, \lambdaL)$ \SM{[TODO]}.
+%
+%The simplified extrinsic likelihood for a frozen LISA then takes the form
+%\be\label{eq:simplelikelihood}
+%	\ln \calL = - \frac{1}{2} \Lambda \left( \left| s_{a} - s_{a}^{\rm inj} \right|^{2} + \left| s_{e} - s_{e}^{\rm inj} \right|^{2} \right) \,,
+%\ee
+%where $\Lambda$ appears as an overall normalization factor, playing the role of the square SNR of the signal.
+
+%We can now explore possible degeneracies in the likelihood given by~\eqref{eq:simplelikelihood}, \eqref{eq:defsase} and \eqref{eq:defa22a2m2}-\eqref{eq:defe22e2m2}.
+We note that the relative contributions of the modes $h_{2,2}$ and $h_{2,-2}$ are strongly separated in the face-on or face-off limit, leading to a simplification of the response and an enhanced degeneracy.
+Due to the symmetry~\eqref{eq:symmetryresponse}, we will focus on the face-on limit $\iota \rightarrow 0$.
+In this limit, the $h_{2,-2}$ contributions die off rapidly as $a_{2,-2}\,, e_{2,-2} \sim \sin^{4} (\iota/2)$.
+If they are assumed to be negligible, then we can rewrite
+%\bsub\label{eq:saseapprox}
+%\begin{align}
+%	s_{a} \simeq i\calA e^{2 i \xi} \left( D_{a}^{+} + i D_{a}^{\times} \right) \,, \\
+%	s_{e} \simeq i\calA e^{2 i \xi} \left( D_{e}^{+} + i D_{e}^{\times} \right) \,,
+%\end{align}
+%\esub
+%where we introduced the notations $\xi (\varphiL, \psiL) \equiv -\varphiL - \psiL$ and $\calA(d, \iota) \equiv 3/4\sqrt{5/\pi}\cos^{4}(\iota/2)/d $ for the common phase and amplitude factors.
+%
+%For a degeneracy to occur, both complex quantities $s_{a}$ and $s_{e}$ must be close to the injection values $s_{a}^{\rm inj}$ and $s_{e}^{\rm inj}$, which can be achieved as follows if~\eqref{eq:saseapprox} is valid.
+Defining the complex ratio
+%\be\label{eq:defr}
+%	r(\lambdaL, \betaL) = \frac{D_{a}^{+} + i D_{a}^{\times}}{D_{e}^{+} + i D_{e}^{\times} } \,,
+%\ee
+%if the position in the sky $(\lambda_{L}^{*}, \beta_{L}^{*})$ can be chosen so that
+%\be\label{eq:rcondition}
+%	r(\lambda_{L}^{*}, \beta_{L}^{*}) = \frac{s_{a}^{\rm inj}}{s_{e}^{\rm inj}} \,,
+%\ee
+%then adjusting $\calA$ and $\xi$ so that
+%\be
+%	i\calA e^{2 i \xi} = \frac{s_{a}^{\rm inj}}{\left( D_{a}^{+} + i D_{a}^{\times} \right)(\lambda_{L}^{*}, \beta_{L}^{*})}
+%\ee
+%generates degenerate points in parameter space with $\ln \calL \simeq 0$.
+%
+%The conditions~\eqref{eq:defr}-\eqref{eq:rcondition} can be inverted for the sky position as follows.
+For a given complex value of $r = s_{a}^{\rm inj} / s_{e}^{\rm inj}$, $r(\lambda_{L}^{*}, \beta_{L}^{*}) = r$ can be rewritten as
+%\be\label{eq:eqtosolveforlambdabetaL}
+%	e^{4i\lambda_{L}^{*}}e^{i\frac{\pi}{3}} \frac{\left( 1 +\sin\beta_{L}^{*} \right)^{2}}{\left( 1 - \sin\beta_{L}^{*} \right)^{2}} = \frac{1+ i r}{1 - i r} \,.
+%\ee
+%Solving for the modulus of~\eqref{eq:eqtosolveforlambdabetaL} then yields
+%\be\label{eq:solutionbetaL}
+%	\sin\beta_{L}^{*} = \frac{\rho - 1}{\rho + 1} \,,
+%\ee
+%for
+%\be
+%	\rho = \sqrt{\left| \frac{1+ i r}{1 - i r} \right|} \,.
+%\ee
+%Solving for the argument of~\eqref{eq:eqtosolveforlambdabetaL} gives four solutions for $\lambda_{L}^{*}$ as
+%\be\label{eq:solutionlambdaL}
+%	\lambda_{L}^{*} = - \frac{\pi}{12} + \frac{1}{4}\mathrm{Arg} \frac{1+ i r}{1 - i r} + \frac{k \pi}{2} \quad \mathrm{for} \quad k = 0,\dots,3 \,.
+%\ee
+%
+%Repeating the argument for the other branch of solutions $\iota \rightarrow \pi$ (or directly considering the symmery~\eqref{eq:symmetryresponse}) gives four other solutions in the sky with identical sky longitudes $\lambda_{L}^{*}$, with symmetrized sky latitudes $\beta_{L}^{*} \rightarrow -\beta_{L}^{*}$, and with $\xi (\varphiL, \psiL) \equiv -\varphiL + \psiL$, $\calA(d, \iota) \equiv 3/4\sqrt{5/\pi}\sin^{4}(\iota/2)/d $.
+%
+%Thus, with~\eqref{eq:solutionbetaL}, \eqref{eq:solutionlambdaL} and $\beta_{L}^{*} \rightarrow -\beta_{L}^{*}$, we arrive at eight degenerate positions in the sky $(\lambda_{L}^{*}, \beta_{L}^{*})$.
+For each of these sky positions, an exact line degeneracy exists between $\varphiL$ and $\psiL$ as long as $\xi = \mathrm{const}$, as well as an approximate line degeneracy between $d$ and $\iota$, as long as $\calA = \mathrm{const}$, limited to inclination values where $\sin^{4}(\iota/2)$ (or $\cos^{4}(\iota/2)$, for the other branch of solutions) remains negligible.
+The combination of these degeneracies generates an extended structure in the multi-dimensional parameter space.
+As a result, when considering the marginal posterior for the sky position, these degenerate spots in the sky can have a lot of probability support.
+
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+
+\section{Massive black holes: accumulation of information with time}
+\label{sec:MBHBPEacctime}
+
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+\subsection{Pre-merger analysis}
+\label{subsec:MBHBacctimePE}
+
+The rate at which parameter information accumulates LISA merger observations is crucial for planning strategies for multi-messenger observations and in guiding the latency demands of the LISA down-link and data processing.
+We explore how parameter information accumulates on approach to merger by performing parameter estimation studies temporally truncated signals, cut at points where the accumulated SNR is about $\{10,42,166\}$ corresponding to $\{1/64,1/16,1/4\}$ of the total SNR=$666.0$, and times $\{41 \mathrm{hr}, 2.5 \mathrm{hr}, 7 \mathrm{min}\}$ before merger.
 
@@
    \end{minipage}
-  \caption{Relations of time, Fourier frequency, and accumulated SNR for $M=2\times 10^6$,$q=3$ merger at $z=4$.
+  \caption{\SM{Redo left plot, right plot redundant with mode-by-mode SNR} Relations of time, Fourier frequency, and accumulated SNR for $M=2\times 10^6$,$q=3$ merger at $z=4$.
 The left panel shows times computed from $\tilde h(f)$ using Eq.~eref{eq:t-of-f}, and are approximately common among the waveform component modes up to minutes before merger.
@@
 \end{figure*}
-
-The rate at which parameter information accumulates LISA merger observations is crucial for planning strategies for multi-messenger observations and in guiding the latency demands of the LISA down-link and data processing.
-We explore how parameter information accumulates on approach to merger by performing parameter estimation studies temporally truncated signals, cut at points where the accumulated SNR is about $\{10,42,166\}$ corresponding to $\{1/64,1/16,1/4\}$ of the total SNR=$666.0$, and times $\{41 \mathrm{hr}, 2.5 \mathrm{hr}, 7 \mathrm{min}\}$ before merger.
 
@@
   \centering
-  \includegraphics[width=.8\linewidth]{../plots/plot_mollweide_tseries.png}
+  \includegraphics[width=.8\linewidth]{../plots/skymap_Lframe_tseries_bambi_22.png}
+  \includegraphics[width=.8\linewidth]{../plots/skymap_Lframe_tseries_ptmcmc_hm.png}
   \caption{.}
@@
 %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-
-\subsection{Investigating degeneracies}
-\label{subsec:SMBHPEdegen}
+\subsection{Degeneracy breaking by the time and frequency dependence in the response}
+\label{subsec:MBHBacctimebreakdegen}
 
@@
   \centering
-  \includegraphics[width=.8\linewidth]{../plots/corner_smbh_case9_simplelike_ptmcmc_bambi.png}
-  \caption{Posterior distributions obtained with the two samplers \texttt{PTMCMC} (blue) and $\texttt{MultiNest}$ (green) for the simplified likelihood~\eqref{eq:simplelikelihoodhm}, using only the dominant mode $(\ell, m) = (2,2)$.
-Masses and time are pinned to the injected value, and the motion of the detector is neglected.
-We also superimpose (red) the contours obtained with \texttt{MultiNest} by including all harmonics $(2,2)$, $(2,1)$, $(3,3)$, $(4,4)$, $(5,5)$.}
-  \label{fig:PEsmbh22hmSimpleLikeCase9}
+  \includegraphics[width=.99\linewidth]{../plots/lnL_skymodes_approx_22.pdf}
+  \includegraphics[width=.99\linewidth]{../plots/lnL_skymodes_approx_hm.pdf}
+  \caption{for different response approximations (from left to right: \textit{Full}, \textit{Frozen}, \textit{Low-$f$}), with 22-mode only (top) and with higher harmonics (bottom).
+A very negative $\ln\calL$ values means that this point is excluded by the inference.
+\SM{do we need a table of post-merger lnL values ?}}
+  \label{fig:lnLskymodesapprox}
 \end{figure*}
 
-\begin{figure}
-  \centering
-  \begin{minipage}{.49\linewidth}
-  \includegraphics[width=.99\linewidth]{../plots/corner_skydown_smbh_case9_22_ptmcmc_bambi_withdegen.png}
-  \end{minipage}
-  \begin{minipage}{.49\linewidth}
-  \includegraphics[width=.99\linewidth]{../plots/corner_skydown_smbh_case9_simplelike_ptmcmc_bambi_withdegen.png}
-  \end{minipage}
-  \caption{Sky position posterior in presence of degeneracies, zooming close to the injected value.
-The left panel shows the complete likelihood, while the right panel shows the simplified likelihood~\eqref{eq:simplelikelihoodhm} with a motionless LISA and fixed masses and time.
-The results of \texttt{PTMCMC} (blue) and \texttt{MultiNest} are superimposed.
-The black cross is the injected value, and the grey cross is the analytic prediction~\eqref{eq:soldegensky} for the offset peak created by degeneracies.}
-  \label{fig:Skydownsmbh22hmSimpleLikeCase9}
-\end{figure}
-
-In this section, we investigate the cause of the degeneracies appearing in the posterior of Case~II shown in~\ref{subsec:SMBHPE}.
-Understanding the structure of these features could prove useful to inform Bayesian samplers, for instance by building adapted jump proposal allowing a much faster mixing of MCMC chains than naive methods.
-a simplified form of the extrinsic likelihood and the structure of its degeneracies.
-
-An important clue about the appearance of these degeneracies is the length of the observable signal.
-As shown in Fig.~\ref{}, for our example SMBH systems the SNR accumulates in a matter of days before merger, which means that there is only little impact of the modulations caused by the motion of the detector on the TDI transfer functions~\eqref{}.
-If we can neglect entirely the motion of the detector (which is not quite true as we will see later on), then we are in a regime where the information about the sky position of the source comes essentially from the pattern functions affecting this high-SNR signal.
-Using the low-frequency approximation (see~\ref{}) and the time-at-LISA $t_{L}$ as a time variable, there is no frequency dependence in the response.
-A further simplification occurs when pinning the masses and the time of coalescence to their true values.
-We found indeed that these parameters correlate only weakly to the other ones, so that we can hope to understand the degeneracy features by restricting the analysis to the six extrinsic parameters $(D_{L}, \iota, \varphi, \lambdaL, \betaL, \psiL)$.
-\TDC{we could make the connection with a similar decoupling in LIGO/Virgo, which is actually what makes low-latency sky localization possible.}
-
-Under all these simplifying assumptions, the likelihood function becomes purely extrinsic, and takes the form of trivial geometric factors multiplying a constant overlap that can be precomputed.
-Using the subscript $\mathrm{inj}$ for the injected signal, the likelihood~\eqref{eq:defliklihood} is
-\be
-	\ln \calL = -\frac{1}{2} \left( \tilde{a} - \tilde{a}_{\rm inj} | \tilde{a} - \tilde{a}_{\rm inj} \right) -\frac{1}{2} \left( \tilde{e} - \tilde{e}_{\rm inj} | \tilde{e} - \tilde{e}_{\rm inj} \right) \,.
-\ee
-From the low-frequency response \eqref{eq:aelowfmodes}, pinning the time-at-LISA~\eqref{} \SM{to be completed for $t_{L}$} so that we can ignore the factor $\exp [2 i \pi f k\cdot p_{0}]$ as a pure constant corresponding to a redefinition of time, we obtain \JGB{Should we use notation $\Phi_R$ here and in \eqref{eq:aelowfmodes}?}
-\begin{align}\label{eq:simplelikelihoodhm}
-	\ln \calL &= -\frac{1}{2} \sum_{\ell m} \sum_{\ell' m'}   \left[ \left( s_{a}^{\ell m} - s_{a, \mathrm{inj}}^{\ell m}\right) \left( s_{a}^{\ell' m'} - s_{a, \mathrm{inj}}^{\ell' m'}\right)^{*} \right.
-\nn\\
-	& \qquad \left.
-+ \left( s_{e}^{\ell m} - s_{e, \mathrm{inj}}^{\ell m}\right) \left( s_{e}^{\ell' m'} - s_{e, \mathrm{inj}}^{\ell' m'}\right)^{*} \right] \left\langle \ell m | \ell' m' \right\rangle \,,
-\end{align}
-where we introduced the following extrinsic signal functions \JGB{Since neither $D_{a,e}$ nor $P_{\ell m}$ are defined in the main text, may be better to write in terms of a version of $F$ as in \eqref{eq:defFapcFepc}}
-\be\label{eq:defsaelm}
-	s_{a,e}^{\ell m} = \frac{1}{d} \frac{2}{3L} D_{a,e} : P_{\ell m} \,,
-\ee
-with $d = D / D^{\rm inj}$ the dimensionless ratio of luminosity distances.
-The prefactors $\left\langle \ell m | \ell' m' \right\rangle$ \sout{are pure normalization constants, and} represent mode overlaps as
-\be\label{eq:innerproductlmlpmp}
-	\left\langle \ell m | \ell' m' \right\rangle = 4\mathrm{Re} \int \frac{df}{S_{n}^{a,e}} \left( 3 \pi f L\right)^{2} \tilde{h}_{\ell m} \tilde{h}_{\ell' m'}^{*} \,.
-\ee
-Here, the noise PSD $S_{n}^{a,e}$ is given by~\eqref{} and is identical between the channels $a$ and $e$.
-Having fixed the masses and the time, these overlaps are to be computed with the modes $\tilde{h}_{\ell m}^{\rm inj}$ of the injected signal.
-
-In the case where we only include the dominant harmonic $h_{22}$, the likelihood~\eqref{eq:simplelikelihoodhm} simplifies to
-\be\label{eq:simplelikelihood22}
-	\ln \calL = -\frac{1}{2} \left\langle 22 | 22 \right\rangle  \left[ \left| s_{a}^{22} - s_{a, \mathrm{inj}}^{22} \right|^{2} + \left| s_{e}^{22} - s_{e, \mathrm{inj}}^{22}\right|^{2} \right] \,.
-\ee
-The functions~\eqref{eq:defsaelm} take the explicit form
-\begin{align}
-	s_{a,e}^{22} &= \frac{1}{4d} \sqrt{\frac{5}{\pi}} \cos^{4}\frac{\iota}{2} e^{2i(\varphi-\psiL)} \left( F_{a,e}^{+} + i F_{a,e}^{\times} \right) \\
-	&+ \frac{1}{4d} \sqrt{\frac{5}{\pi}} \sin^{4}\frac{\iota}{2} e^{2i(\varphi+\psiL)} \left( F_{a,e}^{+} - i F_{a,e}^{\times} \right) \,,
-\end{align}
-with the pattern functions for the channels $a$ and $e$ defined in \eqref{eq:defFapcFepc}-\eqref{eq:FapcFepc}.
-
-In order to make more manifest the degeneracies in the problem, it will be useful to introduce the following notations.
-First, since $\iota \in [0, \pi]$ while $\beta_{L} \in [-\pi/2, \pi/2]$, it will be more convenient to work with $\theta_{L} = \pi/2 - \beta_{L}$.
-We will further abbreviate notations by using the variables
-\be
-	t_{\iota} \equiv \tan \frac{\iota}{2} \,, \quad t_{\theta} \equiv \tan \frac{\theta_{L}}{2} \,.
-\ee
-We also introduce the azimuthal angles $\lambda_{a} = \lambda_{L} - \pi/6$, $\lambda_{e} = \lambda_{L} + \pi/12 = \lambda_{a} + \pi/4$, and form the combinations
-\be
-	\sigma_{\pm} \equiv s_{a}^{22} \pm i s_{e}^{22} \,.
-\ee
-In these notations, we obtain
-\bsub\label{eq:sigmapm}
-\begin{align}
-	\sigma_{+} &= \rho e^{2i\varphi} \left[ t_{\theta}^{4} e^{-2 i \psiL} + t_{\iota}^{4} e^{2 i \psiL} \right] e^{-2i \lambda_{a}} \,, \\
-	\sigma_{-} &= \rho e^{2i\varphi} \left[ e^{-2 i \psiL} + t_{\theta}^{4} t_{\iota}^{4} e^{2 i \psiL} \right] e^{2i \lambda_{a}} \,,
-\end{align}
-\esub
-with a common prefactor \JGB{This seems to need to  be multiplied by -6 to get what I have in my notes. Trace to change in def of $s_{a,e}$?}
-\be\label{eq:sigmafactorrho}
-	\rho(d, \iota, \theta_{L}) = \frac{1}{4d} \sqrt{\frac{5}{\pi}} \frac{1}{\left( 1 + t_{\iota}^{2} \right)^{2} \left(1 + t_{\theta}^{2} \right)^{2}} \,.
-\ee
-Finding points in the parameter space that are degenerate with the injection, i.e.
-with $\ln \calL \simeq 0$, amounts finding choices for the parameters which obtain the same values for both  $\sigma_{+}$ and $\sigma_{-}$ as the injection.
-
-We can now use the simplified response written in the form~\eqref{eq:sigmapm}-\eqref{eq:sigmafactorrho} to look for symmetries and degeneracies.
-First, we note the following manifest symmetry: changing simultaneously
-\be\label{eq:symmetryresponse}
-	\betaL \rightarrow -\betaL\,, \quad \iota \rightarrow \pi - \iota \,, \quad \psiL \rightarrow \pi - \psiL
-\ee
-leaves the likelihood exactly invariant.
-There is also a symmetry based on exchanging $\iota \leftrightarrow \theta_{L}$.
-While leaving $\sigma_-$ unchanged, this conjugates the factor inside brackets for $\sigma_{+}$, which can be compensated using the phase terms $\varphi \pm \lambda_{a}$.
-If $\Phi = \mathrm{Arg} \left[ t_{\theta}^{4} e^{-2 i \psiL} + t_{\iota}^{4} e^{2 i \psiL} \right]$, we obtain the symmetry
-\begin{align}
-	\iota' = \theta_{L} \,, \qquad \theta_{L}' = \iota \,, \nn\\
-	\varphi' = \varphi + \frac{1}{2} \Phi \quad \mathrm{mod} \; \pi\,, \nn\\
-	\lambda_{a}' = \lambda_{a} - \frac{1}{2} \Phi \quad \mathrm{mod} \; \pi\,.
-\end{align}
-
-\JGB{Don't have to resort to limits.}
-\sout{We can now explain the degenerate features we found as certain limits in this simplified likelihood.
-In the face-on or face-off limit $\iota \rightarrow 0$, $|t_{\iota}| \ll 1$.
-This amounts to the $(\ell,m)=(2,2)$ contribution dominating the $(\ell, m)=(2,-2)$ contribution.
-Neglecting the terms in~\eqref{eq:sigmapm} that are quartic in $t_{\iota}$, we have}
-%\bsub\label{eq:sigmapmapprox}
-%\begin{align}
-%	\sigma_{+} &\simeq \rho e^{2i(\varphi - \psiL)} t_{\theta}^{4} e^{-2i \lambda_{a}} \,, \\
-%	\sigma_{-} &\simeq \rho e^{2i(\varphi - \psiL)} e^{2i \lambda_{a}} \,,
-%\end{align}
-%\esub
-
-Beyond these discrete symetries, since likelihood function dependence on the six extrinsic parameters is funneled through just two complex functions, we should expect a two-dimensional degenerate subspace.
-Indeed we can explicitly find a general solution for parameter values that solve $\sigma_{\pm}((D_{L}, t_\iota, t_\theta \varphi, \lambda_a, \psiL)=\sigma_{\pm,{\rm inj}}$, which we just sketch here, as needed to explain features of the degeneracies.
-
-Defining the ratio
-\be
-	r(\lambda_a, t_\iota, t_\theta, \psiL) = \frac{\sigma_{+}}{\sigma_{-}} \,,
-\ee
-makes clear that $r=r_{\rm inj}$ provides one complex condition on 4 real unknown parameters, eliminating parameters $D_{L}$ and $\varphi$ while retaining the features of the full degenerate subspace.
-Going further, $|r|^2=|r_{\rm inj}|^2$ yields one condition on 3 parameters, eliminating $\lambda_a$.
-In fact, with a little rearrangement, this can be written as a quadratic expression for either $t_\iota^4$ or $t_\theta^4$ for an explicit solution.
-Depending on the value of $|r_{\rm inj}|^2$, values of $\psiL$ and the independent variable ($t_\theta$ or $t_i$) can always be chosen to guarantee a real, positive solution.
-
-Given a solution for $(t_\iota,t_\theta,\psi_L)$, the rest of the solution then proceeds backwards.
-An explicit expression for commensurate $\lambda$ comes from solving $r=r_{\rm inj}$, and then $D_L$ and $\varphi$ are obtained from solving, e.g., $s_+=s_{+,{\rm inj}}$.
-
-In the limit that the independent variable is given a value far from unity (near polar $\iota$ ot $\beta_L$), then the solution for the dependent variable approaches a fixed constant special values $t_{\iota,\theta}^4={|r_{\rm inj}|^{\pm1}}$, with leading order corrections proportional to  $t_{\theta,\iota}^{4}$.
-A large part of the degenerate subspace volume then tends to found with parameters $\iota$ ot $\beta_L$ near these special values with corresponding constraints on $\lambda$ coming from the solution for the complex argument of $r$.
-
-In terms of the original parameters then, if we consider inferences with the parameter $\iota$ that is not near the source equator, then we find a significant part of the degenerate maximum-likelihood parameter space volume near the special sky positions
-\bsub
-\begin{align}
-  \beta_L&\approx{\beta_L}^{*}=\pm\left(\frac{\pi}{2} - 2 \mathrm{Arctan} \; \left| r_{\rm inj}\right|^{1/4} \right)\,, \\
-  \lambda_a&\approx {\lambda_a}^{*}=\frac{\pi}{6} -\frac{1}{4} \mathrm{Arg} \; r_{\rm inj} \;\mathrm{mod}\; \frac{\pi}{2} \,,\\
-  \rho e^{2i(\varphi-\psiL)} &\approx e^{-2i\lambda_{a}^{*}} \sigma_{-}^{\rm inj} \,.
-\end{align}
-\esub
-With the $\lambdaL \rightarrow \lambdaL + k \pi/2$ and $\betaL \rightarrow - \betaL$ symmetries, we arrive at eight sky positions where the degeneracy occurs, with an exact degeneracy for the pair of parameters $(\varphi, \psiL)$, as well as an approximate degeneracy for the pair $(d, \iota)$ as long as we remain in the regime $t_{\iota}^{4} \ll 1$ or $t_{\iota}^{4} \gg 1$ (which is quite extended, thanks to the quartic power).
-For each of these sky positions built from~\eqref{eq:soldegensky}, many different values of $(\varphi, \psiL)$ and $(d, \iota)$ produce a waveform very close to the injection, resulting in apparent peaks in the marginalized posterior distribution for the sky position, located at the special sky positions $(\lambda_{a}^{*}, \theta_{L}^{*})$, which are offset from the injected value.
-
-Without special tuning, the injected sky position which led to the given values of $s_{a,\mathrm{inj}}$ and $s_{e,\mathrm{inj}}$ is correspondingly likely to have been near one of these sky locations, but it may not have been.
-In those cases the posterior\ldots
-
-Similarly, considering inferred locations away from the LISA-constellation plane, we find a significant part of the degenerate parameter space near special values for inclination and correspondingly for $\lambda_L+\varphi$
-\bsub
-\begin{align}
-	\iota&\approx\iota^{*} = 2 \mathrm{Arctan} \; \left| r_{\rm inj}\right|^{1/4} \,, \\
-	\lambda_a-\psiL&\approx (\lambda_{a} - \psiL)^{*} = -\frac{1}{4} \mathrm{Arg} \; r_{\rm inj} \;\mathrm{mod}\; \frac{\pi}{2} \,\\
-        \rho e^{2i\varphi} &\approx e^{-2i (\lambda_{a} - \psi_{L})^{*}} \sigma_{-}^{\rm inj} \,.
-\end{align}
-\esub
-
-%finding $(\lambda_{a}^{*}, \theta_{L}^{*})$ verifying
-%\be\label{eq:degenconditionsky}
-%	\frac{\sigma_{+}}{\sigma_{-}} (\lambda_{a}^{*}, \theta_{L}^{*}) = r_{\rm inj} \,,
-%\ee
-%with the approximation~\eqref{eq:sigmapmapprox} for the left-hand side, guarantees that we can then match both $\sigma_{+}$ and $\sigma_{-}$ of the injection with the condition
-%\be\label{eq:degenconditionprefactor}
-%	\rho e^{2i(\varphi - \psiL)} e^{2i \lambda_{a}^{*}} = \sigma_{-}^{\rm inj}\,.
-%\ee
-%In the limit $|t_{\iota}| \ll 1$,
-%\be
-%	\frac{\sigma_{+}}{\sigma_{-}} \simeq t_{\theta}^{4} e^{-4i\lambda_{a}} \,,
-%\ee
-%so that~\eqref{eq:degenconditionsky} is verified for special positions in the sky at (coming back to the original variables $(\lambdaL, \betaL)$)
-%\bsub\label{eq:soldegensky}
-%\begin{align}
-%\beta_{L}^{*} &= \frac{\pi}{2} - 2 \mathrm{Arctan} \; \left| r_{\rm inj}\right|^{1/4} \,, \\
-%	\lambda_{L}^{*} &= \frac{\pi}{6} -\frac{1}{4} \mathrm{Arg} \; r_{\rm inj} \;\mathrm{mod}\; \frac{\pi}{2} \,.
-%\end{align}
-%\esub
-\JGB{Keep something from this?}
-The other condition~\eqref{eq:conditionprefactor} gives then a single phase constraint for $\varphi - \psiL$, and a single amplitude constraint for $\rho(d, \iota, \theta_{L}^{*})$.
-The limit $\iota \rightarrow \pi$ (where the $(2,-2)$ contribution dominates the $(2,2)$) is very similar, as can be seen most easily by the change of variable $\iota \rightarrow \pi - \iota$, with $-\psiL$ and $\pi - \theta_{L}$ playing the roles of $\psiL$ and $\theta_{L}$.
-
-%We can further find the same mechanism, with a few differences, by exchanging the roles of $(\iota, \varphi)$ and $(\theta_{L}, \lambda_{a})$, as follows.
-In the limit $\theta_{L} \rightarrow 0$ with $|t_{\theta}| \ll 1$, we have this time
-%\be
-%	\frac{\sigma_{+}}{\sigma_{-}} \simeq t_{\iota}^{4} e^{-4i(\lambda_{a} - \psi_{L})} \,,
-%\ee
-%which gives in the same way a degeneracy if
-%\bsub\label{eq:soldegeniotaphi}
-%\begin{align}
-%	\iota^{*} &= 2 \mathrm{Arctan} \; \left| r_{\rm inj}\right|^{1/4} \,, \\
-%	(\lambda_{a} - \psiL)^{*} &= -\frac{1}{4} \mathrm{Arg} \; r_{\rm inj} \;\mathrm{mod}\; \frac{\pi}{2} \,,
-%\end{align}
-%\esub
-%with a condition on $\rho$ and $\varphi$ that is now
-%\be
-%\rho e^{2i\varphi} e^{2i (\lambda_{a}^{*} - \psi_{L}^{*})} = \sigma_{-}^{\rm inj} \,.
-%\ee
-\JGB{Keep something from this?}
-Thus, we find again a degenerate pattern, with the difference that \eqref{eq:soldegeniotaphi} imposes a condition on the pair $\lambda_{a} - \psiL$ instead of $\lambda_{a}$, and the extra condition imposes a condition on $\varphi$ instead of the pair $\varphi - \psiL$.
-
-When considering more harmonics beyond the dominant mode $h_{22}$, such degeneracies will be broken easily.
-As in the case of the full likelihood, the simplified likelihood~\eqref{eq:simplelikelihoodhm} will have several terms with different inclination and phase dependencies.
-If the signal is loud enough for at least two modes to be detected, then the inclination $\iota$ and the phase $\varphi$ are fixed by the relative amplitude and phase of these modes.
-This will break the degeneracies $(d, \iota)$ and the degeneracy $(\varphi, \psiL)$.
-
-We illustrate our findings by running a Bayesian parameter estimation for the simplified likelihood~\eqref{eq:simplelikelihood22} with the two samplers \texttt{MultiNest} and \texttt{PTMCMC} described in Sec.~\ref{}.
-Since the inner products~\eqref{eq:innerproductlmlpmp} are constants computed only once, the external factors become trivially inexpensive, and the likelihood is very fast to compute.
-We take advantage of this to run \texttt{PTMCMC} for a very large number of steps, ensuring that we do not miss degenerate regions of the parameter space.
-The resulting posterior distribution is extremely degenerate, as shown in Fig.~\ref{fig:PEsmbh22hmSimpleLikeCase9}.
-The sampler \texttt{MultiNest} fails to resolve this structure, stopping after having only explored the degenerate region $\iota \rightarrow 0$ and $\iota \rightarrow \pi$ and the eight sky positions~\eqref{eq:soldegensky}.
-In accordance with~\eqref{eq:degenconditionsky}-\eqref{eq:degenconditionprefactor}, this corresponds to lines in $\varphi \pm \psiL = \mathrm{const}$.
-The sampler \texttt{PTMCMC} finds more degenerate regions, in particular it finds similar structures with the roles of $\iota$ and $\betaL$ exchanged (as shown notably by the $(d,\iota)$ and $(d,\betaL)$ panels), as well as intermediate regions connecting all these features.
-
-Fig.~\ref{fig:PEsmbh22hmSimpleLikeCase9} also shows the result obtained when including all available harmonics~\eqref{} in~\eqref{eq:simplelikelihoodhm}.
-The degenerate structures collapse to leave only eight modes corresponding to $\betaL \rightarrow - \betaL$ and $\lambdaL \rightarrow \lambdaL + k\pi/2$.
-We found that this degeneracy-breaking occurs already when including a single subdominant harmonic $h_{21}$.
-
-We can now come back to explaining the results of Sec.~\ref{}, and in particular Fig.~\ref{}, obtained with the true likelihood function for the mode $h_{22}$.
-First, the difference between the two Bayesian samplers has been made more extreme by going to the more degenerate likelihood~\eqref{eq:simplelikelihood22}, allowing us to attribute their disagreement to an apparent shortcoming of \texttt{MultiNest}\footnote{We recall that we only used standard values for the metaparameters \SM{[complete]} for this sampler, and we do not exclude that it would be possible to improve results with minor changes.} for resolving extended degenerate regions.
-Focusing on the sky position, we zoom close to the injection in Fig.~\ref{fig:Skydownsmbh22hmSimpleLikeCase9}.
-We compare here the posterior distribution generated with the true likelihood with the one obtained with the simplified likelihood~\eqref{eq:simplelikelihood22}, for both samplers.
-We also superimpose the prediction~\eqref{eq:soldegensky} for the degeneracy-induced peak.
-We see that, although ,
-
-%\be\label{eq:symmetryresponse}
-%\betaL \rightarrow -\betaL\,, \quad \iota \rightarrow \pi - \iota \,, \quad \psiL \rightarrow \pi - \psiL
-%\ee
-%leaves the signal unchanged. We can also rewrite the likelihood in a way that makes clearer the symmetry between the two pairs of parameters $(\iota, \varphiL)$ and $(\betaL, \lambdaL)$ \SM{[TODO]}.
-%
-%The simplified extrinsic likelihood for a frozen LISA then takes the form
-%\be\label{eq:simplelikelihood}
-%	\ln \calL = - \frac{1}{2} \Lambda \left( \left| s_{a} - s_{a}^{\rm inj} \right|^{2} + \left| s_{e} - s_{e}^{\rm inj} \right|^{2} \right) \,,
-%\ee
-%where $\Lambda$ appears as an overall normalization factor, playing the role of the square SNR of the signal.
-
-%We can now explore possible degeneracies in the likelihood given by~\eqref{eq:simplelikelihood}, \eqref{eq:defsase} and \eqref{eq:defa22a2m2}-\eqref{eq:defe22e2m2}.
-We note that the relative contributions of the modes $h_{2,2}$ and $h_{2,-2}$ are strongly separated in the face-on or face-off limit, leading to a simplification of the response and an enhanced degeneracy.
-Due to the symmetry~\eqref{eq:symmetryresponse}, we will focus on the face-on limit $\iota \rightarrow 0$.
-In this limit, the $h_{2,-2}$ contributions die off rapidly as $a_{2,-2}\,, e_{2,-2} \sim \sin^{4} (\iota/2)$.
-If they are assumed to be negligible, then we can rewrite
-%\bsub\label{eq:saseapprox}
-%\begin{align}
-%	s_{a} \simeq i\calA e^{2 i \xi} \left( D_{a}^{+} + i D_{a}^{\times} \right) \,, \\
-%	s_{e} \simeq i\calA e^{2 i \xi} \left( D_{e}^{+} + i D_{e}^{\times} \right) \,,
-%\end{align}
-%\esub
-%where we introduced the notations $\xi (\varphiL, \psiL) \equiv -\varphiL - \psiL$ and $\calA(d, \iota) \equiv 3/4\sqrt{5/\pi}\cos^{4}(\iota/2)/d $ for the common phase and amplitude factors.
-%
-%For a degeneracy to occur, both complex quantities $s_{a}$ and $s_{e}$ must be close to the injection values $s_{a}^{\rm inj}$ and $s_{e}^{\rm inj}$, which can be achieved as follows if~\eqref{eq:saseapprox} is valid.
-Defining the complex ratio
-%\be\label{eq:defr}
-%	r(\lambdaL, \betaL) = \frac{D_{a}^{+} + i D_{a}^{\times}}{D_{e}^{+} + i D_{e}^{\times} } \,,
-%\ee
-%if the position in the sky $(\lambda_{L}^{*}, \beta_{L}^{*})$ can be chosen so that
-%\be\label{eq:rcondition}
-%	r(\lambda_{L}^{*}, \beta_{L}^{*}) = \frac{s_{a}^{\rm inj}}{s_{e}^{\rm inj}} \,,
-%\ee
-%then adjusting $\calA$ and $\xi$ so that
-%\be
-%	i\calA e^{2 i \xi} = \frac{s_{a}^{\rm inj}}{\left( D_{a}^{+} + i D_{a}^{\times} \right)(\lambda_{L}^{*}, \beta_{L}^{*})}
-%\ee
-%generates degenerate points in parameter space with $\ln \calL \simeq 0$.
-%
-%The conditions~\eqref{eq:defr}-\eqref{eq:rcondition} can be inverted for the sky position as follows.
-For a given complex value of $r = s_{a}^{\rm inj} / s_{e}^{\rm inj}$, $r(\lambda_{L}^{*}, \beta_{L}^{*}) = r$ can be rewritten as
-%\be\label{eq:eqtosolveforlambdabetaL}
-%	e^{4i\lambda_{L}^{*}}e^{i\frac{\pi}{3}} \frac{\left( 1 +\sin\beta_{L}^{*} \right)^{2}}{\left( 1 - \sin\beta_{L}^{*} \right)^{2}} = \frac{1+ i r}{1 - i r} \,.
-%\ee
-%Solving for the modulus of~\eqref{eq:eqtosolveforlambdabetaL} then yields
-%\be\label{eq:solutionbetaL}
-%	\sin\beta_{L}^{*} = \frac{\rho - 1}{\rho + 1} \,,
-%\ee
-%for
-%\be
-%	\rho = \sqrt{\left| \frac{1+ i r}{1 - i r} \right|} \,.
-%\ee
-%Solving for the argument of~\eqref{eq:eqtosolveforlambdabetaL} gives four solutions for $\lambda_{L}^{*}$ as
-%\be\label{eq:solutionlambdaL}
-%	\lambda_{L}^{*} = - \frac{\pi}{12} + \frac{1}{4}\mathrm{Arg} \frac{1+ i r}{1 - i r} + \frac{k \pi}{2} \quad \mathrm{for} \quad k = 0,\dots,3 \,.
-%\ee
-%
-%Repeating the argument for the other branch of solutions $\iota \rightarrow \pi$ (or directly considering the symmery~\eqref{eq:symmetryresponse}) gives four other solutions in the sky with identical sky longitudes $\lambda_{L}^{*}$, with symmetrized sky latitudes $\beta_{L}^{*} \rightarrow -\beta_{L}^{*}$, and with $\xi (\varphiL, \psiL) \equiv -\varphiL + \psiL$, $\calA(d, \iota) \equiv 3/4\sqrt{5/\pi}\sin^{4}(\iota/2)/d $.
-%
-%Thus, with~\eqref{eq:solutionbetaL}, \eqref{eq:solutionlambdaL} and $\beta_{L}^{*} \rightarrow -\beta_{L}^{*}$, we arrive at eight degenerate positions in the sky $(\lambda_{L}^{*}, \beta_{L}^{*})$.
-For each of these sky positions, an exact line degeneracy exists between $\varphiL$ and $\psiL$ as long as $\xi = \mathrm{const}$, as well as an approximate line degeneracy between $d$ and $\iota$, as long as $\calA = \mathrm{const}$, limited to inclination values where $\sin^{4}(\iota/2)$ (or $\cos^{4}(\iota/2)$, for the other branch of solutions) remains negligible.
-The combination of these degeneracies generates an extended structure in the multi-dimensional parameter space.
-As a result, when considering the marginal posterior for the sky position, these degenerate spots in the sky can have a lot of probability support.
+We have found that pre-merger analysis of the previous section made apparent the degeneracy pattern that can be predicted analytically from the structure of the \textit{Frozen, low-$f$} response approximation.
+Analyses of IMR signals as in~\ref{sec:MBHBPE} show that only the \textit{reflected} mode~\eqref{eq:degenreflection} survives post-merger.
+In this Section we explain how and why this transition occurs.
+
+We have already presented a discussion of different qualitative effects in the instrument response in Sec.~\ref{MBHBresponseapprox} and Fig.~\ref{responseapproxtdia22}: one hand hand, the motion of LISA leaves an imprint (the time-dependency) on the low frequencies, where the time-to-frequency map is steep enough that a short interval in frequency maps to a large interval of time; on the other hand, the breakdown of the long-wavelength approximation leaves an imprint (the frequency-dependency) at high frequencies.
+Here we look at the quantitative importance of these features on the inference as a function of frequency (or as a function of time).
+We can readily select one or the other feature by using the \textit{Frozen} response, ignoring time dependency, or the \textit{Low-$f$} response, ignoring frequency dependency.
+
+In Fig.~\ref{fig:lnLskymodesapprox} we show log-likelihood values obtained with either the \textit{Full}, \textit{Frozen} or \textit{Low-$f$} response, with and without higher harmonics, for each of the eight modes in the sky described by~\eqref{eq:degenreflection}-\eqref{eq:degenlambdapi2}.
+The likelihood~\eqref{eq:deflnLaet} is computed with the same response approximation is used for the injection and for the template.
+The results are shown as a function of frequency in a cumulative sense: we compute the likelihood by accumulating signal up to the frequency shown in the $x$-axis.
+When including higher harmonics, this cut in frequency is interpreted as a cut in time and propagated to other harmonics according to~\eqref{eq:fscalinglm}.
+
+A value of $\ln \calL = 0$ means that the template signal with shifted parameters is identical to the injection, while a very negative $\ln \calL$ means that these parameter values are excluded by the inference.
+The injection (full black line) has always $\ln \calL = 0$ by construction.
+The \textit{reflected} sky mode~\eqref{eq:degenreflection} is the dashed black line, and the \textit{antipodal} sky mode~\eqref{eq:degenanitpodal} is the dashed red line.
+
+With the \textit{Low-$f$} response, the \textit{antipodal} mode remains almost exactly degenerate with the injection, differing only by the Doppler phase~\eqref{eq:defPhiR}, which has a small effect on our short signals as shown in Fig.~\ref{fig:tranferSMBHCase9}.
+As explained below~\eqref{eq:degenantipodal}, other modes fail to reproduce the injected signal because of the time-dependence of the pattern functions.
+They acquire a moderate penalty at low frequency, but $\ln \calL$ then goes to a constant since the motion becomes negligible at high frequencies, as shown by the transfer functions in Fig.~\ref{fig:responseapproxtdia22}.
+
+With the \textit{Low-$f$} response, the \textit{reflected} mode remains exactly degenerate as anticipated in Sec.~\ref{subsec:MBHBPEdegen}.
+All other modes see their $\ln \calL$ fall rapidly at around $2\mHz$ due to the onset of frequency-dependency in the response.
+Comparing to Fig.~\ref{fig:responseapproxtdia22}, we see that around $2\mHz$ the breaking of the long-wavelength approximation is still mild; however, the SNR start to accumulate a lot from $\SNR/16 \sim 40$ at $2.5\, \mathrm{h}$ to $\SNR/4 \sim 160$ at $7\min$, magnifying the effect of these features.
+Comparing the two rows of the figure, we see that the higher harmonics have an effect here, in making the elimination of secondary modes happen earlier in frequency (or time).
+This is expected as they start to contribute significantly to the SNR (see Fig.\ref{fig:modeoverlapsoftime}) and reach higher in frequency than the $h_{22} harmonic$ according to~\eqref{eq:fscalinglm}.
+
+The \textit{Low-$f$} response case is essentially a superposition of the previous two response approximations.
+The early sky localization is quite poor and relies on the LISA motion.
+Getting closer to merger We stress that the .
+Overall, including the higher harmonics does not change this qualitative picture, despite an earlier onset of frequency-dependent degeneracy-breaking features.
+
+Limitations of such explorations should be made clear: by using a pointwise estimate, we cannot make statistical statements and are missing volume effects discussed in Sec.~\ref{subsec:MBHBPEdegen22}; by transforming the angular parameters with an analytical prescription while keeping the other fixed, we are in a sense overconstraining the degeneracy.
+In particular, it is conceivable that one could find a better match to the injection in the vicinity of a degenerate sky mode, by adjusting slightly all parameters.
+
+This is indeed what happens in multidmensional Bayesian parameter estimation, and is illustrated in Fig.~\ref{fig:lnLskymodesapprox}: in the first panel, for the three pre-merger analyses as well as for the post-merger analysis, we overlay circles indicating the best $\ln \calL$ found among the posterior samples, in each eighth of the sky corresponding to the eight sky modes (when no samples are present in this region of the sky, nothing is displayed).
+A sampler is not an optimizer, so the precise value achieved is not optimal and would vary when repeating runs; nevertheless, this measure gives a good proxy for how much closer of the injected signal we could get by slightly biasing the parameters.
+
+We see that the best $\ln \calL$ among samples can be higher than the point estimate would tell us, especially for the analysis at $\sim 7 \min$ before merger, and higher harmonics make a visible difference.
+This has consequences for approximations like the Fisher analysis, if secondary peaks are shifted from the analytical predictions.
+We leave for future work the investigation of approximate representations of sky degeneracies.
 
@@
 
-\section{The LISA instrument response}
-\label{app:response}
+\section{Conventions for orbits and frames}
+\label{app:conventions}
 
@@
 
+\SM{update: change of plan}
 For completeness, we provide in this Appendix comprehensive definitions and notations for the gravitational wave signal as well as the LISA response.
@@
 
-\subsection{The GW signal}
-\label{sec:gwsignal}
-
-We start by introducing a source-frame $(\hat{x}_{S}, \hat{y}_{S}, \hat{z}_{S})$ attached to the binary system emitting gravitational waves.
-For comparable-mass systems without spin, the natural choice is to take the normal to the orbital plane as $\hat{z}_{S}$, and we will assume that the remaining rotation around $\hat{z}_{S}$ is fixed by the phase convention of the waveform model (see Sec.~\ref{}).
-
-Introducing $k$ the wave propagation vector going from the source towards the observer, we define the inclination $\iota$ and the observer phase $\varphi$ simply as being its spherical angular coordinates $(\theta_{S}, \phi_{S})$, so that in the source-frame
+\subsection{Source frame, wave frame and SSB frame}
+\label{sec:wavessbframe}
+
+We first define a source frame $(\hat{x}_{S}, \hat{y}_{S}, \hat{z}_{S})$ associated to the compact binary system emitting gravitational waves.
+We choose $\hat{z}_{S} = \hat{L}$, the direction of the orbital angular momentum (constant for non-precessing systems).
+The choice of $\hat{x}_{S}$, $\hat{y}_{S}$ amounts to fixing the phase convention.
+
+Introducing $k$ the wave propagation unit vector going from the source towards the observer, we define the inclination $\iota$ and the observer phase $\varphi$ as its spherical angular coordinates in the source frame, so that in that frame
 \be
@@
 \esub
-If $H_{ij} = h_{ij}^{\rm TT}$ represents the gravitational wave signal in the transverse-traceless gauge,
-\be
-	H = h_{+} P_{+} + h_{\times} P_{\times}
-\ee
-with the polarization tensors
+
+...
+
+We start by introducing a source frame $(\hat{x}_{S}, \hat{y}_{S}, \hat{z}_{S})$ attached to the binary system emitting gravitational waves.
+Relating this frame to the physical configuration of the binary is understood as being part of the waveform model.
+For comparable-mass systems without spin or with aligned spins\footnote{For precessing systems, different choices are possible, like $\hat{z}_{S} = \hat{J}$, the direction of the total angular momentum.}, the natural choice is to take the normal to the orbital plane as $\hat{z}_{S}$, and we will assume that the remaining rotation around $\hat{z}_{S}$ is fixed by the phase convention of the waveform model (fixing the waveform phase at a reference frequency, or fixing the orbital phase at a reference time, see Sec.~\ref{} \SM{...}).
+
+Introducing $k$ the wave propagation unit vector going from the source towards the observer, we define the inclination $\iota$ and the observer phase $\varphi$ as its spherical angular coordinates in the source frame, so that in that frame
+\be
+	k_{S} = (\sin\iota \cos\varphi, \sin\iota \sin\varphi, \cos\iota) \,.
+\ee
+
+By convention, for our polarization vectors $p$ and $q$ we will use the spherical coordinate vectors $p = e_{\theta}^{S}$, $q = e_{\phi}^{S}$. In terms of only $\hat{z}_{S}$,
 \bsub
 \begin{align}
-	P_{+} &= p \otimes p - q \otimes q \,,\\
-	P_{\times} &= p \otimes q + q \otimes p \,.
-\end{align}
-\esub
-Conversely, the polarizations are
-\bsub
-\begin{align}
-	h_{+} &= \frac{1}{2} \left(p \otimes p - q \otimes q \right) : H \,,\\
-	h_{\times} &= \frac{1}{2} \left( p \otimes q + q \otimes p \right) : H \,.
-\end{align}
-\esub
-
-We will further decompose the gravitational wave signal, seen as a function of the direction of emission $(\iota, \varphi)$, in spin-weighted spherical harmonics as
-\be\label{eq:apphpcmodes}
-	h_{+} - i h_{\times} = \sum_{\ell \geq 2} \sum_{m = -\ell}^{\ell} {}_{-2}Y_{\ell m} (\iota, \varphi) h_{\ell m} \,,
-\ee
-where the explicit expression of the ${}_{-2}Y_{\ell m}$ is given in~\cite{}.
-
-The vectors $(p,q,k)$ form the Wave-frame, such that the gravitational wave takes in this frame the familiar form
+	p &= q \times k \,,\\
+	q &= \hat{z}_{S} \times k / |\hat{z}_{S} \times k|\,.
+\end{align}
+\esub
+
+The vectors $(p,q,k)$ form the wave frame, such that the gravitational wave takes in this frame the familiar form
 \be
@@
 
-We will both generate the waveforms and apply the response directly in the Fourier domain. Our convention for the Fourier transform is
-\be\label{eq:defFourier}
-	\tilde{F}(f) = \int dt \; e^{2 i \pi f t} F(t) \,.
-\ee
-Note that this differs by a change $f\rightarrow -f$ from the most usual one (used e.g. in \texttt{LAL}).
-
-The definition of the reference time and phase \SM{[\dots COMPLETE]}
+We can now relate the wave frame to a detector frame $(\hat{x}, \hat{y}, \hat{z})$, which will fix our convention for the sky position and polarization angle.
+We choose here this detector frame to be based on the plane of ecliptic, and centered on the Solar System Barycenter; we will call this frame the SSB frame.
+Below, we will introduce another detector frame more suitable for short-lived signals, the LISA frame.
+
+The source position in the sky is given by $(\lambda, \beta)$, the ecliptic longitude and latitude in this SSB-frame.
+Like with the source frame, we can introduce spherical vectors $(e_{r}^{SSB}, e_{\theta}^{SSB}, e_{\phi}^{SSB})$.
+Since the propagation vector is $k = - e_{r}^{SSB}$,
+\be
+	k = (- \cos\beta\cos\lambda, - \cos\beta\sin\lambda, -\sin\beta) \,.
+\ee
+
+The last degree of freedom between the frames represents a rotation along the line-of-sight, parametrized by the polarization angle $\psi$.
+We introduce reference polarization vectors in the SSB-frame as
+\bsub
+\begin{align}
+	u &= \hat{z} \times k / |\hat{z} \times k| \,,\\
+	v &= k \times u \,.
+\end{align}
+\esub
+In terms of spherical vectors, $(k, u, v) = (-e_{r}^{SSB}, -e_{\phi}^{SSB}, -e_{\theta}^{SSB})$.
+
+In the following, we will use the notation $R(v,\varpi)$ to denote the matrix of an active rotation around the vector $v$ by an angle $\varpi$.
+The polarization angle $\psi$ is then defined such that $(p,q)$ are obtained by rotating $(u,v)$ by the angle $\psi$ around $k$, i.e.
+\be
+	(p,q) = R(k, \psi) \cdot (u,v) \,.
+\ee
+
+These relations can be summarized by the active rotation matrix $R_{W}$ from the SSB-frame to the wave frame:
+\be\label{eq:RW}
+	R_{W} = R(z, \lambda - \pi/2) \cdot R(x, \beta + \pi/2 ) \cdot R(z, \psi)\,.
+\ee
 
@@
 \label{sec:lisatraj}
-
-For completeness, we gather in this appendix explicit expressions for the trajectory of the LISA constellation and the rotation matrices entering the response.
 
@@
 \ee
-The rigid approximation for the constellation, at first order in $e$, can be seen as being at first order in $L/R \simeq 0.017$ for a $2.5 \, \mathrm{Gm}$ armlength~\cite{}.
+The rigid approximation for the constellation, at first order in $e$, can be seen as a first-order approximation in the small parameter $L/R \simeq 0.017$ for a $2.5 \, \mathrm{Gm}$ armlength~\cite{}.
+
+\TDC{Something is incorrect in this discussion. From (A11) and the definition of $a$ above, it sounds like the armlength is longer than 1 au.}
 
@@
 It will be very useful to introduce a time-dependent frame $(\hat{x}_{L}, \hat{y}_{L}, \hat{z}_{L})(t)$ following the detector in its motion.
-Specifically, we choose this LISA frame such that at any time,
+Specifically, we choose this LISA frame (L-frame for short) such that at any time,
 \bsub
@@
 
-If $R(v,\varpi)$ denotes the matrix of an active rotation around the vector $v$ by an angle $\varpi$, and if we denote by $R_{L}$ the active rotation from the SSB-frame to the L-frame such that for each of the three basis vectors $(\hat{x}_{L}, \hat{y}_{L}, \hat{z}_{L}) = R_{L} (\hat{x}, \hat{y}, \hat{z})$, we have
+If we denote by $R_{L}$ the active rotation from the SSB-frame to the L-frame such that for each of the three basis vectors $(\hat{x}_{L}, \hat{y}_{L}, \hat{z}_{L}) = R_{L} (\hat{x}, \hat{y}, \hat{z})$, we have
 \be\label{eq:RL}
@@
 
-\subsection{From the Wave frame to the SSB frame}
-\label{sec:wavessbframe}
-
-We can now relate the Wave-frame to the SSB-frame, which will fix our convention for the sky position and polarization angle.
-The source position in the sky is given by $(\lambda, \beta)$, the ecliptic longitude and latitude in the SSB-frame.
-Like with the Source-frame, we can introduce spherical vectors $(e_{r}^{SSB}, e_{\theta}^{SSB}, e_{\phi}^{SSB})$.
-Since the propagation vector is $k = - e_{r}^{SSB}$,
-\be
-	k = (- \cos\beta\cos\lambda, - \cos\beta\sin\lambda, -\sin\beta) \,.
-\ee
-
-The last degree of freedom between the frames represents a rotation along the line-of-sight, parametrized by the polarization angle $\psi$.
-We introduce reference polarization vectors in the SSB-frame as
-\bsub
-\begin{align}
-	u &= \hat{z} \times k / |\hat{z} \times k| \,,\\
-	v &= k \times u \,.
-\end{align}
-\esub
-In terms of spherical vectors, $(k, u, v) = (-e_{r}^{SSB}, -e_{\phi}^{SSB}, -e_{\theta}^{SSB})$.
-The polarization angle $\psi$ is then defined such that $(p,q)$ are obtained by rotating $(u,v)$ by the angle $\psi$ around $k$, i.e.
-\be
-	(p,q) = R(k, \psi) \cdot (u,v) \,.
-\ee
-
-These relations can be summarized by the active rotation matrix $R_{W}$ from the SSB-frame to the Wave-frame:
-\be\label{eq:RW}
-	R_{W} = R(z, \lambda - \pi/2) \cdot R(x, \beta + \pi/2 ) \cdot R(z, \psi)\,.
-\ee
-
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-
 \subsection{Parametrizing the signal in the LISA-frame}
@@
 While long-lasting signals like SOBHs will see a strong imprint of the LISA orbital motion over the course of observation, SMBH signals are dominated in SNR by a short-lived burst of emission at merger.
-For such signals, it will be useful to use a parametrization based on the L-frame at the time of merger instead of the SSB frame.
+For such signals, it will be useful to use a parametrization based on the LISA frame at the time of merger instead of the SSB frame.
 
 The new parameters are defined as playing the same role as the SSB parameters, but relative to the L-frame.
-Namely, we define $R_{LW}$ the active rotation matrix from the L-frame to the Wave-frame, expressed in the L-frame basis, so that for each basis vector expressed in the L-frame basis $(\hat{x}_{W}, \hat{y}_{W}, \hat{z}_{W})_{L} = R_{LW} \cdot (\hat{x}_{L}, \hat{y}_{L}, \hat{z}_{L})_{L}$.
+Namely, we define $R_{LW}$ the active rotation matrix from the L-frame to the wave frame, expressed in the L-frame basis, so that for each basis vector expressed in the L-frame basis $(\hat{x}_{W}, \hat{y}_{W}, \hat{z}_{W})_{L} = R_{LW} \cdot (\hat{x}_{L}, \hat{y}_{L}, \hat{z}_{L})_{L}$.
 The defining condition on $(\lambda_{L}, \beta_{L}, \psi_{L})$ is
@@
 Combining~\eqref{eq:RLW}, \eqref{eq:RL}, \eqref{eq:RW} and~\eqref{eq:defRLW} yields the following expressions for the L-frame angular parameters\footnote{The convention here is that the point of coordinates $(x,y)$ in the plane has for argument $\arctan\left[ x, y \right]$.}:
-\bsub
+\bsub\label{eq:Lframeangles}
 \begin{align}
@@
 	& \qquad\quad\;\; \left. + \sin\frac{\pi}{3} \sin\beta \sin\alpha \right] \,,\\
-	\psi_{L} &= \arctan \left[ \cos\psi \left( \cos\frac{\pi}{3}\cos\beta + \sin\frac{\pi}{3} \sin\beta \cos(\lambda - \alpha)\right) \right. \nn\\
-	& \qquad\quad\;\; \left. + \sin \psi \sin\frac{\pi}{3} \sin(\lambda - \alpha), \right. \nn\\
-	& \qquad\qquad\; \left. \sin\psi \left( \cos\frac{\pi}{3}\cos\beta + \sin\frac{\pi}{3} \sin\beta \cos(\lambda - \alpha)\right) \right. \nn\\
-	& \qquad\quad\;\; \left. - \cos \psi \sin\frac{\pi}{3} \sin(\lambda - \alpha) \right] \,.
-\end{align}
-\esub
+%	\psi_{L} &= \arctan \left[ \cos\psi \left( \cos\frac{\pi}{3}\cos\beta + \sin\frac{\pi}{3} \sin\beta \cos(\lambda - \alpha)\right) \right. \nn\\
+%	& \qquad\quad\;\; \left. + \sin \psi \sin\frac{\pi}{3} \sin(\lambda - \alpha), \right. \nn\\
+%	& \qquad\qquad\; \left. \sin\psi \left( \cos\frac{\pi}{3}\cos\beta + \sin\frac{\pi}{3} \sin\beta \cos(\lambda - \alpha)\right) \right. \nn\\
+%	& \qquad\quad\;\; \left. - \cos \psi \sin\frac{\pi}{3} \sin(\lambda - \alpha) \right] \,.
+	\psi_{L} &= \psi + \arctan \left[\cos\frac{\pi}{3}\cos\beta + \sin\frac{\pi}{3} \sin\beta \cos(\lambda - \alpha), \right. \nn\\
+	&\qquad\qquad\quad\;\; \left.-\sin \frac{\pi}{3} \sin(\lambda - \alpha) \right] \,.
+%gwtools.modpi(psiSSB + arctan2(coslambda*sinalpha*sinzeta -cosalpha*sinzeta*sinlambda, cosbeta*coszeta + cosalpha*coslambda*sinbeta*sinzeta + sinalpha*sinbeta*sinzeta*sinlambda))
+\end{align}
+\esub
+Through $\alpha = \Omega_{0} (t-t_{0}) $, these expressions are time-dependent, reflecting the rotation of this frame to follow LISA on its orbit.
 
 The definition of the reference phase $\phi_{\rm Ref}$, \dots \SM{[COMPLETE]}
-
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-
-\subsection{Mode decomposition and polarization angle}
-\label{sec:modespol}
-
-We now translate the mode decomposition~\eqref{eq:hpcmodes} in the Fourier domain, and make more explicit the dependence of the signal in the polarization $\psi$, so as to make manifest its purely extrinsic nature.
-Indeed, $\psi$ enters the observables only through constant prefactors, which is a bit obscured by the rotation matrix formulation~\eqref{eq:RW}.
-
-The mode decomposition~\eqref{eq:hpcmodes} gives (dropping $(\iota, \varphi)$)
-\bsub
-\begin{align}
-	h_{+} = \frac{1}{2} \sum_{\ell, m} \left( \sYlm h_{\ell m} + \sYlmstar h_{\ell m}^{*} \right) \,,\\
-	h_{\times} = \frac{i}{2} \sum_{\ell, m} \left( \sYlm h_{\ell m} - \sYlmstar h_{\ell m}^{*} \right) \,,
-\end{align}
-\esub
-which is valid in general. Now, for non-precessing binary systems, an exact symmetry relation between modes reads
-\be\label{eq:nonprecsymmetry}
-	h_{\ell, -m} = (-1)^{\ell} h_{\ell, m}^{*} \,.
-\ee
-Using this symmetry, we can write
-\be
-	h_{+,\times} = \sum_{\ell, m} K_{\ell m}^{+, \times} h_{\ell m} \,,
-\ee
-with
-\bsub
-\begin{align}
-	K_{\ell m}^{+} =\frac{1}{2} \left( \sYlm + (-1)^{\ell} \sYlminusmstar \right) \,,\\
-	K_{\ell m}^{\times} = \frac{i}{2} \left( \sYlm - (-1)^{\ell} \sYlminusmstar \right) \,.
-\end{align}
-\esub
-
-Going to the Fourier domain, an approximation often used is to neglect support for negative/positive frequencies according to
-\be\label{eq:approxnegf}
-	\tilde{h}_{\ell m} (f) \simeq 0 \;\; \text{for} \;\; m<0, \; f>0 \;\; ( m>0, \; f<0 )\,,
-\ee
-and neglecting modes $h_{\ell 0}$.
-We will use this approximation throughout this paper.
-Note that we picked our Fourier convention~\eqref{eq:defFourier} to ensure that the mode $\tilde{h}_{22} (f)$ has support for $f>0$.
-Using~\eqref{eq:approxnegf},
-\be
-	\tilde{h}_{+,\times} = \sum_{\ell} \sum_{m>0} K_{\ell m}^{+, \times} \tilde{h}_{\ell m} \,.
-\ee
-
-Next, it is convenient to introduce mode-by-mode polarization matrices
-\be
-	P_{\ell m} = P_{+} K_{\ell m}^{+} + P_{\times} K_{\ell m}^{\times} \,,
-\ee
-so that
-\be\label{eq:Hsummodes}
-	H = \sum_{\ell m} P_{\ell m} h_{\ell m} \,.
-\ee
-To make explicit the dependence in polarization, we can define polarization tensors for 0 polarization angle as
-\bsub
-\begin{align}
-	P_{+}^{0} &= P_{+}(\psi = 0) = u \otimes u - v \otimes v \,,\\
-	P_{\times}^{0} &= P_{\times}(\psi = 0) = u \otimes v + v \otimes u \,.
-\end{align}
-\esub
-This allows to write the dependence in polarization as
-\be
-	P_{+} + i P_{\times} = e^{-2 i \psi} \left( P_{+}^{0} + i P_{\times}^{0} \right) \,,
-\ee
-or explicitly in $P_{\ell m}$ as
-\begin{align}
-	P_{\ell m} &= \frac{1}{2} \sYlm e^{-2 i \psi} \left( P_{+}^{0} + i P_{\times}^{0} \right) \nn\\
-	& + \frac{1}{2} (-1)^{\ell} \sYlminusmstar e^{+2 i \psi} \left( P_{+}^{0} - i P_{\times}^{0} \right) \,.
-\end{align}
-Writing the $P_{\ell m}$ matrices in this way allows to factor out all dependencies in the extrinsic parameters $(\iota, \phi, \psi)$.
-Together with the luminosity distance $D_{L}$, these parameters appear in combinations of constant factors multiplying the whole response.
-\JGB{Would it be better to include the angle dependence explicitly here in the $\sYlm$ here for clarity, since the point is to account for parameter dependencies.}
-
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-
-\subsection{The LISA observables}
-\label{sec:LISAobservables}
-
-\SM{[avoid repetition with methodology section]}
-
-The single-link frequency shift observable $y_{slr} = (\nu_{r} - \nu_{s})/\nu$ between sending spacecraft $s$ and receiving spacecraft $r$ along the link $l$ takes the form
-\be\label{eq:appdefyslr}
-	y_{slr} = \frac{1}{2} \frac{n_{l} \otimes n_{l}}{1 - k\cdot n_{l}} : \left[ H(t - L - k\cdot p_{s}) - H(t - k\cdot p_{r}) \right] \,.
-\ee
-Several assumptions enter this form of the response: \dots \SM{[COMPLETE]} \JGB{Define $n_l$. Also double dot looks like colon, needing some clarification.}.
-
-In the frequency domain, we will use the analysis of~\cite{} and write the response in individual observables $y_{slr}$ with a transfer function for each spherical harmonic mode as
-\be
-	\tilde{y}_{slr} = \sum_{\ell, m}\calT_{slr}^{\ell m}(f) \tilde{h}_{\ell m} \,.
-\ee
-With the decomposition~\eqref{eq:Hsummodes}, the corresponding form of the kernel $G^{\ell m}_{slr}(f, t)$ in the formalism of~\cite{} is
-\begin{align}\label{eq:Gslr}
-	G_{slr}^{\ell m}(f,t) &= \frac{i \pi f L}{2} \sinc \left[ \pi f L\left(1-k\cdot n_{l} \right) \right] \nn\\
-	& \quad \cdot \exp\left[ i \pi f \left( L + k\cdot \left( p_{1} + p_{2} \right) \right) \right]  n_{l} \cdot P_{\ell m} \cdot n_{l} \,.
-\end{align}
-At leading order in the separation of timescales, we have simply
-\be
-	\calT_{slr}^{\ell m}(f) = G_{slr}^{\ell m}(f, t_{f}^{\ell m}) \,,
-\ee
-with $t_{f}^{\ell m}$ the effective time-to-frequency correspondence, defined across the whole frequency band, including in the merger-ringdown, as $t_{f}^{\ell m} = -1/(2\pi) d\Psi_{\ell m} / df$ for a mode $\tilde{h}_{\ell m} = A_{\ell m}(f) \exp\left[-i \Psi_{\ell m}\right]$.
-The analysis of~\cite{} has shown that higher-order corrections are small for SMBH systems, and for SOBH systems that are close enough to merger as will be the case for the examples chosen here.
-The fact that we use the same Fourier-domain treatment of the transfer functions for both the signal and the templates should also mitigate errors.
-
-In our rigid approximation, delays are simply all constant and constant to $L$. Using the shortcut notation $y_{sr,L} = y_{slr}(t - L)$, the first-generation TDI Michelson observable $X$ reads~\cite{}
-\begin{align}
-	X &= y_{31} + y_{13,L} + \left( y_{21} + y_{12,L} \right)_{2L} \nn\\
-	& - \left( y_{21} + y_{12,L} \right) - \left( y_{31} + y_{13,L} \right)_{2L} \,,
-\end{align}
-with the other Michelson observables $Y$, $Z$ being obtained by cyclic permutation. The independent combinations $A$, $E$ and $T$ are then defined as
-\bsub
-\begin{align}
-	A &= \frac{1}{\sqrt{2}} \left( Z - X \right) \,,\\
-	E &= \frac{1}{\sqrt{6}} \left( X - 2Y + Z \right) \,,\\
-	T &= \frac{1}{\sqrt{3}} \left( X + Y + Z \right) \,.
-\end{align}
-\esub
-These channels are independent under the approximation of an uncorrrelated noise between the arms. With constant delays in the rigid approximation, the TDI combinations take the form
-\bsub\label{eq:aet}
-\begin{align}
-	\tilde{a} &= \left[ (1+z) \left( \tilde{y}_{31} + \tilde{y}_{13} \right) - \tilde{y}_{23} - z \tilde{y}_{32} - \tilde{y}_{21} - z \tilde{y}_{12} \right] \,,\\
-	\tilde{e} &= \frac{1}{\sqrt{3}} \left[ (1-z)\left( \tilde{y}_{13} - \tilde{y}_{31} \right) + (2+z) \left( \tilde{y}_{12} - \tilde{y}_{32} \right) \right. \nn\\
-	&\qquad\quad \left. + (1+2z) \left( \tilde{y}_{21} - \tilde{y}_{23} \right) \right] \,,\\
-	\tilde{t} &= \frac{\sqrt{2}}{\sqrt{3}} \left[ \tilde{y}_{21} - \tilde{y}_{12} + \tilde{y}_{32} - \tilde{y}_{23} + \tilde{y}_{13} - \tilde{y}_{31} \right] \,,
-\end{align}
-\esub
-where we have introduced the rescalings
-\bsub\label{eq:scalingAET}
-\begin{align}
-	\tilde{a}, \tilde{e} &= \frac{1}{i \sqrt{2} z \sin 2\pi f L}\times \tilde{A}, \tilde{E} \,,\\
-	\tilde{t} &= \frac{1}{2\sqrt{2} z^{3} \sin \pi f L \sin 2\pi f L} \times \tilde{T} \,.
-\end{align}
-\esub
-Scaling out the same square factors from the noise power spectral density as
-\bsub
-\begin{align}
-	S_{n}^{A}, S_{n}^{E} &= 2 \sin^{2} 2\pi f L \times S_{n}^{a}, S_{n}^{e} \,,\\
-	S_{n}^{T} &= 8 \sin^{2} \pi f L \sin^{2} 2\pi f L S_{n}^{t} \,,
-\end{align}
-\esub
-the reduced PSD for the three channels take the form
-\bsub\label{eq:Snaet}
-\begin{align}
-	S_{n}^{a} = S_{n}^{e} &= 2 \left( 3 + 2\cos 2\pi f L + \cos 4 \pi f L \right) S^{\rm pm}(f) \nn\\
-	& \quad + \left( 2 + \cos 2\pi f L \right) S^{\rm op}(f) \,,\\
-	S_{n}^{t} &= 4 \sin^{2} 2\pi f L S^{\rm pm}(f) + S^{\rm op}(f) \,,
-\end{align}
-\esub
-with $S^{\rm pm}$ the test-mass noise PSD and $S^{\rm op}$ the optical noise PSD.
-Here we ignore the confusion noise coming from the background of galactic binaries in the  LISA band.
-Note that the prefactors~\eqref{eq:scalingAET} are oscillatory and have zero-crossings at high frequencies, and scaling them out from both the signal and the noise is an approximation, as we can expect that the real delays would lead to imperfect cancellations in the vicinity of the zero-crossings.
-
-Finally, the likelihood function that we use, with a zero noise realization, is
-\begin{align}
-	\ln\calL &= -\frac{1}{2} \left[ \left( \tilde{a} - \tilde{a}_{\rm inj} | \tilde{a} - \tilde{a}_{\rm inj}  \right) + \left( \tilde{e} - \tilde{e}_{\rm inj} | \tilde{e} - \tilde{e}_{\rm inj}  \right) \right.
-\nn\\
-	& \qquad\qquad + \left. \left( \tilde{t} - \tilde{t}_{\rm inj} | \tilde{t} - \tilde{t}_{\rm inj} \right) \right] \,,
-\end{align}
-with the $\rm inj$ subscript indicating the injected signals, and where the inner products $(\cdot | \cdot)$ are given by~\eqref{eq:definnerproduct} with the noise PSDs~\eqref{eq:Snaet}.
-
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-
-\subsection{The low-frequency limit}
-\label{sec:low-freq}
-
-As is well known~\cite{Cutler97}, in the low-frequency limit (also called the long-wavelength approximation), the finite-length effects of the arms of LISA disappear, and the response of the detectors falls back to the one of two LIGO-type detectors set in motion, rotated from each other by $\pi/4$.
-
-For $f \ll f_{L}$, we have $2\pi f L \ll 1$ and the kernel~\eqref{eq:defGslr} reduces to
-\be\label{eq:Gslrlowf}
-	G^{\ell m}_{slr} \simeq \frac{i \pi f L}{2} \exp\left[ 2 i \pi f k\cdot p_{0} \right] n_{l} \otimes n_{l} : P_{\ell m}\,.
-\ee
-For even lower frequencies, when $f \ll f_{R}$ as well, $2\pi f R \ll 1$ and the delay term is negligible: $\exp\left[ 2 i \pi f k\cdot p_{0} \right] \simeq 1$.
-It can be convenient to introduce the following notations for this delay phase:
-\be\label{eq:defPhiR}
-	\Phi_{R} \equiv 2 \pi f k\cdot p_{0} \,, \quad \Delta \Phi_{R} = \Phi_{R} - \Phi_{R}(t=t_{f}^{\rm peak}) \,.
-\ee
-Here, $\Phi_{R}$ with $t=t_{f}$ is what is called often the Doppler phase in the literature, and $\Delta \Phi_{R}$ subtracts the contribution that corresponds to shifting the origin of times to the time of the peak seen at the LISA location \SM{[COMPLETE $t_{L}$]}, which is equivalent to using the time variable $t_{L}$ as defined in~\eqref{}.
-
-In the following, we will also apply frequency-domain response at leading order in the separation of timescales~\cite{MB18}, so that the transfer functions are $\calT(f) = G(f, \tf)$ with $\tf$ the time-of-frequency function~\eqref{eq:deftf}.
-For $2\pi f L \ll 1$, we have $z\simeq 1$, and the link reversal symmetry $\tilde{y}_{slr} \simeq \tilde{y}_{r-ls}$, so that~\eqref{eq:aet} becomes
-\bsub\label{eq:aetlowf}
-\begin{align}
-	\tilde{a} &\simeq 4\tilde{y}_{31} - 2 \tilde{y}_{23} - 2 \tilde{y}_{12} \,,\\
-	\tilde{e} &\simeq 2\sqrt{3} \left[ \tilde{y}_{12}  - \tilde{y}_{23} \right] \,,\\
-	\tilde{t} &\simeq 0 \,,
-\end{align}
-\esub
-with the $T$-channel becoming negligible in this limit. Using~\eqref{eq:Gslrlowf}, we can write
-\be\label{eq:aelowfmodes}
-	\tilde{a}, \tilde{e} = \sum_{\ell, m>0} (-2i\pi f) \exp\left[ 2 i \pi f k\cdot p_{0} \right] \tilde{h}_{\ell m} D_{a,e} : P_{\ell m} \,,
-\ee
-where we introduced the detector tensors
-\bsub\label{eq:DaDe}
-\begin{align}
-	D_{a} &= \frac{L}{2} \left( n_{1}\otimes n_{1} + n_{3} \otimes n_{3} - 2 n_{2} \otimes n_{2} \right) \,,\\
-	D_{e} &= \frac{L\sqrt{3}}{2} \left( n_{1} \otimes n_{1} - n_{3} \otimes n_{3} \right) \,.
-\end{align}
-\esub
-Here, we have made apparent factors $(-2i\pi f)$ corresponding to the Fourier-domain translation of a time derivative.
-Indeed, the observables $\Delta \nu / \nu$ used here are time derivatives of the laser phase more commonly used for ground-based detectors.
-
-We can now map these two channels to two fictitious LIGO-type detectors as follows.
-For an orthogonal detector of the same setup as the ground-based LIGO and Virgo with armlength $L_{D}$, rotated by an angle $\epsilon_{D}$ from the basis vectors $x,y$, the detector tensor is
-\begin{align}
-	D &= \frac{L_{D}}{2} \left[ \cos 2\epsilon_{D} \left( x\otimes x - y\otimes y \right) \right. \nn\\
-	& \qquad\quad + \left. \sin 2\epsilon_{D} \left( x\otimes y + y \otimes x\right) \right] \,.
-\end{align}
-By comparison with~\eqref{eq:DaDe}, we obtain
-\be
-	L_{a} = L_{e} = \frac{3L}{2} \,, \; \epsilon_{a} = \frac{2\pi}{3} \,, \; \epsilon_{e} = \frac{5\pi}{12} \,.
-\ee
-Note a degeneracy $\pm \pi$ in how we choose to orient these effective detectors.
-We can recover pattern functions (for $\psi_{L}=0$) by factoring out the effective length, we define
-\JGB{Would it be better to just include a new version of expression \eqref{eq:aelowfmodes} here using $F_{a,e}$, such as:
-  $\tilde{a}, \tilde{e} = (-3i\pi f L)\left[ \tilde{h}_{\ell m}^\times F_{a,e}^+ + \tilde{h}_{\ell m}^\times F_{a,e}^\times\right]$.
-Also an $(\ell,m)$ version of $F_{a,e}$ would be useful in the main text.}
-\be\label{eq:defFapcFepc}
-	F_{a,e}^{+,\times} \equiv \frac{2}{3L} D_{a,e} : P_{+,\times}^{0} \,,
-\ee
-which gives in terms of the LISA-frame sky position angles
-\bsub\label{eq:FapcFepc}
-\begin{align}
-	F_{a}^{+} &= \frac{1}{2} \left( 1 + \sin^{2}\beta_{L} \right) \cos \left(2\lambda_{L} - \frac{\pi}{3} \right) \,,\\
-	F_{a}^{\times} &= \sin\beta_{L} \sin \left(2\lambda_{L} - \frac{\pi}{3} \right) \,,\\
-	F_{e}^{+} &= \frac{1}{2} \left( 1 + \sin^{2}\beta_{L} \right) \cos \left(2\lambda_{L} + \frac{\pi}{6} \right) \,,\\
-	F_{e}^{\times} &= \sin\beta_{L} \sin \left(2\lambda_{L} + \frac{\pi}{6} \right)  \,.
-\end{align}
-\esub
-Note that the expressions for the channel $e$ can be obtained from the expressions for $a$ with the replacement $\lambdaL \rightarrow \lambdaL + \pi/4$.
 
@@
 
-\section{SNR dependence of parameter estimation}
-\label{sec:SMBHPEdistance}
-
--- show series of analyses varying only the distance/SNR
+\section{Sampling the fully degenerate extrinsic posterior}
+\label{app:samplingdegen}
 
@@
   \centering
-  \begin{minipage}{.32\linewidth}
-      \includegraphics[width=.99\linewidth]{../plots/corner_smbh_case9_hm_dseries_m1m2.png}
-   \end{minipage}
-   \begin{minipage}{.32\linewidth}
-      \includegraphics[width=.99\linewidth]{../plots/corner_smbh_case9_hm_dseries_Dinc.png}
-   \end{minipage}
-   \begin{minipage}{.32\linewidth}
-      \includegraphics[width=.99\linewidth]{../plots/corner_smbh_case9_hm_dseries_tphi.png}
-   \end{minipage}
-  \caption{.}
-  \label{fig:smbhCornerZoomHMCase9}
+  \includegraphics[width=.8\linewidth]{../plots/corner_smbh_case9_simplelike_ptmcmc_bambi.png}
+  \caption{Posterior distributions obtained with the two samplers \texttt{PTMCMC} (blue) and $\texttt{MultiNest}$ (green) for the simplified likelihood~\eqref{eq:simplelikelihoodhm}, using only the dominant mode $(\ell, m) = (2,2)$.
+Masses and time are pinned to the injected value, and the motion of the detector is neglected.
+We also superimpose (red) the contours obtained with \texttt{MultiNest} by including all harmonics $(2,2)$, $(2,1)$, $(3,3)$, $(4,4)$, $(5,5)$.}
+  \label{fig:PEsmbh22hmSimpleLikeCase9}
 \end{figure*}
 
-\begin{figure*}
-  \centering
-  \includegraphics[width=.8\linewidth]{../plots/plot_mollweide_dseries.png}
-  \caption{.}
-  \label{fig:MollweidesmbhTserieshmCase9}
-\end{figure*}
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+
+%\section{SNR dependence of parameter estimation}
+%\label{sec:SMBHPEdistance}
+%
+%-- show series of analyses varying only the distance/SNR
+%
+%\begin{figure*}
+%  \centering
+%  \begin{minipage}{.32\linewidth}
+%      \includegraphics[width=.99\linewidth]{../plots/corner_smbh_case9_hm_dseries_m1m2.png}
+%   \end{minipage}
+%   \begin{minipage}{.32\linewidth}
+%      \includegraphics[width=.99\linewidth]{../plots/corner_smbh_case9_hm_dseries_Dinc.png}
+%   \end{minipage}
+%   \begin{minipage}{.32\linewidth}
+%      \includegraphics[width=.99\linewidth]{../plots/corner_smbh_case9_hm_dseries_tphi.png}
+%   \end{minipage}
+%  \caption{.}
+%  \label{fig:smbhCornerZoomHMCase9}
+%\end{figure*}
+%
+%\begin{figure*}
+%  \centering
+%  \includegraphics[width=.8\linewidth]{../plots/plot_mollweide_dseries.png}
+%  \caption{.}
+%  \label{fig:MollweidesmbhTserieshmCase9}
+%\end{figure*}
 

### 2019-11-15 data/t-series-mcmc-results/skyplot_Lframe.py

@@ -22,14 +22,16 @@ for i, label, color, bw in zip([64, 16, 4, 1], labels, colors, bandwidths):
     # create the KDE estimator
     # we could use grid-search cross validation to estimate the bandwidth here
     radec = np.vstack((dec, ra)).T
-    kde = KernelDensity(kernel='tophat', bandwidth=bw, metric='haversine').fit(radec)
+    kde = KernelDensity(kernel='gaussian', bandwidth=bw, metric='haversine').fit(radec)
 
     # find the density levels corresponding to 1, 2, 3 sigmas
     n = 10000
     sample = kde.sample(n)
     sample_densities = np.sort(np.exp(kde.score_samples(sample)))
     # Note: are those levels appropriate for 2d ?
-    levels = [sample_densities[int(n * (1 - p))] for p in [0.9973, 0.9545, 0.6827]]
+    # levels = [sample_densities[int(n * (1 - p))] for p in [0.9973, 0.9545, 0.6827]]
+    sigmalevels2d = 1.0 - np.exp(-0.5 * np.linspace(1.0, 3.0, num=3) ** 2)
+    levels = [sample_densities[int(n * (1 - p))] for p in sigmalevels2d]
 
     gra = np.linspace(-np.pi, np.pi, 300)
     gdec = np.linspace(-np.pi/2, np.pi/2, 300)
@@ -41,4 +43,4 @@ for i, label, color, bw in zip([64, 16, 4, 1], labels, colors, bandwidths):
 
 ax.grid()
 ax.legend(loc=[0.4, 0.75])
-pl.savefig('skymap_Lframe.png', dpi=200)
+pl.savefig('skymap_Lframe_test.png', dpi=200)

### 2020-02-10 paper/paper_PE.tex
Working on the paper

@@
 %\title{Bayesian methods for black hole merger parameter estimation with LISA.}
-\title{Exploring Bayesian parameter estimation of binary black holes with LISA}
+\title{Exploring the Bayesian parameter estimation of binary black holes with LISA}
 
@@
 
-The space-based gravitational wave detector LISA will observe mergers of supermassive black hole binary systems (SMBHs) to cosmological distances, as well as inspiralling stellar-origin (or stellar-mass) binaries (SOBHs) years before they enter the LIGO/Virgo band.
+The space-based gravitational wave detector LISA will observe mergers of supermassive black hole binary systems (SMBHBs) to cosmological distances, as well as inspiralling stellar-origin (or stellar-mass) binaries (SBHBs) years before they enter the LIGO/Virgo band.
 Much remains to be explored for the parameter recovery of both class of systems.
-Previous SMBH analyses relied on inspiral-only signals and/or a simplified Fisher matrix analysis, while SOBHs have not yet been analyzed with Bayesian methods.
+Previous SMBHB analyses relied on inspiral-only signals and/or a simplified Fisher matrix analysis, while SBHBs have not yet been analyzed with Bayesian methods.
 We accelerate likelihood computations by (i) using a Fourier-domain response of the LISA instrument, (ii) using a reduced-order model for non-spinning waveforms that include a merger-ringdown and higher harmonics, (iii) setting the noise realization to zero and computing overlaps in the amplitude/phase representation.
@@
 Including higher harmonics proves to be crucial to break degeneracies and considerably improves the localization of the source, with a surviving bimodality in the sky position.
-We also present the first simulations of Bayesian inference for SOBHs, and show that although unimodal, their posterior distributions can have non-Gaussian features.
+We also present simulations of Bayesian inference for the extrinsic parameters of SBHBs, and show that although unimodal, their posterior distributions can have non-Gaussian features.
 
@@
 By targeting a much lower frequency band, will be sensitive to different sources.
-Among the sources, mergers of SMBHs, seen at high SNR and anywhere in the observable universe, and inspirals of LIGO-type sources (SOBHs), opening the possibility of multi-band gravitational wave astronomy.
+Among the sources, mergers of SMBHs, seen at high SNR and anywhere in the observable universe, and inspirals of LIGO-type sources (SBHBs), opening the possibility of multi-band gravitational wave astronomy.
 %
 %Much remained to be explored for the data analysis of LISA.
-Different challenges : high SNR for merger-dominated SMBH signals, SNR-limited regime for very long SOBH signals.
+Different challenges : high SNR for merger-dominated SMBH signals, SNR-limited regime for very long SBHB signals.
 The motion of the detector leaves an imprint on the signal, while the response also has frequency-dependency at high frequencies where the long-wavelength approx is not valid anymore.
@@
 
-Gravitational waves from the coalescence of black hole binaries are routinely observed \cite{GWTC-1} by the ground-based interferometers Advanced LIGO \cite{LIGO}, Advanced Virgo \cite{Virgo} and, soon, KAGRA \cite{KAGRA}.
-
-Ground-based interferometers are fundamentally limited at low frequency by terrestrial noise, and cannot be used to study mergers of compact objects much heavier than $\approx 10^3 \Msol$.
-The LISA mission \cite{LISA2017} is designed to overcome this limitation, enabling the observation and precise characterization of mergers of black holes ranging in mass from tens of $\Msol$ (stellar-origin black holes, SOBH) to millions of $\Msol$ (supermassive black holes, SMBH).
-Black hole mergers, in fact, will be a major class of sources in the LISA frequency band.
+Gravitational waves from the coalescence of black hole binaries are now regularly observed \cite{GWTC-1} by the ground-based interferometers Advanced LIGO~\cite{LIGO}, Advanced Virgo~\cite{Virgo} and, soon, KAGRA~\cite{KAGRA}.
+
+Ground-based interferometers are fundamentally limited at low frequency by terrestrial noise, and cannot be used to study mergers of compact objects much heavier than a few $ 10^2 \Msol$.
+The spacebourne detector LISA~\cite{LISA2017} will overcome this limitation, enabling the observation and precise characterization of gravitational waves from binary black holes, from coalescences of systems with millions of $\Msol$ (supermassive black hole binaries, SMBHB) to the inspiral of systems with tens of $\Msol$ (stellar-mass black hole binaries, SBHB).
+The observation of black hole mergers and inspirals in the LISA frequency band will yield major scientific rewards, and they form the main target of LISA among other classes of sources.
 
 In order to realize the full scientific objective of LISA with respect to black hole mergers, adequate data analysis tools must be prepared in advance.
-Similarly to what happens in LIGO and Virgo, one has to first identify the presence of a merger waveform in the LISA data, and then infer the distribution of the physical parameters of its source from the data, enabling the construction of a catalog of black hole mergers.
-Inferring the parameters of binaries, in particular their distance and sky location, will also be necessary to organize multimessenger observations of their late inspirals and mergers \cite{TAP-LISA} and for multiband gravitational-wave astronomy \cite{Sesana16}.
-In this paper we focus on the latter part of the analysis problem, i.e.~the inference of the black hole parameters from the LISA data, once we have reasons to believe the presence of the signal in the data.
-
-The inference problem amounts to produce samples from the posterior distribution for sourcee parameters, related by Baye's theroem to the likelihood function, i.e.~the probability of observing the measured data given the source parameters and a model of the source and detector.
-In LISA, this problem is more complex than in kilometer-scale ground-based interferometers.
-Depending on the masses and spins of the black holes, additional challenges arises from the much larger expected signal-to-noise ratios (SNRs) for high-mass systems, or the much longer duration of the waveform for low-mass systems.
-For SMBHBs, the large SNR will require a very accurate modeling of the waveform, including the merger and ringdown regimes, the effects of spin and corrections of higher harmonics in the signal beyond the quadrupole.
-
-The response of the LISA instrument to a given waveform shows a time- and frequency-dependence.
-Contrary to LIGO and Virgo, which are approximately fixed in an inertial frame during the observation of any merger, the LISA constellation will move around the Sun and change its orientation appreciably during the observation of a large fraction of the inspirals, leading to additional modulations of the signals.
+Similarly to what happens in LIGO and Virgo, one has to first identify the presence of a merger waveform in the LISA data, possibly in the presence of other superposed signals (notably galactic white dwarfs binaries~\cite{Nelemans+01}), and then infer the distribution of the physical parameters of its source from the data, enabling the construction of a catalog of black hole binaries.
+Inferring the parameters of binaries, in particular their distance and sky location, and refining the analysis as the signal accumulates with observation time, will also be necessary to organize multimessenger observations of their late inspirals and mergers~\cite{ArmitageNatarajan02, DalCanton+19}.
+Cosmological applications using LISA observations as standard sirens~\cite{Schutz86, Tamanini+16} also depend on the ability to localize individual sources.
+Understanding parameter estimation of SBHBs will also be important to understand the outcomes and challenges of possible multiband gravitational-wave observations~\cite{Sesana16}.
+In this paper we focus on the latter part of the analysis problem, i.e.~the inference of the black hole parameters from the LISA data, once we have reasons to believe the presence of the signal in the data, leaving aside the identification problem.
+We also limit our analysis to extrinsic parameters and ignore the effect of superposed gravitational wave signals from various sources.
+
+The inference problem amounts to produce samples from the posterior distribution for source parameters, related by Baye's theorem to the likelihood function, i.e.~the probability of observing the measured data given the source parameters and a model of the source and detector.
+In LISA, this problem is more complex than in kilometer-scale ground-based interferometers, as additional challenges arises from the much larger expected signal-to-noise ratios (SNRs) for high-mass systems, or the much longer duration of the waveform for low-mass systems.
+In particular, for SMBHBs, the large SNR will require a very accurate modeling of the waveform, including the merger and ringdown regimes, the effects of spin and corrections of higher harmonics in the signal beyond the quadrupole.
+
+The response of the LISA instrument to a given waveform shows a time- and frequency-dependence~\cite{Cutler97, Larson+99, CR02, MB18}.
+Contrary to LIGO and Virgo, which are approximately fixed in an inertial frame during the observation of any merger, the LISA constellation will move around the Sun and change its orientation appreciably if the signal is observable for a significant fraction of a year, leading to additional modulations of the signals.
 Furthermore, due to the LISA arm length, for many sources the long-wavelength approximation will not hold, which introduces a frequency-dependence in the response.
@@
 
-Previous parameter recovery studies with LISA
-
-\TDC{Add literature review on BBH PE here, can base on GOAT study.}
-
-Here we demonstrate that standard Bayesian inference can be performed for SMBHs and SOBHs using a self-consistent waveform model that includes inspiral, merger, ringdown and higher-order modes (albeit neglecting spins) and a full model of the LISA response.
+Previous parameter recovery studies with LISA used mainly simplified signal and response models, and often relied on a Fisher matrix approximation to parameter recovery instead of full Bayesian simulations.
+Numerous works used the combination of inspiral-only post-Newtonian waveforms (sometimes including precession), the low-frequency approximate response and Fisher matrix estimates; examples include~\cite{Cutler97, Vecchio03, Arun06, Berti+04, LangHughes06}.
+The Mock LISA Data Challenges~\cite{MLDC09} used such a setup for the signals and the response, albeit moving towards MCMC tools with a focus on detection.
+Bayesian methods going beyond Fisher albeit still with inspiral signals were developed in~\cite{Brown+07, CornishPorter06a, Crowder+06, Wickham+06, Roever+07, Feroz+09, GairPorter09, Petiteau+09, PorterCarre13, PorterCornish15}.
+The importance of the higher harmonics for LISA was stressed already in several studies~\cite{Arun+07a, TriasSintes07, PorterCornish08, McWilliams+09}.
+To explore the importance of the merger-ringdown using Numerical relativity waveforms, the studies~\cite{Thorpe+08, McWilliams+09, McWilliams+10, McWilliams+11} used a Fisher approach, while~\cite{Babak+08} used Bayesian analyses limited to extrinsic parameters.
+More recently, in the context of the redesign of the LISA mission~\cite{elisa13}, the study~\cite{Klein+15} explored the performance of various LISA instrumental designs using Fisher matrix estimates and inspiral signals, but used a reweighting procedure to represent the role of the merger-ringdown signal.
+To this date, no full Bayesian parameter estimation studies with IMR signals have been performed.
+
+On the other hand, important advances have been registered in recent years in providing fast and accurate waveforms including the merger and ringdown for the LIGO/Virgo data analysis, through phenomenological waveforms~\cite{Khan+15, London+17}, Reduced Order Models (ROM) of Effective One Body Waveforms~\cite{Puerrer14, Bohe+16}, and numerical relativity surrogates~\cite{Blackman+17b, Varma+18}.
+This progress has yet to be transposed to LISA applications.
+
+Here we demonstrate that standard Bayesian inference can be performed for SMBHBs using a self-consistent waveform model that includes inspiral, merger, ringdown and higher-order modes (albeit still simplified as spins are neglected) and a full model of the LISA response.
+We investigate and explain degeneracies in the posterior distribution of the source's parameters, and highlight the crucial role played by the higher harmonics in the signal and the frequency-dependency in the instrument response.
+
 We start by introducing a fast method to calculate the likelihood for a black hole merger.
-The method includes: a fast computation of a Fourier-domain inspiral-merger-ringdown waveform with higher-order modes based on reduced-order modeling; a fast Fourier-domain computation of the LISA response based on \cite{MB18}; and a fast method for computing the noise-weighted product between waveforms.
-We couple our likelihood with two codes for Bayesian inference based on different sampling techniques, constructing an inference engine which allows us to perform a variety of investigations with simulated black hole mergers.
-Both samplers are first shown to converge to comparable posterior distributions of the source parameters for a few case studies, and we highlight the challenges we encounter in the sampling.
-The full posterior distributions are then compared with their Fisher approximations, widely used in gravitational-wave analyses.
-Focusing on massive black holes, we show next the effect of higher-order modes on the degeneracies between some of the parameters.
-We derive an analytic and intuitive explanation of the origin of such degeneracies via various approximations.
-We show examples of how the posterior distributions evolve over time as more and more data becomes available for the inference.
+This method includes: a fast computation of a Fourier-domain inspiral-merger-ringdown waveform with higher-order modes based on reduced-order modeling; a fast Fourier-domain computation of the LISA response based on~\cite{MB18}; and a fast method for computing the noise-weighted product between waveforms when setting the noise realization to zero.
+We couple our likelihood with two codes for Bayesian inference based on different sampling techniques, constructing an inference engine which will allow us in the future to perform a variety of investigations with simulated black hole mergers.
+
+Focusing on two examples of moderately massive black holes binaries, we perform Bayesian parameter estimation simulation for full, inspiral-merger-signals, and we show the crucial role of higher-order modes.
+We highlight the challenges we encounter in the sampling, and compare our two samplers with each other and with the Fisher matrix approximation.
+We explore degeneracies appearing between some of the parameters when ignoring these higher harmonics, and derive an analytic explanation of their origin via a simplified approximations to the instrument response.
+We show for our examples how the posterior distributions evolve over time as more and more data becomes available for the inference, and explain how degeneracies between different positions of the source in the sky are broken by features in the instrument response.
+
 Finally, we repeat some of our investigations for stellar-origin black hole mergers as well.
+Parameter estimation studies for SBHB systems have relied so far on the Fisher matrix approach~\cite{Sesana16, Vitale16, Nishizawa+16a, Nishizawa+16b}, with Bayesian inference tools yet to be developed.
+Focusing on the posterior distribution of extrinsic parameters (fixing the masses and time to coalescence), we demonstrate that Bayesian analyses can be run with a comparable speed for the likelihood computation and discuss the features of the posteriors.
+
+Section~\ref{sec:response} introduces comprehensive notations for the frequency-domain signal and response, and Section~\ref{sec:method} explains our methods for likelihood computations and Bayesian sampling.
+Section~\ref{sec:SMBH} presents the SMBHB example signals that we analyze and their accumulation with time, together with their instrumental response in various approximations.
+In Section~\ref{sec:SMBHPE}, we present our main results for the parameter estimation of SMBHBs; we contrast results obtained with and without higher harmonics, contrast Fisher matrix with Bayesian results, describe and explain degeneracies in parameter space.
+In Section~\ref{sec:MBHBPEacctime}, we turn to the refinement of parameter estimation as the signal accumulates with ongoing observations, describing how multiple inferred sky positions can coexist before features of the full instrument response break the degeneracies.
+In Section~\ref{sec:SBHB}, we turn to the application of our fast likelihood computation to investigate extrinsic parameter inference of SBHBs.
+We summarize and discuss our findings in Section~\ref{sec:summarydiscussion}.
 
@@
 
-\SM{add mini-intro}
-
+In this section, we introduce a complete set of notations for the LISA instrument response that will be useful in the later discussion of degeneracies in parameter space.
 We will use units with $G=c=1$.
@@
 
-One can further decompose the gravitational wave signal, seen as a function of the direction of emission $(\iota, \varphi)$ in the source frame, in spin-weighted spherical harmonics as
-\be\label{eq:apphpcmodes}
-	h_{+} - i h_{\times} = \sum_{\ell \geq 2} \sum_{m = -\ell}^{\ell} {}_{-2}Y_{\ell m} (\iota, \varphi) h_{\ell m} \,,
-\ee
-where the explicit expression of the ${}_{-2}Y_{\ell m}$ is given in~\cite{}.
+One can further decompose the gravitational wave signal, seen as a function of the direction of emission $(\iota, \varphi)$ in the source frame, in spin-weighted spherical harmonics~\cite{Goldberg+67} as
+\be\label{eq:hpcmodes}
+	h_{+} - i h_{\times} = \sum_{\ell \geq 2} \sum_{m = -\ell}^{\ell} {}_{-2}Y_{\ell m} (\iota, \varphi) h_{\ell m} \,.
+\ee
+Explicit expression of the ${}_{-2}Y_{\ell m}$ can be found in e.g.
+Sec.~3 of~\cite{BlanchetLiving}.
 In the following, we will use exclusively this parametrization of the waveform as a set of modes $h_{\ell m}$.
@@
 
-We will both generate the waveforms and apply the response directly in the Fourier domain. Our convention for the Fourier transform of a function $F$ is
+We will both generate the waveforms and apply the response directly in the Fourier domain.
+Our convention for the Fourier transform of a function $F$ is\footnote{Note that this differs by a change $f\rightarrow -f$ from the more usual convention used e.g.
+in \texttt{LAL}~\cite{lal}.}
 \be\label{eq:defFourier}
@@
 \ee
-Note that this differs by a change $f\rightarrow -f$ from the more usual convention used e.g. in \texttt{LAL}~\cite{}.
-
-The definition of the reference time and phase \SM{[\dots COMPLETE]}
 
@@
 Writing the $P_{\ell m}$ matrices in this way allows to factor out explicitly all dependencies in the extrinsic parameters $(\iota, \phi, \psi)$.
-Together with the luminosity distance $D_{L}$ scaling the overall amplitude of the signal, these parameters enter as constant (time and frequency independent) prefactors in the response for each mode $h_{\ell m}$.
+Together with the luminosity distance $D$ scaling the overall amplitude of the signal, these parameters enter as constant (time and frequency independent) prefactors in the response for each mode $h_{\ell m}$.
 
@@
 
-The LISA response can be built from single-link observables $y_{slr} = (\nu_{r} - \nu_{s})/\nu$ respresenting a laser frequency shift between the sending spacecraft $s$ and the receiving spacecraft $r$ along the link $l$.
-We use the expression~\cite{}
+The LISA response~\cite{EW75, Cutler97, Larson+99, CR02, RCP04} can be built from single-link observables $y_{slr} = (\nu_{r} - \nu_{s})/\nu$ representing a laser frequency shift between the sending spacecraft $s$ and the receiving spacecraft $r$ along the link $l$.
+We use the expression~\cite{Vallisneri04, Krolak+04}
 \be\label{eq:defyslr}
@@
 \ee
-with $H$ the transverse-traceless matrix representing the gravitational wave, $L$ the delay along one arm\footnote{We use units with $G=c=1$.}, taken to be fixed, $k$ the propagation vector, $n_{l}$ the link unit vectors (from $s$ to $r$) and $p_{s}$, $p_{r}$ the positions of the spacecrafts.
+with $H$ the transverse-traceless matrix representing the gravitational wave, $L$ the delay along one arm\footnote{We recall that we use units with $G=c=1$.
+The proposed armlength for LISA is $L = 2.5\mathrm{Gm}$~\cite{LISA2017}.}, taken to be fixed, $k$ the propagation vector, $n_{l}$ the link unit vectors (from $s$ to $r$) and $p_{s}$, $p_{r}$ the positions of the spacecrafts.
 Here $n_{l}$, $p_{s}$ and $p_{r}$ are evaluated at the same time $t$.
@@
 \ee
-Applying the perturbative formalism of~\cite{MB18} to leading order, we have simply
+Applying the perturbative formalism of~\cite{MB18} to leading order in the separation of timescales, we have simply
 \be\label{eq:Tlmslr}
@@
 \ee
-is the effective frequency$\to$time correspondence, defined across the whole frequency band and including the merger and ringdown.
-
-The analysis of~\cite{MB18} has shown that higher-order corrections are small for SMBH systems, as well as for SOBH systems provided they are close enough to merger as will be the case for the examples chosen here.
+is the effective time-frequency correspondence, defined across the whole frequency band and including the merger and ringdown. This definition generalizes the Stationary Phase Approximation (SPA).
+
+The analysis of~\cite{MB18} has shown that higher-order corrections are small in general for SMBHB systems, and are also small for SBHB systems provided they are not too far from the coalescence.
 The fact that we use the same Fourier-domain treatment of the transfer functions for both the signal and the templates should also mitigate the importance of modelling errors.
@@
 
-In~\eqref{eq:defyslr} and~\eqref{eq:Gslr}, one can distinguish two types of delays: on one hand the delay associated to the position $p_{0}$ of the center of the constellation on its oarbit around the Sun, with baseline $R=1 \au$, and on the other hand the delays associated to the individual spacecraft positions in the constellation, with baseline $L=2.5 \cdot 10^{6 }\mathrm{km}$.
+In~\eqref{eq:defyslr} and~\eqref{eq:Gslr}, one can distinguish two types of delays: on one hand the delay associated to the position $p_{0}$ of the center of the constellation on its orbit around the Sun, with baseline $R=1 \au$, and on the other hand the delays associated to the individual spacecraft positions in the constellation, with baseline the armlength $L=2.5 \cdot 10^{6 }\mathrm{km}$.
 This defines the transfer frequencies
@@
 \esub
-The transfer frequency $f_{L}$ correspond to fitting a full wavelength in the armlength; as we will see below in Sec.~\ref{}, the long-wavelength approximation starts to break down at significantly lower frequencies.
+The transfer frequency $f_{L}$ correspond to fitting a full wavelength in the armlength; as we will see below in Sec.~\ref{sec:MBHBPEacctime}, departures from the long-wavelength approximation starts to be important at significantly lower frequencies.
 
@@
 \subsection{Time Delay Interferometry}
-\label{sec:TDIresponse}
-
-To mitigate laser noise, LISA requires Time Delay Interferometry (TDI)~\cite{}.
+\label{subsec:TDIresponse}
+
+To mitigate laser noise, LISA requires Time Delay Interferometry (TDI)~\cite{TintoArmstrong99, Armstrong+99, Estabrook+00, Dhurandhar+02, Tintoliving}.
 These TDI observables are built from delayed combinations of the one-arm basic observables~\eqref{eq:defyslr}.
-Detailed cancellations for a flexing and rotating constellation prompted the development of higher-generation TDI schemes~\cite{}.
+Detailed cancellations for a flexing and rotating constellation prompted the development of higher-generation TDI schemes~\cite{Shaddock03, CornishHellings03, Shaddock+03, Tinto+04}.
 However, these refinements affect only marginally the response to the gravitational waves.
 We therefore limit ourselves to first-generation TDI and adopt a rigid approximation for the cancellation, where delays are all constant and set to $L$.
-Using the notation $y_{slr,L} = y_{slr}(t - L)$, the first-generation TDI Michelson observable $X$ reads~\cite{}
+Using the notation $y_{slr,nL} = y_{slr}(t - nL)$, the first-generation TDI Michelson observable $X$ reads~\cite{Vallisneri04}
 \begin{align}
@@
 \end{align}
-with the other Michelson observables $Y$, $Z$ being obtained by cyclic permutation. Uncorrelated combinations $A$, $E$ and $T$ are then defined as
-\bsub
+with the other Michelson observables $Y$, $Z$ being obtained by cyclic permutation. Uncorrelated combinations $A$, $E$ and $T$~\cite{Prince+02} are then defined as
+\bsub\label{eq:defAET}
 \begin{align}
@@
 \esub
-These channels are independent under the assumption of an identical and uncorrrelated \SM{check} noise in the detector arms.
+These channels are independent under the assumption of an identical and uncorrrelated noise in the detector arms.
+Note that various conventions coexist in the literature.
 With constant delays in the rigid approximation, with the notation $z\equiv \exp[2i\pi fL]$ the TDI combinations take the form
@@
 \esub
-where we have eliminated frequency-dependent prefactors that are common to the signal and to the noise by introducing the rescalings \SM{for $\tilde{a}$, $\tilde{e}$ is it $1/z^{2}$ ?}
+where we have eliminated frequency-dependent prefactors that are common to the signal and to the noise by introducing the rescalings
 \bsub\label{eq:scalingAET}
 \begin{align}
-	\tilde{a}, \tilde{e} &= \frac{1}{i \sqrt{2} z \sin (2\pi f L)}\times \tilde{A}, \tilde{E} \,,\\
-	\tilde{t} &= \frac{1}{2\sqrt{2} z^{3} \sin (\pi f L) \sin (2\pi f L)} \times \tilde{T} \,.
+	\tilde{a}, \tilde{e} &= \frac{1}{i \sqrt{2} e^{2i\pi fL} \sin (2\pi f L)}\times \tilde{A}, \tilde{E} \,,\\
+	\tilde{t} &= \frac{1}{2\sqrt{2} e^{3i\pi fL} \sin (\pi f L) \sin (2\pi f L)} \times \tilde{T} \,.
 \end{align}
@@
 \ee
-To make the connection between these observables and the familiar strain observed by ground-based intruments (see Sec.~\ref{} detailing the low-frequency approximation), we also introduce the notation
-\be\label{eq:defhaet}
-	\tilde{h}_{a,e,t} \equiv \frac{1}{-6i \pi f L} \times \tilde{a}, \tilde{e}, \tilde{t} \,,
-\ee
-and similarly the transfer functions $\calT_{h_{a}, h_{e}, h_{t}}^{\ell m} = \calT_{a,e,t}^{\ell m} / (-6 i \pi f L)$.
-
-Scaling out the same square factors from the noise power spectral density as
+To make the connection between these reduced TDI observables and the gravitational strain, more familiar in the context of ground-based intruments, we also introduce the notations
+\bsub\label{eq:defhaet}
+\begin{align}
+	\tilde{h}_{a,e,t} \equiv \frac{1}{(-6i \pi f L)} \times \tilde{a}, \tilde{e}, \tilde{t} \,, \\
+	\calT_{h_{a}, h_{e}, h_{t}}^{\ell m} = \frac{1}{(-6i \pi f L)} \calT_{a,e,t}^{\ell m}\,.
+\end{align}
+\esub
+
+Scaling out the same square factors from the noise power spectral density (PSD) as
 \bsub
@@
 \esub
-with $S^{\rm pm}$ the test-mass noise PSD and $S^{\rm op}$ the optical noise PSD, described in~\ref{} \SM{complete}.
-Note that the confusion noise coming from the background of galactic binaries in the  LISA band is to be added to the point-mass and optical noises.
-We can define strain-like noise PSD associated to the strain-like TDI observables~\eqref{eq:defhaet} as
+with $S^{\rm pm}$ the test-mass noise PSD and $S^{\rm op}$ the optical noise PSD.
+Note that a confusion noise coming from the background of galactic binaries in the  LISA band is to be added to the instrumental noise.
+For the instrument performance defining these noise levels, we take values from~\cite{LISANoiseBudget16} (see App.~\ref{app:LISAnoise}).
+We can define a strain-like noise PSD associated to the strain-like TDI observables~\eqref{eq:defhaet} as
 \be\label{eq:defShaet}
@@
 The prefactors~\eqref{eq:scalingAET} are oscillatory and have zero-crossings at high frequencies, with the first one occuring at $f_{L}/2 = 0.06 \mathrm{Hz}$, which is why it is convenient to eliminate them to avoid $0/0$ numerical instabilities.
-However, scaling these factors out from both the signal and the noise is an approximation, as we can expect that the real delays would lead to imperfect cancellations in the vicinity of the zero-crossings.
+However, scaling these factors out from both the signal and the noise is an approximation, as we can expect that realistic delays would lead to imperfect cancellations; in the vicinity of the zero-crossings, we should therefore expect a localized loss of sensitivity.
 
@@
 
-As is well known~\cite{Cutler97}, in the low-frequency limit (also called the long-wavelength approximation), the finite-length effects of the arms of LISA disappear, and the response of the detectors falls back to the one of two LIGO-type detectors set in motion, rotated from each other by $\pi/4$.
+As is well known~\cite{Cutler97}, in the low-frequency limit (also called the long-wavelength approximation), the finite-armlength effects vanish, and the response of LISA is analogous to the response of of two LIGO-type detectors  rotated from each other by $\pi/4$ and set in motion.
 
@@
 \ee
-For even lower frequencies, when $f \ll f_{R}$ as well, we have $2\pi f R \ll 1$ and the delay term is negligible: $\exp\left[ 2 i \pi f k\cdot p_{0} \right] \simeq 1$.
+
 It can be convenient to introduce the following notations for this delay phase:
@@
 \ee
-The quantity $\Phi_{R}(\tf)$ is often the Doppler phase in the literature; it may appear to be large, but in the limit of short-lived signals it corresponds to the fixed delay between the time of arrival at the SSB and the time of arrival at the LISA constellation.
-Only the variations of the Doppler phase carry useful information about the sky position.
-The phase difference $\Delta \Phi_{R}$ subtracts this constant time delay contribution \SM{[complete for $t_{L}$]}, which is equivalent to parametrizing the signal using the time variable $t_{L}$ as defined in~\eqref{}.
+The quantity $\Phi_{R}$ is often called the Doppler phase in the literature; it may appear to be large for $f \gg f_{R}$, but we should remember that it corresponds in part simply to the fixed delay between the time of arrival at the SSB and the time of arrival at the LISA constellation.
+In the limit of short-lived coalescence signals, only the Doppler phase variation $\Delta \Phi_{R}$ carries useful information about the sky position.
 
@@
 Here, we have made apparent factors $(-2i\pi f)$ corresponding to the Fourier-domain translation of a time derivative.
-Indeed, the observables $\Delta \nu / \nu$ used here are time derivatives of the laser phase more commonly used for ground-based detectors.
+The observables $\tilde{a}$, $\tilde{e}$ are analogous to a time derivative of the strain more commonly used for ground-based detectors.
 
 We can now map these two channels to two fictitious LIGO-type detectors as follows.
-For an orthogonal detector of the same setup as the ground-based LIGO and Virgo with armlength $L_{D}$, rotated by an angle $\epsilon_{D}$ from the basis vectors $x,y$, the detector tensor is
+For an orthogonal detector of the same setup as the ground-based LIGO and Virgo with armlength $L_{D}$, rotated by an angle $\epsilon_{D}$ from the basis vectors $x,y$, the detector tensor is\S
 \begin{align}
@@
 \be
-	L_{a} = L_{e} = \frac{3L}{2} \,, \; \epsilon_{a} = \frac{2\pi}{3} \,, \; \epsilon_{e} = \frac{5\pi}{12} \,.
+	L_{a} = L_{e} = 3L \,, \; \epsilon_{a} = \frac{2\pi}{3} \,, \; \epsilon_{e} = \frac{5\pi}{12} \,.
 \ee
 Note a degeneracy $\pm \pi$ in $\epsilon_{a,e}$, reflecting a freedom in the choice of the orientation of these effective detectors.
-
-We can recover familiar pattern functions analogous to the ones of ground-based detectors (for $\psi_{L}=0$) by factoring out the effective length, defining
+The effective armlength $3L$ is a matter of convention, since it depends on an arbitrary overall scaling in~\eqref{eq:defAET}.
+
+Next, we factor out the effective length, defining
 \be\label{eq:defFapcFepc}
@@
 \ee
-which gives in terms of the LISA-frame sky position angles $\lambdaL$, $\betaL$:
+which gives in terms of the LISA-frame sky position angles $\lambdaL$, $\betaL$ (see App.~\ref{app:LISAframe}):
 \bsub\label{eq:FapcFepc}
@@
 \esub
-One can check that the expressions for the channel $e$ is obtained from the expressions for $a$ with the replacement $\lambdaL \rightarrow \lambdaL + \pi/4$.
+These functions are the familiar pattern functions of ground-based detectors (for $\psi_{L}=0$).
+One can check that the expressions for the channel $e$ are obtained from the expressions for $a$ with the replacement $\lambdaL \rightarrow \lambdaL + \pi/4$.
 
@@
 the mode transfer functions~\eqref{eq:defTlmaet} become
-\be\label{eq:transferlowfae}
-	\tilde{a}, \tilde{e} = \sum_{\ell,m} \calT^{\ell m}_{a,e} \tilde{h}_{\ell m} = \left(-6 i \pi f L\right) \exp[2i \pi f k\cdot p_{0}] F_{a,e}^{\ell m} \tilde{h}_{\ell m}\,.
-\ee
-We see that scaling out $(-6 i \pi fL)$ as in~\eqref{eq:defhaet} brings us back to strain observables.
+\begin{align}\label{eq:transferlowfae}
+	\tilde{a}, \tilde{e} &= \sum_{\ell,m} \calT^{\ell m}_{a,e} \tilde{h}_{\ell m} \nn\\
+	&= \left(-6 i \pi f L\right) \exp[2i \pi f k\cdot p_{0}] \sum_{\ell,m} F_{a,e}^{\ell m} \tilde{h}_{\ell m}\,.
+\end{align}
+We see that scaling out $(-6 i \pi fL)$ as in~\eqref{eq:defhaet} brings us back to strain-like observables.
 The only difference with the response of a ground-based observatory is then the time dependency entering $p_{0}(t)$ and the LISA frame angles $\lambda_{L}(t)$, $\beta_{L}(t)$, $\psi_{L}(t)$ (see~\eqref{eq:Lframeangles}), with the time evaluated at $t_{f}^{\ell m}$.
 
-Finally, we explain how to write a response fully analogous to ground-based detectors.
-Above we have chosen to write $\psi$ in a prefactor outside the pattern functions, but we could write $\psi$-dependent patter functions as
+Finally, we note that we could include the polarization angle in the pattern functions as follows.
+Above we have chosen to write $\psi$ as an outside prefactor, but we could write $\psi$-dependent pattern functions as
 \bsub
 \begin{align}
-	F^{+}_{a,e}(\psi) &= \cos 2\psi F^{+}_{a,e} + \sin 2\psi F^{\times}_{a,e} \,,\\
-	F^{\times}_{a,e}(\psi) &= -\sin 2\psi F^{+}_{a,e} + \cos 2\psi F^{\times}_{a,e} \,,
+	F^{+}_{a,e}(\lambdaL, \betaL, \psiL) &= \cos 2\psiL F^{+}_{a,e} + \sin 2\psiL F^{\times}_{a,e} \,,\\
+	F^{\times}_{a,e}(\lambdaL, \betaL, \psiL) &= -\sin 2\psiL F^{+}_{a,e} + \cos 2\psiL F^{\times}_{a,e} \,.
 \end{align}
@@
 \be
-	\tilde{h}_{a,e} = F^{+}_{a,e}(\psi) \tilde{h}_{+} + F^{\times}_{a,e}(\psi) \tilde{h}_{\times} \,,
-\ee
-if we ignore the delay $k \cdot p_{0}$ and treat the angles as constants.
+	\tilde{h}_{a,e} = F^{+}_{a,e}(\lambdaL, \betaL, \psiL) \tilde{h}_{+} + F^{\times}_{a,e}(\lambdaL, \betaL, \psiL) \tilde{h}_{\times} \,,
+\ee
+if we ignore the delay $k \cdot p_{0}$ and treat the angles as constants. This is the more familiar form of the instrument response used for ground-based detectors.
 
@@
 \subsection{Bayesian setting}
-\label{subsec:defbayes}
+\label{subsec:bayes}
 
@@
 This corresponds simply to performing a Bayesian analysis with~\eqref{eq:deflnL} and ~\eqref{eq:defBayes}, but setting the noise realization to zero, $n=0$.
-It allows us to accelerate the likelihood computation, as will be explained in~\ref{}.
-For exploratory parameter estimation runs, this approximation allows us to explore the structure of the likelihood with its possible parameter degeneracies (\SM{cite past use of this approx.}), and is a significant improvement over Fisher matrix estimates in that sense.
+It allows us to accelerate the likelihood computation, as will be explained in~\ref{subsec:likelihood}.
+For exploratory parameter estimation runs, this approximation allows us to explore the structure of the likelihood with its possible parameter degeneracies, and is a significant improvement over Fisher matrix estimates in that sense.
 We leave the extension of our analysis to include noise realizations for future work.
@@
 
-\SM{complete} The likelihood function that we use, with a zero noise realization, is
+The log-likelihood function that we use, with a zero noise realization, is a sum over the three independent channels $A$, $E$, $T$~\eqref{eq:defAET} rescaled according to~\eqref{eq:scalingAET}: 
 \begin{align}\label{eq:deflnLaet}
@@
 These waveforms are based on the EOB formalism~\cite{BD99, BD00}, and are calibrated on a set of numerical relativity waveforms.
-The model is limited to non-spinning systems on quasicircular orbits, but includes the most important higher harmonics in the signal:
-\be
+The model is limited to non-spinning systems on quasicircular orbits, but includes a set of higher harmonics in the signal, among the most important quantitatively:
+\be\label{eq:listmodes}
 	(\ell, m) = (2,2) ,\, (2,1) ,\, (3,3) ,\, (4,4) ,\, (5,5) \,.
@@
 The waveforms are generated in the time domain by integrating a system of ODEs, an operation requiring up to seconds of computing time for long signals.
-For this reason, we developed a reduced order model (ROM) for these waveforms, \texttt{EOBNRv2HM}, following the methods of~\cite{Puerrer14}, enabling a much faster generation of the Fourier-domain amplitudes and phases of the modes~\eqref{eq:hlmampphase}.
-
-We note that the assumption of non-spinning black holes is an important limitation to a realistic assessment of parameter uncertainties.
-In particular, it means that our results ignore the well-known degeneracy between mass ratio and spin \cite{Baird+2013}, thus underestimating the uncertainty in the recovered mass ratio.
+For this reason, we developed a reduced order model (ROM) for these waveforms, \texttt{EOBNRv2HMROM}, following the methods of~\cite{Puerrer14}, enabling a much faster generation of the Fourier-domain amplitudes and phases of the modes~\eqref{eq:hlmampphase}.
+As the total mass $M$ scales out of the problem, the parameter space is effectively only one-dimensional, with only the mass ratio $q=m_{1}/m_{2}$.
+The parameter space interpolation in this ROM is therefore significantly simpler than the models~\cite{Puerrer14, Puerrer15, Bohe+16}, which was three-dimensional as it included aligned spins.
+The inclusion of higher harmonics, however, imposes to model the relative dephasing of different modes.
+The output of the code consists of a Fourier domain amplitude and phase for each mode, sparsely sampled on $\sim 300$ frequencies.
+The faithfulness with the original waveforms was measured with \texttt{aLIGO} noise curves~\cite{UnofficialNoiseCurves18} to be $\lesssim 10^{-4}$, while the computational cost is submillisecond and subdominant with respect to other stages of our likelihood computation.
+
+We note that the assumption of non-spinning black holes is an important limitation to a realistic assessment of intrinsic parameter uncertainties.
+In particular, it means that our results ignore the well-known degeneracy between mass ratio and spin~\cite{CF94, PW95, Baird+2013}, thus underestimating the uncertainty in the recovered mass ratio.
 By ignoring misaligned spins, we also neglect the possible effects of orbital precession on parameter degeneracies.
-Nevertheless, consistently including merger, ringdown and higher harmonics already represents a significant improvement over previous studies.
-We also expect intrinsic and extrinsic parameters to correlate weakly, at least in the absence of precession, so that the conclusions drawn from Sec.~\ref{sec:SMBHPE} below, notably the analysis of possible degeneracies in the extrinsic part of the likelihood, should hold for aligned spin systems.
-\TDC{the last sentence would fit well in the conclusion.}
-
-Fast waveforms including both the aligned components of the spins and higher harmonics are becoming available in the phenomenological waveform family~\cite{London+17}, as EOB models~\cite{Cotesta+18}, and as Numerical Relativity surrogates~\cite{Varma+18}.
-ROMs to accelerate the waveform model of~\cite{Cotesta+18} are under construction, and we plan to include the effects of aligned and misaligned spins in future studies.
-\TDC{this paragraph also sounds like conclusion material.}
+Nevertheless, consistently including merger, ringdown and higher harmonics already represents a significant improvement over previous studies, as these features in the signal carry a lot of SNR by themselves~\cite{Babak+08, Thorpe+08, McWilliams+09, McWilliams+10, McWilliams+11, Klein+15}.
+In the absence of precession at least, we expect intrinsic and extrinsic parameters to correlate weakly, so that  our investigations of the extrinsic part of the likelihood should be valid for aligned spins.
 
@@
 \subsection{Likelihood computation with zero noise}
-\label{sec:likelihood}
-
-Setting the noise realization to zero in~\eqref{eq:deflnL}, having at our disposal a compact amplitude/phase respresentation of the signals allows us to compute likelihoods in an accelerated fashion.
+\label{subsec:likelihood}
+
+Since typical Bayesian analyses will need to evaluate millions of likelihood values, the computational performance of the likelihood function is crucial.
 In general, both the template and the injection in~\eqref{eq:deflnL} are given as frequency series sampled with the Nyquist criterion $\Delta f = 1/(2T)$ with $T$ the duration of the signal, and the overlaps~\eqref{eq:definnerproduct} are simply computed as discrete sums over those frequency samples.
-Although transforming~\eqref{eq:definnerproduct} in a discrete is a simple operation, the size of the frequency series alone can represent a significant cost.
-In the context of binary neutron star observations.
-In the LISA.
-\SM{complete} \TDC{this paragraph seems to need more work.
-I would start with the existing general case, then say that by using a zero-noise realization we can write the amplitude/phase representation and use a sparse grid and get a large speedup.}
-
-Applying the response as in~\ref{subsec:response} mode-by-mode on the sparsely sampled amplitude and phase generated as in~\ref{subsec:ROM}, we obtain a complete representation of both the waveform and of the instrument transfer functions~\eqref{} on a few hundred points in the Fourier domain, with the full signal being implicitly reconstructed with a standard cubic spline interpolation over frequencies.
+Although transforming~\eqref{eq:definnerproduct} in a discrete sum is a straightforward operation, the size of the frequency series alone can represent a significant cost.
+In the context of LIGO/Virgo observations, this has prompted the development of Reduced Order Quadratures (ROQ)~\cite{Smith+14, Canizares+14, Smith+16} to accelerate the likelihood computation.
+In the LISA case, the computational cost of the standard likelihood implementation depends primarily on the mass of the system and the duration of the signal.
+This cost can range from tens of milliseconds for SMBHBs and short-lived signals (a few days), to impracticable values for SBHBs signals lasting for years and sampled at high frequency.
+Setting the noise realization to zero, however, we can exploit an amplitude/phase representation of the signals, use a sparse frequency grid and get a large speedup.
+
+Applying the response as in~\ref{subsec:FDresponse} mode-by-mode on the sparsely sampled amplitude and phase generated as in~\ref{subsec:ROM}, we obtain a complete representation of both the waveform and of the instrument transfer functions~\eqref{eq:defTlmaet} on a few hundred points in the Fourier domain, with the full signal being implicitly reconstructed with a standard cubic spline interpolation over frequencies.
 Decomposing~\eqref{eq:deflnL} as
@@
 \ee
-we have to compute inner products of the form~\eqref{eq:definnerproduct}.
-Note that individual terms in~\eqref{eq:lnLoverlapstructure} can be individually large, and computing $\ln \calL$ relies on accurate cancellations.
-Decomposing further in TDI channels as in~\eqref{} and in harmonics as in~\eqref{}, we have a sum of terms with the structure (symbolically) \TDC{we need one or two more equations describing explicitly how to go from $\tilde{y}_{slr}$ to $\ln \calL$}
-\be
-	\sum_{\ell m, \ell'm'}\int \frac{df}{S_{n}} \; \calT_{1}^{\ell m} \calT_{2}^{\ell' m'} A_{1,\ell, m} A_{2, \ell' m'} e^{i(\Phi_{1,\ell m} - \Phi_{2,\ell' m'})} \,,
-\ee
-with transfer functions $\calT$, amplitudes $A$, and phases $\Phi$. \TDC{should this notation be aligned to Eq 6 and 7?}
+we have to compute inner products of the form $\left( h(\bm{\theta}) | h(\bm{\theta'}) \right)$.
+Note that individual terms in~\eqref{eq:lnLoverlapstructure} can be individually large (of the order of $\mathrm{SNR}^{2}/2$), and computing $\ln \calL$ relies on accurate cancellations between terms.
+
+Taking the likelihood~\eqref{eq:deflnLaet} decomposed in TDI channels, decomposing further in harmonics as in~\eqref{eq:defTlmaet}, we have a sum of terms with the structure (symbolically)
+\be
+	\sum_{\rm chan.}\sum_{\ell m, \ell'm'}\int \frac{df}{S_{n}} \; \calT_{1, \mathrm{chan}}^{\ell m} \calT_{2, \mathrm{chan}}^{\ell' m'} A_{1}^{\ell m} A_{2}^{\ell' m'} e^{-i(\Psi_{1}^{\ell m} - \Psi_{2}^{\ell' m'})} \,,
+\ee
+with transfer functions $\calT$ for each TDI channel, and Fourier-domain amplitudes $A$ and phases $\Psi$ (see~\eqref{eq:hlmampphase}).
+
 The difficulty of computing numerically such an overlap depends on the phase difference.
 A large phase difference causes the integrand to be very oscillatory, requiring more frequency resolution than non-oscillatory integrands; on the other hand, oscillatory integrands give raise to a small integral due to cancellation effects, while non-oscillatory integrands contribute the most.
-For a single mode, SNR terms of the type $(h|h)$ have a zero phase difference by construction, but including different modes generates cross-terms $(\ell, m) \neq (\ell', m')$ with a large phase difference.
-In a Bayesian analysis, a large fraction of the time (after burn-in) will be spent exploring signals that are rather close to the injection.
-
-Thus, one can envision several approaches to compute integrals of the type~\eqref{eq:lnLoverlapstructure}, such as factoring out a guess of the injected signal~\cite{} and ignoring cross-terms between modes.
-In this work, we use a generic numerical treatment for the integrals~\eqref{eq:lnLoverlapstructure} and keep all terms with no further approximation.
+For a single mode, $\mathrm{SNR}$ terms of the type $(h|h)$ have a zero phase difference by construction, but including different modes generates cross-terms $(\ell, m) \neq (\ell', m')$ with a large phase difference.
+
+Note that in a Bayesian analysis, a large fraction of the time (after burn-in) will be spent exploring signals that are rather close to the injection; with a single mode phase differences will be mostly small.
+In this work, we wish to include oscillatory cross-terms between modes, and we use a generic numerical treatment for the integrals~\eqref{eq:lnLoverlapstructure} and keep all terms with no further approximation.
+Other methods to accelerate likelihoods, applicable in the presence of noise, include heterodyning with a reference signal~\cite{Cornish10}, using a variable frequency resolution (multibanding)~\cite{Porter14, Vinciguerra+17}, and, as discussed before, ROQs~\cite{Smith+14, Canizares+14, Smith+16}.
+We leave for future work the generalization of our likelihood computations to accomodate for noise.
 
@@
 where in each grid interval $P_{j}$ is a cubic polynomial and $\phi_{j}$ a quadratic polynomial.
-We can then compute the elementary integrals~\eqref{eq:fastlikestructure} using a combination of integrations by parts and asymptotic expansions, as explained in App.~\ref{app:numintegration}.
-The number of grid intervals is typically several hundreds \SM{to be completed, report speed}.
+We can then compute the elementary integrals~\eqref{eq:fastlikestructure} using a combination of numerical methods.
+Integrations by parts can reduce the polynomial order in~\eqref{eq:fastlikestructure}, leaving only Fresnel integral functions to be evaluated.
+Numerical instabilities can occur in these integration by parts when the coefficients of the phase polynomial are very small or very large, in which case we resort to asymptotic expansions instead.
+
+We combine this implementation of overlap computations with our fast waveform generation~\ref{subsec:ROM} and fast treatment of the Fourier-domain response~\ref{subsec:FDresponse}.
+Since the number of overlaps to compute in~\eqref{eq:defTlmaet} is quadratic in the number of modes, likelihoods with higher harmonics are significantly more expensive.
+The final likelihood cost depends also on the number of grid intervals, that is chosen to limit spline interpolation errors:
+\begin{itemize}
+	\item for SBHBs with $h_{22}$ only: $\sim 500$ grid points, likelihood cost $\sim 2\mathrm{ms}$;
+	\item for SMBHBs with $h_{22}$ only: $\sim 300$ grid points, likelihood cost $\sim 1.2\mathrm{ms}$;
+	\item for SMBHBs with 5 modes $h_{\ell m}$: $\sim 300$ grid points, likelihood cost $\sim 15\mathrm{ms}$.
+\end{itemize}
+The accuracy of our likelihood computations is typically $|\Delta \ln \calL| \lesssim 0.2$, due to the numerical errors being magnified by the required cancellations between overlaps terms in~\eqref{eq:lnLoverlapstructure}.
 
@@
 \subsection{Bayesian sampling}
-\label{sec:samplers}
-
-For this exploratory work we take a ``naive'' \TDC{``brute force''?} approach to inferring the posterior probability distribution of the parameters, without applying specific knowledge or expectations about distributions and degeneracies.
+\label{subsec:samplers}
+
+For this exploratory work we take a ``brute force'' approach to inferring the posterior probability distribution of the parameters, without applying specific knowledge or expectations about distributions and degeneracies.
 Such assumptions might in fact lead to an incomplete exploration of the posterior probability in ways that would be difficult to recognize and diagnose.
@@
 
-Our MCMC code \texttt{PTMCMC}\footnote{https://github.com/JohnGBaker/ptmcmc} has been developed through several astrophysics projects, tested on several sampling problems, and modularly designed with the aim of minizing opportunities for errors with new applications.
-The code performs parallel tempered MCMC \cite{Swendsen+1986,Littenberg2009} with temperatures slowly adjusted to achieve nearly equal exchange rates among all pairs of adjacent temperatures.
+Our MCMC code \texttt{ptmcmc}\footnote{https://github.com/JohnGBaker/ptmcmc} has been developed through several astrophysics projects, tested on several sampling problems, and modularly designed with the aim of minizing opportunities for errors with new applications.
+The code performs parallel tempered MCMC \cite{SwendsenWang86, LittenbergCornish09} with temperatures slowly adjusted to achieve nearly equal exchange rates among all pairs of adjacent temperatures.
 Most runs here performed comparably with either 80 or 240 parallel temperature chains.
-We applied a general-purpose proposal distribution composed of a weighted set of sub-proposals, including differential-evolution steps~\cite{Vrugt+2008} (although the proposal in this case is based on each single chain, not an ensemble) and a collection of several Gaussian step draws of varying sizes scaled off the prior domain at about 0.01--1\% of its scale and weighted favor the smaller scale proposal draws.
-Effective sample sizes (ESS) were estimated from the number of post-burn-in samples, reduced by the autocorrelation length, with the burn-in size chosen to optimize \TDC{maximize?} the ESS.
+We applied a general-purpose proposal distribution composed of a weighted set of sub-proposals, including differential-evolution steps~\cite{Braak06, BraakVrugt08} (although the proposal in this case is based on each single chain, not an ensemble) and a collection of several Gaussian step draws of varying sizes scaled off the prior domain at about 0.01--1\% of its scale and weighted favor the smaller scale proposal draws.
+Effective sample sizes (ESS) were estimated from the number of post-burn-in samples, reduced by the autocorrelation length, with the burn-in size chosen to maximize the ESS.
 In most cases, the runs were continued until achieving an estimated ESS~$>2000$.
 
-For comparison and verification we also computed posterior samples using the nested sampling code \texttt{BAMBI} \cite{Graff+2012}, a variant of \texttt{MultiNest} \cite{Feroz+2009b} with additional options (unused here).
-The nested sampling algorithm evolves a set of 'live points' toward regions of high probability by iteratively resampling the lowest probability point from within an ever-narrowing region covering the set of live-point samples.
+For comparison and verification we also computed posterior samples using the nested sampling code \texttt{bambi} \cite{Graff+11}, a variant\footnote{\texttt{bambi} comes with the additional option to train a neural network to learn the likelihood, but this was not used in the present study.} of \texttt{multinest} \cite{Feroz+09}.
+The nested sampling algorithm~\cite{Skilling06} evolves a set of 'live points' toward regions of high probability by iteratively resampling the lowest probability point from within an ever-narrowing region covering the set of live-point samples; in \texttt{multinest}, this region is built as a collection of overlapping ellipsoids.
 The computation proceeds until the estimated Bayesian evidence, or marginal probability, within the region covered by the live points is smaller than some threshold.
 Posterior-distributed samples are then drawn based on the set of sub-regions and sample posterior values thereby generated.
-In our runs we used 4000 live points and a stopping evidence threshold of 1.0.
-\JGB{Sylvain needs to check these values.}
+In our runs we used 4000 live points.
 
@@
 \subsection{Fisher matrix parameter estimation}
-\label{sec:Fisher}
-
-For comparison with our Bayesian inference results we also compute estimates of parameter uncertainties using the Fisher information matrix, which despite its limitations is the common workhorse of LISA science studies to date ~\cite{Vallisneri08}.
+\label{subsec:Fisher}
+
+For comparison with our Bayesian inference results we also compute estimates of parameter uncertainties using the Fisher information matrix, which despite its limitations has been the common workhorse of LISA science studies to date ~\cite{Vallisneri08}.
 For measurements with additive Gaussian noise, as assumed here, the Fisher information matrix can be computed by
 \be
-F_{ij}=\left( \partial_{\theta_i} h| \partial_{\theta_i} h \right).
-\ee
-The inverse of this matrix estimates the covariance matrix describing parameter uncertainty.
-The derivatives are computed by second-order finite differences, with the signal inner-products expanded and computed before differencing.
-For these calculations, where numerical smoothness is more of a concern than speed we compute the inner products explicitly on a fine grid.
-
-Generally, finite difference Fisher matrices can be problematic, with small numerical defects potentially having outsized impact, so the finite-difference step size $\epsilon_i$ used in each parameter derivative $\partial_{\theta_i}h$ must be chosen carefully.
-We target the step-size to be scaled off the diagonal Fisher matrix elements by ${\epsilon_i}\approx\delta/\sqrt{F_{ii}}$ using $\delta=0.001$ \JGB{This is \texttt{fisher\_err\_target}, was it different for some runs?}  for the results shown here.
+	F_{ij} = \left( \partial_{i} h | \partial_{i} h \right),
+\ee
+where derivatives are taken with respect to one of the parameter space coordinates $\bm{\theta}$. In the Fisher matrix approach, the likelihood is approximated as
+\be
+	\ln \calL \simeq -\frac{1}{2} F_{ij} \Delta \theta^{i} \Delta \theta^{j} \,,
+\ee
+with $\Delta \theta^{i}$ the parameter deviation from the true signal.
+The inverse of the Fisher matrix, $\Sigma = F^{-1}$, is the Gaussian covariance matrix that can be used as an estimate of the true parameter uncertainties.
+
+We compute the derivatives by second-order finite differences, with the signal inner-products expanded and computed before differencing.
+For these calculations, where numerical smoothness is more of a concern than speed, we compute the inner products~\eqref{eq:definnerproduct} explicitly on a fine grid.
+Generally, finite difference Fisher matrices can be problematic, with small numerical defects potentially having outsized impact, so the finite-difference step size $\epsilon_i$ used in each parameter derivative $\partial_{i}h$ must be chosen carefully.
+We target the step-size to be scaled off the diagonal Fisher matrix elements by ${\epsilon_i}\approx\delta/\sqrt{F_{ii}}$ using $\delta=0.001$ for the results shown here.
 To achieve this we set begin with an initial choice of the $\epsilon_i$ scaled of parameter values or prior widths, then we estimate the Fisher diagonals and iterate until convergence.
 In this process we also constrain that the step is never more than $10^{-10}$ times the initial scaling.
-\JGB{There are some more special-case details, but probably not worth elaborating.
-We could also discuss how we do matrix inversion, but I don't think there is anything interesting there.}
 
@@
 
-We focus our analysis on a single set of mass parameters, representative of canonical MBHB sources expected for LISA~\cite{Proposal}: we pick a total redshifted mass $M=m_{1}+m_{2} = 2\times10^{6} M_{\odot}$ and a mass ratio $q=m_{1}/m_{2} = 3$, and place the source at a redshift $z=4$.
+We focus our analysis on a single set of mass parameters, representative of canonical SMBHB sources expected for LISA~\cite{Barausse12, Klein+15, LISA2017}: we pick a total redshifted mass $M=m_{1}+m_{2} = 2\times10^{6} M_{\odot}$ and a mass ratio $q=m_{1}/m_{2} = 3$, and place the source at a redshift $z=4$.
+The corresponding source-frame total mass is $M/(1+z) = 4\times10^{5} M_{\odot}$.
 The inclination plays an important role in deciding the importance of higher harmonics; we set it to the median value $\iota = \pi/3$.
@@
 The parameters are summarized in Table~\ref{tab:MBHBparams}, along with the SNR of these signals with and without higher harmonics.
-The different randomized orientations and the inclusion of higher harmonics both have a sizeable impact on the SNR.
-
-We note that, by contrast with current ground-based observations that are horizon-limited and that have a strong selection bias towards systems with favorable orientations (face-on or face-off), the LISA horizon for MBHB systems extends to essentially the entirety of the observable universe; it is therefore natural to consider randomized orientations.
-
-\begin{table}\label{tab:MBHBparams}
+The randomized orientations and the inclusion of higher harmonics both have a sizeable impact on the SNR.
+
+We note that, by contrast with current ground-based observations that are horizon-limited and that have a strong selection bias towards systems with favorable orientations (face-on or face-off)~\cite{}, the LISA horizon for SMBHB systems extends to essentially the entirety of the observable universe~\cite{LISA2017} \cite{}; it is therefore natural to consider randomized orientations.
+
+\begin{table}
 	\begin{tabular}{|c||c|c|}
@@
 		\hline
+		Source-frame Mass 1 ($\Msol$) 	& \multicolumn{2}{|c|}{$3\times10^{5}$} \\
+		\hline
+		Source-frame Mass 2 ($\Msol$) 	&  \multicolumn{2}{|c|}{$1\times10^{5}$}\\
+		\hline
 		Redshift 		&  \multicolumn{2}{|c|}{$4.$}  \\
@@
 	\end{tabular}
-	\caption{Parameters of the simulated MBHB mergers. \TDC{Maybe we can show
-			 these on the upper right of each corner plot instead.}}
+	\caption{Parameters of the simulated SMBHB mergers. Angles are given in the SSB-frame. The $\mathrm{SNR}$ labelled $h_{\ell m}$ is obtained including all harmonics listed in~\eqref{eq:listmodes}.}
+	\label{tab:MBHBparams}
 \end{table}
@@
   \includegraphics[width=.99\linewidth]{../plots/plot_hc_tdia_hm.pdf}
-  \caption{.}
+  \caption{Characteristic strain signal~\eqref{eq:defhc}, shown here for the TDI channel $A$, decomposed as a set of $(\ell,m)$ contributions from the modes available in the waveform model.
+The grey curve shows the characteristic noise PSD~\eqref{eq:defhc}.
+Vertical bars indicate for each mode the frequencies corresponding to the time-to-merger cuts of our pre-merger analysis as described in~\ref{subsec:MBHBacctime}, while the dots mark the merger frequency.}
   \label{fig:hctdiahm}
@@
   \includegraphics[width=.98\linewidth]{../plots/plot_smbh9_tdi.pdf}
-  \caption{\SM{TODO: rethink the visualization, maybe plot only transfer functions and Doppler phase}.}
+  \caption{Example of instrument response for System II in Table~\ref{tab:MBHBparams}.
+The first panel isolates the Doppler phase $\Delta\Phi_{R}$ as defined in~\eqref{eq:defPhiR}, while the three other panels show the TDI transfer functions for $A,E,T$.
+We show here the rescaled transfer functions for the mode 22 as defined below~\eqref{eq:defTlmaet}.}
   \label{fig:tranferSMBHCase9}
@@
 
-We illustrate the Fourier-domain signal in Fig.~\ref{hctdiahm}, with its higher harmonics, by showing mode-by-mode contributions to the characteristic strain, as well as the characteristic noise PSD, defined to be
-\bsub
+We illustrate the Fourier-domain signal in Fig.~\ref{fig:hctdiahm}, with mode-by-mode contributions to the characteristic strain, as well as the characteristic noise PSD, defined to be
+\bsub\label{eq:defhc}
 \begin{align}
@@
 \esub
-with the strain-like observables and noise PSD are defind in~\eqref{eq:defhaet} and~\eqref{eq:defShaet}.
-We represent the $A$ TDI channel (the $E$ channel being similar, and the $T$ channel negligible at low frequencies).
-This characteristic strain and associated PSD are often used for plotting~\cite{}, since one can visualize the $\SNR$ by integrating by eye under the curve with a logarithmic $x$-axis.
+with the strain-like observables and noise PSD defined in~\eqref{eq:defhaet} and~\eqref{eq:defShaet}.
+We represent the $A$ TDI channel (the $E$ channel is similar, and the $T$ channel is negligible at low frequencies).
 The full signal shows oscillatory features in frequency that are due to beatings between the harmonics that have smooth amplitudes individually.
-
-We also highlight in Fig.~\ref{fig:hctdiahm} different time intervals before the merger (see the next Section~\ref{subsec:MBHBacctime}).
-These times correspond each to a different frequency across modes, by virtue of the approximate relation between instantaneous mode frequencies in the inspiral:
-\be\label{eq:fscalinglm}
-	f_{\ell m}(t) \simeq \frac{m}{2} \times f_{22}(t) \quad (\mathrm{Inspiral})\,.
-\ee
-We also highlight the frequency at merger, also denoted peak frequency.
-Since we represent mode contributions to the TDI observables, the response is included in this Figure; we see that all modes display structure at low frequencies that corresponds to the time modulation resulting of the LISA motion.
+Since we represent mode contributions to the TDI observables, the response is included; we see that all modes display structure at low frequencies that corresponds to the time modulation resulting of the LISA motion.
 
 Fig.~\ref{fig:tranferSMBHCase9} shows the detail of the transfer functions themselves.
-The Doppler phase variation $\Delta \Phi_{R}$ is a small effect at low frequencies, being suppressed by a factor $\sim 2\pi f R$; it is also small at higher frequencies, because the chirp happens over a short time interval during which $\Phi_{R}$ does not have time to vary.
+The Doppler phase variation $\Delta \Phi_{R}$~\eqref{eq:defPhiR} is a small effect at low frequencies, being suppressed by a factor $\sim 2\pi f R$; it is also small at higher frequencies, because the chirp happens over a short time interval during which $\Phi_{R}$ does not have time to vary.
 We also show transfer functions for the mode $h_{22}$ in the three TDI channels $A,E,T$, ($T$ is suppressed at low frequencies), after factoring out the Doppler phase.
 The low-frequency features show the time-dependency of the response created by the LISA motion, while frequency-dependency in the response appears at high frequencies.
-We will decompose further these different physical effects below in Sec.~\ref{subsec:MBHBresponseapprox}
+We will investigate further these different physical effects in the response in Sec.~\ref{subsec:MBHBresponseapprox}
 
@@
 \label{subsec:MBHBacctime}
+
+\begin{figure}
+   \includegraphics[width=.99\linewidth]{../plots/tflm_smbh.pdf}
+  \caption{Time-to-frequency relation for MBHB systems I and II with $M=2\times 10^6 \; M_{\odot}$, $q=3$ and $z=4$.
+The different lines correspond to the $(\ell, m)$ harmonics available in the waveform model.
+The time-to-frequency is computed according to the defintion in $t_{f}^{\ell m}$ defined in~\eqref{eq:deftflm}, evaluated at $m/2 \times f$ to map to the same time according to~\eqref{eq:fscalinglm}.
+The vertical lines represent the time cuts explained in~\ref{subsec:MBHBacctime} and the merger frequency.}
+  \label{fig:t-f-relation-lm}
+\end{figure}
 
@@
   \includegraphics[width=.98\linewidth]{../plots/overlapmodebymode_maxfscaledlm_t.pdf}
-  \caption{.}
+  \caption{Cumulative contribution to $\mathrm{SNR}^{2}$ as defined in~\eqref{eq:defcontribSNR2}.
+The left panel shows inner products of modes with themselves, the middle panel shows cross-terms involving $h_{22}$, while the right panel shows other cross-terms.
+In all panels the black line represents the total.
+The horizontal axis gives the time to merger, with vertical lines for the cuts of~\ref{subsec:MBHBacctime}, and the dots on the right of each panel show the result for the full signal including the post-merger.}
   \label{fig:modeoverlapsoftime}
@@
 We now illustrate the accumulation of the gravitational wave signal with time, which is crucial to understand the ability of the instrument to indentify and localize MBHB signals in advance of their coalescence, as well as the requirements put on the instrumental configuration (data gaps and downlink cadence).
-As will be shown in more details in a separate publication~\cite{}, within the uncertainties of astrophysical models we can expect the bulk of MBHB signals to be detectable only a few days prior to merger, while a tail of more favorable events could be observable for significantly longer times, up to months.
+As will be shown in more details in a separate publication~\cite{MarsatBabak20} (see also~\cite{KatzLarson18}), within the uncertainties of astrophysical models we can expect the bulk of MBHB signals to be detectable only a few days prior to merger, while a tail of more favorable events could be observable for significantly longer times, up to months.
 
@@
 The time-to-frequency correspondence~\eqref{eq:deftflm} gives for each mode time as a function of frequency.
-It is accurate for the inspiral, as such a correspondence is at the heart of the SPA \SM{check SPA defined}.
-As a function it also extends formally through merger, while losing in accuracy and in physical interpretation, becoming non-monotonous with frequency (see the discussion in~\cite{MB18}).
-Due to the scaling $h_{\ell m} \propto e^{-i m \phi_{\rm orb}}$ with $\phi_{\rm orb}$ the orbital phase, a given time corresponds to different frequencies $f_{\ell m}(t) \simeq m/2 \times f_{22}(t)$ \SM{update with previous section} for different modes during the inspiral.
-At merger and this relation will be lost and the non-monotonicity of $\tf$ makes it impossible to invert $\tf$ to build $f(t)$.
-Equivalently,
-\be\label{eq:relationtflm}
-	t_{f}^{\ell m} \left( \frac{m}{2} f\right) \simeq t_{f}^{22} (f) \quad (\mathrm{Inspiral}) \,.
-\ee
-This relation is illustrated in Fig.~\ref{}. We see that it holds almost up to merger \SM{complete}, while different modes start to depart after this point.
-
-Fig.~\ref{fig:modeoverlapsoftime} shows the accumulated contributions to the total SNR of individual mode combinations.
+Due to the scaling $h_{\ell m} \propto e^{-i m \phi_{\rm orb}}$ with $\phi_{\rm orb}$ the orbital phase, a given time corresponds to different frequencies according to
+\bsub\label{eq:fscalinglm}
+\begin{align}
+	\omega_{\ell m}(t) &\simeq \frac{m}{2} \times \omega_{22}(t) \quad (\mathrm{Inspiral})\,, \\
+	t_{f}^{\ell m} \left( \frac{m}{2} f\right) &\simeq t_{f}^{22} (f) \quad (\mathrm{Inspiral}) \,,
+\end{align}
+\esub
+with $\omega_{\ell m} = \dot{\phi}_{\ell m}$ the instantaneous mode frequency in the time-domain.
+Fig.~\ref{fig:t-f-relation-lm} illustrates these relations and we use them to mark in Fig.~\ref{fig:hctdiahm} the frequencies corresponding to our time cuts, that differ for each mode, together with the merger frequency (also called the peak frequency).
+
+The relations~\eqref{eq:fscalinglm} are accurate for the inspiral, as such a correspondence is at the heart of the SPA.
+As a function, $t_{f}^{\ell m}$~\eqref{eq:deftflm} extends formally through merger, while losing in accuracy and in physical interpretation, becoming non-monotonous with frequency (see the discussion in~\cite{MB18}).
+The departure from the scaling~\eqref{eq:fscalinglm} between modes is illustrated in Fig.~\ref{fig:t-f-relation-lm}, where we see that it holds up to roughly $\sim$7min before merger, with the mode $h_{21}$ showing the earliest departure.
+
+Fig.~\ref{fig:modeoverlapsoftime} shows the cumulative contributions to the total SNR of individual mode combinations.
 For a generic signal $s = \sum_{\ell m} s_{\ell m}$ with mode contributions $s_{\ell m}$ ($s$ is any among the channels $\tilde{a}$, $\tilde{e}$, $\tilde{t}$)
-\be
+\be\label{eq:defcontribSNR2}
 	\SNR^{2} = (s|s) = \sum_{\ell m} \sum_{\ell' m'} (s_{\ell m} | s_{\ell' m'}) \,.
@@
   \includegraphics[width=.98\linewidth]{../plots/responseapprox_tdia_22.pdf}
-  \caption{Response approximations.}
+  \caption{Real and imaginary part of the transfer function $\calT^{22}_{h_{a}}$~\eqref{eq:defhaet} in different response approximations.
+The four lines stand for the four approximation levels (i)-(iv) defined in~\ref{subsec:MBHBresponseapprox}.
+Vertical lines indicate the frequencies of the time cuts of~\ref{subsec:MBHBacctime} and the peak frequency.}
   \label{fig:responseapproxtdia22}
@@
 
-The LISA instrument response, as recalled in Section~\ref{sec:response}, features both time- and frequency-dependency.
+The LISA instrument response, as recalled in Section~\ref{sec:response}, features both a time- and a frequency-dependency.
 As we will see, these features play an important role in breaking degeneracies in the parameter estimation of the source, notably allowing us to localize the source in the sky.
@@
 The time-dependency in the response mirrors the motion of the LISA constellation.
-In the low-frequency picture of~Sec.~\ref{subsec:lowfresponse}, the time dependency enters both in the Doppler phase term~\eqref{} (LISA moves in the wavefront) and in the time-dependent LISA-frame angles~\eqref{} (the LISA arms change orientation).
-However, we have seen in the previous section that the SMBH signals we consider here are quite short (less than two days).
+In the low-frequency picture of~Sec.~\ref{subsec:lowfresponse}, the time dependency enters both in the Doppler phase term~\eqref{eq:defPhiR} (LISA moves in the wavefront) and in the time-dependent LISA-frame angles~\eqref{eq:Lframeangles} (the LISA arms change orientation).
+However, we have seen in the previous section that the SMBHB signals we consider here are quite short (less than two days).
 This limits the effect of the LISA motion: between the time where the signal becomes detectable and the merger, LISA has barely changed in orientation and position.
-Moreover, the region where the signal accumulates most of its SNR is even shorter, as shown in Fig.~{}.
-
-To this end, we distinguish four different versions of the response:
+Moreover, the region where the signal accumulates most of its SNR is even shorter, as shown in Fig.~\ref{fig:modeoverlapsoftime}.
+
+To disentangle these different physical effects, we distinguish four different approximations of the response:
 \begin{enumerate}[(i)]
@@
 Case (iii), \textit{Low-$f$}, has been extensively used in the past for the analysis of MBHB signals.
-In particular, this simplification has been used in previous MLDC studies.
 Being based on the $f \ll f_{L}$ approximation, it is less inaccurate for inspiral-only signals, to which most previous studies were limited, as it is for a full IMR signal.
@@
 Contrarily to networks of ground-based observatories, we have no triangulation information from times of arrival at different detectors.
-This case is the most degenerate, and will be useful to get an anlytical understanding of approximate degeneracies occuring when using the more complicated full response (i).
+This case is the most degenerate, and will be useful to get an analytical understanding of approximate degeneracies occuring when using the more complicated full response (i).
 
@@
 We note finally a coincidence in Fig.~\ref{fig:responseapproxtdia22}: the frequency below which we see the imprint of the LISA motion and the frequency above which we see the breakdown of the long-wavelength approximation appear to be the same.
-This will not be true in general: the former is essentially a measure of the time-to-frequency correspondence $\tf$, with lower-mass signals having support and showing these features at higher frequencies, while the latter only marks the magnitude of $2\pi f L$ factors and is source-independent\footnote{The effect in the transfer functions is source-independent; however, the SNR accumulation with frequency is source-dependent and will influence the impact of these features on the parameter estimation.}.
+This will not be true in general: the former is essentially a measure of the time-to-frequency correspondence $\tf$, with lower-mass signals having support and showing these features at higher frequencies, while the latter only marks the magnitude of $2\pi f L$ factors and is source-independent\footnote{The effect in the transfer functions is source-independent; however, the total SNR will determine the impact of these features on the parameter estimation.}.
 
@@
   \centering
-  \includegraphics[width=.98\linewidth]{../plots/corner_smbh_case0_ptmcmc_22_hm.png}
-  \caption{.}
+  \includegraphics[width=.98\linewidth]{../plots/corner_smbh_case0_ptmcmc_22_hm_withzoom.png}
+  \caption{Inferred parameter posterior distribution for the MBHB system I of Table~\ref{tab:MBHBparams}, with $1$-$2$-$3$-$\sigma$ contours and the black cross indicating the true parameters.
+The result injecting and recovering with only the 22 mode is shown in blue, and the result injecting and recovering with higher harmonics is in red.
+The top right panels zoom on the masses, distance-inclination, and sky position marginalized posteriors.
+The time is centered so that $\Delta t=0$ corresponds to the injection.
+All extrinsic parameters are given in the LISA-frame as defined in App.~\ref{app:LISAframe}.
+\SM{add L subscript to parameters labels for L-frame -- maybe use D and not DL}}
   \label{fig:PEsmbh22hmCase0}
@@
   \centering
-  \includegraphics[width=.98\linewidth]{../plots/corner_smbh_case9_ptmcmc_22_hm.png}
-  \caption{.}
+  \includegraphics[width=.98\linewidth]{../plots/corner_smbh_case9_ptmcmc_22_hm_withzoom.png}
+  \caption{Same as Fig.~\ref{fig:PEsmbh22hmCase0}, for the MBHB system II of Table~\ref{tab:MBHBparams}. In extrinsic parameters, the red 2D contours are barely visible on this scale.}
   \label{fig:PEsmbh22hmCase9}
+\end{figure*}
+
+\begin{figure*}
+  \centering
+  \includegraphics[width=.98\linewidth]{../plots/row_m1m2_ptmcmc_bambi.png}
+  \caption{Comparisons between inferred posteriors obtained with our two different Bayesian samplers described in~\ref{subsec:samplers} and with the Fisher matrix approximation described in~\ref{subsec:Fisher}.
+Results are shown for the systems I and II of Table~\ref{tab:MBHBparams}, with the 22-mode only for the two left panels and with higher harmonics in the two right panels.
+Blue (22) and red (HM) show the $\texttt{ptmcmc}$ result, and green (22) and yellow (HM) the $\texttt{multinest}$ result.
+In all panels, the black ellipses show the Fisher matrix estimate for the posterior.}
+  \label{fig:PEsmbhm1m2ptmcmcbambi}
+\end{figure*}
+
+\begin{figure*}
+  \centering
+  \includegraphics[width=.98\linewidth]{../plots/row_Dinc_ptmcmc_bambi.png}
+  \caption{Same comparisons as in Fig.~\ref{fig:PEsmbhm1m2ptmcmcbambi}, for the distance and inclination parameters, zooming on the relevant regions.}
+  \label{fig:PEsmbhDincptmcmcbambi}
+\end{figure*}
+
+\begin{figure*}
+  \centering
+  \includegraphics[width=.98\linewidth]{../plots/row_sky_ptmcmc_bambi.png}
+  \caption{Same comparisons as in Fig.~\ref{fig:PEsmbhm1m2ptmcmcbambi}, for the sky position parameters, zooming on the relevant regions for inclination.
+The parameters shown are in the LISA-frame as defined in~\ref{app:LISAframe}.}
+  \label{fig:PEsmbhskyptmcmcbambi}
 \end{figure*}
@@
 
+We find that both our samplers \texttt{ptmcmc} and \texttt{multinest}, which as we recall were not really developed or tuned to this particular problem, spend a large number of likelihood evaluations, from a few $10^{7}$ to $10^{8}$, to produce a final set of posterior samples.
+This translates in reasonable computing costs thanks to our fast likelihoods, and is likely to be improvable in the future.
+
 To illustrate the role of higher harmonics in the analysis, we will present two classes of results: ``22'', where we inject and recover signals including only the dominant harmonic $h_{22}$, and ``HM'', where we both inject and recover with the higher harmonics $(h_{22}, h_{21}, h_{33}, h_{44}, h_{55})$.
@@
 A related question would be to assess the mode content necessary for waveform models to mitigate systematic biases when analyzing full GR signals.
-We leave such waveform systematics studies for future work; we only note that, as shown in Fig.~\ref{fig:modeoverlapsoftime}, the set of five modes we include in the present study is quite obviously incomplete.
+We leave such waveform systematics studies for future work; we only note that, as shown in Fig.~\ref{fig:modeoverlapsoftime}, from SNR only the set of five modes we include in the present study is quite obviously incomplete.
 
@@
 All results for MBHB systems are presented using LISA-frame parameters, as defined in App.~\ref{app:conventions}.
-We should keep in mind the difference in SNR of Cases I and II ($\SNR \simeq 945$ versus $\SNR \simeq 666$ with higher harmonics, see Table~\ref{tab:MBHBparams}), resulting from their different orientation, however their posteriors have qualitative differences that go beyond a simple scaling of the errors as $\sim 1/\SNR$ \SM{redundant ?}.
+We should keep in mind the difference in SNR of Cases I and II ($\SNR \simeq 945$ versus $\SNR \simeq 666$ with higher harmonics, see Table~\ref{tab:MBHBparams}), resulting from their different orientation, however their posteriors have qualitative differences that go beyond a simple scaling of the errors as $\sim 1/\SNR$.
 
@@
 Most notably, the sky position, if retaining the same overall bimodality $\betaL \rightarrow -\betaL$ as in Case I, shows here a curious feature: the marginalized sky posterior peaks away from the injected value.
-Looking at other panels, we see that the shifted peak corresponds to the region of high distance (and extremal inclination).
-This is a genuine featuer of the multidimensional zero-noise posterior and will be explained in Sec.~\ref{subsec:MBHBPEdegen22}.
-
-In both Cases I and II, including the higher harmonics has a major effect on the parameter recovery, much beyond what we would expect from the sole gain in total SNR shown in Table.~\ref{tab:MBHBparams} (and a scaling of statistical errors $\sim 1/\SNR$).
+Looking at correlations, we see that the shifted peak corresponds to the region of high distance (and extremal inclination).
+This is a genuine feature of the multidimensional zero-noise posterior and will be explained in Sec.~\ref{subsec:MBHBPEdegen22}.
+
+In both Cases I and II, including the higher harmonics has a major effect on the parameter recovery (as was already stressed in~\cite{Arun+07a, TriasSintes07, PorterCornish08, McWilliams+09}), much beyond what we would expect from the sole gain in total SNR shown in Table.~\ref{tab:MBHBparams} (and a scaling of statistical errors $\sim 1/\SNR$).
 The marginalized posterior of the masses is narrower, although not qualitatively different; the major change is in the extrinsic parameters.
 
-We can understand the effect of the higher harmonics by noting that, in a signal of the form $\sum_{\ell m} \sYlm(\iota, \varphi)$, measuring independently two separate contributions $(\ell m) \neq (\ell' m')$ gives us an independent measurement of the inclination $\iota$ by the relative amplitude of the two mode contributions
-
-First, including the higher harmonics in the analysis breaks the degeneracy between distance and inclination, an effect that is well known in the context of ground-based observations.
-
-Second, the phase $\varphi$ also becomes independently measured.
-
-\SM{Add sky area numbers ? Add table of $\pm 1\sigma$ intervals ?}
-
-\SM{Add comparison between the two samplers + Fisher: Cases I \& II, 22 \& hm ($\times 4$), m1-m2, D-iota, lambda-beta}
-
-\SM{Performance: waveform and likelihood costs, number of sampler evaluations}
+We can understand the effect of the higher harmonics by noting that, in a signal of the form $\sum_{\ell m} \sYlm(\iota, \varphi)h_{\ell m}$ where the $\sYlm$ all have a different dependency with the inclination $\iota$, measuring independently two separate contributions $(\ell m) \neq (\ell' m')$ gives us an independent measurement of $\iota$ by the relative amplitude of the two mode contributions.
+When only the $h_{22}$ mode is included, both the luminosity distance and inclination determine the overall signal amplitude and they are therefore degenerate.
+This is visible in Figs.~\ref{fig:PEsmbh22hmCase0} and~\ref{fig:PEsmbh22hmCase9}, where including the higher harmonics in the analysis gives a better determination of distance and inclination by more than an order of magnitude.
+
+Higher harmonics also lead to an independent determination of the phase $\varphi$, which in turn breaks degeneracies in the other angular parameters.
+In both cases I and II, the sky localization in vastly improved, with a remaining multimodality that we will discuss in the next section.
+
+We compare the Bayesian inference results obtained with our two samplers, \texttt{ptmcmc} and \texttt{multinest}, with the error estimates given by the Fisher matrix approximation~\ref{subsec:Fisher} in Fig.~\ref{fig:PEsmbhm1m2ptmcmcbambi} for the masses, Fig.~\ref{fig:PEsmbhDincptmcmcbambi} for distance/inclination, and Fig.~\ref{fig:PEsmbhskyptmcmcbambi} for the sky position.
+We find a good agreement between the two Bayesian samplers, except for Case II when using only 22-mode signals: there, \texttt{multinest} seems to fail to explore the full posterior, remaining stuck in the large-distance region of the very degenerate parameter space.
+We will futher explore the nature of this degeneracy in~Sec.\ref{subsec:MBHBPEdegen22} and in App.~\ref{app:samplingdegen}.
+The Fisher matrix approach focuses on the vicinity of the true signal's parameters and is by construction unable to handle multimodality.
+It is also insufficient in capturing the degeneracy features of Case II with $h_{22}$ signals.
+It gives good results, however, for the mass parameters and for the main mode of the posterior in non-degenerate cases, in particular when including higher harmonics.
 
@@
 
-The most remarkable feature in Figs.~\ref{fig:PEsmbh22hmCase0} and~\ref{fig:PEsmbh22hmCase9} is the existence of a degenerate mode (or secondary maximum) in the posterior distribution for the sky position $(\lambda_{L}, \beta_{L})$, located at $\beta_{L}^{*} = -\beta_{L}^{\rm inj}$, i.e.
+A remarkable feature in Figs.~\ref{fig:PEsmbh22hmCase0} and~\ref{fig:PEsmbh22hmCase9} is the existence of a degenerate mode (or secondary maximum) in the posterior distribution for the sky position $(\lambda_{L}, \beta_{L})$, located at $\beta_{L}^{*} = -\beta_{L}^{\rm inj}$, i.e.
 by reflecting the wave vector $k$ across the plane of the LISA constellation.
@@
 
-We can gain insight about degeneracies in the parameter space by considering the simple expressions representing the instrument response in the \textit{Frozen low-$f$} approximation described in Sec.~\ref{subsec:responseapprox}, appropriate for the low-frequency limit and signals short enough that the LISA motion can be neglected; we will explore in details the respective influence of these neglected effects in Sec.~\ref{sec:MBHBPEacctime}.
+We can gain insight about degeneracies in the parameter space by considering the simple expressions representing the instrument response in the \textit{Frozen low-$f$} approximation described in Sec.~\ref{subsec:MBHBresponseapprox}, appropriate for the low-frequency limit and signals short enough that the LISA motion can be neglected; we will explore in details the respective influence of these neglected effects in Sec.~\ref{sec:MBHBPEacctime}.
 
@@
 In~\eqref{eq:Flmaerepeat} the $F_{a,e}^{+,\times}$ factors of the two lines are exchanged.
-Since spin-weighted spherical harmonics~\cite{} obey the relation
+Since spin-weighted spherical harmonics~\cite{Goldberg+67} obey the relation
 \be
@@
 \end{align}
-
-Importantly, when moving away from the \textit{Frozen low-$f$} approximation, taking into account the frequency-dependence in the \textit{Frozen} response, the \textit{reflected} is the only mode remaining degenerate.
+How does this situation change when considering a more complete instrument response, moving away from the \textit{Frozen low-$f$} approximation ?
+
+When adding back the frequency-dependence while keeping LISA motionless, in the \textit{Frozen} response, the \textit{reflected} mode is the only mode remaining degenerate.
 Indeed, as noted in Sec.~\ref{subsec:FDresponse}, in~\eqref{eq:Gslr} the frequency-dependent terms feature $k\cdot n_{l}$, $k \cdot p_{s,r}$ projections, that depend only on the projection of the wave vector $k$ in the plane of LISA, invariant for $\betaL \rightarrow -\betaL$.
-The LISA motion, however, will break the degeneracy.
-
-When taking into account the LISA motion in the \textit{Low-$f$} response, the \textit{antipodal} mode is the only one that keeps pattern functions that are exactly degenerate with the injection: all other modes will move together with the time-dependent LISA frame.
-The antipodal mode is not moving, because the antipode of the true direction of the arriving signal  is defined as $k \rightarrow -k$ independently of the orientation of LISA; the only degeneracy-breaking term is then the orbital delay $k \cdot p_{0}$.
+The LISA motion, however, will break this degeneracy in general.
+
+When adding back the LISA motion while still ignoring the frequency-dependence in the response, in the \textit{Low-$f$} response, the \textit{antipodal} mode is the only one that keeps pattern functions that are exactly degenerate with the injection: all other modes will move together with the time-dependent LISA frame.
+The antipodal mode is not moving, because the antipode of the true direction of the arriving signal  is defined as $k \rightarrow -k$ independently of the orientation of LISA; the only degeneracy-breaking term is then the orbital delay $k \cdot p_{0}$ in the Doppler phase~\eqref{eq:defPhiR}.
 With the \textit{Full} response, by contrast, frequency dependency terms in~\eqref{eq:Gslr} featuring $k\cdot n_{l}$, $k \cdot p_{s,r}$ readily break this antipodal degeneracy.
 
-\begin{table}\label{tab:MBHBdegen}
+\begin{table}
 	\begin{tabularx}{.48\textwidth}{|X||c|c|c|c|}
@@
 	\end{tabularx}
-	\caption{Sketch of the degeneracy structure of the eight modes in the sky, with the cases where the degeneracy is exact (degen.), and with the qualitative effects that break the degeneracy with the injected signal: $t$-dependency, $f$-dependency, both, or only the Doppler phase $\Delta \Phi_{R}$.
+	\caption{Sketch of the degeneracy structure of the eight modes in the sky, with the cases where the degeneracy is exact (degen.), and with the qualitative effects that break the degeneracy with the injected signal: $t$-dependency, $f$-dependency, both, or only the Doppler phase $\Delta \Phi_{R}$~\eqref{eq:defPhiR}.
 This qualitative structure is the same with and without higher harmonics.}
+	\label{tab:MBHBdegen}
 \end{table}
@@
 We will see in Sec.~\ref{sec:SMBHPEacctime} that the eight-mode degeneracy pattern indeed appears when doing a pre-merger analysis; on the other hand, in results Figs.~\ref{fig:PEsmbh22hmCase0} and~\ref{fig:PEsmbh22hmCase9} for full IMR signals, out of the eight possible sky modes only the \textit{reflected} mode~\eqref{eq:degenreflection} survives.
-We will investigate in detail in Sec.~\ref{sec:SMBHPEacctime} how time-dependence and frequency-dependence in the response break part, but not all, of the degeneracies.
+We will investigate in detail in Sec.~\ref{sec:MBHBPEacctime} how time-dependence and frequency-dependence in the response break part, but not all, of the degeneracies.
 
@@
 
-\subsection{Apparent biases in the absence of higher harmonics}
+\subsection{Apparent sky position bias for 22-mode signals}
 \label{subsec:MBHBPEdegen22}
@@
   \end{minipage}
-  \caption{Sky position posterior in presence of degeneracies, zooming close to the injected value.
-The left panel shows the complete likelihood, while the right panel shows the simplified likelihood~\eqref{eq:simplelikelihoodhm} with a motionless LISA and fixed masses and time.
-The results of \texttt{PTMCMC} (blue) and \texttt{MultiNest} are superimposed.
+  \caption{Sky position posterior for System II in Table~\ref{tab:MBHBparams} in presence of strong degeneracies, zooming close to the injected value and discarding the secondary mode in the sky.
+The left panel shows the complete likelihood, while the right panel shows the simplified likelihood~\eqref{eq:simplelikelihoodhm} in the \textit{Frozen, low-$f$} approximation.
+and fixed masses and time.
+The results of \texttt{ptmcmc} (blue) and \texttt{multinest} (green) are superimposed.
 The black cross is the injected value, and the grey cross is the analytic prediction~\eqref{eq:soldegensky} for the offset peak created by degeneracies.}
@@
 
-In this section, we investigate the cause of the degeneracies appearing in the posterior of Case~II shown in~\ref{subsec:SMBHPE}.
+In this section, we investigate the cause of the apparent bias in sky position in the posterior distribution of System~II when injecting and recovering with 22-mode only waveforms, as shown in Figs.~\ref{fig:PEsmbh22hmCase9} (and in the middle left panel of~\ref{fig:PEsmbhskyptmcmcbambi}).
+The posterior for the sky forms a peak that appears shifted from the injection.
+This feature is surprising: since we set the noise realization to zero, the maximum likelihood is by construction reached at the injection.
+It occurs only when 22-mode only signals are used, and is present for both Bayesian samplers, although $\texttt{ptmcmc}$ and $\texttt{multinest}$ differ noticeably for this case with $\texttt{ptmcmc}$ recovering more parameter volume.
+We will show that we can understand this feature as a projection effect in the multidimensional degenerate posterior.
 Understanding the structure of these features could prove useful to inform Bayesian samplers, for instance by building adapted jump proposal allowing a much faster mixing of MCMC chains than naive methods.
-a simplified form of the extrinsic likelihood and the structure of its degeneracies.
-
-An important clue about the appearance of these degeneracies is the length of the observable signal.
-As shown in Fig.~\ref{}, for our example SMBH systems the SNR accumulates in a matter of days before merger, which means that there is only little impact of the modulations caused by the motion of the detector on the TDI transfer functions~\eqref{}.
-If we can neglect entirely the motion of the detector (which is not quite true as we will see later on), then we are in a regime where the information about the sky position of the source comes essentially from the pattern functions affecting this high-SNR signal.
-Using the low-frequency approximation (see~\ref{}) and the time-at-LISA $t_{L}$ as a time variable, there is no frequency dependence in the response.
-A further simplification occurs when pinning the masses and the time of coalescence to their true values.
-We found indeed that these parameters correlate only weakly to the other ones, so that we can hope to understand the degeneracy features by restricting the analysis to the six extrinsic parameters $(D_{L}, \iota, \varphi, \lambdaL, \betaL, \psiL)$.
-\TDC{we could make the connection with a similar decoupling in LIGO/Virgo, which is actually what makes low-latency sky localization possible.}
-
-Under all these simplifying assumptions, the likelihood function becomes purely extrinsic, and takes the form of trivial geometric factors multiplying a constant overlap that can be precomputed.
-Using the subscript $\mathrm{inj}$ for the injected signal, the likelihood~\eqref{eq:defliklihood} is
+
+In the following we will aim at explaining these features analytically by building a simplified model for the response. In this \textit{simplified likelihood} approximation, we: 
+\begin{itemize}
+	\item pin the masses and the coalescence time to their injected values;
+	\item use the \textit{Frozen low-$f$} response, ignoring the LISA motion and the frequency-dependence in the response.
+\end{itemize}
+The first point allows us to focus only on the extrinsic parameters, as we find weak correlations between intrinsic and extrinsic parameters.
+Such a decoupling is also at play in LIGO/Virgo, and makes low-latency sky localization possible~\cite{SingerPrice15}.
+The second point is partly justified by the fact that our signal is short, as shown in Fig.~\ref{fig:modeoverlapsoftime}: for the MBHB system that we picked as an example the SNR accumulates in a matter of days, with $\mathrm{SNR}=10$ reached $\sim 40 \,\mathrm{h}$ before merger.
+Neglecting the high-frequency features is in fact a stronger approximation, as will be discussed in Sec.~\ref{subsec:MBHBacctimebreakdegen}.
+We focus on this limit merely to be able to work with the simple analytic expressions for the response given in~\ref{subsec:lowfresponse}.
+
+Under all these simplifying assumptions, the likelihood becomes pure function of the extrinsic parameters $(D, \iota, \varphi, \lambdaL, \betaL, \psiL)$, and takes the form of trivial geometric factors multiplying constant mode overlaps that can be precomputed.
+The \textit{Low-$f$} approximation is detailed in Sec.~\ref{subsec:lowfresponse}.
+The $T$-channel is negligible in this approximation, so that the likelihood~\eqref{eq:deflnLaet} is
 \be
@@
 \ee
-From the low-frequency response \eqref{eq:aelowfmodes}, pinning the time-at-LISA~\eqref{} \SM{to be completed for $t_{L}$} so that we can ignore the factor $\exp [2 i \pi f k\cdot p_{0}]$ as a pure constant corresponding to a redefinition of time, we obtain \JGB{Should we use notation $\Phi_R$ here and in \eqref{eq:aelowfmodes}?}
+Having fixed the intrinsic parameters and the time, in~\eqref{eq:transferlowfae} the modes $h_{\ell m}$ are fixed as well, and it is convenient to introduce the notation
+\be\label{eq:innerproductlmlpmp}
+	\left\langle \ell m | \ell' m' \right\rangle = 4\mathrm{Re} \int \frac{df}{S_{n}^{a,e}} \left( 6 \pi f L\right)^{2} \tilde{h}_{\ell m} \tilde{h}_{\ell' m'}^{*} \,.
+\ee
+for mode overlaps that are constant factors and can be precomputed for a given system. Here, the noise PSD $S_{n}^{a,e}$ is given by~\eqref{eq:Snaet} and is identical between the channels $a$ and $e$.
+
+Using the results of Sec.~\ref{subsec:lowfresponse}, ignoring in~\eqref{eq:transferlowfae} the factor $\exp [2 i \pi f k\cdot p_{0}]$ as a pure constant corresponding to a redefinition of time, we obtain
 \begin{align}\label{eq:simplelikelihoodhm}
-	\ln \calL &= -\frac{1}{2} \sum_{\ell m} \sum_{\ell' m'}   \left[ \left( s_{a}^{\ell m} - s_{a, \mathrm{inj}}^{\ell m}\right) \left( s_{a}^{\ell' m'} - s_{a, \mathrm{inj}}^{\ell' m'}\right)^{*} \right.
+	\ln \calL &= -\frac{1}{2} \sum_{\ell m} \sum_{\ell' m'} \left[ \left( \vphantom{s_{a}^{\ell' m'} } s_{a}^{\ell m} - s_{a, \mathrm{inj}}^{\ell m}\right) \left( s_{a}^{\ell' m'} - s_{a, \mathrm{inj}}^{\ell' m'}\right)^{*} \right.
 \nn\\
 	& \qquad \left.
-+ \left( s_{e}^{\ell m} - s_{e, \mathrm{inj}}^{\ell m}\right) \left( s_{e}^{\ell' m'} - s_{e, \mathrm{inj}}^{\ell' m'}\right)^{*} \right] \left\langle \ell m | \ell' m' \right\rangle \,,
-\end{align}
-where we introduced the following extrinsic signal functions \JGB{Since neither $D_{a,e}$ nor $P_{\ell m}$ are defined in the main text, may be better to write in terms of a version of $F$ as in \eqref{eq:defFapcFepc}}
++ \left( \vphantom{s_{a}^{\ell' m'} } s_{e}^{\ell m} - s_{e, \mathrm{inj}}^{\ell m}\right) \left( s_{e}^{\ell' m'} - s_{e, \mathrm{inj}}^{\ell' m'}\right)^{*} \right] \left\langle \ell m | \ell' m' \right\rangle \,,
+\end{align}
+where we introduced
 \be\label{eq:defsaelm}
-	s_{a,e}^{\ell m} = \frac{1}{d} \frac{2}{3L} D_{a,e} : P_{\ell m} \,,
-\ee
-with $d = D_{L} / D_{L}^{\rm inj}$ the dimensionless ratio of luminosity distances.
-The prefactors $\left\langle \ell m | \ell' m' \right\rangle$ \sout{are pure normalization constants, and} represent mode overlaps as
-\be\label{eq:innerproductlmlpmp}
-	\left\langle \ell m | \ell' m' \right\rangle = 4\mathrm{Re} \int \frac{df}{S_{n}^{a,e}} \left( 3 \pi f L\right)^{2} \tilde{h}_{\ell m} \tilde{h}_{\ell' m'}^{*} \,.
-\ee
-Here, the noise PSD $S_{n}^{a,e}$ is given by~\eqref{} and is identical between the channels $a$ and $e$.
-Having fixed the masses and the time, these overlaps are to be computed with the modes $\tilde{h}_{\ell m}^{\rm inj}$ of the injected signal.
+	s_{a,e}^{\ell m} = \frac{1}{d} F_{a,e}^{\ell m} \,,
+\ee
+with $d = D / D_{\rm inj}$ the dimensionless ratio of luminosity distances, and with mode transfer functions $F_{a,e}^{\ell m}$ given in~\eqref{eq:Flmae}.
 
@@
 \ee
-The functions~\eqref{eq:defsaelm} take the explicit form
+The functions~\eqref{eq:defsaelm} take for $(\ell m)=(22)$ the explicit form
 \begin{align}
@@
 \esub
-with a common prefactor \JGB{This seems to need to  be multiplied by -6 to get what I have in my notes. Trace to change in def of $s_{a,e}$?}
+with a common prefactor
 \be\label{eq:sigmafactorrho}
@@
 Finding points in the parameter space that are degenerate with the injection, i.e.
-with $\ln \calL \simeq 0$, amounts finding choices for the parameters which obtain the same values as the injection for both quantities $\sigma_{+}$ and $\sigma_{-}$.
+with $\ln \calL \simeq 0$, amounts to finding choices for the parameters which obtain the same values as the injection for both quantities $\sigma_{+}$ and $\sigma_{-}$.
 
 We can now use the simplified response written in the form~\eqref{eq:sigmapm}-\eqref{eq:sigmafactorrho} to look for symmetries and degeneracies.
-First, we note the following manifest symmetry: changing simultaneously
-\be\label{eq:symmetryresponse}
-	\betaL \rightarrow -\betaL\,, \quad \iota \rightarrow \pi - \iota \,, \quad \psiL \rightarrow \pi - \psiL
-\ee
-leaves the likelihood exactly invariant.
+First, it is easy to check in these new notations that the transformations~\eqref{eq:degenreflection} and~\eqref{eq:degenlambdapi2} indeed leave the likelihood exactly invariant.
 There is also a symmetry based on exchanging $\iota \leftrightarrow \theta_{L}$.
 While leaving $\sigma_-$ unchanged, this conjugates the factor inside brackets for $\sigma_{+}$, which can be compensated using the phase terms $\varphi \pm \lambda_{a}$.
-If $\Phi = \mathrm{Arg} \left[ t_{\theta}^{4} e^{-2 i \psiL} + t_{\iota}^{4} e^{2 i \psiL} \right]$, we obtain the symmetry
+If $\Phi = \mathrm{Arg} \left[ t_{\theta}^{4} e^{-2 i \psiL} + t_{\iota}^{4} e^{2 i \psiL} \right] $, we obtain the symmetry
 \begin{align}
@@
 
-\JGB{Don't have to resort to limits.}
-\sout{We can now explain the degenerate features we found as certain limits in this simplified likelihood.
-In the face-on or face-off limit $\iota \rightarrow 0$, $|t_{\iota}| \ll 1$.
-This amounts to the $(\ell,m)=(2,2)$ contribution dominating the $(\ell, m)=(2,-2)$ contribution.
-Neglecting the terms in~\eqref{eq:sigmapm} that are quartic in $t_{\iota}$, we have}
-%\bsub\label{eq:sigmapmapprox}
-%\begin{align}
-%	\sigma_{+} &\simeq \rho e^{2i(\varphi - \psiL)} t_{\theta}^{4} e^{-2i \lambda_{a}} \,, \\
-%	\sigma_{-} &\simeq \rho e^{2i(\varphi - \psiL)} e^{2i \lambda_{a}} \,,
-%\end{align}
-%\esub
-
 Beyond these discrete symetries, since likelihood function dependence on the six extrinsic parameters is funneled through just two complex functions, we should expect a two-dimensional degenerate subspace.
-Indeed we can explicitly find a general solution for parameter values that solve $\sigma_{\pm}((D_{L}, t_\iota, t_\theta \varphi, \lambda_a, \psiL)=\sigma_{\pm,{\rm inj}}$, which we just sketch here, as needed to explain features of the degeneracies.
+Indeed we can explicitly find a general solution for parameter values that solve $\sigma_{\pm}((D, t_\iota, t_\theta \varphi, \lambda_a, \psiL)=\sigma_{\pm,{\rm inj}}$, which we just sketch here, as needed to explain features of the degeneracies.
 
@@
 \be
-	r(\lambda_a, t_\iota, t_\theta, \psiL) = \frac{\sigma_{+}}{\sigma_{-}} \,,
-\ee
-makes clear that $r=r_{\rm inj}$ provides one complex condition on 4 real unknown parameters, eliminating parameters $D_{L}$ and $\varphi$ while retaining the features of the full degenerate subspace.
+	r(\lambda_a, t_\iota, t_\theta, \psiL) = \frac{\sigma_{+}}{\sigma_{-}} = \frac{t_{\theta}^{4}  + t_{\iota}^{4} e^{4 i \psiL}}{1+ t_{\theta}^{4} t_{\iota}^{4} e^{4 i \psiL}} e^{-4i \lambda_{a}}\,,
+\ee
+makes clear that $r=r_{\rm inj}$ provides one complex condition on 4 real unknown parameters, eliminating parameters $D$ and $\varphi$ while retaining the features of the full degenerate subspace.
 Going further, $|r|^2=|r_{\rm inj}|^2$ yields one condition on 3 parameters, eliminating $\lambda_a$.
-In fact, with a little rearrangement, this can be written as a quadratic expression for either $t_\iota^4$ or $t_\theta^4$ for an explicit solution.
-Depending on the value of $|r_{\rm inj}|^2$, values of $\psiL$ and the independent variable ($t_\theta$ or $t_i$) can always be chosen to guarantee a real, positive solution.
-
+In fact, with a little rearrangement, this can be written as a quadratic expression for either $t_\iota^4$ or $t_\theta^4$ given the other variable ($t_\theta$ or $t_\iota$) and $\psiL$.
 Given a solution for $(t_\iota,t_\theta,\psi_L)$, the rest of the solution then proceeds backwards.
-An explicit expression for commensurate $\lambda$ comes from solving $r=r_{\rm inj}$, and then $D_L$ and $\varphi$ are obtained from solving, e.g., $s_+=s_{+,{\rm inj}}$.
-
-In the limit that the independent variable is given a value far from unity (near polar $\iota$ ot $\beta_L$), then the solution for the dependent variable approaches a fixed constant special values $t_{\iota,\theta}^4={|r_{\rm inj}|^{\pm1}}$, with leading order corrections proportional to  $t_{\theta,\iota}^{4}$.
-A large part of the degenerate subspace volume then tends to found with parameters $\iota$ ot $\beta_L$ near these special values with corresponding constraints on $\lambda$ coming from the solution for the complex argument of $r$.
-
-In terms of the original parameters then, if we consider inferences with the parameter $\iota$ that is not near the source equator, then we find a significant part of the degenerate maximum-likelihood parameter space volume near the special sky positions
-\bsub
-\begin{align}
-  \beta_L&\approx{\beta_L}^{*}=\pm\left(\frac{\pi}{2} - 2 \mathrm{Arctan} \; \left| r_{\rm inj}\right|^{1/4} \right)\,, \\
-  \lambda_a&\approx {\lambda_a}^{*}=\frac{\pi}{6} -\frac{1}{4} \mathrm{Arg} \; r_{\rm inj} \;\mathrm{mod}\; \frac{\pi}{2} \,,\\
-  \rho e^{2i(\varphi-\psiL)} &\approx e^{-2i\lambda_{a}^{*}} \sigma_{-}^{\rm inj} \,.
-\end{align}
-\esub
-With the $\lambdaL \rightarrow \lambdaL + k \pi/2$ and $\betaL \rightarrow - \betaL$ symmetries, we arrive at eight sky positions where the degeneracy occurs, with an exact degeneracy for the pair of parameters $(\varphi, \psiL)$, as well as an approximate degeneracy for the pair $(d, \iota)$ as long as we remain in the regime $t_{\iota}^{4} \ll 1$ or $t_{\iota}^{4} \gg 1$ (which is quite extended, thanks to the quartic power).
-For each of these sky positions built from~\eqref{eq:soldegensky}, many different values of $(\varphi, \psiL)$ and $(d, \iota)$ produce a waveform very close to the injection, resulting in apparent peaks in the marginalized posterior distribution for the sky position, located at the special sky positions $(\lambda_{a}^{*}, \theta_{L}^{*})$, which are offset from the injected value.
-
-Without special tuning, the injected sky position which led to the given values of $s_{a,\mathrm{inj}}$ and $s_{e,\mathrm{inj}}$ is correspondingly likely to have been near one of these sky locations, but it may not have been.
-In those cases the posterior\ldots
-
-Similarly, considering inferred locations away from the LISA-constellation plane, we find a significant part of the degenerate parameter space near special values for inclination and correspondingly for $\lambda_L+\varphi$
-\bsub
-\begin{align}
-	\iota&\approx\iota^{*} = 2 \mathrm{Arctan} \; \left| r_{\rm inj}\right|^{1/4} \,, \\
-	\lambda_a-\psiL&\approx (\lambda_{a} - \psiL)^{*} = -\frac{1}{4} \mathrm{Arg} \; r_{\rm inj} \;\mathrm{mod}\; \frac{\pi}{2} \,\\
-        \rho e^{2i\varphi} &\approx e^{-2i (\lambda_{a} - \psi_{L})^{*}} \sigma_{-}^{\rm inj} \,.
-\end{align}
-\esub
-
-%finding $(\lambda_{a}^{*}, \theta_{L}^{*})$ verifying
-%\be\label{eq:degenconditionsky}
-%	\frac{\sigma_{+}}{\sigma_{-}} (\lambda_{a}^{*}, \theta_{L}^{*}) = r_{\rm inj} \,,
-%\ee
-%with the approximation~\eqref{eq:sigmapmapprox} for the left-hand side, guarantees that we can then match both $\sigma_{+}$ and $\sigma_{-}$ of the injection with the condition
-%\be\label{eq:degenconditionprefactor}
-%	\rho e^{2i(\varphi - \psiL)} e^{2i \lambda_{a}^{*}} = \sigma_{-}^{\rm inj}\,.
-%\ee
-%In the limit $|t_{\iota}| \ll 1$,
-%\be
-%	\frac{\sigma_{+}}{\sigma_{-}} \simeq t_{\theta}^{4} e^{-4i\lambda_{a}} \,,
-%\ee
-%so that~\eqref{eq:degenconditionsky} is verified for special positions in the sky at (coming back to the original variables $(\lambdaL, \betaL)$)
-%\bsub\label{eq:soldegensky}
-%\begin{align}
-%\beta_{L}^{*} &= \frac{\pi}{2} - 2 \mathrm{Arctan} \; \left| r_{\rm inj}\right|^{1/4} \,, \\
-%	\lambda_{L}^{*} &= \frac{\pi}{6} -\frac{1}{4} \mathrm{Arg} \; r_{\rm inj} \;\mathrm{mod}\; \frac{\pi}{2} \,.
-%\end{align}
-%\esub
-\JGB{Keep something from this?}
-The other condition~\eqref{eq:conditionprefactor} gives then a single phase constraint for $\varphi - \psiL$, and a single amplitude constraint for $\rho(d, \iota, \theta_{L}^{*})$.
-The limit $\iota \rightarrow \pi$ (where the $(2,-2)$ contribution dominates the $(2,2)$) is very similar, as can be seen most easily by the change of variable $\iota \rightarrow \pi - \iota$, with $-\psiL$ and $\pi - \theta_{L}$ playing the roles of $\psiL$ and $\theta_{L}$.
-
-%We can further find the same mechanism, with a few differences, by exchanging the roles of $(\iota, \varphi)$ and $(\theta_{L}, \lambda_{a})$, as follows.
-In the limit $\theta_{L} \rightarrow 0$ with $|t_{\theta}| \ll 1$, we have this time
-%\be
-%	\frac{\sigma_{+}}{\sigma_{-}} \simeq t_{\iota}^{4} e^{-4i(\lambda_{a} - \psi_{L})} \,,
-%\ee
-%which gives in the same way a degeneracy if
-%\bsub\label{eq:soldegeniotaphi}
-%\begin{align}
-%	\iota^{*} &= 2 \mathrm{Arctan} \; \left| r_{\rm inj}\right|^{1/4} \,, \\
-%	(\lambda_{a} - \psiL)^{*} &= -\frac{1}{4} \mathrm{Arg} \; r_{\rm inj} \;\mathrm{mod}\; \frac{\pi}{2} \,,
-%\end{align}
-%\esub
-%with a condition on $\rho$ and $\varphi$ that is now
-%\be
-%\rho e^{2i\varphi} e^{2i (\lambda_{a}^{*} - \psi_{L}^{*})} = \sigma_{-}^{\rm inj} \,.
-%\ee
-\JGB{Keep something from this?}
-Thus, we find again a degenerate pattern, with the difference that \eqref{eq:soldegeniotaphi} imposes a condition on the pair $\lambda_{a} - \psiL$ instead of $\lambda_{a}$, and the extra condition imposes a condition on $\varphi$ instead of the pair $\varphi - \psiL$.
+An explicit expression for commensurate $\lambda$ comes from solving $r=r_{\rm inj}$, and then $D$ and $\varphi$ are obtained from solving, e.g., $\sigma_-=\sigma_{-}^{\rm inj}$.
+
+We can now understand the degeneracies we saw by considering limits of the ratio $r$.
+For instance, when $t_{\iota}^{4} \ll 1$, we have simply $|r| \simeq t_{\theta}^{4}$.
+This means that in this limit all values of $\psiL$ are allowed, with $\theta_{L}$ fixed to a specific value.
+Similarly, for $t_{\iota}^{4} \gg 1$, we have $|r| \simeq t_{\theta}^{-4}$.
+A large part of the degenerate subspace volume then tends to be found with a parameter $\beta_L$ near these special values fixed by the modulus constraint, with a special value of $\lambda_{L}$ coming from the complex argument constraint.
+By contrast, intermediate values of $t_{\iota}^{4}$ will not give as much parameter space allowed for the degeneracy, as illustrated by the case $t_{\iota} = 1$ ($\iota = \pi/2$): in that case $|r|=1$ irrespective of $\theta_{L}$ and $\psiL$, and no solution exists if $|r_{\rm inj}| \neq 1$.
+
+In terms of the original parameters, these special sky positions are
+\bsub\label{eq:soldegensky}
+\begin{align}
+  \betaL &\approx \beta_{L}^{\dagger} = \pm\left(\frac{\pi}{2} - 2 \mathrm{Arctan} \; \left| r_{\rm inj}\right|^{1/4} \right)\,, \\
+  \lambdaL &\approx \lambda_{L}^{\dagger} =\frac{\pi}{6} -\frac{1}{4} \mathrm{Arg} \; r_{\rm inj} \;\mathrm{mod}\; \frac{\pi}{2} \,,\\
+  \rho e^{2i(\varphi-\psiL)} &\approx e^{-2i(\lambda_{L}^{\dagger} - \pi/6)} \sigma_{-}^{\rm inj} \,.
+\end{align}
+\esub
+With the $\lambdaL \rightarrow \lambdaL + k \pi/2$ and $\betaL \rightarrow - \betaL$ symmetries corresponding to the eight-mode sky symmetry discussed in Sec.~\ref{subsec:MBHBPEdegen}.
+The degeneracy for the pair of parameters $(\varphi, \psiL)$ is exact, and the constraint $\rho = |\sigma_{-}^{\rm inj}|$ gives an approximate degeneracy for the pair $(d, \iota)$ as long as we remain in the regime $t_{\iota}^{4} \ll 1$ or $t_{\iota}^{4} \gg 1$ (which is quite extended, thanks to the quartic power).
+Thus, for each of these sky positions built from~\eqref{eq:soldegensky}, many different values of $(\varphi, \psiL)$ and $(d, \iota)$ produce a waveform very close to the injection, resulting in apparent peaks in the marginalized posterior distribution for the sky position, located at the special sky positions $(\lambda_{L}^{\dagger}, \theta_{L}^{\dagger})$, which are offset from the injected value.
+
+Similarly, considering the limits $t_{\theta}^{4} \ll 1$, $t_{\theta}^{4} \gg 1$, we can expect a significant part of the degenerate parameter space near special values for inclination and correspondingly for $\lambda_L+\varphi$
+\bsub\label{eq:soldegeniotaphi}
+\begin{align}
+	\iota&\approx\iota^{\dagger} = \frac{\pi}{2} \mp \left(\frac{\pi}{2} - 2 \mathrm{Arctan} \; \left| r_{\rm inj}\right|^{1/4} \right) \,, \\
+	\lambda_L-\psiL&\approx (\lambda_{L} - \psiL)^{\dagger} = \frac{\pi}{6} -\frac{1}{4} \mathrm{Arg} \; r_{\rm inj} \;\mathrm{mod}\; \frac{\pi}{2} \,\\
+        \rho e^{2i\varphi} &\approx e^{-2i ((\lambda_{L} - \psi_{L})^{\dagger} - \pi/6)} \sigma_{-}^{\rm inj} \,.
+\end{align}
+\esub
 
@@
 
-We illustrate our findings by running a Bayesian parameter estimation for the simplified likelihood~\eqref{eq:simplelikelihood22} with the two samplers \texttt{MultiNest} and \texttt{PTMCMC} described in Sec.~\ref{}.
-Since the inner products~\eqref{eq:innerproductlmlpmp} are constants computed only once, the external factors become trivially inexpensive, and the likelihood is very fast to compute.
-We take advantage of this to run \texttt{PTMCMC} for a very large number of steps, ensuring that we do not miss degenerate regions of the parameter space.
-The resulting posterior distribution is extremely degenerate, as shown in Fig.~\ref{fig:PEsmbh22hmSimpleLikeCase9}.
-The sampler \texttt{MultiNest} fails to resolve this structure, stopping after having only explored the degenerate region $\iota \rightarrow 0$ and $\iota \rightarrow \pi$ and the eight sky positions~\eqref{eq:soldegensky}.
-In accordance with~\eqref{eq:degenconditionsky}-\eqref{eq:degenconditionprefactor}, this corresponds to lines in $\varphi \pm \psiL = \mathrm{const}$.
-The sampler \texttt{PTMCMC} finds more degenerate regions, in particular it finds similar structures with the roles of $\iota$ and $\betaL$ exchanged (as shown notably by the $(d,\iota)$ and $(d,\betaL)$ panels), as well as intermediate regions connecting all these features.
-
-Fig.~\ref{fig:PEsmbh22hmSimpleLikeCase9} also shows the result obtained when including all available harmonics~\eqref{} in~\eqref{eq:simplelikelihoodhm}.
-The degenerate structures collapse to leave only eight modes corresponding to $\betaL \rightarrow - \betaL$ and $\lambdaL \rightarrow \lambdaL + k\pi/2$.
-We found that this degeneracy-breaking occurs already when including a single subdominant harmonic $h_{21}$.
-
-We can now come back to explaining the results of Sec.~\ref{}, and in particular Fig.~\ref{}, obtained with the true likelihood function for the mode $h_{22}$.
-First, the difference between the two Bayesian samplers has been made more extreme by going to the more degenerate likelihood~\eqref{eq:simplelikelihood22}, allowing us to attribute their disagreement to an apparent shortcoming of \texttt{MultiNest}\footnote{We recall that we only used standard values for the metaparameters \SM{[complete]} for this sampler, and we do not exclude that it would be possible to improve results with minor changes.} for resolving extended degenerate regions.
-Focusing on the sky position, we zoom close to the injection in Fig.~\ref{fig:Skydownsmbh22hmSimpleLikeCase9}.
-We compare here the posterior distribution generated with the true likelihood with the one obtained with the simplified likelihood~\eqref{eq:simplelikelihood22}, for both samplers.
-We also superimpose the prediction~\eqref{eq:soldegensky} for the degeneracy-induced peak.
-We see that, although ,
+We illustrate our findings in  by running a Bayesian parameter estimation for the simplified likelihood~\eqref{eq:simplelikelihood22} with our two samplers, \texttt{multinest} and \texttt{ptmcmc}.
+The results display the degeneracy structure that we discussed in this Section.
+Focusing on the sky position in the vicinity of the true source parameters, Fig.~\ref{fig:Skydownsmbh22hmSimpleLikeCase9} contrasts the posterior distribution obtained with the \textit{full} and the \textit{simplified} likelihoods.
+Although differences are visible, we see that both show a similar peak shifted from the true signal's sky position, that agrees well with our prediction~\eqref{eq:soldegensky}.
+More details on the full posterior distribution with the simplified likelihood are given in App.~\ref{app:samplingdegen}.
 
@@
 
-The rate at which parameter information accumulates LISA merger observations is crucial for planning strategies for multi-messenger observations and in guiding the latency demands of the LISA down-link and data processing.
-We explore how parameter information accumulates on approach to merger by performing parameter estimation studies temporally truncated signals, cut at points where the accumulated SNR is about $\{10,42,166\}$ corresponding to $\{1/64,1/16,1/4\}$ of the total SNR=$666.0$, and times $\{41 \mathrm{hr}, 2.5 \mathrm{hr}, 7 \mathrm{min}\}$ before merger.
-
-\begin{figure*}
-  \centering
-  \begin{minipage}{.48\linewidth}
-      \includegraphics[width=.99\linewidth]{../plots/t-of-F.pdf}
-   \end{minipage}
-   \begin{minipage}{.48\linewidth}
-      \includegraphics[width=.99\linewidth]{../plots/t-of-SNR.pdf}
-   \end{minipage}
-  \caption{\SM{Redo left plot, right plot redundant with mode-by-mode SNR} Relations of time, Fourier frequency, and accumulated SNR for $M=2\times 10^6$,$q=3$ merger at $z=4$.
+The rate at which parameter information accumulates LISA merger observations is crucial for planning strategies for multi-messenger observations and in guiding the latency demands of the LISA down-link and data processing~\cite{ArmitageNatarajan02, DalCanton+19}.
+In this Section, we explore how parameter information accumulates on approach to merger by performing parameter estimation studies with temporally truncated signals, as explained in~\ref{subsec:MBHBacctime}: the signals are cut at points where the accumulated SNR is about $\{10,42,166\}$ corresponding to $\{1/64,1/16,1/4\}$ of the total SNR=$666.0$, and times $\{41 \mathrm{hr}, 2.5 \mathrm{hr}, 7 \mathrm{min}\}$ before merger.
+As shown in Fig.~\ref{fig:modeoverlapsoftime}, most of the SNR accumulates over the last few minutes of signal.
+
+%\begin{figure*}
+%  \centering
+%  \begin{minipage}{.48\linewidth}
+%      \includegraphics[width=.99\linewidth]{../plots/t-of-F.pdf}
+%   \end{minipage}
+%   \begin{minipage}{.48\linewidth}
+%      \includegraphics[width=.99\linewidth]{../plots/t-of-SNR.pdf}
+%   \end{minipage}
+%  \caption{\SM{Redo left plot, right plot redundant with mode-by-mode SNR} Relations of time, Fourier frequency, and accumulated SNR for $M=2\times 10^6$,$q=3$ merger at $z=4$.
 The left panel shows times computed from $\tilde h(f)$ using Eq.~eref{eq:t-of-f}, and are approximately common among the waveform component modes up to minutes before merger.
@@
 The stars show points where the SNR has reached $\{1/4,1/16,1/64\}$ of the full-signal total, where we have studied parameter estimate for temporally incomplete signals.}
-  \label{fig:t-f-snr-Case9}
-\end{figure*}
-
-We implement these temporal cuts in the Fourier domain, using Eq.XXXX, the $t(f)$ relation inferred from the smoothly chirping harmonic mode components of the waveforms as part of the LISA response.
-This is a good approximation for temporal cuts, and consistent among the various modes for our cuts, which are sufficiently early before merger, as shown in Fig.~\ref{fig:t-f-snr-Case9}.
-
-The right panel of the figure also shows the times before merger at which a given value of SNR is reached.
-In this case the signal begins being relevant, reaching SNR~1 about 34 days before merger, but reaches a plausible first-detectability threshold about 41 hours before merger.
-Most of the SNR accumulates over the last few minutes of signal.
-
-In Fig.~\ref{fig:smbhCornerZoomHMCase9} and Fig.~\ref{fig:MollweidesmbhTserieshmCase9} we show how parameter information accumulates at these time points.
-Individual masses are within a factor of 2 at first detection and better than $10\%$ by 41 hours before merger.
-Distance within a factor of 2 at 41 hours and within roughly $10\%$ at 2.5 hours prior to merger.
-Merger time is estimated to about an hour at first detection, and within a few minutes by 2.5 hours before merger.
-Sky position...
+%  \label{fig:t-f-snr-Case9}
+%\end{figure*}
+
+We implement these temporal cuts in the Fourier domain, using the time-frequency correspondence~\eqref{eq:deftflm}, adapted for higher harmonics following~\eqref{eq:fscalinglm}.
+Our cuts are sufficiently early before merger for this relation to be a good approximation for temporal cuts, and consistent among the various harmonics, as shown in Fig.~\ref{fig:t-f-relation-lm}.
+
+In Fig.~\ref{fig:smbhCornerZoomHMCase9} we show how parameter information accumulates at these time points.
+The results are shown for Case II, including the higher harmonics in the signal.
+Individual masses are loosely constrained, within a factor 2, at first detection; the chirp mass combination is better determined.
+They start to be individually constrained to better than $10\%$ by 2.5 hours before merger.
+The luminosity distance is poorly determined, within a factor of 4, when reaching the detection threshold 41 hours before merger; the constraint improves to roughly $10\%$ at 2.5 hours prior to merger.
+Merger time is estimated with an uncertainty of roughtly 2hrs at first detection, which improves to a few minutes by 2.5 hours before merger.
+
+The corresponding sky position posteriors are shown in Fig.~\ref{fig:MollweidesmbhTseriesCase922hm}, using LISA frame angles~\ref{eq:Lframeangles}.
+The lines show the 1-2-3-$\sigma$ uncertainty contours on the sky, for each of our time cuts as well as for the full, post-merger signal, with the two panels differing by the inclusion of higher harmonics.
+We see clearly, in the pre-merger analysis, the 8-modes sky degeneracy introduced in~\eqref{eq:degenreflection}, \eqref{eq:degenlambdapi2}.
+The different cuts give us an idea of the continuous evolution from a badly determined sky position at first, when reaching detection, to an 8-modes degeneracy structure, to finally only two modes surviving when reaching merger, the true injection and the \textit{reflected} sky position.
 
@@
    \end{minipage}
-  \caption{.}
+  \caption{Refinement of parameter inference over time for system II.
+The posteriors shown correspond to the time cuts defined in~\ref{subsec:MBHBacctime} at $\mathrm{SNR}/64$ (green), $\mathrm{SNR}/16$, $\mathrm{SNR}/4$ and post-merger (blue)., for the masses (left), distance/inclination (center), and time/phase (right).
+We inject and recover with higher harmonics.
+The black cross shows the true parameters.}
   \label{fig:smbhCornerZoomHMCase9}
@@
   \includegraphics[width=.8\linewidth]{../plots/skymap_Lframe_tseries_ptmcmc_hm.png}
-  \caption{.}
-  \label{fig:MollweidesmbhTserieshmCase9}
+  \caption{Refinement of the LISA-frame sky position inference over time for system II.
+The posteriors shown correspond to the time cuts defined in~\ref{subsec:MBHBacctime} at $\mathrm{SNR}/64$ (green), $\mathrm{SNR}/16$, $\mathrm{SNR}/4$ and post-merger (blue).
+For the top figure, we inject and recover with a 22-mode only waveform, while for the bottom figure we inject and recover with higher harmonics.
+The black cross is the injected sky position; the apparent offset in the 22-mode analysis corresponds to the phenomenon highlighted in~\ref{subsec:MBHBPEdegen22}.
+The blue contour for the analysis with the full signal corresponds to the posterior of Fig.~\ref{fig:PEsmbh22hmCase9} (where blue and red were used for 22 and HM); in the bottom figure with higher harmonics, is almost reduced to two dots on this scale.}
+  \label{fig:MollweidesmbhTseriesCase922hm}
 \end{figure*}
@@
   \includegraphics[width=.99\linewidth]{../plots/lnL_skymodes_approx_hm.pdf}
-  \caption{for different response approximations (from left to right: \textit{Full}, \textit{Frozen}, \textit{Low-$f$}), with 22-mode only (top) and with higher harmonics (bottom).
-A very negative $\ln\calL$ values means that this point is excluded by the inference.
-\SM{do we need a table of post-merger lnL values ?}}
+  \caption{Log-likelihood~\eqref{eq:deflnLaet} at degenerate modes in the sky summarized in Table~\ref{tab:MBHBdegen}, for different response approximations (from left to right: \textit{Full}, \textit{Frozen}, \textit{Low-$f$}), with 22-mode only (top) and with higher harmonics (bottom).
+The log-likelihood is shown as a function of the maximal frequency where we cut the 22-mode signal (with the rescaling~\eqref{eq:fscalinglm} for the other modes), which can be thought of as accumulating signal with time (the vertical lines represent the time cuts of~\ref{subsec:MBHBacctime} and the merger).
+The lines represent the sky positions listed in Table~\ref{tab:MBHBdegen}: in terms of LISA-frame parameters full or dashed stands for the reflection in latitude, and the color stands the $pi/2$-shifted longitudes.
+The dashed black curve corresponds to the \textit{reflected} position, and the dashed red to the \textit{antipodal} position.
+At the injection (solid black), $\ln\calL$ is zero by construction while it is decreasing as a function of $f$ for other sky positions.
+Reaching a very negative $\ln\calL$ means that this sky position is excluded by the signal accumulated thus far.}
   \label{fig:lnLskymodesapprox}
@@
 
-We have found that pre-merger analysis of the previous section made apparent the degeneracy pattern that can be predicted analytically from the structure of the \textit{Frozen, low-$f$} response approximation.
-Analyses of IMR signals as in~\ref{sec:MBHBPE} show that only the \textit{reflected} mode~\eqref{eq:degenreflection} survives post-merger.
+We have found that pre-merger analysis of the previous section made apparent the 8-modes degeneracy pattern that can be predicted analytically from the structure of the \textit{Frozen, low-$f$} response approximation (see~\eqref{eq:degenreflection}, \eqref{eq:degenlambdapi2}).
+Analyses of IMR signals as in~\ref{subsec:SMBHPEfull} show that only the \textit{reflected} mode~\eqref{eq:degenreflection} survives post-merger.
 In this Section we explain how and why this transition occurs.
 
-We have already presented a discussion of different qualitative effects in the instrument response in Sec.~\ref{MBHBresponseapprox} and Fig.~\ref{responseapproxtdia22}: one hand hand, the motion of LISA leaves an imprint (the time-dependency) on the low frequencies, where the time-to-frequency map is steep enough that a short interval in frequency maps to a large interval of time; on the other hand, the breakdown of the long-wavelength approximation leaves an imprint (the frequency-dependency) at high frequencies.
+We have already presented a discussion of different qualitative effects in the instrument response in Sec.~\ref{subsec:MBHBresponseapprox} and Fig.~\ref{fig:responseapproxtdia22}: one hand hand, the motion of LISA leaves an imprint (the time-dependency) on the low frequencies, where the time-frequency map is steep enough that a short interval in frequency maps to a large interval of time; on the other hand, the breakdown of the long-wavelength approximation leaves an imprint (the frequency-dependency) at high frequencies.
 Here we look at the quantitative importance of these features on the inference as a function of frequency (or as a function of time).
@@
 In Fig.~\ref{fig:lnLskymodesapprox} we show log-likelihood values obtained with either the \textit{Full}, \textit{Frozen} or \textit{Low-$f$} response, with and without higher harmonics, for each of the eight modes in the sky described by~\eqref{eq:degenreflection}-\eqref{eq:degenlambdapi2}.
-The likelihood~\eqref{eq:deflnLaet} is computed with the same response approximation is used for the injection and for the template.
+The likelihood~\eqref{eq:deflnLaet} is computed with the same response approximation for the injection and for the template.
 The results are shown as a function of frequency in a cumulative sense: we compute the likelihood by accumulating signal up to the frequency shown in the $x$-axis.
@@
 
-A value of $\ln \calL = 0$ means that the template signal with shifted parameters is identical to the injection, while a very negative $\ln \calL$ means that these parameter values are excluded by the inference.
+A value of $\ln \calL = 0$ means that the template signal with shifted parameters is identical to the injection, while a very negative $\ln \calL$ means that these parameter values are excluded in the inference.
 The injection (full black line) has always $\ln \calL = 0$ by construction.
-The \textit{reflected} sky mode~\eqref{eq:degenreflection} is the dashed black line, and the \textit{antipodal} sky mode~\eqref{eq:degenanitpodal} is the dashed red line.
+The \textit{reflected} sky mode~\eqref{eq:degenreflection} is the dashed black line, and the \textit{antipodal} sky mode~\eqref{eq:degenantipodal} is the dashed red line.
 
@@
 Comparing the two rows of the figure, we see that the higher harmonics have an effect here, in making the elimination of secondary modes happen earlier in frequency (or time).
-This is expected as they start to contribute significantly to the SNR (see Fig.\ref{fig:modeoverlapsoftime}) and reach higher in frequency than the $h_{22} harmonic$ according to~\eqref{eq:fscalinglm}.
-
-The \textit{Low-$f$} response case is essentially a superposition of the previous two response approximations.
-The early sky localization is quite poor and relies on the LISA motion.
-Getting closer to merger We stress that the .
-Overall, including the higher harmonics does not change this qualitative picture, despite an earlier onset of frequency-dependent degeneracy-breaking features.
+This is expected as they start to contribute significantly to the SNR (see Fig.\ref{fig:modeoverlapsoftime}) and reach higher in frequency than the $h_{22}$ harmonic according to~\eqref{eq:fscalinglm}.
+
+The \textit{Full} response case is essentially a superposition of the previous two response approximations, with a successive onset of the time-dependency and frequency-dependency to break degeneracies.
+We note that overall, including the higher harmonics does not change this qualitative picture, despite an earlier onset of frequency-dependent degeneracy-breaking features.
 
 Limitations of such explorations should be made clear: by using a pointwise estimate, we cannot make statistical statements and are missing volume effects discussed in Sec.~\ref{subsec:MBHBPEdegen22}; by transforming the angular parameters with an analytical prescription while keeping the other fixed, we are in a sense overconstraining the degeneracy.
-In particular, it is conceivable that one could find a better match to the injection in the vicinity of a degenerate sky mode, by adjusting slightly all parameters.
+In particular, it is conceivable that one could find a better match to the injected signal in the vicinity of a degenerate sky mode, by adjusting slightly all parameters.
 
@@
 %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-\section{Stellar origin black holes}
-\label{sec:SOBH}
-
--- SOBH system, SNR sufficient for clean analysis + Fisher - stacked SNR contours
--- Performance: waveform and likelihood costs, number of sampler evaluations
-
-Stellar-origin black-hole (SOBH) mergers have been recently recognized as an
-important source of LISA detections \cite{Sesana16} although full parameter
-estimation studies are not available yet. In this section we present two case
-studies of parameter inference for such sources using the formalism and methods
-described earlier. The parameters of the simulated mergers are taken from the
-Radler LISA data challenge and are listed in Table \ref{table:sobh_params}.
-\begin{table}
-	\begin{tabular}{|c||c|c|}
-		\hline
-		Radler entry 		& 1 	& 12   \\
-		\hline
-		Mass 1 ($\Msol$) 	& 21.4	& 63.5 \\
-		\hline
-		Mass 2 ($\Msol$) 	& 20.1	& 45.1 \\
-		\hline
-		Distance (Mpc) 		& 49.1	& 318  \\
-		\hline
-		LISA SNR 			& 26.6	& 11.9 \\
-		\hline
-	\end{tabular}
-	\caption{Parameters of the simulated SOBH mergers. \TDC{Maybe we can show
-			 these on the upper right of each corner plot instead.}}
-	\label{table:sobh_params}
-\end{table}
-
-We caution the reader that our purpose here is to show the feasibility of
-Bayesian inference for SOBH sources in LISA using the methodology presented in
-this paper. It is by no means a final answer on how to do so in an optimal or
-convenient way.  For this reason, the problem is posed in a simplified
-way here: in particular, the SOBH signal is assumed to merge roughly at the end of
-the observation epoch, and the finite observation time is enforced by
-converting it to a minimum analyzed Fourier frequency via the time-frequency
-relation of a simple Newtonian inspiral. In a more realistic situation, the
-merger time would likely be free to range from the beginning of the observation
-to years after its end, and the frequency of the signal at the beginning of the
-observation would be another sampling parameter.
-
-The waveform model used to work with SOBH systems is the same phenomenological
-approximant \texttt{EOBNRv2HMROM} used earlier, extended at low frequency with a post-Newtonian
-frequency-domain waveform. In practice, because LISA only observes the early
-inspiral of SOBH systems, only the post-Newtonian waveform plays a role here.
-We do not expect radiation modes beyond the quadrupole to give a significant
-contribution to the parameter estimates, so we only include the $(2,2)$ mode
-for simplicity. We do not include any effect arising from black-hole spins.
-
-As done in the previous section, we apply our parameter inference procedure to
-these systems using both BAMBI and parallel-tempered MCMC sampling, and we
-numerically evaluate the Fisher matrix at the parameters of the injection.
-
-Due to the duration of an SOBH inspiral in the LISA band, and the complete lack
-of merger and ringdown, the chirp mass $\Mchirp$ is measured much better than
-other mass parameters and the component masses have a much stronger correlation
-than usually found with ground-based gravitational-wave detectors. In order for
-the sampling to converge robustly in a reasonable time, we find it is necessary
-to express the mass parameters in terms of $\Mchirp$ and symmetric mass ratio
-$\eta$, and a relatively tight prior around the injected parameters is required
-as well.
-
-In order to generate these results, our PTMCMC sampler requires between
-$5\times 10^5$ and $3\times 10^6$ steps with 100 temperatures, taking on the
-order of days on a single CPU core without careful optimization. The BAMBI
-sampling, on the other hand, completes in tens of minutes only.
-
-The resulting posterior distributions are shown in Fig.~\ref{fig:sobh_corner_1}
-and \ref{fig:sobh_corner_12}. We find close agreement between the BAMBI and
-PTMCMC samplers. The full joint distributions are fairly complex, but the
-complexity is mostly restricted to multimodalities and degeneracies in the
-angular parameters only. The typical degeneracy between inclination and distance
-is evident. Masses, merger time, distance, inclination and sky location, which
-are arguably the most astrophysically useful parameters, have simple marginal
-distributions with a well-defined mode and close to Gaussian, even at these
-moderate SNR values. The Fisher matrix estimates represent the marginal
-distributions of these parameters at least to the order of magnitude. In some
-cases the Fisher estimates approximate very precisely even the joint
-distributions of pairs of these parameters.
-
-\begin{figure*}
-	\includegraphics[width=\textwidth]{../plots/corner_sobh_tdc1_ptmcmc_bambi}
-	\caption{Inferred parameter posterior distribution for SOBH system 1.}
-	\label{fig:sobh_corner_1}
-\end{figure*}
-
-\begin{figure*}
-	\includegraphics[width=\textwidth]{../plots/corner_sobh_tdc2_ptmcmc_bambi}
-	\caption{Inferred parameter posterior distribution for SOBH system 12.}
-	\label{fig:sobh_corner_12}
-\end{figure*}
-
-Such posterior distributions should not be taken literally as expectations of
-LISA's capability for SOBH systems. One reason is the simplified treatment of
-the finite observation time described earlier. A second reason is the assumption
-of zero black-hole spins, which artificially reduces the width of the posterior
-distribution due to the strong degeneracy between spins and symmetric mass
-ratio \cite{Baird+2013}.
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+
+\section{Stellar mass black holes}
+\label{sec:SBHB}
 
@@
 \subsection{Signals and transfer functions}
-\label{subsec:signaltransferSOBH}
+\label{subsec:signaltransferSBHB}
 
@@
   \includegraphics[width=.98\linewidth]{../plots/plot_sobh1_tdi.pdf}
-  \caption{\SM{TODO: rethink the visualization, maybe plot only transfer functions}.}
-  \label{fig:PEsmbh22hmCase0}
+  \caption{Example of instrument response for the SBHB system I in Table~\ref{tab:SBHBparams}.
+The first panel isolates the Doppler phase $\Delta\Phi_{R}$ as defined in~\eqref{eq:defPhiR}, while the three other panels show the TDI transfer functions for $A,E,T$.
+We show here the transfer functions for the 22-mode contribution to $a$, $e$, $t$ as defined in~\eqref{eq:defTlmaet}.}
+  \label{fig:transferSBHBcaseI}
 \end{figure*}
 
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-
-\section{Discussion}
-\label{sec:discussion}
-
-- Summary/conclusions
-
-----------------
-Other questions:
-- special orientations/sky locations ?
-- Fisher matrix qualification - explore parameter choices
-- continuously increase SNR - include on these examples
-- continuously accumulate information over time
-- full exploration of parameter space for all of the above...
+Stellar-mass black-hole binaries (SBHB) inspirals have been recently recognized as a potentially
+important source of LISA detections \cite{Sesana16, Vitale16}. Full Bayesian parameter
+estimation studies for these LISA signals have not yet been developed, but the Fisher approach has been used in~\cite{Sesana16, Vitale16, Nishizawa+16a, Nishizawa+16b}.
+In this section we present two case
+studies of parameter inference for such sources using the formalism and methods
+described earlier, while limiting the analysis to the extrinsic parameters (excluding the masses, and fixing the time to coalescence).
+The parameters of the simulated mergers were drawn from the Radler LISA Data Challenge (LDC), which itself followed the population of~\cite{Sesana16}, and are listed in Table~\ref{tab:SBHBparams}.
+The system I is a moderately massive system, placed at a favorable close distance; with an SNR of $26.6$ it would be at the tail of the most favorable detections that we can hope for~\cite{Sesana16, Gerosa+19, Moore+19}.
+The system II is a higher-mass, moderate-SNR system.
+Both systems happen to be close to the ecliptic plane.
+
+Fig.~\ref{fig:transferSBHBcaseI} illustrates, for the system I in this table, the instrumental transfer functions~\eqref{eq:defTlmaet} as well as the Doppler phase~\eqref{eq:defPhiR}.
+Contrasting with the SMBHB case, we see that the transfer functions have much more structure in this case.
+Both the time-dependency and the frequency-dependency in the response lead to oscillatory features in the response.
+The Doppler phase~\eqref{eq:defPhiR}, that was negligible for our short SMBHB systems, represents now tens of radians.
+
+%The TDI channel $T$, negligible as well for low-frequency SMBHB signals, makes here a small contribution at high frequencies.
+
+\begin{table}
+	\begin{tabular}{|c||c|c|}
+		\hline
+		Identifier 		& I 	& II   \\
+		\hline
+		\hline
+		Mass 1 ($\Msol$) 	& $21.44$ & $63.51$ \\
+		\hline
+		Mass 2 ($\Msol$) 	& $20.09$ & $45.15$ \\
+		\hline
+		Redshift 		& $0.01$ & $0.07$ \\
+		\hline
+		Lum. Distance (Mpc) 		&  $49.1$ & $318.2$\\
+		\hline
+		Inclination (rad) 		& $0.65$ & $1.92$ \\
+		\hline
+		Phase (rad) 			& $0$	& $0$ \\
+		\hline
+		Ecliptic longitude (rad) 			& $3.44$	& $5.66$ \\
+		\hline
+		Ecliptic latitude (rad) 			& $-0.074$	& $-0.055$ \\
+		\hline
+		Polarization (rad) 			& $1.74$	& $2.20$ \\
+		\hline
+		Time to merger (yr) 			& $2$	& $2$ \\
+		\hline
+		\hline
+		LISA SNR $h_{22}$			& 26.6 & 11.9 \\
+		\hline
+	\end{tabular}
+	\caption{Parameters of the simulated SBHB inspiral signals.}
+	\label{tab:SBHBparams}
+\end{table}
+
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+
+\subsection{Bayesian inference of extrinsic parameters}
+\label{subsec:PESBHB}
+
+We caution the reader that our purpose here is to show the feasibility of
+Bayesian inference for SBHB sources in LISA using the methodology presented in
+this paper, and to illustrate the qualitative features of the posterior distribution for extrinsic parameters, in relation to our previous investigations for SMBHB systems.
+To this end, the problem we pose is simplified
+in several ways: we assume that we know the signal to be present, bypassing the challenges of detection itself~\cite{Moore+19}, we fix the masses and the time to merger to the injected value, and we neglect spins.
+The parameter recovery of the intrinsic parameters, masses, spins, is important to address as one can expect strong degeneracies in mass ratio and spin for these inspiral signals; it will be tackled in a forthcoming publication~\cite{Toubiana+20}.
+
+The waveform model used to work with SBHB systems is a standard post-Newtonian
+frequency-domain waveform (we use a version of the \texttt{TaylorF2} approximant as implemented in \texttt{LAL}~\cite{lal}). In practice, because LISA only observes the early
+inspiral of SBHB systems, the signal is well within the domain of applicability of the post-Newtonian approximation~\cite{Mangiagli+18}.
+In the deep inspiral, harmonics beyond the dominant quadrupole give a negligible
+contribution to the signal, so we only include the $(2,2)$ mode for simplicity.
+
+As done in the previous section, we apply our parameter inference procedure to
+these systems using both the \texttt{multinest} and \texttt{ptmcmc} samplers, and we
+numerically evaluate the Fisher matrix at the parameters of the injection. In order to generate these results, our \texttt{ptmcmc} sampler requires between
+$5\times 10^5$ and $3\times 10^6$ steps with 100 temperatures, taking on the
+order of days on a single CPU core without careful optimization. The \texttt{multinest}
+sampler, on the other hand, completes in a few hours only.
+
+The resulting posterior distributions are shown in Fig.~\ref{fig:sbhb_corner_1}
+and \ref{fig:sbhb_corner_12}. We find close agreement between the \texttt{multinest} and
+\texttt{ptmcmc} samplers.
+The joint distributions for extrinsic parameters are fairly simple, with extended degeneracies occuring only in the angular parameters $(\varphi, \psi)$.
+Thanks to the complexity of the instrument transfer functions shown in Fig.~\ref{fig:transferSBHBcaseI} carrying a lot of information, the sky position determination is very good.
+The typical degeneracy between inclination and distance
+is evident, as there is no degeneracy breaking from higher harmonics.
+Even though the SNR values are much lower than for our SMBHB systems, the Fisher matrix estimates represent quite well the marginal
+distributions of most parameters.
+We note, however, some notable deviation from Gaussianity for the system II: the sky position is almost bimodal, with an approximate symmetry across the ecliptic plane (we recall here that the ecliptic plane is not the plane of the LISA instrument that was used to report SMBHB results), owing probably to the localization being close to this plane; and the distance and inclination posteriors show a long tail extending way beyond the Fisher estimate.
+
+\begin{figure*}
+	\includegraphics[width=0.9\textwidth]{../plots/corner_sobh_tdc1_extr_ptmcmc_bambi}
+	\caption{Inferred parameter posterior distribution for the SBHB system I of Table~\ref{tab:SBHBparams}, with $1$-$2$-$3$-$\sigma$ contours and the black cross indicating the true parameters.
+The black ellipses show the same contours for the Fisher matrix approximation.
+Waveforms include only the 22 mode.
+The results obtained with outr two different samples are shown, $\texttt{ptmcmc}$ in blue and $\texttt{multinest}$ in green.
+The initial frequency is fixed to obtain a time-to-merger of 2 years.
+All extrinsic parameters are given in the SSB-frame.
+\SM{use D and not DL}}
+	\label{fig:sbhb_corner_1}
+\end{figure*}
+
+\begin{figure*}
+	\includegraphics[width=0.9\textwidth]{../plots/corner_sobh_tdc2_extr_ptmcmc_bambi}
+	\caption{Same as Fig.~\ref{fig:sbhb_corner_1}, for the SBHB system II of Table~\ref{tab:SBHBparams}.}
+	\label{fig:sbhb_corner_12}
+\end{figure*}
+
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+
+\section{Summary and discussion}
+\label{sec:summarydiscussion}
+
+In this work, we have explored the Bayesian parameter estimation for full, inspiral-merger-ringdown SMBHB signals including higher harmonics, albeit neglecting the spin degrees of freedom, and we have also produced posterior distributions for the extrinsic parameters of SBHB signals.
+
+We improved the speed of likelihood computations by combining several ingredients: fast IMR Fourier-domain waveforms, a fast Fourier-domain treatment of the instrument response and accelerated overlap computations between amplitude/phase signals in the absence of noise.
+The resulting likelihood costs, of about 15ms for waveforms with higher harmonics and 1ms for waveforms including only the dominant harmonic, allowed us to perform easily Bayesian parameter estimation on high-SNR signals requiring up to $10^{8}$ likelihood evaluations with standard Bayesian samplers.
+
+Using these tools to simulate the parameter recovery of two examples of SMBHB signals, we produced multidimensional posterior distributions.
+While our posterior distributions for masses are not really representative in the absence of know degeneracies between the mass ratio and the spins, we encountered and investigated two different kinds of degeneracies in the extrinsic parameter space.
+
+Considering first IMR signals, we found a \textit{two-fold} degeneracy for the sky position corresponding to the \textit{reflected} sky localization with respect to the plane of the LISA constellation (opposite sign of the LISA-frame latitude).
+We also found that, when limiting the signals to their dominant harmonic $h_{22}$, strong degeneracies in distance-inclination could occur with a posterior distribution in the sky appearing shifted from the injected value.
+We gave an analytic explanation of this degeneracy by eliminating the time and frequency dependencies in the response, reducing it to an explicitly degenerate form.
+However, including higher harmonics in the analysis strongly breaks those degeneracies, and leads to a much better determination of both distance-inclination and sky position.
+Although the reflected sky localization survives with higher harmonics, we also note that the posteriors appear as much more Gaussian in all extrinsic parameters; using a simplified signal with only the dominant harmonic significantly complicates the posterior sampling.
+
+Secondly, considering pre-merger analyses to investigate how information accumulates with time, we found an \textit{eight-fold} degeneracy for the sky position, obtained by duplicating the original and the reflected sky positions with $\pi/2$, $\pi$, $-\pi/2$ shifts in the LISA-frame longitude.
+For our example systems, the posterior only collapses on the two-fold degeneracy, noticed previously for IMR signals, minutes before merger.
+By decomposing the LISA instrument response in different effects, separating its time-dependency from its frequency-dependency, we could pinpoint the role of both effects.
+The time-dependency of the response, consequence of the motion of LISA, leaves a moderate imprint at low frequencies, penalizing moderately other modes in the sky but leaving the \textit{antipodal} sky position degenerate.
+By contrast, the frequency-dependency of the response, not included in previous parameter studies, is responsible for the degeneracy breaking and the collapse from eight to two modes in the sky.
+Note that the relative importance of the time and frequency dependency of the response will change for heavier or lighter systems.
+
+We also explored the recovery of extrinsic parameters of SBHB signals (fixing the masses and spin parameters as well as the merger time), using the same methodology as for SMBHB signals to arrive at a likelihood cost of $\sim 2 \mathrm{ms}$.
+Investigating two example systems, we found that the distribution of extrinsic parameters is mostly Gaussian, with for one case some bimodality in the sky position found for one source lying close to the ecliptic plane, and an extended tail in the distance posterior.
+
+Comparing the parameter estimation results obtained with our two samplers, $\texttt{ptmcmc}$ and $\texttt{multinest}$, we found in general a good agreement.
+We noted however that \texttt{multinest} can fall short of exploring the full degeneracies in parameter space, in the degenerate case of a signal without higher harmonics; this limitation is further illustrated by the example of a fully degenerate response.
+We also found that using Fisher matrix estimates would not reproduce Bayesian results in general: degeneracies in the sky would not be covered by a Fisher computation that is intrinsically unimodal, and neither would the specific features of the $22$-mode degenerate posteriors.
+That degeneracies occur even with the high SNRs of our SMBHB signals is remarkable, and reminds us that the usual statement of the validity of the Fisher matrix approximation in the high-SNR limit is very dependent on the morphology of the signal.
+In many respects, the moderate-SNR posteriors for our SBHB systems are more Gaussian than our high-SNR SMBHB $22$-mode posteriors.
+
+Our exploratory work calls for a number of natural extensions that will help us to better understand the future capabilities of the LISA instrument.
+
+First, we only investigated a handful of sources.
+In particular, we chose our SMBHB systems to be representative of the bulk of the population; they are not ``golden sources'', i.e.
+they are not part of the tail of lower-redshift systems that would provide us with the most interesting science outputs with LISA, notably in terms of electromagnetic counterparts.
+It is likely that the parameter recovery here will be different, with a longer detectable signal and the motion of LISA playing a role in the premerger localization.
+We also limited ourselves to a single redshifted mass, while the morphology of the signals will change strongly between the two ends of the LISA spectrum, from intermediate mass black holes $M\sim 10^{3} M_{\odot}$ to massive systems $M>10^{7} M_{\odot}$.
+
+In the future, it will be important to incorporate more physics in our waveform models.
+In this study our signals were limited to non-spinning, circularized systems.
+The aligned components of the spin will play a role by allowing degeneracies between masses and spin magnitudes, but we can expect that they will not affect qualitatively the posterior distributions for extrinsic parameters.
+Precession and eccentricity, however, will change the harmonic structures of the signals and play an important role.
+
+A limitation of our method is that, in order to accelerate the likelihood computation, set the noise realization to zero.
+Although we do not expect that the noise realization will cause drastic changes, with in particular the structure of degeneracies of signals in parameter space remaining relevant, fast likelihoods for noisy data would be a welcome addition to our methods.
+Analyzing signals with a noise injection will also become more important when considering more realistic instrumental noise models.
+
+We also note that our sampling tools themselves (\texttt{multinest} and \texttt{ptmcmc}) were not tailored for the problem at hand; it is very possible that better sampling algorithm would require much less evaluations of the likelihood.
+The understanding we gained of the parameter space degeneracies will allow us to inform the sampler with e.g.
+tailored jump proposals, effectively telling MCMC chains where to look for degenerate signals in parameter space.
+For instance, the LIGO-Virgo parameter estimation tools~\cite{Veitch+14} make use of an analytical knowledge of degeneracies.
+
+Finally, we recall that many scientific questions about LISA rely on parameter estimation tools, and would benefit from proper Bayesian analyses with realistic signals: pre-merger localization of sources and advance warnings for electromagnetic instruments, cosmography using LISA sources as standard sirens, astrophysical inference from the source population, accuracy requirements on waveform models, trade-offs between instrumental design and scientific outputs.
+We hope that our tools will become useful for all these applications.
 
@@
 
-\SM{update: change of plan}
-For completeness, we provide in this Appendix comprehensive definitions and notations for the gravitational wave signal as well as the LISA response.
-We hope to spare the reader the effort of navigating between the various conventions used in the literature, and to provide a point of reference for future works.
-Having these definitions at hand will also help us derive the approximations linking the full response to the simplified response that we used in Sec.~\ref{subsec:SMBHPEdegen}.
+For completeness, we complete in this Appendix the definitions and convention choices that were sketched in Section~\ref{sec:response}.
 
@@
 \subsection{Source frame, wave frame and SSB frame}
-\label{sec:wavessbframe}
-
-We first define a source frame $(\hat{x}_{S}, \hat{y}_{S}, \hat{z}_{S})$ associated to the compact binary system emitting gravitational waves.
-We choose $\hat{z}_{S} = \hat{L}$, the direction of the orbital angular momentum (constant for non-precessing systems).
-The choice of $\hat{x}_{S}$, $\hat{y}_{S}$ amounts to fixing the phase convention.
-
-Introducing $k$ the wave propagation unit vector going from the source towards the observer, we define the inclination $\iota$ and the observer phase $\varphi$ as its spherical angular coordinates in the source frame, so that in that frame
-\be
-	k_{S} = (\sin\iota \cos\varphi, \sin\iota \sin\varphi, \cos\iota) \,.
-\ee
-
-By convention, for our polarization vectors $p$ and $q$ we will use the spherical coordinate vectors $p = e_{\theta}^{S}$, $q = e_{\phi}^{S}$. In terms of only $\hat{z}_{S}$,
-\bsub
-\begin{align}
-	p &= q \times k \,,\\
-	q &= \hat{z}_{S} \times k / |\hat{z}_{S} \times k|\,.
-\end{align}
-\esub
-
-...
-
-We start by introducing a source frame $(\hat{x}_{S}, \hat{y}_{S}, \hat{z}_{S})$ attached to the binary system emitting gravitational waves.
+\label{app:wavessbframe}
+
+We first introduce a source frame $(\hat{x}_{S}, \hat{y}_{S}, \hat{z}_{S})$ attached to the binary system emitting gravitational waves.
 Relating this frame to the physical configuration of the binary is understood as being part of the waveform model.
-For comparable-mass systems without spin or with aligned spins\footnote{For precessing systems, different choices are possible, like $\hat{z}_{S} = \hat{J}$, the direction of the total angular momentum.}, the natural choice is to take the normal to the orbital plane as $\hat{z}_{S}$, and we will assume that the remaining rotation around $\hat{z}_{S}$ is fixed by the phase convention of the waveform model (fixing the waveform phase at a reference frequency, or fixing the orbital phase at a reference time, see Sec.~\ref{} \SM{...}).
+For comparable-mass systems without spin or with aligned spins\footnote{For precessing systems, different choices are possible, like $\hat{z}_{S} = \hat{J}$, the direction of the total angular momentum.}, the natural choice is to take the normal to the orbital plane as $\hat{z}_{S}$, and we will assume that the remaining rotation around $\hat{z}_{S}$ is fixed by the phase convention of the waveform model.
 
@@
 \subsection{LISA trajectories}
-\label{sec:lisatraj}
+\label{app:lisatraj}
 
@@
 We denote the positions of the spacecrafts as $p_{A}$ for $A=1,2,3$, and the positions relative to the constellation center as $p_{A}^{L} = p_{A} - p_{0}$.
-Setting $\beta_{A} = 2(A-1)\pi/3 + \beta_{0}$ (with $\beta_{0}$ and initial condition set to 0 in our case), the Cartesian coordinates in the SSB-frame of the position of the spacecrafts read~\cite{}
+Setting $\beta_{A} = 2(A-1)\pi/3 + \beta_{0}$ (with $\beta_{0}$ and initial condition set to 0 in our case), the Cartesian coordinates in the SSB-frame of the position of the spacecrafts read
 \bsub
@@
 \ee
-The rigid approximation for the constellation, at first order in $e$, can be seen as a first-order approximation in the small parameter $L/R \simeq 0.017$ for a $2.5 \, \mathrm{Gm}$ armlength~\cite{}.
-
-\TDC{Something is incorrect in this discussion. From (A11) and the definition of $a$ above, it sounds like the armlength is longer than 1 au.}
+The rigid approximation for the constellation, at first order in $e$, can be seen as a first-order approximation in the small parameter $L/R \simeq 0.017$ for a $2.5 \, \mathrm{Gm}$ armlength.
 
@@
 \subsection{The LISA frame}
-\label{sec:LISAframe}
+\label{app:LISAframe}
 
@@
 
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-
-\subsection{Parametrizing the signal in the LISA-frame}
-\label{sec:LISAframeparams}
-
-While long-lasting signals like SOBHs will see a strong imprint of the LISA orbital motion over the course of observation, SMBH signals are dominated in SNR by a short-lived burst of emission at merger.
+While long-lasting signals like SBHBs will see a strong imprint of the LISA orbital motion over the course of observation, SMBHB signals are dominated in SNR by a short-lived burst of emission at merger.
 For such signals, it will be useful to use a parametrization based on the LISA frame at the time of merger instead of the SSB frame.
@@
 
-The definition of the reference phase $\phi_{\rm Ref}$, \dots \SM{[COMPLETE]}
-
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-
-\section{Numerical integration with spline-interpolated phases}
-\label{app:numintegration}
-
-[describe implementation of 0-noise integrals]
-
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-
-\section{Reduced-order model for EOBNRv2HMROM}
-\label{sec:rom}
-
-[describe implementation of the ROM]
-
-\begin{figure}
-  \centering
-  \includegraphics[width=.99\linewidth]{../plots/plot_MM_ROM.pdf}
-  \caption{Unfaithfulness of ROM waveforms compared to the original EOBNRv2HM waveforms, computed as in~\eqref{}.
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+
+\subsection{LISA instrumental nosie}
+\label{app:LISAnoise}
+
+The instrumental noise we consider in this study is constructed from~\cite{LISANoiseBudget16}, which specifies the LISA noise budget for different subsystems.
+The numerical expressions we use are (with all quantities in SI units):
+\begin{widetext}
+\bsub
+\begin{align}
+	S^{\rm pm} (f) &= \left(\frac{1}{2\pi f c}\right)^{2} (3\cdot 10^{-15})^{2} \left[1 + 36 \left( \left(\frac{10^{-4}}{f}\right)^{2} + \left(\frac{3\cdot10^{-5}}{f}\right)^{10} \right)\right] + \frac{1}{4}\left(1.7\cdot 10^{-12}\right)^{2} \left(\frac{2\pi f}{c}\right)^{2} + S^{\rm WD}(f) \,,\\
+	S^{\rm op} (f) &= \left(\frac{2\pi f}{c}\right)^{2} \left[\left(8.9\cdot 10^{-12}\right)^{2} + \left(1.7\cdot 10^{-12}\right)^{2} + \left(2\cdot 10^{-12}\right)^{2}\right] \,.
+\end{align}
+\esub
+\end{widetext}
+
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+
+%\section{Numerical integration with spline-interpolated phases}
+%\label{app:numintegration}
+%
+%[describe implementation of 0-noise integrals]
+
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+
+%\section{Reduced-order model for EOBNRv2HMROM}
+%\label{sec:rom}
+%
+%[describe implementation of the ROM]
+%
+%\begin{figure}
+%  \centering
+%  \includegraphics[width=.99\linewidth]{../plots/plot_MM_ROM.pdf}
+%  \caption{Unfaithfulness of ROM waveforms compared to the original EOBNRv2HM waveforms, computed as in~\eqref{}.
 We show here the results for the orientations \SM{[complete]}, for 200 random mass ratio values between 1 and 12, for masses $10 \Msol$, $50 \Msol$ and $250 \Msol$.
 The inclination angle is $\iota = \pi/2$, maximizing the effect of higher harmonics.}
-  \label{fig:MMROM}
-\end{figure}
+%  \label{fig:MMROM}
+%\end{figure}
 
@@
   \centering
-  \includegraphics[width=.8\linewidth]{../plots/corner_smbh_case9_simplelike_ptmcmc_bambi.png}
-  \caption{Posterior distributions obtained with the two samplers \texttt{PTMCMC} (blue) and $\texttt{MultiNest}$ (green) for the simplified likelihood~\eqref{eq:simplelikelihoodhm}, using only the dominant mode $(\ell, m) = (2,2)$.
+  \includegraphics[width=.9\linewidth]{../plots/corner_smbh_case9_simplelike_ptmcmc_bambi.png}
+  \caption{Posterior distributions obtained with the two samplers \texttt{ptmcmc} (blue) and $\texttt{multinest}$ (green) for the simplified likelihood~\eqref{eq:simplelikelihoodhm}, using only the dominant mode $(\ell, m) = (2,2)$.
 Masses and time are pinned to the injected value, and the motion of the detector is neglected.
-We also superimpose (red) the contours obtained with \texttt{MultiNest} by including all harmonics $(2,2)$, $(2,1)$, $(3,3)$, $(4,4)$, $(5,5)$.}
+We also superimpose (red) the contours obtained with \texttt{multinest} by including all harmonics $(2,2)$, $(2,1)$, $(3,3)$, $(4,4)$, $(5,5)$.}
   \label{fig:PEsmbh22hmSimpleLikeCase9}
 \end{figure*}
+
+In this Appendix, we further illustrate our findings of Sec.~\ref{subsec:MBHBPEdegen22} by running our Bayesian parameter estimation codes with the \textit{simplified likelihood}~\eqref{eq:simplelikelihood22} consisting in pinning intrinsic parameters (masses and time) and using the \textit{Frozen, low-$f$} response approximation.
+
+Since the inner products~\eqref{eq:innerproductlmlpmp} are constants computed only once, the external factors become trivially inexpensive, and the likelihood is very fast to compute.
+We take advantage of this to run \texttt{ptmcmc} for a very large number of steps, ensuring that we do not miss degenerate regions of the parameter space.
+The resulting posterior distribution is extremely degenerate, as shown in Fig.~\ref{fig:PEsmbh22hmSimpleLikeCase9}.
+The sampler \texttt{multinest} fails to resolve this structure, stopping after having only explored the degenerate region $\iota \rightarrow 0$ and $\iota \rightarrow \pi$ and the eight sky positions~\eqref{eq:soldegensky}.
+In accordance with~\eqref{eq:soldegensky}, this corresponds to lines in $\varphi \pm \psiL = \mathrm{const}$.
+The sampler \texttt{ptmcmc} finds more degenerate regions, in particular it finds similar structures with the roles of $\iota$ and $\betaL$ exchanged following~\eqref{eq:soldegeniotaphi} (as shown notably by the $(d,\iota)$ and $(d,\betaL)$ panels), as well as intermediate regions connecting all these features.
+
+Fig.~\ref{fig:PEsmbh22hmSimpleLikeCase9} also shows the result obtained when including all available harmonics~\eqref{eq:listmodes} in~\eqref{eq:simplelikelihoodhm}.
+The degenerate structures collapse to leave the eight modes corresponding to $\betaL \rightarrow - \betaL$ and $\lambdaL \rightarrow \lambdaL + k\pi/2$, this time centered around the injected value and not the shifted value $(\lambda_{L}^{\dagger}, \beta_{L}^{\dagger})$.
+We found that this degeneracy-breaking occurs already when including a single subdominant harmonic (for instance $h_{21}$).
+
+These results also give us a hint to explain the results of Sec.~\ref{subsec:SMBHPEfull}, and in particular the difference between Bayesian samplers shown in the center left panels of Figs.~\ref{fig:PEsmbhDincptmcmcbambi} and~\ref{fig:PEsmbhskyptmcmcbambi}.
+Since making the likelihood more degenerate in the simplified likelihood limit~\eqref{eq:simplelikelihood22} magnifies the difference between \texttt{ptmcmc} and \texttt{multinest}, with \texttt{ptmcmc} exploring a much larger volume in parameter space, their disagreement with the full likelihood can likely be attributed to an apparent shortcoming of \texttt{multinest}\footnote{We only used standard values for the metaparameters for this sampler, and we do not exclude that it would be possible to improve results with minor changes.} for resolving extended degenerate regions.
 
@@
 
-TDC was supported by an appointment to the NASA Postdoctoral Program at the
-Goddard Space Flight Center, administered by Universities Space Research
+We would like to thank Stanislav Babak, Michael Katz, Vivien Raymond and Alexandre Toubiana for useful discussions.
+We are indebted to Philip Graff for his help with the \texttt{bambi} sampler.
+TDC was supported by an appointment to the NASA Postdoctoral Program at the Goddard Space Flight Center, administered by Universities Space Research
 Association under contract with NASA.

### 2020-02-11 paper/paper_PE.tex
Add Makefile, update my affiliation & acks

@@
 \affiliation{Gravitational Astrophysics Laboratory, NASA Goddard Space Flight Center, 8800 Greenbelt Rd., Greenbelt, MD 20771, USA}
+\affiliation{Université Paris-Saclay, CNRS/IN2P3, IJCLab, 91405 Orsay, France}
 
@@
 We are indebted to Philip Graff for his help with the \texttt{bambi} sampler.
-TDC was supported by an appointment to the NASA Postdoctoral Program at the Goddard Space Flight Center, administered by Universities Space Research
-Association under contract with NASA.
+TDC was supported by an appointment to the NASA Postdoctoral Program at the Goddard Space Flight Center, administered by Universities Space Research Association under contract with NASA, during part of this work.
 

### 2020-02-11 paper/paper_PE.tex
Some references

@@
 More recently, in the context of the redesign of the LISA mission~\cite{elisa13}, the study~\cite{Klein+15} explored the performance of various LISA instrumental designs using Fisher matrix estimates and inspiral signals, but used a reweighting procedure to represent the role of the merger-ringdown signal.
+We also note a recent work~\cite{Baibhav+20} investigating parameter recovery for ringdown-dominated signals.
 To this date, no full Bayesian parameter estimation studies with IMR signals have been performed.
@@
 
-We note that, by contrast with current ground-based observations that are horizon-limited and that have a strong selection bias towards systems with favorable orientations (face-on or face-off)~\cite{}, the LISA horizon for SMBHB systems extends to essentially the entirety of the observable universe~\cite{LISA2017} \cite{}; it is therefore natural to consider randomized orientations.
+We note that, by contrast with current ground-based observations that are horizon-limited and that have a strong selection bias towards systems with favorable orientations (face-on or face-off), the LISA horizon for SMBHB systems extends to essentially the entirety of the observable universe~\cite{LISA2017}; it is therefore natural to consider randomized orientations.
 
@@
 Table~\ref{tab:MBHBdegen} summarizes the qualitative picture of degeneracy breaking by the features of the response.
-We will see in Sec.~\ref{sec:SMBHPEacctime} that the eight-mode degeneracy pattern indeed appears when doing a pre-merger analysis; on the other hand, in results Figs.~\ref{fig:PEsmbh22hmCase0} and~\ref{fig:PEsmbh22hmCase9} for full IMR signals, out of the eight possible sky modes only the \textit{reflected} mode~\eqref{eq:degenreflection} survives.
+We note that the recent work~\cite{Baibhav+20}, focusing on low-frequency ringdown-dominated signals for which the LISA motion can be neglected, remarked the same 8-modes sky degeneracy that we described in this section.
+We will see in Sec.~\ref{sec:MBHBPEacctime} that the eight-mode degeneracy pattern indeed appears when doing a pre-merger analysis; on the other hand, in results Figs.~\ref{fig:PEsmbh22hmCase0} and~\ref{fig:PEsmbh22hmCase9} for full IMR signals, out of the eight possible sky modes only the \textit{reflected} mode~\eqref{eq:degenreflection} survives.
 We will investigate in detail in Sec.~\ref{sec:MBHBPEacctime} how time-dependence and frequency-dependence in the response break part, but not all, of the degeneracies.
@@
 In this study our signals were limited to non-spinning, circularized systems.
-The aligned components of the spin will play a role by allowing degeneracies between masses and spin magnitudes, but we can expect that they will not affect qualitatively the posterior distributions for extrinsic parameters.
+The aligned components of the spin will play a role by allowing degeneracies between masses and spin magnitudes, but we can expect that they will not affect qualitatively the posterior distributions for extrinsic parameters; this will be investigated in a forthcoming publication~\cite{Katz+20}.
 Precession and eccentricity, however, will change the harmonic structures of the signals and play an important role.

### 2020-02-12 paper/paper_PE.tex
Proposed edits on latest version

@@
 Cosmological applications using LISA observations as standard sirens~\cite{Schutz86, Tamanini+16} also depend on the ability to localize individual sources.
-Understanding parameter estimation of SBHBs will also be important to understand the outcomes and challenges of possible multiband gravitational-wave observations~\cite{Sesana16}.
-In this paper we focus on the latter part of the analysis problem, i.e.~the inference of the black hole parameters from the LISA data, once we have reasons to believe the presence of the signal in the data, leaving aside the identification problem.
+Parameter estimation of SBHBs will also be necessary for multiband gravitational-wave astronomy~\cite{Sesana16}.
+In this paper we focus on the latter part of the analysis problem, i.e.~the inference of the black hole parameters from the LISA data, leaving aside the detection problem and assuming the presence of the signal in the data.
 We also limit our analysis to extrinsic parameters and ignore the effect of superposed gravitational wave signals from various sources.
 
-The inference problem amounts to produce samples from the posterior distribution for source parameters, related by Baye's theorem to the likelihood function, i.e.~the probability of observing the measured data given the source parameters and a model of the source and detector.
-In LISA, this problem is more complex than in kilometer-scale ground-based interferometers, as additional challenges arises from the much larger expected signal-to-noise ratios (SNRs) for high-mass systems, or the much longer duration of the waveform for low-mass systems.
+The inference problem amounts to drawing samples from the posterior distribution for source parameters, related by Bayes' theorem to the likelihood function, i.e.~the probability of observing the measured data given the source parameters and a model of the source and detector.
+In LISA, this problem is more complex than in kilometer-scale ground-based interferometers, as additional challenges arise from the much larger expected signal-to-noise ratios (SNRs) for high-mass systems, or the much longer duration of the waveform for low-mass systems.
 In particular, for SMBHBs, the large SNR will require a very accurate modeling of the waveform, including the merger and ringdown regimes, the effects of spin and corrections of higher harmonics in the signal beyond the quadrupole.
@@
 Previous parameter recovery studies with LISA used mainly simplified signal and response models, and often relied on a Fisher matrix approximation to parameter recovery instead of full Bayesian simulations.
-Numerous works used the combination of inspiral-only post-Newtonian waveforms (sometimes including precession), the low-frequency approximate response and Fisher matrix estimates; examples include~\cite{Cutler97, Vecchio03, Arun06, Berti+04, LangHughes06}.
+Numerous works used the combination of inspiral-only post-Newtonian waveforms (sometimes including precession), the low-frequency approximate response and Fisher matrix estimates~\cite{Cutler97, Vecchio03, Arun06, Berti+04, LangHughes06}.
 The Mock LISA Data Challenges~\cite{MLDC09} used such a setup for the signals and the response, albeit moving towards MCMC tools with a focus on detection.
-Bayesian methods going beyond Fisher albeit still with inspiral signals were developed in~\cite{Brown+07, CornishPorter06a, Crowder+06, Wickham+06, Roever+07, Feroz+09, GairPorter09, Petiteau+09, PorterCarre13, PorterCornish15}.
+Bayesian methods going beyond Fisher, albeit still restricted to inspiral signals, were developed in~\cite{Brown+07, CornishPorter06a, Crowder+06, Wickham+06, Roever+07, Feroz+09, GairPorter09, Petiteau+09, PorterCarre13, PorterCornish15}.
 The importance of the higher harmonics for LISA was stressed already in several studies~\cite{Arun+07a, TriasSintes07, PorterCornish08, McWilliams+09}.
@@
 
-Here we demonstrate that standard Bayesian inference can be performed for SMBHBs using a self-consistent waveform model that includes inspiral, merger, ringdown and higher-order modes (albeit still simplified as spins are neglected) and a full model of the LISA response.
+Here we demonstrate that standard Bayesian inference can be performed for SMBHBs using a self-consistent waveform model that includes inspiral, merger, ringdown and higher-order modes (albeit no spins) and a full model of the LISA response.
 We investigate and explain degeneracies in the posterior distribution of the source's parameters, and highlight the crucial role played by the higher harmonics in the signal and the frequency-dependency in the instrument response.
@@
 We start by introducing a fast method to calculate the likelihood for a black hole merger.
-This method includes: a fast computation of a Fourier-domain inspiral-merger-ringdown waveform with higher-order modes based on reduced-order modeling; a fast Fourier-domain computation of the LISA response based on~\cite{MB18}; and a fast method for computing the noise-weighted product between waveforms when setting the noise realization to zero.
-We couple our likelihood with two codes for Bayesian inference based on different sampling techniques, constructing an inference engine which will allow us in the future to perform a variety of investigations with simulated black hole mergers.
+This method includes: a fast computation of a Fourier-domain inspiral-merger-ringdown waveform with higher-order modes based on ROM; a fast Fourier-domain computation of the LISA response based on~\cite{MB18}; and a fast method for computing the noise-weighted product between waveforms when setting the noise realization to zero.
+We couple our likelihood with two codes for Bayesian inference based on different sampling techniques, constructing an inference engine which enables us to perform a variety of investigations with simulated black hole mergers.
 
@@
 
-Finally, we repeat some of our investigations for stellar-origin black hole mergers as well.
+Finally, we repeat some of our investigations for stellar-origin black hole mergers.
 Parameter estimation studies for SBHB systems have relied so far on the Fisher matrix approach~\cite{Sesana16, Vitale16, Nishizawa+16a, Nishizawa+16b}, with Bayesian inference tools yet to be developed.
@@
 In this section, we introduce a complete set of notations for the LISA instrument response that will be useful in the later discussion of degeneracies in parameter space.
-We will use units with $G=c=1$.
+We will use units with $G=c=1$ and assume the proposed armlength for LISA $L = 2.5\mathrm{Gm}$~\cite{LISA2017}.
 
@@
 \end{align}
-Writing the $P_{\ell m}$ matrices in this way allows to factor out explicitly all dependencies in the extrinsic parameters $(\iota, \phi, \psi)$.
+Writing the $P_{\ell m}$ matrices in this way allows us to factor out explicitly all dependencies in the extrinsic parameters $(\iota, \phi, \psi)$.
 Together with the luminosity distance $D$ scaling the overall amplitude of the signal, these parameters enter as constant (time and frequency independent) prefactors in the response for each mode $h_{\ell m}$.
 
-Since only $2\psi$ intervenes, the polarization $\psi$ has an exact $\pi$-degeneracy; by convention we set $\psi \in [0,\pi]$.
+Since $\psi$ always appears with a factor of 2, it has an exact $\pi$-degeneracy, and we choose the convention of restricting $\psi \in [0,\pi]$.
 
@@
 
-The LISA response~\cite{EW75, Cutler97, Larson+99, CR02, RCP04} can be built from single-link observables $y_{slr} = (\nu_{r} - \nu_{s})/\nu$ representing a laser frequency shift between the sending spacecraft $s$ and the receiving spacecraft $r$ along the link $l$.
+The LISA response~\cite{EW75, Cutler97, Larson+99, CR02, RCP04} can be built from single-link observables $y_{slr} = (\nu_{r} - \nu_{s})/\nu$ representing a laser frequency shift between the transmitting spacecraft $s$ and the receiving spacecraft $r$ along the link $l$.
 We use the expression~\cite{Vallisneri04, Krolak+04}
@@
 \ee
-with $H$ the transverse-traceless matrix representing the gravitational wave, $L$ the delay along one arm\footnote{We recall that we use units with $G=c=1$.
-The proposed armlength for LISA is $L = 2.5\mathrm{Gm}$~\cite{LISA2017}.}, taken to be fixed, $k$ the propagation vector, $n_{l}$ the link unit vectors (from $s$ to $r$) and $p_{s}$, $p_{r}$ the positions of the spacecrafts.
+with $H$ the transverse-traceless matrix representing the gravitational wave, $L$ the delay along one arm, taken to be fixed, $n_{l}$ the link unit vectors (from $s$ to $r$) and $p_{s}$, $p_{r}$ the positions of the spacecrafts.
 Here $n_{l}$, $p_{s}$ and $p_{r}$ are evaluated at the same time $t$.
 
-In the frequency domain, modes will be represented with their amplitude and phase as\footnote{Note our Fourier convention~\eqref{eq:defFourier}.}
+In the frequency domain, modes will be represented with their amplitude and phase as
 \be\label{eq:hlmampphase}
@@
 
-In~\eqref{eq:defyslr} and~\eqref{eq:Gslr}, one can distinguish two types of delays: on one hand the delay associated to the position $p_{0}$ of the center of the constellation on its orbit around the Sun, with baseline $R=1 \au$, and on the other hand the delays associated to the individual spacecraft positions in the constellation, with baseline the armlength $L=2.5 \cdot 10^{6 }\mathrm{km}$.
+In~\eqref{eq:defyslr} and~\eqref{eq:Gslr}, one can distinguish two types of delays: on one hand the delay associated to the position $p_{0}$ of the center of the constellation on its orbit around the Sun, with baseline $R=1 \au$, and on the other hand the delays associated to the individual spacecraft positions in the constellation, with baseline the armlength $L$.
 This defines the transfer frequencies

### 2020-02-17 paper/paper_PE.tex
More proposed edits

@@
 
-To mitigate laser noise, LISA requires Time Delay Interferometry (TDI)~\cite{TintoArmstrong99, Armstrong+99, Estabrook+00, Dhurandhar+02, Tintoliving}.
-These TDI observables are built from delayed combinations of the one-arm basic observables~\eqref{eq:defyslr}.
-Detailed cancellations for a flexing and rotating constellation prompted the development of higher-generation TDI schemes~\cite{Shaddock03, CornishHellings03, Shaddock+03, Tinto+04}.
+The basic one-arm observables of Eq.~\eqref{eq:defyslr} are affected by laser noise whose amplitude is orders of magnitude larger than the astrophysical signal.
+However, time delay interferometry (TDI) allows one to construct a new set of observables from delayed combinations of $y_{slr}$, where laser noise is suppressed by orders of magnitude~\cite{TintoArmstrong99, Armstrong+99, Estabrook+00, Dhurandhar+02, Tintoliving}.
+Various generations of TDI schemes have been proposed in order to deal with a non-rigid and rotating LISA constellation~\cite{Shaddock03, CornishHellings03, Shaddock+03, Tinto+04}.
 However, these refinements affect only marginally the response to the gravitational waves.
-We therefore limit ourselves to first-generation TDI and adopt a rigid approximation for the cancellation, where delays are all constant and set to $L$.
+Hence, in this work we only consider first-generation TDI and adopt a rigid approximation for the constellation, where delays are all constant and equal to $L$.
 Using the notation $y_{slr,nL} = y_{slr}(t - nL)$, the first-generation TDI Michelson observable $X$ reads~\cite{Vallisneri04}
@@
 \end{align}
-with the other Michelson observables $Y$, $Z$ being obtained by cyclic permutation. Uncorrelated combinations $A$, $E$ and $T$~\cite{Prince+02} are then defined as
+with the other Michelson observables $Y$, $Z$ being obtained by cyclic permutation.
+The uncorrelated-noise combinations $A$, $E$ and $T$~\cite{Prince+02} are then defined as
 \bsub\label{eq:defAET}
@@
 \esub
-These channels are independent under the assumption of an identical and uncorrrelated noise in the detector arms.
-Note that various conventions coexist in the literature.
-With constant delays in the rigid approximation, with the notation $z\equiv \exp[2i\pi fL]$ the TDI combinations take the form
+These channels are independent under the assumption of an identical and uncorrrelated noise in the detector arms. Note that various conventions coexist in the literature.
+With constant delays in the rigid approximation, and using the notation $z\equiv \exp[2i\pi fL]$, the TDI combinations in the frequency domain take the form
 \bsub\label{eq:aet}
 \begin{align}
-	\tilde{a} &= \left[ (1+z) \left( \tilde{y}_{31} + \tilde{y}_{13} \right) - \tilde{y}_{23} - z \tilde{y}_{32} - \tilde{y}_{21} - z \tilde{y}_{12} \right] \,,\\
+	\tilde{a} &= (1+z) \left( \tilde{y}_{31} + \tilde{y}_{13} \right) - \tilde{y}_{23} - z \tilde{y}_{32} - \tilde{y}_{21} - z \tilde{y}_{12} \,,\\
 	\tilde{e} &= \frac{1}{\sqrt{3}} \left[ (1-z)\left( \tilde{y}_{13} - \tilde{y}_{31} \right) + (2+z) \left( \tilde{y}_{12} - \tilde{y}_{32} \right) \right. \nn\\
@@
 
-As is well known~\cite{Cutler97}, in the low-frequency limit (also called the long-wavelength approximation), the finite-armlength effects vanish, and the response of LISA is analogous to the response of of two LIGO-type detectors  rotated from each other by $\pi/4$ and set in motion.
+In the low-frequency limit, also called the long-wavelength approximation, the effects due the finite LISA armlength become negligible and the LISA response becomes equivalent to the response of two coplanar LIGO-like detectors rotated by $\pi/4$ relative to each other and set in motion~\cite{Cutler97}.
 
@@
 \esub
-Here, we have made apparent factors $(-2i\pi f)$ corresponding to the Fourier-domain translation of a time derivative.
-The observables $\tilde{a}$, $\tilde{e}$ are analogous to a time derivative of the strain more commonly used for ground-based detectors.
+Here, we have made apparent factors $(-2i\pi f)$, which correspond to Fourier-domain time derivatives.
+The observables $\tilde{a}$, $\tilde{e}$ are therefore analogous to time derivatives of the strain commonly used for ground-based detectors.
 
 We can now map these two channels to two fictitious LIGO-type detectors as follows.
-For an orthogonal detector of the same setup as the ground-based LIGO and Virgo with armlength $L_{D}$, rotated by an angle $\epsilon_{D}$ from the basis vectors $x,y$, the detector tensor is\S
+For an orthogonal detector of the same setup as the ground-based LIGO and Virgo with armlength $L_{D}$, rotated by an angle $\epsilon_{D}$ from the basis vectors $x,y$, the detector tensor is
 \begin{align}
@@
 \ee
-for stationary Gaussian noise with noise PSD $S_{n}$, the likelihood $\calL = p(d | \bm{\theta})$ for a given gravitational wave channel takes the form
+for stationary Gaussian noise with PSD $S_{n}$, the likelihood $\calL = p(d | \bm{\theta})$ for a given gravitational wave channel takes the form
 \be\label{eq:deflnL}
@@
 with the data stream being a superposition of the gravitational wave signal for the true parameters $\bm{\theta}_{0}$ and the noise realization in the experiment, $d = h(\bm{\theta}_{0}) + n$.
-In general, the templates $h(\theta)$ used for the analysis will only be an approximation of the physical signals entering the data stream $d$; we will ignore entirely this distinction and neglect the systematic errors in the waveform model.
-The posterior distribution for the physical parameters $\bm{\theta}$ given the observed data $d$ is then
-\be\label{eq:defBayes}
+The posterior distribution for the physical parameters $\bm{\theta}$ given the observed data $d$ is then \be\label{eq:defBayes}
 	p\left( \bm{\theta} | d \right) = \frac{p(d | \bm{\theta})p(\bm{\theta})}{p(d)}\,,
 \ee
-with $p(\bm{\theta})$ the priors on the parameters, and $p(d)$ the evidence that we will ignore here as it enters as a multiplicative constant in the posterior.
-
-In our analysis, we will use the so-called zero-noise approximation.
-This corresponds simply to performing a Bayesian analysis with~\eqref{eq:deflnL} and ~\eqref{eq:defBayes}, but setting the noise realization to zero, $n=0$.
-It allows us to accelerate the likelihood computation, as will be explained in~\ref{subsec:likelihood}.
-For exploratory parameter estimation runs, this approximation allows us to explore the structure of the likelihood with its possible parameter degeneracies, and is a significant improvement over Fisher matrix estimates in that sense.
-We leave the extension of our analysis to include noise realizations for future work.
-We will also ignore waveform systematic errors and assume our waveform templates match exactly the real signals.
+with $p(\bm{\theta})$ the priors on the parameters, and $p(d)$ the evidence.
+The focus of this work is not selecting between different models, so we will treat $p(d)$ as a normalization constant and do not consider it further.
+
+In general, the templates $h(\bm{\theta})$ used for the analysis will only be an approximation of the physical signals present in the data stream $d$.
+In this work we will ignore entirely this distinction and assume that our template waveforms exactly match the astrophysical waveforms.
+In other words, we will not explore the issue of systematic errors arising from incomplete modeling of compact binary merger waveforms.
+
+In our analysis, we will simulate signals and perform Bayesian analyses of them with Eq.~\eqref{eq:deflnL} and ~\eqref{eq:defBayes} to obtain the parameter posteriors.
+In doing so, we will use the so-called zero-noise approximation, i.e.~setting $n=0$.
+This allows us to accelerate the likelihood computation as explained in Sec.~\ref{subsec:likelihood}.
+The approximation is sufficient to explore the structure of the likelihood and the parameter degeneracies, greatly improving with respect to the Fisher matrix approach.
+The extension of our analysis to include nonzero noise realizations will be discussed in future work.
 
@@
 \end{align}
-with the $\rm inj$ subscript indicating the injected signals, and where the inner products $(\cdot | \cdot)$ are given by~\eqref{eq:definnerproduct} with the noise PSDs~\eqref{eq:Snaet}.
+with the ``$\rm inj$'' subscript indicating the simulated signals, and where the inner products $(\cdot | \cdot)$ are given by~\eqref{eq:definnerproduct} with the noise PSDs~\eqref{eq:Snaet}.
 
@@
 Since higher harmonics will play a crucial role in our analysis, in this study we use the Effective-One-Body (EOB) waveform model \texttt{EOBNRv2HM}~\cite{Pan+11}.
-These waveforms are based on the EOB formalism~\cite{BD99, BD00}, and are calibrated on a set of numerical relativity waveforms.
+These waveforms are based on the EOB formalism~\cite{BD99, BD00}, and are calibrated to numerical relativity waveforms.
 The model is limited to non-spinning systems on quasicircular orbits, but includes a set of higher harmonics in the signal, among the most important quantitatively:
@@
 \ee
-The waveforms are generated in the time domain by integrating a system of ODEs, an operation requiring up to seconds of computing time for long signals.
-For this reason, we developed a reduced order model (ROM) for these waveforms, \texttt{EOBNRv2HMROM}, following the methods of~\cite{Puerrer14}, enabling a much faster generation of the Fourier-domain amplitudes and phases of the modes~\eqref{eq:hlmampphase}.
-As the total mass $M$ scales out of the problem, the parameter space is effectively only one-dimensional, with only the mass ratio $q=m_{1}/m_{2}$.
-The parameter space interpolation in this ROM is therefore significantly simpler than the models~\cite{Puerrer14, Puerrer15, Bohe+16}, which was three-dimensional as it included aligned spins.
-The inclusion of higher harmonics, however, imposes to model the relative dephasing of different modes.
+The waveforms are generated in the time domain by integrating a system of ordinary differential equations, an operation requiring up to seconds of computing time for long signals.
+For this reason, we developed a reduced order model (ROM) for these waveforms, \texttt{EOBNRv2HMROM}, following the methods of~\cite{Puerrer14}, enabling a much faster generation of the Fourier-domain amplitudes and phases of the modes in Eq.~\eqref{eq:hlmampphase}.
+As the total mass $M$ scales out of the problem, the parameter space reduces to the mass ratio $q=m_{1}/m_{2}$ only.
+The parameter space interpolation in this ROM is therefore significantly simpler than the models~\cite{Puerrer14, Puerrer15, Bohe+16}, which include aligned spins.
+The inclusion of higher harmonics, however, requires modeling the relative dephasing of different modes.
 The output of the code consists of a Fourier domain amplitude and phase for each mode, sparsely sampled on $\sim 300$ frequencies.
-The faithfulness with the original waveforms was measured with \texttt{aLIGO} noise curves~\cite{UnofficialNoiseCurves18} to be $\lesssim 10^{-4}$, while the computational cost is submillisecond and subdominant with respect to other stages of our likelihood computation.
+The faithfulness with the original waveforms, assuming Advanced LIGO noise curves~\cite{UnofficialNoiseCurves18}, is $\lesssim 10^{-4}$.
+The computational cost is submillisecond and smaller than other stages of our likelihood computation.
 
@@
 Nevertheless, consistently including merger, ringdown and higher harmonics already represents a significant improvement over previous studies, as these features in the signal carry a lot of SNR by themselves~\cite{Babak+08, Thorpe+08, McWilliams+09, McWilliams+10, McWilliams+11, Klein+15}.
-In the absence of precession at least, we expect intrinsic and extrinsic parameters to correlate weakly, so that  our investigations of the extrinsic part of the likelihood should be valid for aligned spins.
+In the absence of precession, we expect intrinsic and extrinsic parameters to be weakly correlated, so that  our investigations of the extrinsic part of the likelihood should be valid for aligned spins.
 
@@
 In general, both the template and the injection in~\eqref{eq:deflnL} are given as frequency series sampled with the Nyquist criterion $\Delta f = 1/(2T)$ with $T$ the duration of the signal, and the overlaps~\eqref{eq:definnerproduct} are simply computed as discrete sums over those frequency samples.
-Although transforming~\eqref{eq:definnerproduct} in a discrete sum is a straightforward operation, the size of the frequency series alone can represent a significant cost.
+Although transforming~\eqref{eq:definnerproduct} in a discrete sum is a straightforward operation, the size of the frequency series can lead to a significant cost.
 In the context of LIGO/Virgo observations, this has prompted the development of Reduced Order Quadratures (ROQ)~\cite{Smith+14, Canizares+14, Smith+16} to accelerate the likelihood computation.
@@
 This cost can range from tens of milliseconds for SMBHBs and short-lived signals (a few days), to impracticable values for SBHBs signals lasting for years and sampled at high frequency.
-Setting the noise realization to zero, however, we can exploit an amplitude/phase representation of the signals, use a sparse frequency grid and get a large speedup.
+By setting the noise realization to zero, however, we can represent the amplitude and phase of the signals over a sparse frequency grid and obtain a large speedup.
 
@@
 
-The difficulty of computing numerically such an overlap depends on the phase difference.
-A large phase difference causes the integrand to be very oscillatory, requiring more frequency resolution than non-oscillatory integrands; on the other hand, oscillatory integrands give raise to a small integral due to cancellation effects, while non-oscillatory integrands contribute the most.
+The difficulty of numerically computing such an overlap depends on the phase difference.
+A large phase difference causes the integrand to be very oscillatory, requiring more frequency resolution than non-oscillatory integrands.
+On the other hand, oscillatory integrands result in a small integral due to cancellation effects, while non-oscillatory integrands contribute the most.
 For a single mode, $\mathrm{SNR}$ terms of the type $(h|h)$ have a zero phase difference by construction, but including different modes generates cross-terms $(\ell, m) \neq (\ell', m')$ with a large phase difference.
@@
 Since the number of overlaps to compute in~\eqref{eq:defTlmaet} is quadratic in the number of modes, likelihoods with higher harmonics are significantly more expensive.
-The final likelihood cost depends also on the number of grid intervals, that is chosen to limit spline interpolation errors:
-\begin{itemize}
-	\item for SBHBs with $h_{22}$ only: $\sim 500$ grid points, likelihood cost $\sim 2\mathrm{ms}$;
-	\item for SMBHBs with $h_{22}$ only: $\sim 300$ grid points, likelihood cost $\sim 1.2\mathrm{ms}$;
-	\item for SMBHBs with 5 modes $h_{\ell m}$: $\sim 300$ grid points, likelihood cost $\sim 15\mathrm{ms}$.
-\end{itemize}
+The final likelihood cost depends also on the number of grid intervals, that is chosen to limit spline interpolation errors (Table~\ref{tab:likecost}).
+\begin{table}
+	\begin{tabular}{|c|c|c|c|}
+		\hline
+		Source type & Modes & Grid points & Cost \\
+		\hline
+		SBHB & $h_{22}$ & $500$ & $2$ ms \\
+		SMBHB & $h_{22}$ & $300$ & $1.2$ ms \\
+		SMBHB & $5$ & $300$ & $15$ ms \\
+		\hline
+	\end{tabular}
+	\label{tab:likecost}
+	\caption{Approximate frequency grid resolution and likelihood cost for different choices of source type and number of modes in the waveform.}
+\end{table}
+%\begin{itemize}
+%	\item for SBHBs with $h_{22}$ only: $\sim 500$ grid points, likelihood cost $\sim 2\mathrm{ms}$;
+%	\item for SMBHBs with $h_{22}$ only: $\sim 300$ grid points, likelihood cost $\sim 1.2\mathrm{ms}$;
+%	\item for SMBHBs with 5 modes: $\sim 300$ grid points, likelihood cost $\sim 15\mathrm{ms}$.
+%\end{itemize}
 The accuracy of our likelihood computations is typically $|\Delta \ln \calL| \lesssim 0.2$, due to the numerical errors being magnified by the required cancellations between overlaps terms in~\eqref{eq:lnLoverlapstructure}.

### 2020-02-18 paper/paper_PE.tex
Continue proposing edits

@@
 
-For comparison with our Bayesian inference results we also compute estimates of parameter uncertainties using the Fisher information matrix, which despite its limitations has been the common workhorse of LISA science studies to date ~\cite{Vallisneri08}.
+For comparison with our Bayesian inference results we also compute estimates of parameter uncertainties using the Fisher information matrix, which despite its limitations has been the common workhorse of LISA science studies to date \cite{Vallisneri08}.
 For measurements with additive Gaussian noise, as assumed here, the Fisher information matrix can be computed by
 \be
-	F_{ij} = \left( \partial_{i} h | \partial_{i} h \right),
-\ee
-where derivatives are taken with respect to one of the parameter space coordinates $\bm{\theta}$. In the Fisher matrix approach, the likelihood is approximated as
+	F_{ij} = \left( \partial_{i} h | \partial_{j} h \right),
+\ee
+where $\partial_i$ is the derivative with respect to component $i$ of the parameter vector $\bm{\theta}$. In the Fisher matrix approach, the likelihood is approximated as
 \be
@@
 We target the step-size to be scaled off the diagonal Fisher matrix elements by ${\epsilon_i}\approx\delta/\sqrt{F_{ii}}$ using $\delta=0.001$ for the results shown here.
-To achieve this we set begin with an initial choice of the $\epsilon_i$ scaled of parameter values or prior widths, then we estimate the Fisher diagonals and iterate until convergence.
-In this process we also constrain that the step is never more than $10^{-10}$ times the initial scaling.
+To achieve this, we begin with an initial choice of $\epsilon_i$ scaled off the parameter values or prior widths, then we estimate the Fisher diagonals and iterate until convergence.
+In this process we also impose the constrain that the step is never larger than $10^{-10}$ times the initial scaling.
 
@@
 
-We focus our analysis on a single set of mass parameters, representative of canonical SMBHB sources expected for LISA~\cite{Barausse12, Klein+15, LISA2017}: we pick a total redshifted mass $M=m_{1}+m_{2} = 2\times10^{6} M_{\odot}$ and a mass ratio $q=m_{1}/m_{2} = 3$, and place the source at a redshift $z=4$.
+We focus our analysis on a single choice of mass parameters, representative of canonical SMBHB sources expected for LISA~\cite{Barausse12, Klein+15, LISA2017}: we pick a total redshifted mass $M=m_{1}+m_{2} = 2\times10^{6} M_{\odot}$ and a mass ratio $q=m_{1}/m_{2} = 3$, and place the source at a redshift $z=4$.
 The corresponding source-frame total mass is $M/(1+z) = 4\times10^{5} M_{\odot}$.
@@
 with the strain-like observables and noise PSD defined in~\eqref{eq:defhaet} and~\eqref{eq:defShaet}.
-We represent the $A$ TDI channel (the $E$ channel is similar, and the $T$ channel is negligible at low frequencies).
-The full signal shows oscillatory features in frequency that are due to beatings between the harmonics that have smooth amplitudes individually.
-Since we represent mode contributions to the TDI observables, the response is included; we see that all modes display structure at low frequencies that corresponds to the time modulation resulting of the LISA motion.
-
-Fig.~\ref{fig:tranferSMBHCase9} shows the detail of the transfer functions themselves.
+We only show the $A$ TDI channel, as the $E$ channel is qualitatively similar and the $T$ channel is negligible at low frequencies.
+Although the individual harmonics have fairly smooth amplitudes as a function of frequency, the full signal shows strong oscillations caused by the beating between the harmonics.
+The effect of the LISA response can be seen in the amplitude oscillations at low frequency, which are caused by the modulation resulting from the LISA motion.
+
+Fig.~\ref{fig:tranferSMBHCase9} focuses on the details of the transfer functions.
 The Doppler phase variation $\Delta \Phi_{R}$~\eqref{eq:defPhiR} is a small effect at low frequencies, being suppressed by a factor $\sim 2\pi f R$; it is also small at higher frequencies, because the chirp happens over a short time interval during which $\Phi_{R}$ does not have time to vary.
@@
 The low-frequency features show the time-dependency of the response created by the LISA motion, while frequency-dependency in the response appears at high frequencies.
-We will investigate further these different physical effects in the response in Sec.~\ref{subsec:MBHBresponseapprox}
+Sec.~\ref{subsec:MBHBresponseapprox} will further investigate these different physical effects in the response.
 
@@
    \includegraphics[width=.99\linewidth]{../plots/tflm_smbh.pdf}
-  \caption{Time-to-frequency relation for MBHB systems I and II with $M=2\times 10^6 \; M_{\odot}$, $q=3$ and $z=4$.
+  \caption{Relation between inspiral time and frequency for MBHB systems I and II with $M=2\times 10^6 \; M_{\odot}$, $q=3$ and $z=4$.
 The different lines correspond to the $(\ell, m)$ harmonics available in the waveform model.
-The time-to-frequency is computed according to the defintion in $t_{f}^{\ell m}$ defined in~\eqref{eq:deftflm}, evaluated at $m/2 \times f$ to map to the same time according to~\eqref{eq:fscalinglm}.
+The time-to-frequency correspondence is computed according to $t_{f}^{\ell m}$ defined in~\eqref{eq:deftflm}, evaluated at $m f / 2$ to map to the same time according to~\eqref{eq:fscalinglm}.
 The vertical lines represent the time cuts explained in~\ref{subsec:MBHBacctime} and the merger frequency.}
@@
 In all panels the black line represents the total.
-The horizontal axis gives the time to merger, with vertical lines for the cuts of~\ref{subsec:MBHBacctime}, and the dots on the right of each panel show the result for the full signal including the post-merger.}
+The horizontal axis gives the time to merger, with vertical lines for the cuts of~\ref{subsec:MBHBacctime}, and the dots on the right of each panel show the result for the full signal including the ringdown.}
   \label{fig:modeoverlapsoftime}
@@
 
-We will focus on the system II of Table~\ref{tab:MBHBparams}, and we will highlight four different epochs, corresponding to the time before merger where a certain fraction of the total $\SNR = 666$ has been accumulated.
+We will focus on system II of Table~\ref{tab:MBHBparams}, and we will highlight four different epochs, corresponding to the time before merger where a certain fraction of the total $\SNR = 666$ has been accumulated.
 We include higher harmonics here.
@@
 \begin{itemize}
-	\item $\SNR/64$: $\sim 42\mathrm{h}$ before merger, which corresponds roughly, with $\SNR \simeq 10$, to the first time we could confidently claim a detection;
-	\item $\SNR/16$: $\sim 2.5\mathrm{h}$ before merger;
-	\item $\SNR/4$: $\sim 7\mathrm{min}$ before merger.
+	\item $\SNR/64$: $\sim 42$ hr before merger, which corresponds roughly, with $\SNR \simeq 10$, to the first time we could confidently claim a detection;
+	\item $\SNR/16$: $\sim 2.5$ hr before merger;
+	\item $\SNR/4$: $\sim 7$ min before merger.
 \end{itemize}
 
-To represent the observations stopping with a certain time to merger left, we will approximate the cut in time using simply the time-to-frequency correspondence~\eqref{eq:deftflm} to translate it as a cut in frequency.
-In a more realistic setting, an abrupt interruption of the signal would translate into non-local features in the Fourier domain unless tapering is applied, but we ignore these complications here.
-
-The time-to-frequency correspondence~\eqref{eq:deftflm} gives for each mode time as a function of frequency.
+We approximate the abrupt interruption of the observation as an upper frequency cutoff.
+The cutoff frequency is derived from the end time of the observation via the time-to-frequency correspondence~\eqref{eq:deftflm}.
+In reality, an abrupt interruption of the signal would produce non-local features in the Fourier domain (unless tapering is applied) which we do not consider here.
+
+The time-to-frequency correspondence~\eqref{eq:deftflm} gives time as a function of frequency for each $(\ell, m)$ mode.
 Due to the scaling $h_{\ell m} \propto e^{-i m \phi_{\rm orb}}$ with $\phi_{\rm orb}$ the orbital phase, a given time corresponds to different frequencies according to
@@
 
-The relations~\eqref{eq:fscalinglm} are accurate for the inspiral, as such a correspondence is at the heart of the SPA.
-As a function, $t_{f}^{\ell m}$~\eqref{eq:deftflm} extends formally through merger, while losing in accuracy and in physical interpretation, becoming non-monotonous with frequency (see the discussion in~\cite{MB18}).
-The departure from the scaling~\eqref{eq:fscalinglm} between modes is illustrated in Fig.~\ref{fig:t-f-relation-lm}, where we see that it holds up to roughly $\sim$7min before merger, with the mode $h_{21}$ showing the earliest departure.
+The relations~\eqref{eq:fscalinglm} are only accurate for the inspiral regime, as such a correspondence is at the heart of the SPA.
+Although $t_{f}^{\ell m}$~\eqref{eq:deftflm} can be formally extended even past the merger time, close to merger it starts to lose accuracy and physical interpretation, eventually becoming non-monotonic with frequency~\cite{MB18}.
+The departure from the scaling~\eqref{eq:fscalinglm} between modes can be seen in Fig.~\ref{fig:t-f-relation-lm}: the scaling holds up to a few minutes before merger, and the mode $h_{21}$ shows the earliest signs of a deviation.
 
@@
 
-The LISA instrument response, as recalled in Section~\ref{sec:response}, features both a time- and a frequency-dependency.
+The LISA instrument response, as recalled in Section~\ref{sec:response}, is both time- and frequency-dependent.
 As we will see, these features play an important role in breaking degeneracies in the parameter estimation of the source, notably allowing us to localize the source in the sky.
@@
 Case (iii), \textit{Low-$f$}, has been extensively used in the past for the analysis of MBHB signals.
-Being based on the $f \ll f_{L}$ approximation, it is less inaccurate for inspiral-only signals, to which most previous studies were limited, as it is for a full IMR signal.
-
-In the case (iv), \textit{Frozen low-$f$}, the response is equivalent to two LIGO-type detectors, lying motionless in the same location and rotated from each other by $\pi/4$, and there is no other information on the source's sky position other than the pattern functions of the two effective detectors.
+Since it is based on the $f \ll f_{L}$ approximation, it is more appropriate for inspiral-only signals (to which most previous studies were limited) than for a full IMR signal.
+
+In the case (iv), \textit{Frozen low-$f$}, the response is equivalent to two coplanar LIGO-type detectors, lying motionless at the same location and rotated with respect to each other by $\pi/4$, and there is no other information on the source's sky position other than the pattern functions of the two effective detectors.
 Contrarily to networks of ground-based observatories, we have no triangulation information from times of arrival at different detectors.
-This case is the most degenerate, and will be useful to get an analytical understanding of approximate degeneracies occuring when using the more complicated full response (i).
-
-The effect on the response of adopting these approximations is shown in Fig.~\ref{fig:responseapproxtdia22}.
+This is the most degenerate case, and will be useful to get an analytical understanding of the approximate degeneracies occuring when using the more complicated full response (i).
+
+The effect of these approximations on the response is shown in Fig.~\ref{fig:responseapproxtdia22}.
 We display the strain transfer function $\calT_{h_{a}}^{22}(f)$ as defined in~\eqref{eq:defhaet} for the mode $h_{22}$, with the four lines showing the four approximations (i)-(iv).
@@
   \includegraphics[width=.98\linewidth]{../plots/corner_smbh_case9_ptmcmc_22_hm_withzoom.png}
-  \caption{Same as Fig.~\ref{fig:PEsmbh22hmCase0}, for the MBHB system II of Table~\ref{tab:MBHBparams}. In extrinsic parameters, the red 2D contours are barely visible on this scale.}
+  \caption{Same as Fig.~\ref{fig:PEsmbh22hmCase0}, for the MBHB system II of Table~\ref{tab:MBHBparams}.
+           In the extrinsic parameters, the red 2D contours are barely visible on this scale.}
   \label{fig:PEsmbh22hmCase9}
@@
 
-We performed Bayesian parameter estimation for the two MBHB sources listed in Table.~\ref{tab:MBHBparams}, using the methodology summarized in Sec.~\ref{sec:method}.
-In particular, we set the noise realization to zero, used accelerated likelihood based on an amplitude and phase representation of the signals.
+Using the methodology summarized in Sec.~\ref{sec:method}, here we present the results of Bayesian parameter estimation for the two MBHB sources listed in Table~\ref{tab:MBHBparams}.
 The priors used are logarithmic in mass, flat in luminosity distance, and flat for the angles: flat on the sphere for the pairs of angles $(\iota, \varphi)$ and $(\lambda, \beta)$, and flat for the polarization $\psi$.
@@
 
-We find that both our samplers \texttt{ptmcmc} and \texttt{multinest}, which as we recall were not really developed or tuned to this particular problem, spend a large number of likelihood evaluations, from a few $10^{7}$ to $10^{8}$, to produce a final set of posterior samples.
-This translates in reasonable computing costs thanks to our fast likelihoods, and is likely to be improvable in the future.
-
-To illustrate the role of higher harmonics in the analysis, we will present two classes of results: ``22'', where we inject and recover signals including only the dominant harmonic $h_{22}$, and ``HM'', where we both inject and recover with the higher harmonics $(h_{22}, h_{21}, h_{33}, h_{44}, h_{55})$.
-We do not attempt to recover an injection with higher modes using 22-mode waveforms; one expects parameter biases in this case, induced by the inadequacy of the simplified waveforms.
-A related question would be to assess the mode content necessary for waveform models to mitigate systematic biases when analyzing full GR signals.
-We leave such waveform systematics studies for future work; we only note that, as shown in Fig.~\ref{fig:modeoverlapsoftime}, from SNR only the set of five modes we include in the present study is quite obviously incomplete.
+We find that both our samplers, \texttt{ptmcmc} and \texttt{multinest}, require between $\sim 10^{7}$ and $\sim 10^{8}$ likelihood evaluations to produce a final set of posterior samples.
+Thanks to our fast likelihood implementation, this already represents a manageable computing cost.
+We stress again that the settings of both samplers were not optimized by taking into account the characteristics of the specific problem or the expected correlations between parameters.
+Therefore, we expect that the cost can be reduced even further with future optimizations.
+
+To illustrate the role of higher harmonics in the analysis, we will present two classes of results: ``22'', where we perform both the simulation and the inference by only including the dominant harmonic $h_{22}$, and ``HM'', where we perform both the simulation and the inference with all available harmonics $(h_{22}, h_{21}, h_{33}, h_{44}, h_{55})$.
+We do not consider the case where the simulation includes all harmonics and the inference is restricted to the 22 mode, as this will most likely bias the inferred parameters by an amount larger than the statistical uncertainties.
+In fact, the SNR calculation in Fig.~\ref{fig:modeoverlapsoftime} suggests that even the set of five modes included in this study may not be sufficient.
+The question of how many harmonics are necessary to keep the systematic errors below the statistical uncertainties is important and will be investigated in the future.
 
 Figs.~\ref{fig:PEsmbh22hmCase0} and~\ref{fig:PEsmbh22hmCase9} show posterior distributions for all parameters for Cases I and II respectively.
-We overlay ``22'' and ``HM'' posteriors.
+We overlay the ``22'' and ``HM'' posteriors.
 All results for MBHB systems are presented using LISA-frame parameters, as defined in App.~\ref{app:conventions}.
@@
 
-In both Cases I and II, including the higher harmonics has a major effect on the parameter recovery (as was already stressed in~\cite{Arun+07a, TriasSintes07, PorterCornish08, McWilliams+09}), much beyond what we would expect from the sole gain in total SNR shown in Table.~\ref{tab:MBHBparams} (and a scaling of statistical errors $\sim 1/\SNR$).
-The marginalized posterior of the masses is narrower, although not qualitatively different; the major change is in the extrinsic parameters.
-
-We can understand the effect of the higher harmonics by noting that, in a signal of the form $\sum_{\ell m} \sYlm(\iota, \varphi)h_{\ell m}$ where the $\sYlm$ all have a different dependency with the inclination $\iota$, measuring independently two separate contributions $(\ell m) \neq (\ell' m')$ gives us an independent measurement of $\iota$ by the relative amplitude of the two mode contributions.
+In both Cases I and II, including the higher harmonics greatly reduces the uncertainties in the inferred parameters, consistently with previous studies~\cite{Arun+07a, TriasSintes07, PorterCornish08, McWilliams+09}.
+The improvement is much larger than we would expect from the gain in total SNR and a scaling of statistical errors $\sim 1/\SNR$.
+The largest effect is found for the angular parameters and luminosity distance.
+The mass posteriors are slightly more precise, but remain qualitatively similar.
+
+We can understand the dramatic effect of the higher harmonics on distance and inclination by noting that, in a signal of the form $\sum_{\ell m} \sYlm(\iota, \varphi)h_{\ell m}$ where the $\sYlm$ all have a different dependency with the inclination $\iota$, measuring independently two separate contributions $(\ell m) \neq (\ell' m')$ gives us an independent measurement of $\iota$ by the relative amplitude of the two mode contributions.
 When only the $h_{22}$ mode is included, both the luminosity distance and inclination determine the overall signal amplitude and they are therefore degenerate.
-This is visible in Figs.~\ref{fig:PEsmbh22hmCase0} and~\ref{fig:PEsmbh22hmCase9}, where including the higher harmonics in the analysis gives a better determination of distance and inclination by more than an order of magnitude.
 

### 2020-02-18 paper/paper_PE.tex
More edits

@@
 by reflecting the wave vector $k$ across the plane of the LISA constellation.
-This secondary mode has less weight than the main mode around the injection, but is present in both cases.
-In particular, it survives when including higher harmonics.
-
-We can gain insight about degeneracies in the parameter space by considering the simple expressions representing the instrument response in the \textit{Frozen low-$f$} approximation described in Sec.~\ref{subsec:MBHBresponseapprox}, appropriate for the low-frequency limit and signals short enough that the LISA motion can be neglected; we will explore in details the respective influence of these neglected effects in Sec.~\ref{sec:MBHBPEacctime}.
+The secondary mode contains less probability than the ``correct'' mode close to the true parameters, but it is present in both simulations and it survives the inclusion of higher harmonics.
+
+We can gain insight about degeneracies in the parameter space by considering the simple expressions of the instrument response in the \textit{Frozen low-$f$} approximation described in Sec.~\ref{subsec:MBHBresponseapprox}, appropriate for the low-frequency limit and signals short enough that the LISA motion can be neglected; we will explore in details the respective influence of these neglected effects in Sec.~\ref{sec:MBHBPEacctime}.
 
@@
 
-Changing the sign of $\beta_{L}$ in the pattern functions $F_{a,e}^{+,\times}$ given in~\eqref{eq:FapcFepc} changes the sign of $F_{a,e}^{\times}$ for both channels, leaving $F_{a,e}^{+}$ invariant.
-In~\eqref{eq:Flmaerepeat} the $F_{a,e}^{+,\times}$ factors of the two lines are exchanged.
-Since spin-weighted spherical harmonics~\cite{Goldberg+67} obey the relation
+Changing the sign of $\beta_{L}$ in the pattern functions $F_{a,e}^{+,\times}$ given in~\eqref{eq:FapcFepc} has no effect on $F_{a,e}^{+}$ and it changes the sign of $F_{a,e}^{\times}$ for both channels.
+Thus, the $\left( F_{a,e}^{+} \pm i F_{a,e}^{\times}\right)$ factors in the two terms of Eq.~\eqref{eq:Flmaerepeat} are exchanged.
+Moreover, since spin-weighted spherical harmonics~\cite{Goldberg+67} obey the relation
 \be
@@
 \ee
-we see that changing $\iota \rightarrow \pi-\iota$, $\psi_{L} \rightarrow \pi - \psi_{L}$ leaves $F_{a}^{\ell m}$, $F_{e}^{\ell m}$ invariant.
+we see that simultaneously changing $\iota \rightarrow \pi-\iota$ and $\psi_{L} \rightarrow \pi - \psi_{L}$ leaves $F_{a}^{\ell m}$, $F_{e}^{\ell m}$ unchanged.
 
@@
 \end{align}
-How does this situation change when considering a more complete instrument response, moving away from the \textit{Frozen low-$f$} approximation ?
+How does this situation change when considering a more complete instrument response, moving away from the \textit{Frozen low-$f$} approximation?
 
@@
 
-Thus, we have found that in the \textit{Frozen low-$f$} approximation for the response we expect a pattern of degeneracy of eight degenerate positions in the sky (with certain rules for inclination and polarization).
+Thus, we have found that in the \textit{Frozen low-$f$} approximation for the response we expect a pattern of eight degenerate positions in the sky (with certain rules for inclination and polarization).
 Table~\ref{tab:MBHBdegen} summarizes the qualitative picture of degeneracy breaking by the features of the response.

### 2020-02-21 paper/paper_PE.tex
Small edits

--- 2020-02-17 paper/paper_PE.tex
@@
 Cosmological applications using LISA observations as standard sirens~\cite{Schutz86, Tamanini+16} also depend on the ability to localize individual sources.
-Parameter estimation of SBHBs will also be necessary for multiband gravitational-wave astronomy~\cite{Sesana16}.
-In this paper we focus on the latter part of the analysis problem, i.e.~the inference of the black hole parameters from the LISA data, leaving aside the detection problem and assuming the presence of the signal in the data.
+Understanding parameter estimation of SBHBs will also be important to understand the outcomes and challenges of possible multiband gravitational-wave observations~\cite{Sesana16}.
+In this paper we focus on the latter part of the analysis problem, i.e.~the inference of the black hole parameters from the LISA data, once we have reasons to believe the presence of the signal in the data, leaving aside the identification problem.
 We also limit our analysis to extrinsic parameters and ignore the effect of superposed gravitational wave signals from various sources.
 
-The inference problem amounts to drawing samples from the posterior distribution for source parameters, related by Bayes' theorem to the likelihood function, i.e.~the probability of observing the measured data given the source parameters and a model of the source and detector.
-In LISA, this problem is more complex than in kilometer-scale ground-based interferometers, as additional challenges arise from the much larger expected signal-to-noise ratios (SNRs) for high-mass systems, or the much longer duration of the waveform for low-mass systems.
+The inference problem amounts to produce samples from the posterior distribution for source parameters, related by Baye's theorem to the likelihood function, i.e.~the probability of observing the measured data given the source parameters and a model of the source and detector.
+In LISA, this problem is more complex than in kilometer-scale ground-based interferometers, as additional challenges arises from the much larger expected signal-to-noise ratios (SNRs) for high-mass systems, or the much longer duration of the waveform for low-mass systems.
 In particular, for SMBHBs, the large SNR will require a very accurate modeling of the waveform, including the merger and ringdown regimes, the effects of spin and corrections of higher harmonics in the signal beyond the quadrupole.
@@
 Previous parameter recovery studies with LISA used mainly simplified signal and response models, and often relied on a Fisher matrix approximation to parameter recovery instead of full Bayesian simulations.
-Numerous works used the combination of inspiral-only post-Newtonian waveforms (sometimes including precession), the low-frequency approximate response and Fisher matrix estimates~\cite{Cutler97, Vecchio03, Arun06, Berti+04, LangHughes06}.
+Numerous works used the combination of inspiral-only post-Newtonian waveforms (sometimes including precession), the low-frequency approximate response and Fisher matrix estimates; examples include~\cite{Cutler97, Vecchio03, Arun06, Berti+04, LangHughes06}.
 The Mock LISA Data Challenges~\cite{MLDC09} used such a setup for the signals and the response, albeit moving towards MCMC tools with a focus on detection.
-Bayesian methods going beyond Fisher, albeit still restricted to inspiral signals, were developed in~\cite{Brown+07, CornishPorter06a, Crowder+06, Wickham+06, Roever+07, Feroz+09, GairPorter09, Petiteau+09, PorterCarre13, PorterCornish15}.
+Bayesian methods going beyond Fisher albeit still with inspiral signals were developed in~\cite{Brown+07, CornishPorter06a, Crowder+06, Wickham+06, Roever+07, Feroz+09, GairPorter09, Petiteau+09, PorterCarre13, PorterCornish15}.
 The importance of the higher harmonics for LISA was stressed already in several studies~\cite{Arun+07a, TriasSintes07, PorterCornish08, McWilliams+09}.
@@
 
-Here we demonstrate that standard Bayesian inference can be performed for SMBHBs using a self-consistent waveform model that includes inspiral, merger, ringdown and higher-order modes (albeit no spins) and a full model of the LISA response.
+Here we demonstrate that standard Bayesian inference can be performed for SMBHBs using a self-consistent waveform model that includes inspiral, merger, ringdown and higher-order modes (albeit still simplified as spins are neglected) and a full model of the LISA response.
 We investigate and explain degeneracies in the posterior distribution of the source's parameters, and highlight the crucial role played by the higher harmonics in the signal and the frequency-dependency in the instrument response.
@@
 We start by introducing a fast method to calculate the likelihood for a black hole merger.
-This method includes: a fast computation of a Fourier-domain inspiral-merger-ringdown waveform with higher-order modes based on ROM; a fast Fourier-domain computation of the LISA response based on~\cite{MB18}; and a fast method for computing the noise-weighted product between waveforms when setting the noise realization to zero.
-We couple our likelihood with two codes for Bayesian inference based on different sampling techniques, constructing an inference engine which enables us to perform a variety of investigations with simulated black hole mergers.
+This method includes: a fast computation of a Fourier-domain inspiral-merger-ringdown waveform with higher-order modes based on reduced-order modeling; a fast Fourier-domain computation of the LISA response based on~\cite{MB18}; and a fast method for computing the noise-weighted product between waveforms when setting the noise realization to zero.
+We couple our likelihood with two codes for Bayesian inference based on different sampling techniques, constructing an inference engine which will allow us in the future to perform a variety of investigations with simulated black hole mergers.
 
@@
 
-Finally, we repeat some of our investigations for stellar-origin black hole mergers.
+Finally, we repeat some of our investigations for stellar-origin black hole mergers as well.
 Parameter estimation studies for SBHB systems have relied so far on the Fisher matrix approach~\cite{Sesana16, Vitale16, Nishizawa+16a, Nishizawa+16b}, with Bayesian inference tools yet to be developed.
@@
 In this section, we introduce a complete set of notations for the LISA instrument response that will be useful in the later discussion of degeneracies in parameter space.
-We will use units with $G=c=1$ and assume the proposed armlength for LISA $L = 2.5\mathrm{Gm}$~\cite{LISA2017}.
+We will use units with $G=c=1$.
 
@@
 \end{align}
-Writing the $P_{\ell m}$ matrices in this way allows us to factor out explicitly all dependencies in the extrinsic parameters $(\iota, \phi, \psi)$.
+Writing the $P_{\ell m}$ matrices in this way allows to factor out explicitly all dependencies in the extrinsic parameters $(\iota, \phi, \psi)$.
 Together with the luminosity distance $D$ scaling the overall amplitude of the signal, these parameters enter as constant (time and frequency independent) prefactors in the response for each mode $h_{\ell m}$.
 
-Since $\psi$ always appears with a factor of 2, it has an exact $\pi$-degeneracy, and we choose the convention of restricting $\psi \in [0,\pi]$.
+Since only $2\psi$ intervenes, the polarization $\psi$ has an exact $\pi$-degeneracy; by convention we set $\psi \in [0,\pi]$.
 
@@
 
-The LISA response~\cite{EW75, Cutler97, Larson+99, CR02, RCP04} can be built from single-link observables $y_{slr} = (\nu_{r} - \nu_{s})/\nu$ representing a laser frequency shift between the transmitting spacecraft $s$ and the receiving spacecraft $r$ along the link $l$.
+The LISA response~\cite{EW75, Cutler97, Larson+99, CR02, RCP04} can be built from single-link observables $y_{slr} = (\nu_{r} - \nu_{s})/\nu$ representing a laser frequency shift between the sending spacecraft $s$ and the receiving spacecraft $r$ along the link $l$.
 We use the expression~\cite{Vallisneri04, Krolak+04}
@@
 \ee
-with $H$ the transverse-traceless matrix representing the gravitational wave, $L$ the delay along one arm, taken to be fixed, $n_{l}$ the link unit vectors (from $s$ to $r$) and $p_{s}$, $p_{r}$ the positions of the spacecrafts.
+with $H$ the transverse-traceless matrix representing the gravitational wave, $L$ the delay along one arm\footnote{We recall that we use units with $G=c=1$.
+The proposed armlength for LISA is $L = 2.5\mathrm{Gm}$~\cite{LISA2017}.}, taken to be fixed, $k$ the propagation vector, $n_{l}$ the link unit vectors (from $s$ to $r$) and $p_{s}$, $p_{r}$ the positions of the spacecrafts.
 Here $n_{l}$, $p_{s}$ and $p_{r}$ are evaluated at the same time $t$.
 
-In the frequency domain, modes will be represented with their amplitude and phase as
+In the frequency domain, modes will be represented with their amplitude and phase as\footnote{Note our Fourier convention~\eqref{eq:defFourier}.}
 \be\label{eq:hlmampphase}
@@
 
-In~\eqref{eq:defyslr} and~\eqref{eq:Gslr}, one can distinguish two types of delays: on one hand the delay associated to the position $p_{0}$ of the center of the constellation on its orbit around the Sun, with baseline $R=1 \au$, and on the other hand the delays associated to the individual spacecraft positions in the constellation, with baseline the armlength $L$.
+In~\eqref{eq:defyslr} and~\eqref{eq:Gslr}, one can distinguish two types of delays: on one hand the delay associated to the position $p_{0}$ of the center of the constellation on its orbit around the Sun, with baseline $R=1 \au$, and on the other hand the delays associated to the individual spacecraft positions in the constellation, with baseline the armlength $L=2.5 \cdot 10^{6 }\mathrm{km}$.
 This defines the transfer frequencies
@@
 
-The basic one-arm observables of Eq.~\eqref{eq:defyslr} are affected by laser noise whose amplitude is orders of magnitude larger than the astrophysical signal.
-However, time delay interferometry (TDI) allows one to construct a new set of observables from delayed combinations of $y_{slr}$, where laser noise is suppressed by orders of magnitude~\cite{TintoArmstrong99, Armstrong+99, Estabrook+00, Dhurandhar+02, Tintoliving}.
-Various generations of TDI schemes have been proposed in order to deal with a non-rigid and rotating LISA constellation~\cite{Shaddock03, CornishHellings03, Shaddock+03, Tinto+04}.
+To mitigate laser noise, LISA requires Time Delay Interferometry (TDI)~\cite{TintoArmstrong99, Armstrong+99, Estabrook+00, Dhurandhar+02, Tintoliving}.
+These TDI observables are built from delayed combinations of the one-arm basic observables~\eqref{eq:defyslr}.
+Detailed cancellations for a flexing and rotating constellation prompted the development of higher-generation TDI schemes~\cite{Shaddock03, CornishHellings03, Shaddock+03, Tinto+04}.
 However, these refinements affect only marginally the response to the gravitational waves.
-Hence, in this work we only consider first-generation TDI and adopt a rigid approximation for the constellation, where delays are all constant and equal to $L$.
+We therefore limit ourselves to first-generation TDI and adopt a rigid approximation for the constellaion, where delays are all constant and set to $L$.
 Using the notation $y_{slr,nL} = y_{slr}(t - nL)$, the first-generation TDI Michelson observable $X$ reads~\cite{Vallisneri04}
@@
 \end{align}
-with the other Michelson observables $Y$, $Z$ being obtained by cyclic permutation.
-The uncorrelated-noise combinations $A$, $E$ and $T$~\cite{Prince+02} are then defined as
+with the other Michelson observables $Y$, $Z$ being obtained by cyclic permutation. Uncorrelated combinations $A$, $E$ and $T$~\cite{Prince+02} are then defined as
 \bsub\label{eq:defAET}
@@
 \esub
-These channels are independent under the assumption of an identical and uncorrrelated noise in the detector arms. Note that various conventions coexist in the literature.
-With constant delays in the rigid approximation, and using the notation $z\equiv \exp[2i\pi fL]$, the TDI combinations in the frequency domain take the form
+These channels are independent under the assumption of an identical and uncorrrelated noise in the detector arms.
+Note that various conventions coexist in the literature.
+With constant delays in the rigid approximation, with the notation $z\equiv \exp[2i\pi fL]$ the TDI combinations take the form
 \bsub\label{eq:aet}
 \begin{align}
-	\tilde{a} &= (1+z) \left( \tilde{y}_{31} + \tilde{y}_{13} \right) - \tilde{y}_{23} - z \tilde{y}_{32} - \tilde{y}_{21} - z \tilde{y}_{12} \,,\\
+	\tilde{a} &= \left[ (1+z) \left( \tilde{y}_{31} + \tilde{y}_{13} \right) - \tilde{y}_{23} - z \tilde{y}_{32} - \tilde{y}_{21} - z \tilde{y}_{12} \right] \,,\\
 	\tilde{e} &= \frac{1}{\sqrt{3}} \left[ (1-z)\left( \tilde{y}_{13} - \tilde{y}_{31} \right) + (2+z) \left( \tilde{y}_{12} - \tilde{y}_{32} \right) \right. \nn\\
@@
 
-In the low-frequency limit, also called the long-wavelength approximation, the effects due the finite LISA armlength become negligible and the LISA response becomes equivalent to the response of two coplanar LIGO-like detectors rotated by $\pi/4$ relative to each other and set in motion~\cite{Cutler97}.
+As is well known~\cite{Cutler97}, in the low-frequency limit (also called the long-wavelength approximation), the finite-armlength effects vanish, and the response of LISA is analogous to the response of of two LIGO-type detectors  rotated from each other by $\pi/4$ and set in motion.
 
@@
 \esub
-Here, we have made apparent factors $(-2i\pi f)$, which correspond to Fourier-domain time derivatives.
-The observables $\tilde{a}$, $\tilde{e}$ are therefore analogous to time derivatives of the strain commonly used for ground-based detectors.
+Here, we have made apparent factors $(-2i\pi f)$ corresponding to the Fourier-domain translation of a time derivative.
+The observables $\tilde{a}$, $\tilde{e}$ are analogous to a time derivative of the strain more commonly used for ground-based detectors.
 
 We can now map these two channels to two fictitious LIGO-type detectors as follows.
-For an orthogonal detector of the same setup as the ground-based LIGO and Virgo with armlength $L_{D}$, rotated by an angle $\epsilon_{D}$ from the basis vectors $x,y$, the detector tensor is
+For an orthogonal detector of the same setup as the ground-based LIGO and Virgo with armlength $L_{D}$, rotated by an angle $\epsilon_{D}$ from the basis vectors $x,y$, the detector tensor is\S
 \begin{align}
@@
 \ee
-for stationary Gaussian noise with PSD $S_{n}$, the likelihood $\calL = p(d | \bm{\theta})$ for a given gravitational wave channel takes the form
+for stationary Gaussian noise with noise PSD $S_{n}$, the likelihood $\calL = p(d | \bm{\theta})$ for a given gravitational wave channel takes the form
 \be\label{eq:deflnL}
@@
 with the data stream being a superposition of the gravitational wave signal for the true parameters $\bm{\theta}_{0}$ and the noise realization in the experiment, $d = h(\bm{\theta}_{0}) + n$.
-The posterior distribution for the physical parameters $\bm{\theta}$ given the observed data $d$ is then \be\label{eq:defBayes}
+In general, the templates $h(\theta)$ used for the analysis will only be an approximation of the physical signals entering the data stream $d$; we will ignore entirely this distinction and neglect the systematic errors in the waveform model.
+The posterior distribution for the physical parameters $\bm{\theta}$ given the observed data $d$ is then
+\be\label{eq:defBayes}
 	p\left( \bm{\theta} | d \right) = \frac{p(d | \bm{\theta})p(\bm{\theta})}{p(d)}\,,
 \ee
-with $p(\bm{\theta})$ the priors on the parameters, and $p(d)$ the evidence.
-The focus of this work is not selecting between different models, so we will treat $p(d)$ as a normalization constant and do not consider it further.
-
-In general, the templates $h(\bm{\theta})$ used for the analysis will only be an approximation of the physical signals present in the data stream $d$.
-In this work we will ignore entirely this distinction and assume that our template waveforms exactly match the astrophysical waveforms.
-In other words, we will not explore the issue of systematic errors arising from incomplete modeling of compact binary merger waveforms.
-
-In our analysis, we will simulate signals and perform Bayesian analyses of them with Eq.~\eqref{eq:deflnL} and ~\eqref{eq:defBayes} to obtain the parameter posteriors.
-In doing so, we will use the so-called zero-noise approximation, i.e.~setting $n=0$.
-This allows us to accelerate the likelihood computation as explained in Sec.~\ref{subsec:likelihood}.
-The approximation is sufficient to explore the structure of the likelihood and the parameter degeneracies, greatly improving with respect to the Fisher matrix approach.
-The extension of our analysis to include nonzero noise realizations will be discussed in future work.
+with $p(\bm{\theta})$ the priors on the parameters, and $p(d)$ the evidence that we will ignore here as it enters as a multiplicative constant in the posterior.
+
+In our analysis, we will use the so-called zero-noise approximation.
+This corresponds simply to performing a Bayesian analysis with~\eqref{eq:deflnL} and ~\eqref{eq:defBayes}, but setting the noise realization to zero, $n=0$.
+It allows us to accelerate the likelihood computation, as will be explained in~\ref{subsec:likelihood}.
+For exploratory parameter estimation runs, this approximation allows us to explore the structure of the likelihood with its possible parameter degeneracies, and is a significant improvement over Fisher matrix estimates in that sense.
+We leave the extension of our analysis to include noise realizations for future work.
+We will also ignore waveform systematic errors and assume our waveform templates match exactly the real signals.
 
@@
 \end{align}
-with the ``$\rm inj$'' subscript indicating the simulated signals, and where the inner products $(\cdot | \cdot)$ are given by~\eqref{eq:definnerproduct} with the noise PSDs~\eqref{eq:Snaet}.
+with the $\rm inj$ subscript indicating the injected signals, and where the inner products $(\cdot | \cdot)$ are given by~\eqref{eq:definnerproduct} with the noise PSDs~\eqref{eq:Snaet}.
 
@@
 Since higher harmonics will play a crucial role in our analysis, in this study we use the Effective-One-Body (EOB) waveform model \texttt{EOBNRv2HM}~\cite{Pan+11}.
-These waveforms are based on the EOB formalism~\cite{BD99, BD00}, and are calibrated to numerical relativity waveforms.
+These waveforms are based on the EOB formalism~\cite{BD99, BD00}, and are calibrated on a set of numerical relativity waveforms.
 The model is limited to non-spinning systems on quasicircular orbits, but includes a set of higher harmonics in the signal, among the most important quantitatively:
@@
 \ee
-The waveforms are generated in the time domain by integrating a system of ordinary differential equations, an operation requiring up to seconds of computing time for long signals.
-For this reason, we developed a reduced order model (ROM) for these waveforms, \texttt{EOBNRv2HMROM}, following the methods of~\cite{Puerrer14}, enabling a much faster generation of the Fourier-domain amplitudes and phases of the modes in Eq.~\eqref{eq:hlmampphase}.
-As the total mass $M$ scales out of the problem, the parameter space reduces to the mass ratio $q=m_{1}/m_{2}$ only.
-The parameter space interpolation in this ROM is therefore significantly simpler than the models~\cite{Puerrer14, Puerrer15, Bohe+16}, which include aligned spins.
-The inclusion of higher harmonics, however, requires modeling the relative dephasing of different modes.
+The waveforms are generated in the time domain by integrating a system of ODEs, an operation requiring up to seconds of computing time for long signals.
+For this reason, we developed a reduced order model (ROM) for these waveforms, \texttt{EOBNRv2HMROM}, following the methods of~\cite{Puerrer14}, enabling a much faster generation of the Fourier-domain amplitudes and phases of the modes~\eqref{eq:hlmampphase}.
+As the total mass $M$ scales out of the problem, the parameter space is effectively only one-dimensional, with only the mass ratio $q=m_{1}/m_{2}$.
+The parameter space interpolation in this ROM is therefore significantly simpler than the models~\cite{Puerrer14, Puerrer15, Bohe+16}, which was three-dimensional as it included aligned spins.
+The inclusion of higher harmonics, however, imposes to model the relative dephasing of different modes.
 The output of the code consists of a Fourier domain amplitude and phase for each mode, sparsely sampled on $\sim 300$ frequencies.
-The faithfulness with the original waveforms, assuming Advanced LIGO noise curves~\cite{UnofficialNoiseCurves18}, is $\lesssim 10^{-4}$.
-The computational cost is submillisecond and smaller than other stages of our likelihood computation.
+The faithfulness with the original waveforms was measured with \texttt{aLIGO} noise curves~\cite{UnofficialNoiseCurves18} to be $\lesssim 10^{-4}$, while the computational cost is submillisecond and subdominant with respect to other stages of our likelihood computation.
 
@@
 Nevertheless, consistently including merger, ringdown and higher harmonics already represents a significant improvement over previous studies, as these features in the signal carry a lot of SNR by themselves~\cite{Babak+08, Thorpe+08, McWilliams+09, McWilliams+10, McWilliams+11, Klein+15}.
-In the absence of precession, we expect intrinsic and extrinsic parameters to be weakly correlated, so that  our investigations of the extrinsic part of the likelihood should be valid for aligned spins.
+In the absence of precession at least, we expect intrinsic and extrinsic parameters to correlate weakly, so that  our investigations of the extrinsic part of the likelihood should be valid for aligned spins.
 
@@
 In general, both the template and the injection in~\eqref{eq:deflnL} are given as frequency series sampled with the Nyquist criterion $\Delta f = 1/(2T)$ with $T$ the duration of the signal, and the overlaps~\eqref{eq:definnerproduct} are simply computed as discrete sums over those frequency samples.
-Although transforming~\eqref{eq:definnerproduct} in a discrete sum is a straightforward operation, the size of the frequency series can lead to a significant cost.
+Although transforming~\eqref{eq:definnerproduct} in a discrete sum is a straightforward operation, the size of the frequency series alone can represent a significant cost.
 In the context of LIGO/Virgo observations, this has prompted the development of Reduced Order Quadratures (ROQ)~\cite{Smith+14, Canizares+14, Smith+16} to accelerate the likelihood computation.
@@
 This cost can range from tens of milliseconds for SMBHBs and short-lived signals (a few days), to impracticable values for SBHBs signals lasting for years and sampled at high frequency.
-By setting the noise realization to zero, however, we can represent the amplitude and phase of the signals over a sparse frequency grid and obtain a large speedup.
+Setting the noise realization to zero, however, we can exploit an amplitude/phase representation of the signals, use a sparse frequency grid and get a large speedup.
 
@@
 
-The difficulty of numerically computing such an overlap depends on the phase difference.
-A large phase difference causes the integrand to be very oscillatory, requiring more frequency resolution than non-oscillatory integrands.
-On the other hand, oscillatory integrands result in a small integral due to cancellation effects, while non-oscillatory integrands contribute the most.
+The difficulty of computing numerically such an overlap depends on the phase difference.
+A large phase difference causes the integrand to be very oscillatory, requiring more frequency resolution than non-oscillatory integrands; on the other hand, oscillatory integrands give raise to a small integral due to cancellation effects, while non-oscillatory integrands contribute the most.
 For a single mode, $\mathrm{SNR}$ terms of the type $(h|h)$ have a zero phase difference by construction, but including different modes generates cross-terms $(\ell, m) \neq (\ell', m')$ with a large phase difference.
@@
 Since the number of overlaps to compute in~\eqref{eq:defTlmaet} is quadratic in the number of modes, likelihoods with higher harmonics are significantly more expensive.
-The final likelihood cost depends also on the number of grid intervals, that is chosen to limit spline interpolation errors (Table~\ref{tab:likecost}).
-\begin{table}
-	\begin{tabular}{|c|c|c|c|}
-		\hline
-		Source type & Modes & Grid points & Cost \\
-		\hline
-		SBHB & $h_{22}$ & $500$ & $2$ ms \\
-		SMBHB & $h_{22}$ & $300$ & $1.2$ ms \\
-		SMBHB & $5$ & $300$ & $15$ ms \\
-		\hline
-	\end{tabular}
-	\label{tab:likecost}
-	\caption{Approximate frequency grid resolution and likelihood cost for different choices of source type and number of modes in the waveform.}
-\end{table}
-%\begin{itemize}
-%	\item for SBHBs with $h_{22}$ only: $\sim 500$ grid points, likelihood cost $\sim 2\mathrm{ms}$;
-%	\item for SMBHBs with $h_{22}$ only: $\sim 300$ grid points, likelihood cost $\sim 1.2\mathrm{ms}$;
-%	\item for SMBHBs with 5 modes: $\sim 300$ grid points, likelihood cost $\sim 15\mathrm{ms}$.
-%\end{itemize}
+The final likelihood cost depends also on the number of grid intervals, that is chosen to limit spline interpolation errors:
+\begin{itemize}
+	\item for SBHBs with $h_{22}$ only: $\sim 500$ grid points, likelihood cost $\sim 2\mathrm{ms}$;
+	\item for SMBHBs with $h_{22}$ only: $\sim 300$ grid points, likelihood cost $\sim 1.2\mathrm{ms}$;
+	\item for SMBHBs with 5 modes $h_{\ell m}$: $\sim 300$ grid points, likelihood cost $\sim 15\mathrm{ms}$.
+\end{itemize}
 The accuracy of our likelihood computations is typically $|\Delta \ln \calL| \lesssim 0.2$, due to the numerical errors being magnified by the required cancellations between overlaps terms in~\eqref{eq:lnLoverlapstructure}.
@@
 In all panels the black line represents the total.
-The horizontal axis gives the time to merger, with vertical lines for the cuts of~\ref{subsec:MBHBacctime}, and the dots on the right of each panel show the result for the full signal including the post-merger.}
+The horizontal axis gives the time to merger, with vertical lines for the cuts of~\ref{subsec:MBHBacctime}, and the dots on the right of each panel show the result for the full signal including the post-merger.
+\SM{Make clear it is $|(h|h')|$ for cross terms -- changes sign}}
   \label{fig:modeoverlapsoftime}

### 2020-02-21 paper/paper_PE.tex

--- 2020-02-18 paper/paper_PE.tex
@@
 In all panels the black line represents the total.
-The horizontal axis gives the time to merger, with vertical lines for the cuts of~\ref{subsec:MBHBacctime}, and the dots on the right of each panel show the result for the full signal including the ringdown.}
+The horizontal axis gives the time to merger, with vertical lines for the cuts of~\ref{subsec:MBHBacctime}, and the dots on the right of each panel show the result for the full signal including the ringdown.
+\SM{Make clear it is $|(h|h')|$ for cross terms -- changes sign}}
   \label{fig:modeoverlapsoftime}

### 2020-02-21 paper/paper_PE.tex
Remove package ulem
Small edits

@@
 \usepackage{enumerate}
-\usepackage{ulem}
+%\usepackage{ulem}
 \usepackage{tabularx}
@@
 We start by introducing a fast method to calculate the likelihood for a black hole merger.
-This method includes: a fast computation of a Fourier-domain inspiral-merger-ringdown waveform with higher-order modes based on ROM; a fast Fourier-domain computation of the LISA response based on~\cite{MB18}; and a fast method for computing the noise-weighted product between waveforms when setting the noise realization to zero.
+This method includes: a fast computation of a Fourier-domain inspiral-merger-ringdown waveform with higher-order modes based on a reduced order model; a fast Fourier-domain computation of the LISA response based on~\cite{MB18}; and a fast method for computing the noise-weighted product between waveforms when setting the noise realization to zero.
 We couple our likelihood with two codes for Bayesian inference based on different sampling techniques, constructing an inference engine which enables us to perform a variety of investigations with simulated black hole mergers.
@@
 \ee
-with $H$ the transverse-traceless matrix representing the gravitational wave, $L$ the delay along one arm, taken to be fixed, $n_{l}$ the link unit vectors (from $s$ to $r$) and $p_{s}$, $p_{r}$ the positions of the spacecrafts.
+with $H$ the transverse-traceless matrix representing the gravitational wave, $k$ the wave propagation vector, $L$ the delay along one arm, taken to be fixed, $n_{l}$ the link unit vectors (from $s$ to $r$) and $p_{s}$, $p_{r}$ the positions of the spacecrafts.
 Here $n_{l}$, $p_{s}$ and $p_{r}$ are evaluated at the same time $t$.
@@
 However, scaling these factors out from both the signal and the noise is an approximation, as we can expect that realistic delays would lead to imperfect cancellations; in the vicinity of the zero-crossings, we should therefore expect a localized loss of sensitivity.
+\SM{make clear approximation already baked in}
 
@@
 \esub
-Here, we have made apparent factors $(-2i\pi f)$, which correspond to Fourier-domain time derivatives.
+Here, we have made apparent factors $(-2i\pi f)$, which correspond to a time derivative in the Fourier-domain.
 The observables $\tilde{a}$, $\tilde{e}$ are therefore analogous to time derivatives of the strain commonly used for ground-based detectors.
@@
 \esub
-These functions are the familiar pattern functions of ground-based detectors (for $\psi_{L}=0$).
+These functions are the familiar pattern functions of ground-based detectors (for $\psi_{L}=0$) \SM{[add citation]}.
 One can check that the expressions for the channel $e$ are obtained from the expressions for $a$ with the replacement $\lambdaL \rightarrow \lambdaL + \pi/4$.
@@
 
-Introducing the standard matched-filter inner product (also called overlap) as
+Introducing the standard matched-filter inner \SM{[add reference]} product (also called overlap) as
 \be\label{eq:definnerproduct}
@@
 		\hline
-		Source type & Modes & Grid points & Cost \\
-		\hline
-		SBHB & $h_{22}$ & $500$ & $2$ ms \\
-		SMBHB & $h_{22}$ & $300$ & $1.2$ ms \\
-		SMBHB & $5$ & $300$ & $15$ ms \\
+		Source type & Modes & Grid & Cost \\
+		\hline
+		SBHB & $(2,2)$ & $500$ & $2$ ms \\
+		SMBHB & $(2,2)$ & $300$ & $1.2$ ms \\
+		SMBHB & $(2,2),(2,1),(3,3),(4,4),(5,5)$ & $300$ & $15$ ms \\
 		\hline
@@
    \includegraphics[width=.99\linewidth]{../plots/tflm_smbh.pdf}
-  \caption{Relation between inspiral time and frequency for MBHB systems I and II with $M=2\times 10^6 \; M_{\odot}$, $q=3$ and $z=4$.
+  \caption{Relation between time and frequency for MBHB systems I and II with $M=2\times 10^6 \; M_{\odot}$, $q=3$ and $z=4$.
 The different lines correspond to the $(\ell, m)$ harmonics available in the waveform model.
@@
 We stress again that the settings of both samplers were not optimized by taking into account the characteristics of the specific problem or the expected correlations between parameters.
-Therefore, we expect that the cost can be reduced even further with future optimizations.
+Therefore, we expect that the cost can be reduced with future optimizations.
 

### 2020-02-21 paper/paper_PE.tex
Proposed fixes/edits on Feb 21

--- 2020-02-21 paper/paper_PE.tex
@@
 However, these refinements affect only marginally the response to the gravitational waves.
-We therefore limit ourselves to first-generation TDI and adopt a rigid approximation for the constellaion, where delays are all constant and set to $L$.
+We therefore limit ourselves to first-generation TDI and adopt a rigid approximation for the cancellation, where delays are all constant and set to $L$.
 Using the notation $y_{slr,nL} = y_{slr}(t - nL)$, the first-generation TDI Michelson observable $X$ reads~\cite{Vallisneri04}
@@
 In all panels the black line represents the total.
-The horizontal axis gives the time to merger, with vertical lines for the cuts of~\ref{subsec:MBHBacctime}, and the dots on the right of each panel show the result for the full signal including the post-merger.
-\SM{Make clear it is $|(h|h')|$ for cross terms -- changes sign}}
+The horizontal axis gives the time to merger, with vertical lines for the cuts of~\ref{subsec:MBHBacctime}, and the dots on the right of each panel show the result for the full signal including the post-merger.}
   \label{fig:modeoverlapsoftime}
@@
 We will show that we can understand this feature as a projection effect in the multidimensional degenerate posterior.
-Understanding the structure of these features could prove useful to inform Bayesian samplers, for instance by building adapted jump proposal allowing a much faster mixing of MCMC chains than naive methods.
+Understanding the structure of these features could prove useful to inform Bayesian samplers, for instance by adapting jump proposals to speed up mixing of MCMC chains.
 
@@
 
-The rate at which parameter information accumulates LISA merger observations is crucial for planning strategies for multi-messenger observations and in guiding the latency demands of the LISA down-link and data processing~\cite{ArmitageNatarajan02, DalCanton+19}.
+The rate at which parameter information accumulates during the observation of an inspiral is crucial for establishing the LISA downlink and data processing requirements, as well as for planning multimessenger observations~\cite{ArmitageNatarajan02, DalCanton+19}.
 In this Section, we explore how parameter information accumulates on approach to merger by performing parameter estimation studies with temporally truncated signals, as explained in~\ref{subsec:MBHBacctime}: the signals are cut at points where the accumulated SNR is about $\{10,42,166\}$ corresponding to $\{1/64,1/16,1/4\}$ of the total SNR=$666.0$, and times $\{41 \mathrm{hr}, 2.5 \mathrm{hr}, 7 \mathrm{min}\}$ before merger.
-As shown in Fig.~\ref{fig:modeoverlapsoftime}, most of the SNR accumulates over the last few minutes of signal.
+As shown in Fig.~\ref{fig:modeoverlapsoftime}, most of the SNR accumulates over the last few minutes of this signal.
 
@@
   \caption{Refinement of parameter inference over time for system II.
-The posteriors shown correspond to the time cuts defined in~\ref{subsec:MBHBacctime} at $\mathrm{SNR}/64$ (green), $\mathrm{SNR}/16$, $\mathrm{SNR}/4$ and post-merger (blue)., for the masses (left), distance/inclination (center), and time/phase (right).
+The posteriors shown correspond to the time cuts defined in~\ref{subsec:MBHBacctime} at $\mathrm{SNR}/64$ (green), $\mathrm{SNR}/16$, $\mathrm{SNR}/4$ and post-merger (blue) for the masses (left), distance/inclination (center), and time/phase (right).
 We inject and recover with higher harmonics.
@@
 The dashed black curve corresponds to the \textit{reflected} position, and the dashed red to the \textit{antipodal} position.
-At the injection (solid black), $\ln\calL$ is zero by construction while it is decreasing as a function of $f$ for other sky positions.
-Reaching a very negative $\ln\calL$ means that this sky position is excluded by the signal accumulated thus far.}
+At the true position (solid black), $\ln\calL$ is zero by construction while it is decreasing as a function of $f$ for other sky positions.
+Reaching a very negative $\ln\calL$ means that this sky position is excluded by the data accumulated thus far.}
   \label{fig:lnLskymodesapprox}
@@
 
-We have found that pre-merger analysis of the previous section made apparent the 8-modes degeneracy pattern that can be predicted analytically from the structure of the \textit{Frozen, low-$f$} response approximation (see~\eqref{eq:degenreflection}, \eqref{eq:degenlambdapi2}).
-Analyses of IMR signals as in~\ref{subsec:SMBHPEfull} show that only the \textit{reflected} mode~\eqref{eq:degenreflection} survives post-merger.
+The pre-merger analysis of the previous section made apparent the 8-modes degeneracy pattern that can be predicted analytically from the structure of the \textit{Frozen, low-$f$} response approximation (see~\eqref{eq:degenreflection}, \eqref{eq:degenlambdapi2}).
+However, the analysis of IMR signals in~\ref{subsec:SMBHPEfull} shows that only the \textit{reflected} mode~\eqref{eq:degenreflection} survives post-merger.
 In this Section we explain how and why this transition occurs.
 
-We have already presented a discussion of different qualitative effects in the instrument response in Sec.~\ref{subsec:MBHBresponseapprox} and Fig.~\ref{fig:responseapproxtdia22}: one hand hand, the motion of LISA leaves an imprint (the time-dependency) on the low frequencies, where the time-frequency map is steep enough that a short interval in frequency maps to a large interval of time; on the other hand, the breakdown of the long-wavelength approximation leaves an imprint (the frequency-dependency) at high frequencies.
+We have already presented a discussion of different qualitative effects in the instrument response in Sec.~\ref{subsec:MBHBresponseapprox} and Fig.~\ref{fig:responseapproxtdia22}: on one hand, the motion of LISA leaves an imprint (the time-dependency) on the low frequencies, where the time-frequency map is steep enough that a short interval in frequency maps to a large interval of time; on the other hand, the breakdown of the long-wavelength approximation leaves an imprint (the frequency-dependency) at high frequencies.
 Here we look at the quantitative importance of these features on the inference as a function of frequency (or as a function of time).
@@
 
-A value of $\ln \calL = 0$ means that the template signal with shifted parameters is identical to the injection, while a very negative $\ln \calL$ means that these parameter values are excluded in the inference.
+A value of $\ln \calL = 0$ means that the template signal with shifted parameters is identical to the injection, while a very negative $\ln \calL$ means that these parameter values are ruled out.
 The injection (full black line) has always $\ln \calL = 0$ by construction.
@@
 
-With the \textit{Low-$f$} response, the \textit{reflected} mode remains exactly degenerate as anticipated in Sec.~\ref{subsec:MBHBPEdegen}.
+With the \textit{Frozen} response, the \textit{reflected} mode remains exactly degenerate as anticipated in Sec.~\ref{subsec:MBHBPEdegen}.
 All other modes see their $\ln \calL$ fall rapidly at around $2\mHz$ due to the onset of frequency-dependency in the response.
@@
 
-This is indeed what happens in multidmensional Bayesian parameter estimation, and is illustrated in Fig.~\ref{fig:lnLskymodesapprox}: in the first panel, for the three pre-merger analyses as well as for the post-merger analysis, we overlay circles indicating the best $\ln \calL$ found among the posterior samples, in each eighth of the sky corresponding to the eight sky modes (when no samples are present in this region of the sky, nothing is displayed).
-A sampler is not an optimizer, so the precise value achieved is not optimal and would vary when repeating runs; nevertheless, this measure gives a good proxy for how much closer of the injected signal we could get by slightly biasing the parameters.
+This is indeed what happens in multidimensional Bayesian parameter estimation, and is illustrated in Fig.~\ref{fig:lnLskymodesapprox}: in the first panel, for the three pre-merger analyses as well as for the post-merger analysis, we overlay circles indicating the best $\ln \calL$ found among the posterior samples, in each eighth of the sky corresponding to the eight sky modes (when no samples are present in this region of the sky, nothing is displayed).
+A sampler is not an optimizer, so the precise value achieved is not optimal and would vary when repeating runs; nevertheless, this measure gives a good proxy for how much closer to the injected signal we could get by slightly biasing the parameters.
 
@@
 The parameters of the simulated mergers were drawn from the Radler LISA Data Challenge (LDC), which itself followed the population of~\cite{Sesana16}, and are listed in Table~\ref{tab:SBHBparams}.
-The system I is a moderately massive system, placed at a favorable close distance; with an SNR of $26.6$ it would be at the tail of the most favorable detections that we can hope for~\cite{Sesana16, Gerosa+19, Moore+19}.
-The system II is a higher-mass, moderate-SNR system.
+System I is moderately massive, placed at a favorable close distance; with an SNR of $26.6$ it would be at the tail of the most favorable detections that we can hope for~\cite{Sesana16, Gerosa+19, Moore+19}.
+System II is a higher-mass, moderate-SNR system.
 Both systems happen to be close to the ecliptic plane.
 
-Fig.~\ref{fig:transferSBHBcaseI} illustrates, for the system I in this table, the instrumental transfer functions~\eqref{eq:defTlmaet} as well as the Doppler phase~\eqref{eq:defPhiR}.
+Fig.~\ref{fig:transferSBHBcaseI} illustrates the instrumental transfer functions~\eqref{eq:defTlmaet} as well as the Doppler phase~\eqref{eq:defPhiR} for system I.
 Contrasting with the SMBHB case, we see that the transfer functions have much more structure in this case.
 Both the time-dependency and the frequency-dependency in the response lead to oscillatory features in the response.
-The Doppler phase~\eqref{eq:defPhiR}, that was negligible for our short SMBHB systems, represents now tens of radians.
+The Doppler phase~\eqref{eq:defPhiR}, negligible for our short SMBHB systems, now varies by tens of radians.
 
@@
 inspiral of SBHB systems, the signal is well within the domain of applicability of the post-Newtonian approximation~\cite{Mangiagli+18}.
-In the deep inspiral, harmonics beyond the dominant quadrupole give a negligible
-contribution to the signal, so we only include the $(2,2)$ mode for simplicity.
+In the early inspiral, harmonics beyond the dominant quadrupole give a negligible contribution, so we only include the $(2,2)$ mode.
 
@@
 The resulting posterior distributions are shown in Fig.~\ref{fig:sbhb_corner_1}
-and \ref{fig:sbhb_corner_12}. We find close agreement between the \texttt{multinest} and
-\texttt{ptmcmc} samplers.
+and \ref{fig:sbhb_corner_12}.
+Similarly to the MBHB systems, we find close agreement between the \texttt{multinest} and \texttt{ptmcmc} samplers.
 The joint distributions for extrinsic parameters are fairly simple, with extended degeneracies occuring only in the angular parameters $(\varphi, \psi)$.
-Thanks to the complexity of the instrument transfer functions shown in Fig.~\ref{fig:transferSBHBcaseI} carrying a lot of information, the sky position determination is very good.
+Thanks to the complexity of the instrument transfer functions shown in Fig.~\ref{fig:transferSBHBcaseI} carrying a lot of information, the sky localization is very good.
 The typical degeneracy between inclination and distance
 is evident, as there is no degeneracy breaking from higher harmonics.
-Even though the SNR values are much lower than for our SMBHB systems, the Fisher matrix estimates represent quite well the marginal
-distributions of most parameters.
-We note, however, some notable deviation from Gaussianity for the system II: the sky position is almost bimodal, with an approximate symmetry across the ecliptic plane (we recall here that the ecliptic plane is not the plane of the LISA instrument that was used to report SMBHB results), owing probably to the localization being close to this plane; and the distance and inclination posteriors show a long tail extending way beyond the Fisher estimate.
+Even though the SNR values are much lower than for our SMBHB systems, the Fisher matrix represents quite well the marginal distributions of most parameters.
+We note, however, some notable deviation from Gaussianity for system II: the sky position is almost bimodal, with an approximate symmetry across the ecliptic plane (we recall here that the ecliptic plane is not the plane of the LISA instrument that was used to report SMBHB results), owing probably to the localization being close to this plane; and the distance and inclination posteriors show a long tail extending way beyond the Fisher estimate.
 
@@
 Using these tools to simulate the parameter recovery of two examples of SMBHB signals, we produced multidimensional posterior distributions.
-While our posterior distributions for masses are not really representative in the absence of know degeneracies between the mass ratio and the spins, we encountered and investigated two different kinds of degeneracies in the extrinsic parameter space.
+Because our study did not consider spins and their well-known degeneracy with masses, our posterior distributions for masses cannot yet be taken as representative of the actual LISA capabilities.
+Nevertheless, we encountered and investigated two different kinds of degeneracies in the extrinsic parameter space.
 
 Considering first IMR signals, we found a \textit{two-fold} degeneracy for the sky position corresponding to the \textit{reflected} sky localization with respect to the plane of the LISA constellation (opposite sign of the LISA-frame latitude).
-We also found that, when limiting the signals to their dominant harmonic $h_{22}$, strong degeneracies in distance-inclination could occur with a posterior distribution in the sky appearing shifted from the injected value.
+We also found that, when limiting the signals to their dominant harmonic $h_{22}$, strong degeneracies in distance-inclination can occur, causing the posterior distribution in the sky to appear shifted from the injected value.
 We gave an analytic explanation of this degeneracy by eliminating the time and frequency dependencies in the response, reducing it to an explicitly degenerate form.
@@
 
-Secondly, considering pre-merger analyses to investigate how information accumulates with time, we found an \textit{eight-fold} degeneracy for the sky position, obtained by duplicating the original and the reflected sky positions with $\pi/2$, $\pi$, $-\pi/2$ shifts in the LISA-frame longitude.
+Secondly, considering pre-merger analyses to investigate how information accumulates with time, we found an \textit{eight-fold} degeneracy for the sky position, obtained by rotating the original and reflected sky positions by integer multiples of $\pi/2$ around the axis of the LISA constellation.
 For our example systems, the posterior only collapses on the two-fold degeneracy, noticed previously for IMR signals, minutes before merger.
-By decomposing the LISA instrument response in different effects, separating its time-dependency from its frequency-dependency, we could pinpoint the role of both effects.
+By decomposing the LISA instrument response into different effects, separating its time-dependency from its frequency-dependency, we could pinpoint the role of both effects.
 The time-dependency of the response, consequence of the motion of LISA, leaves a moderate imprint at low frequencies, penalizing moderately other modes in the sky but leaving the \textit{antipodal} sky position degenerate.
@@
 We also explored the recovery of extrinsic parameters of SBHB signals (fixing the masses and spin parameters as well as the merger time), using the same methodology as for SMBHB signals to arrive at a likelihood cost of $\sim 2 \mathrm{ms}$.
-Investigating two example systems, we found that the distribution of extrinsic parameters is mostly Gaussian, with for one case some bimodality in the sky position found for one source lying close to the ecliptic plane, and an extended tail in the distance posterior.
-
-Comparing the parameter estimation results obtained with our two samplers, $\texttt{ptmcmc}$ and $\texttt{multinest}$, we found in general a good agreement.
+Investigating two example systems, we found that the distribution of extrinsic parameters is mostly Gaussian, with a bimodality in the sky position found for one source lying close to the ecliptic plane, and an extended tail in the distance posterior.
+
+Comparing the parameter estimation results obtained with our two samplers, $\texttt{ptmcmc}$ and $\texttt{multinest}$, we found an overall good agreement.
 We noted however that \texttt{multinest} can fall short of exploring the full degeneracies in parameter space, in the degenerate case of a signal without higher harmonics; this limitation is further illustrated by the example of a fully degenerate response.
-We also found that using Fisher matrix estimates would not reproduce Bayesian results in general: degeneracies in the sky would not be covered by a Fisher computation that is intrinsically unimodal, and neither would the specific features of the $22$-mode degenerate posteriors.
+We also found that using Fisher matrix estimates does not reproduce Bayesian results in general: the complicated degeneracies in the sky cannot be covered by a Fisher computation that is intrinsically unimodal, and neither can the specific features of the $22$-mode degenerate posteriors.
 That degeneracies occur even with the high SNRs of our SMBHB signals is remarkable, and reminds us that the usual statement of the validity of the Fisher matrix approximation in the high-SNR limit is very dependent on the morphology of the signal.
@@
 First, we only investigated a handful of sources.
-In particular, we chose our SMBHB systems to be representative of the bulk of the population; they are not ``golden sources'', i.e.
-they are not part of the tail of lower-redshift systems that would provide us with the most interesting science outputs with LISA, notably in terms of electromagnetic counterparts.
+In particular, we chose our SMBHB systems to be representative of the bulk of the expected population; they are not ``golden sources'', i.e.
+they are not part of the tail of lower-redshift systems that will deliver the most interesting science outputs with LISA, notably in terms of electromagnetic counterparts.
 It is likely that the parameter recovery here will be different, with a longer detectable signal and the motion of LISA playing a role in the premerger localization.
-We also limited ourselves to a single redshifted mass, while the morphology of the signals will change strongly between the two ends of the LISA spectrum, from intermediate mass black holes $M\sim 10^{3} M_{\odot}$ to massive systems $M>10^{7} M_{\odot}$.
-
-In the future, it will be important to incorporate more physics in our waveform models.
-In this study our signals were limited to non-spinning, circularized systems.
-The aligned components of the spin will play a role by allowing degeneracies between masses and spin magnitudes, but we can expect that they will not affect qualitatively the posterior distributions for extrinsic parameters; this will be investigated in a forthcoming publication~\cite{Katz+20}.
-Precession and eccentricity, however, will change the harmonic structures of the signals and play an important role.
-
-A limitation of our method is that, in order to accelerate the likelihood computation, set the noise realization to zero.
-Although we do not expect that the noise realization will cause drastic changes, with in particular the structure of degeneracies of signals in parameter space remaining relevant, fast likelihoods for noisy data would be a welcome addition to our methods.
+We also limited ourselves to a single redshifted mass, while the morphology of the signals will change strongly between the two ends of the LISA spectrum, from intermediate mass black holes ($M\sim 10^{3} M_{\odot}$) to massive systems ($M>10^{7} M_{\odot}$).
+
+Second, it will be necessary to incorporate more physics in our waveform models.
+In this study our signals were limited to non-spinning, quasicircular systems.
+The spin components along the orbital angular momentum will be degenerate with the masses, but we do not expect this to qualitatively change the posterior distributions for the extrinsic parameters.
+This will be investigated in a forthcoming publication~\cite{Katz+20}.
+Orbital precession and eccentricity, however, will change the harmonic structure of the signals and we expect they will play a much more important role.
+
+Third, we used zero-noise realizations in order to accelerate the likelihood computation.
+Although we do not expect that the noise realization will cause qualitative changes, with in particular the structure of degeneracies of signals in parameter space remaining relevant, fast likelihoods for pre-sampled noisy data will eventually be necessary.
 Analyzing signals with a noise injection will also become more important when considering more realistic instrumental noise models.
@@
 tailored jump proposals, effectively telling MCMC chains where to look for degenerate signals in parameter space.
-For instance, the LIGO-Virgo parameter estimation tools~\cite{Veitch+14} make use of an analytical knowledge of degeneracies.
+For instance, parameter estimation tools used in ground-based gravitational-wave astronomy make use of an analytical knowledge of degeneracies~\cite{Veitch+14}.
 

### 2020-02-21 paper/paper_PE.tex
Double parenthesis

@@
 Beyond these discrete symetries, since likelihood function dependence on the six extrinsic parameters is funneled through just two complex functions, we should expect a two-dimensional degenerate subspace.
-Indeed we can explicitly find a general solution for parameter values that solve $\sigma_{\pm}((D, t_\iota, t_\theta \varphi, \lambda_a, \psiL)=\sigma_{\pm,{\rm inj}}$, which we just sketch here, as needed to explain features of the degeneracies.
+Indeed we can explicitly find a general solution for parameter values that solve $\sigma_{\pm}(D, t_\iota, t_\theta \varphi, \lambda_a, \psiL)=\sigma_{\pm,{\rm inj}}$, which we just sketch here, as needed to explain features of the degeneracies.
 

### 2020-02-21 paper/paper_PE.tex
John's edits and some comments

--- 2020-02-21 paper/paper_PE.tex
@@
 The space-based gravitational wave detector LISA will observe mergers of supermassive black hole binary systems (SMBHBs) to cosmological distances, as well as inspiralling stellar-origin (or stellar-mass) binaries (SBHBs) years before they enter the LIGO/Virgo band.
-Much remains to be explored for the parameter recovery of both class of systems.
-Previous SMBHB analyses relied on inspiral-only signals and/or a simplified Fisher matrix analysis, while SBHBs have not yet been analyzed with Bayesian methods.
+Much remains to be explored for the parameter recovery of both classes of systems.
+Previous SMBHB analyses relied on inspiral-only signals and/or a simplified Fisher matrix analysis, while SBHBs have not yet been extensively analyzed with Bayesian methods.
 We accelerate likelihood computations by (i) using a Fourier-domain response of the LISA instrument, (ii) using a reduced-order model for non-spinning waveforms that include a merger-ringdown and higher harmonics, (iii) setting the noise realization to zero and computing overlaps in the amplitude/phase representation.
 We present the first simulations of Bayesian inference for the parameters of supermassive black hole systems including consistently the merger and ringdown of the signal, as well as higher harmonics.
-We find that restricting the merger-dominated signal to its dominant harmonic can make the extrinsic likelihood very degenerate.
+We clarify the roles of LISA response time and frequency dependencies in breaking degeneracies and illustrate how degeneracy breaking unfolds over time.
+We also find that restricting the merger-dominated signal to its dominant harmonic can make the extrinsic likelihood very degenerate.
 Including higher harmonics proves to be crucial to break degeneracies and considerably improves the localization of the source, with a surviving bimodality in the sky position.
@@
 
-In order to realize the full scientific objective of LISA with respect to black hole mergers, adequate data analysis tools must be prepared in advance.
+In order to realize the full scientific objectives of LISA with respect to black hole mergers, adequate data analysis tools must be prepared in advance.
 Similarly to what happens in LIGO and Virgo, one has to first identify the presence of a merger waveform in the LISA data, possibly in the presence of other superposed signals (notably galactic white dwarfs binaries~\cite{Nelemans+01}), and then infer the distribution of the physical parameters of its source from the data, enabling the construction of a catalog of black hole binaries.
@@
 
-The inference problem amounts to produce samples from the posterior distribution for source parameters, related by Baye's theorem to the likelihood function, i.e.~the probability of observing the measured data given the source parameters and a model of the source and detector.
+The inference problem amounts to producing samples from the posterior distribution for source parameters, related by Bayes theorem to the likelihood function, i.e.~the probability of observing the measured data given the source parameters and a model of the source and detector.
 In LISA, this problem is more complex than in kilometer-scale ground-based interferometers, as additional challenges arises from the much larger expected signal-to-noise ratios (SNRs) for high-mass systems, or the much longer duration of the waveform for low-mass systems.
@@
 
-The polarization angle $\psi$ can be seen as detailed in App~\ref{app:conventions} as a degree of freedom in the relation between the source frame and the detector frame, parametrizing a rotation around the wave vector $k$.
+The polarization angle $\psi$ can be seen (see App~\ref{app:conventions}) as a degree of freedom in the relation between the source frame and the detector frame, parametrizing a rotation around the wave vector $k$.
 To define this angle, one introduces reference vectors $(u,v)$ orthogonal to $k$, that define the zero of the polarization angle as $(p,q)(\psi=0) = (u,v)$.
@@
 
-In the frequency domain, modes will be represented with their amplitude and phase as\footnote{Note our Fourier convention~\eqref{eq:defFourier}.}
+Our response formalism applies to waveforms which can be represented in component modes with
+slowly varying amplitude and phase as\footnote{Note our Fourier convention~\eqref{eq:defFourier}.}
 \be\label{eq:hlmampphase}
@@
 \end{align}
-with the other Michelson observables $Y$, $Z$ being obtained by cyclic permutation. Uncorrelated combinations $A$, $E$ and $T$~\cite{Prince+02} are then defined as
+with the other Michelson observables $Y$, $Z$ being obtained by cyclic permutation. Uncorrelated combinations $A$, $E$ and $T$~\cite{Prince+02} are then experessed as
 \bsub\label{eq:defAET}
@@
 \esub
-where we have eliminated frequency-dependent prefactors that are common to the signal and to the noise by introducing the rescalings
+where we have eliminated frequency-dependent prefactors that are common to the signal and to the noise by introducing the rescalings \JGB{Would be neater to use $z$ here or put exponential in numerator.}
 \bsub\label{eq:scalingAET}
@@
 
-The prefactors~\eqref{eq:scalingAET} are oscillatory and have zero-crossings at high frequencies, with the first one occuring at $f_{L}/2 = 0.06 \mathrm{Hz}$, which is why it is convenient to eliminate them to avoid $0/0$ numerical instabilities.
-However, scaling these factors out from both the signal and the noise is an approximation, as we can expect that realistic delays would lead to imperfect cancellations; in the vicinity of the zero-crossings, we should therefore expect a localized loss of sensitivity.
+The prefactors~\eqref{eq:scalingAET} are oscillatory and have zero-crossings at high frequencies, with the first one occuring at $f_{L}/2 = 0.06 \mathrm{Hz}$, which is why it is convenient to factor them out to avoid $0/0$ numerical instabilities.
+This treatment would not apply directly with a more realitic model for LISA with varying arm-lengths and residual laser noise.
+In that case imperfect cancellations in the vicinity of the zero-crossings would likely result in localized loss of sensitivity.
+\JGB{Note this as discussed in call.}
 
@@
 \label{subsec:lowfresponse}
+
+Though our parameter estimation calculation primarily applies the full LISA response, it will be useful to consider some some simplifying asymptotic limits to understand parameter degeneracies.
 
@@
 
-It can be convenient to introduce the following notations for this delay phase:
+The essential effect is a delay phase factor. It can be convenient to introduce the following notations for this delay phase:
 \be\label{eq:defPhiR}
@@
 The parameter space interpolation in this ROM is therefore significantly simpler than the models~\cite{Puerrer14, Puerrer15, Bohe+16}, which was three-dimensional as it included aligned spins.
-The inclusion of higher harmonics, however, imposes to model the relative dephasing of different modes.
+The inclusion of higher harmonics, however, mandates modeling the relative dephasing of different modes.
 The output of the code consists of a Fourier domain amplitude and phase for each mode, sparsely sampled on $\sim 300$ frequencies.
@@
 The difficulty of computing numerically such an overlap depends on the phase difference.
-A large phase difference causes the integrand to be very oscillatory, requiring more frequency resolution than non-oscillatory integrands; on the other hand, oscillatory integrands give raise to a small integral due to cancellation effects, while non-oscillatory integrands contribute the most.
+A large phase difference causes the integrand to be very oscillatory, requiring more frequency resolution than non-oscillatory integrands; on the other hand, oscillatory integrands give rise to a small integral due to cancellation effects, while non-oscillatory integrands contribute the most.
 For a single mode, $\mathrm{SNR}$ terms of the type $(h|h)$ have a zero phase difference by construction, but including different modes generates cross-terms $(\ell, m) \neq (\ell', m')$ with a large phase difference.
@@
 Effective sample sizes (ESS) were estimated from the number of post-burn-in samples, reduced by the autocorrelation length, with the burn-in size chosen to maximize the ESS.
-In most cases, the runs were continued until achieving an estimated ESS~$>2000$.
+In most cases, the runs were continued until achieving ESS~$>2000$.
 
@@
 The inclination plays an important role in deciding the importance of higher harmonics; we set it to the median value $\iota = \pi/3$.
-We will study in detail two such systems, named cases I and II, stemming from a series of twelve runs with randomized orientation parameters, and chosen to exemplify qualitative differences in the parameter estimation.
+We will study two such systems in detail, named System I and System II, stemming from a series of twelve runs with randomized orientation parameters, and chosen to exemplify qualitative differences in the parameter estimation.
 The parameters are summarized in Table~\ref{tab:MBHBparams}, along with the SNR of these signals with and without higher harmonics.
@@
 The grey curve shows the characteristic noise PSD~\eqref{eq:defhc}.
-Vertical bars indicate for each mode the frequencies corresponding to the time-to-merger cuts of our pre-merger analysis as described in~\ref{subsec:MBHBacctime}, while the dots mark the merger frequency.}
+Vertical bars indicate for each mode the frequencies corresponding to the time-to-merger cuts marking the points where (1/64,1/16,1/4) of the final SNR has accumulated as in our pre-merger analysis as described in~\ref{subsec:MBHBacctime}, while the dots mark the merger frequency.\JGB{Which System is this?}}
   \label{fig:hctdiahm}
@@
 In all panels the black line represents the total.
-The horizontal axis gives the time to merger, with vertical lines for the cuts of~\ref{subsec:MBHBacctime}, and the dots on the right of each panel show the result for the full signal including the post-merger.}
+The horizontal axis gives the time to merger, with vertical lines for the cuts of~\ref{subsec:MBHBacctime}, and the dots on the right of each panel show the result for the full signal including the post-merger.\JGB{Note absolute value.
+Which system is this?}}
   \label{fig:modeoverlapsoftime}
@@
 We choose to show contributions to $\SNR^{2}$ because this is the scale on which the contributions of different modes and channels are additive, and because it is the relevant scale for the log-likelihood~\eqref{eq:deflnL}, since $\ln \calL \sim \SNR^{2}$.
-We also sum over the three channels ($T$ being essentially negligible).
+We also sum over the three channels.
 Together with the total contribution to $\SNR^{2}$, the three panels of Fig.~\ref{fig:modeoverlapsoftime} show diagonal terms $(s_{\ell m} | s_{\ell m})$, cross-terms involving the dominant quadrupolar mode $(s_{22} | s_{\ell m})$, and finally other cross-terms between subdominant modes.
@@
 We can draw several conclusions from Fig.~\ref{fig:modeoverlapsoftime}.
-The signal reaches detectability at $\SNR=10$ about two days before merger.
+The signal reaches a roughly detectable level of $\SNR=10$ about two days before merger.
 We see that the SNR accumulates rapidly in the last instants before merger, as shown in particular by reaching $\SNR/4$ only 7 minutes before coalescence.
-Stopping the signal at a given time in the inspiral changes somewhat the hierarchy between subdominant modes: contrasting Fig.~\ref{fig:modeoverlapsoftime} with Fig.~\ref{fig:hctdiahm}, we see e.g that $h_{21}$ is now subdominant before merger (and increases significantly in strength when including the post-merger signal).
+For a period following first detection most or all of the higher modes are not significant.
+Stopping the signal at a given time in the inspiral somewhat alters the hierarchy between subdominant modes that was seen in their power spectra: contrasting Fig.~\ref{fig:modeoverlapsoftime} with Fig.~\ref{fig:hctdiahm}, we see e.g that $h_{21}$ is now subdominant before merger (and increases significantly in strength when including the post-merger signal).
 This is because modes $h_{\ell m}$ with a higher $m$ will reach a higher frequency at any given time, while the weighting by the noise favors higher frequencies.
@@
 
-We also see that, as even the most subdominant modes contribute significantly to $\SNR^{2}$, our set $(22, 21, 33, 44, 55)$ is incomplete; we will need waveforms with a richer set of higher harmonics to analyze real LISA data.
+We also see that, since even the most subdominant modes contribute significantly to $\SNR^{2}$, our set $(22, 21, 33, 44, 55)$ is seems to be incomplete; we will need waveforms with a richer set of higher harmonics to analyze real LISA data.
 Diagonal terms $(s_{\ell m} | s_{\ell m})$ and cross terms with $(\ell m) \neq (\ell' m')$ are qualitatively different.
@@
 
-The time-dependency in the response mirrors the motion of the LISA constellation.
+The time-dependency in the response follows the motion of the LISA constellation.
 In the low-frequency picture of~Sec.~\ref{subsec:lowfresponse}, the time dependency enters both in the Doppler phase term~\eqref{eq:defPhiR} (LISA moves in the wavefront) and in the time-dependent LISA-frame angles~\eqref{eq:Lframeangles} (the LISA arms change orientation).
@@
 This limits the effect of the LISA motion: between the time where the signal becomes detectable and the merger, LISA has barely changed in orientation and position.
-Moreover, the region where the signal accumulates most of its SNR is even shorter, as shown in Fig.~\ref{fig:modeoverlapsoftime}.
+Moreover, the period when the signal accumulates most of its SNR is even shorter, as shown in Fig.~\ref{fig:modeoverlapsoftime}.
 
@@
 
-In the case (iv), \textit{Frozen low-$f$}, the response is equivalent to two LIGO-type detectors, lying motionless in the same location and rotated from each other by $\pi/4$, and there is no other information on the source's sky position other than the pattern functions of the two effective detectors.
+In the case (iv), \textit{Frozen low-$f$}, the response is equivalent to two LIGO-type detectors, lying motionless in the same location and rotated from each other by $\pi/4$, and there is no other information on the source's sky position other than the frequency-independent pattern functions of the two effective detectors.
 Contrarily to networks of ground-based observatories, we have no triangulation information from times of arrival at different detectors.
 This case is the most degenerate, and will be useful to get an analytical understanding of approximate degeneracies occuring when using the more complicated full response (i).
+This limit can also be a representative approximation for some short-duration premerger LISA MBHB observations.
 
@@
 Vertical lines also indicate the times to merger highlighted in Sec.~\ref{subsec:MBHBacctime}, as well as the peak frequency.
-Paralleling this figure to Figs.~\ref{fig:hctdiahm} and~\ref{fig:modeoverlapsoftime}, we stress that most of the SNR is accumulated in the very last instants before merger and at the merger itself, due to the noise normalization not visible at the level of the transfer function.
+Relating this figure to Figs.~\ref{fig:hctdiahm} and~\ref{fig:modeoverlapsoftime}, we stress that most of the SNR is accumulated in the very last instants before merger and at the merger itself, due to the noise normalization not visible at the level of the transfer function.
 
@@
 To illustrate the role of higher harmonics in the analysis, we will present two classes of results: ``22'', where we inject and recover signals including only the dominant harmonic $h_{22}$, and ``HM'', where we both inject and recover with the higher harmonics $(h_{22}, h_{21}, h_{33}, h_{44}, h_{55})$.
-We do not attempt to recover an injection with higher modes using 22-mode waveforms; one expects parameter biases in this case, induced by the inadequacy of the simplified waveforms.
+We do not attempt to recover an injection with higher modes using 22-mode waveforms; one would expect parameter biases in this case, induced by the inadequacy of the simplified waveforms.
 A related question would be to assess the mode content necessary for waveform models to mitigate systematic biases when analyzing full GR signals.
 We leave such waveform systematics studies for future work; we only note that, as shown in Fig.~\ref{fig:modeoverlapsoftime}, from SNR only the set of five modes we include in the present study is quite obviously incomplete.
+
+While the 22-mode-only signals here are non-physical, studying them is instructive not only because  22-only waveforms have been a common approximation in previous work but also because (as seen in Fig.~\ref{fig:modeoverlapsoftime}) upon first detection LISA MBHB signals will often be fairly approximated by 22-mode-only signals.
 
@@
 Looking at correlations, we see that the shifted peak corresponds to the region of high distance (and extremal inclination).
-This is a genuine feature of the multidimensional zero-noise posterior and will be explained in Sec.~\ref{subsec:MBHBPEdegen22}.
-
-In both Cases I and II, including the higher harmonics has a major effect on the parameter recovery (as was already stressed in~\cite{Arun+07a, TriasSintes07, PorterCornish08, McWilliams+09}), much beyond what we would expect from the sole gain in total SNR shown in Table.~\ref{tab:MBHBparams} (and a scaling of statistical errors $\sim 1/\SNR$).
+This is a genuine feature of the multidimensional posterior, even without noise,  and will be explained in Sec.~\ref{subsec:MBHBPEdegen22}.
+
+In both Cases I and II, including the higher harmonics has a major effect on the parameter recovery (as was already stressed in~\cite{Arun+07a, TriasSintes07, PorterCornish08, McWilliams+09}), much beyond what we would expect from the solely from the modest gain in total SNR shown in Table.~\ref{tab:MBHBparams} (and a scaling of statistical errors $\sim 1/\SNR$).
 The marginalized posterior of the masses is narrower, although not qualitatively different; the major change is in the extrinsic parameters.
@@
 
-We can gain insight about degeneracies in the parameter space by considering the simple expressions representing the instrument response in the \textit{Frozen low-$f$} approximation described in Sec.~\ref{subsec:MBHBresponseapprox}, appropriate for the low-frequency limit and signals short enough that the LISA motion can be neglected; we will explore in details the respective influence of these neglected effects in Sec.~\ref{sec:MBHBPEacctime}.
+We can gain insight about degeneracies in the parameter space by considering the simple expressions representing the instrument response in the \textit{Frozen low-$f$} approximation described in Sec.~\ref{subsec:MBHBresponseapprox}, appropriate for the low-frequency limit and signals short enough that the LISA motion can be neglected; we will explore the respective influence of these sometimes neglected effects in Sec.~\ref{sec:MBHBPEacctime}.
 
@@
 \end{align}
-How does this situation change when considering a more complete instrument response, moving away from the \textit{Frozen low-$f$} approximation ?
+How does this situation change when considering a more complete instrument response, moving away from the \textit{Frozen low-$f$} approximation ? As in Sec IV.C
+we separately consider relaxing each of the qualifiers ``Frozen'' and ``Low-f''.
 
@@
 
-When adding back the LISA motion while still ignoring the frequency-dependence in the response, in the \textit{Low-$f$} response, the \textit{antipodal} mode is the only one that keeps pattern functions that are exactly degenerate with the injection: all other modes will move together with the time-dependent LISA frame.
+When adding back the LISA motion while still ignoring the frequency-dependence in the response, in the \textit{Low-$f$} response, the \textit{antipodal} mode is the only one that keeps pattern functions that are exactly degenerate with the injection: all other modes will evolve with the time-dependent LISA frame.
 The antipodal mode is not moving, because the antipode of the true direction of the arriving signal  is defined as $k \rightarrow -k$ independently of the orientation of LISA; the only degeneracy-breaking term is then the orbital delay $k \cdot p_{0}$ in the Doppler phase~\eqref{eq:defPhiR}.
-With the \textit{Full} response, by contrast, frequency dependency terms in~\eqref{eq:Gslr} featuring $k\cdot n_{l}$, $k \cdot p_{s,r}$ readily break this antipodal degeneracy.
+%With the \textit{Full} response, by contrast, frequency dependency terms in~\eqref{eq:Gslr} featuring $k\cdot n_{l}$, $k \cdot p_{s,r}$ readily break this antipodal degeneracy.
 
@@
   \caption{Sky position posterior for System II in Table~\ref{tab:MBHBparams} in presence of strong degeneracies, zooming close to the injected value and discarding the secondary mode in the sky.
-The left panel shows the complete likelihood, while the right panel shows the simplified likelihood~\eqref{eq:simplelikelihoodhm} in the \textit{Frozen, low-$f$} approximation.
-and fixed masses and time.
+The left panel shows the complete likelihood, while the right panel shows the simplified likelihood~\eqref{eq:simplelikelihoodhm} in the \textit{Frozen, low-$f$} approximation and with fixed masses and time.
 The results of \texttt{ptmcmc} (blue) and \texttt{multinest} (green) are superimposed.
@@
 We will show that we can understand this feature as a projection effect in the multidimensional degenerate posterior.
-Understanding the structure of these features could prove useful to inform Bayesian samplers, for instance by adapting jump proposals to speed up mixing of MCMC chains.
+Understanding the structure of these features will be useful to inform Bayesian samplers, for instance by building adapted jump proposals allowing a much faster mixing of MCMC chains than naive methods.
 
@@
 \begin{itemize}
-	\item pin the masses and the coalescence time to their injected values;
+	\item pin the masses and the LISA-frame coalescence time to their injected values;
 	\item use the \textit{Frozen low-$f$} response, ignoring the LISA motion and the frequency-dependence in the response.
@@
 The second point is partly justified by the fact that our signal is short, as shown in Fig.~\ref{fig:modeoverlapsoftime}: for the MBHB system that we picked as an example the SNR accumulates in a matter of days, with $\mathrm{SNR}=10$ reached $\sim 40 \,\mathrm{h}$ before merger.
-Neglecting the high-frequency features is in fact a stronger approximation, as will be discussed in Sec.~\ref{subsec:MBHBacctimebreakdegen}.
-We focus on this limit merely to be able to work with the simple analytic expressions for the response given in~\ref{subsec:lowfresponse}.
+Neglecting the high-frequency features is in fact a stronger approximation, as will be explored in Sec.~\ref{subsec:MBHBacctimebreakdegen}.
+In this limit we can apply the simple analytic expressions for the response given in~\ref{subsec:lowfresponse}.
 
 Under all these simplifying assumptions, the likelihood becomes pure function of the extrinsic parameters $(D, \iota, \varphi, \lambdaL, \betaL, \psiL)$, and takes the form of trivial geometric factors multiplying constant mode overlaps that can be precomputed.
-The \textit{Low-$f$} approximation is detailed in Sec.~\ref{subsec:lowfresponse}.
-The $T$-channel is negligible in this approximation, so that the likelihood~\eqref{eq:deflnLaet} is
+Eq.~\ref{eq:aetlowf} shows that the $T$-channel is negligible in this approximation, so that the likelihood~\eqref{eq:deflnLaet} is
 \be
@@
 with $d = D / D_{\rm inj}$ the dimensionless ratio of luminosity distances, and with mode transfer functions $F_{a,e}^{\ell m}$ given in~\eqref{eq:Flmae}.
+In each term the intrinsic/extrinsic parameter dependence is thus separated with intrinsic parameters showing up only $\left\langle \ell m | \ell' m' \right\rangle$ which we will thereby neglect below.
 
@@
 \ee
-The functions~\eqref{eq:defsaelm} take for $(\ell m)=(22)$ the explicit form
+For $(\ell m)=(22)$ the functions~\eqref{eq:defsaelm} take the explicit form
 \begin{align}
@@
 
-In order to make more manifest the degeneracies in the problem, it will be useful to introduce the following notations.
-First, since $\iota \in [0, \pi]$ while $\beta_{L} \in [-\pi/2, \pi/2]$, it will be more convenient to work with $\theta_{L} = \pi/2 - \beta_{L}$.
-We will further abbreviate notations by using the variables
+In order to elucidate the degeneracies in the problem, it will be useful to introduce the following notation.
+First, since $\iota \in [0, \pi]$ while $\beta_{L} \in [-\pi/2, \pi/2]$, it will be more convenient to work with the colatitude $\theta_{L} = \pi/2 - \beta_{L}$.
+We will further abbreviate notation by using the variables
 \be
@@
 \ee
-In these notations, we obtain
+In this notation, we obtain
 \bsub\label{eq:sigmapm}
@@
 \ee
-Finding points in the parameter space that are degenerate with the injection, i.e.
-with $\ln \calL \simeq 0$, amounts to finding choices for the parameters which obtain the same values as the injection for both quantities $\sigma_{+}$ and $\sigma_{-}$.
+Finding points in the parameter space that are degenerate with the injection, {i.e.} with $\ln \calL \simeq 0$, amounts to finding choices for the parameters which obtain the same values as the injection for both quantities $\sigma_{+}$ and $\sigma_{-}$.
 
 We can now use the simplified response written in the form~\eqref{eq:sigmapm}-\eqref{eq:sigmafactorrho} to look for symmetries and degeneracies.
-First, it is easy to check in these new notations that the transformations~\eqref{eq:degenreflection} and~\eqref{eq:degenlambdapi2} indeed leave the likelihood exactly invariant.
+First, it is easy to check in this new notation that the transformations~\eqref{eq:degenreflection} and~\eqref{eq:degenlambdapi2} indeed leave the likelihood exactly invariant.
 There is also a symmetry based on exchanging $\iota \leftrightarrow \theta_{L}$.
 While leaving $\sigma_-$ unchanged, this conjugates the factor inside brackets for $\sigma_{+}$, which can be compensated using the phase terms $\varphi \pm \lambda_{a}$.
-If $\Phi = \mathrm{Arg} \left[ t_{\theta}^{4} e^{-2 i \psiL} + t_{\iota}^{4} e^{2 i \psiL} \right] $, we obtain the symmetry
+\JGB{Should this just be ``+'' not ``$\pm$''?} If $\Phi = \mathrm{Arg} \left[ t_{\theta}^{4} e^{-2 i \psiL} + t_{\iota}^{4} e^{2 i \psiL} \right] $, we obtain the symmetry
 \begin{align}
@@
 A large part of the degenerate subspace volume then tends to be found with a parameter $\beta_L$ near these special values fixed by the modulus constraint, with a special value of $\lambda_{L}$ coming from the complex argument constraint.
-By contrast, intermediate values of $t_{\iota}^{4}$ will not give as much parameter space allowed for the degeneracy, as illustrated by the case $t_{\iota} = 1$ ($\iota = \pi/2$): in that case $|r|=1$ irrespective of $\theta_{L}$ and $\psiL$, and no solution exists if $|r_{\rm inj}| \neq 1$.
+By contrast, intermediate values of $t_{\iota}^{4}$ will not give as much parameter space allowed for the degeneracy, as illustrated by the case $t_{\iota} = 1$ (i.e.
+$\iota = \pi/2$).
+In that case $|r|=1$ irrespective of $\theta_{L}$ and $\psiL$, and no solution exists if $|r_{\rm inj}| \neq 1$.
 
@@
 
-The rate at which parameter information accumulates during the observation of an inspiral is crucial for establishing the LISA downlink and data processing requirements, as well as for planning multimessenger observations~\cite{ArmitageNatarajan02, DalCanton+19}.
+The rate at which parameter information accumulates LISA merger observations is crucial for planning strategies for multi-messenger observations and in guiding the latency demands of the LISA down-link and data processing~\cite{ArmitageNatarajan02, DalCanton+19}.
 In this Section, we explore how parameter information accumulates on approach to merger by performing parameter estimation studies with temporally truncated signals, as explained in~\ref{subsec:MBHBacctime}: the signals are cut at points where the accumulated SNR is about $\{10,42,166\}$ corresponding to $\{1/64,1/16,1/4\}$ of the total SNR=$666.0$, and times $\{41 \mathrm{hr}, 2.5 \mathrm{hr}, 7 \mathrm{min}\}$ before merger.
-As shown in Fig.~\ref{fig:modeoverlapsoftime}, most of the SNR accumulates over the last few minutes of this signal.
+As shown in Fig.~\ref{fig:modeoverlapsoftime}, most of the SNR accumulates over the last few minutes of signal.
 
@@
 The results are shown for Case II, including the higher harmonics in the signal.
-Individual masses are loosely constrained, within a factor 2, at first detection; the chirp mass combination is better determined.
+Individual masses are loosely constrained, within a factor of 2, at first detection; the chirp mass combination is better determined.
 They start to be individually constrained to better than $10\%$ by 2.5 hours before merger.
 The luminosity distance is poorly determined, within a factor of 4, when reaching the detection threshold 41 hours before merger; the constraint improves to roughly $10\%$ at 2.5 hours prior to merger.
-Merger time is estimated with an uncertainty of roughtly 2hrs at first detection, which improves to a few minutes by 2.5 hours before merger.
+Merger time is estimated with an uncertainty of roughly 2hrs at first detection, which improves to a few minutes by 2.5 hours before merger.
 
@@
 The lines show the 1-2-3-$\sigma$ uncertainty contours on the sky, for each of our time cuts as well as for the full, post-merger signal, with the two panels differing by the inclusion of higher harmonics.
-We see clearly, in the pre-merger analysis, the 8-modes sky degeneracy introduced in~\eqref{eq:degenreflection}, \eqref{eq:degenlambdapi2}.
+We see clearly, in the pre-merger analysis, the 8-modes sky degeneracy introduced in~\eqref{eq:degenreflection}-\eqref{eq:degenlambdapi2}.
 The different cuts give us an idea of the continuous evolution from a badly determined sky position at first, when reaching detection, to an 8-modes degeneracy structure, to finally only two modes surviving when reaching merger, the true injection and the \textit{reflected} sky position.
@@
   \caption{Refinement of parameter inference over time for system II.
-The posteriors shown correspond to the time cuts defined in~\ref{subsec:MBHBacctime} at $\mathrm{SNR}/64$ (green), $\mathrm{SNR}/16$, $\mathrm{SNR}/4$ and post-merger (blue) for the masses (left), distance/inclination (center), and time/phase (right).
+The posteriors shown correspond to the time cuts defined in~\ref{subsec:MBHBacctime} at $\mathrm{SNR}/64$ (green), $\mathrm{SNR}/16$, $\mathrm{SNR}/4$ and post-merger (blue)., for the masses (left), distance/inclination (center), and time/phase (right).
 We inject and recover with higher harmonics.
@@
 The dashed black curve corresponds to the \textit{reflected} position, and the dashed red to the \textit{antipodal} position.
-At the true position (solid black), $\ln\calL$ is zero by construction while it is decreasing as a function of $f$ for other sky positions.
-Reaching a very negative $\ln\calL$ means that this sky position is excluded by the data accumulated thus far.}
+At the injection (solid black), $\ln\calL$ is zero by construction while it is decreasing as a function of $f$ for other sky positions.
+Reaching a very negative $\ln\calL$ means that this sky position is excluded by the signal accumulated thus far.
+The circles indicate the maximum log-likelihood when the parameters are allowed to vary.}
   \label{fig:lnLskymodesapprox}
@@
 
-The pre-merger analysis of the previous section made apparent the 8-modes degeneracy pattern that can be predicted analytically from the structure of the \textit{Frozen, low-$f$} response approximation (see~\eqref{eq:degenreflection}, \eqref{eq:degenlambdapi2}).
-However, the analysis of IMR signals in~\ref{subsec:SMBHPEfull} shows that only the \textit{reflected} mode~\eqref{eq:degenreflection} survives post-merger.
+We have found that pre-merger analysis of the previous section made apparent the 8-modes degeneracy pattern that can be predicted analytically from the structure of the \textit{Frozen, low-$f$} response approximation (see~\eqref{eq:degenreflection}, \eqref{eq:degenlambdapi2}).
+Analyses of IMR signals as in~\ref{subsec:SMBHPEfull} show that only the \textit{reflected} mode~\eqref{eq:degenreflection} survives post-merger.
 In this Section we explain how and why this transition occurs.
 
-We have already presented a discussion of different qualitative effects in the instrument response in Sec.~\ref{subsec:MBHBresponseapprox} and Fig.~\ref{fig:responseapproxtdia22}: on one hand, the motion of LISA leaves an imprint (the time-dependency) on the low frequencies, where the time-frequency map is steep enough that a short interval in frequency maps to a large interval of time; on the other hand, the breakdown of the long-wavelength approximation leaves an imprint (the frequency-dependency) at high frequencies.
+We have already presented a discussion of different qualitative effects in the instrument response in Sec.~\ref{subsec:MBHBresponseapprox} and Fig.~\ref{fig:responseapproxtdia22}: one hand hand, the motion of LISA leaves an imprint (the time-dependency) on the low frequencies, where the time-frequency map is steep enough that a short interval in frequency maps to a large interval of time; on the other hand, the breakdown of the long-wavelength approximation leaves an imprint (the frequency-dependency) at high frequencies.
 Here we look at the quantitative importance of these features on the inference as a function of frequency (or as a function of time).
@@
 
-A value of $\ln \calL = 0$ means that the template signal with shifted parameters is identical to the injection, while a very negative $\ln \calL$ means that these parameter values are ruled out.
+A value of $\ln \calL = 0$ means that the template signal with shifted parameters is identical to the injection, while a very negative $\ln \calL$ means that these parameter values are excluded in the inference.
 The injection (full black line) has always $\ln \calL = 0$ by construction.
@@
 All other modes see their $\ln \calL$ fall rapidly at around $2\mHz$ due to the onset of frequency-dependency in the response.
-Comparing to Fig.~\ref{fig:responseapproxtdia22}, we see that around $2\mHz$ the breaking of the long-wavelength approximation is still mild; however, the SNR start to accumulate a lot from $\SNR/16 \sim 40$ at $2.5\, \mathrm{h}$ to $\SNR/4 \sim 160$ at $7\min$, magnifying the effect of these features.
-Comparing the two rows of the figure, we see that the higher harmonics have an effect here, in making the elimination of secondary modes happen earlier in frequency (or time).
+Although the frequency dependency vanishes in the limit $f\ll f_L\approx 0.12$ Hz, note that the effect is crucial even for $f/f_L\approx 0.02$.
+Comparing to Fig.~\ref{fig:responseapproxtdia22}, we see that around $2\mHz$ the breaking of the long-wavelength approximation still appears mild; however, the SNR starts to accumulate a lot from $\SNR/16 \sim 40$ at $2.5\, \mathrm{h}$ to $\SNR/4 \sim 160$ at $7\min$, magnifying the effect of these features.
+Comparing the two rows of the figure, we see that the higher harmonics have an effect here, in making the elimination of secondary modes happen slightly earlier in frequency (or time).
 This is expected as they start to contribute significantly to the SNR (see Fig.\ref{fig:modeoverlapsoftime}) and reach higher in frequency than the $h_{22}$ harmonic according to~\eqref{eq:fscalinglm}.
@@
 
-This is indeed what happens in multidimensional Bayesian parameter estimation, and is illustrated in Fig.~\ref{fig:lnLskymodesapprox}: in the first panel, for the three pre-merger analyses as well as for the post-merger analysis, we overlay circles indicating the best $\ln \calL$ found among the posterior samples, in each eighth of the sky corresponding to the eight sky modes (when no samples are present in this region of the sky, nothing is displayed).
-A sampler is not an optimizer, so the precise value achieved is not optimal and would vary when repeating runs; nevertheless, this measure gives a good proxy for how much closer to the injected signal we could get by slightly biasing the parameters.
+This is indeed what happens in multidmensional Bayesian parameter estimation, and is illustrated in Fig.~\ref{fig:lnLskymodesapprox}: in the first panel, for the three pre-merger analyses as well as for the post-merger analysis, we overlay circles indicating the best $\ln \calL$ found among the posterior samples, in each eighth of the sky corresponding to the eight sky modes (when no samples are present in this region of the sky, nothing is displayed).
+A sampler is not an optimizer, so the precise value achieved is not optimal and would vary when repeating runs; nevertheless, this measure gives a good proxy for how much closer of the injected signal we could get by slightly biasing the parameters.
 
@@
 
-Stellar-mass black-hole binaries (SBHB) inspirals have been recently recognized as a potentially
+Stellar-mass black-hole binary (SBHB) inspirals have been recently recognized as a potentially
 important source of LISA detections \cite{Sesana16, Vitale16}. Full Bayesian parameter
@@
 The parameters of the simulated mergers were drawn from the Radler LISA Data Challenge (LDC), which itself followed the population of~\cite{Sesana16}, and are listed in Table~\ref{tab:SBHBparams}.
-System I is moderately massive, placed at a favorable close distance; with an SNR of $26.6$ it would be at the tail of the most favorable detections that we can hope for~\cite{Sesana16, Gerosa+19, Moore+19}.
-System II is a higher-mass, moderate-SNR system.
+The system I is a moderately massive system, placed at a favorable close distance; with an SNR of $26.6$ it would be at the tail of the most favorable detections that we can hope for~\cite{Sesana16, Gerosa+19, Moore+19}.
+The system II is a higher-mass, moderate-SNR system.
 Both systems happen to be close to the ecliptic plane.
 
-Fig.~\ref{fig:transferSBHBcaseI} illustrates the instrumental transfer functions~\eqref{eq:defTlmaet} as well as the Doppler phase~\eqref{eq:defPhiR} for system I.
+Fig.~\ref{fig:transferSBHBcaseI} illustrates, for the system I in this table, the instrumental transfer functions~\eqref{eq:defTlmaet} as well as the Doppler phase~\eqref{eq:defPhiR}.
 Contrasting with the SMBHB case, we see that the transfer functions have much more structure in this case.
-Both the time-dependency and the frequency-dependency in the response lead to oscillatory features in the response.
-The Doppler phase~\eqref{eq:defPhiR}, negligible for our short SMBHB systems, now varies by tens of radians.
+Both the time-dependency and the frequency-dependency in the response lead to oscillatory features in the transfer functions.
+The Doppler phase~\eqref{eq:defPhiR}, that was negligible for our short SMBHB systems, represents now tens of radians.
 
@@
 inspiral of SBHB systems, the signal is well within the domain of applicability of the post-Newtonian approximation~\cite{Mangiagli+18}.
-In the early inspiral, harmonics beyond the dominant quadrupole give a negligible contribution, so we only include the $(2,2)$ mode.
+In the deep inspiral, harmonics beyond the dominant quadrupole give a negligible
+contribution to the signal, so we only include the $(2,2)$ mode for simplicity.
 
@@
 The resulting posterior distributions are shown in Fig.~\ref{fig:sbhb_corner_1}
-and \ref{fig:sbhb_corner_12}.
-Similarly to the MBHB systems, we find close agreement between the \texttt{multinest} and \texttt{ptmcmc} samplers.
+and \ref{fig:sbhb_corner_12}. We find close agreement between the \texttt{multinest} and
+\texttt{ptmcmc} samplers.
 The joint distributions for extrinsic parameters are fairly simple, with extended degeneracies occuring only in the angular parameters $(\varphi, \psi)$.
-Thanks to the complexity of the instrument transfer functions shown in Fig.~\ref{fig:transferSBHBcaseI} carrying a lot of information, the sky localization is very good.
+Thanks to the complexity of the instrument transfer functions shown in Fig.~\ref{fig:transferSBHBcaseI} carrying a lot of information, the sky position determination is very good.
 The typical degeneracy between inclination and distance
 is evident, as there is no degeneracy breaking from higher harmonics.
-Even though the SNR values are much lower than for our SMBHB systems, the Fisher matrix represents quite well the marginal distributions of most parameters.
-We note, however, some notable deviation from Gaussianity for system II: the sky position is almost bimodal, with an approximate symmetry across the ecliptic plane (we recall here that the ecliptic plane is not the plane of the LISA instrument that was used to report SMBHB results), owing probably to the localization being close to this plane; and the distance and inclination posteriors show a long tail extending way beyond the Fisher estimate.
+Even though the SNR values are much lower than for our SMBHB systems, the Fisher matrix estimates represent quite well the marginal
+distributions of most parameters.
+We note, however, some notable deviation from Gaussianity for the system II: the sky position is almost bimodal, with an approximate symmetry across the ecliptic plane (we recall here that the ecliptic plane is not the plane of the LISA instrument that was used to report SMBHB results), owing probably to the localization being close to this plane; and the distance and inclination posteriors show a long tail extending way beyond the Fisher estimate.
 
@@
 Using these tools to simulate the parameter recovery of two examples of SMBHB signals, we produced multidimensional posterior distributions.
-Because our study did not consider spins and their well-known degeneracy with masses, our posterior distributions for masses cannot yet be taken as representative of the actual LISA capabilities.
-Nevertheless, we encountered and investigated two different kinds of degeneracies in the extrinsic parameter space.
+While our posterior distributions for masses are not really representative in the absence of known degeneracies between the mass ratio and the spins, we encountered and investigated two different kinds of degeneracies in the extrinsic parameter space.
 
 Considering first IMR signals, we found a \textit{two-fold} degeneracy for the sky position corresponding to the \textit{reflected} sky localization with respect to the plane of the LISA constellation (opposite sign of the LISA-frame latitude).
-We also found that, when limiting the signals to their dominant harmonic $h_{22}$, strong degeneracies in distance-inclination can occur, causing the posterior distribution in the sky to appear shifted from the injected value.
+We also found that, when limiting the signals to their dominant harmonic $h_{22}$, strong degeneracies in distance-inclination could occur with a posterior distribution in the sky appearing shifted from the injected value.
 We gave an analytic explanation of this degeneracy by eliminating the time and frequency dependencies in the response, reducing it to an explicitly degenerate form.
@@
 
-Secondly, considering pre-merger analyses to investigate how information accumulates with time, we found an \textit{eight-fold} degeneracy for the sky position, obtained by rotating the original and reflected sky positions by integer multiples of $\pi/2$ around the axis of the LISA constellation.
+Secondly, considering pre-merger analyses to investigate how information accumulates with time, we found an \textit{eight-fold} degeneracy for the sky position, obtained by duplicating the original and the reflected sky positions with $\pi/2$, $\pi$, $-\pi/2$ shifts in the LISA-frame longitude.
 For our example systems, the posterior only collapses on the two-fold degeneracy, noticed previously for IMR signals, minutes before merger.
-By decomposing the LISA instrument response into different effects, separating its time-dependency from its frequency-dependency, we could pinpoint the role of both effects.
-The time-dependency of the response, consequence of the motion of LISA, leaves a moderate imprint at low frequencies, penalizing moderately other modes in the sky but leaving the \textit{antipodal} sky position degenerate.
-By contrast, the frequency-dependency of the response, not included in previous parameter studies, is responsible for the degeneracy breaking and the collapse from eight to two modes in the sky.
-Note that the relative importance of the time and frequency dependency of the response will change for heavier or lighter systems.
+By decomposing the LISA instrument response in different effects, separating its time-dependency from its frequency-dependency, we could pinpoint the role of both effects.
+The time-dependency of the response, consequence of the motion of LISA, leaves a moderate imprint at low frequencies, moderately penalizing other modes in the sky but leaving the \textit{antipodal} sky position degenerate.
+By contrast, the frequency-dependency of the response, not included in many previous parameter studies, is responsible for the degeneracy breaking and the collapse from eight to two modes in the sky.
+Note that the relative importance of the time and frequency dependency of the response will change for heavier or lighter systems..
 
 We also explored the recovery of extrinsic parameters of SBHB signals (fixing the masses and spin parameters as well as the merger time), using the same methodology as for SMBHB signals to arrive at a likelihood cost of $\sim 2 \mathrm{ms}$.
-Investigating two example systems, we found that the distribution of extrinsic parameters is mostly Gaussian, with a bimodality in the sky position found for one source lying close to the ecliptic plane, and an extended tail in the distance posterior.
-
-Comparing the parameter estimation results obtained with our two samplers, $\texttt{ptmcmc}$ and $\texttt{multinest}$, we found an overall good agreement.
+Investigating two example systems, we found that the distribution of extrinsic parameters is mostly Gaussian, with for one case some bimodality in the sky position found for one source lying close to the ecliptic plane, and an extended tail in the distance posterior.
+
+Comparing the parameter estimation results obtained with our two samplers, $\texttt{ptmcmc}$ and $\texttt{multinest}$, we found in general a good agreement.
 We noted however that \texttt{multinest} can fall short of exploring the full degeneracies in parameter space, in the degenerate case of a signal without higher harmonics; this limitation is further illustrated by the example of a fully degenerate response.
-We also found that using Fisher matrix estimates does not reproduce Bayesian results in general: the complicated degeneracies in the sky cannot be covered by a Fisher computation that is intrinsically unimodal, and neither can the specific features of the $22$-mode degenerate posteriors.
+We also found that using Fisher matrix estimates would not reproduce Bayesian results in general: degeneracies in the sky would not be covered by a Fisher computation that is intrinsically unimodal, and neither would the specific features of the $22$-mode degenerate posteriors.
 That degeneracies occur even with the high SNRs of our SMBHB signals is remarkable, and reminds us that the usual statement of the validity of the Fisher matrix approximation in the high-SNR limit is very dependent on the morphology of the signal.
@@
 First, we only investigated a handful of sources.
-In particular, we chose our SMBHB systems to be representative of the bulk of the expected population; they are not ``golden sources'', i.e.
-they are not part of the tail of lower-redshift systems that will deliver the most interesting science outputs with LISA, notably in terms of electromagnetic counterparts.
+In particular, we chose our SMBHB systems to be representative of the bulk of the population; they are not ``golden sources'', i.e.
+they are not part of the tail of lower-redshift systems that would provide us with the most interesting science outputs with LISA, notably in terms of electromagnetic counterparts.
 It is likely that the parameter recovery here will be different, with a longer detectable signal and the motion of LISA playing a role in the premerger localization.
-We also limited ourselves to a single redshifted mass, while the morphology of the signals will change strongly between the two ends of the LISA spectrum, from intermediate mass black holes ($M\sim 10^{3} M_{\odot}$) to massive systems ($M>10^{7} M_{\odot}$).
-
-Second, it will be necessary to incorporate more physics in our waveform models.
-In this study our signals were limited to non-spinning, quasicircular systems.
-The spin components along the orbital angular momentum will be degenerate with the masses, but we do not expect this to qualitatively change the posterior distributions for the extrinsic parameters.
-This will be investigated in a forthcoming publication~\cite{Katz+20}.
-Orbital precession and eccentricity, however, will change the harmonic structure of the signals and we expect they will play a much more important role.
-
-Third, we used zero-noise realizations in order to accelerate the likelihood computation.
-Although we do not expect that the noise realization will cause qualitative changes, with in particular the structure of degeneracies of signals in parameter space remaining relevant, fast likelihoods for pre-sampled noisy data will eventually be necessary.
-Analyzing signals with a noise injection will also become more important when considering more realistic instrumental noise models.
+We also limited ourselves to a single redshifted mass, while the morphology of the signals will change strongly between the two ends of the LISA spectrum, from intermediate mass black holes $M\sim 10^{3} M_{\odot}$ to massive systems $M>10^{7} M_{\odot}$.
+
+In the future, it will be important to incorporate more physics in our waveform models.
+In this study our signals were limited to non-spinning, circularized systems.
+The aligned components of the spin will play a role by allowing degeneracies between masses and spin magnitudes, but we can expect that they will not affect qualitatively the posterior distributions for extrinsic parameters; this will be investigated in a forthcoming publication~\cite{Katz+20}.
+Precession and eccentricity, however, will change the harmonic structures of the signals and play an important role.
+
+A limitation of our method is that, in order to accelerate the likelihood computation, set the noise realization to zero.
+Although we do not expect that the noise realization will cause drastic changes, with in particular the structure of degeneracies of signals in parameter space remaining relevant, fast likelihoods for noisy data would be a welcome addition to our methods.
+Analyzing signals with a noise injection will also become more important when considering more realistic instrumental noise models and in prototyping algorithms for analysis of actual LISA data.
 
@@
 tailored jump proposals, effectively telling MCMC chains where to look for degenerate signals in parameter space.
-For instance, parameter estimation tools used in ground-based gravitational-wave astronomy make use of an analytical knowledge of degeneracies~\cite{Veitch+14}.
+For instance, the LIGO-Virgo parameter estimation tools~\cite{Veitch+14} make use of an analytical knowledge of degeneracies.
 

### 2020-02-28 paper/paper_PE.tex
Abstract

--- 2020-02-21 paper/paper_PE.tex
@@
 Previous SMBHB analyses relied on inspiral-only signals and/or a simplified Fisher matrix analysis, while SBHBs have not yet been analyzed with Bayesian methods.
-We accelerate likelihood computations by (i) using a Fourier-domain response of the LISA instrument, (ii) using a reduced-order model for non-spinning waveforms that include a merger-ringdown and higher harmonics, (iii) setting the noise realization to zero and computing overlaps in the amplitude/phase representation.
 We present the first simulations of Bayesian inference for the parameters of supermassive black hole systems including consistently the merger and ringdown of the signal, as well as higher harmonics.
+We use an instrumental response that includes finite-armlength effects at high frequencies, but we ignore the spin degrees of freedom and set the noise realization to zero.
 We find that restricting the merger-dominated signal to its dominant harmonic can make the extrinsic likelihood very degenerate.
 Including higher harmonics proves to be crucial to break degeneracies and considerably improves the localization of the source, with a surviving bimodality in the sky position.
+We explore the pre-merger parameter recovery by cutting the signals at different times before merger, and find a degenerate structure with eight modes in the sky reducing to two modes only shortly before merger.
+We analyze the respective roles of the time- and frequency-dependence of the LISA response in breaking degeneracies, and find that finite-armlength effects prove crucial in eliminating six of the eight modes in the sky, while the effect of the LISA motion is limited for our short-lived example signals.
 We also present simulations of Bayesian inference for the extrinsic parameters of SBHBs, and show that although unimodal, their posterior distributions can have non-Gaussian features.
@@
 We start by introducing a fast method to calculate the likelihood for a black hole merger.
-This method includes: a fast computation of a Fourier-domain inspiral-merger-ringdown waveform with higher-order modes based on reduced-order modeling; a fast Fourier-domain computation of the LISA response based on~\cite{MB18}; and a fast method for computing the noise-weighted product between waveforms when setting the noise realization to zero.
-We couple our likelihood with two codes for Bayesian inference based on different sampling techniques, constructing an inference engine which will allow us in the future to perform a variety of investigations with simulated black hole mergers.
+This method includes: a fast computation of a Fourier-domain inspiral-merger-ringdown waveform with higher-order modes based on a reduced order model; a fast Fourier-domain computation of the LISA response based on~\cite{MB18}; and a fast method for computing the noise-weighted product between waveforms when setting the noise realization to zero.
+We couple our likelihood with two codes for Bayesian inference based on different sampling techniques, constructing an inference engine which enables us to perform a variety of investigations with simulated black hole mergers.
 
@@
 \ee
-with $H$ the transverse-traceless matrix representing the gravitational wave, $L$ the delay along one arm\footnote{We recall that we use units with $G=c=1$.
-The proposed armlength for LISA is $L = 2.5\mathrm{Gm}$~\cite{LISA2017}.}, taken to be fixed, $k$ the propagation vector, $n_{l}$ the link unit vectors (from $s$ to $r$) and $p_{s}$, $p_{r}$ the positions of the spacecrafts.
+with $H$ the transverse-traceless matrix representing the gravitational wave, $k$ the wave propagation vector, $L$ the delay along one arm, taken to be fixed, $n_{l}$ the link unit vectors (from $s$ to $r$) and $p_{s}$, $p_{r}$ the positions of the spacecrafts.
 Here $n_{l}$, $p_{s}$ and $p_{r}$ are evaluated at the same time $t$.
 
-In the frequency domain, modes will be represented with their amplitude and phase as\footnote{Note our Fourier convention~\eqref{eq:defFourier}.}
+In the frequency domain, modes will be represented with their amplitude and phase as
 \be\label{eq:hlmampphase}
@@
 \esub
-Here, we have made apparent factors $(-2i\pi f)$ corresponding to the Fourier-domain translation of a time derivative.
-The observables $\tilde{a}$, $\tilde{e}$ are analogous to a time derivative of the strain more commonly used for ground-based detectors.
+Here, we have made apparent factors $(-2i\pi f)$, which correspond to a time derivative in the Fourier-domain.
+The observables $\tilde{a}$, $\tilde{e}$ are therefore analogous to time derivatives of the strain commonly used for ground-based detectors.
 
 We can now map these two channels to two fictitious LIGO-type detectors as follows.
-For an orthogonal detector of the same setup as the ground-based LIGO and Virgo with armlength $L_{D}$, rotated by an angle $\epsilon_{D}$ from the basis vectors $x,y$, the detector tensor is\S
+For an orthogonal detector of the same setup as the ground-based LIGO and Virgo with armlength $L_{D}$, rotated by an angle $\epsilon_{D}$ from the basis vectors $x,y$, the detector tensor is
 \begin{align}
@@
 Since the number of overlaps to compute in~\eqref{eq:defTlmaet} is quadratic in the number of modes, likelihoods with higher harmonics are significantly more expensive.
-The final likelihood cost depends also on the number of grid intervals, that is chosen to limit spline interpolation errors:
-\begin{itemize}
-	\item for SBHBs with $h_{22}$ only: $\sim 500$ grid points, likelihood cost $\sim 2\mathrm{ms}$;
-	\item for SMBHBs with $h_{22}$ only: $\sim 300$ grid points, likelihood cost $\sim 1.2\mathrm{ms}$;
-	\item for SMBHBs with 5 modes $h_{\ell m}$: $\sim 300$ grid points, likelihood cost $\sim 15\mathrm{ms}$.
-\end{itemize}
+The final likelihood cost depends also on the number of grid intervals, that is chosen to limit spline interpolation errors (Table~\ref{tab:likecost}).
+\begin{table}
+	\begin{tabular}{|c|c|c|c|}
+		\hline
+		Source type & Modes & Grid & Cost \\
+		\hline
+		SBHB & $(2,2)$ & $500$ & $2$ ms \\
+		SMBHB & $(2,2)$ & $300$ & $1.2$ ms \\
+		SMBHB & $(2,2),(2,1),(3,3),(4,4),(5,5)$ & $300$ & $15$ ms \\
+		\hline
+	\end{tabular}
+	\label{tab:likecost}
+	\caption{Approximate frequency grid resolution and likelihood cost for different choices of source type and number of modes in the waveform.}
+\end{table}
+%\begin{itemize}
+%	\item for SBHBs with $h_{22}$ only: $\sim 500$ grid points, likelihood cost $\sim 2\mathrm{ms}$;
+%	\item for SMBHBs with $h_{22}$ only: $\sim 300$ grid points, likelihood cost $\sim 1.2\mathrm{ms}$;
+%	\item for SMBHBs with 5 modes: $\sim 300$ grid points, likelihood cost $\sim 15\mathrm{ms}$.
+%\end{itemize}
 The accuracy of our likelihood computations is typically $|\Delta \ln \calL| \lesssim 0.2$, due to the numerical errors being magnified by the required cancellations between overlaps terms in~\eqref{eq:lnLoverlapstructure}.
@@
    \includegraphics[width=.99\linewidth]{../plots/tflm_smbh.pdf}
-  \caption{Time-to-frequency relation for MBHB systems I and II with $M=2\times 10^6 \; M_{\odot}$, $q=3$ and $z=4$.
+  \caption{Relation between time and frequency for MBHB systems I and II with $M=2\times 10^6 \; M_{\odot}$, $q=3$ and $z=4$.
 The different lines correspond to the $(\ell, m)$ harmonics available in the waveform model.
-The time-to-frequency is computed according to the defintion in $t_{f}^{\ell m}$ defined in~\eqref{eq:deftflm}, evaluated at $m/2 \times f$ to map to the same time according to~\eqref{eq:fscalinglm}.
+The time-to-frequency correspondence is computed according to $t_{f}^{\ell m}$ defined in~\eqref{eq:deftflm}, evaluated at $m f / 2$ to map to the same time according to~\eqref{eq:fscalinglm}.
 The vertical lines represent the time cuts explained in~\ref{subsec:MBHBacctime} and the merger frequency.}
@@
 In all panels the black line represents the total.
-The horizontal axis gives the time to merger, with vertical lines for the cuts of~\ref{subsec:MBHBacctime}, and the dots on the right of each panel show the result for the full signal including the post-merger.}
+The horizontal axis gives the time to merger, with vertical lines for the cuts of~\ref{subsec:MBHBacctime}, and the dots on the right of each panel show the result for the full signal including the ringdown.
+\SM{Make clear it is $|(h|h')|$ for cross terms -- changes sign}}
   \label{fig:modeoverlapsoftime}
@@
 
-We find that both our samplers \texttt{ptmcmc} and \texttt{multinest}, which as we recall were not really developed or tuned to this particular problem, spend a large number of likelihood evaluations, from a few $10^{7}$ to $10^{8}$, to produce a final set of posterior samples.
-This translates in reasonable computing costs thanks to our fast likelihoods, and is likely to be improvable in the future.
-
-To illustrate the role of higher harmonics in the analysis, we will present two classes of results: ``22'', where we inject and recover signals including only the dominant harmonic $h_{22}$, and ``HM'', where we both inject and recover with the higher harmonics $(h_{22}, h_{21}, h_{33}, h_{44}, h_{55})$.
-We do not attempt to recover an injection with higher modes using 22-mode waveforms; one expects parameter biases in this case, induced by the inadequacy of the simplified waveforms.
-A related question would be to assess the mode content necessary for waveform models to mitigate systematic biases when analyzing full GR signals.
-We leave such waveform systematics studies for future work; we only note that, as shown in Fig.~\ref{fig:modeoverlapsoftime}, from SNR only the set of five modes we include in the present study is quite obviously incomplete.
+We find that both our samplers, \texttt{ptmcmc} and \texttt{multinest}, require between $\sim 10^{7}$ and $\sim 10^{8}$ likelihood evaluations to produce a final set of posterior samples.
+Thanks to our fast likelihood implementation, this already represents a manageable computing cost.
+We stress again that the settings of both samplers were not optimized by taking into account the characteristics of the specific problem or the expected correlations between parameters.
+Therefore, we expect that the cost can be reduced with future optimizations.
+
+To illustrate the role of higher harmonics in the analysis, we will present two classes of results: ``22'', where we perform both the simulation and the inference by only including the dominant harmonic $h_{22}$, and ``HM'', where we perform both the simulation and the inference with all available harmonics $(h_{22}, h_{21}, h_{33}, h_{44}, h_{55})$.
+We do not consider the case where the simulation includes all harmonics and the inference is restricted to the 22 mode, as this will most likely bias the inferred parameters by an amount larger than the statistical uncertainties.
+In fact, the SNR calculation in Fig.~\ref{fig:modeoverlapsoftime} suggests that even the set of five modes included in this study may not be sufficient.
+The question of how many harmonics are necessary to keep the systematic errors below the statistical uncertainties is important and will be investigated in the future.
 
 Figs.~\ref{fig:PEsmbh22hmCase0} and~\ref{fig:PEsmbh22hmCase9} show posterior distributions for all parameters for Cases I and II respectively.
-We overlay ``22'' and ``HM'' posteriors.
+We overlay the ``22'' and ``HM'' posteriors.
 All results for MBHB systems are presented using LISA-frame parameters, as defined in App.~\ref{app:conventions}.
@@
 We will show that we can understand this feature as a projection effect in the multidimensional degenerate posterior.
-Understanding the structure of these features could prove useful to inform Bayesian samplers, for instance by adapting jump proposals to speed up mixing of MCMC chains.
+Understanding the structure of these features could prove useful to inform Bayesian samplers, for instance by building adapted jump proposal allowing a much faster mixing of MCMC chains than naive methods.
 
@@
 The dashed black curve corresponds to the \textit{reflected} position, and the dashed red to the \textit{antipodal} position.
-At the true position (solid black), $\ln\calL$ is zero by construction while it is decreasing as a function of $f$ for other sky positions.
-Reaching a very negative $\ln\calL$ means that this sky position is excluded by the data accumulated thus far.}
+At the injection (solid black), $\ln\calL$ is zero by construction while it is decreasing as a function of $f$ for other sky positions.
+Reaching a very negative $\ln\calL$ means that this sky position is excluded by the signal accumulated thus far.}
   \label{fig:lnLskymodesapprox}
@@
 
-With the \textit{Frozen} response, the \textit{reflected} mode remains exactly degenerate as anticipated in Sec.~\ref{subsec:MBHBPEdegen}.
+With the \textit{Low-$f$} response, the \textit{reflected} mode remains exactly degenerate as anticipated in Sec.~\ref{subsec:MBHBPEdegen}.
 All other modes see their $\ln \calL$ fall rapidly at around $2\mHz$ due to the onset of frequency-dependency in the response.
@@
 The parameters of the simulated mergers were drawn from the Radler LISA Data Challenge (LDC), which itself followed the population of~\cite{Sesana16}, and are listed in Table~\ref{tab:SBHBparams}.
-System I is moderately massive, placed at a favorable close distance; with an SNR of $26.6$ it would be at the tail of the most favorable detections that we can hope for~\cite{Sesana16, Gerosa+19, Moore+19}.
-System II is a higher-mass, moderate-SNR system.
+The system I is a moderately massive system, placed at a favorable close distance; with an SNR of $26.6$ it would be at the tail of the most favorable detections that we can hope for~\cite{Sesana16, Gerosa+19, Moore+19}.
+The system II is a higher-mass, moderate-SNR system.
 Both systems happen to be close to the ecliptic plane.
 
-Fig.~\ref{fig:transferSBHBcaseI} illustrates the instrumental transfer functions~\eqref{eq:defTlmaet} as well as the Doppler phase~\eqref{eq:defPhiR} for system I.
+Fig.~\ref{fig:transferSBHBcaseI} illustrates, for the system I in this table, the instrumental transfer functions~\eqref{eq:defTlmaet} as well as the Doppler phase~\eqref{eq:defPhiR}.
 Contrasting with the SMBHB case, we see that the transfer functions have much more structure in this case.
 Both the time-dependency and the frequency-dependency in the response lead to oscillatory features in the response.
-The Doppler phase~\eqref{eq:defPhiR}, negligible for our short SMBHB systems, now varies by tens of radians.
+The Doppler phase~\eqref{eq:defPhiR}, that was negligible for our short SMBHB systems, represents now tens of radians.
 
@@
 Using these tools to simulate the parameter recovery of two examples of SMBHB signals, we produced multidimensional posterior distributions.
-Because our study did not consider spins and their well-known degeneracy with masses, our posterior distributions for masses cannot yet be taken as representative of the actual LISA capabilities.
-Nevertheless, we encountered and investigated two different kinds of degeneracies in the extrinsic parameter space.
+While our posterior distributions for masses are not really representative in the absence of know degeneracies between the mass ratio and the spins, we encountered and investigated two different kinds of degeneracies in the extrinsic parameter space.
 
 Considering first IMR signals, we found a \textit{two-fold} degeneracy for the sky position corresponding to the \textit{reflected} sky localization with respect to the plane of the LISA constellation (opposite sign of the LISA-frame latitude).
-We also found that, when limiting the signals to their dominant harmonic $h_{22}$, strong degeneracies in distance-inclination can occur, causing the posterior distribution in the sky to appear shifted from the injected value.
+We also found that, when limiting the signals to their dominant harmonic $h_{22}$, strong degeneracies in distance-inclination could occur with a posterior distribution in the sky appearing shifted from the injected value.
 We gave an analytic explanation of this degeneracy by eliminating the time and frequency dependencies in the response, reducing it to an explicitly degenerate form.
@@
 
-Secondly, considering pre-merger analyses to investigate how information accumulates with time, we found an \textit{eight-fold} degeneracy for the sky position, obtained by rotating the original and reflected sky positions by integer multiples of $\pi/2$ around the axis of the LISA constellation.
+Secondly, considering pre-merger analyses to investigate how information accumulates with time, we found an \textit{eight-fold} degeneracy for the sky position, obtained by duplicating the original and the reflected sky positions with $\pi/2$, $\pi$, $-\pi/2$ shifts in the LISA-frame longitude.
 For our example systems, the posterior only collapses on the two-fold degeneracy, noticed previously for IMR signals, minutes before merger.
-By decomposing the LISA instrument response into different effects, separating its time-dependency from its frequency-dependency, we could pinpoint the role of both effects.
+By decomposing the LISA instrument response in different effects, separating its time-dependency from its frequency-dependency, we could pinpoint the role of both effects.
 The time-dependency of the response, consequence of the motion of LISA, leaves a moderate imprint at low frequencies, penalizing moderately other modes in the sky but leaving the \textit{antipodal} sky position degenerate.
@@
 We also explored the recovery of extrinsic parameters of SBHB signals (fixing the masses and spin parameters as well as the merger time), using the same methodology as for SMBHB signals to arrive at a likelihood cost of $\sim 2 \mathrm{ms}$.
-Investigating two example systems, we found that the distribution of extrinsic parameters is mostly Gaussian, with a bimodality in the sky position found for one source lying close to the ecliptic plane, and an extended tail in the distance posterior.
-
-Comparing the parameter estimation results obtained with our two samplers, $\texttt{ptmcmc}$ and $\texttt{multinest}$, we found an overall good agreement.
+Investigating two example systems, we found that the distribution of extrinsic parameters is mostly Gaussian, with for one case some bimodality in the sky position found for one source lying close to the ecliptic plane, and an extended tail in the distance posterior.
+
+Comparing the parameter estimation results obtained with our two samplers, $\texttt{ptmcmc}$ and $\texttt{multinest}$, we found in general a good agreement.
 We noted however that \texttt{multinest} can fall short of exploring the full degeneracies in parameter space, in the degenerate case of a signal without higher harmonics; this limitation is further illustrated by the example of a fully degenerate response.
-We also found that using Fisher matrix estimates does not reproduce Bayesian results in general: the complicated degeneracies in the sky cannot be covered by a Fisher computation that is intrinsically unimodal, and neither can the specific features of the $22$-mode degenerate posteriors.
+We also found that using Fisher matrix estimates would not reproduce Bayesian results in general: degeneracies in the sky would not be covered by a Fisher computation that is intrinsically unimodal, and neither would the specific features of the $22$-mode degenerate posteriors.
 That degeneracies occur even with the high SNRs of our SMBHB signals is remarkable, and reminds us that the usual statement of the validity of the Fisher matrix approximation in the high-SNR limit is very dependent on the morphology of the signal.
@@
 First, we only investigated a handful of sources.
-In particular, we chose our SMBHB systems to be representative of the bulk of the expected population; they are not ``golden sources'', i.e.
-they are not part of the tail of lower-redshift systems that will deliver the most interesting science outputs with LISA, notably in terms of electromagnetic counterparts.
+In particular, we chose our SMBHB systems to be representative of the bulk of the population; they are not ``golden sources'', i.e.
+they are not part of the tail of lower-redshift systems that would provide us with the most interesting science outputs with LISA, notably in terms of electromagnetic counterparts.
 It is likely that the parameter recovery here will be different, with a longer detectable signal and the motion of LISA playing a role in the premerger localization.
-We also limited ourselves to a single redshifted mass, while the morphology of the signals will change strongly between the two ends of the LISA spectrum, from intermediate mass black holes ($M\sim 10^{3} M_{\odot}$) to massive systems ($M>10^{7} M_{\odot}$).
-
-Second, it will be necessary to incorporate more physics in our waveform models.
-In this study our signals were limited to non-spinning, quasicircular systems.
-The spin components along the orbital angular momentum will be degenerate with the masses, but we do not expect this to qualitatively change the posterior distributions for the extrinsic parameters.
-This will be investigated in a forthcoming publication~\cite{Katz+20}.
-Orbital precession and eccentricity, however, will change the harmonic structure of the signals and we expect they will play a much more important role.
-
-Third, we used zero-noise realizations in order to accelerate the likelihood computation.
-Although we do not expect that the noise realization will cause qualitative changes, with in particular the structure of degeneracies of signals in parameter space remaining relevant, fast likelihoods for pre-sampled noisy data will eventually be necessary.
+We also limited ourselves to a single redshifted mass, while the morphology of the signals will change strongly between the two ends of the LISA spectrum, from intermediate mass black holes $M\sim 10^{3} M_{\odot}$ to massive systems $M>10^{7} M_{\odot}$.
+
+In the future, it will be important to incorporate more physics in our waveform models.
+In this study our signals were limited to non-spinning, circularized systems.
+The aligned components of the spin will play a role by allowing degeneracies between masses and spin magnitudes, but we can expect that they will not affect qualitatively the posterior distributions for extrinsic parameters; this will be investigated in a forthcoming publication~\cite{Katz+20}.
+Precession and eccentricity, however, will change the harmonic structures of the signals and play an important role.
+
+A limitation of our method is that, in order to accelerate the likelihood computation, set the noise realization to zero.
+Although we do not expect that the noise realization will cause drastic changes, with in particular the structure of degeneracies of signals in parameter space remaining relevant, fast likelihoods for noisy data would be a welcome addition to our methods.
 Analyzing signals with a noise injection will also become more important when considering more realistic instrumental noise models.

### 2020-02-28 paper/paper_PE.tex

@@
 Previous SMBHB analyses relied on inspiral-only signals and/or a simplified Fisher matrix analysis, while SBHBs have not yet been analyzed with Bayesian methods.
+We accelerate likelihood computations by (i) using a Fourier-domain response of the LISA instrument, (ii) using a reduced-order model for non-spinning waveforms that include a merger-ringdown and higher harmonics, (iii) setting the noise realization to zero and computing overlaps in the amplitude/phase representation.
 We present the first simulations of Bayesian inference for the parameters of supermassive black hole systems including consistently the merger and ringdown of the signal, as well as higher harmonics.
-We use an instrumental response that includes finite-armlength effects at high frequencies, but we ignore the spin degrees of freedom and set the noise realization to zero.
 We find that restricting the merger-dominated signal to its dominant harmonic can make the extrinsic likelihood very degenerate.
 Including higher harmonics proves to be crucial to break degeneracies and considerably improves the localization of the source, with a surviving bimodality in the sky position.
-We explore the pre-merger parameter recovery by cutting the signals at different times before merger, and find a degenerate structure with eight modes in the sky reducing to two modes only shortly before merger.
-We analyze the respective roles of the time- and frequency-dependence of the LISA response in breaking degeneracies, and find that finite-armlength effects prove crucial in eliminating six of the eight modes in the sky, while the effect of the LISA motion is limited for our short-lived example signals.
 We also present simulations of Bayesian inference for the extrinsic parameters of SBHBs, and show that although unimodal, their posterior distributions can have non-Gaussian features.
@@
 We start by introducing a fast method to calculate the likelihood for a black hole merger.
-This method includes: a fast computation of a Fourier-domain inspiral-merger-ringdown waveform with higher-order modes based on a reduced order model; a fast Fourier-domain computation of the LISA response based on~\cite{MB18}; and a fast method for computing the noise-weighted product between waveforms when setting the noise realization to zero.
+This method includes: a fast computation of a Fourier-domain inspiral-merger-ringdown waveform with higher-order modes based on ROM; a fast Fourier-domain computation of the LISA response based on~\cite{MB18}; and a fast method for computing the noise-weighted product between waveforms when setting the noise realization to zero.
 We couple our likelihood with two codes for Bayesian inference based on different sampling techniques, constructing an inference engine which enables us to perform a variety of investigations with simulated black hole mergers.
@@
 \ee
-with $H$ the transverse-traceless matrix representing the gravitational wave, $k$ the wave propagation vector, $L$ the delay along one arm, taken to be fixed, $n_{l}$ the link unit vectors (from $s$ to $r$) and $p_{s}$, $p_{r}$ the positions of the spacecrafts.
+with $H$ the transverse-traceless matrix representing the gravitational wave, $L$ the delay along one arm, taken to be fixed, $n_{l}$ the link unit vectors (from $s$ to $r$) and $p_{s}$, $p_{r}$ the positions of the spacecrafts.
 Here $n_{l}$, $p_{s}$ and $p_{r}$ are evaluated at the same time $t$.
@@
 However, scaling these factors out from both the signal and the noise is an approximation, as we can expect that realistic delays would lead to imperfect cancellations; in the vicinity of the zero-crossings, we should therefore expect a localized loss of sensitivity.
-\SM{make clear approximation already baked in}
 
@@
 \esub
-Here, we have made apparent factors $(-2i\pi f)$, which correspond to a time derivative in the Fourier-domain.
+Here, we have made apparent factors $(-2i\pi f)$, which correspond to Fourier-domain time derivatives.
 The observables $\tilde{a}$, $\tilde{e}$ are therefore analogous to time derivatives of the strain commonly used for ground-based detectors.
@@
 \esub
-These functions are the familiar pattern functions of ground-based detectors (for $\psi_{L}=0$) \SM{[add citation]}.
+These functions are the familiar pattern functions of ground-based detectors (for $\psi_{L}=0$).
 One can check that the expressions for the channel $e$ are obtained from the expressions for $a$ with the replacement $\lambdaL \rightarrow \lambdaL + \pi/4$.
@@
 
-Introducing the standard matched-filter inner \SM{[add reference]} product (also called overlap) as
+Introducing the standard matched-filter inner product (also called overlap) as
 \be\label{eq:definnerproduct}
@@
 		\hline
-		Source type & Modes & Grid & Cost \\
-		\hline
-		SBHB & $(2,2)$ & $500$ & $2$ ms \\
-		SMBHB & $(2,2)$ & $300$ & $1.2$ ms \\
-		SMBHB & $(2,2),(2,1),(3,3),(4,4),(5,5)$ & $300$ & $15$ ms \\
+		Source type & Modes & Grid points & Cost \\
+		\hline
+		SBHB & $h_{22}$ & $500$ & $2$ ms \\
+		SMBHB & $h_{22}$ & $300$ & $1.2$ ms \\
+		SMBHB & $5$ & $300$ & $15$ ms \\
 		\hline
@@
    \includegraphics[width=.99\linewidth]{../plots/tflm_smbh.pdf}
-  \caption{Relation between time and frequency for MBHB systems I and II with $M=2\times 10^6 \; M_{\odot}$, $q=3$ and $z=4$.
+  \caption{Relation between inspiral time and frequency for MBHB systems I and II with $M=2\times 10^6 \; M_{\odot}$, $q=3$ and $z=4$.
 The different lines correspond to the $(\ell, m)$ harmonics available in the waveform model.
@@
 We stress again that the settings of both samplers were not optimized by taking into account the characteristics of the specific problem or the expected correlations between parameters.
-Therefore, we expect that the cost can be reduced with future optimizations.
+Therefore, we expect that the cost can be reduced even further with future optimizations.
 
@@
 The resulting posterior distributions are shown in Fig.~\ref{fig:sbhb_corner_1}
-and \ref{fig:sbhb_corner_12}. We find close agreement between the \texttt{multinest} and
+and \ref{fig:sbhb_corner_12}. Similarly to the MBHB systems, we find close agreement between the \texttt{multinest} and
 \texttt{ptmcmc} samplers.

### 2020-02-28 paper/paper_PE.tex
Remove ulem
Clean up merge

--- 2020-02-28 paper/paper_PE.tex
@@
 The space-based gravitational wave detector LISA will observe mergers of supermassive black hole binary systems (SMBHBs) to cosmological distances, as well as inspiralling stellar-origin (or stellar-mass) binaries (SBHBs) years before they enter the LIGO/Virgo band.
-Much remains to be explored for the parameter recovery of both class of systems.
-Previous SMBHB analyses relied on inspiral-only signals and/or a simplified Fisher matrix analysis, while SBHBs have not yet been analyzed with Bayesian methods.
+Much remains to be explored for the parameter recovery of both classes of systems.
+Previous SMBHB analyses relied on inspiral-only signals and/or a simplified Fisher matrix analysis, while SBHBs have not yet been extensively analyzed with Bayesian methods.
+We accelerate likelihood computations by (i) using a Fourier-domain response of the LISA instrument, (ii) using a reduced-order model for non-spinning waveforms that include a merger-ringdown and higher harmonics, (iii) setting the noise realization to zero and computing overlaps in the amplitude/phase representation.
 We present the first simulations of Bayesian inference for the parameters of supermassive black hole systems including consistently the merger and ringdown of the signal, as well as higher harmonics.
-We use an instrumental response that includes finite-armlength effects at high frequencies, but we ignore the spin degrees of freedom and set the noise realization to zero.
-We find that restricting the merger-dominated signal to its dominant harmonic can make the extrinsic likelihood very degenerate.
+We clarify the roles of LISA response time and frequency dependencies in breaking degeneracies and illustrate how degeneracy breaking unfolds over time.
+We also find that restricting the merger-dominated signal to its dominant harmonic can make the extrinsic likelihood very degenerate.
 Including higher harmonics proves to be crucial to break degeneracies and considerably improves the localization of the source, with a surviving bimodality in the sky position.
-We explore the pre-merger parameter recovery by cutting the signals at different times before merger, and find a degenerate structure with eight modes in the sky reducing to two modes only shortly before merger.
-We analyze the respective roles of the time- and frequency-dependence of the LISA response in breaking degeneracies, and find that finite-armlength effects prove crucial in eliminating six of the eight modes in the sky, while the effect of the LISA motion is limited for our short-lived example signals.
 We also present simulations of Bayesian inference for the extrinsic parameters of SBHBs, and show that although unimodal, their posterior distributions can have non-Gaussian features.
@@
 Ground-based interferometers are fundamentally limited at low frequency by terrestrial noise, and cannot be used to study mergers of compact objects much heavier than a few $ 10^2 \Msol$.
-The spacebourne detector LISA~\cite{LISA2017} will overcome this limitation, enabling the observation and precise characterization of gravitational waves from binary black holes, from coalescences of systems with millions of $\Msol$ (supermassive black hole binaries, SMBHB) to the inspiral of systems with tens of $\Msol$ (stellar-mass black hole binaries, SBHB).
+The spacebourne detector LISA~\cite{LISA2017} will overcome this limitation, enabling the observation and precise characterization of gravitational waves from binary black holes, from coalescences of systems with millions of $\Msol$ (supermassive black hole binaries, SMBHB~\cite{Klein+15}) to the inspiral of systems with tens of $\Msol$ (stellar-mass black hole binaries, SBHB~\cite{Sesana16}).
 The observation of black hole mergers and inspirals in the LISA frequency band will yield major scientific rewards, and they form the main target of LISA among other classes of sources.
 
-In order to realize the full scientific objective of LISA with respect to black hole mergers, adequate data analysis tools must be prepared in advance.
+In order to realize the full scientific objectives of LISA with respect to black hole mergers, adequate data analysis tools must be prepared in advance.
 Similarly to what happens in LIGO and Virgo, one has to first identify the presence of a merger waveform in the LISA data, possibly in the presence of other superposed signals (notably galactic white dwarfs binaries~\cite{Nelemans+01}), and then infer the distribution of the physical parameters of its source from the data, enabling the construction of a catalog of black hole binaries.
@@
 Cosmological applications using LISA observations as standard sirens~\cite{Schutz86, Tamanini+16} also depend on the ability to localize individual sources.
-Parameter estimation of SBHBs will also be necessary for multiband gravitational-wave astronomy~\cite{Sesana16}.
-In this paper we focus on the latter part of the analysis problem, i.e.~the inference of the black hole parameters from the LISA data, leaving aside the detection problem and assuming the presence of the signal in the data.
+Understanding parameter estimation of SBHBs will also be important to understand the outcomes and challenges of possible multiband gravitational-wave observations~\cite{Sesana16}.
+In this paper we focus on the latter part of the analysis problem, i.e.~the inference of the black hole parameters from the LISA data, once we have reasons to believe the presence of the signal in the data, leaving aside the identification problem.
 We also limit our analysis to extrinsic parameters and ignore the effect of superposed gravitational wave signals from various sources.
 
-The inference problem amounts to drawing samples from the posterior distribution for source parameters, related by Bayes' theorem to the likelihood function, i.e.~the probability of observing the measured data given the source parameters and a model of the source and detector.
-In LISA, this problem is more complex than in kilometer-scale ground-based interferometers, as additional challenges arise from the much larger expected signal-to-noise ratios (SNRs) for high-mass systems, or the much longer duration of the waveform for low-mass systems.
+The inference problem amounts to producing samples from the posterior distribution for source parameters, related by Bayes theorem to the likelihood function, i.e.~the probability of observing the measured data given the source parameters and a model of the source and detector.
+In LISA, this problem is more complex than in kilometer-scale ground-based interferometers, as additional challenges arises from the much larger expected signal-to-noise ratios (SNRs) for high-mass systems, or the much longer duration of the waveform for low-mass systems.
 In particular, for SMBHBs, the large SNR will require a very accurate modeling of the waveform, including the merger and ringdown regimes, the effects of spin and corrections of higher harmonics in the signal beyond the quadrupole.
@@
 Parameter estimation studies for SBHB systems have relied so far on the Fisher matrix approach~\cite{Sesana16, Vitale16, Nishizawa+16a, Nishizawa+16b}, with Bayesian inference tools yet to be developed.
-Focusing on the posterior distribution of extrinsic parameters (fixing the masses and time to coalescence), we demonstrate that Bayesian analyses can be run with a comparable speed for the likelihood computation and discuss the features of the posteriors.
+Focusing on the posterior distribution of extrinsic parameters (fixing the masses and time to coalescence), we demonstrate that Bayesian analyses can be run with a speed for the likelihood computation that is comparable to the SMBHB case, and discuss the features of the posteriors.
 
@@
 Here $n_{l}$, $p_{s}$ and $p_{r}$ are evaluated at the same time $t$.
-
-In the frequency domain, modes will be represented with their amplitude and phase as
+In the following, we will use interchangeably the notation $y_{sr}$ instead of $y_{slr}$, since $n_{l}$ can be deduced from the sending and receiving indices $s,r$.
+
+Our response formalism applies to waveforms which can be represented as a combination of harmonics with
+slowly varying amplitude and phase as\footnote{Note our Fourier convention~\eqref{eq:defFourier}.}
 \be\label{eq:hlmampphase}
@@
 
-The analysis of~\cite{MB18} has shown that higher-order corrections are small in general for SMBHB systems, and are also small for SBHB systems provided they are not too far from the coalescence.
+The analysis of~\cite{MB18} has shown that higher-order corrections in the separation of timescales in the LISA response are small in general for SMBHB systems, and are also small for SBHB systems provided they are not too far from the coalescence.
 The fact that we use the same Fourier-domain treatment of the transfer functions for both the signal and the templates should also mitigate the importance of modelling errors.
@@
 
-We note an important feature of~\eqref{eq:Gslr}: frequency-dependent terms only feature the projections
-\be
-	k_{\parallel} \cdot n_{l}, \; k_{\parallel} \cdot p_{s,r} \,,
+In~\eqref{eq:Gslr} it is convenient to decompose the spacecraft positions as
+\be\label{eq:defp0}
+	p_{r,s} = p_{0} + p_{r,s}^{L} \,,
+\ee
+with $p_{0}$ the position of the center of the LISA constellation.
+This allows us to note an important feature of~\eqref{eq:Gslr}: apart from a global phase delay factor $\exp[i \pi f k\cdot p_{0}]$, frequency-dependent terms only feature the projections
+\be
+	k_{\parallel} \cdot n_{l}, \; k_{\parallel} \cdot p_{s,r}^{L} \,,
 \ee
 with $k_{\parallel}$ the projection of the wave vector $k$ in the (instantaneous) LISA plane, thus the frequency-dependent factors are invariant when we reflect $k$ across this plane.
+The delay $k\cdot p_{0}$ is the only one with a baseline outside the LISA plane (see~\eqref{eq:defPhiR} below).
 This will remain true when constructing TDI variables below\footnote{Strictly speaking it will not be true in the most general response formalism of~\cite{MB18}, where corrections are either slightly non-local in time or involve velocities, that are out of the LISA plane.}.
@@
 \end{align}
-with the other Michelson observables $Y$, $Z$ being obtained by cyclic permutation.
-The uncorrelated-noise combinations $A$, $E$ and $T$~\cite{Prince+02} are then defined as
+with the other Michelson observables $Y$, $Z$ being obtained by cyclic permutation. Uncorrelated combinations $A$, $E$ and $T$~\cite{Prince+02} are then expressed as
 \bsub\label{eq:defAET}
@@
 \begin{align}
-	\tilde{a}, \tilde{e} &= \frac{1}{i \sqrt{2} e^{2i\pi fL} \sin (2\pi f L)}\times \tilde{A}, \tilde{E} \,,\\
-	\tilde{t} &= \frac{1}{2\sqrt{2} e^{3i\pi fL} \sin (\pi f L) \sin (2\pi f L)} \times \tilde{T} \,.
+	\tilde{a}, \tilde{e} &= \frac{e^{-2i\pi fL}}{i \sqrt{2} \sin (2\pi f L)}\times \tilde{A}, \tilde{E} \,,\\
+	\tilde{t} &= \frac{e^{-3i\pi fL}}{2\sqrt{2}  \sin (\pi f L) \sin (2\pi f L)} \times \tilde{T} \,.
 \end{align}
@@
 with $S^{\rm pm}$ the test-mass noise PSD and $S^{\rm op}$ the optical noise PSD.
-Note that a confusion noise coming from the background of galactic binaries in the  LISA band is to be added to the instrumental noise.
+We also include a confusion noise coming from the background of galactic binaries in the LISA band that is added to the instrumental noise.
 For the instrument performance defining these noise levels, we take values from~\cite{LISANoiseBudget16} (see App.~\ref{app:LISAnoise}).
@@
 
-The prefactors~\eqref{eq:scalingAET} are oscillatory and have zero-crossings at high frequencies, with the first one occuring at $f_{L}/2 = 0.06 \mathrm{Hz}$, which is why it is convenient to eliminate them to avoid $0/0$ numerical instabilities.
-However, scaling these factors out from both the signal and the noise is an approximation, as we can expect that realistic delays would lead to imperfect cancellations; in the vicinity of the zero-crossings, we should therefore expect a localized loss of sensitivity.
-\SM{make clear approximation already baked in}
+The prefactors~\eqref{eq:scalingAET} are oscillatory and have zero-crossings at high frequencies, with the first one occuring at $f_{L}/2 = 0.06 \mathrm{Hz}$, which is why it is convenient to factor them out to avoid $0/0$ numerical instabilities.
+This treatment would not apply directly with a more realitic model for LISA with varying arm-lengths and residual laser noise.
+In that case imperfect cancellations in the vicinity of the zero-crossings would likely result in localized loss of sensitivity.
 
@@
 
-In the low-frequency limit, also called the long-wavelength approximation, the effects due the finite LISA armlength become negligible and the LISA response becomes equivalent to the response of two coplanar LIGO-like detectors rotated by $\pi/4$ relative to each other and set in motion~\cite{Cutler97}.
+Though our parameter estimation calculation primarily applies the full LISA response, it will be useful to consider some some simplifying asymptotic limits to understand parameter degeneracies.
+
+As is well known~\cite{Cutler97}, in the low-frequency limit (also called the long-wavelength approximation), the finite-armlength effects vanish, and the response of LISA is analogous to the response of of two LIGO-type detectors rotated from each other by $\pi/4$ and set in motion.
 
@@
 
-It can be convenient to introduce the following notations for this delay phase:
+The exponential factor is a delay phase, for which we introduce the notation:
 \be\label{eq:defPhiR}
@@
 \be\label{eq:aelowfmodes}
-	\tilde{a}, \tilde{e} = \sum_{\ell, m>0} (-2i\pi f) \exp\left[ 2 i \pi f k\cdot p_{0} \right] \tilde{h}_{\ell m} D_{a,e} : P_{\ell m} \,,
+	\tilde{a}, \tilde{e} = (-2i\pi f) \exp\left[ 2 i \pi f k\cdot p_{0} \right] \sum_{\ell, m>0} \tilde{h}_{\ell m} D_{a,e} : P_{\ell m} \,,
 \ee
@@
 \esub
-These functions are the familiar pattern functions of ground-based detectors (for $\psi_{L}=0$) \SM{[add citation]}.
+These functions are the familiar pattern functions of ground-based detectors (for $\psi_{L}=0$) (see e.g.~\cite{FlanaganHughes05}).
 One can check that the expressions for the channel $e$ are obtained from the expressions for $a$ with the replacement $\lambdaL \rightarrow \lambdaL + \pi/4$.
@@
 
-Introducing the standard matched-filter inner \SM{[add reference]} product (also called overlap) as
+Introducing the standard matched-filter inner product~\cite{FlanaganHughes05} (also called overlap) as
 \be\label{eq:definnerproduct}
@@
 with the ``$\rm inj$'' subscript indicating the simulated signals, and where the inner products $(\cdot | \cdot)$ are given by~\eqref{eq:definnerproduct} with the noise PSDs~\eqref{eq:Snaet}.
+\SM{check nothing was git-lost here}
 
@@
 The parameter space interpolation in this ROM is therefore significantly simpler than the models~\cite{Puerrer14, Puerrer15, Bohe+16}, which include aligned spins.
-The inclusion of higher harmonics, however, requires modeling the relative dephasing of different modes.
+The inclusion of higher harmonics, however, mandates modeling the relative dephasing of different modes.
 The output of the code consists of a Fourier domain amplitude and phase for each mode, sparsely sampled on $\sim 300$ frequencies.
-The faithfulness with the original waveforms, assuming Advanced LIGO noise curves~\cite{UnofficialNoiseCurves18}, is $\lesssim 10^{-4}$.
+The unfaithfulness with the original waveforms, assuming Advanced LIGO noise curves~\cite{UnofficialNoiseCurves18}, is $\lesssim 10^{-4}$.
 The computational cost is submillisecond and smaller than other stages of our likelihood computation.
@@
 Most runs here performed comparably with either 80 or 240 parallel temperature chains.
-We applied a general-purpose proposal distribution composed of a weighted set of sub-proposals, including differential-evolution steps~\cite{Braak06, BraakVrugt08} (although the proposal in this case is based on each single chain, not an ensemble) and a collection of several Gaussian step draws of varying sizes scaled off the prior domain at about 0.01--1\% of its scale and weighted favor the smaller scale proposal draws.
+We applied a general-purpose proposal distribution composed of a weighted set of sub-proposals, including differential-evolution steps~\cite{Braak06, BraakVrugt08} (although the proposal in this case is based on each single chain, not an ensemble) and a collection of several Gaussian step draws of varying sizes scaled off the prior domain at about 0.01--1\% of its scale and weighted to favor the smaller scale proposal draws.
 Effective sample sizes (ESS) were estimated from the number of post-burn-in samples, reduced by the autocorrelation length, with the burn-in size chosen to maximize the ESS.
-In most cases, the runs were continued until achieving an estimated ESS~$>2000$.
+In most cases, the runs were continued until achieving ESS~$>2000$.
 
@@
 
-We focus our analysis on a single choice of mass parameters, representative of canonical SMBHB sources expected for LISA~\cite{Barausse12, Klein+15, LISA2017}: we pick a total redshifted mass $M=m_{1}+m_{2} = 2\times10^{6} M_{\odot}$ and a mass ratio $q=m_{1}/m_{2} = 3$, and place the source at a redshift $z=4$.
+We focus our analysis on a single choice of mass parameters, representative of canonical SMBHB sources expected for LISA~\cite{Barausse12, Klein+15, LISA2017}: we pick a total redshifted mass $M=m_{1}+m_{2} = 2\times10^{6} M_{\odot}$ and a mass ratio $q=m_{1}/m_{2} = 3$, and place the source at a redshift $z=4$\footnote{We convert between luminosity distances and redshifts using the cosmological parameters of~\cite{Planck15param}}.
 The corresponding source-frame total mass is $M/(1+z) = 4\times10^{5} M_{\odot}$.
-The inclination plays an important role in deciding the importance of higher harmonics; we set it to the median value $\iota = \pi/3$.
-We will study in detail two such systems, named cases I and II, stemming from a series of twelve runs with randomized orientation parameters, and chosen to exemplify qualitative differences in the parameter estimation.
+The inclination plays an important role in deciding the importance of higher harmonics; we set it to the value $\iota = \pi/3$.
+We will study two such systems in detail, named System I and System II, stemming from a series of twelve runs with randomized orientation parameters, and chosen to exemplify qualitative differences in the parameter estimation.
 The parameters are summarized in Table~\ref{tab:MBHBparams}, along with the SNR of these signals with and without higher harmonics.
@@
 We note that, by contrast with current ground-based observations that are horizon-limited and that have a strong selection bias towards systems with favorable orientations (face-on or face-off), the LISA horizon for SMBHB systems extends to essentially the entirety of the observable universe~\cite{LISA2017}; it is therefore natural to consider randomized orientations.
+Here, although we focus on just two systems, we take $\iota = \pi/3$, the median value when inclination is randomized.
 
@@
   \includegraphics[width=.99\linewidth]{../plots/plot_hc_tdia_hm.pdf}
-  \caption{Characteristic strain signal~\eqref{eq:defhc}, shown here for the TDI channel $A$, decomposed as a set of $(\ell,m)$ contributions from the modes available in the waveform model.
+  \caption{Characteristic strain signal~\eqref{eq:defhc}, shown here for the TDI channel $A$ of System II, decomposed as a set of $(\ell,m)$ contributions from the modes available in the waveform model.
 The grey curve shows the characteristic noise PSD~\eqref{eq:defhc}.
-Vertical bars indicate for each mode the frequencies corresponding to the time-to-merger cuts of our pre-merger analysis as described in~\ref{subsec:MBHBacctime}, while the dots mark the merger frequency.}
+Vertical bars indicate for each mode the frequencies corresponding to the time-to-merger cuts marking the points where (1/64,1/16,1/4) of the final SNR has accumulated as in our pre-merger analysis as described in~\ref{subsec:MBHBacctime}, while the dots mark the merger frequency.}
   \label{fig:hctdiahm}
@@
 The first panel isolates the Doppler phase $\Delta\Phi_{R}$ as defined in~\eqref{eq:defPhiR}, while the three other panels show the TDI transfer functions for $A,E,T$.
-We show here the rescaled transfer functions for the mode 22 as defined below~\eqref{eq:defTlmaet}.}
+We show here the rescaled transfer functions for the mode 22 as defined below~\eqref{eq:defTlmaet}.
+The vertical lines correspond to the frequency at merger.}
   \label{fig:tranferSMBHCase9}
@@
 The time-to-frequency correspondence is computed according to $t_{f}^{\ell m}$ defined in~\eqref{eq:deftflm}, evaluated at $m f / 2$ to map to the same time according to~\eqref{eq:fscalinglm}.
-The vertical lines represent the time cuts explained in~\ref{subsec:MBHBacctime} and the merger frequency.}
+The vertical lines represent the time cuts explained in~\ref{subsec:MBHBacctime} and the merger frequency.
+\SM{fix 40h -> 42h}}
   \label{fig:t-f-relation-lm}
@@
   \includegraphics[width=.98\linewidth]{../plots/overlapmodebymode_maxfscaledlm_t.pdf}
-  \caption{Cumulative contribution to $\mathrm{SNR}^{2}$ as defined in~\eqref{eq:defcontribSNR2}.
+  \caption{Cumulative contribution to $\mathrm{SNR}^{2}$ as defined in~\eqref{eq:defcontribSNR2} for system II.
 The left panel shows inner products of modes with themselves, the middle panel shows cross-terms involving $h_{22}$, while the right panel shows other cross-terms.
+Note that oscillatory cross-terms change sign, we show their absolute value.
 In all panels the black line represents the total.
 The horizontal axis gives the time to merger, with vertical lines for the cuts of~\ref{subsec:MBHBacctime}, and the dots on the right of each panel show the result for the full signal including the ringdown.
-\SM{Make clear it is $|(h|h')|$ for cross terms -- changes sign}}
+\SM{also change title $(h|h)$, these are mode contributions to TDI not $h$}}
   \label{fig:modeoverlapsoftime}
@@
 
-We also see that, as even the most subdominant modes contribute significantly to $\SNR^{2}$, our set $(22, 21, 33, 44, 55)$ is incomplete; we will need waveforms with a richer set of higher harmonics to analyze real LISA data.
+We also see that, since even the most subdominant modes contribute significantly to $\SNR^{2}$, our set $(22, 21, 33, 44, 55)$ seems to be incomplete; we will need waveforms with a richer set of higher harmonics to analyze real LISA data.
 Diagonal terms $(s_{\ell m} | s_{\ell m})$ and cross terms with $(\ell m) \neq (\ell' m')$ are qualitatively different.
@@
 
-In the case (iv), \textit{Frozen low-$f$}, the response is equivalent to two coplanar LIGO-type detectors, lying motionless at the same location and rotated with respect to each other by $\pi/4$, and there is no other information on the source's sky position other than the pattern functions of the two effective detectors.
+In the case (iv), \textit{Frozen low-$f$}, the response is equivalent to two LIGO-type detectors, lying motionless in the same location and rotated from each other by $\pi/4$, and there is no other information on the source's sky position other than the frequency-independent pattern functions of the two effective detectors.
 Contrarily to networks of ground-based observatories, we have no triangulation information from times of arrival at different detectors.
-This is the most degenerate case, and will be useful to get an analytical understanding of the approximate degeneracies occuring when using the more complicated full response (i).
-
-The effect of these approximations on the response is shown in Fig.~\ref{fig:responseapproxtdia22}.
+This case is the most degenerate, and will be useful to get an analytical understanding of approximate degeneracies occuring when using the more complicated full response (i).
+This limit can also be a representative approximation for some short-duration premerger LISA SMBHB observations.
+
+The effect on the response of adopting these approximations is shown in Fig.~\ref{fig:responseapproxtdia22}.
 We display the strain transfer function $\calT_{h_{a}}^{22}(f)$ as defined in~\eqref{eq:defhaet} for the mode $h_{22}$, with the four lines showing the four approximations (i)-(iv).
 Vertical lines also indicate the times to merger highlighted in Sec.~\ref{subsec:MBHBacctime}, as well as the peak frequency.
-Paralleling this figure to Figs.~\ref{fig:hctdiahm} and~\ref{fig:modeoverlapsoftime}, we stress that most of the SNR is accumulated in the very last instants before merger and at the merger itself, due to the noise normalization not visible at the level of the transfer function.
+Relating this figure to Figs.~\ref{fig:hctdiahm} and~\ref{fig:modeoverlapsoftime}, we stress that most of the SNR is accumulated in the very last instants before merger and at the merger itself, due to the noise normalization not visible at the level of the transfer function.
 
@@
   \caption{Same as Fig.~\ref{fig:PEsmbh22hmCase0}, for the MBHB system II of Table~\ref{tab:MBHBparams}.
-           In the extrinsic parameters, the red 2D contours are barely visible on this scale.}
+           In the extrinsic parameters, the red 2D contours are barely visible on this scale. \SM{try fill contours}}
   \label{fig:PEsmbh22hmCase9}
@@
 Using the methodology summarized in Sec.~\ref{sec:method}, here we present the results of Bayesian parameter estimation for the two MBHB sources listed in Table~\ref{tab:MBHBparams}.
-The priors used are logarithmic in mass, flat in luminosity distance, and flat for the angles: flat on the sphere for the pairs of angles $(\iota, \varphi)$ and $(\lambda, \beta)$, and flat for the polarization $\psi$.
+The priors used are logarithmic in mass, flat in luminosity distance, and uniform for the angles: flat on the sphere for the pairs of angles $(\iota, \varphi)$ and $(\lambda, \beta)$, and flat for the polarization $\psi$.
 We expect priors to be unimportant for the masses, since they are well determined in this very high SNR limit.
@@
 
-To illustrate the role of higher harmonics in the analysis, we will present two classes of results: ``22'', where we perform both the simulation and the inference by only including the dominant harmonic $h_{22}$, and ``HM'', where we perform both the simulation and the inference with all available harmonics $(h_{22}, h_{21}, h_{33}, h_{44}, h_{55})$.
-We do not consider the case where the simulation includes all harmonics and the inference is restricted to the 22 mode, as this will most likely bias the inferred parameters by an amount larger than the statistical uncertainties.
-In fact, the SNR calculation in Fig.~\ref{fig:modeoverlapsoftime} suggests that even the set of five modes included in this study may not be sufficient.
-The question of how many harmonics are necessary to keep the systematic errors below the statistical uncertainties is important and will be investigated in the future.
+To illustrate the role of higher harmonics in the analysis, we will present two classes of results: ``22'', where we inject and recover signals including only the dominant harmonic $h_{22}$, and ``HM'', where we both inject and recover with the higher harmonics $(h_{22}, h_{21}, h_{33}, h_{44}, h_{55})$.
+We do not attempt to recover an injection with higher modes using 22-mode waveforms; one would expect parameter biases in this case, induced by the inadequacy of the simplified waveforms.
+A related question would be to assess the mode content necessary for waveform models to mitigate systematic biases when analyzing full GR signals.
+We leave such waveform systematics studies for future work; we only note that, as shown in Fig.~\ref{fig:modeoverlapsoftime}, from SNR only the set of five modes we include in the present study is quite obviously incomplete.
+
+While the 22-mode-only signals here are non-physical, studying them is instructive not only because  22-only waveforms have been a common approximation in previous work but also because (as seen in Fig.~\ref{fig:modeoverlapsoftime}) upon first detection LISA SMBHB signals will often be fairly approximated by 22-mode-only signals.
 
@@
 Looking at correlations, we see that the shifted peak corresponds to the region of high distance (and extremal inclination).
-This is a genuine feature of the multidimensional zero-noise posterior and will be explained in Sec.~\ref{subsec:MBHBPEdegen22}.
-
-In both Cases I and II, including the higher harmonics greatly reduces the uncertainties in the inferred parameters, consistently with previous studies~\cite{Arun+07a, TriasSintes07, PorterCornish08, McWilliams+09}.
-The improvement is much larger than we would expect from the gain in total SNR and a scaling of statistical errors $\sim 1/\SNR$.
-The largest effect is found for the angular parameters and luminosity distance.
-The mass posteriors are slightly more precise, but remain qualitatively similar.
+This is a genuine feature of the multidimensional posterior, even without noise,  and will be explained in Sec.~\ref{subsec:MBHBPEdegen22}.
+
+In both Cases I and II, including the higher harmonics has a major effect on the parameter recovery (as was already stressed in~\cite{Arun+07a, TriasSintes07, PorterCornish08, McWilliams+09}), much beyond what we would expect solely from the modest gain in total SNR shown in Table.~\ref{tab:MBHBparams} (and a scaling of statistical errors $\sim 1/\SNR$).
+The marginalized posterior of the masses is narrower, although not qualitatively different; the major change is in the extrinsic parameters.
 
@@
 
-We can gain insight about degeneracies in the parameter space by considering the simple expressions of the instrument response in the \textit{Frozen low-$f$} approximation described in Sec.~\ref{subsec:MBHBresponseapprox}, appropriate for the low-frequency limit and signals short enough that the LISA motion can be neglected; we will explore in details the respective influence of these neglected effects in Sec.~\ref{sec:MBHBPEacctime}.
+We can gain insight about degeneracies in the parameter space by considering the simple expressions representing the instrument response in the \textit{Frozen low-$f$} approximation described in Sec.~\ref{subsec:MBHBresponseapprox}, appropriate for the low-frequency limit and signals short enough that the LISA motion can be neglected; we will explore the respective influence of these sometimes neglected effects in Sec.~\ref{sec:MBHBPEacctime}.
 
@@
 \ee
-we see that simultaneously changing $\iota \rightarrow \pi-\iota$ and $\psi_{L} \rightarrow \pi - \psi_{L}$ leaves $F_{a}^{\ell m}$, $F_{e}^{\ell m}$ unchanged.
+we see that simultaneously changing $\beta_{L} \rightarrow -\beta_{L}$, $\iota \rightarrow \pi-\iota$ and $\psi_{L} \rightarrow \pi - \psi_{L}$ leaves $F_{a}^{\ell m}$, $F_{e}^{\ell m}$ unchanged.
 
@@
 \end{align}
-How does this situation change when considering a more complete instrument response, moving away from the \textit{Frozen low-$f$} approximation?
+How does this situation change when considering a more complete instrument response, moving away from the \textit{Frozen low-$f$} approximation ? As in Sec IV.C
+we separately consider relaxing each of the qualifiers \textit{Frozen} and \textit{Low-f}.
 
@@
 
-When adding back the LISA motion while still ignoring the frequency-dependence in the response, in the \textit{Low-$f$} response, the \textit{antipodal} mode is the only one that keeps pattern functions that are exactly degenerate with the injection: all other modes will move together with the time-dependent LISA frame.
+When adding back the LISA motion while still ignoring the frequency-dependence in the response, in the \textit{Low-$f$} response, the \textit{antipodal} mode is the only one that keeps pattern functions that are exactly degenerate with the injection: all other modes will evolve with the time-dependent LISA frame.
 The antipodal mode is not moving, because the antipode of the true direction of the arriving signal  is defined as $k \rightarrow -k$ independently of the orientation of LISA; the only degeneracy-breaking term is then the orbital delay $k \cdot p_{0}$ in the Doppler phase~\eqref{eq:defPhiR}.
-With the \textit{Full} response, by contrast, frequency dependency terms in~\eqref{eq:Gslr} featuring $k\cdot n_{l}$, $k \cdot p_{s,r}$ readily break this antipodal degeneracy.
+When considering the \textit{Full} response, the frequency dependency terms in~\eqref{eq:Gslr} featuring $k\cdot n_{l}$, $k \cdot p_{s,r}$ will break this antipodal degeneracy.
 
@@
 We will show that we can understand this feature as a projection effect in the multidimensional degenerate posterior.
-Understanding the structure of these features could prove useful to inform Bayesian samplers, for instance by building adapted jump proposal allowing a much faster mixing of MCMC chains than naive methods.
+Understanding the structure of these features will be useful to inform Bayesian samplers, for instance by building adapted jump proposals allowing a much faster mixing of MCMC chains than naive methods.
 
@@
 with $d = D / D_{\rm inj}$ the dimensionless ratio of luminosity distances, and with mode transfer functions $F_{a,e}^{\ell m}$ given in~\eqref{eq:Flmae}.
+In each term the intrinsic/extrinsic parameter dependence is thus separated with the intrinsic parameters (that we keep fixed) intervening only in $\left\langle \ell m | \ell' m' \right\rangle$.
 
@@
 The log-likelihood is shown as a function of the maximal frequency where we cut the 22-mode signal (with the rescaling~\eqref{eq:fscalinglm} for the other modes), which can be thought of as accumulating signal with time (the vertical lines represent the time cuts of~\ref{subsec:MBHBacctime} and the merger).
-The lines represent the sky positions listed in Table~\ref{tab:MBHBdegen}: in terms of LISA-frame parameters full or dashed stands for the reflection in latitude, and the color stands the $pi/2$-shifted longitudes.
+The lines represent the sky positions listed in Table~\ref{tab:MBHBdegen}: in terms of LISA-frame parameters full or dashed stands for the reflection in latitude, and the color stands for the $\pi/2$-shifted longitudes.
 The dashed black curve corresponds to the \textit{reflected} position, and the dashed red to the \textit{antipodal} position.
 At the injection (solid black), $\ln\calL$ is zero by construction while it is decreasing as a function of $f$ for other sky positions.
-Reaching a very negative $\ln\calL$ means that this sky position is excluded by the signal accumulated thus far.}
+Reaching a very negative $\ln\calL$ means that this sky position is excluded by the signal accumulated thus far.
+In the left panel, the circles indicate the maximum log-likelihood found among the posterior samples, when the parameters are allowed to deviate slightly from the theoretical degenerate point.}
   \label{fig:lnLskymodesapprox}
@@
 
-With the \textit{Low-$f$} response, the \textit{reflected} mode remains exactly degenerate as anticipated in Sec.~\ref{subsec:MBHBPEdegen}.
+With the \textit{Frozen} response, the \textit{reflected} mode remains exactly degenerate as anticipated in Sec.~\ref{subsec:MBHBPEdegen}.
 All other modes see their $\ln \calL$ fall rapidly at around $2\mHz$ due to the onset of frequency-dependency in the response.
-Comparing to Fig.~\ref{fig:responseapproxtdia22}, we see that around $2\mHz$ the breaking of the long-wavelength approximation is still mild; however, the SNR start to accumulate a lot from $\SNR/16 \sim 40$ at $2.5\, \mathrm{h}$ to $\SNR/4 \sim 160$ at $7\min$, magnifying the effect of these features.
-Comparing the two rows of the figure, we see that the higher harmonics have an effect here, in making the elimination of secondary modes happen earlier in frequency (or time).
+Although the frequency dependency vanishes in the limit $f\ll f_L\approx 0.12$ Hz, note that the effect is crucial even for $f/f_L\approx 0.02$.
+Comparing to Fig.~\ref{fig:responseapproxtdia22}, we see that around $2\mHz$ the breaking of the long-wavelength approximation still appears mild; however, the SNR starts to accumulate a lot from $\SNR/16 \sim 40$ at $2.5\, \mathrm{h}$ to $\SNR/4 \sim 160$ at $7\min$, magnifying the effect of these features.
+Comparing the two rows of the figure, we see that the higher harmonics have an effect here, in making the elimination of secondary modes happen slightly earlier in frequency (or time).
 This is expected as they start to contribute significantly to the SNR (see Fig.\ref{fig:modeoverlapsoftime}) and reach higher in frequency than the $h_{22}$ harmonic according to~\eqref{eq:fscalinglm}.
@@
 Contrasting with the SMBHB case, we see that the transfer functions have much more structure in this case.
-Both the time-dependency and the frequency-dependency in the response lead to oscillatory features in the response.
-The Doppler phase~\eqref{eq:defPhiR}, that was negligible for our short SMBHB systems, represents now tens of radians.
+Both the time-dependency and the frequency-dependency in the response lead to oscillatory features in the transfer functions.
+The Doppler phase~\eqref{eq:defPhiR}, that was negligible for our short SMBHB systems, represents now already tens of radians at the lowest frequency.
 
@@
 Using these tools to simulate the parameter recovery of two examples of SMBHB signals, we produced multidimensional posterior distributions.
-While our posterior distributions for masses are not really representative in the absence of know degeneracies between the mass ratio and the spins, we encountered and investigated two different kinds of degeneracies in the extrinsic parameter space.
+While our posterior distributions for masses are not really representative in the absence of known degeneracies between the mass ratio and the spins, we encountered and investigated two different kinds of degeneracies in the extrinsic parameter space.
 
@@
 By decomposing the LISA instrument response in different effects, separating its time-dependency from its frequency-dependency, we could pinpoint the role of both effects.
-The time-dependency of the response, consequence of the motion of LISA, leaves a moderate imprint at low frequencies, penalizing moderately other modes in the sky but leaving the \textit{antipodal} sky position degenerate.
-By contrast, the frequency-dependency of the response, not included in previous parameter studies, is responsible for the degeneracy breaking and the collapse from eight to two modes in the sky.
-Note that the relative importance of the time and frequency dependency of the response will change for heavier or lighter systems.
+The time-dependency of the response, consequence of the motion of LISA, leaves a moderate imprint at low frequencies, moderately penalizing other modes in the sky but leaving the \textit{antipodal} sky position degenerate.
+By contrast, the frequency-dependency of the response, not included in many previous parameter studies, is responsible for the degeneracy breaking and the collapse from eight to two modes in the sky.
+Note that the relative importance of the time and frequency dependency of the response will change for heavier or lighter systems..
 
@@
 Although we do not expect that the noise realization will cause drastic changes, with in particular the structure of degeneracies of signals in parameter space remaining relevant, fast likelihoods for noisy data would be a welcome addition to our methods.
-Analyzing signals with a noise injection will also become more important when considering more realistic instrumental noise models.
+Analyzing signals with a noise injection will also become more important when considering more realistic instrumental noise models and in prototyping algorithms for analysis of actual LISA data.
 
@@
 Finally, we recall that many scientific questions about LISA rely on parameter estimation tools, and would benefit from proper Bayesian analyses with realistic signals: pre-merger localization of sources and advance warnings for electromagnetic instruments, cosmography using LISA sources as standard sirens, astrophysical inference from the source population, accuracy requirements on waveform models, trade-offs between instrumental design and scientific outputs.
-We hope that our tools will become useful for all these applications.
+We will keep extending our tools to address all these applications.
 
@@
 
-We would like to thank Stanislav Babak, Michael Katz, Vivien Raymond and Alexandre Toubiana for useful discussions.
+We would like to thank Stanislav Babak, Michael Katz, Antoine Klein, Vivien Raymond and Alexandre Toubiana for useful discussions.
 We are indebted to Philip Graff for his help with the \texttt{bambi} sampler.

### 2020-02-29 2020-02-29_submission/paper_PE.tex
Added submission dir

--- 2020-02-28 paper/paper_PE.tex
@@
 
-\newcommand{\SM}[1]{{\color{Blue} [SM: #1]}}
-\newcommand{\JGB}[1]{{\color{DarkGreen} [JGB: #1]}}
-\newcommand{\TDC}[1]{{\color{red} [TDC: #1]}}
+%\newcommand{\SM}[1]{{\color{Blue} [SM: #1]}}
+%\newcommand{\JGB}[1]{{\color{DarkGreen} [JGB: #1]}}
+%\newcommand{\TDC}[1]{{\color{red} [TDC: #1]}}
 
@@
 
-The space-based gravitational wave detector LISA will observe mergers of supermassive black hole binary systems (SMBHBs) to cosmological distances, as well as inspiralling stellar-origin (or stellar-mass) binaries (SBHBs) years before they enter the LIGO/Virgo band.
+The space-based gravitational wave detector LISA will observe mergers of massive black hole binary systems (MBHBs) to cosmological distances, as well as inspiralling stellar-origin (or stellar-mass) binaries (SBHBs) years before they enter the LIGO/Virgo band.
 Much remains to be explored for the parameter recovery of both classes of systems.
-Previous SMBHB analyses relied on inspiral-only signals and/or a simplified Fisher matrix analysis, while SBHBs have not yet been extensively analyzed with Bayesian methods.
+Previous MBHB analyses relied on inspiral-only signals and/or a simplified Fisher matrix analysis, while SBHBs have not yet been extensively analyzed with Bayesian methods.
 We accelerate likelihood computations by (i) using a Fourier-domain response of the LISA instrument, (ii) using a reduced-order model for non-spinning waveforms that include a merger-ringdown and higher harmonics, (iii) setting the noise realization to zero and computing overlaps in the amplitude/phase representation.
-We present the first simulations of Bayesian inference for the parameters of supermassive black hole systems including consistently the merger and ringdown of the signal, as well as higher harmonics.
+We present the first simulations of Bayesian inference for the parameters of massive black hole systems including consistently the merger and ringdown of the signal, as well as higher harmonics.
 We clarify the roles of LISA response time and frequency dependencies in breaking degeneracies and illustrate how degeneracy breaking unfolds over time.
@@
 Ground-based interferometers are fundamentally limited at low frequency by terrestrial noise, and cannot be used to study mergers of compact objects much heavier than a few $ 10^2 \Msol$.
-The spacebourne detector LISA~\cite{LISA2017} will overcome this limitation, enabling the observation and precise characterization of gravitational waves from binary black holes, from coalescences of systems with millions of $\Msol$ (supermassive black hole binaries, SMBHB~\cite{Klein+15}) to the inspiral of systems with tens of $\Msol$ (stellar-mass black hole binaries, SBHB~\cite{Sesana16}).
+The spacebourne detector LISA~\cite{LISA2017} will overcome this limitation, enabling the observation and precise characterization of gravitational waves from binary black holes, from coalescences of systems with millions of $\Msol$ (massive black hole binaries, MBHB~\cite{Klein+15}) to the inspiral of systems with tens of $\Msol$ (stellar-mass black hole binaries, SBHB~\cite{Sesana16}).
 The observation of black hole mergers and inspirals in the LISA frequency band will yield major scientific rewards, and they form the main target of LISA among other classes of sources.
@@
 In LISA, this problem is more complex than in kilometer-scale ground-based interferometers, as additional challenges arises from the much larger expected signal-to-noise ratios (SNRs) for high-mass systems, or the much longer duration of the waveform for low-mass systems.
-In particular, for SMBHBs, the large SNR will require a very accurate modeling of the waveform, including the merger and ringdown regimes, the effects of spin and corrections of higher harmonics in the signal beyond the quadrupole.
+In particular, for MBHBs, the large SNR will require a very accurate modeling of the waveform, including the merger and ringdown regimes, the effects of spin and corrections of higher harmonics in the signal beyond the quadrupole.
 
@@
 
-Here we demonstrate that standard Bayesian inference can be performed for SMBHBs using a self-consistent waveform model that includes inspiral, merger, ringdown and higher-order modes (albeit no spins) and a full model of the LISA response.
+Here we demonstrate that standard Bayesian inference can be performed for MBHBs using a self-consistent waveform model that includes inspiral, merger, ringdown and higher-order modes (albeit no spins) and a full model of the LISA response.
 We investigate and explain degeneracies in the posterior distribution of the source's parameters, and highlight the crucial role played by the higher harmonics in the signal and the frequency-dependency in the instrument response.
@@
 Parameter estimation studies for SBHB systems have relied so far on the Fisher matrix approach~\cite{Sesana16, Vitale16, Nishizawa+16a, Nishizawa+16b}, with Bayesian inference tools yet to be developed.
-Focusing on the posterior distribution of extrinsic parameters (fixing the masses and time to coalescence), we demonstrate that Bayesian analyses can be run with a speed for the likelihood computation that is comparable to the SMBHB case, and discuss the features of the posteriors.
+Focusing on the posterior distribution of extrinsic parameters (fixing the masses and time to coalescence), we demonstrate that Bayesian analyses can be run with a speed for the likelihood computation that is comparable to the MBHB case, and discuss the features of the posteriors.
 
 Section~\ref{sec:response} introduces comprehensive notations for the frequency-domain signal and response, and Section~\ref{sec:method} explains our methods for likelihood computations and Bayesian sampling.
-Section~\ref{sec:SMBH} presents the SMBHB example signals that we analyze and their accumulation with time, together with their instrumental response in various approximations.
-In Section~\ref{sec:SMBHPE}, we present our main results for the parameter estimation of SMBHBs; we contrast results obtained with and without higher harmonics, contrast Fisher matrix with Bayesian results, describe and explain degeneracies in parameter space.
+Section~\ref{sec:SMBH} presents the MBHB example signals that we analyze and their accumulation with time, together with their instrumental response in various approximations.
+In Section~\ref{sec:SMBHPE}, we present our main results for the parameter estimation of MBHBs; we contrast results obtained with and without higher harmonics, contrast Fisher matrix with Bayesian results, describe and explain degeneracies in parameter space.
 In Section~\ref{sec:MBHBPEacctime}, we turn to the refinement of parameter estimation as the signal accumulates with ongoing observations, describing how multiple inferred sky positions can coexist before features of the full instrument response break the degeneracies.
@@
 
-The analysis of~\cite{MB18} has shown that higher-order corrections in the separation of timescales in the LISA response are small in general for SMBHB systems, and are also small for SBHB systems provided they are not too far from the coalescence.
+The analysis of~\cite{MB18} has shown that higher-order corrections in the separation of timescales in the LISA response are small in general for MBHB systems, and are also small for SBHB systems provided they are not too far from the coalescence.
 The fact that we use the same Fourier-domain treatment of the transfer functions for both the signal and the templates should also mitigate the importance of modelling errors.
@@
 with the ``$\rm inj$'' subscript indicating the simulated signals, and where the inner products $(\cdot | \cdot)$ are given by~\eqref{eq:definnerproduct} with the noise PSDs~\eqref{eq:Snaet}.
-\SM{check nothing was git-lost here}
 
@@
 In the LISA case, the computational cost of the standard likelihood implementation depends primarily on the mass of the system and the duration of the signal.
-This cost can range from tens of milliseconds for SMBHBs and short-lived signals (a few days), to impracticable values for SBHBs signals lasting for years and sampled at high frequency.
+This cost can range from tens of milliseconds for MBHBs and short-lived signals (a few days), to impracticable values for SBHBs signals lasting for years and sampled at high frequency.
 By setting the noise realization to zero, however, we can represent the amplitude and phase of the signals over a sparse frequency grid and obtain a large speedup.
@@
 		SBHB & $(2,2)$ & $500$ & $2$ ms \\
-		SMBHB & $(2,2)$ & $300$ & $1.2$ ms \\
-		SMBHB & $(2,2),(2,1),(3,3),(4,4),(5,5)$ & $300$ & $15$ ms \\
+		MBHB & $(2,2)$ & $300$ & $1.2$ ms \\
+		MBHB & $(2,2),(2,1),(3,3),(4,4),(5,5)$ & $300$ & $15$ ms \\
 		\hline
@@
 %	\item for SBHBs with $h_{22}$ only: $\sim 500$ grid points, likelihood cost $\sim 2\mathrm{ms}$;
-%	\item for SMBHBs with $h_{22}$ only: $\sim 300$ grid points, likelihood cost $\sim 1.2\mathrm{ms}$;
-%	\item for SMBHBs with 5 modes: $\sim 300$ grid points, likelihood cost $\sim 15\mathrm{ms}$.
+%	\item for MBHBs with $h_{22}$ only: $\sim 300$ grid points, likelihood cost $\sim 1.2\mathrm{ms}$;
+%	\item for MBHBs with 5 modes: $\sim 300$ grid points, likelihood cost $\sim 15\mathrm{ms}$.
 %\end{itemize}
@@
 
-We focus our analysis on a single choice of mass parameters, representative of canonical SMBHB sources expected for LISA~\cite{Barausse12, Klein+15, LISA2017}: we pick a total redshifted mass $M=m_{1}+m_{2} = 2\times10^{6} M_{\odot}$ and a mass ratio $q=m_{1}/m_{2} = 3$, and place the source at a redshift $z=4$\footnote{We convert between luminosity distances and redshifts using the cosmological parameters of~\cite{Planck15param}}.
+We focus our analysis on a single choice of mass parameters, representative of canonical MBHB sources expected for LISA~\cite{Barausse12, Klein+15, LISA2017}: we pick a total redshifted mass $M=m_{1}+m_{2} = 2\times10^{6} M_{\odot}$ and a mass ratio $q=m_{1}/m_{2} = 3$, and place the source at a redshift $z=4$\footnote{We convert between luminosity distances and redshifts using the cosmological parameters of~\cite{Planck15param}}.
 The corresponding source-frame total mass is $M/(1+z) = 4\times10^{5} M_{\odot}$.
@@
 
-We note that, by contrast with current ground-based observations that are horizon-limited and that have a strong selection bias towards systems with favorable orientations (face-on or face-off), the LISA horizon for SMBHB systems extends to essentially the entirety of the observable universe~\cite{LISA2017}; it is therefore natural to consider randomized orientations.
+We note that, by contrast with current ground-based observations that are horizon-limited and that have a strong selection bias towards systems with favorable orientations (face-on or face-off), the LISA horizon for MBHB systems extends to essentially the entirety of the observable universe~\cite{LISA2017}; it is therefore natural to consider randomized orientations.
 Here, although we focus on just two systems, we take $\iota = \pi/3$, the median value when inclination is randomized.
@@
 	\end{tabular}
-	\caption{Parameters of the simulated SMBHB mergers. Angles are given in the SSB-frame. The $\mathrm{SNR}$ labelled $h_{\ell m}$ is obtained including all harmonics listed in~\eqref{eq:listmodes}.}
+	\caption{Parameters of the simulated MBHB mergers. Angles are given in the SSB-frame. The $\mathrm{SNR}$ labelled $h_{\ell m}$ is obtained including all harmonics listed in~\eqref{eq:listmodes}.}
 	\label{tab:MBHBparams}
@@
 The time-to-frequency correspondence is computed according to $t_{f}^{\ell m}$ defined in~\eqref{eq:deftflm}, evaluated at $m f / 2$ to map to the same time according to~\eqref{eq:fscalinglm}.
-The vertical lines represent the time cuts explained in~\ref{subsec:MBHBacctime} and the merger frequency.
-\SM{fix 40h -> 42h}}
+The vertical lines represent the time cuts explained in~\ref{subsec:MBHBacctime} and the merger frequency.}
   \label{fig:t-f-relation-lm}
@@
 In all panels the black line represents the total.
-The horizontal axis gives the time to merger, with vertical lines for the cuts of~\ref{subsec:MBHBacctime}, and the dots on the right of each panel show the result for the full signal including the ringdown.
-\SM{also change title $(h|h)$, these are mode contributions to TDI not $h$}}
+The horizontal axis gives the time to merger, with vertical lines for the cuts of~\ref{subsec:MBHBacctime}, and the dots on the right of each panel show the result for the full signal including the ringdown.}
   \label{fig:modeoverlapsoftime}
@@
 In the low-frequency picture of~Sec.~\ref{subsec:lowfresponse}, the time dependency enters both in the Doppler phase term~\eqref{eq:defPhiR} (LISA moves in the wavefront) and in the time-dependent LISA-frame angles~\eqref{eq:Lframeangles} (the LISA arms change orientation).
-However, we have seen in the previous section that the SMBHB signals we consider here are quite short (less than two days).
+However, we have seen in the previous section that the MBHB signals we consider here are quite short (less than two days).
 This limits the effect of the LISA motion: between the time where the signal becomes detectable and the merger, LISA has barely changed in orientation and position.
@@
 This case is the most degenerate, and will be useful to get an analytical understanding of approximate degeneracies occuring when using the more complicated full response (i).
-This limit can also be a representative approximation for some short-duration premerger LISA SMBHB observations.
+This limit can also be a representative approximation for some short-duration premerger LISA MBHB observations.
 
@@
 The time is centered so that $\Delta t=0$ corresponds to the injection.
-All extrinsic parameters are given in the LISA-frame as defined in App.~\ref{app:LISAframe}.
-\SM{add L subscript to parameters labels for L-frame -- maybe use D and not DL}}
+All extrinsic parameters are given in the LISA-frame as defined in App.~\ref{app:LISAframe}.}
   \label{fig:PEsmbh22hmCase0}
@@
   \caption{Same as Fig.~\ref{fig:PEsmbh22hmCase0}, for the MBHB system II of Table~\ref{tab:MBHBparams}.
-           In the extrinsic parameters, the red 2D contours are barely visible on this scale. \SM{try fill contours}}
+           In the extrinsic parameters, the red 2D contours are barely visible on this scale.}
   \label{fig:PEsmbh22hmCase9}
@@
 
-While the 22-mode-only signals here are non-physical, studying them is instructive not only because  22-only waveforms have been a common approximation in previous work but also because (as seen in Fig.~\ref{fig:modeoverlapsoftime}) upon first detection LISA SMBHB signals will often be fairly approximated by 22-mode-only signals.
+While the 22-mode-only signals here are non-physical, studying them is instructive not only because  22-only waveforms have been a common approximation in previous work but also because (as seen in Fig.~\ref{fig:modeoverlapsoftime}) upon first detection LISA MBHB signals will often be fairly approximated by 22-mode-only signals.
 
@@
 We will show that we can understand this feature as a projection effect in the multidimensional degenerate posterior.
-Understanding the structure of these features will be useful to inform Bayesian samplers, for instance by building adapted jump proposals allowing a much faster mixing of MCMC chains than naive methods.
+Understanding the structure of these features could prove useful to inform Bayesian samplers, for instance by adapting jump proposals to speed up mixing of MCMC chains.
 
@@
 There is also a symmetry based on exchanging $\iota \leftrightarrow \theta_{L}$.
-While leaving $\sigma_-$ unchanged, this conjugates the factor inside brackets for $\sigma_{+}$, which can be compensated using the phase terms $\varphi \pm \lambda_{a}$.
-\JGB{Should this just be ``+'' not ``$\pm$''?} If $\Phi = \mathrm{Arg} \left[ t_{\theta}^{4} e^{-2 i \psiL} + t_{\iota}^{4} e^{2 i \psiL} \right] $, we obtain the symmetry
+While leaving $\sigma_-$ unchanged, this conjugates the factor inside brackets for $\sigma_{+}$, which can be compensated using the phase term $\varphi - \lambda_{a}$.
+If $\Phi = \mathrm{Arg} \left[ t_{\theta}^{4} e^{-2 i \psiL} + t_{\iota}^{4} e^{2 i \psiL} \right] $, we obtain the symmetry
 \begin{align}
@@
 Beyond these discrete symetries, since likelihood function dependence on the six extrinsic parameters is funneled through just two complex functions, we should expect a two-dimensional degenerate subspace.
-Indeed we can explicitly find a general solution for parameter values that solve $\sigma_{\pm}((D, t_\iota, t_\theta \varphi, \lambda_a, \psiL)=\sigma_{\pm,{\rm inj}}$, which we just sketch here, as needed to explain features of the degeneracies.
+Indeed we can explicitly find a general solution for parameter values that solve $\sigma_{\pm}(D, t_\iota, t_\theta, \varphi, \lambda_a, \psiL)=\sigma_{\pm,{\rm inj}}$, which we just sketch here, as needed to explain features of the degeneracies.
 
@@
    \end{minipage}
+   \begin{minipage}{.32\linewidth}
+      \includegraphics[width=.99\linewidth]{../plots/corner_smbh_case9_hm_tseries_m1m2_zoom.png}
+   \end{minipage}
+   \begin{minipage}{.32\linewidth}
+      \includegraphics[width=.99\linewidth]{../plots/corner_smbh_case9_hm_tseries_Dinc_zoom.png}
+   \end{minipage}
+   \begin{minipage}{.32\linewidth}
+      \includegraphics[width=.99\linewidth]{../plots/corner_smbh_case9_hm_tseries_tphi_zoom.png}
+   \end{minipage}
   \caption{Refinement of parameter inference over time for system II.
-The posteriors shown correspond to the time cuts defined in~\ref{subsec:MBHBacctime} at $\mathrm{SNR}/64$ (green), $\mathrm{SNR}/16$, $\mathrm{SNR}/4$ and post-merger (blue)., for the masses (left), distance/inclination (center), and time/phase (right).
+The posteriors shown correspond to the time cuts defined in~\ref{subsec:MBHBacctime} at $\mathrm{SNR}/64$ (green), $\mathrm{SNR}/16$ (yellow), $\mathrm{SNR}/4$ (red) and post-merger (blue) for the masses (left), distance/inclination (center), and time/phase (right).
+On the first row, all four posteriors are shown, while in the second row we display only $\mathrm{SNR}/4$ and post-merger, zooming closer to the true parameters.
 We inject and recover with higher harmonics.
@@
 The dashed black curve corresponds to the \textit{reflected} position, and the dashed red to the \textit{antipodal} position.
-At the injection (solid black), $\ln\calL$ is zero by construction while it is decreasing as a function of $f$ for other sky positions.
-Reaching a very negative $\ln\calL$ means that this sky position is excluded by the signal accumulated thus far.
+At the true position (solid black), $\ln\calL$ is zero by construction while it is decreasing as a function of $f$ for other sky positions.
+Reaching a very negative $\ln\calL$ means that this sky position is excluded by the data accumulated thus far.
 In the left panel, the circles indicate the maximum log-likelihood found among the posterior samples, when the parameters are allowed to deviate slightly from the theoretical degenerate point.}
@@
 The parameters of the simulated mergers were drawn from the Radler LISA Data Challenge (LDC), which itself followed the population of~\cite{Sesana16}, and are listed in Table~\ref{tab:SBHBparams}.
-The system I is a moderately massive system, placed at a favorable close distance; with an SNR of $26.6$ it would be at the tail of the most favorable detections that we can hope for~\cite{Sesana16, Gerosa+19, Moore+19}.
-The system II is a higher-mass, moderate-SNR system.
+System I is moderately massive, placed at a favorable close distance; with an SNR of $26.6$ it would be at the tail of the most favorable detections that we can hope for~\cite{Sesana16, Gerosa+19, Moore+19}.
+System II is a higher-mass, moderate-SNR system.
 Both systems happen to be close to the ecliptic plane.
@@
 Fig.~\ref{fig:transferSBHBcaseI} illustrates, for the system I in this table, the instrumental transfer functions~\eqref{eq:defTlmaet} as well as the Doppler phase~\eqref{eq:defPhiR}.
-Contrasting with the SMBHB case, we see that the transfer functions have much more structure in this case.
+Contrasting with the MBHB case, we see that the transfer functions have much more structure in this case.
 Both the time-dependency and the frequency-dependency in the response lead to oscillatory features in the transfer functions.
-The Doppler phase~\eqref{eq:defPhiR}, that was negligible for our short SMBHB systems, represents now already tens of radians at the lowest frequency.
-
-%The TDI channel $T$, negligible as well for low-frequency SMBHB signals, makes here a small contribution at high frequencies.
+The Doppler phase~\eqref{eq:defPhiR}, that was negligible for our short MBHB systems, now varies by tens of radians already at the lowest frequency.
+
+%The TDI channel $T$, negligible as well for low-frequency MBHB signals, makes here a small contribution at high frequencies.
 
@@
 Bayesian inference for SBHB sources in LISA using the methodology presented in
-this paper, and to illustrate the qualitative features of the posterior distribution for extrinsic parameters, in relation to our previous investigations for SMBHB systems.
+this paper, and to illustrate the qualitative features of the posterior distribution for extrinsic parameters, in relation to our previous investigations for MBHB systems.
 To this end, the problem we pose is simplified
@@
 The typical degeneracy between inclination and distance
-is evident, as there is no degeneracy breaking from higher harmonics.
-Even though the SNR values are much lower than for our SMBHB systems, the Fisher matrix estimates represent quite well the marginal
+is evident, as there is no degeneracy breaking from higher harmonics. Even though the SNR values are much lower than for our MBHB systems, the Fisher matrix estimates represent quite well the marginal
 distributions of most parameters.
-We note, however, some notable deviation from Gaussianity for the system II: the sky position is almost bimodal, with an approximate symmetry across the ecliptic plane (we recall here that the ecliptic plane is not the plane of the LISA instrument that was used to report SMBHB results), owing probably to the localization being close to this plane; and the distance and inclination posteriors show a long tail extending way beyond the Fisher estimate.
+We note, however, some notable deviation from Gaussianity for the system II: the sky position is almost bimodal, with an approximate symmetry across the ecliptic plane (we recall here that the ecliptic plane is not the plane of the LISA instrument that was used to report MBHB results), owing probably to the localization being close to this plane; and the distance and inclination posteriors show a long tail extending way beyond the Fisher estimate.
 
@@
 The initial frequency is fixed to obtain a time-to-merger of 2 years.
-All extrinsic parameters are given in the SSB-frame.
-\SM{use D and not DL}}
+All extrinsic parameters are given in the SSB-frame.}
 	\label{fig:sbhb_corner_1}
@@
 
-In this work, we have explored the Bayesian parameter estimation for full, inspiral-merger-ringdown SMBHB signals including higher harmonics, albeit neglecting the spin degrees of freedom, and we have also produced posterior distributions for the extrinsic parameters of SBHB signals.
+In this work, we have explored the Bayesian parameter estimation for full, inspiral-merger-ringdown MBHB signals including higher harmonics, albeit neglecting the spin degrees of freedom, and we have also produced posterior distributions for the extrinsic parameters of SBHB signals.
 
@@
 
-Using these tools to simulate the parameter recovery of two examples of SMBHB signals, we produced multidimensional posterior distributions.
-While our posterior distributions for masses are not really representative in the absence of known degeneracies between the mass ratio and the spins, we encountered and investigated two different kinds of degeneracies in the extrinsic parameter space.
+Using these tools to simulate the parameter recovery of two examples of MBHB signals, we produced multidimensional posterior distributions.
+Because our study did not consider spins and their well-known degeneracy with masses, our posterior distributions for masses cannot yet be taken as representative of the actual LISA capabilities.
+Nevertheless, we encountered and investigated two different kinds of degeneracies in the extrinsic parameter space.
 
 Considering first IMR signals, we found a \textit{two-fold} degeneracy for the sky position corresponding to the \textit{reflected} sky localization with respect to the plane of the LISA constellation (opposite sign of the LISA-frame latitude).
-We also found that, when limiting the signals to their dominant harmonic $h_{22}$, strong degeneracies in distance-inclination could occur with a posterior distribution in the sky appearing shifted from the injected value.
+We also found that, when limiting the signals to their dominant harmonic $h_{22}$, strong degeneracies in distance-inclination can occur, causing the posterior distribution in the sky to appear shifted from the injected value.
 We gave an analytic explanation of this degeneracy by eliminating the time and frequency dependencies in the response, reducing it to an explicitly degenerate form.
@@
 
-Secondly, considering pre-merger analyses to investigate how information accumulates with time, we found an \textit{eight-fold} degeneracy for the sky position, obtained by duplicating the original and the reflected sky positions with $\pi/2$, $\pi$, $-\pi/2$ shifts in the LISA-frame longitude.
+Secondly, considering pre-merger analyses to investigate how information accumulates with time, we found an \textit{eight-fold} degeneracy for the sky position, obtained by rotating the original and reflected sky positions by integer multiples of $\pi/2$ around the axis of the LISA constellation.
 For our example systems, the posterior only collapses on the two-fold degeneracy, noticed previously for IMR signals, minutes before merger.
-By decomposing the LISA instrument response in different effects, separating its time-dependency from its frequency-dependency, we could pinpoint the role of both effects.
-The time-dependency of the response, consequence of the motion of LISA, leaves a moderate imprint at low frequencies, moderately penalizing other modes in the sky but leaving the \textit{antipodal} sky position degenerate.
-By contrast, the frequency-dependency of the response, not included in many previous parameter studies, is responsible for the degeneracy breaking and the collapse from eight to two modes in the sky.
-Note that the relative importance of the time and frequency dependency of the response will change for heavier or lighter systems..
-
-We also explored the recovery of extrinsic parameters of SBHB signals (fixing the masses and spin parameters as well as the merger time), using the same methodology as for SMBHB signals to arrive at a likelihood cost of $\sim 2 \mathrm{ms}$.
-Investigating two example systems, we found that the distribution of extrinsic parameters is mostly Gaussian, with for one case some bimodality in the sky position found for one source lying close to the ecliptic plane, and an extended tail in the distance posterior.
-
-Comparing the parameter estimation results obtained with our two samplers, $\texttt{ptmcmc}$ and $\texttt{multinest}$, we found in general a good agreement.
+By decomposing the LISA instrument response into different effects, separating its time-dependency from its frequency-dependency, we could pinpoint the role of both effects.
+The time-dependency of the response, consequence of the motion of LISA, leaves a moderate imprint at low frequencies, penalizing moderately other modes in the sky but leaving the \textit{antipodal} sky position degenerate.
+By contrast, the frequency-dependency of the response, not included in previous parameter studies, is responsible for the degeneracy breaking and the collapse from eight to two modes in the sky.
+Note that the relative importance of the time and frequency dependency of the response will change for heavier or lighter systems.
+
+We also explored the recovery of extrinsic parameters of SBHB signals (fixing the masses and spin parameters as well as the merger time), using the same methodology as for MBHB signals to arrive at a likelihood cost of $\sim 2 \mathrm{ms}$.
+Investigating two example systems, we found that the distribution of extrinsic parameters is mostly Gaussian, with a bimodality in the sky position found for one source lying close to the ecliptic plane, and an extended tail in the distance posterior.
+
+Comparing the parameter estimation results obtained with our two samplers, $\texttt{ptmcmc}$ and $\texttt{multinest}$, we found an overall good agreement.
 We noted however that \texttt{multinest} can fall short of exploring the full degeneracies in parameter space, in the degenerate case of a signal without higher harmonics; this limitation is further illustrated by the example of a fully degenerate response.
-We also found that using Fisher matrix estimates would not reproduce Bayesian results in general: degeneracies in the sky would not be covered by a Fisher computation that is intrinsically unimodal, and neither would the specific features of the $22$-mode degenerate posteriors.
-That degeneracies occur even with the high SNRs of our SMBHB signals is remarkable, and reminds us that the usual statement of the validity of the Fisher matrix approximation in the high-SNR limit is very dependent on the morphology of the signal.
-In many respects, the moderate-SNR posteriors for our SBHB systems are more Gaussian than our high-SNR SMBHB $22$-mode posteriors.
+We also found that using Fisher matrix estimates does not reproduce Bayesian results in general: the complicated degeneracies in the sky cannot be covered by a Fisher computation that is intrinsically unimodal, and neither can the specific features of the $22$-mode degenerate posteriors.
+That degeneracies occur even with the high SNRs of our MBHB signals is remarkable, and reminds us that the usual statement of the validity of the Fisher matrix approximation in the high-SNR limit is very dependent on the morphology of the signal.
+In many respects, the moderate-SNR posteriors for our SBHB systems are more Gaussian than our high-SNR MBHB $22$-mode posteriors.
 
@@
 First, we only investigated a handful of sources.
-In particular, we chose our SMBHB systems to be representative of the bulk of the population; they are not ``golden sources'', i.e.
-they are not part of the tail of lower-redshift systems that would provide us with the most interesting science outputs with LISA, notably in terms of electromagnetic counterparts.
+In particular, we chose our MBHB systems to be representative of the bulk of the expected population; they are not ``golden sources'', i.e.
+they are not part of the tail of lower-redshift systems that will deliver the most interesting science outputs with LISA, notably in terms of electromagnetic counterparts.
 It is likely that the parameter recovery here will be different, with a longer detectable signal and the motion of LISA playing a role in the premerger localization.
-We also limited ourselves to a single redshifted mass, while the morphology of the signals will change strongly between the two ends of the LISA spectrum, from intermediate mass black holes $M\sim 10^{3} M_{\odot}$ to massive systems $M>10^{7} M_{\odot}$.
-
-In the future, it will be important to incorporate more physics in our waveform models.
-In this study our signals were limited to non-spinning, circularized systems.
-The aligned components of the spin will play a role by allowing degeneracies between masses and spin magnitudes, but we can expect that they will not affect qualitatively the posterior distributions for extrinsic parameters; this will be investigated in a forthcoming publication~\cite{Katz+20}.
-Precession and eccentricity, however, will change the harmonic structures of the signals and play an important role.
-
-A limitation of our method is that, in order to accelerate the likelihood computation, set the noise realization to zero.
-Although we do not expect that the noise realization will cause drastic changes, with in particular the structure of degeneracies of signals in parameter space remaining relevant, fast likelihoods for noisy data would be a welcome addition to our methods.
+We also limited ourselves to a single redshifted mass, while the morphology of the signals will change strongly between the two ends of the LISA spectrum, from intermediate mass black holes ($M\sim 10^{3} M_{\odot}$) to massive systems ($M>10^{7} M_{\odot}$).
+
+Second, it will be necessary to incorporate more physics in our waveform models.
+In this study our signals were limited to non-spinning, quasicircular systems.
+The spin components along the orbital angular momentum will be degenerate with the masses, but we do not expect this to qualitatively change the posterior distributions for the extrinsic parameters.
+This will be investigated in a forthcoming publication~\cite{Katz+20}.
+Orbital precession and eccentricity, however, will change the harmonic structure of the signals and we expect they will play a much more important role.
+
+Third, we used zero-noise realizations in order to accelerate the likelihood computation.
+Although we do not expect that the noise realization will cause drastic changes, with in particular the structure of degeneracies of signals in parameter space remaining relevant, fast likelihoods for pre-sampled noisy data will eventually be necessary.
 Analyzing signals with a noise injection will also become more important when considering more realistic instrumental noise models and in prototyping algorithms for analysis of actual LISA data.
@@
 
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-%
-%\subsection{The LISA response - OLD}
-%\label{sec:lisaresponse}
-%
-%\SM{------------------------------------------------------}
-%
-%\SM{[copy-paste from previous paper -- just here to copy needed formulas / refs]}
-%
-%We begin by detailing the model that we use for the response of a LISA-like detector, together with the assumptions used and their limitations.
-Since our aim is to assess the accuracy of our direct Fourier-domain treatment of the response, we can focus on the gravitational-wave contribution to the basic single-link observables.
-We therefore use a somewhat simplified model for the time-domain response~\cite{Krolak+04}, ignoring corrections that would be crucial from the point of view of noise cancellations, keeping in mind that the response model can be enriched later without affecting the conclusions of the present analysis.
-%
-%The frequency-shift response for a single link~\eqref{eq:yslr} was derived in Ref.~\cite{EW75} (see also~\cite{CR02, RCP04, Finn08, Cornish09}).
-Several assumptions enter the result as written in~\eqref{eq:yslr}: (i) effects of the order $v/c$ are neglected, including for instance the special relativistic Doppler effect created by the relative speeds of the spacecraft on their orbits (ii) the propagation is assumed to take place in a flat spacetime, perturbed only by the gravitational wave; thus the gravitational redshift as well as the deflection of light created by the gravitational potential of the Sun is ignored (iii) all geometric factors are evaluated at a single time, whereas one should consider the beam as propagating from the position of the first spacecraft at the time of emission to the position of the second scapecraft at the time of reception, leading to a point-ahead effect.
-%
-%Additionally, we limit ourselves to a rigid model for the orbits of the constellation, namely we assume that the constellation remains in an equilateral configuration with fixed armlengths.
-These simplified orbits neglect (iv) effects of order $e^{2}$ from the eccentricity of the individual Keplerian orbits, (v) the effect of gravitational perturbations coming from other celestial bodies, such as the Earth, the quadrupole of the Sun, and the other planets.
-Note that although we can neglect all the effects (i)-(v) for our present study, keeping track of these corrections is crucial for the purpose of laser noise cancellations, and led to the development of new generations of TDI observables~\cite{Tintoliving}.
-%
-%It is natural to split the response~\eqref{eq:yslr} into two steps: first the orbital delay related to the orbit around the Sun of the whole constellation, and then the constellation response.
-The baselines for the delays are indeed very different in the two cases.
-Geometrical projection factors aside, we have for the orbit around the Sun $R=1\text{au}=1.5\times 10^{8} \text{km}$, while the detector armlength is $L=2.5\times 10^{6}\text{km}$ (in the configuration proposed in~\cite{LISA17}).
-It is natural to define two transfer frequencies for the two relevant length scales for the delays, defined such that a wavelength fits within this length scale, i.e.
-$2\pi f d = 1$.
-This gives
-%\begin{subequations}\label{eq:transferfrequencies}
-%\begin{align}
-%	f_{R} &= 3.2\times10^{-4}\Hz \,,\\
-%	f_{L} &= 1.9\times 10^{-2}\Hz \,.
-%\end{align}
-%\end{subequations}
-%The LISA response will behave qualitatively differently on the three frequency bands $f \leq f_{R}$, $f_{R} \leq f \leq f_{L}$ and $f_{L} \leq f$.
-%
-%The first stage of the response, the orbital delay, consists simply in applying the varying time delay to bring the wavefront sampling point from the SSB reference to the center of the LISA triangular constellation, common to all $y_{slr}$ observables.
-For $h^{\rm TT}$ the transverse-traceless gravitational waveform in matrix form, we write this orbital time delay as
-%\be\label{eq:defresponse0}
-%	h_{0}^{\rm TT} (t) = h^{\rm TT}(t-\hatk\cdot p_{0}) \,,
-%\ee
-%with $p_{0}$ the position of the constellation center, which follows the Earth orbit around the Sun.
-The second stage of the response calculation comprises the remaining, constellation-centered response.
-For the single-link contribution to the response, for the laser link from spacecraft $s$ to spacecraft $r$ along a path in direction $n_l$, we write
-%\begin{align}\label{eq:defresponseL}
-%	y_{slr} &= \frac{1}{2} \frac{1}{1 - \hatk\cdot n_{l}} \nn\\
-%	& \cdot n_{l}\cdot \left[ h_{0}^{\rm TT}(t - L - \hatk\cdot p^{L}_{s}) - h_{0}^{\rm TT}(t - \hatk\cdot p^{L}_{r}) \right] \cdot n_{l}\,,
-%\end{align}
-%where we reference the positions of the spacecraft relative to the center of the constellation, $p^{L}_{A} \equiv p_{A} - p_{0}$.
-%
-%As described in App.~\ref{app:notation} We will decompose the full signal in the contributions of the individual spin-weighted spherical modes $h_{\ell m}$, whose Fourier transforms are assumed to have a smooth amplitude and phase.
-First, we define the matrices $P_{+},P_{\times}$ such that, in the sense of matrices,
-%\be
-%	h^{\rm TT} = h_{+}P_{+} + h_{\times}P_{\times} \,.
-%\ee
-%We focus only on positive frequencies.
-Assuming that the approximation~\eqref{eq:zeronegativef} applies, we consider a single mode contribution, $h=h_{\ell m}$ with $m>0$.
-For each given mode we define a complex matrix $P_{\ell m}$ incorporating the spin-weighted spherical harmonic constant factor as
-%\be
-%	P_{\ell m} =
-%	\begin{cases}
-%	\frac{1}{2} {}_{-2}Y_{\ell m} \left( P_{+} + i P_{\times} \right) \text{ for } m>0\,,\\
-%	\frac{1}{2} {}_{-2}Y_{\ell m}^{*} \left( P_{+} - i P_{\times} \right) \text{ for } m<0\,.
-%	\end{cases}
-%\ee
-%
-%We now turn to the transformation of~\eqref{eq:defresponse0} and~\eqref{eq:defresponseL} to the Fourier domain. Applying a pure delay as in~\eqref{eq:defresponse0} translates into
-%\be\label{eq:G0}
-%	G_{0}(f, t) = e^{-2i\pi f d_{0}(t)} \,,
-%\ee
-%with $d_{0} = -\hatk \cdot p_{0}$ the delay associated to the orbit around the Sun.
-For the leading-order response~\eqref{eq:transferlocal}, this gives a Fourier-domain transfer function common to all modes, that is a pure phase factor, proportional to the frequency but also $t_{f}$-dependent:
-%\be\label{eq:transfer0local}
-%	\calT_{0}^{\rm local}(f) = G_{0}(f, \tf)\,.
-%\ee
-%If $(\lambda, \beta)$ are the ecliptic longitude and latitude of the source in the sky, and if the orbital phase is set by convention to $0$ at $t=0$, the orbital delay has the simple expression
-%\be\label{eq:delay0}
-%	d_{0}(t) = -R \cos\beta \cos\left(\Omega_{0}t - \lambda\right)\,.
-%\ee
-%
-%For the constellation part of response~\eqref{eq:defresponseL}, treated separately from the delay~\eqref{eq:delay0}, we write
-%\begin{align}\label{eq:decomposeGslr}
-%	F_{slr}^{L}(t) &= \frac{1}{2} \frac{1}{1 - \hatk\cdot n_{l}(t)} n_{l}(t) \cdot P_{\ell m} \cdot n_{l} (t) \,,\nn\\
-%	d_{s}(t) &= - k\cdot p_{s}^{L}(t) \,, \quad d_{r}(t) = - k\cdot p_{r}^{L}(t) \,,\nn\\
-%	G_{slr}^{L}(f,t) &=  F_{slr}^{L}(t) \left( e^{-2i\pi f (d_{s,L}(t) + L)} - e^{-2i\pi f d_{r}(t)} \right) \,.
-%\end{align}
-%The superscript $L$ indicates that the orbital delay~\eqref{eq:delay0} is not included.
-Since we also assume the rigid approximation for the constellation, where the armlengths are fixed, a particular simplification occurs when combining these individual delays, thanks to the relation $p^{L}_{r} - p^{L}_{s} =  L n_{l}$:
-%\begin{align}\label{eq:GslrL}
-%	G_{slr}^{L}(f,t) &= \frac{i \pi f L}{2} \sinc \left[ \pi f L\left(1-\hatk\cdot n_{l} \right) \right] \nn\\
-%	& \quad \cdot \exp\left[ i \pi f \left( L + \hatk\cdot \left( p_{1}^{L} + p_{2}^{L} \right) \right) \right]  n_{l} \cdot P_{\ell m} \cdot n_{l} \,,
-%\end{align}
-%with all time-dependent vectors evaluated at $t$.
-This expression is well known as describing the frequency-dependency in the LISA response~\cite{Larson+99, Cornish01, CR02, RCP04}.
-In the local approximation~\eqref{eq:transferlocal}, the Fourier-domain transfer function then reads
-%\begin{align}\label{eq:transferLlocal}
-%	\calT_{slr}^{L, \mathrm{local}}(f) &= G_{slr}^{L}(f, \tf) \,.
-%\end{align}
-%For plotting purposes, we will also define
-%\be\label{eq:transferLenvelope}
-%	\overline{\calT}_{slr}^{L} (f) = \frac{i \pi f L}{2} n_{l} \cdot P_{\ell m} \cdot n_{l} (\tf)
-%\ee
-%which will serve as an estimate for the enveloppe function of the response, devoid of the zero-crossings at high frequencies of the $\sinc$ term in~\eqref{eq:GslrL}.
-%
-%Note that, if the corrections of Sec.~\ref{subsec:delays}, for non-negligible $\dot d$, are included for the constellation delays, the transfer function will not have this simple form anymore, as $\dot{d}$ will have a different velocity-dependent expression for the sending and receiving spacecraft.
-One must then separately handle $d_{s}$ and $d_{r}$ in~\eqref{eq:decomposeGslr} to compute the corrections.
-%
-%The orbital response~\eqref{eq:transfer0local}-\eqref{eq:delay0} takes a simple analytic form, but the phase contribution of this delay is significant across most of the frequency band and can be large for $f \gg f_{R}$.
-%
-%The constellation response~\eqref{eq:GslrL}-\eqref{eq:transferLlocal} can be interpreted as the Fourier-domain translation of a discrete derivative taken on the waveform.
-The leading factor in~\eqref{eq:GslrL} shows that the amplitude of the response is proportional to $f$ in the low-frequency limit $f\ll f_{L}$, where the other factors are essentially unity.
-For $f\gtrsim f_{L}$, the $\sinc$ and the phase of the exponential generate additional structure in the response, including zero-crossings when the projected armlength is an integer number of wavelengths.
-From~\eqref{eq:transferLlocal}, an expansion for small $f\ll f_{L}$ yields back a Fourier-domain analog of the low-frequency approximation of the response~\cite{Cutler97, RCP04}, which is equivalent to having two LIGO-type interferometers turned by $\pi/4$ and set in motion.
-%
-%For analysis of the response we need a concrete set of gravitational waveforms.
-We will use the PhenomD model~\cite{Khan+15,Husa+15}, which provides Fourier-domain inspiral-merger-ringdown waveforms for aligned spins.
-We refer to App.~\ref{app:precLISA} for a brief discussion of the prospects for applying our formalism for the LISA response to precessing Fourier-domain waveforms.
-%
-%\SM{------------------------------------------------------}
-
-For completeness, we complete in this Appendix the definitions and convention choices that were sketched in Section~\ref{sec:response}.
+For completeness, we give in this Appendix the definitions and convention choices that were sketched in Section~\ref{sec:response} but were not essential for the discussion.
 
@@
 
-While long-lasting signals like SBHBs will see a strong imprint of the LISA orbital motion over the course of observation, SMBHB signals are dominated in SNR by a short-lived burst of emission at merger.
+While long-lasting signals like SBHBs will see a strong imprint of the LISA orbital motion over the course of observation, MBHB signals are dominated in SNR by a short-lived burst of emission at merger.
 For such signals, it will be useful to use a parametrization based on the LISA frame at the time of merger instead of the SSB frame.
@@
 Through $\alpha = \Omega_{0} (t-t_{0}) $, these expressions are time-dependent, reflecting the rotation of this frame to follow LISA on its orbit.
+
+Finally, it is useful to introduce a ``time-at-LISA'' parameter $t_{L}$, taken into account the propagation delay from the SSB to the center of the LISA constellation, as
+\be\label{eq:deftL}
+	t_{L} = t + k\cdot p_{0} \,.
+\ee
+Labeling short-lived MBHB signals by their time of arrival at the SSB is a bad parametrization.
+For a given measured arrival time at the LISA instrument $t_{L}$, different sky positions will give separate peaks in the ``time-at-SSB'' variable $t$.
+The use of $t_{L}$ eliminates this issue.
 

### 2020-02-29 2020-02-29_submission/paper_PE.tex
arXiv submission

@@
   \centering
-  \includegraphics[width=.99\linewidth]{../plots/plot_hc_tdia_hm.pdf}
+  \includegraphics[width=.99\linewidth]{plot_hc_tdia_hm.pdf}
   \caption{Characteristic strain signal~\eqref{eq:defhc}, shown here for the TDI channel $A$ of System II, decomposed as a set of $(\ell,m)$ contributions from the modes available in the waveform model.
@@
   \centering
-  \includegraphics[width=.98\linewidth]{../plots/plot_smbh9_tdi.pdf}
+  \includegraphics[width=.98\linewidth]{plot_smbh9_tdi.pdf}
   \caption{Example of instrument response for System II in Table~\ref{tab:MBHBparams}.
@@
 \begin{figure}
-   \includegraphics[width=.99\linewidth]{../plots/tflm_smbh.pdf}
+   \includegraphics[width=.99\linewidth]{tflm_smbh.pdf}
   \caption{Relation between time and frequency for MBHB systems I and II with $M=2\times 10^6 \; M_{\odot}$, $q=3$ and $z=4$.
@@
   \centering
-  \includegraphics[width=.98\linewidth]{../plots/overlapmodebymode_maxfscaledlm_t.pdf}
+  \includegraphics[width=.98\linewidth]{overlapmodebymode_maxfscaledlm_t.pdf}
   \caption{Cumulative contribution to $\mathrm{SNR}^{2}$ as defined in~\eqref{eq:defcontribSNR2} for system II.
@@
   \centering
-  \includegraphics[width=.98\linewidth]{../plots/responseapprox_tdia_22.pdf}
+  \includegraphics[width=.98\linewidth]{responseapprox_tdia_22.pdf}
   \caption{Real and imaginary part of the transfer function $\calT^{22}_{h_{a}}$~\eqref{eq:defhaet} in different response approximations.
@@
   \centering
-  \includegraphics[width=.98\linewidth]{../plots/corner_smbh_case0_ptmcmc_22_hm_withzoom.png}
+  \includegraphics[width=.98\linewidth]{corner_smbh_case0_ptmcmc_22_hm_withzoom.png}
   \caption{Inferred parameter posterior distribution for the MBHB system I of Table~\ref{tab:MBHBparams}, with $1$-$2$-$3$-$\sigma$ contours and the black cross indicating the true parameters.
@@
   \centering
-  \includegraphics[width=.98\linewidth]{../plots/corner_smbh_case9_ptmcmc_22_hm_withzoom.png}
+  \includegraphics[width=.98\linewidth]{corner_smbh_case9_ptmcmc_22_hm_withzoom.png}
   \caption{Same as Fig.~\ref{fig:PEsmbh22hmCase0}, for the MBHB system II of Table~\ref{tab:MBHBparams}.
@@
   \centering
-  \includegraphics[width=.98\linewidth]{../plots/row_m1m2_ptmcmc_bambi.png}
+  \includegraphics[width=.98\linewidth]{row_m1m2_ptmcmc_bambi.png}
   \caption{Comparisons between inferred posteriors obtained with our two different Bayesian samplers described in~\ref{subsec:samplers} and with the Fisher matrix approximation described in~\ref{subsec:Fisher}.
@@
   \centering
-  \includegraphics[width=.98\linewidth]{../plots/row_Dinc_ptmcmc_bambi.png}
+  \includegraphics[width=.98\linewidth]{row_Dinc_ptmcmc_bambi.png}
   \caption{Same comparisons as in Fig.~\ref{fig:PEsmbhm1m2ptmcmcbambi}, for the distance and inclination parameters, zooming on the relevant regions.}
@@
   \centering
-  \includegraphics[width=.98\linewidth]{../plots/row_sky_ptmcmc_bambi.png}
+  \includegraphics[width=.98\linewidth]{row_sky_ptmcmc_bambi.png}
   \caption{Same comparisons as in Fig.~\ref{fig:PEsmbhm1m2ptmcmcbambi}, for the sky position parameters, zooming on the relevant regions for inclination.
@@
   \begin{minipage}{.49\linewidth}
-  \includegraphics[width=.99\linewidth]{../plots/corner_skydown_smbh_case9_22_ptmcmc_bambi_withdegen.png}
+  \includegraphics[width=.99\linewidth]{corner_skydown_smbh_case9_22_ptmcmc_bambi_withdegen.png}
   \end{minipage}
   \begin{minipage}{.49\linewidth}
-  \includegraphics[width=.99\linewidth]{../plots/corner_skydown_smbh_case9_simplelike_ptmcmc_bambi_withdegen.png}
+  \includegraphics[width=.99\linewidth]{corner_skydown_smbh_case9_simplelike_ptmcmc_bambi_withdegen.png}
   \end{minipage}
@@
 %  \begin{minipage}{.48\linewidth}
-%      \includegraphics[width=.99\linewidth]{../plots/t-of-F.pdf}
+%      \includegraphics[width=.99\linewidth]{t-of-F.pdf}
 %   \end{minipage}
 %   \begin{minipage}{.48\linewidth}
-%      \includegraphics[width=.99\linewidth]{../plots/t-of-SNR.pdf}
+%      \includegraphics[width=.99\linewidth]{t-of-SNR.pdf}
 %   \end{minipage}
@@
   \begin{minipage}{.32\linewidth}
-      \includegraphics[width=.99\linewidth]{../plots/corner_smbh_case9_hm_tseries_m1m2.png}
+      \includegraphics[width=.99\linewidth]{corner_smbh_case9_hm_tseries_m1m2.png}
    \end{minipage}
    \begin{minipage}{.32\linewidth}
-      \includegraphics[width=.99\linewidth]{../plots/corner_smbh_case9_hm_tseries_Dinc.png}
+      \includegraphics[width=.99\linewidth]{corner_smbh_case9_hm_tseries_Dinc.png}
    \end{minipage}
    \begin{minipage}{.32\linewidth}
-      \includegraphics[width=.99\linewidth]{../plots/corner_smbh_case9_hm_tseries_tphi.png}
+      \includegraphics[width=.99\linewidth]{corner_smbh_case9_hm_tseries_tphi.png}
    \end{minipage}
    \begin{minipage}{.32\linewidth}
-      \includegraphics[width=.99\linewidth]{../plots/corner_smbh_case9_hm_tseries_m1m2_zoom.png}
+      \includegraphics[width=.99\linewidth]{corner_smbh_case9_hm_tseries_m1m2_zoom.png}
    \end{minipage}
    \begin{minipage}{.32\linewidth}
-      \includegraphics[width=.99\linewidth]{../plots/corner_smbh_case9_hm_tseries_Dinc_zoom.png}
+      \includegraphics[width=.99\linewidth]{corner_smbh_case9_hm_tseries_Dinc_zoom.png}
    \end{minipage}
    \begin{minipage}{.32\linewidth}
-      \includegraphics[width=.99\linewidth]{../plots/corner_smbh_case9_hm_tseries_tphi_zoom.png}
+      \includegraphics[width=.99\linewidth]{corner_smbh_case9_hm_tseries_tphi_zoom.png}
    \end{minipage}
@@
   \centering
-  \includegraphics[width=.8\linewidth]{../plots/skymap_Lframe_tseries_bambi_22.png}
-  \includegraphics[width=.8\linewidth]{../plots/skymap_Lframe_tseries_ptmcmc_hm.png}
+  \includegraphics[width=.8\linewidth]{skymap_Lframe_tseries_bambi_22.png}
+  \includegraphics[width=.8\linewidth]{skymap_Lframe_tseries_ptmcmc_hm.png}
   \caption{Refinement of the LISA-frame sky position inference over time for system II.
@@
   \centering
-  \includegraphics[width=.99\linewidth]{../plots/lnL_skymodes_approx_22.pdf}
-  \includegraphics[width=.99\linewidth]{../plots/lnL_skymodes_approx_hm.pdf}
+  \includegraphics[width=.99\linewidth]{lnL_skymodes_approx_22.pdf}
+  \includegraphics[width=.99\linewidth]{lnL_skymodes_approx_hm.pdf}
   \caption{Log-likelihood~\eqref{eq:deflnLaet} at degenerate modes in the sky summarized in Table~\ref{tab:MBHBdegen}, for different response approximations (from left to right: \textit{Full}, \textit{Frozen}, \textit{Low-$f$}), with 22-mode only (top) and with higher harmonics (bottom).
@@
   \centering
-  \includegraphics[width=.98\linewidth]{../plots/plot_sobh1_tdi.pdf}
+  \includegraphics[width=.98\linewidth]{plot_sobh1_tdi.pdf}
   \caption{Example of instrument response for the SBHB system I in Table~\ref{tab:SBHBparams}.
@@
 \begin{figure*}
-	\includegraphics[width=0.9\textwidth]{../plots/corner_sobh_tdc1_extr_ptmcmc_bambi}
+	\includegraphics[width=0.9\textwidth]{corner_sobh_tdc1_extr_ptmcmc_bambi}
 	\caption{Inferred parameter posterior distribution for the SBHB system I of Table~\ref{tab:SBHBparams}, with $1$-$2$-$3$-$\sigma$ contours and the black cross indicating the true parameters.
@@
 \begin{figure*}
-	\includegraphics[width=0.9\textwidth]{../plots/corner_sobh_tdc2_extr_ptmcmc_bambi}
+	\includegraphics[width=0.9\textwidth]{corner_sobh_tdc2_extr_ptmcmc_bambi}
 	\caption{Same as Fig.~\ref{fig:sbhb_corner_1}, for the SBHB system II of Table~\ref{tab:SBHBparams}.}
@@
 %  \centering
-%  \includegraphics[width=.99\linewidth]{../plots/plot_MM_ROM.pdf}
+%  \includegraphics[width=.99\linewidth]{plot_MM_ROM.pdf}
 %  \caption{Unfaithfulness of ROM waveforms compared to the original EOBNRv2HM waveforms, computed as in~\eqref{}.
@@
   \centering
-  \includegraphics[width=.9\linewidth]{../plots/corner_smbh_case9_simplelike_ptmcmc_bambi.png}
+  \includegraphics[width=.9\linewidth]{corner_smbh_case9_simplelike_ptmcmc_bambi.png}
   \caption{Posterior distributions obtained with the two samplers \texttt{ptmcmc} (blue) and $\texttt{multinest}$ (green) for the simplified likelihood~\eqref{eq:simplelikelihoodhm}, using only the dominant mode $(\ell, m) = (2,2)$.
@@
 %  \begin{minipage}{.32\linewidth}
-%      \includegraphics[width=.99\linewidth]{../plots/corner_smbh_case9_hm_dseries_m1m2.png}
+%      \includegraphics[width=.99\linewidth]{corner_smbh_case9_hm_dseries_m1m2.png}
 %   \end{minipage}
 %   \begin{minipage}{.32\linewidth}
-%      \includegraphics[width=.99\linewidth]{../plots/corner_smbh_case9_hm_dseries_Dinc.png}
+%      \includegraphics[width=.99\linewidth]{corner_smbh_case9_hm_dseries_Dinc.png}
 %   \end{minipage}
 %   \begin{minipage}{.32\linewidth}
-%      \includegraphics[width=.99\linewidth]{../plots/corner_smbh_case9_hm_dseries_tphi.png}
+%      \includegraphics[width=.99\linewidth]{corner_smbh_case9_hm_dseries_tphi.png}
 %   \end{minipage}
@@
 %  \centering
-%  \includegraphics[width=.8\linewidth]{../plots/plot_mollweide_dseries.png}
+%  \includegraphics[width=.8\linewidth]{plot_mollweide_dseries.png}
 %  \caption{.}

### 2021-02-03 resubmission/reply_referee.tex
Added resubmission

\documentclass[11pt]{letter}

\usepackage{amsmath}
\usepackage{eufrak}

\begin{document}

\textbf{Reply to the Referee report on DQ12326/Marsat}

We would like to thank the anonymous referee for this report and apologize for the delay in implementing the corrections. We describe below the steps taken to address the referee's remarks, as well as other minor corrections.

\begin{itemize}
	\item {\it For the supermassive BH signals, the inferences in this
paper use a waveform model with no spins. (Described in
sec III B). This very significant simplification; it greatly
simplifies the modelling (e.g. the ROM only interpolates in 1D)
and may obscure some parameter degeneracies. I ask the authors
to consider emphasising this point more prominently in the
introduction and/or abstract. The other key simplification
(namely the accelerated likelihood calculation of a sparse grid
of frequencies) is already clearly pointed out in the abstract.}
\end{itemize}

We emphasized more this point in the abstract and introduction. It should be noted that the inclusion of aligned spins, although important quantitatively (especially for merger-dominated signals), is not expected to change the qualitative picture of degeneracies in the set of extrinsic parameters (as we saw in further studies), due to the weak correlations between intrinsic and extrinsic parameters. Misaligned spins causing precession, on the other hand, could change qualitatively the inference and also affect extrinsic parameters, since they change the mode content of the waveform and its structure. To our knowledge, this remains to be explored. We added a few sentences in the introduction to reflect this.

Edit in the abstract: {\it We present the first simulations of Bayesian inference for the parameters of massive black hole systems including consistently the merger and ringdown of the signal, as well as higher harmonics, albeit neglecting spins.}

Edit in the introduction: {\it In the present study, we will consider a somewhat reduced parameter space, not including the effects of the spins of the two black holes. These could lead to more complicated waveforms, notably through precession effects, which we leave for future work. It should be noted, however, that aligned spin components, though quantitatively important, are not expected to change many of our conclusions due to the relative disconnect between the inference of intrinsic and extrinsic parameters.}

\begin{itemize}
	\item {\it Eq. 20 describes an approximation to the transfer function
(single link redshift observables) based on a separation of
timescales. I would find useful if this approximation could
be explained in more detail in the text. For example,
what are the different timescales involved?}
\end{itemize}

We expanded around the reference to [16], where this is discussed in extensive details, and clarified this point in the text.

Edit above eq. (20) : {\it How to take the Fourier transform of~\eqref{eq:defyslr} was investigated in details in the perturbative formalism of~\cite{MB18}. The main timescales to compare are the radiation-reaction timescale $\propto 1/\sqrt{\dot{\omega}}$ of the binary and the LISA orbital timescale $1\mathrm{yr}$. At  leading order in the separation of these timescales, for a signal that is chirping fast enough, we have simply}

\begin{itemize}
	\item {\it When discussing degeneracies in the sky. Is it possible to
compare the match between the injected signal and the best
fitting signal from the other side of the sky? Currently in
this is discussed in section VIB where the authors use
log-likelihood values to quantify the importance of the second
mode. I think using match values might be more familiar to
many in the community.}
\end{itemize}

The reason we chose a likelihood measure here is that it is both easy to compute and easy to interpret in terms of probability (density). Because of the scaling of likelihoods with $\mathrm{SNR}^{2}$, the meaning of a given mismatch figure depends on the SNR of the signal, and for LISA this can be in the thousands (vs. few tens for LIGO/Virgo). Mismatches are also usually computed by optimizing over time, phase and polarization, which we are not doing with our estimates. We did not compute non-optimized mismatches but added a discussion in the text to emphasize these points.

Edit in Sec. VI B: {\it We choose here to show a likelihood measure instead of the perhaps more familiar mismatch between signals. Mismatches are by definition normalized and therefore SNR-insensitive, while the log-likelihood scales with $\mathrm{SNR}^{2}$. Since LISA could observe SNR values in the hundreds or thousands, intuition based on previous familiarity with mismatch measures used in the LIGO/Virgo context might be lost.}

\begin{itemize}
	\item {\it In sec III E: "In this process we also impose the
constrain...", constraint?.}
\end{itemize}

Thanks for noticing this, typo corrected.

\textbf{Other changes}

There was a typo in eqs. (67)-(68), with the real part being taken in the wrong place. The $\langle \ell m | \ell' m'\rangle$ mode overlaps are defined as complex quantities. This does not affect our results.

We added $m>0$ in eq. (13).

\end{document}

### 2021-02-03 resubmission/2021-02-03_resubmission/paper_PE.tex
Added resubmission

--- 2020-02-29 2020-02-29_submission/paper_PE.tex
@@
 \affiliation{APC, AstroParticule et Cosmologie, Universit\'{e} Paris Diderot, CNRS/IN2P3, CEA/Irfu, Observatoire de Paris, Sorbonne Paris Cit\'{e}, 10, rue Alice Domon et L\'{e}onie Duquet 75205 PARIS Cedex 13, France}
+\affiliation{Max Planck Institute for Gravitational Physics (Albert Einstein Institute),
+Am M\"uhlenberg 1, Potsdam-Golm, 14476, Germany}
 \author{John G. Baker}
@@
 
+In the present study, we will consider a somewhat reduced parameter space, not including the effects of the spins of the two black holes.
+These could lead to more complicated waveforms, notably through precession effects, which we leave for future work.
+It should be noted, however, that aligned spin components, though quantitatively important, are not expected to change many of our conclusions due to the relative disconnect between the inference of intrinsic and extrinsic parameters.
+
 We start by introducing a fast method to calculate the likelihood for a black hole merger.
@@
 \be\label{eq:Hsummodes}
-	H = \sum_{\ell, m} P_{\ell m} h_{\ell m} \,.
+	H = \sum_{\ell, m > 0} P_{\ell m} h_{\ell m} \,.
 \ee
@@
 \ee
-Applying the perturbative formalism of~\cite{MB18} to leading order in the separation of timescales, we have simply
+How to take the Fourier transform of~\eqref{eq:defyslr} was investigated in details in the perturbative formalism of~\cite{MB18}.
+The main timescales to compare are the radiation-reaction timescale $\propto 1/\sqrt{\dot{\omega}}$ of the binary and the LISA orbital timescale $1\mathrm{yr}$.
+At  leading order in the separation of these timescales, for a signal that is chirping fast enough, we have simply
 \be\label{eq:Tlmslr}
@@
 To achieve this, we begin with an initial choice of $\epsilon_i$ scaled off the parameter values or prior widths, then we estimate the Fisher diagonals and iterate until convergence.
-In this process we also impose the constrain that the step is never larger than $10^{-10}$ times the initial scaling.
+In this process we also impose the constraint that the step is never larger than $10^{-10}$ times the initial scaling.
 
@@
 \ee
-Having fixed the intrinsic parameters and the time, in~\eqref{eq:transferlowfae} the modes $h_{\ell m}$ are fixed as well, and it is convenient to introduce the notation
+Having fixed the intrinsic parameters and the time, in~\eqref{eq:transferlowfae} the modes $h_{\ell m}$ are fixed as well, and it is convenient to introduce the notation (note that these quantities are defined as complex)
 \be\label{eq:innerproductlmlpmp}
-	\left\langle \ell m | \ell' m' \right\rangle = 4\mathrm{Re} \int \frac{df}{S_{n}^{a,e}} \left( 6 \pi f L\right)^{2} \tilde{h}_{\ell m} \tilde{h}_{\ell' m'}^{*} \,.
+	\left\langle \ell m | \ell' m' \right\rangle = 4 \int \frac{df}{S_{n}^{a,e}} \left( 6 \pi f L\right)^{2} \tilde{h}_{\ell m} \tilde{h}_{\ell' m'}^{*} \,.
 \ee
@@
 \begin{align}\label{eq:simplelikelihoodhm}
-	\ln \calL &= -\frac{1}{2} \sum_{\ell m} \sum_{\ell' m'} \left[ \left( \vphantom{s_{a}^{\ell' m'} } s_{a}^{\ell m} - s_{a, \mathrm{inj}}^{\ell m}\right) \left( s_{a}^{\ell' m'} - s_{a, \mathrm{inj}}^{\ell' m'}\right)^{*} \right.
+	\ln \calL &= -\frac{1}{2} \mathrm{Re} \sum_{\ell m} \sum_{\ell' m'} \left[ \left( \vphantom{s_{a}^{\ell' m'} } s_{a}^{\ell m} - s_{a, \mathrm{inj}}^{\ell m}\right) \left( s_{a}^{\ell' m'} - s_{a, \mathrm{inj}}^{\ell' m'}\right)^{*} \right.
 \nn\\
@@
 \begin{align}
-	s_{a,e}^{22} &= \frac{1}{4d} \sqrt{\frac{5}{\pi}} \cos^{4}\frac{\iota}{2} e^{2i(\varphi-\psiL)} \left( F_{a,e}^{+} + i F_{a,e}^{\times} \right) \\
+	s_{a,e}^{22} &= \frac{1}{4d} \sqrt{\frac{5}{\pi}} \cos^{4}\frac{\iota}{2} e^{2i(\varphi-\psiL)} \left( F_{a,e}^{+} + i F_{a,e}^{\times} \right) \nn \\
 	&+ \frac{1}{4d} \sqrt{\frac{5}{\pi}} \sin^{4}\frac{\iota}{2} e^{2i(\varphi+\psiL)} \left( F_{a,e}^{+} - i F_{a,e}^{\times} \right) \,,
@@
 
+We choose here to show a likelihood measure instead of the perhaps more familiar mismatch between signals.
+Mismatches are by definition normalized and therefore SNR-insensitive, while the log-likelihood scales with $\mathrm{SNR}^{2}$.
+Since LISA could observe SNR values in the hundreds or thousands, intuition based on previous familiarity with mismatch measures used in the LIGO/Virgo context might be lost.
+
 With the \textit{Low-$f$} response, the \textit{antipodal} mode remains almost exactly degenerate with the injection, differing only by the Doppler phase~\eqref{eq:defPhiR}, which has a small effect on our short signals as shown in Fig.~\ref{fig:tranferSMBHCase9}.
@@
 We are indebted to Philip Graff for his help with the \texttt{bambi} sampler.
+This study benefited strongly from the use of the HPC clusters Hypatia and Minerva at the Max Planck Institute for Gravitational Physics.
 TDC was supported by an appointment to the NASA Postdoctoral Program at the Goddard Space Flight Center, administered by Universities Space Research Association under contract with NASA, during part of this work.

### 2021-02-09 resubmission/2021-02-03_resubmission/paper_PE.tex
Add missing definition of \Omega_0; add Makefile

@@
 \affiliation{Gravitational Astrophysics Laboratory, NASA Goddard Space Flight Center, 8800 Greenbelt Rd., Greenbelt, MD 20771, USA}
-\author{Tito Dal Canton}
+\author[0000-0001-5078-9044]{Tito Dal Canton}
 \thanks{NASA Postdoctoral Program fellow}
@@
 The orbits of the three spacecrafts around the Sun can be chosen so that, at leading order in the eccentricity of the orbits, the constellation retains the shape of an equilateral triangle in a cartwheeling motion following the Earth orbit.
-We denote by $a$ the semi-major axis, $e$ the eccentricity and pose $\alpha = \Omega_{0} (t-t_{0}))$, with $t_{0}$ a reference time for the initial position ($t_{0} = 0$ in our case).
+We denote by $a$ the semi-major axis, $e$ the eccentricity and pose $\alpha = \Omega_{0} (t-t_{0})$, with $t_{0}$ a reference time for the initial position ($t_{0} = 0$ in our case) and $\Omega_{0} = 2\pi / (1\ \mathrm{yr})$.
 The reference SSB frame is $(\hat{x}, \hat{y}, \hat{z})$.
